---
title: "EP playground – 06 categ dummy"
output: 
  html_notebook:
    toc: true
---

```{r setup}
library(tidyverse)
library(patchwork)
library(emmeans)
library(simglm)

theme_update(base_size = 5)
```


# Simulate data

```{r message=FALSE, warning=FALSE}
set.seed(3119) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation + study + method,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1),
               study = list(var_type = 'factor', 
                            levels = c('alone', 'others'),
                            prob = c(0.53, 0.47)),
               method = list(var_type = 'factor', 
                            levels = c('read', 'summarise', 'self-test'),
                            prob = c(0.3, 0.4, 0.3))),
  error = list(variance = 20),
  sample_size = 250,
  reg_weights = c(0.6, 1.4, 1.5, 6, 6, 2)
)

df3 <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study3 <- df3 %>%
  dplyr::select(y, hours, motivation, study, method) %>%
  mutate(
    ID = paste("ID", 101:350, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2),
    study = factor(study),
    method = factor(method)
  ) %>%
  dplyr::select(ID, score, hours, motivation, study, method)

```


# `study` (two levels)

## Group means

```{r}
test_study3 |>
  group_by(study) |>
  summarise(
    avg_score = round(mean(score), 2)
  )

mean_alone <- filter(test_study3, study == 'alone')$score |> mean()
mean_others <- filter(test_study3, study == 'others')$score |> mean()
```


## Plot `study` data

```{r}
test_study3 |>
  ggplot(aes(x = study, y = score, fill = study, colour = study)) +
  geom_violin(alpha = 0.5) +
  geom_sina(alpha = 0.5) +
  theme(legend.position = 'none') +
  stat_summary(colour = 'black', fun = mean, geom = 'point', size = 2) +
  geom_segment(aes(x = 'alone', xend = 'others', y = mean_alone, yend = mean_others), colour = 'black') +
  NULL
```

Plot on xy plane, dummy-coded.

```{r}
contrasts(test_study3$study)
```


`alone` = ref level.

```{r}
xlim_lower <- -2.2
xlim_upper <-  2.2
ylim_lower <- -25
ylim_upper <-  55

test_study3 |>
  mutate(study_num = ifelse(study == 'alone', 0, 1)) |>
  ggplot(aes(x = study_num, y = score, fill = study, colour = study)) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_x_continuous(limits = c(xlim_lower, xlim_upper), expand = c(0, 0)) +
  scale_y_continuous(limits = c(ylim_lower, ylim_upper), expand = c(0, 0)) +
  geom_segment(aes(x = xlim_lower, xend = xlim_upper, y = 0, yend = 0), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 0, y = ylim_lower, yend = ylim_upper), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_alone, yend = mean_others), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_alone, yend = mean_alone), colour = 'red', linewidth = 2) +
  geom_segment(aes(x = 1, xend = 1, y = mean_alone, yend = mean_others), colour = 'red', linewidth = 2) +
  NULL
```


Add note that the manual stuff is so you understand how coding works.
But you'll never have to do it this laboriously.

Exact same outcome:

- manual ifelse
- `contr.treatment()`




## Model `score ~ study`

```{r}
mod1 <- lm(score ~ study, data = test_study3)
summary(mod1)
```

### t-test

```{r}
t.test(score ~ study, var.equal = T, data = test_study3)
```

## emmeans

```{r}
(mod1_emm <- emmeans(mod1, ~study))
```


```{r}
(p_mod1_emm <- mod1_emm |> plot() +
  coord_flip())
```

### overlay emmeans on data plot

```{r}
mod1_emm_tib <- mod1_emm |>
  as_tibble() |>
  rename(score = emmean)
```

```{r}
test_study3 |>
  ggplot(aes(x = study, y = score, fill = study, colour = study)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  theme(legend.position = 'none') +
  geom_errorbar(data = mod1_emm_tib, aes(ymin = lower.CL, ymax = upper.CL), colour = 'black', width = 0.2) +
  geom_point(data = mod1_emm_tib, colour = 'black', size = 3) +
  NULL
```


### pairwise comparisons

```{r}
levels(test_study3$study)
```

For each comparison we want to make, assign 1 to the level we're interested in, and assign -1 to the level you want to compare it to.
If there are any other levels in there, assign them 0.

```{r}
mod1_comparisons <- list(
  'alone vs. others' = c(
    1,   # alone
    -1   # others
  )
)

contrast(
  object = mod1_emm,
  method = mod1_comparisons
)
```

Basically reconstructs the model's estimates.
Not so interesting.


## Change ref level: coefs are diff but emmeans are same

Original contrasts:

```{r}
contrasts(test_study3$study)
```

Now, let's flip it:

```{r}
contrasts(test_study3$study) <- contr.treatment(
  levels(test_study3$study), 
  base = 2)
contrasts(test_study3$study)
```

### plot flipped data

```{r}
xlim_lower <- -2.2
xlim_upper <-  2.2
ylim_lower <- -25
ylim_upper <-  55

test_study3 |>
  mutate(study_num = ifelse(study == 'alone', 1, 0)) |>
  ggplot(aes(x = study_num, y = score, fill = study, colour = study)) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_x_continuous(limits = c(xlim_lower, xlim_upper), expand = c(0, 0)) +
  scale_y_continuous(limits = c(ylim_lower, ylim_upper), expand = c(0, 0)) +
  geom_segment(aes(x = xlim_lower, xend = xlim_upper, y = 0, yend = 0), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 0, y = ylim_lower, yend = ylim_upper), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_others, yend = mean_alone), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_others, yend = mean_others), colour = 'red', linewidth = 2) +
  geom_segment(aes(x = 1, xend = 1, y = mean_others, yend = mean_alone), colour = 'red', linewidth = 2) +
  NULL
```

### fit model (different)

```{r}
mod1b <- lm(score ~ study, data = test_study3)
coef(mod1b)
```

- What is intercept? (mean when study = others)
- What is studyalone? (diff between levels: mean score when study = alone minus mean score when study = others)

Compare this to the original coefs from `mod1`:

```{r}
coef(mod1)
```

- What is intercept? (mean when study = alone)
- What is studyothers? (diff between levels: mean score when study = others minus mean score when study = alone)


### get emmeans (same)

```{r}
(mod1b_emm <- emmeans(mod1b, ~study))
```

```{r}
p_mod1b_emm <- mod1b_emm |> plot() +
  coord_flip() +
  ggtitle('With reference level = others')

p_mod1_emm <- p_mod1_emm + ggtitle('With reference level = alone')

p_mod1b_emm + p_mod1_emm
```

Identical!


# `method` (three levels)

## Group means

```{r}
test_study3 |>
  group_by(method) |>
  summarise(
    avg_score = round(mean(score), 2)
  )

mean_read <- filter(test_study3, method == 'read')$score |> mean()
mean_self <- filter(test_study3, method == 'self-test')$score |> mean()
mean_summ <- filter(test_study3, method == 'summarise')$score |> mean()
```


## Plot data

```{r}
test_study3 |>
  ggplot(aes(x = method, y = score, fill = method, colour = method)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  theme(legend.position = 'none') +
  NULL
```

Goal: Compare the means of each group to one another.
That was easy when we had only two groups: one straight line can go from mean to mean.
But now one straight line won't cut it anymore.

The smallest number of lines that will connect all the means = 2.
Instead of estimating one line, we will estimate two.

And in general, if we have $n$ groups, we will be estimating $n-1$ lines.

The lines that we're going to be estimating now:

```{r}
test_study3 |>
  ggplot(aes(x = method, y = score, fill = method, colour = method)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  theme(legend.position = 'none') +
  stat_summary(colour = 'black', fun = mean, geom = 'point', size = 2) +
  # line from read to self-test:
  geom_segment(colour = 'black', aes(x = 'read', xend = 'self-test', y = mean_read, yend = mean_self)) +
  # line from read to summarise:
  geom_segment(colour = 'black', aes(x = 'read', xend = 'summarise', y = mean_read, yend = mean_summ)) +
  NULL
```


But I'm lying a little bit.
This looks like going to go from 0 to 1 in one line, and 0 to 2 in the other line.
But that's not true -- this is just the intuition about what groups we are comparing.
Really what we're going to do is this:


## Plot dummy-coded `method` in xy space

Two panels: the first will be the difference between `read` and `self-test` (the first variable), the second will be difference between `read` and `summarise` (the second variable).

```{r fig.width = 10, fig.height = 4}
xlim_lower <- -2.2
xlim_upper <-  2.2
ylim_lower <- -25
ylim_upper <-  55

p1 <- test_study3 |> 
  filter(method %in% c('read', 'self-test')) |>
  mutate(method_num = ifelse(method == 'read', 0, 1)) |>
  ggplot(aes(x = method_num, y = score, fill = method, colour = method)) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_x_continuous(limits = c(xlim_lower, xlim_upper), expand = c(0, 0)) +
  scale_y_continuous(limits = c(ylim_lower, ylim_upper), expand = c(0, 0)) +
  geom_segment(aes(x = xlim_lower, xend = xlim_upper, y = 0, yend = 0), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 0, y = ylim_lower, yend = ylim_upper), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_read, yend = mean_self), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_read, yend = mean_read), colour = 'red', linewidth = 2) +
  geom_segment(aes(x = 1, xend = 1, y = mean_read, yend = mean_self), colour = 'red', linewidth = 2) +
  labs(
    title = 'First line: self-test vs. read'
  ) +
  NULL

p2 <- test_study3 |> 
  filter(method %in% c('read', 'summarise')) |>
  mutate(method_num = ifelse(method == 'read', 0, 1)) |>
  ggplot(aes(x = method_num, y = score, fill = method, colour = method)) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_x_continuous(limits = c(xlim_lower, xlim_upper), expand = c(0, 0)) +
  scale_y_continuous(limits = c(ylim_lower, ylim_upper), expand = c(0, 0)) +
  geom_segment(aes(x = xlim_lower, xend = xlim_upper, y = 0, yend = 0), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 0, y = ylim_lower, yend = ylim_upper), arrow = arrow(ends = 'both', length = unit(12, 'pt')), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_read, yend = mean_summ), colour = 'black') +
  geom_segment(aes(x = 0, xend = 1, y = mean_read, yend = mean_read), colour = 'red', linewidth = 2) +
  geom_segment(aes(x = 1, xend = 1, y = mean_read, yend = mean_summ), colour = 'red', linewidth = 2) +
  labs(
    title = 'Second line: summarise vs. read'
  ) +  
  NULL

p1 + p2
```


## Dummy-code `method` variable

Manual dummy-coding:

```{r}
dummyCoded <- test_study3 %>%
  select(ID, score, method) %>%
  mutate(
    method1 = ifelse(method == "self-test", 1, 0),
    method2 = ifelse(method == "summarise", 1, 0))

dummyCoded
```


```{r}
contrasts(test_study3$method)
```




## Model `score ~ method`

```{r}
mod2 <- lm(score ~ method, data = dummyCoded)
summary(mod2)
```


### t-tests

(Nearly) equivalent t-tests (different bc diff degrees of freedom, bc filtering dataset. but close).

```{r}
dC1 <- test_study3 |> 
  filter(method %in% c('read', 'self-test')) |>
  mutate(method = factor(method, levels = c('read', 'self-test')))
contrasts(dC1$method) <- contr.treatment(2)
contrasts(dC1$method)

t.test(score ~ method, var.equal = T, data = dC1)
```


```{r}
dC2 <- test_study3 |> 
  filter(method %in% c('read', 'summarise')) |>
  mutate(method = factor(method, levels = c('read', 'summarise')))
contrasts(dC2$method) <- contr.treatment(2)
contrasts(dC2$method)

t.test(score ~ method, var.equal = T, data = dC2)
```


## emmeans 

Before, I said that the smallest number of lines we need to compare three groups is two.
But this actually means that one of the possible comparisons is left out.

(??? above the missing line between self-test and summarise)

Options:

- make something else ref level. Then we get that comparison. But then we're missing a different third comparison.
- get emmeans, and once we're in the outcome space, we can compute differences between whatever we want! yay!


```{r}
(mod2_emm <- emmeans(mod2, ~method))
```


```{r}
mod2_emm |> plot() +
  coord_flip()
```


### overlay emmeans on data plot

```{r}
mod2_emm_tib <- mod2_emm |>
  as_tibble() |>
  rename(score = emmean)
```

```{r}
test_study3 |>
  ggplot(aes(x = method, y = score, fill = method, colour = method)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, width = 0.2) +
  theme(legend.position = 'none') +
  geom_errorbar(data = mod2_emm_tib, aes(ymin = lower.CL, ymax = upper.CL), colour = 'black', width = 0.2) +
  geom_point(data = mod2_emm_tib, colour = 'black', size = 3) +
  NULL
```


### pairwise comparisons

```{r}
levels(test_study3$method)
```


Assign 1 to the level we are interested in, and -1 to the baseline we want to compare it to.

```{r}
mod2_comparisons <- list(
  'Self-test vs. Read' = c(-1, 1, 0),  # read self-test summarise, in that order
  'Summarise vs. Read' = c(-1, 0, 1),
  'Self-test vs. Summarise' = c(0, 1, -1),
  'mean(Self-test, summarise) vs. Read' = c(-1, 0.5, 0.5)  # weights must sum to 0 -- pools self-test and summarise tog and pits that mean against mean of Read
)
```


1 - 1 

`mean(Self-test) - mean(Read)` = positive, bc self-test is larger than read

```{r}
(mod2_comparisons_test <- contrast(mod2_emm, mod2_comparisons))
```

`contrast()` gives us the p-values for each difference.
We can also get the 95% CIs of each difference by giving the outcome of `contrast()` into `confint()`.

```{r}
confint(mod2_comparisons_test)
```


