---
title: "<b> Interactions 3 </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(ggmosaic)
library(kableExtra)
library(performance)
library(interactions)


set.seed(19981)
dept <- rbinom(100, 1, 0.4)
location <- rbinom(100, 1, 0.35)
serv <- rnorm(100, 0, 1.4)
serv <- serv+3.5
dl <- dept*location

omit <- rbinom(100, 1, .3)

b0 = 20
b1 = 3.1
b2 = -4.5
b3 = -4.2
b4 = -6.5
b5 = 2

eps <- rnorm(100, 0, 4)

salary <- b0 + (b1*serv) + (b2*dept) + (b3*location) + (b4*dl) + (b5*omit) + eps
salary <- salary+20

salary3 <- tibble(
  ID = paste("ID", 101:200, sep=""),
  salary = salary,
  department = factor(dept,labels = c("Accounts", "Manager")),
  location = factor(location, labels = c("London", "Birmingham")),
  serv = serv
)

test <- lm(salary ~ department*location, salary3)
```

# Weeks Learning Objectives
1. Interpret interactions between two categorical variables (dummy coding).

2. Visualize and probe interactions.

3. Be able to read interaction plots.

---
# Topics for today

+ Categorical-categorical interaction.
  + Begin with 2x2 (two binary variables).
  + Move on to look at examples with more levels.

+ Visualising and understanding categorical interactions.


---
#  General definition 

+ When the effects of one predictor on the outcome differ across levels of another predictor.

+ Categorical*categorical interaction:
	+ There is a difference in the differences between groups across levels of a second factor.

+ This idea of a difference in differences can be quite tricky to think about.
  + So we will start with some visualisation, and then look at two examples.

---
# Visualizing interactions
+ In our class example we have been looking at predicting salary from years of service and department.

+ Suppose our company also had two sites, London and Birmingham, and we wanted to see if the salaries across department differed depending on location.

+ The table below contains the average salary in thousands of pounds for each group.

```{r,, echo=FALSE}
eg1 <- matrix(c(50,40,30,20),ncol=2,byrow=TRUE)
colnames(eg1) <- c("London","Birmingham")
rownames(eg1) <- c("Accounts","Manager")
eg1 <- as.table(eg1)
kable(eg1)
```

---
# Basic plot

.pull-left[
```{r,, echo=FALSE}
dat1 <- tibble(
  location = c("London", "London", "Birmingham", "Birmingham"),
  department = c("Accounts","Manager","Accounts","Manager"),
  salary = c(50, 30, 40, 20)
)
```

```{r, echo = FALSE}
kable(eg1)
```

+ And now let's look at the plot:
  + x-axis shows locations
  + y-axis is our salaries
  + The colours represent departments
  
]

.pull-right[

```{r, echo=FALSE, fig.height=6}
dat1 %>%
  ggplot(., aes(x = location, y = salary, group = department, colour = department, shape = department)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,60))
```
]


---
# Difference in differences (1)


.pull-left[

```{r,, echo=FALSE}
dat1 <- tibble(
  location = c("London", "London", "Birmingham", "Birmingham"),
  department = c("Accounts","Manager","Accounts","Manager"),
  salary = c(50, 30, 40, 20)
)
```

```{r, echo = FALSE}
kable(eg1)
```

+ In each plot we look at, think about subtracting the average store managers salary (blue triangle) from the average accounts salary (red circle).

+ In both cases it is £20,000.

+ Note, the lines are parallel
  + Remember what we have said about parallel lines...no interaction.

]


.pull-right[

```{r, echo=FALSE, fig.height=6}
dat1 %>%
  ggplot(., aes(x = location, y = salary, group = department, colour = department, shape = department)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,60)) +
  geom_segment(aes(x= 0.9 , xend = 0.9 , y=20, yend=40), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  geom_segment(aes(x= 2.1 , xend = 2.1 , y=30, yend=50), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  annotate("text", x = 0.8 , y = 30, label = "20") +
  annotate("text", x = 2.2 , y = 40, label = "20") +
  ggtitle("No difference")

```

]


---
# Difference in differences (2)


.pull-left[
```{r,, echo=FALSE}
dat2 <- tibble(
  location = c("London", "London", "Birmingham", "Birmingham"),
  department = c("Accounts","Manager","Accounts","Manager"),
  salary = c(50, 40, 40, 20)
)
```

```{r, echo = FALSE}
eg2 <- matrix(c(50,40,40,20),ncol=2,byrow=TRUE)
colnames(eg2) <- c("London","Birmingham")
rownames(eg2) <- c("Accounts","Manager")
eg2 <- as.table(eg2)
kable(eg2)
```

+ This time we can see the difference differs
  + £20,000 in Birmingham
  + £10,000 in London
  
+ Note the lines are no longer parallel
  + Suggests interaction
  + But not crossing (so ordinal interaction)

]

.pull-right[

```{r, echo=FALSE, fig.height=6}
dat2 %>%
  ggplot(., aes(x = location, y = salary, group = department, colour = department, shape = department)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,60)) +
  geom_segment(aes(x= 0.9 , xend = 0.9 , y=20, yend=40), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  geom_segment(aes(x= 2.1 , xend = 2.1 , y=40, yend=50), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  annotate("text", x = 0.8 , y = 30, label = "20") +
  annotate("text", x = 2.2 , y = 45, label = "10") +
  ggtitle("Difference")

```

]

---
# Difference in differences (3)

.pull-left[
```{r,, echo=FALSE}
dat3 <- tibble(
  location = c("London", "London", "Birmingham", "Birmingham"),
  department = c("Accounts","Manager","Accounts","Manager"),
  salary = c(40, 60, 40, 20)
)
```

```{r, echo = FALSE}
eg3 <- matrix(c(40,60,40,20),ncol=2,byrow=FALSE)
colnames(eg3) <- c("London","Birmingham")
rownames(eg3) <- c("Accounts","Manager")
eg3 <- as.table(eg3)
kable(eg3)
```

+ This time we can see the difference differs
  + £20,000 in Birmingham
  + £20,000 in London
  
+ Note the lines are no longer parallel
  + Suggests interaction.
  + Now crossing (so disordinal interaction)

]

.pull-right[

```{r, echo=FALSE, fig.height=6}
dat3 %>%
  ggplot(., aes(x = location, y = salary, group = department, colour = department, shape = department)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,60)) +
  geom_segment(aes(x= 0.9 , xend = 0.9 , y=20, yend=40), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  geom_segment(aes(x= 2.1 , xend = 2.1 , y=40, yend=60), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  annotate("text", x = 0.8 , y = 30, label = "20") +
  annotate("text", x = 2.2 , y = 50, label = "-20") +
  ggtitle("Big difference")

```

]

---
class: center, middle
# Question break ...


---
#  Lecture notation 

$$y_i = \beta_0 + \beta_1 x_{i} + \beta_2 z_{i} + \beta_3 xz_{i} + \epsilon_i$$

+ Lecture notation:
  + $y$ is a continuous outcome
  
  + $x$ is a categorical predictor ( `location` )
  
  + $z$ is a categorical predictor (`department` )
  
  + $xz$ is their product or interaction predictor
  
---
# Product for dummy variables

```{r, echo=FALSE}
tibble(
  location = c("London", "London", "Birmingham", "Birmingham"),
  department = rep(c("Accounts", "Manager"),2),
  x = c(0,0,1,1),
  z = c(0,1,0,1),
  xz = c(0,0,0,1)
) %>%
  kable(.) 
```

---
#  Interpretation: Categorical*categorical interaction (dummy codes)

$$y_i = \beta_0 + \beta_1 x_{i} + \beta_2 z_{i} + \beta_3 xz_{i} + \epsilon_i$$

+ $\beta_0$ = Value of $y$ when $x$ and $z$ are 0

+ Expected salary for Accounts in London.


---
#  Interpretation: Categorical*categorical interaction (dummy codes)

$$y_i = \beta_0 + \beta_1 x_{i} + \beta_2 z_{i} + \beta_3 xz_{i} + \epsilon_i$$

+ $\beta_1$ = Difference between levels of $x$ when $z$ = 0

+ The difference in salary between Accounts in London and Birmingham



---
#  Interpretation: Categorical*categorical interaction (dummy codes)

$$y_i = \beta_0 + \beta_1 x_{i} + \beta_2 z_{i} + \beta_3 xz_{i} + \epsilon_i$$

+ $\beta_2$ = Difference between levels of $z$ when $x$ = 0.

+ The difference in salary between Accounts and Store managers in London.


---
#  Interpretation: Categorical*categorical interaction (dummy codes)

$$y_i = \beta_0 + \beta_1 x_{i} + \beta_2 z_{i} + \beta_3 xz_{i} + \epsilon_i$$

+  $\beta_3$ = Difference between levels of $x$ across levels of $z$

+ The difference between salary in Accounts and Store managers between London and Birmingham



---
#  Example: Categorical*categorical

.pull-left[
```{r, eval=FALSE, warning=FALSE}
salary3 %>%
  group_by(location, department) %>%
  summarise(
    Salary = mean(salary)
  )
```
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}
group_means <- salary3 %>%
  group_by(location, department) %>%
  summarise(
    Salary = mean(salary)
  )
group_means
```
]

---
#  Example: Categorical*categorical
```{r}
m1 <- lm(salary ~ location*department, salary3)
```

```{r, echo=FALSE}
summary(m1)
```

---
#  Example: Categorical*categorical

.pull-left[

+ We can visualise categorical interactions using `cat_plot()` from `interactions`. 

+ `probe_interaction()` does not work with two categorical predictors.

```{r, eval=FALSE}
cat_plot(m1,
         pred = location,
         modx = department)
```
]

.pull-right[
```{r, echo=FALSE}
cat_plot(m1,
         pred = location,
         modx = department)
```

]


---
#  Example: Categorical*categorical
```{r, echo=FALSE}
res <- summary(m1)
round(res$coefficients,2)
```

.pull-left[
+ $\beta_0$ = Value of $y$ when $x$ and $z$ are 0

+ Expected salary for Accounts in London is £50,670.
]


.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}

ex1_means <- matrix(c(round(group_means$Salary,2)),ncol=2,byrow=FALSE)
colnames(ex1_means) <- c("London","Birmingham")
rownames(ex1_means) <- c("Accounts","Manager")
ex1_means <- as.table(ex1_means)
kable(ex1_means) 

```
]

---
#  Example: Categorical*categorical
```{r, echo=FALSE}
res <- summary(m1)
round(res$coefficients,2)
```

.pull-left[
+ $\beta_1$ = Difference between levels of $x$ when $z$ = 0

+ The difference in salary between Accounts in London and Birmingham is £1,980. The salary is lower in Birmingham. 

+ These are calculated as:

$$\text{Group} - \text{Reference group (intercept)}$$

]


.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(ex1_means) 
```
]


---
#  Example: Categorical*categorical
```{r, echo=FALSE}
res <- summary(m1)
round(res$coefficients,2)
```

.pull-left[
+ $\beta_2$ = Difference between levels of $z$ when $x$ = 0.

+ The difference in salary between Accounts and Store managers in London is £3,460. The salary is lower for Store Managers.

+ These are calculated as:

$$\text{Group} - \text{Reference group (intercept)}$$

]


.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(ex1_means) 
```
]

---
#  Example: Categorical*categorical
```{r, echo=FALSE}
res <- summary(m1)
round(res$coefficients,2)
```

.pull-left[
+  $\beta_3$ = Difference between levels of $x$ across levels of $z$

+ The difference between salary for Accounts and Store managers between London and Birmingham, differs by £8,420. The difference is greater in Birmingham than in London.

+ These are fiddly (handout online that goes through this in detail)

]


.pull-right[
```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(ex1_means) 
```
]

---
class: center, middle
# Good time to pause

---
# Extending past 2x2
+ When fitting an interaction to categorical variables with more than 2 levels, we need additional interaction terms.

+ Remember, I categorical variable with 3 levels is represented by 2 dummy coded variables.
  + So that means we have two variables to create products with.

+ The general rule on the number of interaction terms is:

$$(r-1)*(c-1)$$

+ Where 
  + $r$ (row) = number of levels of the first categorical variable.
  + $c$ (column) = number of levels of the second categorical variable.


---
# Example
+ The data comes from a study into patient care in a paediatric wards. 

+ A researcher was interested in whether the subjective well-being of patients differed dependent on the post-operation treatment schedule they were given, and the hospital in which they were staying. 

+ **Condition 1**: `Treatment` (Levels: TreatA, TreatB, TreatC).
  
+ **Condition 2**: `Hosp` (Levels: Hosp1, Hosp2). 
  
+ Total sample n = 180 (30 patients in each of 6 groups).
  + Between person design. 

+ **Outcome**: Subjective well-being (SWB)
  + An average of multiple raters (the patient, a member of their family, and a friend). 
  + SWB score ranged from 0 to 20.


---
# The data
```{r}
hosp_tbl <- read_csv("data/hospital.csv", col_types = "dff")
hosp_tbl %>%
  slice(1:10)
```

---
# The group means

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, 
  "TreatA", "10.80",  "7.85",
  "TreatB", "9.43",  "13.11",
  "TreatC", "10.10",  "7.98",
  ), col.names = c("", "Hosp1",  "Hosp2"))
```

---
# Model equations and coding

$$y_{ijk} = b_0 + \underbrace{(b_1D_1 + b_2D_2)}_{\text{Treatment}} + \underbrace{b_3D_3}_{\text{Hospital}} + \underbrace{b_4D_{13} + b_5D_{23}}_{\text{Interactions}} + \epsilon_{i}$$


```{r, echo=FALSE}
tibble(
  Treatment = c("A", "A", "B", "B", "C", "C"),
  Hospital = rep(c("Hosp1", "Hosp2"),3),
  D1 = c(0,0,1,1,0,0),
  D2 = c(0,0,0,0,1,1),
  D3 = c(0,1,0,1,0,1),
  D13 = c(0,0,0,1,0,0),
  D23 = c(0,0,0,0,0,1)
)
```

---
# Interpretation with dummy coding

$$y_{ijk} = b_0 + \underbrace{(b_1D_1 + b_2D_2)}_{\text{Treatment}} + \underbrace{b_3D_3}_{\text{Hospital}} + \underbrace{b_4D_{13} + b_5D_{23}}_{\text{Interactions}} + \epsilon_{i}$$

+ $b_0$ = Mean of treatment A in hospital 1.
+ $b_1$ = Difference between Treatment B and Treatment A in Hospital 1.
+ $b_2$ = Difference between Treatment C and Treatment A in Hospital 1.
+ $b_3$ = Difference between Treatment A in Hospital 1 and Hospital 2.
+ $b_4$ = Difference between Treatment A and Treatment B between Hospital 1 and Hospital 2
+ $b_5$ = Difference between Treatment A and Treatment C between Hospital 1 and Hospital 2

---
# Our results
```{r}
m4 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl)
m4sum <- summary(m4)
m4res <- round(m4sum$coefficients,2)
m4res
```

---
# Interpretation with dummy coding

```{r, echo=FALSE}
m4res
```

.pull-left[
+ $b_0$ = Mean of treatment A in hospital 1.
+ $b_1$ = Difference between Treatment B and Treatment A in Hospital 1.
+ $b_2$ = Difference between Treatment C and Treatment A in Hospital 1.
+ $b_3$ = Difference between Treatment A in Hospital 1 and Hospital 2.
+ $b_4$ = Difference between Treatment A and Treatment B between Hospital 1 and Hospital 2
+ $b_5$ = Difference between Treatment A and Treatment C between Hospital 1 and Hospital 2
]

.pull-right[
```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, 
  "TreatA", "10.80",  "7.85",
  "TreatB", "9.43",  "13.11",
  "TreatC", "10.10",  "7.98",
  ), col.names = c("", "Hosp1",  "Hosp2"))
```

]
---
# Brief comment: 3-way interactions

+ **In principle** we can extend interactions to more than two variables.

+ Consider our salary example:
  + We could ask does the change in difference in the effect of years of service across departments, differ dependent on whether you are based in London or Birmingham.
  
+ Note we have three predictors here:
  + Years of service (continuous)
  + department (binary; Store Manager, Accounts)
  + location (binary; London, Birmingham)

+ This is a plausible question concerning organisatonal fairness.

---
# Brief comment: 3-way interactions

+ In general, three-way interactions are tricky to interpret.

+ Require lots of data to test as the effects are often very small.
  + Extends the issues of power already discussed.

+ **Only test a three-way interaction if you have strong theoretical reason for doing so**

---
# Summary
+ We have considered how we fit and interpret linear models with categorical interactions.

+ We have focused on dummy coded (0,1) variables.

+ Next semester, we will return to this topic and consider other ways of treating categorical data when we have specific hypotheses we want to test.

---
class: center, middle
# Thanks for listening and interacting! :)