---
title: "Bootstrapping demo"
subtitle: "DAPR2 09 2026/27"
output: 
  html_notebook:
    toc: true
---

```{r setup, message = F}
library(tidyverse)
```


## Read in data from Qualtrics

Export settings:

- CSV
- tick "Download all fields"
- select "Export labels"

Read in CSV data downloaded from Qualtrics. 

```{r message = F}
DATAFILE <- "DAPR2 09 Bootstrapping_10 September 2026_10.02.csv"

# Wrangle Qualtrics data so it's formatted in a less horrible way
headers      <- read_csv(DATAFILE, col_names = FALSE, n_max = 1)
df           <- read_csv(DATAFILE, skip = 3, col_names = FALSE)
colnames(df) <- headers
df_tall <- df |>
  select(ResponseId, starts_with('Q')) |>
  pivot_longer(cols = starts_with('Q'),names_to = 'q', values_to = 'sampled_value')

# Save a vector containing each response ID
ppts <- df$ResponseId
```

How many respondents do we have data from?

```{r}
length(ppts)
```

Show a random handful of bootstrapped samples:

```{r}
df_tall |>
  # for nicer plotting, convert sampled value into a factor
  mutate(sampled_value = factor(sampled_value)) |>
  # select only a random sample of three participants
  filter(ResponseId %in% sample(ppts, size = 3)) |>
  # make the bar plot showing how many times each value was chosen per sample
  ggplot(aes(x = sampled_value, fill = ResponseId))+
  geom_bar() +
  facet_wrap(~ ResponseId) +
  # hide the legend bc the info already comes from the facet headers
  theme(legend.position = 'none') +
  # show all numbers on the x axis in every plot
  scale_x_discrete(drop = FALSE)
```


## Making the sampling distribution 

Compute the mean of each bootstrapped sample.
The result is the **sampling distribution of sample means.**

```{r}
sampling_distrib <- df_tall |>
  group_by(ResponseId) |>
  summarise(sample_mean = mean(sampled_value))

sampling_distrib
```


```{r warning=F}
p_samp_dist_dens <- sampling_distrib |>
  ggplot(aes(x = sample_mean)) +
  geom_density(fill = 'black', colour = 'black', alpha = 0.2) +
  geom_rug(colour = 'black') +
  labs(
    title = 'Sampling distribution of sample means',
    x = 'Mean of each bootstrapped sample'
  ) +
  xlim(0, 100)

p_samp_dist_dens
```




## Mean of the sampling distribution

What is the mean of our sampling distribution of means?

```{r}
samp_distrib_mean <- mean( sampling_distrib$sample_mean )
samp_distrib_mean
```

```{r}
p_samp_dist_dens +
  geom_vline(xintercept = samp_distrib_mean, linewidth = 2, colour = 'red') +
  geom_label(
    x = samp_distrib_mean+2, 
    y = 0.005, 
    colour = 'red', 
    label = paste('mean of bootstrapped\nsample means =', 
                  round(samp_distrib_mean, 2)),
    hjust = 0)
```


Probably pretty close to the mean of our main sample:

```{r}
main_sample <- c(6, 8, 17, 32, 70, 76, 79, 81, 85, 93)
mean(main_sample)
```



## The REALLY useful thing: Variability of the sampling distribution

```{r}
sd( sampling_distrib$sample_mean )
```


Why is this number useful?

We use the **standard deviation of the sampling distribution** to represent the **standard error of the mean** in situations where the formula for the standard error can't be used (e.g., because the assumptions about errors are not met.)
