---
title: "16 prep"
output: html_notebook
---

```{r}
library(tidyverse)
```

# bad lm


Why is it weird to fit a linear model to binary outcome data?

```{r}
mallow <- read_csv("https://uoepsy.github.io/data/mallow.csv")
```

```{r}
mallow_lm <- lm(taken ~ age, data = mallow)
```

## check assumptions of bad lm

```{r}
car::crPlots(mallow_lm)
```

non-linear



```{r}
tibble(residuals=mallow_lm$residuals) |> 
  ggplot(aes(x = residuals)) +
  geom_histogram()
```

```{r}
plot(mallow_lm, which = 2)
```

not normal!
pattern consistent with bounded outcome data.


```{r}
plot(mallow_lm, which = 1)
```

again, consistent with only two possible outcome values.


## try ot interpret coefs?

```{r}
summary(mallow_lm)
```

if you either took a marshmallow = 1 or didn't = 0, then what does it mean for the model to estimate that at age zero, the estimated outcome is 1.29? you SUPER took the marshmallow? weird.


## look at model predictions


```{r}
effects::effect(term = 'age', mod = mallow_lm, xlevels = list(age = 0:10)) |>
    as.data.frame() |>
    ggplot(aes(x = age, y = fit)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)
```

imposs values.
(do this or coefs. i think this is better bc impossibility is clearer.)

# plot logistic curve

copy code

```{r}
p_logistic <- tibble(
  logodds = seq(-5, 5, length.out = 100),
  prob = plogis(logodds)
) |>
  ggplot(aes(x = logodds, y = prob)) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  labs(
    y = 'Probability',
    x = 'Log-odds'
  ) +
  ggtitle('The logistic function')

p_logistic
```



# transforming probs to log-odds

Based on the plot of the logistic curve, guess the approximate log-odds value that corresponds to each of these probabilities.
Write down your guess.

```{r}
probs <- c(0.33, 0.05, 0.99, 0.2, 0, 1)
```


Now compute the actual log-odds value for each probability and compare your guesses.

```{r}
qlogis(probs)
```

Try running the following lines of code. why does each one fail?

```{r}
qlogis(-5)

qlogis(1.5)

qlogis(100)
```


# transforming log-odds to probs


guess

```{r}
logodds <- c(-10, -2, 0.2, 5, 10)
```

```{r}
plogis(logodds)
```

Can you find any values that make `plogis()` fail?

If you find some values, what values are they, and why does `plogis()` fail?

If there are no values, why does `plogis()` not fail?

```{r}
plogis(-Inf)
```

```{r}
plogis(Inf)
```


# why back-transforming indiv model coefs into prob space does not work

Imagine a line in log-odds space.
That line has an intercept of 1 and a slope of 0.5.

Here's some data that corresponds to that line:

```{r}
logodds_line_df <- tibble(
    predictor = 0:5,
    logodds_success = 1 + 0.5 * predictor
)

logodds_line_df
```

To plot the line:

```{r}
logodds_line_df |>
    ggplot(aes(x = predictor, y = logodds_success)) +
    geom_line()
```

back-transform intercept to prob space.

```{r}
plogis(1)
```


do the wrong thing and back-transform slope to prob space using `plogis()` as well, we get 

```{r}
plogis(0.5)
```


what value would this line predict for predictor = 1? predictor = 2?

```{r}
0.73 + 0.62 * 1
0.73 + 0.62 * 2
```



so if that was a line in probability space, it would look like

```{r}
tibble(
    predictor = 0:5,
    bad_prob_success = plogis(1) + plogis(0.5) * predictor
) |>
    ggplot(aes(x = predictor, y = bad_prob_success)) +
    geom_line()
```


how to back-transform this line to probabilities *correctly*:

we can only back-transform actual log-odds values, NOT differences between log-odds values.

```{r}
logodds_line_df |>
    mutate(
        prob_success = plogis(logodds_success)
    )
```


and if you plot it, notice that it's not a straight line, so it doesn't make sense to have a constant slope in probability space!

```{r}
logodds_line_df |>
    mutate(
        prob_success = plogis(logodds_success)
    ) |>
    ggplot(aes(x = predictor, y = prob_success)) +
    geom_line()
```


```{r}
plogis(1 + 0.5*0)
plogis(1 + 0.5*1)
plogis(1 + 0.5*2)
```
