---
title: "11: Numeric/categorical interactions"
params: 
    SHOW_SOLS: TRUE
    # TOGGLE: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')
# source('assets/setup.R')

library(tidyverse)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll TODO

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab11.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
   
:::

---

:::frame

RQ: **Does the association between mental wellbeing and the number of social interactions differ between rural and non-rural residents?**

Data dictionary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ruraldata  <- read_csv("https://uoepsy.github.io/data/wellbeing_location_rural.csv")
tibble(
variable = names(ruraldata),
description = c("Age in years of respondent","Self report estimated number of hours per week spent outdoors", "Self report estimated number of social interactions per week (both online and in-person)", "Binary 1=Yes/0=No response to the question 'Do you follow a daily routine throughout the week?'", "Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and well-being. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70", "Location of primary residence (Rural, Not Rural)")
) %>% gt::gt()
```
  
:::{.callout-note collapse='true'}
### More detail about this dataset

Researchers were specifically interested in how the number of social interactions influenced mental health and wellbeing differently for those living in rural communities compared to those in non rural locations. They wanted to assess whether the effect of social interactions on wellbeing _was moderated by_ (depends upon, interacts with) whether or not a person lived in a rural area.

From the Edinburgh & Lothians, 100 non-rural (e.g., city/suburb) residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), was used to measure mental health and well-being. 

Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), and whether a daily routine is followed (yes/no).  
  
The data in `wellbeing_location_rural.csv` contain six attributes collected from a random sample of $n=200$ hypothetical residents over Edinburgh & Lothians.

:::

:::

# q1

`r qbegin(qcounter())`

Read in the data from <https://uoepsy.github.io/data/wellbeing_location_rural.csv> and store it in a variable named `rural_data`.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
rural_data <- read_csv("https://uoepsy.github.io/data/wellbeing_location_rural.csv")
```

`r solend()`


