---
title: "02: Multiple regression"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: FALSE  # TODO replace w TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->


```{r}
#| label: setup
#| include: false

source('../assets/setup.R')

library(tidyverse)
library(patchwork)
library(sjPlot)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll add another predictor into the same model you fit last week.
You'll practice interpreting coefficients in this multiple regression model (i.e., a model with more than one predictor), and you'll learn how to make a nice plot of one of the resulting model-estimated associations.

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab02.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
    - `patchwork`
    - `sjPlot`
   
:::

---

:::frame

Research question (RQ): **Is there an association between wellbeing and time spent outdoors, when controlling for the effect that social interaction has on wellbeing?**

Data dictionary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
mwdata  <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
tibble(
variable = names(mwdata),
description = c("Age in years of respondent","Self report estimated number of hours per week spent outdoors ", "Self report estimated number of social interactions per week (both online and in-person)", "Binary 1=Yes/0=No response to the question 'Do you follow a daily routine throughout the week?'", "Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and wellbeing. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70", "Location of primary residence (City, Suburb, Rural)", "Average weekly number of steps in thousands (as given by activity tracker if available)")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

From the Edinburgh & Lothians, 100 city/suburb residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS) was used to measure mental health and wellbeing. 

Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), whether a daily routine is followed (yes/no). For those respondents who had an activity tracker app or smart watch, they were asked to provide their average weekly number of steps.  

:::

:::

# Read in and explore data

<!-- # q1 -->

`r qbegin(qcounter())`

Read in the data from  <https://uoepsy.github.io/data/wellbeing_rural.csv> and store it in a variable named `mwdata` ("mw" stands for "mental wellbeing").

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mwdata <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
```

`r solend()`


<!-- # q2 -->

`r qbegin(qcounter())`

Recall our RQ: **Is there an association between wellbeing and time spent outdoors, when controlling for the effect that social interaction has on wellbeing?**

Identify the three relevant variables (that is, the three columns in `mwdata`) which we'll use to address this question.

Produce two scatterplots, showing how each of the independent (predictor) variables is individually associated with the outcome variable.
Use `geom_smooth(method = 'lm', se = FALSE)` to add the line of best fit to each scatterplot.

Bonus: use `patchwork` to create a single graphic that shows the two scatterplots beside each other.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The relevant variables in `mwdata`:

- `wellbeing`: outcome
- `outdoor_time`: one predictor
- `social_int`: another predictor (included so that we can hold it constant)


```{r}
wellbeing_outdoor <- 
  ggplot(data = mwdata, aes(x = outdoor_time, y = wellbeing)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Time spent outdoors \nper week (hours)", y = "Wellbeing score (WEMWBS)")

wellbeing_social <- 
  ggplot(data = mwdata, aes(x = social_int, y = wellbeing)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Number of social interactions \nper week", y = "Wellbeing score (WEMWBS)")
```


```{r}
# place plots adjacent to one another using notation from the `patchwork` package
wellbeing_outdoor + wellbeing_social
```

`r solend()`

<!-- # q3 -->


`r qbegin(qcounter())`

When we include multiple predictors in a linear model, we want to make sure that they're aren't too highly correlated.
(In Week 8, we'll go into more detail about why.)

Use the `cor()` function to find the correlation coefficient for the correlation between the two predictor variables.

What kind of correlation does this coefficient represent?

:::{.callout-tip collapse='true'}
### Code hint

Here's the basic usage of `cor()` for getting the correlation between two variables:

```{r eval=F}
cor(variable1, variable2)
```

Or, if the two variables are the only two columns in a dataframe called `df`:

```{r eval=F}
cor(df)
```

:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

There are a couple ways to get the correlation coefficient for the correlation between `outdoor_time` and `social_int`.

You might use the dollar sign notation to extract each column from `mwdata`:

```{r}
cor(mwdata$outdoor_time, mwdata$social_int)
```

<br>

You might also subset `mwdata` using `select()` and then pipe the resulting two columns into `cor()`:

```{r}
mwdata |>
  select(outdoor_time, social_int) |>
  cor()
```

(This option gives you a correlation matrix.
The 1s on the diagonal indicate that each variable is perfectly correlated with itself, and the off-diagonals represent the correlations between different variables.)

<br>

The correlation coefficient is –0.04, which is a small/weak negative correlation: as outdoor time increases, social interaction tends to decrease (or vice versa), but only a tiny bit.



`r solend()`


# Set up and fit linear model

<!-- # q4 -->

`r qbegin(qcounter())`

Write the mathematical model formulation for the linear model that will address our RQ.

Which $\beta$ coefficient will we interpret in order to address our RQ about the relationship between outdoor time and wellbeing, when controlling for social interaction?

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

$$
\text{wellbeing} = \beta_0 + 
(\beta_1 \cdot \text{outdoor\_time}) +
(\beta_2 \cdot \text{social\_int}) +
\epsilon
$$

Write:

```
$$
\text{wellbeing} = \beta_0 + 
(\beta_1 \cdot \text{outdoor\_time}) +
(\beta_2 \cdot \text{social\_int}) +
\epsilon
$$
```

The $\beta$ coefficient that we'll use to address our RQ is $\beta_1$, which is the slope of the line that associates outdoor time with wellbeing while holding social interaction constant.

`r solend()`


<!-- # q5 -->

`r qbegin(qcounter())`

Use the function `lm()` to fit this linear model.
Name the result `mdl`.

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mdl <- lm(wellbeing ~ outdoor_time + social_int, data = mwdata)
```

`r solend()`

# Look at model estimates

<!-- # q6 -->

`r qbegin(qcounter())`

Look at the model summary using `summary()`.

Write one sentence per $\beta$ coefficient, interpreting what each coefficient means in the context of this data and RQ.

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
summary(mdl)
```

<br>

- **Intercept ($\beta_0$):** The estimated wellbeing score of somebody with zero hours of outdoor time and zero social interactions per week is 28.62 points.
- **Slope over `outdoor_time` ($\beta_1$):** Increasing one hour of outdoor time per week is associated with an increase in wellbeing score of 0.20 points, *holding the number of weekly social interactions constant*.
- **Slope over `social_int` ($\beta_2$):** Increasing one social interaction per week is associated with an increase in wellbeing score of 0.33 points, *holding the number of outdoor hours constant*.

(Remember to name the units after you've given the number! That is, write "28.62 points" and not just "28.62".)


`r solend()`

<!-- # q7 -->

`r qbegin(qcounter())`

Now you'll use a function called `tab_model()` to create a pretty-printed table of coefficient estimates.
It comes from the package `sjPlot`.

Run the following code:

```{r eval=F}
tab_model(mdl)
```

**Caution:** `tab_model()` is very nice, but if you include it in an Rmd file which you then want to knit to PDF, it will cause `knitr` to crash.
**So it's best to use `tab_model()` only in Rmd files that you want to knit to HTML.**
If you need to knit to PDF, then you can recreate the output of `tab_model()` manually.

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
tab_model(mdl)
```

`r solend()`



<!-- # q8 -->

`r qbegin(qcounter())`

Here you'll try out a new function called `plot_model()`, also from the package `sjPlot`.
Run the following code to generate a plot showing the association between `outdoor_time` and `wellbeing`, for a few different values of `social_int`.

```{r eval=F}
plot_model(
    mdl, 
    type = 'eff', 
    terms = c(
        'outdoor_time', 
        'social_int'
    ),
    show.data = TRUE
)
```

(If you're not sure what one of the lines of this code does, then comment it out and run the rest of the code again.
Pay attention to what changes.)

<br>

**Based on this plot and the relevant model coefficient you identified above, how would you respond to the RQ?**

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
plot_model(
    mdl, 
    type = 'eff', 
    terms = c(
        'outdoor_time', 
        'social_int'
    ),
    show.data = TRUE
)
```

This plot shows a positive association between outdoor time and wellbeing, even when accounting for social interaction: no matter how much social interaction people get, people with more outdoor time are estimated to have higher wellbeing scores.
And the $\beta$ coefficient that quantifies the slope of this line is also positive at 0.20.
Therefore I'd conclude that yes, there seems to be a positive association between wellbeing and time spent outdoors, controlling for the effect that social interaction also has on wellbeing.

(Next week, we'll learn how to report the corresponding significance test!)



`r solend()`


# Use the model to make predictions

<!-- # q9 -->

`r qbegin(qcounter())`

Take the coefficients estimated in `mdl`, rounded to two decimal points, and substitute them appropriately into the mathematical model expression you wrote in Q4.

:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

$$
\text{wellbeing} = 28.62 + 
(0.20 \cdot \text{outdoor\_time}) +
(0.33 \cdot \text{social\_int}) +
\epsilon
$$

Write:

```
$$
\text{wellbeing} = 28.62 + 
(0.20 \cdot \text{outdoor\_time}) +
(0.33 \cdot \text{social\_int}) +
\epsilon
$$
```

`r solend()`


<!-- # q10 -->

`r qbegin(qcounter())`

Last week, you used your linear expression to reconstruct the estimated `wellbeing` score of somebody whose data we observed (the person whose data was in the third row of `mwdata`).

Finally, this week, you'll use the linear expression you wrote in Q9 to predict `wellbeing` scores for people that we *haven't* observed any data from.

Calculate the `wellbeing` scores of the following three people (remember you can use the R console as a calculator):

- Shuting:
    - Outdoor time = 36 hours
    - Social interactions = 19
- Fatima:
    - Outdoor time = 20 hours
    - Social interactions = 15
- Donna: 
    - Outdoor time = 1 hour
    - Social interactions = 7

Who has the highest predicted `wellbeing` score?
Who has the lowest?


:::red
🗂️
See [TODO](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

For Shuting:

$$
\begin{align}
\text{wellbeing}_{\text{Shuting}} &= 28.62 + (0.20 \cdot 36) + (0.33 \cdot 19) \\
  &= 42.09 \\
\end{align}
$$

Write:

```
\begin{align}
\text{wellbeing}_{\text{Shuting}} &= 28.62 + (0.20 \cdot 36) + (0.33 \cdot 19) \\
  &= 42.09 \\
\end{align}
```

<br>


For Fatima:

$$
\begin{align}
\text{wellbeing}_{\text{Fatima}} &= 28.62 + (0.20 \cdot 20) + (0.33 \cdot 15) \\
  &= 37.57 \\
\end{align}
$$

<br>


For Donna:

$$
\begin{align}
\text{wellbeing}_{\text{Donna}} &= 28.62 + (0.20 \cdot 1) + (0.33 \cdot 7) \\
  &= 31.13 \\
\end{align}
$$


<br>

Shuting has the highest predicted wellbeing score at 42.09, and Donna has the lowest at 31.13.

`r solend()`



