---
title: "05 playground"
output: 
    html_notebook:
        toc: true
---

# RQ1

RQ: How is people's sense of Fear of Missing Out (FOMO) associated with their age and how many Instagram followers they have?

Predictions:

- As age increases, FOMO will decrease.
- As the number of Instagram followers increases, FOMO will increase.


## 1a: Set up your code and data

Load required packages.

```{r}
library(tidyverse)   # for wrangling
library(psych)       # for summarising
library(sjPlot)      # for nice tables/plots
library(kableExtra)  # for nicely-formatted tables
library(patchwork)   # for combining plots into one graphic
```

Read in data.

```{r}
fomo <- read_csv("https://uoepsy.github.io/data/FOMOdataset.csv")
```

Check for missingness or impossible values. (Ranges all look reasonable here, and no NAs, so we're good to move on.)

```{r}
summary(fomo)
```

## 1b: Set up the variables you'll model

> RQ: How is people's sense of Fear of Missing Out (FOMO) associated with their age and how many Instagram followers they have?

RQ is about associations between variables, so we'll test it using significance test for coef estimates.

<br>

Outcome: `FOMO`
Predictors: `Age`, `TotalFollowers`

<br>

Transformations?
One option might be to z-score Age and TotalFollowers.
This would be an important step if Age = 0 and TotalFollowers = 0 were totally meaningless quantities (imagine Height = 0 or ShoeSize = 0 – impossible and nonexistent).
But Age = 0 means an infant, and TotalFollowers = 0 means someone with no followers.
So they are reasonable quantities to generate predictions for.

<br>

Plot outcome and predictors together.

```{r}
p_age <- fomo |>
    ggplot(aes(x = Age, y = FOMO)) +
    geom_point() +
    geom_smooth(method = 'lm', se = FALSE)
# p_age
```

```{r}
p_followers <- fomo |>
    ggplot(aes(x = TotalFollowers, y = FOMO)) +
    geom_point() +
    geom_smooth(method = 'lm', se = FALSE)
# p_followers
```

```{r}
p_age + p_followers
```


<br>

Nice table of descriptives:

- the describe() function is from the psych package, and kable() from kableExtra which is used to make a nice table where the values are rounded to 2 decimal places using digits = 2. 
- We are first renaming our variables to give more appropriate / informative names.
- Next we are selecting just the cols we care about and renaming them 

```{r}
fomo |> 
    select(FOMO, Age, TotalFollowers) |>
    rename(
        "Fear of Missing Out" = FOMO, 
        "Age (in years)" = Age, 
        "Number of Instagram Followers" = TotalFollowers
    ) |>
    describe() |>
    select(n, mean, sd, min, max) |>
    rename("N" = n, "Mean" = mean, "SD" = sd, "Minimum" = min, "Maximum" = max) |>
    kable(caption = "Descriptive statistics for FOMO and socio-demographic factors", digits = 2) |>
    kable_styling()
```



## 2: Model

```{r}
m1 <- lm(
    FOMO ~ Age + TotalFollowers,
    data = fomo
)
```

## 3a: Plot and interpret model estimates

Interpret coef estims:

```{r}
summary(m1)
```

```{r}
confint(m1) |> round(3)
```

- `Intercept`: The estimated FOMO score for someone aged zero with zero total followers is 26.52 points.
- `Age`: Increasing one year of age is associated with a decrease in FOMO score of 0.17 points (holding follower count constant). This estimate is significantly different from zero (b = –0.17, 95% CI [–0.18, –0.15], p < .001).
    - prediction borne out!
- `TotalFollowers`: Increasing one follower is associated with an increase in FOMO score of 0.018 points (holding age constant). This estimate is significantly different from zero (b = 0.018, 95% CI [0.016, 0.020], p < .001).
    - prediction borne out!
    - (rounded to 3 dp because otherwise estim and CI bounds are all 0.02)

<br>

Nice regression table:

```{r}
tab_model(
    m1,
    dv.labels = 'Fear of missing out (FOMO)',
    pred.labels = c("Age" = "Age (in years)"),
    title = "RQ1: Linear model coefficient estimates"
)
```

<br>

Plot model-fitted values:

```{r}
plot_model(
    m1, 
    type = 'eff', 
    terms = c('Age', 'TotalFollowers'), 
    show.data = TRUE
)
```


# RQ2 – Bonus challenge

RQ: Do the Big 5 personality traits significantly predict FOMO, in addition to the effects of age and total Instagram followers?


## 1b: Set up variables

This is a model comparison kind of question.

<br>

Outcome: `FOMO`
Predictors: `Age`, `TotalFollowers`, and each OCEAN variable

<br>

No great good way to plot data to set the scene for a model comparison question.

<br>

Descriptives table:

```{r}
fomo |> 
    select(O, C, E, A, N) |>
    rename(
        "Outgoingness" = O,
        "Conscientiousness" = C,
        "Extraversion" = E,
        "Agreeableness" = A,
        "Neuroticism" = N
    ) |>
    describe() |>
    select(n, mean, sd, min, max) |>
    rename("N" = n, "Mean" = mean, "SD" = sd, "Minimum" = min, "Maximum" = max) |>
    kable(caption = "Descriptive statistics for Big 5 personality traits", digits = 2) |>
    kable_styling()
```



## 2: Model

```{r}
m2 <- lm(
    FOMO ~ Age + TotalFollowers + O + C + E + A + N,
    data = fomo
)
summary(m2)
```


## 3a: Plot and interpret model estimates


Compare models. Because they're nested, we can use the incremental F-test via the `anova()` function.

```{r}
anova(m1, m2)
```

Significant result, so yes, Big 5 personality traits do predict FOMO over and above age and total Instagram followers (F(5, 3362) = 146.45, p < .001).




# Example write-up for RQ1

## Set the scene

In this study, we wanted to test how people's Fear of Missing Out (FOMO) is associated with socio-demographic factors: specifically, people's age and how many followers they have on Instagram.
We predicted that as age increases, FOMO will decrease, but that as the number of followers increases, FOMO will increase.

## Describe and show sample data

To test these predictions, we are using data gathered from 3370 individuals across the UK.
The data contains a FOMO score, as measured by a 10-item FOMO scale which produces scores ranging from 10 to 50.
The data also contains each respondent's age in years and the number of Instagram followers they have.
Descriptive statistics for the three relevant variables are shown in Table 1.

```{r}
#| code-fold: true

fomo |> 
    select(FOMO, Age, TotalFollowers) |>
    rename(
        "Fear of Missing Out" = FOMO, 
        "Age (in years)" = Age, 
        "Number of Instagram Followers" = TotalFollowers
    ) |>
    describe() |>
    select(n, mean, sd, min, max) |>
    rename("N" = n, "Mean" = mean, "SD" = sd, "Minimum" = min, "Maximum" = max) |>
    kable(
        caption = "Table 1: Descriptive statistics for FOMO and socio-demographic factors", 
        digits = 2
    ) |>
    kable_styling()
```


## Visualise and summarise  data patterns that address RQ

As shown in Figure 1, age appears to be negatively associated with FOMO (left), while number of Instagram followers appears positively associated with FOMO (right).

```{r fig.cap='ehlo'}
#| fig_caption: true
#| code-fold: true

p_age <- fomo |>
    ggplot(aes(x = Age, y = FOMO)) +
    geom_point() +
    geom_smooth(method = 'lm', se = FALSE) +
    labs(
        x = 'Age (in years)'
    )
p_followers <- fomo |>
    ggplot(aes(x = TotalFollowers, y = FOMO)) +
    geom_point() +
    geom_smooth(method = 'lm', se = FALSE) +
    labs(
        x = 'Number of Instagram followers'
    )

p_age + p_followers
```

> Figure 1: Associations between socio-demographic factors and FOMO. Age appears to have a positive association with FOMO, while the number of instagram followers appears to have a negative association.


## Introduce statistical analysis

We conducted all statistical analyses using R (R Core Team, 2025), and we will consider effects significant at $\alpha = .05$.

To address the RQ, we fit a linear model which predicts FOMO as a function of age in years and number of Instagram followers.
The mathematical model formulation of the model is as follows:

$$
\text{FOMO} = \beta_0 + (\beta_1 \cdot \text{Age}) + (\beta_2 \cdot \text{Followers}) + \epsilon
$$

The null hypothesis is that neither age nor number of followers are significantly associated with FOMO.
The alternative hypothesis is that either age or number of followers (or both) are significantly associated with FOMO.

$$
\begin{align}
H_0 &: \text{ All } \beta_j = 0 \text{ for } j = 1, 2 \\
H_1 &: \text{ Any } \beta_j \neq 0 \\
\end{align}
$$

## Report results of analysis

If our predictions were borne out, we would expect $\beta_1$ (the slope coefficient for age) to be negative, and we would expect $\beta_2$ (the slope coefficient for the number of followers) to be positive.

Indeed, $\beta_1$ is negative and significantly different from zero ($b$ = –0.17, 95% CI [–0.18, –0.15], $p$ < .001).
Thus age is significantly negatively associated with FOMO: for every year increased in age (holding follower count constant), FOMO is estimated to decrease by 0.17 points.
Further, $\beta_2$ is positive and also significantly different from zero ($b$ = 0.018, 95% CI [0.016, 0.020], $p$ < .001), indicating that follower count is significantly positively associated with FOMO.
Specifically, for an increase in one follower, FOMO is estimated to increase by 0.018 points.
Thus we can reject the null hypothesis defined above.

For an overview of all regression coefficients, see Table 2, and for a plot of model-fitted values, see Figure 2.

```{r}
tab_model(
    m1,
    dv.labels = 'Fear of missing out (FOMO)',
    pred.labels = c("Age" = "Age (in years)"),
    title = "Table 2: Linear model coefficient estimates"
)
```


```{r}
plot_model(
    m1, 
    type = 'eff', 
    terms = c('Age', 'TotalFollowers'), 
    show.data = TRUE
)
```

> Observed data (individual points) and well as model-fitted values (lines and 95% CI error ribbons) indicating that as age increases, FOMO tends to decrease, holding follower count constant; additionally, as total follower count increases, FOMO tends to also increase, holding age constant.


# Example write-up for RQ2

## Set the scene

Additionally, we are interested in the additional role that personality factors may play in predicting FOMO.
Specifically, we want to know: Do the Big 5 personality traits predict FOMO, over and above the roles played by age and number of Instagram followers?


## Describe and show sample data

Our data also includes measures of the Big 5 personality traits for each respondent: Openness, Conscientiousness, Extraversion, Agreeableness, and Neuroticism.
Descriptive statistics for these traits are shown in Table 3.

```{r}
fomo |> 
    select(O, C, E, A, N) |>
    rename(
        "Outgoingness" = O,
        "Conscientiousness" = C,
        "Extraversion" = E,
        "Agreeableness" = A,
        "Neuroticism" = N
    ) |>
    describe() |>
    select(n, mean, sd, min, max) |>
    rename("N" = n, "Mean" = mean, "SD" = sd, "Minimum" = min, "Maximum" = max) |>
    kable(caption = "Table 3: Descriptive statistics for Big 5 personality traits", digits = 2) |>
    kable_styling()
```


## Visualise and summarise data patterns that address RQ

Figure 3 shows how each personality trait is associated with FOMO in our data.
The largest association appears to be a positive association with Neuroticism, though Agreeableness and Conscientiousness may also be more weakly negatively associated with FOMO.

```{r}
#| code-fold: true

# you could accomplish the same thing by making five plots and patchworking
# them together, but this is a bit less repetitive

fomo |>
    rename(
        "Outgoingness" = O,
        "Conscientiousness" = C,
        "Extraversion" = E,
        "Agreeableness" = A,
        "Neuroticism" = N
    ) |>
    pivot_longer(
        cols = Neuroticism:Conscientiousness, 
        names_to = 'Trait', 
        values_to = 'Score'
    ) |>
    ggplot(aes(x = Score, y = FOMO)) +
    geom_point(alpha = 0.1) +
    geom_smooth(method = 'lm', se = F) +
    facet_wrap(~ Trait)
```

> Figure 3: Associations between each Big 5 personality trait and FOMO. Most personality traits appear to have a small negative association with FOMO, although Neuroticism appears to have a larger positive association.


## Introduce statistical analysis

To address this research question, we conducted an incremental F-test to compare the model from RQ1 to a new model which contains those same variables as well as each of the Big 5 traits.
Specifically, the new model is the following:

$$
\text{FOMO} = \beta_0 + 
(\beta_1 \cdot \text{Age}) + 
(\beta_2 \cdot \text{Followers}) + 
(\beta_3 \cdot \text{O}) + 
(\beta_4 \cdot \text{C}) + 
(\beta_5 \cdot \text{E}) + 
(\beta_6 \cdot \text{A}) + 
(\beta_7 \cdot \text{N}) + 
\epsilon
$$

The null hypothesis that corresponds to this model comparison is that none of Big 5 predictors are significantly associated with FOMO.
The alternative hypothesis is that any one (or more) of the Big 5 predictors are significantly associated with FOMO.

$$
\begin{align}
H_0 &: \text{ All } \beta_j = 0 \text{ for } j = 3, 4, 5, 6, 7 \\
H_1 &: \text{ Any } \beta_j \neq 0 \\
\end{align}
$$

## Report results of analysis 

Based on an incremental F-test, we conclude that the Big 5 personality traits do significantly predict FOMO over and above the effects of age and follower count ($F$(5, 3362) = 146.45, $p$ < .001).
We therefore reject the null hypothesis defined above.

