---
title: "Block 4 playground"
output: html_notebook
---

```{r}
library(tidyverse)
```


# mallow data

```{r}
mallow <- read_csv("https://uoepsy.github.io/data/mallow.csv")
```

## mallow with categ age-group var

```{r}
mallow <- mallow |>
  mutate(
    agegroup = factor(
      ifelse(age < 6, 'below 6', '6 or older'),
      levels = c('below 6', '6 or older')
    )
  )
```



# week 2: mallow continuous

## intercept-only model

```{r}
mallow_m0 <- glm(taken ~ 1, data = mallow, family = binomial)
summary(mallow_m0)
```

```{r}
confint(mallow_m0)
```



## plot taken ~ age

```{r}
mallow |>
  ggplot(aes(x = age, y = taken)) +

  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # add curve
  geom_smooth(
    method = "glm",  # not a straight line
    method.args = list(family = binomial),   # specify type of model
    se = FALSE  # remove standard error ribbon around line
  ) +
  
  labs(
    y = 'Took marshmallow immediately (0 = No, 1 = Yes)',
    x = 'Age (years)'
  )
```

## model taken ~ age

```{r}
mallow_m1 <- glm(taken ~ age, data = mallow, family = binomial)
summary(mallow_m1)
```

```{r}
confint(mallow_m1)
```

## plot model-fitted values

```{r}
library(effects)

effect(term = "age", mod = mallow_m1, xlevels = 20) |>
  as.data.frame() |>
  ggplot(aes(x = age, y = fit, ymin = lower, ymax = upper))+
  geom_line()+
  geom_ribbon(alpha = .3) +
  labs(
    y = 'Probability of taking the marshmallow immediately',
    x = 'Age (years)'
  ) +
  ylim(0, 1)
```

# week 2: mallow categorical

## plot taken ~ agegroup

```{r}
mallow |>
  ggplot(aes(x = factor(taken), fill = agegroup)) +
  geom_bar(position = 'fill') +
  labs(
    y = 'Proportion of children',
    x = 'Took marshmallow immediately (0 = No, 1 = Yes)',
    fill = 'Age group'
  )
```


## model taken ~ agegroup

```{r}
mallow_m2 <- glm(taken ~ agegroup, data = mallow, family = binomial)
summary(mallow_m2)
```

```{r}
confint(mallow_m2)
```

## plot model-fitted values

```{r}
effect(term = 'agegroup', mod = mallow_m2) |>
  as.data.frame() |>
  ggplot(aes(x = agegroup, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) +
  labs(
    y = 'Probability of taking the marshmallow immediately',
    x = 'Age group'
  ) +
  ylim(0, 1)
```



# week 3: mallow continuous

## Model comparison – Difference in deviance

```{r}
anova(mallow_m0, mallow_m1, test = 'Chisq')
```

Model with predictor significantly better than null model.
Big whoop.


## Assump: check std dev resids

```{r}
rstandard(mallow_m1, type = 'deviance') |> plot()
```

all good, all within |3|

## Diag: check Cook's Distance

```{r}
cooks.distance(mallow_m1) |> plot()
```

all good, all below 0.5


# week 3: mallow  categ

## Model comparison – Difference in deviance

```{r}
anova(mallow_m0, mallow_m2, test = 'Chisq')
```

Model with predictor significantly better than null model.
Big whoop.


## Assump: check std dev resids

```{r}
rstandard(mallow_m2, type = 'deviance') |> plot()
```

all good, all within |3|

## Diag: check Cook's Distance

```{r}
cooks.distance(mallow_m2) |> plot()
```

all good, all well below 0.5


