---
title: "Calculating the Slope and Intercept"
author: "Emma Waterston"
format: html
---


In this document we are going to work through the hand calculations for a simple linear model. 

## Packages

For this example we will only need the `tidyverse` package.

```{r, message=FALSE, warning=FALSE}

```


## Data

Below is code to create the data we used in lecture. As this is a small example data set, we can make the `tibble` manually, rather than import it. The structure is pretty simple. On the left of the `=` we name the variable, on the right we enter the data.

```{r}

```

The `tibble` function is pretty smart, and it defaults to setting the type of variable based on how you enter values. The `glimpse()` function allows us to look at the structure of our dataset. The `str()` function is pretty useful too.

```{r}
```

```{r}
```


We can see `student` is being recognised as a character vector, and our variable `hours` and `score` variables as `dbl` or numeric.

## Calculating the Slope

From lecture, we can calculate the slope using the following:

$$
\hat \beta_1 = \frac{SP_{xy}}{SS_x}
$$

Where $SP_{xy}$ = sum of cross-products:


$$
SP_{xy} = \sum_{i = 1}^{n}(x_i - \bar{x})(y_i - \bar{y})
$$

and $SS_x$ = sums of squared deviations of $x$:

$$
SS_x = \sum_{i = 1}^{n}(x_i - \bar{x})^2
$$

### Sum of cross-products

$$
SP_{xy} = \sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})
$$

Let's break down what this is doing. The $\sum_{i=1}^{n}$ is telling us that we need to perform the action that follows for each individual ($i$) for all rows of data ($n$). 

The two terms in parentheses require us to subtract the mean of $x$ ($\bar{x}$) and the mean of $y$ ($\bar{y}$) from the individual values of $x$ and $y$ for each row of data or participant. We know this because these are subscripted with $i$. In other words, for each person, we are calculating the deviation of their individual scores from the means.

We then multiply the deviations for each person. Once we have these for everyone, we sum them up.

What we are going to do now is add columns to our data set that represent the steps in this calculation. In our example, $x$ = `hours` and $y$ = `score`. First, lets add both means.

```{r}

```

We use `mutate()` to create new variables and add them to our data set. Notice for both `m_hours` and `m_score` the values in each row are the same. That is because there is only one mean. By default, `mutate()` includes this for every row. I have used `select()` to reorder the columns so the variable containing the values and the mean are next to one another. The numbers refer to the column numbers.

Next, we can calculate the deviances:

```{r}

```

Again, we calculated two new variables and reordered the data set.

The last step is to multiply the two deviances for each person.

```{r}

```

And there we go. Our `SPXYi` variable contains the product of the deviations of each observation from the means of hours studied and test score. Our final step is just to add this up.

```{r}

```

### Sum of square deviations

So now to calculate $SS_x$. Let's look at the equation again:

$$
SS_x = \sum_{i=1}^{n}(x_i - \bar{x})^2
$$

What do you notice?

Well, we have already calculated the individual deviations $(x_1 - \bar{x}), \dots, (x_n - \bar{x})$. They are the column called `dev_hours` in the `calc` table. So all we need to do is square these for each observation and sum them. 

```{r}

```

and

```{r}

```

### Return to the Slope

We now have the two values we need to calculate the slope:

$$
\hat \beta_1 = \frac{SP_{xy}}{SS_x} = \frac{21.75}{20.625} = 1.054545
$$

```{r}

```


## Calculating the Intercept

Calculations for intercept:

$$
\hat \beta_0 = \bar{y} - \hat \beta_1 \bar{x}
$$

So this one is much easier as all we need are:

- $\hat \beta_1$ = slope estimate
- $\bar{y}$ = mean of $y$
- $\bar{x}$ = mean of $x$

So...


```{r}

```

Keep a note of these values.

```{r}

```

```{r}

```

