---
title: "Week 2: Multiple Linear Regression & Standardisation"
author: "Emma Waterston"
format: html
---

In this document we are going to work through an example of a multiple linear regression model with standardised variables. 

## Packages

For this example we will need a few different packages:

```{r, message=FALSE, warning=FALSE}

```


## Data

Working with a subset of data used in the lecture, and reading this data into `R` using a link and assigning it to a dataset named dat:

```{r, message=FALSE, warning=FALSE}

```

Look at the structure of our data using `str()`:

```{r}

```

We can see `ID` is being recognised as a character vector, whilst `score`, `hours`, and `motivation` as numeric.

## Visualisations


```{r}

```


## Building Multiple Linear Regression Models

### Only motivation standardised

`motivation` was already a z-score.

```{r}

```

#### Interpretation 

We can interpret our coefficients as follows:

+ $\beta_0$ = A student who did not study who had average school motivation was expected to score 6.38 points on the test
+ $\beta_1$ = Controlling for motivation, for every 1 additional hour studied, there was a 1.39 point increase in test score
+ $\beta_2$ = Controlling for hours of study, for every 1 SD increase in motivation, there was a 0.98 point increase in test score


### Hours and motivation standardised

```{r}

```

#### Interpretation 

We can interpret our coefficients as follows:

+ $\beta_0$ = A student studied for an average number of hours who had average school motivation was expected to score 16.10 points on the test
+ $\beta_1$ = Controlling for motivation, for every 1 SD increase in hours studied, there was a 6.41 point increase in test score
+ $\beta_2$ = Controlling for hours of study, for every 1 SD increase in motivation, there was a 0.98 point increase in test score


### All variables (DV + 2 IVs) standardised

```{r}

```

#### Interpretation 

The intercept often becomes less practically informative because it mainly reflects the standardisation of the data rather than meaningful zero points of the predictors. Here we are often more interested in interpreting the estimated coefficients of the predictors themselves.

We can interpret our coefficients as follows:

+ $\beta_1$ = Controlling for motivation, for every 1 SD increase in hours studied, there was a 0.81 SD increase in test score
+ $\beta_2$ = Controlling for hours of study, for every 1 SD increase in motivation, there was a 0.12 SD increase in test score
