---
title: "Testing Predictors and Evaluating Linear Models"
author: "Dr. Patrick Sturt (Patrick.Sturt@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
---


```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

dapr2red <- "#BF1932" 

theme_set(
  theme_minimal(
    base_size = 28
  )
)

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  hours = seq(0.5,5,.5),
  score = c(1,3,1,2,2,6,3,3,4,8)
)

res <- lm(score ~ hours, data = test)
```


# Course overview {background-color="white" style="font-size:60%;"}

TODO

## This Week's Learning Objectives

1. Understand how to interpret significance tests for $\beta$ coefficients

2. Understand how to calculate and interpret $R^2$ and adjusted- $R^2$ as a measure of model quality

3. Be able to locate information on the significance of individual predictors and overall model fit in R `lm` model output

# Part 1: Recap & Overview

## Recap

+ Last week we expanded the general linear model equation to include multiple predictors:

$$
y_i = \beta_0 + (\beta_1 \cdot x_{1i}) + (\beta_2 \cdot x_{2i}) + (\beta_j \cdot x_{ji}) + \epsilon_i
$$

+ And we ran an example concerning test scores:

$$
\text{Score}_i = \beta_0 + (\beta_1 \cdot \text{Hours}_{i}) + (\beta_2 \cdot \text{Motivation}_{i}) + \epsilon_i
$$

+ And we looked at how to run this model in `R`:

```{r, eval=FALSE}
lm(score ~ hours + motivation, data = test_study2)
```


## Evaluating our Model

So far we have estimated values for the key parameters of our model ( $\beta$s )

+ Now we have to think about how we evaluate the model

. . .

+ Evaluating a model will consist of:

  1. Evaluating the individual coefficients
  
  2. Evaluating the overall model quality
  
  3. Evaluating the model assumptions


. . .

**Important:** Before accepting a set of results, all three of these aspects of evaluation must be considered

+ We will talk about evaluating individual coefficients and model quality today
+ Model assumptions covered later in the course (Semester 1, Week 8)


##  Significance of Individual Effects 

+ A general way to ask this question would be to state: 

> **Is our model informative about the association between X and Y?**

. . .

+ In the context of our example from last lecture, we could ask, 

> **Is study time a useful predictor of test score?**

. . .

+ This is kind of vague, what does it mean to be a "useful predictor"?

+ We need to turn this into a testable statistical hypothesis


##  Evaluating Individual Predictors

Steps in hypothesis testing:
  
. . .
  
  + Research question
  
. . .

  + Statistical hypothesis

. . .
  
  + Define the null hypothesis
  
. . .
  
  + Calculate an estimate of effect of interest
  
. . .
  
  + Calculate an appropriate test statistic
  
. . .
  
  + Evaluate the test statistic against the null
  

## Research Questions and Hypotheses

**Research questions** are statements of what we intend to study    
    
<br>
    
A good question defines:

  + constructs under study
  + the association being tested
  + a direction of association
  + target populations etc.

For example:

> **Does increased study time improve test scores in school-age children?**

## Research Questions and Hypotheses

**Statistical hypotheses** are testable mathematical statements  
    
+ In typical testing in Psychology, we define a **null hypothesis ($H_0$)** and an **alternative hypothesis ($H_1$)**.
  
  + $H_0$ is precise: it specifically states that the effect of interest is equal to zero.
  + $H_1$ is not specific: it simply says that the effect of interest is *not* equal to zero (but it makes no claims about what the effect of interest might really be equal to).


## Statistical Significance: Review from DAPR1

+ Remember, we can only ever test the null hypothesis

+ We select a significance level, $\alpha$ (typically .05)

+ Then we calculate the $p$-value associated with our test statistic

  + A $p$-value is the probability of observing our test statistic or a value even more extreme than our test statistic, assuming that $H_0$ is true (that is, assuming that the true effect really is zero)

+ If the associated $p$ is smaller than $\alpha$, then we **reject** the null

+ If it is larger, then we **fail to reject** the null


# Part 2: Steps in Significance Testing

## Defining the Null  {.smaller}

:::: {.columns}

::: {.column width="50%"}

In this example data, there's no pattern in how values of $x$ are associated with values of $y$.

We can't make any guesses about $y$ based on the value of $x$.

In this scenario, the slope $\beta_1$ of the line associating $x$ and $y$ is 0.

**Why is the slope 0?**
  
+ $\beta$ gives the predicted change in $y$ for a unit change in $x$
+ If $x$ and $y$ are unrelated, then a change in $x$ will not result in any change to the predicted value of $y$
+ So for a unit change in $x$, there is no change (zero change) in $y$
	
**Our null hypothesis is that there is no association between $x$ and $y$. Formally:**
	
$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$

:::

::: {.column width="50%"}

```{r, echo = F, message = F, fig.height=7, fig.width=7}
set.seed(2010)
dat <- data.frame(x=1:25, y=rnorm(25))
ggplot(dat, aes(x, y)) + 
  geom_point(size=3) +
  geom_smooth(method = 'lm', se=F, colour = '#BF1932') + 
  # theme(axis.title=element_text(size = 14, face = 'bold'))
  NULL
```

:::

::::

## Significance tests for $\beta$ coefficients

+ The test statistic associated with $\beta$ coefficients is a $t$-statistic:

$$t = \frac{\hat \beta}{SE(\hat \beta)} = \frac{\text{estimated coefficient}}{\text{standard error estimated for that coefficient}}$$


+ Recall from DAPR1 that the standard error describes the spread of the sampling distribution
  + The standard error $(SE)$ measures how variable the sample is
  + A smaller SE suggests a more precise estimate (= good)
  + If you like maths, check the appendix of these slides for details about how to compute SE by hand

## Let's work out a $t$-value

:::{style="font-size:65%;"}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(7284) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1)),
  error = list(variance = 20),
  sample_size = 150,
  reg_weights = c(0.6, 1.4, 1.5)
)

df <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study2 <- df %>%
  dplyr::select(y, hours, motivation) %>%
  mutate(
    ID = paste("ID", 101:250, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2)
  ) %>%
  dplyr::select(ID, score, hours, motivation)

m1 <- lm(score ~ hours + motivation, data = test_study2)
res <- lm(score ~ hours, data = test)
```

```{r}
summary(m1)
```


<br>

For `motivation`, which we'll represent as $\beta_2$:

$$
t = \frac{\hat \beta_2}{SE(\hat \beta_2)} = \frac{0.9163}{0.3838} = 2.388
$$

:::

## Compare $t$ to a Null Distribution {.smaller}

- We have an observed $t$-statistic, 2.388. 
Now we want to find out how likely it is to observe a $t$-value this extreme or more extreme, assuming that $H_0$ is true.

- **We know the distribution that $t$-values follow when $H_0$ is true: a $t$-distribution.**

- $t$-distributions can vary, depending on their degrees of freedom (df):

```{r, echo = F, warning = F, message = F, fig.height = 3.5}
ggplot() + 
  xlim(-4, 4) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=2), 
                colour = '#BF1932',
                linewidth = 1) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=5),
                colour = '#0F4C81',
                linewidth = 1) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=500),
                colour = '#88B04B',
                linewidth = 1) +
  labs(x='t', y = '') +
  annotate(hjust = 0, geom = 'text', label = 'df = 2', colour = '#BF1932', x = 2.2, y = .15, size = 8) +
  annotate(hjust = 0, geom = 'text', label = 'df = 5', colour = '#0F4C81', x = 2.2, y = .25, size = 8) +
  annotate(hjust = 0, geom = 'text', label = 'df = 500', colour = '#88B04B', x = 2.2, y = .35, size = 8) +
  # theme(axis.title = element_text(size = 12, face = 'bold'))
  NULL
```

+ For $\beta$, the null distribution is a $t$-distribution with **$n-k-1$ degrees of freedom**
	+ $n$ = sample size
  + $k$ = number of predictors (that is, the number of $\beta$ coefficients that are not the intercept $beta_0$)
	+ The additional –1 represents the intercept
	

## $n-k-1$ degrees of freedom

- $n$ = sample size **= 150**
- $k$ = number of predictors **= 2 (`hours` and `motivation`)**
	
So: $150 - 2 - 1 = 147$ degrees of freedom.

```{r, echo = F}
(p <- ggplot() + 
  xlim(-5, 5) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=147),
                linewidth = 1) +
  labs(x='\n t', y = '') + 
  ggtitle("t-distribution (df=147)") +
  # theme(axis.title=element_text(size = 14, face = 'bold'),
        # plot.title=element_text(size = 14, face = 'bold'))
   NULL
 )

```



## Use null distribution to find critical values

+ As with all tests we need to set our $\alpha$: our probability of incorrectly rejecting the null hypothesis, even when it's true
	
	+ Let's set $\alpha$ = 0.05 (two tailed)
	+ This is a 5% risk of incorrectly rejecting the null

+ We'll use this $\alpha$ value, together with the null distribution from the last slide, to find the critical values that we'll compare our observed $t$-value to


## Critical values {.smaller}


```{r, echo = F, fig.align = 'center', message = F, fig.height = 5, fig.width = 10}

(p <- p + stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.025, 147), -4),
                alpha=.25,
                fill = "blue",
                args = list(df=147)) +
    stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.975, 147), 4),
                alpha=.25,
                fill = "blue",
                args = list(df=147)) +
  geom_vline(xintercept = qt(0.025, 147), linewidth = 1, linetype='dashed') +
  geom_vline(xintercept = qt(0.975, 147), linewidth = 1, linetype='dashed'))

```

+ Critical values $(t^*)$ establish a boundary for significance
+ The probability that a $t$-value will fall within these extreme regions of the distribution given $H_0$ is true is equal to $\alpha$
+ Because we are performing a two-tailed test, $\alpha$ is split between each tail, and the critical values are the edges of that shaded region (the dashed lines).

```{r}
c(
  lower = round(qt(0.025, 147), 3),
  upper = round(qt(0.975, 147), 3)
)
```

## How our $t$-value of 2.388 compares

```{r, echo = F, fig.align = 'center', message = F, fig.height = 5, fig.width = 10}

p + 
  geom_vline(xintercept = 2.388, colour = '#BF1932') +
  annotate(geom = 'text', x = 2.5, y = 0.3, label = 't = 2.388', hjust = 0, colour = dapr2red) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(2.388, 4),
                alpha= .7,
                fill = '#BF1932',
                args = list(df=147))

```


```{r include=F}
pt(2.388, 147, lower.tail=FALSE) * 2
```

<br>

The shaded red area, times 2 because it's a two-tailed test, gives us a $p$-value of .018.

$p = .018$ is less than $\alpha = 0.05$, so we can reject the null hypothesis that the association between `motivation` and `score` is equal to zero.


<!-- # Part 3: Alternative Approach: Using Confidence Intervals -->

<!-- ## Refresher: What is a Confidence Interval? -->

<!-- + When we perform these analyses, we obtain a parameter estimate from our sample (e.g. $\beta_2 = 0.92$) -->

<!-- + It's unlikely that the true value is exactly equal to our parameter estimate -->


<!-- + We can be much more certain we've captured the true value if we report **confidence intervals** -->

<!--   + Range of plausible values for the parameter -->

<!--   + The wider the range (that is, the higher the percentage of the confidence interval), the more confident we can be that our interval captures the true value  -->


<!-- ## Refresher: What is a Confidence Level? -->

<!-- + To create a confidence interval we must decide on a **confidence level** -->

<!--   + A number between 0 and 1 specified by us -->

<!--   + How confident do you want to be that the confidence interval will contain the true parameter value? -->

<!-- + Typical confidence levels are 90%, 95%, or 99% -->

<!-- . . .  -->

<!-- > **Test your understanding:** If we select a 90% confidence level, will the range of values included in our CI be smaller or larger than if we selected a 99% confidence level? -->


<!-- ##  Confidence Intervals for $\beta$ -->

<!-- + We can also compute confidence intervals for $\hat \beta$ -->

<!-- + The $100 (1 - \alpha)$, e.g., 95%, confidence interval for the slope is: -->

<!-- $$ -->
<!-- \hat \beta_1 \pm t^* \times SE(\hat \beta_1) -->
<!-- $$ -->

<!-- . . . -->

<!-- + So, the 95% confidence interval for the effect of `motivation` would be: -->

<!-- ```{r} -->
<!-- LowerCI <- round(0.91634 - (qt(0.975, 147) * 0.38376), 3) -->
<!-- LowerCI -->
<!-- ``` -->

<!-- ```{r} -->
<!-- UpperCI <-  round(0.91634 + (qt(0.975, 147)* 0.38376), 3) -->
<!-- UpperCI  -->
<!-- ``` -->

<!-- . . .  -->

<!-- + We can be 95% confident that the range 0.158 and 1.675 contains the true value of our $\beta_2$ -->



<!-- ## `confint` Function -->

<!-- + We can get confidence intervals using the `confint()` function: -->

<!-- ```{r} -->
<!-- confint(m1) -->
<!-- ``` -->


<!-- + The confidence intervals for both `motivation` and `hours` do not include the null value (in this case, 0)  -->

<!-- + This provides support (beyond $p<.05$) that **motivation and hours are statistically significant predictors of test scores** -->


# Part 3: Measure Model Quality using the Coefficient of Determination ( $R^2$ )

<!-- ## Revisit Model Output -->

<!-- ```{r} -->
<!-- m1 <- lm(score ~ hours + motivation, data = test_study2) -->
<!-- summary(m1) -->
<!-- ``` -->


## Quality of the Overall Model

- When we measure an outcome variable $y$, there will be variability in that outcome.

  - Some variability arises (we hope) because an independent variable $x$ impacts $y$.
    - This is variability our model can explain.
  - Other variability arises because there is just random variability in the world that we can't possibly account for.
    - This is variability our model *cannot* explain.

- We quantify how good our model is based on **what proportion of the variability in our data it can explain.**


## What proportion of variability the model can explain {.smaller}

To work out this proportion, we need to quantify three different kinds of variability:

<br>


| Type of variability in $y$         | How we quantify it                                                |
| ---------------------------------- | ----------------------------------------------------------------- |
| Variability explained by model     | $SS_{Model}$ <br> "Model sum of squares"                              |
| Variability not explained by model | $SS_{Residual}$ <br> "Residual sum of squares"                        |
| Total variability                  | $SS_{Model} + SS_{Residual} = SS_{Total}$ <br> "Total sum of squares" |

<br>

The proportion of variability that the model explains is then computed as

$$\frac{SS_{Model}}{SS_{Total}} \quad \text{or} \quad 1 - \frac{SS_{Residual}}{SS_{Total}}$$

We call this proportion the **coefficient of determination** and represent it as $R^2$.


## Interpreting the Coefficient of Determination 

$$
\begin{align}
R^2 &= \frac{SS_{Model}}{SS_{Total}}\\[1em]
  &=  1 - \frac{SS_{Residual}}{SS_{Total}} \\
\end{align}
$$

+ Quantifies the amount of variability in the outcome accounted for by the predictors
+ The more variance accounted for, the better the model fit
+ Represents the extent to which the prediction of $y$ is improved when predictions are based on the linear relation between $x$ and $y$, compared to not considering $x$

Let's see how to calculate each of these three sums of squares.



## Total Sum of Squares: Distance from mean

$$SS_{Total} = \sum_{i=1}^{n}(y_i - \bar{y})^2$$

In words: For every observation, find the difference between the observed value and the mean of all observed $y$ values, square that difference, and add up all the squares.

```{r, echo=FALSE, fig.height = 6, fig.width = 10, fig.align="center"}
test |>
  ggplot(aes(hours, score)) +
  geom_point(color='#BF1932', size = 3) +
  labs(x="Hours of Study", y = "Test Score") +
  ggtitle(latex2exp::TeX(r'($SS_{Total}$)')) + 
  geom_hline(aes(yintercept = mean(score)),color='#0F4C81', linewidth=2) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(rep(mean(score),10))), color = '#BF1932', lty = 2)
```


## Residual Sum of Squares: Distance from model

$$SS_{Residual} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2$$

In words: For every observation, find the difference between that observation and the value predicted by the model, square that difference, and add up all the squares.


```{r, echo=FALSE, fig.height = 6, fig.width = 10, fig.align="center"}
test |>
  ggplot(aes(hours, score)) +
  geom_point(color='#BF1932', size = 3) +
  labs(x="Hours of Study", y = "Test Score") +
  ggtitle(latex2exp::TeX(r'($SS_{Residual}$)')) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", linewidth=2) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(res$fitted.values)), color = '#BF1932', lty =2)
```



## Model Sum of Squares: Distance between model and mean

$$SS_{Model} = \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2$$

In words: For every observation, find the difference between the predicted value for that observation and the mean of all observed $y$ values, square that difference, and add them all up.

```{r, echo=FALSE, fig.height = 6, fig.width = 10, fig.align="center"}
test |>
  ggplot(aes(hours, score)) +
  # geom_point(color='#BF1932', size = 3) +  
  geom_blank() +
  labs(x="Hours of Study", y = "Test Score") +
  ggtitle(latex2exp::TeX(r'($SS_{Model}$)')) + 
  geom_hline(aes(yintercept = mean(score)),color='#0F4C81', linewidth=2) +
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", linewidth=2) + 
  geom_segment(aes(x = hours, y = res$fitted.values, xend = hours, yend = c(rep(mean(score),10))), color = '#BF1932', lty =2) +
  NULL
```



## Our model's $R^2$

+ In the current example, these values are:

  + $SS_\text{total}$ = 8556.06
  + $SS_\text{residual}$ = 2826.83
  + $SS_\text{model}$ = 5729.23



$$
R^2 = \frac{SS_{Model}}{SS_{Total}}
$$

. . . 

$$
\begin{align}
R^2 &= \frac{5729.23}{8556.06} \\[.5em]
  &= 0.6695 \\
\end{align}
$$

. . .

**$R^2$ = 0.6695 means that 66.95% of the variation in test scores is accounted for by hours of revision and student motivation**


## This $R^2$ = `lm()`'s Multiple R-squared! <br> (with some rounding error)

:::{style="font-size:80%;"}

```{r}
summary(m1)
```

:::


## What's Adjusted $R^2$?

+ When there are two or more predictors, $R^2$ tends to be an inflated estimate of the corresponding population value

  + Due to random sampling fluctuation, even when $R^2 = 0$ in the population, it's value in the sample may $\neq 0$ 

  + In **smaller samples,** the fluctuations from zero will be larger on average

  + With **more predictors,** there are more opportunities to add to the positive fluctuation


+ We therefore compute an adjusted $R^2$

$$
\hat R^2 = 1 - (1 - R^2)\frac{n-1}{n-k-1}
$$

+ Adjusted $R^2$ adjusts for both sample size $n$ and number of predictors $k$


## Our Multiple R-squared vs. Adjusted R-squared

<br>

```{r echo=F}
cat('', paste0(capture.output(
  summary(m1)
), '\n')[17:20])
```

+ **Based on adjusted R-squared, hours studying and student motivation explain `r round(summary(m1)$adj.r.squared,3)*100`% of the variance in test scores**

+ As the sample size is large and the number of predictors small, unadjusted (`r round(summary(m1)$r.squared,3)`) and adjusted R-squared (`r round(summary(m1)$adj.r.squared,3)`) are similar

+ If there were even more predictors, then adjusted R-squared would be a much more accurate representation of proportion variance explained


## Summary

+ We have an inferential test, based on a $t$-distribution, for the significance of $\beta$ 

+ We can compute confidence intervals that give us more certainty that we have captured the true value of $\beta$  

+ We are more likely to find a statistically significant effect when residuals are small and we have a large sample  

+ We can assess the degree to which our model explains variance in the outcome based on $R^2$  

+ When we have multiple predictors, we should use the adjusted $R^2$ to get a more conservative estimate


# Back matter

## This week {.smaller}

<br>

::::{.columns}
:::{.column width="50%"}
**Tasks:**

<br>

![](../figs/labs.svg){width=80px style="margin:10px;margin-bottom:-50px"} Work on exercises in labs

<br>

![](../figs/exam.svg){width=80px style="margin:10px;margin-bottom:-45px"} Complete the weekly quiz 


:::

:::{.column width="50%"}
**Get support:**

<br>

![](../figs/flashcards.svg){width=80px style="margin:10px;margin-bottom:-30px"}
Consult the [flash cards](https://uoepsy.github.io/dapr2/2627/flashcards/){target="_blank"}

<br>

![](../figs/forum.svg){width=80px style="margin:10px;margin-bottom:-50px"}
Ask questions anonymously on Piazza

<br>

![](../figs/oh.svg){width=80px style="margin:10px;margin-bottom:-40px"} 
We really like seeing you in office hours!

:::
::::


# Appendix {.appendix}

## Standard error: The maths {.smaller}

+ The formula for the standard error of the slope is:

$$
SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{j}})^2(1-R_{xj}^2)}}
$$

+ Where:
	+ $SS_{Residual}$ is the residual sum of squares
	+ $n$ is the sample size
	+ $k$ is the number of predictors
	+ $x_{ij}$ is the observed value of a predictor ( $j$ ) for an individual ( $i$ )
	+ $\bar{x_{j}}$ is the mean of a predictor
	+ $R_{xj}^2$ derives from the multiple correlation coefficient of the predictors

+ $R_{xj}^2$ captures to degree to which all of our predictors are related to each other
  + For simple linear models, $R_{xj}^2$ = 0 as there is only 1 predictor
  

## SE( $\hat \beta_j$ )

$$
SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{j}})^2(1-R_{xj}^2)}}
$$  

+ We want our $SE$ to be smaller - this means our estimate is precise


+ Examining the above formula we can see that:
	+ $SE$ is smaller when residual variance ( $SS_{residual}$ ) is smaller
	+ $SE$ is smaller when sample size ( $n$ ) is larger
	+ $SE$ is larger when the number of predictors ( $k$ ) is larger
	+ $SE$ is larger when a predictor is strongly correlated with other predictors ( $R_{xj}^2$ )


