---
title: "F-Ratio & Model Comparisons"
author: "Emma Waterston"
format: html
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(simglm)
library(kableExtra)

set.seed(7284) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1)),
  error = list(variance = 20),
  sample_size = 150,
  reg_weights = c(0.6, 1.4, 1.5)
)

df <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study2 <- df %>%
  dplyr::select(y, hours, motivation) %>%
  mutate(
    ID = paste("ID", 101:250, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2)
  ) %>%
  dplyr::select(ID, score, hours, motivation)

```

In this document we are going to work through the hand calculations for:
  
- $F$-statistic 
- Incremental $F$-test

## Packages

For this example we will only need the `tidyverse`package.

```{r, message=FALSE, warning=FALSE}

```

## Data

```{r}

```


We will make our calculations for the model we have been using in class this week:

```{r}

```
 
## $F$-Statistic

$$
F = \frac{\frac{SS_{model}}{df_{model}}}{\frac{SS_{residual}}{df_{residual}}} = \frac{MS_{Model}}{MS_{Residual}}
$$

### Sums of Squares Residual

$$
SS_{Residual} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2
$$


```{r}

```

As we did previously, we can then sum this final column to give the sums of squares residual:

```{r}

```

### Sums of Squares Model

$$
SS_{Model} = \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2
$$

```{r}

```


```{r}

```

### DF Model

$$
df_\text{model} = k = 2
$$

### DF Residual

$$
df_\text{residual} = n - k - 1= 150 - 2 - 1 = 147
$$

### Back to $F$-Statistic

$$
\begin{align}
F = \frac{\frac{SS_{model}}{df_{model}}}{\frac{SS_{residual}}{df_{residual}}} = \frac{MS_{Model}}{MS_{Residual}} \\
\\
F = \frac{\frac{5728.79}{2}}{\frac{2826.83}{147}} = \frac{2864.40}{19.23} \\
\\
F = 148.95
\end{align}
$$

We can compare this to the value at the bottom of our model `summary()`:

```{r}

```

If we wanted to save this specific value from the summary output, we can do the following:

```{r}

```


## Model Comparisons

**Research Question: Does motivation significantly predict score over and above the effects of study hours?**

Note that this is not a very useful model comparison (because the two models being compared only differ by one predictor).
But it works to illustrate the mechanics of doing model comparison in R.

### Step 1: Fit & Run Models

```{r}

```

### Step 2: Statistically Compare Models

#### Using `anova()`

Note: we are using the `anova()` function, but we are not running an ANOVA.
We must report this as a "model comparison", not an ANOVA.

```{r}

```

#### By Hand (ADVANCED/OPTIONAL)

We can extract the information we need from the `summary()` output given the relationship between the residual sum of squares and residual standard error:

$$
\text{SS}_\text{Residual} = \text{Residual Standard Error}^2 \cdot df_\text{Residual}
$$

Therefore we can calculate our $SS_\text{Residual}$ as follows:

*Restricted Model*

```{r}

```

$$
\begin{align} 
SSR_R = (4.4547^2) \cdot 148 \\  
\\  
SSR_R = 2936.964 \\
\end{align}
$$

*Full Model*

```{r}

```

$$
\begin{align}
SSR_F = (4.3856^2) \cdot 147 \\  
\\  
SSR_F = 2827.323 \\
\end{align}
$$

<br> 

Once we have obtained these values, we can substitute into the formula:

$$
\begin{align}
F_{(df_R-df_F),df_F} = \frac{(SSR_R-SSR_F)/(df_R-df_F)}{SSR_F / df_F} \\
\\
F_{(148-147),147} = \frac{(2936.964-2827.323)/(148-147)}{2827.323 / 147} \\
\\
F_{1,147} = \frac{109.641/1}{ 19.2335} \\
\\
F_{1,147} = \frac{109.641}{19.2335} \\
\\
F_{1,147} = 5.70
\end{align}
$$


### Step 3: Write Up

Motivation was found to explain a significant amount of variance in scores over and above the effects of study hours alone $F(1, 147) . = 5.70, p < .05)$.
