---
title: "Exploring data visually"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(psych)
```


Visual exploration of our data allows us to visualize the distributions of our data, and to identify potential associations among variables.

## How to Visualise Data

To visualise (i.e., plot) our data, we can use `ggplot()` from the **tidyverse** package. Note the key components of the `ggplot()` code:

+ `data =` where we provide the name of the dataframe
+ `aes =` where we provide the aesthetics. These are things which we map from the data to the graph. For instance, the $x$-axis, or if we wanted to colour the columns/bars according to some aspect of the data
+ `+ geom_...` = where we add (using `+`) some geometry. These are the shapes (e.g., bars, points, etc.), which will be put in the correct place according to what we specified in `aes()`  
+ `labs()` = where we provide labels for our plot (e.g., the $x$- and $y$-axis)

:::{.callout-note}

There are lots of arguments that you can further customise, some of which are specified in the examples below e.g., `bins = `, `alpha = `, `fill = `, `linewidth = `. `linetype = `, `size = ` etc. For these, you can look up the helper function to see the range of arguments they can take using `?` - e.g., `?fill`.

If you'd like to read more about `ggplot()`, there is a handy [cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf).

:::

One other thing to consider when visualising your data is how you are going to arrange your plots. Some handy tips on this:  
  
- Use \n to wrap text in your titles and or axis labels  
- The **patchwork** package allows us to arrange multiple plots in two ways - `|` arranges the plots adjacent to one another, and `/` arranges the plots on top of one another  


<br> 

## Marginal distributions – Examples

::: {.panel-tabset}

#### Histogram

A histogram shows the frequency of values which fall within bins of an equal width. 

**Basic:**

- x-axis: possible values of some variable, grouped into bins
- y-axis: frequency of a given value or values within bins
- *What are bins?*: A bin represents a range of scores

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_histogram() +
    labs(x = "Sepal Length (in cm)")
```


**Updating Bins:**

Within `geom_histogram()`, we can specify `bins = ` to specify the number of columns we want (for this example, lets say we want 10):

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_histogram(bins = 10) +
    labs(x = "Sepal Length (in cm)")
```

Alternatively, we can specify `binwidth = ` to specify the width of each bin (it is very helpful to be aware of the scale of your variable here!):

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_histogram(binwidth = 0.1) +
    labs(x = "Sepal Length (in cm)")
```


**Outline Columns with Color:**

Within `geom_histogram()`, we can specify `color = ` to set a colored outline of the columns:

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_histogram(color = "darkred") +
    labs(x = "Sepal Length (in cm)")
```


**Fill Columns with Color:**

Within `geom_histogram()`, we can specify `fill = ` to fill the columns with a color:

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_histogram(fill = "darkred") +
    labs(x = "Sepal Length (in cm)")
```

#### Density

“Density” is a bit similar to the notion of “relative frequency” (or “proportion”), in that for a density curve, the values on the y-axis are scaled so that the total area under the curve is equal to 1. In creating a curve for which the total area underneath is equal to one, we can use the area under the curve in a range of values to indicate the proportion of values in that range.

**Basic:**

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_density() +
    labs(x = "Sepal Length (in cm)")
```

**Filled:**

We can fill our plot with colour by specifying `fill = ` within `geom_density()`:

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_density(fill = "darkred") +
    labs(x = "Sepal Length (in cm)")
```

**Line Type & Width:**

We can change the type and width of the line by specifying `linetype = ` and `linewidth = ` within `geom_density()`:

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
    geom_density(linetype = 6, linewidth = 3) +
    labs(x = "Sepal Length (in cm)")
```

#### Boxplot

Boxplots provide a useful way of visualising the __interquartile range (IQR).__ You can see what each part of the boxplot represents in Figure @fig-boxplotdesc.  


```{r}
#| label: fig-boxplotdesc
#| echo: false
#| fig-cap: "Anatomy of a Boxplot"
#| out-width: '100%'

set.seed(34875)
px = rnorm(100,10,10)
mn=min(px[px>quantile(px, .25)-(1.5*IQR(px))])
mx=max(px[px<quantile(px, .75)+(1.5*IQR(px))])
outs_up = px[px>quantile(px, .75)+(1.5*IQR(px))]
outs_lw = px[px<quantile(px, .25)-(1.5*IQR(px))]
ggplot(data = NULL, aes(x = px)) +
  geom_boxplot(col="red")+
  ylim(-1,1)+xlim(-30,45)+
  theme_classic()+
  theme(axis.title = element_blank(), axis.text = element_blank(),
      axis.ticks = element_blank(), axis.line = element_blank())+
  
  # Quartiles
  annotate("text",x=median(px), y=0.7, label="Q2\n(Median)", vjust=1,hjust=0.5, col="grey70")+
  annotate("text",x=quantile(px, .25), y=0.6, label="Q1", vjust=1,hjust=1, col="grey70")+
  annotate("text",x=quantile(px, .75), y=0.6, label="Q3", vjust=1,hjust=0, col="grey70")+
  geom_segment(aes(x=median(px), xend=median(px), y=0.48, yend=0.38), col="grey70",lty="dashed")+
  geom_segment(aes(x=quantile(px, .25), xend=quantile(px, .25), y=0.48, yend=0.38), col="grey70",lty="dashed")+
  geom_segment(aes(x=quantile(px, .75), xend=quantile(px, .75), y=0.48, yend=0.38), col="grey70",lty="dashed")+
  
  # Whiskers
  annotate("text",x=mn, y=-0.2, label="Minimum value in data\nwhich is > Q1 - (1.5 * IQR)", vjust=1,hjust=0.5, col="grey70")+
  annotate("text",x=mx, y=-0.2, label="Maximum value in data\nwhich is < Q3 + (1.5 * IQR)", vjust=1,hjust=0.5, col="grey70")+
  geom_segment(aes(x=mn, xend=mn, y=0, yend=-0.2),col="grey70",lty="dashed")+
  geom_segment(aes(x=mx, xend=mx, y=0, yend=-0.2),col="grey70",lty="dashed")+
  
  #IQR
  geom_segment(aes(x=quantile(px, .25), xend=quantile(px, .25), y=0.78, yend=0.85), col="grey70")+
  geom_segment(aes(x=quantile(px, .75), xend=quantile(px, .75), y=0.78, yend=0.85), col="grey70")+
  geom_segment(aes(x=quantile(px, .75), xend=quantile(px, .25), y=0.85, yend=0.85), col="grey70")+
  annotate("label",x=median(px), y=0.9, label="IQR", vjust=1,hjust=0.5, col="grey70")+
  
  # Outliers +
  geom_segment(aes(x=outs_up, xend=mean(outs_up), y=0, yend=0.27), col="grey70", lty="dashed")+
  annotate("text",x=mean(outs_up), y=0.5, label="Outliers\nDatapoints > Q3 + (1.5 * IQR)", vjust=1,hjust=0.5, col="grey70")+
  # Outliers -
  geom_segment(aes(x=outs_lw, xend=mean(outs_lw), y=0, yend=0.27), col="grey70", lty="dashed")+
  annotate("text",x=mean(outs_lw), y=0.5, label="Outliers\nDatapoints < Q1 - (1.5 * IQR)", vjust=1,hjust=0.5, col="grey70")+
  
  NULL
  
rm(mn,mx,outs_lw,outs_up,px)
```


**Basic:**

We need to specify `+ geom_boxplot()` to get a boxplot: 

```{r}
ggplot(data = iris, aes(x = Sepal.Length)) +
  geom_boxplot() +
    labs(x = "Sepal Length (in cm)")
```

**Rotate Boxplot:**

If we had set `aes(y = Sepal.Length)` instead, then it would simply be rotated 90 degrees:

```{r}
ggplot(data = iris, aes(y = Sepal.Length)) +
  geom_boxplot() +
    labs(y = "Sepal Length (in cm)")
```

:::


<br> 

## Bivariate associations – Examples

Unlike in our marginal plots where we specified our x-axis variable within `aes()`, to visualise bivariate associations, we need to specify what variables we want on both our x- and y-axis. 

::: {.panel-tabset}

#### Scatterplot

We can use a scatterplot (since the variables are numeric and continuous) to visualise the association between the two numeric variables - these will be our x- and y-axis values. 

We plot these values for each row of our dataset, and we should end up with a *cloud* of scattered points. 

Here we will want to comment on any key observations that we notice, including if we detect outliers or points that do not fit with the pattern in the rest of the data. Outliers are extreme observations that are not possible values of a variable or that do not seem to fit with the rest of the data. This could either be:

+ *marginally* along one axis: points that have an unusual (too high or too low) x-coordinate or y-coordinate
+ *jointly*: observations that do not fit with the rest of the point cloud 

**Basic:**

We need to specify `+ geom_point()` to get a scatterplot: 

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point() +
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

**Fill points with color:**

Within `geom_point()`, we can specify `color = ` to fill the points with a color:

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point(color = "darkred") +
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

**Change size and opacity:**

We can change the size (using `size =`) and the opacity (using `alpha =`) of our geom elements on the plot. Let’s include this below:

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point(size = 3, alpha = 0.5) +
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

**Add a line of best fit:**

We can superimpose (i.e., add) a line of best fit by including the argument `+ geom_smooth()`. Since we want to fit a straight line, we want to use `method = "lm"`. We can also specify whether we want to display confidence intervals around our line by specifying `se = TRUE / FALSE`. 

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```


#### Scatterplot of Matrices (SPLOM)

Using `pairs.panels()` is likely the most useful way to visualise the associations among numeric variables. It returns a scatterplot of matrices (SPLOM) returning you (1) the marginal distribution of each variable via a histogram, (2) the correlation between variables, and (3) bivariate scatterplots. 

```{r}
#| label: fig-pairs-p
#| fig-cap: "Iris SPLOM"

iris |>
    select(Sepal.Length, Petal.Length, Petal.Width) |>
    pairs.panels(main = "Iris SPLOM")
```

#### Boxplot

We can use a boxplot to visualise the association between one numeric variable and one categorical variable - these will be our y- and x-axis values respectively. This can be helpful to visually compare the distribution of multiple groups.

**Basic:**

We need to specify `+ geom_boxplot()` to get a boxplot: 

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot() +
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Change boxplot fill colours by group:**

Within `aes()`, we can specify `fill = ` to fill the boxes with a color:

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
    geom_boxplot() +
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Change boxplot line colours by group:**

Within `aes()`, we can specify `color = ` to colour the lines with a color:

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, color = Species)) +
    geom_boxplot() +
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Adding jitter:**

We can add jittered points to a boxplot to better see the underlying distribution of the data (by adding a little random variation to each data point) via `geom_jitter()`: 

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, color = Species)) +
    geom_boxplot() +
    geom_jitter() + 
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Change legend position:**

We can add the argument `+ theme(legend.position = )` to move (or even remove) the legend by specifying, for example, ` "right" `, ` "left" `, ` "top" `, ` "bottom" `, or  `"none" ` to remove. 

```{r}
# legend at bottom of plot
ggplot(data = iris, aes(x = Species, y = Sepal.Length, color = Species)) +
    geom_boxplot() +
    labs(x = "Species", y = "Sepal Length (in cm)") + 
    theme(legend.position = "bottom")
```


#### Violin Plot

Like a boxplot, we can use a violin plot to visualise the association between one numeric variable and one categorical variable - these will be our y- and x-axis values respectively. It combines a summary of the data's range and a kernel density estimation, providing a detailed view of the distribution. 
  
**Basic:**

We need to specify `+ geom_violin()` to get a violin plot: 

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
    geom_violin() +
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Change violin colours by group:**

Within `aes()`, we can specify `fill = ` to fill the violins with a color:

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
    geom_violin() +
    labs(x = "Species", y = "Sepal Length (in cm)")
```


**Change Size and opacity:**

We can change the size (using `size =`) and the opacity (using `alpha =`) of our geom elements on the plot. Let’s include this below: 

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
    geom_violin(size = 1, alpha = 0.5) +
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Adding jitter:**

We can add jittered points to a violin plot to better see the underlying distribution of the data (by adding a little random variation to each data point) via `geom_jitter()`: 

```{r}
ggplot(data = iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
    geom_violin() + 
    geom_jitter(alpha = 0.5) + 
    labs(x = "Species", y = "Sepal Length (in cm)")
```

**Change legend position:**

We can add the argument `+ theme(legend.position = )` to move (or even remove) the legend by specifying, for example, ` "right" `, ` "left" `, ` "top" `, ` "bottom" `, or  `"none" ` to remove. 

```{r}
# no legend
ggplot(data = iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
    geom_violin() + 
    geom_jitter(alpha = 0.5) + 
    theme(legend.position = 'none') +
    labs(x = "Species", y = "Sepal Length (in cm)")
```


#### Facets

When we have two numeric variables, as well as categorical variables, we can use `facet_wrap()` / `facet_grid()` to help divide/arrange our plots. If we had two categorical variables, by simply stringing them together to further group our plots by specifying `facet_wrap( ~ cat_variable1 + cat_variable2)`

**Basic:**

We need to specify `+ geom_point()` to get a scatterplot, and *either* `+ facet_wrap()` or  `+ facet_grid()` to separate by your categorical variable: 

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point() +
    facet_wrap(~Species) + 
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

**Add a line of best fit:**

We can superimpose (i.e., add) a line of best fit by including the argument `+ geom_smooth()`. Since we want to fit a straight line, we want to use `method = "lm"`. We can also specify whether we want to display confidence intervals around our line by specifying `se = TRUE / FALSE`. Note that a line is fitted for every level of your categorical variable:

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
        facet_wrap(~Species) + 
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

**Subplot layout:**

You can change the overall layout of the subplots by specifying `dir = ` within the `facet_wrap()` argument, where `“h”` will return a horizontal layout (this is the default) and `“v”` for vertical.

You can also change the layout of the subplot labels by specifying `strip.position = ` within the `facet_wrap()` argument, where labels can be arranged to display at the `“top”` (this is the default), `“bottom”`, `“left”` or `“right”`. 

```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Sepal.Length)) +
    geom_point() +
    facet_wrap(~Species, dir = "v", strip.position = "right") + 
    labs(x = "Petal Length (in cm)", y = "Sepal Length (in cm)")
```

:::

