---
title: "Functions and mathematical models"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


Basic functions and mathematical models are foundational tools used to describe and predict associations between variables.

`r optbegin("Identification & Specification", olabel=FALSE,toggle=params$TOGGLE)`

Consider the function $y = 2 + 5 \ x$. From this, we can do the following:

- Identify the Dependent Variable (DV)
- Identify the Independent Variable (IV)
- Describe in words what the function does, and compute the output for the following input:

$$
x = \begin{bmatrix}
2 \\
6
\end{bmatrix}
$$

The function says that the $y$ value is obtained as a transformation of the $x$ value.

- The dependent variable is $y$
- The independent variable is $x$
- The $y$ value is calculated as two plus five times $x$

<br />
Example (1): If $x$ equals 2, the corresponding value of $y$ will be $2 + (5 \cdot 2) = 12$.
<br />
Example (2): If $x$ equals 6, the corresponding value of $y$ will be $2 + (5 \cdot 6) = 32$.

$$
y = \begin{bmatrix}
2 + 5 \cdot 2 \\
2 + 5 \cdot 6
\end{bmatrix}
= \begin{bmatrix}
12 \\
32
\end{bmatrix}
$$

`r optend()`

## Deterministic Models

`r optbegin("Description & Specification", olabel=FALSE,toggle=params$TOGGLE)`

We come across functions *a lot* in daily life, and probably don't think much about it. In a slightly more mathematical setting, we can write down in words and in symbols the function describing the association between the side of a square and its perimeter (e.g., to capture *how* the perimeter *varies* as a *function* of its side). In this case, the perimeter is the dependent variable, and the side is the independent variable.

This is what we would refer to as a *deterministic model*, as it is a model of an *exact relationship* - there can be no deviation.

**Model Specification**

::: {.panel-tabset}

## In words

The perimeter of a square is four times the length of its side.


## In symbols

The relationship between side and perimeter of squares is given by:

$$
\text{Perimeter} = 4 \cdot \text{Side}
$$

If you denote $y$ as the dependent variable _Perimeter_, and $x$ as the independent variable _Side_ we can rewrite as: 

$$
y = 4 \cdot x
$$
:::

`r optend()`

<br>

`r optbegin("Visualisation", olabel=FALSE,toggle=params$TOGGLE)`

Let's create a dataset called `squares`, containing the perimeter of four squares having sides of length $0, 2, 5, 9$ metres, and then plot the `squares` data as points on a scatterplot. 

First, let's make our squares data. Here we will use two important functions - `tibble()` and `c()`. The `tibble()` function allows us to construct a data frame. To store a sequence of numbers into R, we can **c**ombine the values using `c()`. A sequence of elements all of the same type is called a *vector*.

```{r}
#create data frame named squares
squares <- tibble(
  side = c(0, 2, 5, 9), 
  perimeter = 4 * side
)

#check that our values are contained within squares
squares
```

Now we know how `ggplot()` works, we can start to build our plot. First we specify our data (we want to use the *squares* data frame), and then our aesthetics. Since the perimeter varies as a function of side, we want side on the $x$-axis, and perimeter on the $y$-axis. We want to create a scatterplot, so we need to specify our `geom_...` argument as `geom_point()`. Lastly, we will provide clearer axis labels, and include the units of measurement. 

```{r}
#| label: fig-squares1
#| fig-cap: "Perimeter = 4*Side" 

ggplot(data = squares, aes(x = side, y = perimeter)) +
  geom_point() +  
  labs(x = 'Side (m)', y = 'Perimeter (m)',  title = 'Perimeter = 4*Side')  
```

We could also visualise the functional relationship by connecting the individual points with a line. To do so, we need to add a new argument - `geom_line()`. If you would like to change the colour of the line from the default, you can specify `geom_line(colour = "insert colour name")`. 
```{r}
#| label: fig-squares2
#| fig-cap: "Perimeter = 4*Side" 

ggplot(data = squares, aes(x = side, y = perimeter)) +
  geom_point() +
  geom_line(colour = "darkred") + 
  labs(x = 'Side (m)', y = 'Perimeter (m)',  title = 'Perimeter = 4*Side')  
```

`r optend()`

<br>

`r optbegin("Predicted Values", olabel=FALSE,toggle=params$TOGGLE)`

Sometimes we can directly read a predicted value from the graph of the functional relationship.

Consider the plot created above. For example, first we need to check where $x$ = 2.5. Then, we draw a vertical dashed line until it meets the blue line. The $y$ value corresponding to $x$ = 2.5 can be read off the $y$-axis. In our case, we would say a side of 2.5m corresponds to a perimeter of 10m. 

```{r}
#| label: fig-squares3
#| fig-cap: "Perimeter = 4*Side" 

ggplot(data = squares, aes(x = side, y = perimeter)) +
  geom_point() +
  geom_line(colour = "blue") + 
  geom_vline(xintercept = 2.5, colour = "darkred", lty = "dashed", lwd = 1) + 
  labs(x = 'Side (m)', y = 'Perimeter (m)',  title = 'Perimeter = 4 * Side')  
```

However, in this case it is not that easy to read it from @fig-squares3 (especially without the superimposed dashed red line)... This leads us to the algebraic approach:

We can substitute the $x$ value in the formula and calculate the corresponding $y$ value where we would conclude that the predicted perimeter of squared paintings having a 2.5m side is 10m:

$$
y = 4 \cdot x  \\    
$$

$$
y = 4 \cdot 2.5 \\  
$$

$$
y = 10  \\
$$

`r optend()`

`r optbegin("hello my optional friend")`

it's nice to see you again

`r optend()`


## Linked flash cards

### Outgoing links

- TODO


### Backlinks

- TODO
