---
title: "Linear functions"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

To describe and predict how variables are associated, a key mathematical tool is the **function.**

A function is a mathematical expression that 

- takes some value as input,
- transforms that input in some specified way, and
- returns a single value as output.

In DAPR2, we focus on **linear functions**, which have the form $y = c + mx$.

- $y$: outcome variable, dependent variable
- $c$: y-intercept (where the line intersects the y-axis, or the value of the function when $x = 0$)
- $m$: the slope of the line
- $x$: input variable, independent variable

Take, for example, the function $y = 2 + 5x$.
This function describes a line with a slope of 5 and a y-intercept (in stats we just call it an "intercept") of 2.
That function looks like this:

```{r}
#| code-fold: true

len_out <- 10

p_line <- tibble(
  x = seq(0, 10, length.out = len_out),
  y = seq(0, 40, length.out = len_out)
) |>
  ggplot(aes(x = x, y = y)) +
  geom_blank() +
  geom_abline(intercept = 2, slope = 5) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(breaks = seq(0, 10, 2), expand = c(0,0)) +
  NULL

p_line
```


Here's how we use this linear function to predict $y$ based on $x$:

- It takes some value of $x$ as input.
- It multiplies that value by 5.
- It adds 2 to that product.
- It returns that sum as $y$, the outcome.


Example 1: $x = 2$

$$
\begin{align}
y &= 2 + 5(2) \\
&= 2 + 10 \\
&= 12
\end{align}
$$
```{r}
#| code-fold: true
p_line +
  # arrow from x to line
  geom_segment(x = 2, xend = 2, y = 0, yend = 12, colour = dapr2red, arrow = arrow(length = unit(0.5,"cm"))) +
  geom_text(x = 2 + 0.1, y = 6, label = 'if x = 2 ...', colour = dapr2red, hjust = 0) +
  # arrow from line to y
  geom_segment(x = 2, xend = 0, y = 12, yend = 12, colour = dapr2red, arrow = arrow(length = unit(0.5,"cm"))) +
  geom_text(x = 1, y = 12 + 3, label = '... then y = 12', colour = dapr2red, hjust = 0.5) + 
  NULL
```

Example 2: $x = 6$

$$
\begin{align}
y &= 2 + 5(6) \\
&= 2 + 30 \\
&= 32
\end{align}
$$

```{r}
#| code-fold: true
p_line +
  # arrow from x to line
  geom_segment(x = 6, xend = 6, y = 0, yend = 32, colour = dapr2red, arrow = arrow(length = unit(0.5,"cm"))) +
  geom_text(x = 6 + 0.1, y = 16, label = 'if x = 6 ...', colour = dapr2red, hjust = 0) +
  # arrow from line to y
  geom_segment(x = 6, xend = 0, y = 32, yend = 32, colour = dapr2red, arrow = arrow(length = unit(0.5,"cm"))) +
  geom_text(x = 3, y = 32 + 3, label = '... then y = 32', colour = dapr2red, hjust = 0.5) + 
  NULL
```

Because you always get the same output every time you give the function the same input, a linear function like this is an example of a [deterministic function](deterministic-probabilistic.html){target="_blank"}.
