---
title: "Diagnostics for influential observations"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

Just like in standard linear regression, we can check for influential observations using the measure of Cook's Distance.

For logistic regression, the thresholds we typically use are the following:

- Cook's Distance > 0.5 indicates moderately influential values.
- Cook's Distance > 1 indicates highly influential values.


Let's check for influential observations in our senility symptoms example:

```{r}
sendata <- read_csv("https://uoepsy.github.io/data/SenilityWAIS.csv")
sen_mdl1 <- glm(senility ~ wais, family = "binomial", data = sendata)
```


To get the Cook's Distance of each observation, we can use `cooks.distance()`:

```{r}
cooks.distance(sen_mdl1)
```


We can visually assess these values by specifying `which = 4` when plotting your fitted model (i.e., `plot(model, which = 4)`.

```{r}
plot(sen_mdl1, which = 4)
```

All observations have Cook's Distance values below 0.5, so we have no unduly influential observations in this analysis.
