---
title: "Checking assumptions"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

Logistic regression models do not make the same assumptions as standard linear models.
There's only one assumption we need to check: **no outlying standardised deviance residuals.**

- An observation's standardised deviance residual is small if the model would predict with high probability the same outcome that we actually observed (e.g., if the model thinks a success is very likely, and the observation is indeed a success).
- And an observation's standardised deviance residual is large if the model would predict with high probability the *opposite* outcome (e.g., if the model thinks a success is very likely, but the observation is actually a failure).

::: {.callout-caution collapse="true"}
### What exactly are deviance residuals?

If you're curious about how deviance residuals are computed and why we use them, check out [this really fantastic explainer from the University of Virginia](https://library.virginia.edu/data/articles/understanding-deviance-residuals){target="_blank"}.

:::

If the model fits well, then standardised deviance residuals roughly follow a standard normal distribution (i.e., a normal distribution with mean 0 and SD 1).
Therefore we'd expect the following:

- About 5% of standardised deviance residuals may be smaller than –2 or larger than 2.
- About 0.3% of standardised deviance residuals may be smaller than –3 or larger than 3.

**If our model contains any standardised deviance residuals smaller than –3 or larger than 3, then we should be a bit suspicious that the data we're modelling may not be suitable for the model.**

Let's check this using our senility symptoms example:

```{r}
sendata <- read_csv("https://uoepsy.github.io/data/SenilityWAIS.csv")
sen_mdl1 <- glm(senility ~ wais, family = "binomial", data = sendata)
```


We obtain standardised deviance residuals using `rstandard()` with `type = 'deviance'`.

```{r}
rstandard(sen_mdl1, type = 'deviance')
```

Let's plot these to visually check whether we've got any values larger than 3 in absolute value (i.e., smaller than –3 or larger than 3):

```{r}
plot(rstandard(sen_mdl1, type = "deviance"), ylab = "Standardised deviance residuals")
```

No standardised deviance residuals with values more extreme than –3 or 3, so it looks like we can be satisfied that the logistic regression model's assumptions have been met.