---
title: "Logistic regression example"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(sjPlot)
```

> **Research Question** 
>
> Does the probability of elderly adults having symptoms of senility change as a function of their score on the Wechsler Adult Intelligence Scale?


## Overview

A small sample ($n$ = 54) of elderly people were given a psychiatric examination to determine if symptoms of senility were present (`senility`: 0 = not present; 1 = present).
Other measurements taken at the same time included the score on a subset of the Wechsler Adult Intelligence Scale (`wais`).

```{r}
sendata <- read_csv("https://uoepsy.github.io/data/SenilityWAIS.csv")
head(sendata)
```


<br>

## Visualise Data

Just the observed data:

```{r}
sendata |>
  ggplot(aes(x = wais, y = senility)) +
  
  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # rename the axes for clarity
  labs(
    y = 'Presence of senility symptoms',
    x = 'WAIS score'
  )
```

And we can also add a line of best fit using `+ geom_smooth()`.
But this shouldn't be a straight line, as we know.
To add a logistic curve, we specify `method = 'glm'` and specify the binomial family using `method.args = list(family = binomial)`.
And to get rid of the ribbon showing the standard error, we include `se = FALSE`.

```{r}
sendata |>
  ggplot(aes(x = wais, y = senility)) +
  
  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # rename the axes for clarity
  labs(
    y = 'Presence of senility symptoms',
    x = 'WAIS score'
  ) +
  
  # add logistic curve
  geom_smooth(
    method = "glm",  # not a straight line
    method.args = list(family = binomial),   # specify logistic curve
    se = FALSE  # remove standard error ribbon around line
  )
```

Looks like the probability of senility symptoms decreases as WAIS score increases.

To see if this decrease is statistically significant, let's model it.


## Mathematical model formulation

<!-- We're modelling the log-odds of senility symptoms as a function of WAIS. -->

$$
\begin{aligned}
\log \left( \frac{p}{1 - p}\right) = \beta_0 + (\beta_1 \cdot \text{WAIS})
\end{aligned}
$$

where $p$ is the probability of having senility symptoms.


## State hypotheses

Same as always!

$$
\begin{align}
H_0 &: \beta_1 = 0 \\
H_1 &: \beta_1 \neq 0\\
\end{align}
$$

## Fit model

```{r}
sen_mdl1 <- glm(
  senility ~ wais, 
  family = "binomial",   # this tells glm() to use logistic regression
  data = sendata
)
```


```{r}
summary(sen_mdl1)  # check out model summary
confint(sen_mdl1)  # get confidence intervals for the coefficient
```


## Interpreting the coefficients

<!-- Our outcome variable is the log-odds of having symptoms of senility, so all of these model coefficients are on the scale of log-odds. -->

### `(Intercept)` = $\beta_0$

For people with a WAIS score of 0, the estimated log-odds of having symptoms of senility is 2.4.

This estimate is significantly different from 0 (95% CI: [0.21, 4.98], p = 0.04).

**For intercepts only (not for slopes), we can back-transform the coefficient estimate to probability space:**

```{r}
plogis(2.4)
```

So in other words, for people with a WAIS score of 0, the estimated probability of having symptoms of senility is 91.7%.



#### Why are we interpreting the significance of the intercept now?

When the outcome scale is log-odds, it's actually interesting if the intercept is significantly different from zero!

A log-odds of zero equals a probability of 0.5, which is the same as 50/50 chance, like flipping a coin.

**So if our intercept is significantly different from zero, then we know that the log-odds of a successful outcome (whatever that is for the given model) is significantly different from chance.**

For the senility model, the significant intercept tells us that people with a WAIS score of 0 have a probability of showing senility symptoms that's significantly higher than chance.


### `wais` = $\beta_1$

Increasing WAIS score by one point is associated with the log-odds of senility symptoms decreasing by 0.32.

This decrease is significantly different from zero (95% CI [–0.58, –0.12], p = 0.005).

We cannot back-transform the slopes to probabilities the way we can with the intercept.


### Why can we back-transform the intercept but not the slopes to probabilities?

We can only back-transform the intercept into probabilities because the intercept represents the log-odds of something happening.
In contrast, the slope represents a *change* in log-odds, or differences between two log-odds values.
And because the logit link function is not linear, the differences between two log-odds values mean different things, depending on what the values are that we're comparing.

::: {.callout-caution collapse="true"}
#### More detail on why we can't back-transform slopes

For example, imagine a slope coefficient of 1, that is, a difference in log-odds of one unit.

If that difference is between a log-odds of 0 and a log-odds of 1, then it maps to a difference in probabilities of 23%:

```{r}
plogis(1) - plogis(0)
```

But if that difference is between a log-odds of 3 and a log-odds of 4, then it maps to a difference in probabilities of just 3%:

```{r}
plogis(4) - plogis(3)
```


To give a visual intuition:

```{r}
#| code-fold: true

tibble(
  logodds = seq(-5, 5, length.out = 100),
  prob = plogis(logodds)
) |>
  ggplot(aes(x = logodds, y = prob)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), breaks = -4:4) +
  labs(x = 'Log-odds', y = 'Probability') +
  theme(panel.grid = element_blank()) +
  # horiz green
  geom_segment(x = 0, y = plogis(0), xend = -5, yend = plogis(0), colour = '#2B8654') +
  geom_segment(x = 1, y = plogis(1), xend = -5, yend = plogis(1), colour = '#2B8654') +
  geom_text(
    x = -4.8,
    y = 0.62,
    colour = '#2B8654',
    label = paste0('diff: ', 100*round(plogis(1)-plogis(0), 2), '%'),
    hjust = 0,
    size = 5
  ) +
  # vert green
  geom_segment(x = 0, y = 0, xend = 0, yend = plogis(0), colour = '#2B8654') +
  geom_segment(x = 1, y = 0, xend = 1, yend = plogis(1), colour = '#2B8654') +
  geom_text(
    x = 0.5,
    y = 0.15,
    colour = '#2B8654',
    label = paste0('diff:\n 1 log-\nodd'),
    size = 5
  ) +
  
  # horiz pink
  geom_segment(x = 3, y = plogis(3), xend = -5, yend = plogis(3), colour = '#FF6D9A') +
  geom_segment(x = 4, y = plogis(4), xend = -5, yend = plogis(4), colour = '#FF6D9A') +
  geom_text(
    x = -4.8,
    y = 0.91,
    colour = '#FF6D9A',
    label = paste0('^ diff: ', 100*round(plogis(4)-plogis(3), 2), '%'),
    hjust = 0,
    size = 5
  ) +
  # vert pink
  geom_segment(x = 3, y = 0, xend = 3, yend = plogis(3), colour = '#FF6D9A') +
  geom_segment(x = 4, y = 0, xend = 4, yend = plogis(4), colour = '#FF6D9A') +
  geom_text(
    x = 3.5,
    y = 0.15,
    colour = '#FF6D9A',
    label = paste0('diff:\n 1 log-\nodd'),
    size = 5
  ) +

  geom_point() +
  NULL
```


Because the slope coefficient is 1, presented in isolation from the actual values that it's the difference between, we can't back-transform it without knowing what those other values are, because the difference in probability space could be 3% or 23% or anywhere in between.

:::


## Plotting model-fitted values

To display the predicted probability of senility symptoms across all WAIS values, we can use the `effect()` function from the **effects** package. 

        
```{r}
library(effects)

effect(term = "wais", mod = sen_mdl1, xlevels = 20) |>
  as.data.frame() |>
  ggplot(aes(x = wais, y = fit, ymin = lower, ymax = upper))+
  geom_line()+
  geom_ribbon(alpha = .3) +
  labs(
    y = 'Probability of senility symptoms',
    x = 'WAIS score'
  )
```

Here's what's going on in the code above:

Within `effect()`:   

- `term`: the predictor we want on the x axis.
- `mod`: the name of the fitted model  
- `xlevels`: return X number of evenly spaced fitted values across the predictor (in the below example, we are asking for 20)        
    
Within `ggplot()`:

-  `x`: predictor on x-axis        
-  `y`: predicted outcome on y-axis       
-  `ymin`: lower CI bound      
-  `ymax`: upper CI bound
-  `geom_line()`: predicted effect line (i.e., how the DV varies as a function of the continuous IV)
-  `geom_pointrange()`: predicted probability for each group (i.e., how the DV differs across categorical IV)
-  `geom_ribbon() `: shaded band representing the confidence interval                        
    

<!-- ## Get model-fitted values -->

<!-- We can also use `R` to extract predicted probabilities from our models. -->

<!-- -   Calculate the predicted log-odds (probabilities on the logit scale): `predict(model, type="link")` -->
<!-- -   Calculate the predicted probabilities: `predict(model, type="response")` -->

