---
title: "Plotting binary outcome data"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```

This flash card is a gallery of different ways you might plot binary outcome data, depending on what kinds of predictors you have.

For all plots, we'll use `Effect()` from the `effects` package.

```{r}
library(effects)
```

## One continuous predictor

::: {.callout-tip collapse="true"}
#### The `msleep` data

The `msleep` dataset from `ggplot2` records the amount of sleep of different mammals, along with other information about each species.
We'll only look at what predicts whether a mammal is a carnivore rather than a herbivore.

```{r}
msleep <- msleep |>
  filter(
    vore %in% c('herbi', 'carni')
  ) |>
  mutate(
    carnivore = ifelse(vore == 'carni', 1, 0)
  )

head(msleep)
```
:::

Visualise how likely an animal is to be a carnivore as a function of hours of REM sleep:

```{r}
msleep |> 
  ggplot(aes(x = sleep_rem, y = carnivore)) +
  geom_jitter(height = 0.01) +
  scale_y_continuous(breaks = c(0, 1)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = F, linewidth = 2) +
  labs(
    y = 'Type of animal \n(0 = Herbivore, 1 = Carnivore)',
    x = 'Amount of REM sleep (hours)'
  ) +
  NULL
```

^ a good illustration of how one data point can really influence a model!


## One categorical predictor

::: {.callout-tip collapse="true"}
#### The `Titanic` data

The `Titanic` dataset from the package `datasets` records the survival of passengers on the Titanic by age (child or adult), sex (male or female), and class (1st, 2nd, 3rd, and Crew).

```{r}
Titanic <- Titanic |> 
  as.data.frame() |>
  uncount(Freq)

Titanic |> head()
```

:::


Visualise proportion of each passenger class aboard the Titanic that survived:

```{r}
Titanic |>
  ggplot(aes(x = Survived, fill = Class)) +
  geom_bar(position = 'fill') +
  labs(
    y = 'Proportion'
  )
```





## One continuous and one categorical predictor

Tip: If you have a continuous predictor, it's a good idea to put that on the x axis.
Then you can use facets or colours to distinguish the levels of the categorical predictor.

Visualise carnivore vs. herbivore status as a function of amount of REM sleep, dividing for different conservation statuses (domesticated and lc = least concern).

Using facets:

```{r}
msleep |> 
  filter(conservation %in% c('lc', 'domesticated')) |>  # filter for ease of visualisation
  ggplot(aes(x = sleep_rem, y = carnivore)) +
  geom_jitter(height = 0.01) +
  scale_y_continuous(breaks = c(0, 1)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = F) +
  labs(
    y = 'Type of animal \n(0 = Herbivore, 1 = Carnivore)',
    x = 'Amount of REM sleep (hours)'
  ) +
  facet_wrap(~ conservation) 
```

<br>

Using colour:

```{r}
msleep |> 
  filter(conservation %in% c('lc', 'domesticated')) |>  # filter for ease of visualisation
  ggplot(aes(x = sleep_rem, y = carnivore, colour = conservation)) +
  geom_jitter(height = 0.01) +
  scale_y_continuous(breaks = c(0, 1)) +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = F) +
  labs(
    y = 'Type of animal \n(0 = Herbivore, 1 = Carnivore)',
    x = 'Amount of REM sleep (hours)',
    colour = 'Conservation status'
  )
```


## Two continuous predictors

It's not easy to show how a binary outcome changes as a function of two continuous predictors.

For multiple regression models, it suffices to plot each predictor against the outcome on its own, and then to use `patchwork` to join the plots to create a single figure.

For interaction models, where it is important to see how one variable's effect might depend on the other variable, then using `cut()` to divide one variable into discrete chunks and then treat it as a categorical variable is probably the nicest way (though still a bit awkward, since the data being visualised appears different from the data being modelled).

```{r}
msleep |>
  mutate(
    sleep_cycle_bin = cut(
      sleep_cycle,
      # define the cut points
      breaks = c( 
        min(sleep_cycle , na.rm = T),   # the bottom cut point = the min
        mean(sleep_cycle, na.rm = T), 
        max(sleep_cycle , na.rm = T)   # the top cut point = the max
      ),
      include.lowest = TRUE,
      # label each interval between the cut points
      labels = c(
        'Sleep cycle duration,\n from min to mean', 
        'Sleep cycle duration,\n from mean to max' 
      )
    )
  ) |>
  filter(!is.na(sleep_cycle_bin)) |>  # drop any NAs in this column
  ggplot(aes(x = sleep_rem, y = carnivore)) +
  facet_wrap(~sleep_cycle_bin, nrow = 1) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = binomial), se = F) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    y = 'Type of animal \n(0 = Herbivore, 1 = Carnivore)',
    x = 'Amount of REM sleep (hours)'
  )

```




## Two (or more) categorical predictors

Visualise proportion of each passenger class that survived, faceting by sex:

```{r}
Titanic |>
  ggplot(aes(x = Survived, fill = Class)) +
  geom_bar(position = 'fill') +
  labs(
    y = 'Proportion'
  ) +
  facet_wrap(~ Sex)
```


<br>


Visualise proportion of each passenger class that survived, faceting by sex and age:

```{r}
Titanic |>
  ggplot(aes(x = Survived, fill = Class)) +
  geom_bar(position = 'fill') +
  labs(
    y = 'Proportion'
  ) +
  facet_grid(Age ~ Sex)
```
