---
title: "Probabilities and log-odds"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(sjPlot)
```

We want to model the probability with which a "success" occurs (where "success" is the standard shorthand way of talking about the level of our [binary outcome](binary-outcomes.html){target="_blank"} which is represented as 1).

But probabilities are bounded between 0 and 1.
So instead of modelling the probability of success, logistic regression models will model the **log-odds** of success.
Log-odds are continuous and range between $-\infty$ and $\infty$, so they are the right kind of mathematical object that linear models can work with.

This plot shows the so-called "logistic" or "inverse logit" function, which maps log-odds to probabilities:

```{r}
#| code-fold: true

p_logistic <- tibble(
  logodds = seq(-5, 5, length.out = 100),
  prob = plogis(logodds)
) |>
  ggplot(aes(x = logodds, y = prob)) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  labs(
    y = 'Probability',
    x = 'Log-odds'
  )

p_logistic
```



## Converting from probabilities to log-odds

The so-called **logit** function (**log**istic un**it**) converts probabilities to log-odds.
In R, the logit function is `qlogis()`.

For example:

```{r}
qlogis(0.333)  # 33.3% probability
```

```{r}
#| code-fold: true

prob <- 0.333
p_logistic +
  # horiz
  geom_segment( x = -5, xend = qlogis(prob), y = prob, yend = prob, colour = dapr2red, arrow = arrow() ) +
  # vert
  geom_segment( x = qlogis(prob),  xend = qlogis(prob),  y = prob,  yend = 0,  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0(prob*100, '% probability = ', round(qlogis(prob), 2), ' log-odds'))
```

<br>

```{r}
qlogis(0.5)  # 50% probability, aka 50/50 chance
```

```{r}
#| code-fold: true

prob <- 0.5
p_logistic +
  # horiz
  geom_segment( x = -5, xend = qlogis(prob), y = prob, yend = prob, colour = dapr2red, arrow = arrow() ) +
  # vert
  geom_segment( x = qlogis(prob),  xend = qlogis(prob),  y = prob,  yend = 0,  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0(prob*100, '% probability = ', round(qlogis(prob), 2), ' log-odds'))
```


<br>


```{r}
qlogis(0.99)  # 99% probability
```

```{r}
#| code-fold: true

prob <- 0.99
p_logistic +
  # horiz
  geom_segment( x = -5, xend = qlogis(prob), y = prob, yend = prob, colour = dapr2red, arrow = arrow() ) +
  # vert
  geom_segment( x = qlogis(prob),  xend = qlogis(prob),  y = prob,  yend = 0,  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0(prob*100, '% probability = ', round(qlogis(prob), 2), ' log-odds'))
```


## Converting from log-odds to probabilities

The **logistic** function, aka the **inverse logit** function, converts log-odds to probabilities.
In R, the logistic function / inverse logit function is `plogis()`.

For example:

```{r}
plogis(3)
```


```{r}
#| code-fold: true

logodds <- 3

p_logistic +
  # vert
  geom_segment( x = logodds, xend = logodds, yend = plogis(logodds), y = 0, colour = dapr2red, arrow = arrow() ) +
  # horiz
  geom_segment( x = logodds,  xend = -5,  y =  plogis(logodds),  yend =  plogis(logodds),  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0( logodds, ' log-odds = ', 100 * round(plogis(logodds), 3), '% probability'))
```

<br>

```{r}
plogis(0)
```


```{r}
#| code-fold: true

logodds <- 0

p_logistic +
  # vert
  geom_segment( x = logodds, xend = logodds, yend = plogis(logodds), y = 0, colour = dapr2red, arrow = arrow() ) +
  # horiz
  geom_segment( x = logodds,  xend = -5,  y =  plogis(logodds),  yend =  plogis(logodds),  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0( logodds, ' log-odds = ', 100 * round(plogis(logodds), 3), '% probability'))
```


<br>

```{r}
plogis(-1.5)
```


```{r}
#| code-fold: true

logodds <- -1.5

p_logistic +
  # vert
  geom_segment( x = logodds, xend = logodds, yend = plogis(logodds), y = 0, colour = dapr2red, arrow = arrow() ) +
  # horiz
  geom_segment( x = logodds,  xend = -5,  y =  plogis(logodds),  yend =  plogis(logodds),  colour = dapr2red, arrow = arrow() ) +
  ggtitle(paste0( logodds, ' log-odds = ', 100 * round(plogis(logodds), 3), '% probability'))
```