---
title: "16: Probabilities and log-odds"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: inline
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')

library(tidyverse)
library(patchwork)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```


In this lab, you'll practice transforming probabilities into log-odds as well as log-odds into probabilities.

You'll also work through an example that aims to show you why transforming the slope of a line from log-odds into probabilities leads to mathematically impossible predictions.

<!-- The goal of this lab is to reinforce that you shouldn't directly back-transform the slopes of a line from log-odds into probabilities. -->


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab16.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
:::

---


<!-- # q1 -->

`r qbegin(qcounter())`

Copy and run the following code to generate for yourself a plot of the logistic function.

You'll use this plot in the next few questions.

```{r eval=F}
p_logistic <- tibble(
  logodds = seq(-5, 5, length.out = 100),
  prob = plogis(logodds)
) |>
  ggplot(aes(x = logodds, y = prob)) +
  geom_line() +
  labs(
    y = 'Probability',
    x = 'Log-odds'
  ) +
  ggtitle('The logistic function')

p_logistic
```

(Optional: To find out what a given line of code does, you can comment it out and try running the code chunk again.)

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The plot should look something like this:

```{r echo=F}
p_logistic <- tibble(
  logodds = seq(-5, 5, length.out = 100),
  prob = plogis(logodds)
) |>
  ggplot(aes(x = logodds, y = prob)) +
  geom_line() +
  labs(
    y = 'Probability',
    x = 'Log-odds'
  ) +
  ggtitle('The logistic function')

p_logistic
```


`r solend()`


# Converting probabilities to log-odds

<!-- # q2 -->

`r qbegin(qcounter())`

Based on the logistic curve you plotted in Q1, guess the approximate log-odds value that corresponds to each of the following probabilities.
Write down your guesses.

Probabilities:

- 33%
- 99%
- 20%
- 0%
- 100%



`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Eyeballing a path from the probabilities on the y axis via the logistic function to the log-odds values on the x axis, I'd say:

- 33% probability $\approx$ –0.5 log-odds
- 99% probability $\approx$ 4.5 log-odds
- 20% probability $\approx$ –1.5 log-odds
- 0% probability is never met, only approached, as the logistic function goes down to $-\infty$
- 100% probability is never met, only approached, as the logistic function goes up to $\infty$

Your guesses might be different from mine, and that's OK!

`r solend()`

<!-- # q3 -->

`r qbegin(qcounter())`

Now compute the actual log-odds value for each probability given in Q2.

If any of your guesses were really different from the actual log-odds, why do you think that is?

:::red
🗂️
See [TODO probs and log-odds](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
qlogis(0.33)
```

<br>

```{r}
qlogis(0.99)
```

<br>

```{r}
qlogis(0.2)
```

<br>

```{r}
qlogis(0)
```

`-Inf`(inity) is R's way of telling us that the logistic function never really reaches 0% probability, but as it gets closer, it gets bigger and more negative.

<br>

```{r}
qlogis(1)
```

Likewise for 100% probability: this value is never reached by the logistic function, but it gets ever closer as log-odds get larger and more positive.

<!-- (The mathematical term for a value always approached but never attained is an "asymptote". -->
<!-- Probabilities of 0% and 100% are asymptotes for the logistic function.) -->

`r solend()`

<!-- # q4 -->

`r qbegin(qcounter())`

Try running the following three lines of code.
They will fail to return sensible log-odds values.

Explain why each line fails.

```{r eval=F}
qlogis(-0.5)
qlogis(1.5)
qlogis(100)
```


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

None of the given values are possible probabilities, so there is no way to map them to log-odds.

- –0.5: Probabilities cannot be negative.
- 1.5 and 100: Probabilities cannot be greater than 1.

`r solend()`


# Converting log-odds to probabilities

<!-- # q5 -->

`r qbegin(qcounter())`

Based on the logistic curve you plotted in Q1, guess the approximate probability value that corresponds to each of the following log-odds.
Write down your guesses.

Log-odds:

- 0.5
- –0.5
- 10
- –10
- 4


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Eyeballing a path from the log-odds on the x axis via the logistic function to the probabilities on the y-axis, I'd say:

- 0.5 log-odds $\approx$ 60% probability
- –0.5 log-odds $\approx$ 40% probability
- 10 log-odds $\approx$ 99.9% probability, something really big
- –10 log-odds $\approx$ 0.1% probability, something really small
- 4 log-odds $\approx$ 98% probability

Your guesses might be different from mine, and that's OK!

`r solend()`

<!-- # q6 -->

`r qbegin(qcounter())`

Now compute the actual probabilities for each log-odds value given in Q2.

If any of your guesses were really different from the actual probability, why do you think that is?

:::red
🗂️
See [TODO probs and log-odds](https://uoepsy.github.io/dapr2/2627/flashcards/block1/TODO.html){target="_blank"} flash cards.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
plogis(0.5)
```

<br>

```{r}
plogis(-0.5)
```

<br>

```{r}
plogis(10)
```

<br>

```{r}
plogis(-10)
```

<br>

```{r}
plogis(4)
```

`r solend()`

<!-- # q7 -->

`r qbegin(qcounter())`

Can you find any input values that cause `plogis()` to fail?

If you do find some values, which are they, and why do they make `plogis()` fail?

If you cannot find any such values, why are there no values that make `plogis()` fail?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

There are no values that make `plogis()` fail.
It will take any number (even `-Inf` and `Inf`) and map it via the logistic function to a probability between 0 and 1.

This is because the input to `plogis()` are values on the log-odds scale.
And log-odds can be any number between $-\infty$ and $\infty$.
(This property is exactly what makes log-odds a suitable scale for linear modelling!)

`r solend()`

# Why we can't back-transform slopes from log-odds space to probability space

<!-- # q8 -->

`r qbegin(qcounter())`
Imagine a line in log-odds space which has an intercept of 1 and a slope of 0.5.
Here's a plot of that line:

```{r}
#| code-fold: true

tibble(
  x = 0:5,
  logodds_success = 1 + 0.5 * x
) |>
  ggplot(aes(x = x, y = logodds_success)) +
  geom_line()
```

Now, **imagine doing the wrong thing and back-transforming *both* model coefficients from log-odds space to probability space.**
This is a "proof by contradiction": we'll intentionally do the wrong thing and see how it leads us to nonsense results.

- For the intercept, `plogis(1)` = 0.73.
- For the slope, `plogis(0.5)` = 0.62.

The linear expression for this model is $y = 0.73 + 0.62 \cdot x$.

What outcome value (i.e., $y$ value) does this model predict for $x = 0$? for $x = 1$? for $x=2$?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

For $x = 0$:

$$
\begin{align}
y &= 0.73 + 0.62 \cdot 0 \\
y &= 0.73 \\
\end{align}
$$

<br>


For $x = 1$:

$$
\begin{align}
y &= 0.73 + 0.62 \cdot 1 \\
y &= 1.35 \\
\end{align}
$$
<br>


For $x = 2$:

$$
\begin{align}
y &= 0.73 + 0.62 \cdot 2 \\
y &= 0.73 + 1.24 \\
y &= 1.97 \\
\end{align}
$$

`r solend()`

<!-- # q9 -->

`r qbegin(qcounter())`

Our goal is for the outcome values of this model to be probabilities: the probability of success.

Are the outcome values you computed in Q8 possible probabilities?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Probabilities can range between 0 and 1. Therefore:

- 0.73 is a possible probability (this is the y-intercept in probability space).
- 1.35 is not a possible probability, because it's greater than 1.
- 1.97 is also not a possible probability, because it's greater than 1.

Because our model generates nonsense predictions (i.e., impossible probability values), the assumptions behind the model must be incorrect.
Specifically, our assumption that a linear relationship in log-odds translates exactly to a linear relationship in probability space is not right.


`r solend()`


<!-- # q10 -->


`r qbegin(qcounter())`

Below are two plots.

Plot A (left) shows the intentionally wrong way we tried to convert the log-odds line into probability space: by transforming each parameter individually to end up with $y = 0.73 + 0.62 \cdot x$.

Plot B (right) shows the correct way of converting the log-odds line into probability space.

```{r fig.width = 10, fig.height = 5}
#| code-fold: true

p_prob_bad <- tibble(
  x = 0:5,
  prob_success = 0.73 + 0.62 * x
) |>
  ggplot(aes(x = x, y = prob_success)) +
  geom_line() +
  ggtitle('Bad transformation to prob. space')


p_prob_good <- tibble(
  x = 0:5,
  logodds_success = 1 + 0.5 * x,
  prob_success = plogis(logodds_success)
) |>
  ggplot(aes(x = x, y = prob_success)) +
  geom_line() +
  ggtitle('Good transformation to prob. space')

p_prob_bad + p_prob_good + plot_annotation(tag_levels = 'A')
```


Notice that we get the same value in both plots when $x = 0$: the probability is indeed 0.73 = `plogis(1)`.
So transforming the intercept alone is okay!

Based on these two plots, try to explain in your own words why transforming the slope directly is not a mathematically appropriate way to convert a line in log-odds space into probability space.


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

You might observe that the bad transformation gives us impossible probabilities (> 1), while the good transformation only gives us reasonable probabilities (between 0 and 1).

That's absolutely true, and that observation is the goal of this lab.

If you want to go a bit deeper, you might also look at Plot B and observe that the good line isn't a straight line at all: in fact, the slope is constantly changing as $x$ increases.
This is because the logistic function has to squish the straight line down so that it fits in between 0 and 1.

Because the line in probability space is not straight, there's no single slope value that applies to the whole line.
Consequently, it doesn't make sense to transform the slope from log-odds space to probability space and expect the resulting straight line to accurately describe the probability of success.

`r solend()`

<!-- `r qbegin(qcounter())` -->

<!-- **Optional bonus challenge:** If we can't back-transform slopes to probability space, how *can* we describe the effect that a predictor has on the probability of success? -->

<!-- Here's the trick:  -->

<!-- 1. Use the model to predict outcome values in log-odds space for whatever values of the predictor we are interested in. -->
<!--     - Typically for a continuous predictor, these values are the predictor's mean and one SD above the predictor's mean. -->
<!-- 1. Transform each of those log-odds values into probability space. -->
<!-- 1. Subtract the estimated probability of success at the predictor's mean from the estimated probability of success at mean + 1 SD. -->


<!-- - Find the mean of `mallow$age` and store it in a variable called `age_mean`. -->
<!-- - Find the SD of `mallow$age` and store it in a variable called `age_sd`. -->
<!-- - Add the mean and SD together and store the result in a variable called `age_mean_plus_1`. -->
<!-- - -->


<!-- `r qend()` -->
<!-- `r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)` -->

<!-- ```{r} -->
<!-- age_mean <- mean(mallow$age) -->
<!-- age_sd   <- sd(mallow$age) -->
<!-- age_mean_plus_1 <- age_mean + age_sd -->


<!-- ``` -->


<!-- `r solend()` -->
