---
title: "17: Logistic regression with one predictor"
params: 
    SHOW_SOLS: TRUE
    # TOGGLE: TRUE
    TOGGLE: FALSE
editor_options: 
  chunk_output_type: inline
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')

library(tidyverse)
library(effects)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```


In this lab, you'll practice fitting and interpreting a logistic regression model with that contains one four-level categorical predictor.


:::{.callout-note}
### Get set up

1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab17.Rmd`).
1. In the first code chunk, load the packages you'll need this week:
    - `tidyverse`
    - `effects`
:::

---

:::frame

RQ: **How are current intentions to quit smoking associated with the probability that an individual has previously attempted to quit smoking?**

Data dictionary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
quit_data <- read_csv("https://uoepsy.github.io/data/QuitAttempts.csv")
tibble(
variable = names(quit_data),
description = c("Subject Number","Whether the subject made an attempt to quit using tobacco products (0 = no attempt to quit; 1 = attempted to quit).", "Subject's intention to quit (1 = Never intend to quit; 2 = May intend to quit eventually; 3 = Intend to quit in the next 6 months; 4 = Intend to quit in the next 30 days)")
) %>% gt::gt()
```

:::

# Read in and set up data

# q1

`r qbegin(qcounter())`

Read in the data from <https://uoepsy.github.io/data/scs_study.csv> and store it in a variable named `quit_data`.

Then copy and run the following code to make the dataset clearer and more human-readable.

```{r eval=F}
quit_data <- quit_data|>
    rename(
        pptID = S,
        QuitAttempt = Q,
        Intent = I
    ) |>
    mutate(
        Intent = case_when(
            Intent == 1 ~ 'Never',
            Intent == 2 ~ 'Maybe',
            Intent == 3 ~ 'Yes6mo',
            Intent == 4 ~ 'Yes30days'
        )
    )
```


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
quit_data <- read_csv('https://uoepsy.github.io/data/QuitAttempts.csv')

quit_data <- quit_data|>
    rename(
        pptID = S,
        QuitAttempt = Q,
        Intent = I
    ) |>
    mutate(
        Intent = case_when(
            Intent == 1 ~ 'Never',
            Intent == 2 ~ 'Maybe',
            Intent == 3 ~ 'Yes6mo',
            Intent == 4 ~ 'Yes30days'
        )
    )

head(quit_data)
```


`r solend()`

# q2

`r qbegin(qcounter())`

We'll model people's previous quitting behaviour as a function of their current intent to quit, so `Intent` is going to be our predictor in this model.
And the `Never` level would be a good reference level to compare all the other levels back to.

Convert `Intent` into a factor and set `Never` as its first level.
Then use `contrasts()` to generate a contrast matrix.
If you've done it right, you should see something like the following output:

```{r echo=F}
quit_data <- quit_data |>
    mutate(
        Intent = factor(Intent, levels = c('Never', 'Maybe', 'Yes6mo', 'Yes30days'))
    )

contrasts(quit_data$Intent)
```

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
quit_data <- quit_data |>
    mutate(
        Intent = factor(Intent, levels = c('Never', 'Maybe', 'Yes6mo', 'Yes30days'))
    )

contrasts(quit_data$Intent)
```

`r solend()`


# q3

`r qbegin(qcounter())`

Make a plot that visualises, for each value of `QuitAttempt`, the proportion of different `Intent`s to quit.

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
quit_data |>
    ggplot(aes(
        x = factor(QuitAttempt), 
        fill = Intent)
    ) +
    geom_bar(position = 'fill') +
    
    # update labels (the automatic label "count" on the y axis is wrong!)
    labs(
        y = 'Proportion of individuals',
        x = 'Previously attempted to quit smoking (0 = No, 1 = Yes)',
        fill = 'Intent to quit'
    )
```

<br>

Bonus, if you want to get fancy and relabel the levels of `QuitAttempt` and `Intent` in the plot:

```{r}
quit_data |>
    ggplot(aes(
        x = factor(QuitAttempt), 
        fill = Intent)
    ) +
    geom_bar(position = 'fill') +
    
    # Change how the fill variable's levels are labelled
    # (careful to match the original order!)
    scale_fill_discrete(labels = c('Never', 'Maybe', 'Yes (within 6 months)', 'Yes (within 30 days)')) +
    
    # Change how the x axis variable's levels are labelled
    # (careful to match the original order!)
    scale_x_discrete(labels = c('No', 'Yes')) +
    
    # update labels (the automatic label "count" on the y axis is wrong!)
    labs(
        y = 'Proportion of individuals',
        x = 'Previously attempted to quit smoking?',
        fill = 'Intent to quit'
    )
```



`r solend()`


# Set up and fit model

# q4

`r qbegin(qcounter())`

The logistic regression model we will fit has the formula `QuitAttempt ~ Intent`.

Based on the contrast matrix you generated above, write the mathematical expression that represents this model.

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

$$
\text{logodds(QuitAttempt)} = \beta_0 + 
(\beta_1 \cdot \text{Intent}_\text{Maybe}) + 
(\beta_2 \cdot \text{Intent}_\text{Yes6mo}) + 
(\beta_3 \cdot \text{Intent}_\text{Yes30days})
$$

Write:

```
$$
\text{logodds(QuitAttempt)} = \beta_0 + 
(\beta_1 \cdot \text{Intent}_\text{Maybe}) + 
(\beta_2 \cdot \text{Intent}_\text{Yes6mo}) + 
(\beta_3 \cdot \text{Intent}_\text{Yes30days})
$$
```

`r solend()`

# q5

`r qbegin(qcounter())`

For each $\beta$ coefficient in the model expression above, write a sentence to describe what it will mean.

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

- $\beta_0$ (the intercept): the estimated log-odds of having previously attempted to quit smoking for people who currently intend to never quit smoking.
- $\beta_1$: the difference in the log-odds of having a previous quit attempt between people who intend to maybe quit smoking and people who intend to never quit smoking.
- $\beta_2$: the difference in the log-odds of having a previous quit attempt between people who intend to quit smoking within the next six months and people who intend to never quit smoking.
- $\beta_3$: the difference in the log-odds of having a previous quit attempt between people who intend to quit smoking within the next 30 days and people who intend to never quit smoking.

`r solend()`

# q6

`r qbegin(qcounter())`

Use `glm()` to fit this logistic regression model.
Save the output as `quit_mdl` and generate the model summary.


:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
quit_mdl <- glm(
    QuitAttempt ~ Intent,
    data = quit_data,
    family = binomial
)

summary(quit_mdl)
```


`r solend()`

# q7

`r qbegin(qcounter())`

For each coefficient estimated in `quit_mdl`, write a sentence that interprets it and another sentence that reports its statistical significance.
(Remember that for logistic regression, we want to report the significance of the intercept as well because now it tests whether a particular outcome is significantly more or less likely than random chance!)

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`(Intercept)`:

- The log-odds of having a previous quit attempt for somebody who currently never intends to quit smoking is –1.95; this corresponds to a probability of about 12.5%.
- This estimate (in log-odds space) is not significantly different from zero, that is, from 50% chance ($p$ = .069).


`IntentMaybe`

- The log-odds of having a previous quit attempt for somebody who currently intends to maybe quit smoking is estimated to increase by 2.58 log-odds compared to someone who never intends to quit.
- This estimate is significantly different from zero ($p$ = .024).


`IntentYes6mo`

- The log-odds of having a previous quit attempt for somebody who currently intends to quit smoking within the next six months is estimated to increase by 3.23 log-odds compared to someone who never intends to quit.
- This estimate is significantly different from zero ($p$ = .006).



`IntentYes30days`

- The log-odds of having a previous quit attempt for somebody who currently intends to quit smoking within the next 30 months is estimated to increase by 3.33 log-odds compared to someone who never intends to quit.
- This estimate is significantly different from zero ($p$ = .031).


`r solend()`

# Predict probabilities of quit attempts

# q8

`r qbegin(qcounter())`

What is the predicted log-odds of each group (that is, each level of `Intent`) to have previously attempted to quit smoking?

Use the mathematical model expression from Q4 and the contrast matrix from Q2 to calculate the model-predicted log-odds by hand (that is, not using `predict()`).

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The mathematical model expression, with the model-estimated coefficients substituted in (and "QuitAttempt" shortened to "QA" for convenience), is:

$$
\text{logodds(QA)} = –1.95 + 
(2.58 \cdot \text{Intent}_\text{Maybe}) + 
(3.23 \cdot \text{Intent}_\text{Yes6mo}) + 
(3.33 \cdot \text{Intent}_\text{Yes30days})
$$

The contrast matrix tells us what values to substitute in for each predictor in order to calculate the estimated log-odds of having a previous quit attempt for each level of `Intent`.

```{r}
contrasts(quit_data$Intent)
```

<br>

**For `Intent = Never`, the values we substitute in for each dummy variable are:**

```{r}
contrasts(quit_data$Intent)['Never',]
```


So:

$$
\begin{align}
\text{logodds(QA)}_\text{Never} &= –1.95 + 
(2.58 \cdot \text{Intent}_\text{Maybe}) + 
(3.23 \cdot \text{Intent}_\text{Yes6mo}) + 
(3.33 \cdot \text{Intent}_\text{Yes30days}) \\
 &= –1.95 + 
(2.58 \cdot 0 ) + 
(3.23 \cdot 0 ) + 
(3.33 \cdot 0 ) \\
  &= –1.95 \\
\end{align}
$$

<br>

**For `Intent = Maybe`:**

```{r}
contrasts(quit_data$Intent)['Maybe',]
```

So:

$$
\begin{align}
\text{logodds(QA)}_\text{Maybe} &= –1.95 + 
(2.58 \cdot \text{Intent}_\text{Maybe}) + 
(3.23 \cdot \text{Intent}_\text{Yes6mo}) + 
(3.33 \cdot \text{Intent}_\text{Yes30days}) \\
 &= –1.95 + 
(2.58 \cdot 1 ) + 
(3.23 \cdot 0 ) + 
(3.33 \cdot 0 ) \\
  &= –1.95 + 2.58 \\
  &= 0.63 \\
\end{align}
$$


<br>

**For `Intent = Yes6mo`:**


```{r}
contrasts(quit_data$Intent)['Yes6mo',]
```

So:

$$
\begin{align}
\text{logodds(QA)}_\text{Yes6mo} &= –1.95 + 
(2.58 \cdot \text{Intent}_\text{Maybe}) + 
(3.23 \cdot \text{Intent}_\text{Yes6mo}) + 
(3.33 \cdot \text{Intent}_\text{Yes30days}) \\
 &= –1.95 + 
(2.58 \cdot 0 ) + 
(3.23 \cdot 1 ) + 
(3.33 \cdot 0 ) \\
  &= –1.95 + 3.23 \\
  &= 1.28 \\
\end{align}
$$

<br>

**For `Intent = Yes30days`:**

```{r}
contrasts(quit_data$Intent)['Yes30days',]
```

So:

$$
\begin{align}
\text{logodds(QA)}_\text{Yes30days} &= –1.95 + 
(2.58 \cdot \text{Intent}_\text{Maybe}) + 
(3.23 \cdot \text{Intent}_\text{Yes6mo}) + 
(3.33 \cdot \text{Intent}_\text{Yes30days}) \\
 &= –1.95 + 
(2.58 \cdot 0 ) + 
(3.23 \cdot 0 ) + 
(3.33 \cdot 1 ) \\
  &= –1.95 + 3.33 \\
  &= 1.38 \\
\end{align}
$$

`r solend()`



# q9

`r qbegin(qcounter())`

Convert each of the log-odds values you calculated in Q8 to probabilities.

How much more probable is it to have a previous quit attempt for someone who intends to quit within the next 30 days, compared to someone who intends never to quit?

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

**For `Intent = Never`:**

```{r}
plogis(-1.95) |> round(3) * 100
```

The probability of a previous quit attempt for people who intend never to quit is 12.5%.
(notice that this works out to the same value as the model's intercept, which matches our interpretation above!)

<br>

**For `Intent = Maybe`:**

```{r}
plogis(0.63) |> round(3) * 100
```

The probability of a previous quit attempt for people who intend maybe to quit is 65.2%.

<br>

**For `Intent = Yes6mo`:**

```{r}
plogis(1.28) |> round(3) * 100
```

The probability of a previous quit attempt for people who intend to quit within the next six months is 78.2%.

<br>

**For `Intent = Yes30days`:**

```{r}
plogis(1.38) |> round(3) * 100
```

The probability of a previous quit attempt for people who intend to quit within the next 30 days is 79.9%.


If someone intends to quit within 30 days, they're 79.9% – 12.5% = 67.4% more likely to have a previous quitting attempt, compared to someone who never intends to quit.
(Fun fact: *This* is how you can calculate the difference between groups in probability space! Not by converting the slope from log-odds to probabilities, but rather by calculating the model-predicted probabilities for the two groups you want to compare, and then subtracting them :) )

`r solend()`


# q10

`r qbegin(qcounter())`

Use `Effect()` to plot the model-predicted probabilities of a previous quit attempt for each level of `Intent`.
Do the values you calculated in Q9 match the estimates shown in the resulting plot?
If they don't, can you identify where your calculations went wrong?

:::red
🗂️
See [TODO](TODO){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
Effect(focal.predictors = 'Intent', mod = quit_mdl) |>
    as.data.frame() |>
    ggplot(aes(x = Intent, y = fit)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) +
    labs(
        y = 'Probability of having a previous quit attempt',
        x = 'Intent to quit smoking'
    ) +
    ylim(0, 1) +
    # fancy bonus line to make x axis labels nice
    scale_x_discrete(labels = c('Never', 'Maybe', 'Yes (within\n6 months)', 'Yes (within\n30 days)'))
```

`r solend()`