---
title: "19: Practice write-up"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
    # TOGGLE: FALSE
editor_options: 
  chunk_output_type: inline
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->


```{r}
#| label: setup
#| include: false

source('../assets/setup.R')

library(tidyverse)
library(psych)
library(sjPlot)
library(kableExtra)
library(patchwork)
library(car)
library(interactions)
library(effects)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll practice writing up the analysis that was developed live during lectures.

Your goal is to present your analysis in a journal-article-style Results section that a reader could reproduce in any statistical software, not just R.

Quick links to key flash cards:

- 🗂️   [Analysis workflow (Block 4)](https://uoepsy.github.io/dapr2/2627/flashcards/resources/workflow-b4.html){target="_blank"}
- 🗂️   [Reporting LM analysis checklist](https://uoepsy.github.io/dapr2/2627/flashcards/resources/reporting-lm.html){target="_blank"}


:::{.callout-important}

The example write-ups that will be provided **are not perfect.**

You must not copy them directly for this write-up or any future assessment, as copying and presenting them as your own work would constitute academic misconduct.
But you may refer to them for stylistic guidance.

:::

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab19.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
    - `psych`
    - `sjPlot`
    - `kableExtra`
    - `car`
    - `effects`
   
:::

---

:::frame

The data in `mallow2_cleaned.csv` contain five attributes collected from a simulated sample of $n=301$ hypothetical individuals, and includes:

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
delay_data <- read_csv("https://uoepsy.github.io/data/mallow2_cleaned.csv")
tibble(
  Variable = names(delay_data),
  Description = c("Participant name", "Age (in months)", "Time of day that the experiment took place ('am' = morning, 'pm' = afternoon)", "Experimental condition – whether the marshmallow was 'visible' or 'hidden' for the 10 minutes", "Whether or not the participant took ('taken') the marshmallow within the 10 minutes or left it ('waited')", "Age (in years)")
) %>% gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

The data used for this write-up exercise are simulated, drawing on the Stanford marshmallow experiment - a study on delayed gratification in children. The simulated data are loosely based on the findings of this work, and acted to expand upon the methods and results reported in the paper:

Mischel, W., Ebbesen, E. B., & Raskoff Zeiss, A. (1972). Cognitive and attentional mechanisms in delay of gratification. *Journal of Personality and Social Psychology, 21*(2), 204–218. [https://doi.org/10.1037/h0032198](https://doi.org/10.1037/h0032198)

*NOTE: You are not expected to write an introduction, so you do not have to read this article.*

In the current study, a sample of 304 children, ranging in ages from 3 to 10 years old, took part. Each child was shown a marshmallow, and it was explained that they were about to be left alone for 10 minutes. They were told that they were welcome to eat the marshmallow while they were waiting, but if the marshmallow was still there after 10 minutes, they would be rewarded with **two** marshmallows.

For half of the children who took part, the marshmallow was visible for the entire 10 minutes (or until they ate it!). For the other half, the marshmallow was placed under a plastic cup.

The experiment took part at various times throughout the working day, and researchers were worried about children being more hungry at certain times of day, so they kept track of whether each child completed the task in the morning or the afternoon, so that they could control for this in their analyses.

:::

:::

:::frame

RQ: Does the probability of delaying gratification change as a function of age, marshmallow visibility, and time of day?

:::

:::{.callout-tip collapse='true'}
### Provided analysis code

This code should resemble the analysis developed during lectures this week.

```{r eval=F}
# Logistic regression: Practice analysis
# Follows workflow here: dapr2/2627/dapr2_flashcards/resources/workflow-b4.html

# Phase 1: Before model fitting ---- 

## 1a: Set up code and data ---- 

# Load the required R packages
library(tidyverse)
library(psych)
library(sjPlot)
library(kableExtra)
library(car)
library(effects)

# Read in data
delay_data <- read_csv('https://uoepsy.github.io/data/mallow2_cleaned.csv')

# Tidy data (e.g., any missingness, any implausible values? are data types set correctly?).
glimpse(delay_data)

# - describe continuous variables
describe(delay_data$age)  # looks plausible – min 3.25, max 10, mean 6.56 years

# - get frequency distributions for categorical variables
table(delay_data$timeofday)   # fine, no weird values
table(delay_data$visibility)  # fine, no weird values
table(delay_data$taken)       # fine, no weird values


## 1b: Set up the variables you'll model ----

# Based on the RQ, decide whether you'll test your hypotheses using coefficient significance tests or model comparison.
# - RQ: Does the probability of delaying gratification change as a function of age, marshmallow visibility, and time of day?
# - RQ asks about how a probability changes as a function of some predictors – so that's coefficients

# Based on the RQ, decide whether your model requires an interaction between predictors.
# - RQ doesn't ask about whether one predictor's effect depends on another predictor
# - So we can answer the RQ with a multiple regression model only

# Based on the RQ, identify your outcome variable and the predictors.
# - Outcome: `taken`
# - Predictors: `age`, `visibility`, `timeofday`

# Based on the outcome variable, decide whether to use regular regression or logistic regression.
# - The outcome variable `taken` is binary (taken vs. waited), so we'll use logistic regression

# If logistic regression, set up the outcome variable so that desired level will be modelled as "success".
# - The RQ wants us to estimate the probability of a "waited" response.
# - Therefore "waited" needs to be the "success" level, coded as 1
delay_data <- delay_data |>
  mutate(
    taken = ifelse(taken == 'waited', 1, 0)
  )

# Set up categorical predictors (e.g., factor levels? which contrast coding scheme?)
# - No specific requirements for contrast coding scheme from RQ
# - So let's just do R's default treatment coding, and use the alphabetically first level for each variable
# - Can just convert to factors and check contrast matrices
delay_data <- delay_data |>
  mutate(
    visibility = factor(visibility),
    timeofday = factor(timeofday),
  )
contrasts(delay_data$visibility)  # ref level = hidden
contrasts(delay_data$timeofday)   # ref level = am


# Set up continuous predictors
# - It's reasonable to mean-centre age, because we have no observations of 0-year-olds in the data
# - It would also be reasonable not to, because 0 is a possible age, but let's do it for the practice
delay_data <- delay_data |>
  mutate(
    age_c = age - mean(age)
  )

# Explore patterns in the data by plotting outcome and predictor variables together
# - We'll have the continuous variable on the x axis, the responses on the y axis at 0 and 1,
#   and facets for each combination of visibility and timeofday
delay_data |>
  ggplot(aes(x = age_c, y = taken)) +
  
  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # facet by categorical predictors
  facet_grid(visibility ~ timeofday) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # add logistic curve
  geom_smooth(
    method = "glm",                          # not a straight line
    method.args = list(family = binomial),   # specify logistic curve
    se = FALSE          # remove standard error ribbon around line
  ) +
  
  # add labels for clarity (currently taken = 1 looks like they took it, not waited!)
  labs(
    x = 'Age (in years, mean-centered)',
    y = 'Delayed gratification? (0 = took marshmallow, 1 = waited)'
  )


# Phase 2: Model fitting ----

# Write the mathematical model formulation for your model(s)
# - Refer back to contrast matrices to know what the reference levels will be.
#   $$
#   \text{logodds(waited)} = \beta_0 + 
#   (\beta_1 \cdot \text{visibility}_\text{visible}) + 
#   (\beta_2 \cdot \text{timeofday}_\text{pm}) + 
#   (\beta_3 \cdot \text{age})
#   $$
  
  
# Explicitly define the hypotheses that your RQ is aiming to test.
# - RQ asks whether any predictors affect the outcome, so the H0 is that none of them do
#   $$
#   \begin{align}
#   H_0 &: \text{All } \beta_j = 0 \text{ for } j = 2, 3, 4 \\
#   H_1 &: \text{Any } \beta_j \neq 0\\
#   \end{align}
#   $$

# Fit your model(s) using `lm()` or `glm()` as appropriate.
delay_mdl <- glm(
  taken ~ age_c + visibility + timeofday, 
  data = delay_data,
  family = binomial
)
delay_mdl

# Phase 3: After model fitting ----

## 3a: Check assumptions and diagnostics ----

# Check model assumptions
rstandard(delay_mdl, type = "deviance") |> plot() # nothing outside -/+3, good
  
# Run diagnostics for multicollinearity.
vif(delay_mdl)  # all close to 1, good

# Run diagnostics for influential observations.
plot(delay_mdl, which = 4)  # all well below 0.5 threshold

# If you find extreme influential observations: run sensitivity analysis.
# - NA
   

## 3b: Plot and interpret model estimates ----
   
# Interpret the coefficient estimates
summary(delay_mdl)

# - Intercept:
#   - For a child of average age in the morning when the marshmallow is hidden, 
#     the log-odds of delaying gratification are 0.63.
plogis( coef(delay_mdl)[[1]] )  # or 65% probability
#   - This estimate is significantly different from zero (p = .011), 
#     so significantly more likely than chance.

# age_c:
#   - Holding visibility and time of day constant, increasing one year in age
#     is associated with an increase of 0.58 in the log-odds of delaying gratification.
#   - This estimate is significantly different from zero (p < .001), 


# - visibilityvisible:
#   - Holding age and time of day constant, changing the marshmallow from hidden to visible 
#     is associated with a decrease of 1.20 in the log-odds of delaying gratification.
#   - This estimate is significantly different from zero (p < .001), 

# timeofdaypm:
#   - Holding age and visibility constant, changing the time of day from am to pm
#     is associated with an increase of 0.49 in the log-odds of delaying gratification.
#   - This estimate is not significantly different from zero (p = .096).


# Generate a nicely-formatted regression table.
tab_model(
  delay_mdl,
  transform = NULL,  # display coefficients as log-odds
  dv.labels = 'Log-odds of delaying gratification',
  pred.labels = c(
    'Intercept', 'Visibility [visible]', 'Time of day [pm]', 'Age (mean-centered)'
  ),
  title = "Logistic regression model coefficient estimates"
)

# Compare models and interpret model comparison statistics (if appropriate for your RQ) 
# - NA.

# Plot model-fitted values and/or simple slopes/simple effects.
# - We'll plot the data in probability space, since that's what more readers will understand.
# - To create swoopy-line plots in the 2x2 facet grid:
Effect(
  focal.predictors = c('age_c', 'visibility', 'timeofday'), 
  mod = delay_mdl,
  xlevels = 20  # how many values of `age_c` to estimate outcomes for
) |>
  as.data.frame() |>
  ggplot(aes(x = age_c, y = fit)) +
  # a line for the model-fitted estimates
  geom_line() +
  # a ribbon for the 95% CIs
  geom_ribbon( aes( ymin = lower,  ymax = upper ),  alpha = .3 ) +
  # facet by visibility and timeofday
  facet_grid(visibility ~ timeofday) +
  # axis labels for clarity
  labs(
    y = 'Probability of delaying gratification',
    x = 'Age (years, mean-centered)'
  ) 


## 3c: Write up -- in labs! ----

```

```{r include=F}
delay_data <- read_csv('https://uoepsy.github.io/data/mallow2_cleaned.csv')

delay_data <- delay_data |>
    mutate(
        taken = ifelse(taken == 'waited', 1, 0),
        visibility = factor(visibility),
        timeofday = factor(timeofday),
        age_c = age - mean(age)
    )

delay_mdl <- glm(
  taken ~ visibility + timeofday + age_c,
  data = delay_data,
  family = binomial
)
```


:::

<!-- # q1 -->

`r qbegin(qcounter())`

**Set the scene (so the reader knows what you're talking about).**

- [ ] Remind the reader briefly what the study aimed to test (e.g., what the predictions were).

`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

This study investigates how children delay gratification using the classic marshmallow paradigm.
We are interested in how children's age, the visibility of the marshmallow, and the time of day might play into children's delay gratification behaviour.

`r solend()`

<!-- # q2 -->

`r qbegin(qcounter())`

**Describe and show the sample data (so that the reader knows how generalisable your analysis might be).**

- [ ] Introduce the variables the study is using to test this question, but avoid using R variable names; describe the variables in human-readable terms instead (e.g., what scale are they on? how were they measured?)
- [ ] Briefly describe the data set: how many observations? what does each observation represent? 
- [ ] If applicable, mention any data cleaning / outlier removal / transformations.
  
`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

We have data from 301 children, including their ages in years, whether the marshmallow was visible to the child or hidden from them ("visible" vs. "hidden"), the time of day the experiment was run (morning "am" vs. afternoon "pm"), and whether or not the child waited to take the marshmallow (i.e., whether or not they delayed gratification).
Because we have no observations of children aged zero, we chose to mean-centre the age variable to make the model coefficients' interpretations more intuitive.

`r solend()`

<!-- # q3 -->

`r qbegin(qcounter())`

**Visualise and summarise data patterns that address the RQ (so the reader knows the basic pattern of what you found).**

- [ ] Include a plot of the data which includes all relevant variables (and make sure to refer to it in the text).
  - Tip: If you transform any continuous predictors for analysis, use the transformed version of the predictor in the plot.
  - Tip: Academic writing styles vary, but Elizabeth likes it when figure captions give the reader a brief take-home message about what they should see in the plot.
- [ ] In a sentence or two, describe the overall pattern of results.
  - Tip: No inferential stats have been done yet, so we can't make any claims just yet about whether the predictions were supported. But it is OK to say whether the results appear broadly in line with the prediction.


`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Whether children delayed gratification across ages, marshmallow visibility, and time of day is shown in Figure 1.

```{r fig.cap = "Figure 1: Delay of gratification as a function of age (mean-centered, on x axis), time of day (am or pm, in columns), and marshmallow visibility (hidden or visible, in rows). Each black dot represents one child's action, and the blue lines are the estimated curves of best fit. Children seem to delay gratification more when they are older as well as when the marshmallow is hidden."}
#| code-fold: true
delay_data |>
  ggplot(aes(x = age_c, y = taken)) +
  
  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # facet by categorical predictors
  facet_grid(visibility ~ timeofday) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # add logistic curve
  geom_smooth(
    method = "glm",                          # not a straight line
    method.args = list(family = binomial),   # specify logistic curve
    se = FALSE          # remove standard error ribbon around line
  ) +
  
  # add labels for clarity (currently taken = 1 looks like they took it, not waited!)
  labs(
    x = 'Age (in years, mean-centered)',
    y = 'Delayed gratification?\n(0 = took marshmallow, 1 = waited)'
  )
```

<br>

Figure 1 shows a tendency to delay gratification more often with increasing age.
Additionally, children appear to delay gratification more when the marshmallow is hidden (top row), but we observe no obvious difference between morning and afternoon sessions.



`r solend()`


<!-- # q4 -->

`r qbegin(qcounter())`

**Introduce your statistical analysis (so the reader can interpret the numbers you will report).**

- [ ] State that you're running your analysis using R, and cite the developers (run `citation()` in the console to get the citation for your version of R). 
- [ ] State the $\alpha$ level that you're using for all your analyses (in Psychology, the convention is 5%).
- [ ] Describe the structure of your model in words, e.g.: "We fit a linear model predicting Y as a function of X1, X2, and their interaction".
- [ ] If fitting a logistic regression model, state this clearly.
- [ ] Include the mathematical model formulation of your model.
- [ ] Explain the hypotheses that your analysis will test and include them in mathematical notation.
- [ ] If categorical predictors, describe contrast coding, e.g.: "X1 has two levels, A (coded as –1) and B (coded as 1)".
- [ ] If continuous predictors have been transformed, describe the transformation, e.g.: "X2 has been z-scored".
- [ ] For each model assumption, state whether you are satisfied that the assumption has been met.
- [ ] Include plots for each assumption check as figures in the Appendix section of the document (and refer to them in the main text).
- [ ] If applicable, check for multicollinearity and report VIF or GVIF.
- [ ] If you detected any high-influence observations and then conducted a sensitivity analysis, mention whether the pattern of results stayed the same even after removing each influential observation.
  - Tip: If you'll report your results using the full regression table for your coefficients, then also include the full regression table for the sensitivity analysis in another Appendix section.


`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

We conducted all statistical analyses using R (R Core Team, 2025), and we will consider effects significant at $\alpha = .05$.

To analyse this data, we fit a logistic regression model to estimate the probability of children delaying gratification as a function of children's age, marshmallow visibility, and time of day.
As mentioned above, age has been mean-centered.
Marshmallow visibility has been treatment-coded, with "hidden" as the reference level, and time of day has also been treatment-coded, with "am" as the reference level.

The mathematical model formulation is as follows:

$$
\text{logodds(waited)} = \beta_0 + 
(\beta_1 \cdot \text{visibility}_\text{visible}) + 
(\beta_2 \cdot \text{timeofday}_\text{pm}) + 
(\beta_3 \cdot \text{age})
$$


The null hypothesis we are testing is that all slope parameters are equal to zero, that is, that there is no association between any predictors and the probability of delaying gratification:

$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 2, 3, 4 \\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$

After fitting this model, we ran assumption checks and verified that all standardised deviance residuals fall between –3 and 3 (see Figure A.1 in Appendix A).
To check for problematic correlations between predictors, we calculated each predictor's VIF and verified that all were close to 1, which indicates that multicollinearity is not a problem for this analysis.
Finally, we checked for high-influence observations and verified that no data points have a Cook's Distance greater than 0.5 (see Figure A.2 in Appendix A). 

`r solend()`

<!-- # q5 -->

`r qbegin(qcounter())`

**Report the results of your analysis (so the reader knows exactly what you found).**

- [ ] State which model coefficient(s) are related to which prediction(s), and what direction/sign (positive or negative) you would expect the coefficients to have if the predictions were borne out, if applicable.
- [ ] State the key coefficient estimates along with their 95% CI and p-value, e.g.: "The slope associating X1 and Y was significantly different from zero ($b$ = 0.26, 95% CI [0.23, 0.29], $p$ < .001)" or "The slope associating X1 and Y was not significantly different from zero ($b$ = 0.26, 95% CI [–0.04, 0.56], $p$ = .18)", if applicable.
  - Tip: You don't need to report every single parameter estimate. Focus on the one(s) you identified as being relevant to your prediction(s).
  - Tip: Briefly link the direction and significance of the effect back to your earlier predictions—is the prediction supported by the parameter estimate?
- [ ] Include a table of all coefficients estimated by the model (and make sure to refer to it in the text), if applicable. The function `tab_model()` from the package `sjPlot` may be useful here, but note that `tab_model()` causes Rmd files to fail to knit to PDF, so you might want to manually recreate the table it produces.
  - Tip: if you DO report all coefficient estimates in the text, then there's no need to include the coefficient table because the information is redundant.
- [ ] Plot the model-fitted values (and refer to the plot in the text), if applicable.
- [ ] Report the results of a model comparison, if applicable.
- [ ] Report any other information you consider important for your story!

`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The model indicates that age and marshmallow visibility are both significantly associated with the log-odds of delaying gratification, though time of day is not.
Specifically, a visible marshmallow is estimated to lower the log-odds of delaying gratification by 1.20, holding age and time of day constant ($b$ = –1.20, 95% CI [–1.75, –0.67], $p$ < .001).
**[To do an extra-nice and reader-friendly interpretation, include predicted probabilities!]** In terms of probabilities, the model suggests that children of average age in a morning session when the marshmallow is hidden delay gratification with a probability of about 65.2%, while children of average age in a morning session when the marshmallow is *visible* have a much lower probability of delaying gratification, at only 36.1%.

```{r eval=F}
#| code-fold: true
# to calculate predicted probability with all predictors at 0/ref level:
plogis(
    coef(delay_mdl)[['(Intercept)']]
) |> round(3) * 100

# predicted probability with visibility = 1, all else = 0
plogis(
    coef(delay_mdl)[['(Intercept)']] + coef(delay_mdl)[['visibilityvisible']]
) |> round(3) * 100
```


Further, increasing one year of age is estimated to increase the log-odds of delaying gratification by 0.58, holding visibility and time of day constant ($b$ = 0.58, 95% CI [0.42, 0.76], $p$ < .001).
Compared to the probability of 65.2% that children of average age will delay gratification in morning sessions with a hidden marshmallow, children who are one year older than average (in morning sessions with a hidden marshmallow) are estimated to delay gratification with a probability of 77.1%.

```{r eval=F}
#| code-fold: true

# predicted probability with age_c = 1, all else = 0
plogis(
    coef(delay_mdl)[['(Intercept)']] + coef(delay_mdl)[['age_c']]
) |> round(3) * 100
```

See Table 1 for all estimated model coefficients.
Because at least one slope coefficient is significantly different from zero, we reject the null hypothesis defined above.

```{r}
#| code-fold: true
tab_model(
  delay_mdl,
  transform = NULL,  # display coefficients as log-odds
  dv.labels = 'Log-odds of delaying gratification',
  pred.labels = c(
    'Intercept', 'Visibility [visible]', 'Time of day [pm]', 'Age (mean-centered)'
  ),
  title = "Table 1: Logistic regression model coefficient estimates"
)
```

<br>

Finally, Figure 2 illustrates the estimated probability of delaying gratification for all combinations of age, marshmallow visibility, and time of day.

```{r fig.cap="Figure 2: Model-estimated probabilities of delaying gratification. The black lines indicates the model's estimates, and the shaded ribbons span the 95% CIs. All else held constant, children's probability of delaying gratification increases as they get older and decreases if they can see the marshmallow."}
#| code-fold: true
Effect(
  focal.predictors = c('age_c', 'visibility', 'timeofday'), 
  mod = delay_mdl,
  xlevels = 20  # how many values of `age_c` to estimate outcomes for
) |>
  as.data.frame() |>
  ggplot(aes(x = age_c, y = fit)) +
  # a line for the model-fitted estimates
  geom_line() +
  # a ribbon for the 95% CIs
  geom_ribbon( aes( ymin = lower,  ymax = upper ),  alpha = .3 ) +
  # facet by visibility and timeofday
  facet_grid(visibility ~ timeofday) +
  # axis labels for clarity
  labs(
    y = 'Probability of delaying gratification',
    x = 'Age (years, mean-centered)'
  ) 
```


`r solend()`

<br>


<!-- # q6 -->

`r qbegin(qcounter())`

**Appendices might include:**

- [ ] Figures that visualise assumption checks
- [ ] Sensitivity analysis regression table / any other useful details

`r qend()`
`r solbegin(label='example', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

**Appendix A: Model assumption and diagnostic checks**

```{r fig.cap = "Figure A.1: None of the model's standardised deviance residuals should be outwith the range of –3 to 3. All residuals are within the acceptable range."}
#| code-fold: true
rstandard(delay_mdl, type = "deviance") |> plot()
```

```{r fig.cap = "Figure A.2: No data points appear to exert undue influence on the model's estimates; all Cook's Distance values are well below the threshold of 0.5 for moderate influence."}
#| code-fold: true
plot(delay_mdl, which = 4)
```

`r solend()`

<br>


`r solbegin(label='example – full write-up', show=params$SHOW_SOLS, toggle=params$TOGGLE)`

This study investigates how children delay gratification using the classic marshmallow paradigm.
We are interested in how children's age, the visibility of the marshmallow, and the time of day might play into children's delay gratification behaviour.

We have data from 301 children, including their ages in years, whether the marshmallow was visible to the child or hidden from them ("visible" vs. "hidden"), the time of day the experiment was run (morning "am" vs. afternoon "pm"), and whether or not the child waited to take the marshmallow (i.e., whether or not they delayed gratification).
Because we have no observations of children aged zero, we chose to mean-centre the age variable to make the model coefficients' interpretations more intuitive.

Whether children delayed gratification across ages, marshmallow visibility, and time of day is shown in Figure 1.

```{r fig.cap = "Figure 1: Delay of gratification as a function of age (mean-centered, on x axis), time of day (am or pm, in columns), and marshmallow visibility (hidden or visible, in rows). Each black dot represents one child's action, and the blue lines are the estimated curves of best fit. Children seem to delay gratification more when they are older as well as when the marshmallow is hidden.", echo=F}
delay_data |>
  ggplot(aes(x = age_c, y = taken)) +
  
  # wiggle the points around a little bit so they don't all overlap
  geom_jitter(height = 0.01) +
  
  # facet by categorical predictors
  facet_grid(visibility ~ timeofday) +
  
  # label only 0 and 1 on the y axis
  scale_y_continuous(breaks = c(0, 1)) +
  
  # add logistic curve
  geom_smooth(
    method = "glm",                          # not a straight line
    method.args = list(family = binomial),   # specify logistic curve
    se = FALSE          # remove standard error ribbon around line
  ) +
  
  # add labels for clarity (currently taken = 1 looks like they took it, not waited!)
  labs(
    x = 'Age (in years, mean-centered)',
    y = 'Delayed gratification?\n(0 = took marshmallow, 1 = waited)'
  )
```


<br>

Figure 1 shows a tendency to delay gratification more often with increasing age.
Additionally, children appear to delay gratification more when the marshmallow is hidden (top row), but we observe no obvious difference between morning and afternoon sessions.

We conducted all statistical analyses using R (R Core Team, 2025), and we will consider effects significant at $\alpha = .05$.

To analyse this data, we fit a logistic regression model to estimate the probability of children delaying gratification as a function of children's age, marshmallow visibility, and time of day.
As mentioned above, age has been mean-centered.
Marshmallow visibility has been treatment-coded, with "hidden" as the reference level, and time of day has also been treatment-coded, with "am" as the reference level.

The mathematical model formulation is as follows:

$$
\text{logodds(waited)} = \beta_0 + 
(\beta_1 \cdot \text{visibility}_\text{visible}) + 
(\beta_2 \cdot \text{timeofday}_\text{pm}) + 
(\beta_3 \cdot \text{age})
$$


The null hypothesis we are testing is that all slope parameters are equal to zero, that is, that there is no association between any predictors and the probability of delaying gratification:

$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 2, 3, 4 \\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$

After fitting this model, we ran assumption checks and verified that all standardised deviance residuals fall between –3 and 3 (see Figure A.1 in Appendix A).
To check for problematic correlations between predictors, we calculated each predictor's VIF and verified that all were close to 1, which indicates that multicollinearity is not a problem for this analysis.
Finally, we checked for high-influence observations and verified that no data points have a Cook's Distance greater than 0.5 (see Figure A.2 in Appendix A).


The model indicates that age and marshmallow visibility are both significantly associated with the log-odds of delaying gratification, though time of day is not.
Specifically, a visible marshmallow is estimated to lower the log-odds of delaying gratification by 1.20, holding age and time of day constant ($b$ = –1.20, 95% CI [–1.75, –0.67], $p$ < .001).
In terms of probabilities, the model suggests that children of average age in a morning session when the marshmallow is hidden delay gratification with a probability of about 65.2%, while children of average age in a morning session when the marshmallow is *visible* have a much lower probability of delaying gratification, at only 36.1%.

Further, increasing one year of age is estimated to increase the log-odds of delaying gratification by 0.58, holding visibility and time of day constant ($b$ = 0.58, 95% CI [0.42, 0.76], $p$ < .001).
Compared to the probability of 65.2% that children of average age will delay gratification in morning sessions with a hidden marshmallow, children who are one year older than average (in morning sessions with a hidden marshmallow) are estimated to delay gratification with a probability of 77.1%.

See Table 1 for all estimated model coefficients.
Because at least one slope coefficient is significantly different from zero, we reject the null hypothesis defined above.

```{r echo=F}
tab_model(
  delay_mdl,
  transform = NULL,  # display coefficients as log-odds
  dv.labels = 'Log-odds of delaying gratification',
  pred.labels = c(
    'Intercept', 'Visibility [visible]', 'Time of day [pm]', 'Age (mean-centered)'
  ),
  title = "Table 1: Logistic regression model coefficient estimates"
)
```

<br>

Finally, Figure 2 illustrates the estimated probability of delaying gratification for all combinations of age, marshmallow visibility, and time of day.

```{r fig.cap="Figure 2: Model-estimated probabilities of delaying gratification. The black lines indicates the model's estimates, and the shaded ribbons span the 95% CIs. All else held constant, children's probability of delaying gratification increases as they get older and decreases if they can see the marshmallow.", echo=F}
Effect(
  focal.predictors = c('age_c', 'visibility', 'timeofday'), 
  mod = delay_mdl,
  xlevels = 20  # how many values of `age_c` to estimate outcomes for
) |>
  as.data.frame() |>
  ggplot(aes(x = age_c, y = fit)) +
  # a line for the model-fitted estimates
  geom_line() +
  # a ribbon for the 95% CIs
  geom_ribbon( aes( ymin = lower,  ymax = upper ),  alpha = .3 ) +
  # facet by visibility and timeofday
  facet_grid(visibility ~ timeofday) +
  # axis labels for clarity
  labs(
    y = 'Probability of delaying gratification',
    x = 'Age (years, mean-centered)'
  ) 
```

<br>


**Appendix A: Model assumption and diagnostic checks**

```{r fig.cap = "Figure A.1: None of the model's standardised deviance residuals should be outwith the range of –3 to 3. All residuals are within the acceptable range.", echo=F}
rstandard(delay_mdl, type = "deviance") |> plot()
```

```{r fig.cap = "Figure A.2: No data points appear to exert undue influence on the model's estimates; all Cook's Distance values are well below the threshold of 0.5 for moderate influence.", echo=F}
plot(delay_mdl, which = 4)
```


<br>

**References**

R Core Team (2025). _R: A Language and Environment for Statistical Computing_. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.

`r solend()`

