---
title: "Categorical/categorical example"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


RQ: **Do differences in body mass between species differ by sex?**

```{r}
library(palmerpenguins)
```

Here we are using the *palmer penguins* dataset^[Horst, A. M., Hill, A. P., & Gorman, K. B. (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/ doi: 10.5281/zenodo.3960218]. The dataset contains a variety of information for a total of 344 adult penguins, including body measurements (bill length (`bill_length_mm`; measured in mm)/depth (`bill_length_mm`; measured in mm), flipper length (`flipper_length_mm`; measured in mm), body mass (`body_mass_g`; measured in grams)) for male and female (`sex`) penguins from three species (`species`; categorised as Adelie, Chinstrap, or Gentoo) across three Antarctic islands (`island`; categorised as Biscoe, Dream, or Torgersen) from 2007-2009 (`year`).

The variables we'll use to address the RQ are `body_mass_g`, `species`, and `sex`.


## Summarise/describe data

We can summarise the continuous outcome variables, grouped by the two categorical predictor variables, as follows.

```{r}
penguins |>
  group_by(species, sex) |>
  summarise(
    BodyMass_M  = mean(body_mass_g),
    BodyMass_SD = sd(body_mass_g),
  ) |>
  drop_na() |> # some rows have missing values – comment out this line and see what changes!
  kableExtra::kable(
    caption = "Descriptive statistics for body mass (in g) for different penguin species and different sexes", 
    digits = 2
  ) |>
  kableExtra::kable_styling(full_width = FALSE)
```


## Visualise data

To plot a continuous outcome  based on two categorical predictors, we can use violin plots again to plot the first predictor against the outcome, and distinguish levels of the second predictor using either facets or colour.

Facetting by `species`, colouring by `sex` (try it yourself facetting by `sex` and colouring by `species` instead):

```{r}
penguins |>
  drop_na() |>  # drop any rows containing NAs
  ggplot(aes(x = sex, y = body_mass_g, fill = sex, colour = sex)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 3) +
  theme(legend.position = 'none') +
  facet_wrap(~ species) +
  labs(
    y = 'Body mass (in g)',
    x = 'Sex',
  )
```


Only colouring by `sex` (try it yourself colouring by `species` instead):

```{r}
penguins |>
  drop_na() |>  # drop any rows containing NAs
  ggplot(aes(x = species, y = body_mass_g, fill = sex, colour = sex)) +
  geom_violin(alpha = 0.5) +
  geom_point(
      alpha = 0.5, 
      position = position_jitterdodge(dodge.width = 0.9)  # align points + violins
  ) +
  stat_summary(
      fun = mean, 
      geom = 'point', 
      colour = 'black', 
      show.legend = F,  # drop from legend
      size = 3,
      position = position_dodge(width = 0.9)  # align means + violins
  ) +
  labs(
    y = 'Body mass (in g)',
    x = 'Penguin species',
    fill = 'Sex',
    colour = 'Sex'
  )
```


## Set up predictors

Before we specify our model, we must select (and provide justification for this choice) the reference group(s) for our categorical variable(s). 

In this specific example, there is no natural reference category, nor one that maps to our RQ, so we will go with `R`'s default coding and have Adelie as our reference group for species, and female for sex.

```{r}
contrasts(penguins$species)
contrasts(penguins$sex)
```


## How many interaction terms will the model contain?

The number of interaction terms is given by

$$
(r - 1) \times (c - 1)
$$

- $r$ (which stands for "rows") is the number of levels in the first interacting predictor;
- $c$ (which stands for "columns") is the number of levels in the second interacting predictor.

To see why we are talking about "rows" and "columns", imagine the variables arranged in a table like this:

|              | female | male |
| ------------ | --- | ---- |
| Adelie |  ...   |   ...   |
| Chinstrap | ...    |     ... |
| Gentoo |...     |...      |

<br>

So in our 3x2 data, where $r=3$ and $c=2$, we have

$$
\begin{align}
  & (r - 1) \times (c - 1)\\
  =~& (3 - 1) \times (2 - 1) \\
  =~& 2 \times 1 \\
  =~& 2
\end{align}
$$

interaction terms.



## Mathematical model specification

$$
\begin{align}
\text{body\_mass\_g} ~=~ & \beta_0 + 
(\beta_1 \cdot \text{species}_\text{Chinstrap}) + 
(\beta_2 \cdot \text{species}_\text{Gentoo}) + 
(\beta_3 \cdot \text{sex}_\text{male}) ~+  \\
& (\beta_4 \cdot \text{species}_\text{Chinstrap} \cdot \text{sex}_\text{male}) +
(\beta_5 \cdot \text{species}_\text{Gentoo} \cdot \text{sex}_\text{male}) + 
\epsilon
\end{align}
$$


Write:

```
$$
\begin{align}
\text{body\_mass\_g} ~=~ & \beta_0 + 
(\beta_1 \cdot \text{species}_\text{Chinstrap}) + 
(\beta_2 \cdot \text{species}_\text{Gentoo}) + 
(\beta_3 \cdot \text{sex}_\text{male}) ~+  \\
& (\beta_4 \cdot \text{species}_\text{Chinstrap} \cdot \text{sex}_\text{male}) +
(\beta_5 \cdot \text{species}_\text{Gentoo} \cdot \text{sex}_\text{male}) + 
\epsilon
\end{align}
$$
```

## H0 and H1


For most RQs that require interaction models, we're interested in testing the hypothesis that there's no difference in how some predictor `x` is associated with the outcome `y` for different values of `z`.
In other words, we are interested in testing the null hypothesis that the **interaction term is zero**.

For the model above:

$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 4, 5\\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$

Write:

```
$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 4, 5\\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$
```



## Fit the model

```{r}
mdl_cc <- lm(body_mass_g ~ species * sex, data = penguins)
```

Each row in the `summary()` output of the model will correspond to one of the estimated $\beta$’s in the equation above.  

```{r}
summary(mdl_cc)
```


## Interpret results

### $\beta_0$ = `(Intercept)` = `r round(coef(mdl_cc)[1],2)`  

- A female Adelie penguin is expected to have a body mass of `r round(coef(mdl_cc)[1],2)` g.  
    
###  $\beta_1$ = `speciesChinstrap` = `r round(coef(mdl_cc)[2],2)`  

- A female Chinstrap penguin is estimated to have a body mass `r round(coef(mdl_cc)[2],2)` g greater than a female Adelie penguin.
- This estimate is significantly different from zero $(p = .014)$.
    
### $\beta_2$ = `speciesGentoo` = `r round(coef(mdl_cc)[3],2)`  

- A female Gentoo penguin is estimated to have a body mass `r round(coef(mdl_cc)[3],2)` g greater than a female Adelie penguin.
- This estimate is significantly different from zero $(p < .001)$.

### $\beta_3$ = `sexmale` = `r round(coef(mdl_cc)[4],2)`  

- A male Adelie penguin is estimated to have a body mass `r round(coef(mdl_cc)[4],2)` g greater than a female Adelie penguin.
- This estimate is significantly different from zero $(p < .001)$.


###  $\beta_4$ = `speciesChinstrap:sexmale` = `r round(coef(mdl_cc)[5],2)`  

- For Chinstrap penguins, the difference between male and female penguins is estimated to decrease by 262.9 g compared to the male/female difference for Adelie penguins.
- This estimate is significantly different from zero $(p = .004)$.


###  $\beta_5$ = `speciesGentoo:sexmale` = `r round(coef(mdl_cc)[6],2)`  

- For Gentoo penguins, the difference between male and female penguins is estimated to increase by `r round(coef(mdl_cc)[6],2)` g compared to the male/female difference for Adelie penguins.



## Visualise the model


We can do this using the `cat_plot()` function from the **interactions** package.

In terms of of specification, it might be useful to look up the helper function  (i.e., `?cat_plot`). As a quick guide:

- `model = `: The name model to be used  
- `pred = `: The categorical predictor variable that will appear on the x-axis  
- `modx = `: The categorical moderator variable  

Remember to give your plot informative titles/labels. You, for example, likely want to give your plot:

- a clear and concise title (specify `main.title = `)  
- axis labels with units or scale included (specify `x.label = ` and `y.label = `)  
- a legend title (specify `legend.main = `)  

```{r}
library(interactions)

plt_mdl_cc <- cat_plot(
  model = mdl_cc,
  pred = species,
  modx = sex,
  main.title = "Body mass across species and sex",
  x.label = "Species",
  y.label = "Body mass (in g)",
  legend.main = "Sex"
)
plt_mdl_cc
```



## Testing simple effects

`probe_interactions()` would test simple slopes for continuous predictors, but it doesn't work quite the same way for categorical predictors, so we need some new machinery.
Specifically, we'll use the `emmeans` package and its function `emmeans()`.

```{r}
library(emmeans)
```


Imagine we want to find the simple effects of sex for each species of penguin: that is, how much do males and females differ for Adelie penguins, and for Chinstrap penguins, and for Gentoo penguins, and are each of those differences significantly different from zero?

You can run this analysis step by step:

```{r}
# Use `emmeans()` to get estimated body mass 
# for all combinations of species and sex
mdl_cc_emm <- emmeans(mdl_cc, ~ species * sex)

## Use `pairs()` to find the simple effects of sex for each species
pairs(mdl_cc_emm, simple = "sex")
```

Or you can do the same analysis all in one command:

```{r}
pairs(
  emmeans(mdl_cc, ~ species * sex), 
  simple = "sex"
)
```

In this case, the differences between males and females for all penguin species are significantly different from zero.
