---
title: "Correcting for multiple comparisons"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

## The multiple comparisons problem

Think back to Type I errors from DAPR1.
When we conduct an hypothesis test, and we set $\alpha = .05$, we will reject the null hypothesis $H_0$ when we find a $p < .05$. 

Now remember what a $p$-value represents: the chance of observing a statistic at least as extreme as the one we do have, assuming the null hypothesis to be true.
This means that *if* $H_0$ **is** true, then we will still observe a $p < .05$ 5\% of the time.
So, our chance of making a Type I error = the threshold ($\alpha$) at which below a $p$-value results in us rejecting $H_0$. 

We accept a 5% error rate overall.
But this error rate applies to each statistical test we run.
So if we conduct an analysis in which we plan on conducting lots of tests of different comparisons, the chance of an error being made increases substantially.
Across all the tests performed, that chance will be much higher than our accepted risk level of 5\%.


## When to correct for multiple comparisons

Whenever you are running more than one test using the same data and using them both to draw conclusions.
This could be:

- two linear models
- one linear model and one t-test
- one linear model and one chi-squared test
- ...

In this scenario, the simplest correction to use is Bonferroni (see below).

(We see this a lot in Honours dissertations: many tests are run to address several related RQs, but the p-values are not corrected, so the level of risk is higher than our accepted 5%.)


## How to correct for multiple comparisons

### When testing simple effects: Tukey and Bonferroni

Think back to our `penguins` example from [the categorical/categorical interactions flash card](cat-cat.html){target="_blank"}.

```{r}
library(palmerpenguins)

# predict body mass based on species, sex, and their interaction
mdl_cc <- lm(body_mass_g ~ species * sex, data = penguins)

# test the simple effects of species for each sex
pairs(
  emmeans::emmeans(mdl_cc, ~ species * sex), 
  simple = "species"
)
```

Within each `sex`, we are running three statistical tests, and so the $p$-values of those tests need to be adjusted.

By default, `pairs()` uses the Tukey correction.

::: {.callout-caution collapse="true"}
#### How does the Tukey correction work?

- Tukey's HSD compares the means of the members of one pair in a pairwise comparison.
- The difference between group means is represented as a **q-statistic.**
- The q-statistic is compared to a critical value from the **studentised range distribution** to decide whether the difference is significant.

To see how to calculate this by hand, here's [a YouTube video that walks through the calculation](https://www.youtube.com/watch?v=EBGnAJHs3G4&t=218s){target="_blank"}.
(The main action is between 3:41–7:42.)

:::

You can also tell `pairs()` to use the Bonferroni correction, which divides the $\alpha$ level by the number of tests / multiplies the $p$-value by the number of tests.

```{r}
pairs(
  emmeans::emmeans(mdl_cc, ~ species * sex), 
  simple = "species",
  adjust = 'bonferroni'
)
```


When testing simple effects, it's fine to just use the Tukey correction, because it's specifically designed for comparing pairs of group means.


### Any other scenario: Bonferroni

As mentioned above, the Bonferroni correction divides the $\alpha$ level by the number of tests / multiplies the $p$-value by the number of tests.

This means you can apply it yourself to the results of any tests that give you $p$-values.

