---
title: "Numeric/categorical example"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
```


RQ: **Does the association between body mass and flipper length differ between species of penguin?**

```{r}
library(palmerpenguins)
```

Here we are using the *palmer penguins* dataset^[Horst, A. M., Hill, A. P., & Gorman, K. B. (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/ doi: 10.5281/zenodo.3960218]. The dataset contains a variety of information for a total of 344 adult penguins, including body measurements (bill length (`bill_length_mm`; measured in mm)/depth (`bill_length_mm`; measured in mm), flipper length (`flipper_length_mm`; measured in mm), body mass (`body_mass_g`; measured in grams)) for male and female (`sex`) penguins from three species (`species`; categorised as Adelie, Chinstrap, or Gentoo) across three Antarctic islands (`island`; categorised as Biscoe, Dream, or Torgersen) from 2007-2009 (`year`).

The variables that are relevant to us are `body_mass_g`, `flipper_length_mm`, and `species`.

## Summarise/describe data

We can summarise the two continuous variables, grouped by the one categorical variable, as follows.

```{r}
penguins |>
    group_by(species) |>
    summarise(
        BodyMass_M = mean(body_mass_g, na.rm = T),
        BodyMass_SD = sd(body_mass_g, na.rm = T),
        FlipperLength_M = mean(flipper_length_mm, na.rm = T),
        FlipperLength_SD = sd(flipper_length_mm, na.rm = T)
    ) |>
    kableExtra::kable(
        caption = "Descriptive statistics for body mass and flipper length in each penguin species",
        digits = 2
    ) |>
    kableExtra::kable_styling(full_width = FALSE)
```




## Visualise data

There are several options for visualising a continuous outcome variable with one continuous and one categorical predictor.

We can make a scatterplot for the two continuous variables, adding colour to distinguish the levels of each categorical variable, which will by default generate lines of best fit for each level separately.

```{r}
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Flipper Length (in mm)", y = "Body Mass (in g)")
```

We can also distinguish the three species using facetting instead of (or in addition to) colour.

```{r, warning=FALSE}
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_wrap(~species) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Flipper Length (in mm)", y = "Body Mass (in g)")
```


## Set up predictors

Before we specify our model, we must select (and provide justification for this choice) the reference group(s) for our categorical variable(s). In this specific example, there is no natural reference category, nor one that maps to our RQ, so we will go with `R`'s default coding and have Adelie as our reference group for species. 

```{r}
contrasts(penguins$species)
```

The contrast matrix shows we'll have two dummy variables, one for `Chinstrap` and one for `Gentoo`.

Next, flipper length.
We could use the flipper length variable as-is, since the millimetres scale is meaningful.
However, no penguins in our dataset had a flipper length of zero mm, so it might be more intuitive to mean-centre `flipper_length_mm` so that zero instead represents the average flipper length.

```{r}
penguins <- penguins |>
  mutate(
    flipper_length_mm_c = flipper_length_mm - mean(flipper_length_mm, na.rm = T)
  )

mean(penguins$flipper_length_mm_c, na.rm=T) |> round()
```


## Mathematical model specification

$$
\begin{align}
\text{body\_mass} ~=~ & \beta_0 + (\beta_1 \cdot \text{flipper\_length\_c}) + 
(\beta_2 \cdot \text{species}_\text{Chinstrap}) + 
(\beta_3 \cdot \text{species}_\text{Gentoo}) ~+ \\
& (\beta_4 \cdot \text{flipper\_length\_c} \cdot \text{species}_\text{Chinstrap}) ~+ \\
& (\beta_5 \cdot \text{flipper\_length\_c} \cdot \text{species}_\text{Gentoo}) + \epsilon \\
\end{align}
$$

Write:

```
$$
\begin{align}
\text{body\_mass} ~=~ & \beta_0 + (\beta_1 \cdot \text{flipper\_length\_c}) + 
(\beta_2 \cdot \text{species}_\text{Chinstrap}) + 
(\beta_3 \cdot \text{species}_\text{Gentoo}) ~+ \\
& (\beta_4 \cdot \text{flipper\_length\_c} \cdot \text{species}_\text{Chinstrap}) ~+ \\
& (\beta_5 \cdot \text{flipper\_length\_c} \cdot \text{species}_\text{Gentoo}) + \epsilon \\
\end{align}
$$
```

## H0 and H1

For most RQs that require interaction models, we're interested in testing the hypothesis that there's no difference in how some predictor `x` is associated with the outcome `y` for different values of `z`.
In other words, we are interested in testing the null hypothesis that the **interaction term is zero**.

For the model above:

$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 4, 5\\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$

Write:

```
$$
\begin{align}
H_0 &: \text{All } \beta_j = 0 \text{ for } j = 4, 5\\
H_1 &: \text{Any } \beta_j \neq 0\\
\end{align}
$$
```



## Fit the model

```{r}
mdl_nc <- lm(body_mass_g ~ flipper_length_mm_c * species, data = penguins)
```

Each row in the `summary()` output of the model will correspond to one of the estimated $\beta$’s in the equation above.  

```{r}
summary(mdl_nc)
```


## Interpret results

### $\beta_0$ = `(Intercept)` = `r round(coef(mdl_nc)[1],2)`  

- An Adelie penguin with an average flipper length was expected to have a body mass of `r round(coef(mdl_nc)[1],2)` g.  
    
###  $\beta_1$ = `flipper_length_mm_c` = `r round(coef(mdl_nc)[2],2)`  

- For an Adelie penguin, every 1 additional mm in flipper length was associated with a `r round(coef(mdl_nc)[2],2)` g increase in their body mass.
- This estimate is significantly different from zero $(p < .001)$.
    
### $\beta_2$ = `speciesChinstrap` = `r round(coef(mdl_nc)[3],2)`  

- A Chinstrap penguin with an average flipper length is estimated to have a body mass 151.42 g lighter than an Adelie penguin with the same flipper length.
- This estimate is not statistically different from zero $(p = .062)$.

### $\beta_3$ = `speciesGentoo` = `r round(coef(mdl_nc)[4],2)`  

- A Gentoo penguin with an average flipper length is estimated to have a body mass `r round(coef(mdl_nc)[4],2)`g heavier than an Adelie penguin with the same flipper length. 
- This estimate is not statistically different from zero $(p = .242)$.  

###  $\beta_4$ = `flipper_length_mm_c:speciesChinstrap` = `r round(coef(mdl_nc)[5],2)`  

- For a Chinstrap penguin, every additional 1 mm in flipper length is associated with an *additional* `r round(coef(mdl_nc)[5],2)` g increase to their body mass, compared to Adelie penguins.
- This estimate is not statistically different from zero $(p = .825)$.  

###  $\beta_5$ = `flipper_length_mm_c:speciesGentoo` = `r round(coef(mdl_nc)[6],2)`  

- For a Gentoo penguin, every additional 1 mm in flipper length is associated with an *additional* `r round(coef(mdl_nc)[6],2)` g increase to their body mass, compared to Adelie penguins.
- This estimate is statistically different from zero $(p = .002)$.  


## Visualise the model

We can do this using the `probe_interaction()` function from the **interactions** package.  

In terms of of specification, it might be useful to look up the helper function (i.e., `?probe_interaction`). As a quick guide:

- `model = `: The name model to be used  
- `pred = `: The continuous predictor variable that will appear on the x-axis  
- `modx = `: The categorical moderator variable  
- `interval = `: If we say `TRUE`, then confidence/prediction intervals will be plotted around the line   

Remember to give your plot informative titles/labels. You, for example, likely want to give your plot:

- a clear and concise title (specify `main.title = `)  
- axis labels with units or scale included (specify `x.label = ` and `y.label = `)  
- a legend title (specify `legend.main = `)  

```{r}
library(interactions)

probe_mdl_nc <- probe_interaction(
  model = mdl_nc,
  pred = flipper_length_mm_c,
  modx = species,
  interval = T,
  main.title = "Predicted body mass across flipper length by species",
  x.label = "Flipper Length (in mm; mean-centered)",
  y.label = "Body mass (in g)",
  legend.main = "Penguin species"
)
probe_mdl_nc$interactplot
```

Alternatively, we can directly use the `interact_plot()` function and simply write

```{r}
plot_mdl_nc <- interact_plot(
  model = mdl_nc,
  pred = flipper_length_mm_c,
  modx = species,
  interval = T,
  main.title = "Predicted body mass across flipper length by species",
  x.label = "Flipper Length (in mm; mean-centered)",
  y.label = "Body mass (in g)",
  legend.main = "Penguin species"
)
plot_mdl_nc
```

