---
title: "Interaction models overview"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
```

## Specifying interaction models

Interaction models allow us to model the idea of “the association between $x_1$ and $y$ differs depending on the level of $x_2$” by including a product (multiplication) term between the two predictors.

**Formula:**  

$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 (x_1 \cdot x_2) + \epsilon
$$


If we fit the interaction `x1:x2`, we almost always want to also fit the separate effects `x1` and `x2`:

:::quote
"Except in special circumstances, a model including a product term for interaction between two explanatory variables should also include terms with each of the explanatory variables individually, even though their coefficients may not be significantly different from zero. Following this rule avoids the logical inconsistency of saying that the effect of $X_1$ depends on the level of $X_2$ but that there is no effect of $X_1$."  
--- Ramsey, 2012
:::


:::blue


In **R**:

There are basically two pieces of information that we need to pass to the `lm()` function:

1. The formula: The regression formula should be specified in the form `y ~ x` where $y$ is the dependent variable (DV), $x_1$ the first independent variable (IV1), and $x_2$ the second independent variable (IV2). We then also need to include the interaction term between $x_1$ and $x_2$.

2. The data: Specify which dataframe (`data = `) contains the variables specified in the formula.

+ run interaction model via `lm()` function

```{r eval=FALSE}
model_name <- lm(DV ~ IV1*IV2, data = data_name)
```

**OR**

```{r eval=FALSE}
model_name <- lm(DV ~ IV1 + IV2 + IV1:IV2, data = data_name)
```

:::

## Interpreting coefficients


::: {.panel-tabset}

#### Interpreting coefficients for A and B in the presence of an interaction A:B

When you include an interaction between $x_1$ and $x_2$ in a regression model, you are estimating the extent to which the effect of $x_1$ on $y$ is different across the values of $x_2$.  

What this means is that the effect of $x_1$ on $y$ *depends on/is conditional upon* the value of $x_2$ (and vice versa, the effect of $x_2$ on $y$ is different across the values of $x_1$).  
  
This means that we can no longer talk about the "effect of $x_1$ _holding $x_2$ constant_". Instead we can talk about a _marginal effect_ of $x_1$ on $y$ at a specific value of $x_2$. 

:::frame
When we fit the model $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 (x_1 \cdot x_2) + \epsilon$ using `lm()`:  

- the parameter estimate $\hat \beta_1$ is the _marginal effect_ of $x_1$ on $y$ where $x_2 = 0$  
- the parameter estimate $\hat \beta_2$ is the _marginal effect_ of $x_2$ on $y$ where $x_1 = 0$ 
In other words, when we fit a model with an interaction in R, we get out coefficients for both predictors, and for the interaction. The coefficients for each individual predictor reflect the effect on the outcome when the other predictor is zero.

:::

<div style="margin-left: 15px">
<small>
__N.B.__ Regardless of whether or not there is an interaction term in our model, all parameter estimates in multiple regression are "conditional" in the sense that they are dependent upon the inclusion of other variables in the model. For instance, in $y = \beta_0 + \beta_1 \cdot x_1 + \beta_2 \cdot x_2 + \epsilon$ the coefficient $\hat \beta_1$ is conditional upon holding $x_2$ constant. 
</small>
</div>

#### Interpreting the interaction term A:B 

The coefficient for an interaction term can be thought of as providing an _adjustment to the slope._   

Let's take the following model as an example:

```
lm(formula = y ~ x1 * x2, data = df)
...

...
Coefficients:
             Estimate Std. Error t value Pr(>|t|)  
(Intercept)  ...        ....       ...      ...  
x1           ...        ....       ...      ...  
x2           ...        ....       ...      ...  
x1:x2        ...        ....       ...      ...  
---
```

These coefficients can be interpreted, in turn as:  

| Coefficient      | Interpretation |
| ----------- | ----------- |
| `(Intercept)` | the estimated $y$ when all predictors ($x_1$ and $x_2$) are zero is [*estimate*] |
| `x1`  | **when $x_2$ is zero,** a 1 unit increase in $x_1$ is associated with a [*estimate*] change in $y$ |
| `x2`  | **when $x_1$ is zero,** a 1 unit increase in $x_2$ is associated with a [*estimate*] change in $y$. |
| `x1:x2`  | as $x_2$ increases by 1, the association between $x_1$ and $y$ changes by [*estimate*]<br>_**or**_<br>as $x_1$ increases by 1, the association between $x_2$ and $y$ changes by [*estimate*] |

:::

::: {.callout-note}
#### What if there are other things (e.g., other predictors/covariates) in the model too?

Note that the interaction `x1:x2` changes how we interpret the individual coefficients for `x1` and `x2`.  

It does __*not*__ change how we interpret coefficients for other predictors that might be in our model. For variables that **aren't** involved in the interaction term, these are still held constant.  

For example, suppose we _also_ had another predictor $c_1$ in our model: 

```
lm(y ~ c1 + x1 + x2 + x1:x2)
```

| Coefficient      | Interpretation |
| ----------- | ----------- |
| `(Intercept)` | the estimated $y$ when all predictors ($c_1$, $x_1$ and $x_2$) are zero is [*estimate*] |
| `c1` | a 1 unit increase in $c_1$ is associated with a [*estimate*] increase in $y$, holding constant all other variables in the model ($x_1$ and $x_2$) |
| `x1`  | holding $c_1$ constant, **when $x_2$ is zero,** a 1 unit increase in $x_1$ is associated with a [*estimate*] change in $y$ |
| `x2`  | holding $c_1$ constant, **when $x_1$ is zero,** a 1 unit increase in $x_2$ is associated with a [*estimate*] change in $y$. |
| `x1:x2`  | holding $c_1$ constant, as $x_2$ increases by 1, the association between $x_1$ and $y$ changes by [*estimate*]<br>_**or**_<br>holding $c_1$ constant, as $x_1$ increases by 1, the association between $x_2$ and $y$ changes by [*estimate*] |

:::



