---
title: "<b>Course Introduction </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #  base_color = "#0F4C81", # DAPR1
  base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  # base_color = "#95A5A6", #intro
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)

library(tidyverse)
```


```{r welcome, echo=FALSE}
cohort <- 2021
team <- c("Aja", "Alex", "Emma", "Josiah", "Tom", "Umberto", "Zach", "all the tutors")
lecture_hours <- 10+10+50
lab_hours <- 20+20
# leaving this here for posterity :) MK, 2024
```


# dapR2 Team

--

+ Marju Kaps (CO, Lectures, Labs)

+ Emma Waterston (CO, Labs)

+ Zach Horne (Lectures)

+ Wonderful tutors (Labs)


---
# What is R? 
--

+ A very flexible programming language for all things data.
  + It does pretty much any statistical method you can think of
  + But it does a lot more

+ We will also continue to teach you R Markdown 
  + this is a really neat way to integrate text and analysis
  + and to write reproducible documents and analyses
  
+ Open source (free) software

+ Huge community worldwide which means lots of free resources in addition to your course materials


---
# dapR2 

--
+ In dapR2, we will teach you how to...

--

  + Deal with data in R, tidy it, describe it and visualise it

--

  + Run, interpret and use linear models for observational designs
  
--

  + Run, interpret and use linear models for experimental designs
  
--

  + Introduce the basic concepts of the generalised linear model

--

  + Talk about some common issues when applying the linear model to data.


---
# dapR2 building from dapR1

--
+ dapR2 builds on dapR1

--

+ Core concepts:
  + types of data
  + plotting
  + significance testing
  + bootstrap
  + correlation
  + independent t-tests
  
--

+ You can access these materials via the welcome week folder in LEARN
  
---
# dapR2 

--

+ The course will help develop a large number of transferable skills:

--

  + Working with data

--

  + Logical thinking and problem solving

--

  + Collaborative working (in labs solving the lab problems)
  
--

  + Report writing (taking something complex, and presenting it in a digestible way)

---
# dapR2 Topics

--

+ **Semester 1:**

  + Weeks 1-4: Basics of the linear model
  + Week 5: Application of weeks 1-4
  + Week 6: Catch-up week (no lectures, labs, quiz, office hours)
  + Weeks 7-8: Working with categorical data
  + Weeks 9-10: Robustness checks for our models
  + Week 11: Application of weeks 7-10 
  
--

+ **Semester 2:**

  + Weeks 1-4: Interactions & Analysing Experiments
  + Week 5: Application of weeks 1-4
  + Flexible Learning Week (no lectures, labs, quiz, office hours)
  + Week 6: Power Analyses
  + Weeks 7-8: Generalised Linear Model
  + Week 9: Application of weeks 6-8
  + Week 10: Revision
  

---
# dapR2 Course Delivery

--

+ **Lectures**

  + Two 50 minute lectures per week.
  + These will be a mix of new content and practical demonstration/Q&A (like the Live Rs last year)
  + ***Core purpose***: Learning new material

--

+ **Labs**

  + Check your personal timetable for your lab allocation
  + 1 hour per week in-person labs (additional self-study expected)
  + Structured reading and exercises with R
  + ***Core purpose***: VERY important sessions for practical skills

  
---
# dapR2 Course Delivery

--

+ **Supported by**

  + Office hours (see LEARN for times)
  + Piazza peer discussion boards
  + ***Core purpose***: Resolving ambiguities and questions, helping answer others' questions

--

+ **Reading**

  + All reading is from free-to-access sources
  + You do not need to buy a book and should be able to access everything electronically
  + ***Core purpose***: Scaffolding knowledge


---
# dapR2: Assessments (Overview)

--

+ **Weekly quizzes** (10%)

  + 20 quizzes in total (Quizzes 1 and 2 for practice, the rest marked)
  + Four lowest quizzes dropped - mark is the average of your best 14/18 scores
  + Released Mondays at 9am, due Sunday at 5pm
  + Time limit 60 minutes
  + Make sure to complete, in order to get feedback

--

+ **Coursework report** (30%)

  + Group-work based (working in groups of approx. 5)
  + Groups formed early Semester 2 - if you are not present in your Lab session, you will be randomly allocated
  + Report completed in your groups outside of teaching time (not in the Lab sessions)
  + Set Thursday 13th February at 12 noon (Semester 2 Week 5), Due 12 noon Thursday 6th March 2025 (Semester 2 Week 7)
  + Grades will be peer-adjusted - you will rate each group member on a number of criteria
  + Provided a dataset and some questions, your task is to construct models to answer the questions, describe them, run them, report the results

---
# dapR2: Assessments (Overview)

--

+ **Exam** (60%)

  + Summer exam block
  + Mix of MCQ, short answer R/calculation, and larger interpretation questions
  + Closed book, though you will be given an equation sheet to refer to within the exam
  + Sessions for example exam questions and Q&A taking place at the end of Semester 2


---
# dapR2: Assessments: Why?

--

+ **Weekly quizzes** (10%)

  + Encourage continual engagement
  + Act as a knowledge check for you each week to monitor your progress

--

+ **Coursework report** (30%)

  + To practice working with data, converting questions to models, and interpreting results
  + Continue to develop your group work skills
  + This is what you will be doing in your Mini-dissertation (year 3) and Dissertation (year 4)

--

+ **Exam** (60%)

  + Primary assessment of individual learning
  
---
# dapR2: Academic Integrity

--
 **All assessments are individual assessments (with the exception of the group report)**

--

+ What is acceptable (encouraged):

  + Going over your quizzes together once your marks and feedback have been released
  + Discussing the approach you plan to take to the report
  + Helping one another think through and solve problems (as in the lab)
  + Revising with one another and helping each other prepare for the exam

--

+ Some examples of academic misconduct (not acceptable!):

  + Copying code for the report
  + Copying blocks of text for the report
  + Sharing questions and answers on quizzes
  + Cheating in the exam
  + Use of AI tools (such as GPT) for any assessment on this course

--

+ We will report cases of misconduct

  
---
# Materials

--
+ All of your materials for each week will be within the weekly folders on LEARN

--

+ It will contain:

  + Slide decks and links to lecture recordings
  + Links to lab material
  + Links to the weekly quizzes
  + Any comments or specific instructions for the week
  + Links to discussion boards
  + Reading (all freely available online)

--

+ The weekly folders will become available incrementally.

--




---
# Expectations

--

**What you can expect from us**

1. We will work hard to help you learn

2. We will be open and communicate with you

3. We will be polite, respectful and treat you like adults


--

**What we expect of you**

1. You work hard

2. That you talk to us

3. That you are polite, and respect the teaching team and your classmates

4. Try and have fun

---
# A brief word on engagement

--

+ dapR courses require consistent work across the course

  + You can not really cram this content

--

+ We want to make sure everyone keeps on top of the course

  + To do that, we need to know when people are struggling

--

+ Please help us by being proactive:

  + Come to all sessions
  + Use office hours to clarify any questions you have
  + Talk to peers at your lab session and consider meeting for study groups to learn from one another
  + Monitor your quiz performance


---
# Installing R and RStudio

--
**Please install a version of R and RStudio on your own computer**

+ You will need both R and RStudio installed before your first Lab in Week 1

+ Instructions for installation and updating available at <a href="https://edin.ac/3B0oi5A" target="_blank">https://edin.ac/3B0oi5A</a>
+ Please follow these instructions carefully


--

+ For those of you who have Chromebooks:
  + Local installation is not possible so you should (continue to) use the PPLS RStudio Server at <a href="https://rstudio.ppls.ed.ac.uk/" target="_blank">https://rstudio.ppls.ed.ac.uk/</a>
  + If you need to request access, fill out the form here: <a href="https://edin.ac/3Le1mEW" target="_blank">https://edin.ac/3Le1mEW</a>

---
class: center, middle
# See you in class!
