---
title: "Numeric/numeric interactions"
author: "Dr. Elizabeth Pankratz (elizabeth.pankratz@ed.ac.uk)"
editor_options: 
  chunk_output_type: inline
format:
  revealjs:
    smaller: true
---



```{r}
#| label: setup
#| include: false

library(simglm)
library(tidyverse)
library(interactions)  # for probe_interactions() plot

dapr2red <- "#BF1932" 

source('../_theme/theme_quarto.R')

palette_probe <- c('#4ab8fc', '#ff7b01')
```


```{r include=F}
# simulate the data
set.seed(3119) 

sim_arguments <- list(
  formula = wellbeing ~ 1 + outdoor_time + mm_rain + outdoor_time:mm_rain,
  fixed = list(
    outdoor_time = list(var_type = 'continuous', mean = 5, sd = 2),
    mm_rain = list(var_type = 'continuous', mean = 100, sd = 20)
  ),
  error = list(variance = 20),
  sample_size = 60,
  reg_weights = c(60, 7, -0.01, -0.05)
)

outdoors_rain <- simulate_fixed(data = NULL, sim_arguments) |>
  simulate_error(sim_arguments) |>
  generate_response(sim_arguments) |>
  select(mm_rain, outdoor_time, wellbeing) |>
  mutate(
    mm_rain_bin = cut(
      mm_rain, 
      breaks = c( min(mm_rain), mean(mm_rain) - sd(mm_rain), mean(mm_rain), mean(mm_rain) + sd(mm_rain), max(mm_rain) ),
      include.lowest = TRUE,
      labels = c( 'mm_rain,\n from min \nto mean – 1 SD', 'mm_rain,\n from mean – 1 SD \nto mean', 'mm_rain,\n from mean \nto mean + 1 SD', 'mm_rain,\n from mean + 1 SD \nto max' )
    ),
    outdoor_time_bin = cut(
      outdoor_time, 
      breaks = c( min(outdoor_time), mean(outdoor_time) - sd(outdoor_time), mean(outdoor_time), mean(outdoor_time) + sd(outdoor_time), max(outdoor_time) ),
      include.lowest = TRUE,
      labels = c( 'outdoor_time,\n from min \nto mean – 1 SD', 'outdoor_time,\n from mean – 1 SD \nto mean', 'outdoor_time,\n from mean \nto mean + 1 SD', 'outdoor_time,\n from mean + 1 SD \nto max' )
    )
  )
outdoors_rain$wellbeing <- outdoors_rain$wellbeing[,1]

outdoors_rain |>
  ggplot(aes(x = mm_rain, y = wellbeing)) +
  geom_point() +
  facet_wrap(~ outdoor_time_bin, nrow = 1)+
  geom_smooth(se = F, method = 'lm', formula = 'y~x')

outdoors_rain |>
  ggplot(aes(x = outdoor_time, y = wellbeing)) +
  geom_point() +
  facet_wrap(~ mm_rain_bin, nrow = 1) +
  geom_smooth(se = F, method = 'lm', formula = 'y~x')

outdoors_rain |>
  ggplot(aes(x = outdoor_time, y = wellbeing, colour = mm_rain)) +
  geom_point(size = 3)

lm(wellbeing ~ outdoor_time*mm_rain, data = outdoors_rain) |> summary()
```


# Course overview {background-color="white" style="font-size:80%;"}

<br>

:::: {.columns}
::: {.column width="50%"}

```{r echo = FALSE, results='asis', warning = FALSE}
source('../dapr2_course_outline.R')
source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")
course_table(block1_name, block2_name, block1_lecs, block2_lecs, week = 0)
```

:::

::: {.column width="50%"}

```{r echo = FALSE, results='asis', warning = FALSE}
course_table(block3_name, block4_name, block3_lecs, block4_lecs, week = 2)
```

:::
::::







## This week's learning objectives

<br>

::::: {style="font-size: 125%;"}

::: {.dapr2callout .fragment}
How do we specify an interaction between two numeric predictors?
:::

::: {.dapr2callout .fragment}
What does it mean when we say "interactions are symmetrical"?
:::

::: {.dapr2callout .fragment}
What is a simple slope?
:::

::: {.dapr2callout .fragment}
When we have an interaction between two numeric predictors, what values do we typically use to compute simple slopes?
:::

:::::


## Where we are in the analysis plan today

![](figs/workflow-11.svg){fig-align="center"}


## What's an interaction again?

An interaction is how we allow a model to estimate that **the association between one predictor and the outcome is different, _depending on_ the value of another predictor**.

<br>

![](figs/depends-on-11.svg){fig-align="center"}


## Today's data


<br>

:::: {.columns}
::: {.column width="60%"}
```{r fig.align = 'center', fig.width = 7, fig.height = 5}
#| code-fold: true
outdoors_rain |>
  ggplot(aes(x = outdoor_time, y = wellbeing, colour = mm_rain)) +
  geom_point(size = 5)
```
:::

::: {.column width="40%"}

<!-- ::::{ style="font-size:85%;"} -->
```{r eval=F}
outdoors_rain |>
  head(12)
```

```{r echo=F}
outdoors_rain |>
  select(-ends_with('bin')) |>
  head(12)
```

<!-- :::: -->

:::

::::

How does the association between **outdoor time** and people's **wellbeing** change, _depending on_ the **amount of rainfall** in a day?


## A plot to make interaction more obvious

```{r fig.align = 'center', fig.width = 12, fig.height = 6, warning = F}
#| code-fold: true
p_rain_panel <- outdoors_rain |>
  ggplot(aes(x = outdoor_time, y = wellbeing)) +
  geom_point(size = 5) +
  facet_wrap(~ mm_rain_bin, nrow = 1) +
  geom_smooth(se = F, method = 'lm', formula = 'y~x', linewidth = 2) +
  theme(strip.text = element_text(size = 20))
p_rain_panel
```


<br>


:::{style="font-size:125%;"}
Different slopes in different panels $\rightarrow$ The **association** between outdoor time and wellbeing appears **different for different amounts of rainfall.**
:::


# Specifying the interaction model

## Retrieval practice: Specifying the interaction model

<br>

:::{.hcenter style="font-size: 125%;"}
`wooclap.com`, enter code `BIAHTI`
:::


## Specifying the interaction model

<br>

**In mathematical notation:**

$$
\text{wellbeing} = \beta_0 + (\beta_1 \cdot \text{outdoor_time}) +
(\beta_2 \cdot \text{mm_rain}) +
(\beta_3 \cdot \text{outdoor_time} \cdot \text{mm_rain}) +
\epsilon
$$

<br>

**In R,** where both of these options have the same result:

<br>

```{r eval=F}
lm(wellbeing ~ outdoor_time + mm_rain + outdoor_time:mm_rain, data = outdoors_rain)

lm(wellbeing ~ outdoor_time * mm_rain, data = outdoors_rain)
```

<br>

:::{.dapr2callout .fragment}
**Does an interaction model need to contain each of the interacting predictors on their own, too?**

Yes—otherwise we'd only see how one predictor's effect depends on another predictor, with no idea how they behave individually.

- Good: `Y ~ A + B + A:B`
- Good: `Y ~ A * B`
- Not good: `Y ~ A:B`

<!-- Including each interacting predictor as well as the interaction (the "Better" example above) is an example of the **"principle of marginality"** ... yet another terrible name for a stats concept :(  -->
:::

## Think–Pair–Share: Num/num interaction coefs

<br>

The interaction model &nbsp; `wellbeing ~ outdoor_time * mm_rain` &nbsp; will have four coefficients:

- `Intercept`
- `outdoor_time`
- `mm_rain`
- `outdoor_time:mm_rain`

**Without looking ahead, try to figure out: What will each coefficient represent?**


<br>

:::{.dapr2callout style="font-size:100%"}

- **Think** to yourself about each coefficient's meaning. / **Pair** up with your neighbour and think together.

- Then **share** on `wooclap.com`, enter code `BIAHTI`

:::


<!-- ## Think ahead, using our key points about multiple regression/interactions (1) -->

<!-- <br> -->

<!-- :::{.hcenter style="font-size: 125%;"} -->
<!-- `wooclap.com`, enter code `BIAHTI` -->
<!-- ::: -->

<!-- <br> -->

<!-- :::{.dapr2callout} -->
<!-- **For every predictor that goes into your model, know what zero represents.** -->

<!-- - What does `outdoor_time = 0` represent? -->
<!-- - What does `mm_rain = 0` represent? -->
<!-- ::: -->


<!-- :::{.dapr2callout} -->

<!-- A model's intercept is the estimated mean outcome **when every predictor in the model is at zero.** -->

<!-- - What will the intercept represent for `wellbeing ~ outdoor_time * mm_rain`? -->
<!-- ::: -->



<!-- ## Think ahead, using our key points about multiple regression/interactions (2) -->

<!-- <br> -->

<!-- :::{.hcenter style="font-size: 125%;"} -->
<!-- `wooclap.com`, enter code `BIAHTI` -->
<!-- ::: -->

<!-- <br> -->

<!-- :::{.dapr2callout} -->

<!-- Each predictor's slope represents the association of the given predictor with the outcome **when every other predictor is at zero.**  -->

<!-- If predictors interact, then slope estimates are **conditional** on the other predictor being zero. (In other words, each predictor's slope estimate **is _only_ true when the other interacting predictor is zero.**) -->

<!-- - What will the slope of `outdoor_time` represent? -->
<!-- - What will the slope of `mm_rain` represent? -->

<!-- ::: -->



<!-- :::{.dapr2callout} -->

<!-- The interaction term tells us **how much the association between _one_ predictor and the outcome changes, when the _other_ predictor moves from 0 to 1.** -->

<!-- - What will the interaction term `outdoor_time:mm_rain` represent? -->

<!-- ::: -->



## Fitting the model

```{r}
m1 <- lm(wellbeing ~ outdoor_time * mm_rain, data = outdoors_rain)
summary(m1)
```



## Interpreting the model (1)

:::: {.columns}
::: {.column width="60%"}
```{r echo=F, fig.align='center', fig.width=10, fig.height=6}
p_rain_panel
```

::::{ style="font-size:80%;"}
```{r echo=F}
cat(paste0(capture.output(summary(m1)), '\n')[10:14])
```
::::
:::
::: {.column width="40%"}

:::{.fragment}

**`Intercept`**

- When `outdoor_time = 0` and `mm_rain = 0`, the estimated average `wellbeing` is 67.83 points.

:::

:::{.fragment}

**`outdoor_time`**

- Specifically when `mm_rain = 0`, spending an additional hour outdoors is associated with an increase in wellbeing of 5.33 points.
- This estimate is significantly different from zero ($p$ = .002).

:::

:::{.fragment}

**`mm_rain`**

- Specifically when `outdoor_time = 0`, an additional mm of rain is associated with a decrease in wellbeing of 0.08 points.
- This estimate is not significantly different from zero ($p$ = 0.363).

:::

:::

::::


## Interpreting the model (2)

:::: {.columns}
::: {.column width="60%"}
```{r echo=F, fig.align='center', fig.width=10, fig.height=6}
p_rain_panel
```

::::{ style="font-size:80%;"}
```{r echo=F}
cat(paste0(capture.output(summary(m1)), '\n')[10:14])
```
::::
:::
::: {.column width="40%"}

**`outdoor_time:mm_rain`**

- When `mm_rain` increases by 1, the association of `outdoor_time` with `wellbeing` decreases by 0.03 points.

- So, the association of `outdoor_time` with `wellbeing` when ...
  - `mm_rain = 0` is 5.33
  - `mm_rain = 1` is 5.33 – 0.03 = 5.30
  - `mm_rain = 2` is 5.33 – 0.03 – 0.03 = 5.27 
  - ...
  
**The association was positive to begin with (5.33), so the negative interaction term <br> (–0.33) brings the slope ever closer to zero as `mm_rain` increases.**

:::
::::


## But wait! There's another way to interpret the interaction term

```{r fig.align = 'center', fig.width = 13, fig.height = 5, warning = F}
#| code-fold: true
p_out_panel <- outdoors_rain |>
  ggplot(aes(x = mm_rain, y = wellbeing)) +
  geom_point(size = 5) +
  facet_wrap(~ outdoor_time_bin, nrow = 1)+
  geom_smooth(se = F, method = 'lm', formula = 'y~x', linewidth = 2) +
  theme(
    axis.text.x = element_text(size = 22),
    strip.text = element_text(size = 20)
  )

p_out_panel
```

<br>

Now the x axis shows mm of rainfall, and each panel shows a group of similar `outdoor_time`s.

<!-- We've flipped the interacting variables, but the pattern of the interaction is still there. -->

This plot shows that **the association between sunlight hours and wellbeing is different, when different amounts of time are spent outdoors.** The interacting variables are flipped, but the interaction is still there.


## Both ways of interpreting the interaction term

**Interpretation 1:**

:::: {.columns}
::: {.column width="70%"}
```{r echo=F, fig.dim = c(10, 4)}
p_rain_panel
```
:::

::: {.column width="30%"}

<br>

Increasing `mm_rain` by 1 means that the association between `outdoor_time` and `wellbeing` decreases by 0.03 points.
:::
::::


**Interpretation 2 flips the two interacting variables and is still true:**

:::: {.columns}
::: {.column width="70%"}
```{r echo=F, fig.dim = c(10, 4)}
p_out_panel
```
:::

::: {.column width="30%"}

<br>

Increasing `outdoor_time` by 1 means that the association between `mm_rain` and `wellbeing` decreases by 0.03 points.
:::
::::



## Interactions are symmetrical

A common example of symmetry: **When a shape looks the same, even when it is mirrored.**

![Image from Wikimedia Commons](figs/Asymmetric_(PSF).png){fig-align="center"}


<br>

**If we have an interaction between predictors A and B, then both "mirror images" are always true:**

- The association between **A** and the outcome is different for different values of **B**.
- The association between **B** and the outcome is different for different values of **A**.

**So we say that interactions are symmetrical.**

<br>

In practice, usually we choose the interpretation that's narratively simpler or makes more intuitive sense (like I secretly did last week!)


## Back to interpreting the interaction term

:::: {.columns}
::: {.column width="60%"}
```{r echo=F, fig.align='center', fig.width=10, fig.height=6}
p_rain_panel
```

::::{ style="font-size:80%;"}
```{r echo=F}
cat(paste0(capture.output(summary(m1)), '\n')[10:14])
```
::::
:::
::: {.column width="40%"}

**`outdoor_time:mm_rain`**

- When `mm_rain` increases by 1, the association of `outdoor_time` with `wellbeing` decreases by 0.03 points.

- So, the association of `outdoor_time` with `wellbeing` when ...
  - `mm_rain=0` is 5.33
  - `mm_rain=1` is 5.33 – 0.03 = 5.30
  - `mm_rain=2` is 5.33 – 0.03 – 0.03 = 5.27 
  - ...
  
:::{.fragment}

- **Equivalently:** When `outdoor_time` increases by 1, the association of `mm_rain` with `wellbeing` decreases by 0.03 points.

- So, the association of `mm_rain` with `wellbeing` when ...
  - `out_t=0` is –0.08
  - `out_t=1` is –0.08 – 0.03 = –0.11
  - `out_t=2` is –0.08 – 0.03 – 0.03 = –0.14
  - ...
  
:::

:::
::::


# Simple slopes

## Simple slopes

**A simple slope is the slope of the association between one predictor and the outcome, at a specific value of another predictor.**

I've been showing you simple slopes when interpreting interaction coefficients. 
For example:

- the association of `outdoor_time` with `wellbeing` when ...
  - `mm_rain=0` is 5.33
  - `mm_rain=1` is 5.33 – 0.03 = 5.30
  - `mm_rain=2` is 5.33 – 0.03 – 0.03 = 5.27 
  - ...
  
^ each of those is a simple slope!

<br>

But now let's see how to calculate simple slopes by hand for *any* predictor value.

What we need to know:

- The model's linear expression
- The values of each $\beta$ coefficient


## Calculating simple slopes by hand (Example 1)

<br>

**The model's linear expression** (where w stands for wellbeing):
 
$$
\text{w} = \beta_0 + (\beta_1 \cdot \text{outdoor_time}) +
(\beta_2 \cdot \text{mm_rain}) +
(\beta_3 \cdot \text{outdoor_time} \cdot \text{mm_rain}) +
\epsilon
$$ 

**The model's coefficients:**

```{r echo=F}
m1_coef_tab <- coef(m1) |> 
  round(2) |> 
  as.data.frame() |> 
  cbind(paste0('beta', 0:3)) |>
  rownames_to_column()
names(m1_coef_tab) <- c('parameter', 'estim', 'beta')
select(m1_coef_tab, parameter, beta, estim) |>
  kableExtra::kable()
```



<br>

:::{.dapr2callout}

**Step 1:** Substitute the estimated coefficients into the linear expression.

$$
\text{w} = 67.83 + (5.33 \cdot \text{outdoor_time}) +
(-0.08 \cdot \text{mm_rain}) +
(-0.03 \cdot \text{outdoor_time} \cdot \text{mm_rain}) +
\epsilon
$$ 

:::

## Calculating simple slopes by hand (Example 1)

<br>


:::{.dapr2callout}

**Step 1:** Substitute the estimated coefficients into the linear expression.

$$
\text{w} = 67.83 + (5.33 \cdot \text{outdoor_time}) +
(-0.08 \cdot \text{mm_rain}) +
(-0.03 \cdot \text{outdoor_time} \cdot \text{mm_rain}) +
\epsilon
$$ 

:::

<br>

:::{.dapr2callout .fragment}


**Step 2:** Decide what specific value of `mm_rain` we want to find the slope over `outdoor_time` for.
Substitute that value in for `mm_rain`.

**Let's start with `mm_rain = 0`.**

$$
\text{w}_{\text{mm_rain} = 0} = 67.83 + (5.33 \cdot \text{outdoor_time}) + (-0.08 \cdot 0) + (-0.03 \cdot \text{outdoor_time} \cdot 0) + \epsilon
$$ 
:::

<br>

:::{.dapr2callout .fragment}

**Step 3:** Simplify this expression by working out the multiplication and addition step by step.

$$
\begin{align}
\text{w}_{\text{mm_rain} = 0}  &= 67.83 + (5.33 \cdot \text{outdoor_time}) + (-0.08 \cdot 0) + (-0.03 \cdot \text{outdoor_time} \cdot 0) + \epsilon \\
\text{w}_{\text{mm_rain} = 0} &= 67.83 + (5.33 \cdot \text{outdoor_time}) + \epsilon \\
\end{align}
$$ 

So the line that associates `outdoor_time` with `wellbeing`, when `mm_rain = 0`, has an intercept of 67.83 and a slope of 5.33. **5.33 is the simple slope.**

:::


## Calculating simple slopes by hand (Example 2)

<br>

:::{.dapr2callout}

**Step 1:** Substitute the estimated coefficients into the linear expression.

$$
\text{w} = 67.83 + (5.33 \cdot \text{outdoor_time}) +
(-0.08 \cdot \text{mm_rain}) +
(-0.03 \cdot \text{outdoor_time} \cdot \text{mm_rain}) +
\epsilon
$$ 

:::

<br>

:::{.dapr2callout .fragment}


**Step 2:** Decide what specific value of `mm_rain` we want to find the slope over `outdoor_time` for.
Substitute that value in for `mm_rain`.

**Let's do `mm_rain` = 100.**

:::{style="font-size:95%;"}

$$
\text{w}_{\text{mm_rain} = 100} = 67.83 + (5.33 \cdot \text{outdoor_time}) + (-0.08 \cdot 100) + (-0.03 \cdot \text{outdoor_time} \cdot 100) + \epsilon
$$ 
:::
:::

<br>

:::{.dapr2callout .fragment}

**Step 3:** Simplify this expression by working out the multiplication and addition step by step.

Breakdown of steps on next slide.

:::

## Step 3 breakdown

<br>

**The starting expression that we want to simplify:**

$$
\text{w}_{\text{mm_rain} = 100}  = 67.83 + (5.33 \cdot \text{outdoor_time}) + (-0.08 \cdot 100) + (-0.03 \cdot \text{outdoor_time} \cdot 100) + \epsilon
$$

:::{.fragment}

**Multiply the 100s together with the numbers in the same brackets.**

$$
\text{w}_{\text{mm_rain} = 100}  = 67.83 + (5.33 \cdot \text{outdoor_time}) + (-8.00) + (-3.00 \cdot \text{outdoor_time}) + \epsilon
$$

:::

:::{.fragment}

**Add together the freestanding numbers (the numbers not being multiplied with `outdoor_time`).**

$$
\begin{align}
\text{w}_{\text{mm_rain} = 100}  &= 67.83 - 8.00 + (5.33 \cdot \text{outdoor_time}) + (-3.00 \cdot \text{outdoor_time}) + \epsilon \\
 &= 59.83 + (5.33 \cdot \text{outdoor_time}) + (-3.00 \cdot \text{outdoor_time}) + \epsilon \\
\end{align}
$$

:::

:::{.fragment}

**Add together the numbers that are multiplied with $\text{outdoor_time}$. (In math terms, we're factoring out the variable $\text{outdoor_time}$ from both terms that contain it, then adding the constants together.)**

$$
\begin{align}
\text{w}_{\text{mm_rain} = 100}  &= 59.83 + ((5.33 - 3.00) \cdot \text{outdoor_time}) + \epsilon
 \\
  &= 59.83 + (2.33 \cdot \text{outdoor_time}) + \epsilon \\
\end{align}
$$

:::{.dapr2callout}
So the line that associates `outdoor_time` with `wellbeing`, when `mm_rain` is set to 100, has an intercept of 59.83 and a slope of 2.33. **2.33 is the simple slope.**
:::

:::



## Plotting simple slopes in R

```{r eval=F}
library(interactions)

probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
  interval = T
)$interactplot
```

```{r echo=F, fig.dim = c(10, 8), fig.align='center'}

probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
  interval = T
)$interactplot +
  theme(text = element_text(size = 28))

```



## Computing simple slopes in R 

`probe_interaction()` can compute simple slopes _and_ test if they're significantly different from zero.

<br>

:::: {.columns}
::: {.column width="40%"}
```{r eval=F}
probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
)$simslopes
```
:::
::: {.column width="60%" style = "font-size:80%"}
```{r echo=F}
m1_simslopes <- probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
)$simslopes

m1_simslopes
```
:::
::::


## The Johnson-Neyman interval

The JN interval tells us **the values for one predictor when the simple slope of the other predictor is significantly different from zero.**
(This is also called a **region of significance analysis.**)

```{r echo=F}
cat(paste0(capture.output(m1_simslopes), '\n')[1:7])
```

<!-- <br> -->

**To plot the JN interval:**

:::: {.columns}

::: {.column width="50%"}

```{r echo=F, fig.dim = c(8, 6), fig.align = 'center'}
m1_jnp <- probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
  jnplot = T
)$simslopes$jnplot
m1_jnp +
  theme(text = element_text(size = 18))
```


:::

::: {.column width="50%"}

```{r eval=F}
probe_interaction(
  m1,
  pred = outdoor_time,
  modx = mm_rain,
  jnplot = T
)$simslopes$jnplot
```

- x: `mm_rain`
- y: Slope of `outdoor_time`
- Red area: Slope of `outdoor_time` not significantly diff. from zero
- Blue area: Slope of `outdoor_time` **is** significantly diff. from zero

:::

::::


# Numeric/numeric interactions, Round 2: With mean-centering

## Retrieval practice: Mean-centering

<br>

:::{.hcenter style="font-size: 125%;"}
`wooclap.com`, enter code `BIAHTI`
:::



## What's mean-centering again?

To mean-centre a variable means to **transform it so that the mean of the centered version is zero.**

Specifically, this involves **subtracting the mean** of a variable from every observation of that variable.


```{r eval=F}
outdoors_rain <- outdoors_rain |>
  mutate(
    outdoor_time_c = outdoor_time - mean(outdoor_time),
    mm_rain_c      = mm_rain - mean(mm_rain),
  )

head(outdoors_rain)
```


```{r echo=F}
outdoors_rain <- outdoors_rain |>
  select(-ends_with('bin')) |>
  mutate(
    outdoor_time_c = outdoor_time - mean(outdoor_time),
    mm_rain_c      = mm_rain - mean(mm_rain),
  )

head(outdoors_rain)
```


<br>

:::: {.columns}
::: {.column width="50%"}
```{r}
mean(outdoors_rain$outdoor_time) |> round(2)
mean(outdoors_rain$mm_rain) |> round(2)
```
:::
::: {.column width="50%"}
```{r}
mean(outdoors_rain$outdoor_time_c) |> round(2)
mean(outdoors_rain$mm_rain_c) |> round(2)
```
:::
::::




## Before and after mean-centering

**Before**:

```{r echo=F, fig.align = 'center', fig.width=12, fig.height = 4, warning = F}
p_rain_panel
```


**After**:

```{r echo=F, fig.align = 'center', fig.width=12, fig.height = 4, warning = F}
p_rain_panel_c <- outdoors_rain |>
  mutate(
    mm_rain_c_bin = cut(
      mm_rain_c, 
      breaks = c( min(mm_rain_c), mean(mm_rain_c) - sd(mm_rain_c), mean(mm_rain_c), mean(mm_rain_c) + sd(mm_rain_c), max(mm_rain_c) ),
      include.lowest = TRUE,
      labels = c( 'mm_rain_c,\n from min \nto mean – 1 SD', 'mm_rain_c,\n from mean – 1 SD \nto mean', 'mm_rain_c,\n from mean \nto mean + 1 SD', 'mm_rain_c,\n from mean + 1 SD \nto max' )
    )
  ) |>
  ggplot(aes(x = outdoor_time_c, y = wellbeing)) +
  geom_point(size = 5)+
  facet_wrap(~ mm_rain_c_bin, nrow = 1) +
  geom_smooth(se = F, method = 'lm', linewidth=2) +
  theme(strip.text = element_text(size = 20))
p_rain_panel_c
```



## Think–Pair–Share: Mean-centered interactions

<br>

The interaction model &nbsp; `wellbeing ~ outdoor_time_c * mm_rain_c` &nbsp; will have four coefficients:

- `Intercept`
- `outdoor_time_c`
- `mm_rain_c`
- `outdoor_time_c:mm_rain_c`

**What will each coefficient represent?**


<br>

:::{.dapr2callout style="font-size:100%"}

- **Think** to yourself about each coefficient's meaning. / **Pair** up with your neighbour and think together.

- Then **share** on `wooclap.com`, enter code `BIAHTI`

:::


## Fit the model

$$
\text{wellbeing} = \beta_0 + (\beta_1 \cdot \text{outdoor_time_c}) +
(\beta_2 \cdot \text{mm_rain_c}) +
(\beta_3 \cdot \text{outdoor_time_c} \cdot \text{mm_rain_c}) +
\epsilon
$$

```{r}
m2 <- lm(wellbeing ~ outdoor_time_c * mm_rain_c, data = outdoors_rain)
```

:::{.fragment}
```{r}
summary(m2)
```
:::


## Interpreting the mean-centered model (1)

:::: {.columns}
::: {.column width="60%"}
```{r echo=F, fig.align='center', fig.dim=c(12, 6)}
p_rain_panel_c
```

::::{style="font-size:80%;"}
```{r echo=F}
# cat(paste0(capture.output(summary(m2)), '\n')[10:14])
summary(m2)$coefficients |> round(2)
```
::::

:::

::: {.column width="40%"}

:::{.fragment}

**`Intercept`**

- When `outdoor_time = 0` (its mean) and `mm_rain = 0` (its mean), the estimated average `wellbeing` is 70.39.

:::

:::{.fragment}

**`outdoor_time_c`**

- Specifically when `mm_rain = 0` (at the mean rainfall), spending an additional hour outdoors is associated with an increase in wellbeing of 2.00 points ($p$ < .001).

:::

:::{.fragment}

**`mm_rain_c`**

- Specifically when `outdoor_time = 0` (at the mean outdoor time), an additional mm of rain is associated with a decrease in wellbeing of 0.26 points ($p$ < .001).

:::

:::
::::

## Interpreting the mean-centered model (2)

:::: {.columns}
::: {.column width="60%"}
```{r echo=F, fig.align='center', fig.dim=c(12, 6)}
p_rain_panel_c
```

::::{style="font-size:80%;"}
```{r echo=F}
# cat(paste0(capture.output(summary(m2)), '\n')[10:14])
summary(m2)$coefficients |> round(2)
```
::::


:::

::: {.column width="40%"}

<!-- ::::{style="font-size:90%"} -->

**`outdoor_time_c:mm_rain_c`**

- Increasing centered rainfall by 1 mm changes the association of `outdoor_time_c` with `wellbeing` by 0.03 points.

- So, the association of `outdoor_time_c` with `wellbeing` when ...
  - `mm_r_c=0` is 2
  - `mm_r_c=1` is 2 – 0.03 = 1.97
  - `mm_r_c=2` is 2 – 0.03 – 0.03 = 1.94
  - ...
  
:::{.fragment}

- **Equivalently:** Increasing centered outdoor time by 1 changes the association of `mm_rain_c` with `wellbeing` by 3.3.

- So, the association of `mm_rain_c` with `wellbeing` when ...
  - `o_t_c=0` is –0.26
  - `o_t_c=1` is –0.26 – 0.03 = –0.23
  - `o_t_c=2` is –0.26 – 0.03 – 0.03 = –0.20
  - ...

:::

<!-- :::: -->

:::
::::




# Probing the interaction

## Simple slopes

```{r fig.align = "center"}
probe_interaction(
  m2,
  pred = outdoor_time_c,
  modx = mm_rain_c,
  interval = T,
)$interactplot
```

Compare to the old plot of simple slopes: the shape is the same!
Mean-centering doesn't change the relationship between simple slopes.


## Region of significance analysis

```{r fig.align="center"}
probe_interaction(
  m2,
  pred = outdoor_time_c,
  modx = mm_rain_c,
  jnplot = T
)$simslopes$jnplot
```

Compare to the old plot: the shape is the same!
Mean-centering doesn't change the regions of (non)significance in how the predictors interact.



# The big picture: Interactions, now with numeric predictors

## Interactions, now with numeric predictors

<br>

![](figs/depends-on-12.svg){fig-align="center"}




# Back matter


## Revisiting this week's learning objectives

<br>

::: {.dapr2callout}
**How do we specify an interaction between two numeric predictors?**

- Exactly the same as we would for two categorical predictors!
- Mathematically: Interactions are always the product of the two interacting predictors.
- In R: If we have two interacting predictors A and B, we can write `A*B` (which stands for the individual predictors _and_ their interaction) or `A:B` (which stands for just the interaction).
:::

::: {.dapr2callout}
**What does it mean when we say "interactions are symmetrical"?**

- Imagine we have two interacting predictors, A and B.
- We can see the symmetry of this interaction because both of these "mirrored" interpretations are true:
	- The association between A and the outcome is different for different values of B.
	- The association between B and the outcome is different for different values of A.
:::

## Revisiting this week's learning objectives

<br>


::: {.dapr2callout}
**What is a simple slope?**

- The slope of the association between one predictor and the outcome, at a specific value of another predictor.
- If we have two interacting predictors, A and B, then we could look at (for example) 
  - the simple slope of A, when B = 0,
  - or the simple slope of B, when A = 100,
  - or the simple slope of A, when B = –5,
  - or ...
:::

::: {.dapr2callout}
**When we have an interaction between two numeric predictors, what values do we typically use to compute simple slopes?**

- We typically look at the simple slope of one predictor at three specific values of the other one: its mean, its mean + 1 SD, its mean – 1 SD.
- If we have two interacting predictors, A and B, then we might look at the simple slope of A
  - when B is at its mean minus one SD (i.e., `mean(B) - sd(B)`),
  - when B is at its mean (i.e., `mean(B)`), and
  - when B is at its mean plus one SD (i.e., `mean(B) + sd(B)`).
:::




## This week 

<br>

::::{.columns}
:::{.column width="50%"}
**Tasks:**

<br>

![](../figs/labs.svg){width=80px style="margin:10px;margin-bottom:-50px"} Work on exercises in labs

<br>

![](../figs/exam.svg){width=80px style="margin:10px;margin-bottom:-45px"} Complete the weekly quiz 


:::

:::{.column width="50%"}
**Get support:**

<br>

![](../figs/flashcards.svg){width=80px style="margin:10px;margin-bottom:-30px"}
Consult the [flash cards](https://uoepsy.github.io/dapr2/2627/flashcards/){target="_blank"}

<br>

![](../figs/forum.svg){width=80px style="margin:10px;margin-bottom:-50px"}
Ask questions anonymously on Piazza

<br>

![](../figs/oh.svg){width=80px style="margin:10px;margin-bottom:-40px"} 
We really like seeing you in office hours!

:::
::::


# Appendix {.appendix} 



## Compare the coefficients

<!-- - Why is the intercept larger in the mean-centered model than in the non-centered model? -->
<!-- - Why is outdoor_time_c so much closer to zero in the mean-centered model? -->
<!-- - Why is mm_rain_c so much larger in the mean-centered model? -->
<!-- - Why are the interaction terms the same in both models? -->

<!-- <br> -->

**The old coefficients in the non-centered model, `m1`:**

::::{.columns}
::: {.column width="30%"}

```{r echo=F}
round(coef(m1)[1], 1)
round(coef(m1)[2], 1)
round(coef(m1)[3], 1)
round(coef(m1)[4], 1)
```

:::

::: {.column width="70%"}


```{r echo=F, fig.align = 'center', fig.dim = c(12, 4), warning = F}
p_rain_panel
```

:::

::::



**The new coefficients in the mean-centered model, `m2`:**

::::{.columns}

::: {.column width="30%"}

```{r echo=F}
round(coef(m2)[1], 1)
round(coef(m2)[2], 1)
round(coef(m2)[3], 1)
round(coef(m2)[4], 1)
```

:::

::: {.column width="70%"}

```{r echo=F, fig.align = 'center', fig.dim = c(12, 4), warning = F}
p_rain_panel_c
```

:::

::::



<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- a -->
<!-- ::: -->
<!-- ::: {.column width="50%"} -->
<!-- b -->
<!-- ::: -->
<!-- :::: -->
