---
title: "10 Practice analysis: Code template"
output: 
  html_notebook:
    toc: true
---

Follows workflow here: dapr2/2627/dapr2_flashcards/resources/workflow-b2.html

# Phase 1: Before model fitting

## 1a: Set up code and data

- **Load the required R packages.** You'll probably need at least:
  - `tidyverse` (for managing and wrangling data)
  - `psych` (for descriptives)
  - `sjPlot` (for nice regression tables and plots)
  - `kableExtra` (for nicely-formatted tables)
  - `patchwork` (for combining multiple plots into one graphic)
  
```{r}

```
  
- **Read in your data.**

```{r}

```

- **Tidy data** (e.g., any missingness, any implausible values? are data types set correctly?).

```{r}

```


## 1b: Set up the variables you'll model

- **Based on the RQ, decide whether you'll test your hypotheses using coefficient significance tests or model comparison.**


- **Based on the RQ, identify your outcome variable and the predictors.**



- **Set up categorical predictors.** (e.g., factor levels? which contrast coding scheme?)

```{r}

```


- **Set up continuous predictors.** (e.g., any transformations?)

NA


- **Explore patterns in the data by plotting outcome and predictor variables together.** (It's most reader-friendly to plot the transformed version of the data, the version that you'll also be modelling.)

```{r}

```


- **Make a nice descriptives table of all relevant variables for the write-up.**

```{r}

```



# Phase 2: Model fitting

- **Write the mathematical model formulation for your model(s).**



- **Explicitly define the hypotheses that your RQ is aiming to test.**



- **Fit your model(s) using `lm()`.**



# Phase 3: After model fitting

## 3a: Check assumptions and diagnostics

- **Check model assumptions.**

```{r}

```


- **If model assumptions about normality/equal variance of errors are violated, bootstrap model estimates.**

```{r}

```


- **Run diagnostics for multicollinearity.**

```{r}

```


- **Run diagnostics for influential observations.**

```{r}

```


- **If you find extreme influential observations: run sensitivity analysis.**
   
```{r}

```
   
   
## 3b: Plot and interpret model estimates
   
- **Interpret the coefficient estimates** (if appropriate for your RQ).

```{r}

```


- **Generate a nicely-formatted regression table** (if appropriate for your RQ).

```{r}

```


- **Compare models and interpret model comparison statistics** (if appropriate for your RQ).

```{r}

```


- **Plot model-fitted values** (if appropriate for your RQ).

```{r}

```

  
  
## 3c: Write up

In labs!