---
title: "10 Practice analysis: Code EE"
output: 
  html_notebook:
    toc: true
---

Follows workflow here: dapr2/2627/dapr2_flashcards/resources/workflow-b2.html

# Phase 1: Before model fitting

## 1a: Set up code and data

- **Load the required R packages.** You'll probably need at least:
  - `tidyverse` (for managing and wrangling data)
  - `psych` (for descriptives)
  - `sjPlot` (for nice regression tables and plots)
  - `kableExtra` (for nicely-formatted tables)
  - `patchwork` (for combining multiple plots into one graphic)
  - `performance` (for assumption checks)
  - `car` (for multicollinearity check)
  
```{r}
library(tidyverse)
library(psych)
library(sjPlot)
library(kableExtra)
library(patchwork)
library(performance)
library(car)
```
  
- **Read in your data.**

```{r}
attend_df <- read_csv("https://uoepsy.github.io/data/DapR2_S1B2_PracticalPart1.csv")

# rename Conscientiousness because it's a lot to type
attend_df <- attend_df |>
  rename(Consc = Conscientiousness)
```

- **Tidy data** (e.g., any missingness, any implausible values? are data types set correctly?).

```{r}
summary(attend_df)  # no NAs, no wacky values
```


```{r}
glimpse(attend_df)  # lets us see example values for categorical preds
```


## 1b: Set up the variables you'll model

- **Based on the RQ, decide whether you'll test your hypotheses using coefficient significance tests or model comparison.**

RQ: Do people with greater conscientiousness and more advanced years of study attend university courses more?

This question is about associations between predictors and outcomes, so we can test it using coef significance tests.


- **Based on the RQ, identify your outcome variable and the predictors.**

- Outcome: `Attendance`
- Predictors: `Consc` and `Year`


- **Set up categorical predictors.** (e.g., factor levels? which contrast coding scheme?)

First, `Consc` has three levels: `Low`, `Moderate`, and `High`.
This is not their alphabetical order, so let's specify this order using `mutate()`.

```{r}
attend_df <- attend_df |>
  mutate(
    Consc = factor(Consc, levels = c('Low', 'Moderate', 'High'))
  )
```

We are interested in comparing people with higher conscientiousness to people with lower conscientiousness, so treatment coding is fine.
How will `Consc` be coded?

```{r}
contrasts(attend_df$Consc)
```

Reference level = `Low`, and two dummy variables which compare `Low` to `Moderate` and `Low` to `High`.



Next, `Year` has many levels: Y1, Y2, Y3, Y4, MSc, and PhD.
This is also not alphabetical, so we'll specify manually.


```{r}
attend_df <- attend_df |>
  mutate(
    Year = factor(Year, c('Y1', 'Y2', 'Y3', 'Y4', 'MSc', 'PhD'))
  )

contrasts(attend_df$Year)
```

`Y1` is the ref level, and we have five dummy variables which compare Y1 to each following year.


- **Set up continuous predictors.** (e.g., any transformations?)

NA


- **Explore patterns in the data by plotting outcome and predictor variables together.** (It's most reader-friendly to plot the transformed version of the data, the version that you'll also be modelling.)

Consc:

```{r}
# fine for a quick look, but doesn't show the data itself
attend_df |>
  ggplot(aes(x = Consc, y = Attendance)) +
  geom_boxplot()
```

```{r}
# better for a report
p_consc <- attend_df |>
  ggplot(aes(x = Consc, y = Attendance, fill = Consc, colour = Consc)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 3) +
  theme(legend.position = 'none')
p_consc
```

<br>

Year:

```{r}
# fine for a quick look, but doesn't show the data itself
attend_df |>
  ggplot(aes(x = Year, y = Attendance)) +
  geom_boxplot()
```

```{r}
# better for a report
p_year <- attend_df |>
  ggplot(aes(x = Year, y = Attendance, fill = Year, colour = Year)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 3) +
  theme(legend.position = 'none')
p_year
```

Patchwork them together:

```{r}
p_consc + p_year
```



- **Make a nice descriptives table of all relevant variables for the write-up.**

```{r}
attend_df |>
  group_by(Year, Consc) |>
  summarise(n = n(), 
            Mean = mean(Attendance), 
            SD = sd(Attendance),
            Minimum = min(Attendance),
            Maximum = max(Attendance)) |>
    kable(caption = "Descriptive statistics about attendance by year of study and level of conscientiousness", digits = 2) |>
    kable_styling()   
```



# Phase 2: Model fitting

- **Write the mathematical model formulation for your model(s).**

Refer to contrast matrices to know what the dummy variables will be.

$$
\text{Attendance} = \beta_0 + 
(\beta_1 \cdot \text{Consc}_\text{Moderate}) + 
(\beta_2 \cdot \text{Consc}_\text{High}) +  \\
(\beta_3 \cdot \text{Year}_\text{Y2}) + 
(\beta_4 \cdot \text{Year}_\text{Y3}) + 
(\beta_5 \cdot \text{Year}_\text{Y4}) + 
(\beta_6 \cdot \text{Year}_\text{MSc}) + 
(\beta_7 \cdot \text{Year}_\text{PhD}) + 
\epsilon
$$

Incorrect model formula (because disregards dummy variables):

$$
\text{Attendance} = \beta_0 + 
(\beta_1 \cdot \text{Consc}) + 
(\beta_2 \cdot \text{Year}) + 
\epsilon
$$


- **Explicitly define the hypotheses that your RQ is aiming to test.**

We want to know whether there are significant associations between any dummy variable and Attendance, so:

$$
\begin{align}
H_0 &: \text{ All } \beta_j = 0 \text{ for } j = 1, 2, 3, 4, 5, 6, 7\\
H_1 &: \text{ Any } \beta_j \neq 0\\
\end{align}
$$


- **Fit your model(s) using `lm()`.**

```{r}
attend_mdl <- lm(
  Attendance ~ Consc + Year,
  data = attend_df
)
attend_mdl
```



# Phase 3: After model fitting

## 3a: Check assumptions and diagnostics

- **Check model assumptions.**

```{r}
performance::check_model(attend_mdl)
```

Looks all good!

- **If model assumptions about normality/equal variance of errors are violated, bootstrap model estimates.**

NA


- **Run diagnostics for multicollinearity.**

```{r}
vif(attend_mdl)
```

Close to 1, so we're good!


- **Run diagnostics for influential observations.**

Compute Cook's distance for each observation and plot:

```{r}
cooks.distance(attend_mdl)

plot(attend_mdl, which = 4)
```

There are a few data points with relatively larger Cook's Distance values.


- **If you find extreme influential observations: run sensitivity analysis.**

If time allows, run three sensitivity analyses, one for each of the three most extreme points.
(Will need to look at original model first anyway, so come back to this)

```{r}
attend_mdl_no355 <- lm(
  Attendance ~ Consc + Year,
  data = attend_df[-355,]
)
summary(attend_mdl_no355)  # no change to significance pattern
```


```{r}
attend_mdl_no365 <- lm(
  Attendance ~ Consc + Year,
  data = attend_df[-365,]
)
summary(attend_mdl_no365) # no change to significance pattern
```


```{r}
attend_mdl_no161 <- lm(
  Attendance ~ Consc + Year,
  data = attend_df[-161,]
)
summary(attend_mdl_no161)  # no change to significance pattern
```
   
   
## 3b: Plot and interpret model estimates
   
- **Interpret the coefficient estimates** (if appropriate for your RQ).

```{r}
summary(attend_mdl)
```

- `(Intercept)`
  - The estimated mean attendance for people with low conscientiousness (Consc = 0 = Low) in first-year (Year = 0 = Y1) is 14.57 days.

- `ConscModerate`
  - Holding year of study constant, moving from low conscientiousness to moderate conscientiousness is associated with an estimated increase in attendance of 10.80 days.
  - This estimate is significantly different from zero (p < .001).

- `ConscHigh`
  - Holding year of study constant, moving from low conscientiousness to high conscientiousness is associated with an estimated increase in attendance of 18.24 days.
  - This estimate is significantly different from zero (p < .001).

- `YearY2`
  - Holding conscientiousness constant, moving from Y1 to Y2 is associated with an estimated increase in attendance of 4.76 days.
  - This estimate is significantly different from zero (p = 0.004).

- `YearY3`
  - Holding conscientiousness constant, moving from Y1 to Y3 is associated with an estimated increase in attendance of 3.44 days.
  - This estimate is not significantly different from zero (p = 0.07).

- `YearY4`
  - Holding conscientiousness constant, moving from Y1 to Y4 is associated with an estimated increase in attendance of 4.60 days.
  - This estimate is significantly different from zero (p = 0.013).

- `YearMSc`
  - Holding conscientiousness constant, moving from Y1 to MSc is associated with an estimated increase in attendance of 5.96 days.
  - This estimate is significantly different from zero (p = 0.004).

- `YearPhD`   
  - Holding conscientiousness constant, moving from Y1 to PhD is associated with an estimated increase in attendance of 12.57 days.
  - This estimate is significantly different from zero (p < 0.001).


- **Generate a nicely-formatted regression table** (if appropriate for your RQ).

```{r}
tab_model(
  attend_mdl,
  dv.labels = 'Attendance (days)',
  pred.labels = c(
    "ConscModerate" = "Conscientiousness [Moderate]",
    "ConscHigh" = "Conscientiousness [High]"
    ),
  title = "Linear model coefficient estimates"
)
```


- **Compare models and interpret model comparison statistics** (if appropriate for your RQ).

NA


- **Plot model-fitted values** (if appropriate for your RQ).

```{r}
plot_model(
  attend_mdl, 
  type = 'eff', 
  terms = c('Year', 'Consc')   # first term is the one on x axis
)
```

  
  
## 3c: Write up

In labs!