---
title: "<b>Week 3: Model evaluation</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)

test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  hours = seq(0.5,5,.5),
  score = c(1,3,1,2,2,6,3,3,4,8)
)

res <- lm(score ~ hours, data = test)
```

# Weeks Learning Objectives
1. Understand the calculation and interpretation of the coefficient of determination. 

2. Understand the calculation and interpretation of the F-test of model utility.

3. Understand how to standardized model coefficients and when this is appropriate to do.

4. Understand the relationship between the correlation coefficient and the regression slope.

---
# Topics for today
+ Overall model evaluation
  + Coefficent of determination ($R^2$)
  + F-test for the model


---
#  Quality of the overall model 
+ When we measure an outcome ($y$) in some data, the scores will vary (we hope).
  + This we can think of as the total variation we are interested in.

--

+ The aim of our linear model is to build a model which describes $Y$ as a function of $X$.
	+ That is we are trying to explain variation in $Y$ using $X$.
	+ So our model will explain (again, we hope) some variation in $Y$

--

+ But it wont explain it all.
  + What is left is the residual variance.

--

+ We calculate the total, model and residual variances using sums of squares.

$$SS_{total} = SS_{model} + SS_{residual}$$

---
#  Coefficient of determination 
+ One way to consider how good our model is, would be to consider the proportion of total variance our model accounts for. 

+ Enter $R^2$.

+ $R^2$ = coefficient of determination

+ Quantifies the amount of variability in the outcome accounted for by the predictors.
  + More variance accounted for, the better.
  
+ Represents the extent to which the prediction of $Y$ is improved when predictions are based on the linear relation between $X$ and $Y$.

+ Let's see how it works.
  + To do so, we need to calculate the different sums of squares.


---
# Total Sums of Squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{total} = \sum_{i}^{n}(y_i - \bar{Y})^2$$

+ Squared distance of each data point from the mean of $y$.

+ Mean is our baseline. 

+ Without any other information, our best guess at the value of $y$ for any person is the mean.

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle("SS Total") +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]

---
# Calculations

.pull-left[
```{r}
ss_tab <- test %>%
  mutate(
  y_dev = score - mean(score), 
  y_dev2 = y_dev^2
  )
```

```{r}
ss_tab %>%
  summarize(
    ss_tot = sum(y_dev2)
  )
```

]

.pull-right[

```{r, echo=FALSE}
kable(ss_tab) %>%
  kable_styling(., full_width = F)
```

]

---
# Residual sums of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{residual} = \sum_{i}^{n}(y_i - \hat{y}_i)^2$$

+ Which you may recognise.

+ $SS_{residual} = SSE$

+ Squared distance of each point from the predicted value.
]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle("SS Residual") +
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(res$fitted.values)), color = "red", lty =2)

```

]

---
# Calculations

.pull-left[
```{r}
ss_tab <- ss_tab %>%
  mutate(
    y_pred = round(res$fitted.values,2),
    pred_dev = round((score - y_pred),2),
    pred_dev2 = round(pred_dev^2,2)
  )
```

```{r}
ss_tab %>%
  summarize(
    ss_tot = sum(y_dev2),
    ss_resid = sum(pred_dev2) #<<
  )
```

]

.pull-right[

```{r, echo=FALSE}
kable(ss_tab[-c(4,5)]) %>%
  kable_styling(., full_width = F)
```

]


---
# Model sums of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{model} = \sum_{i}^{n}(\hat{y}_i - \bar{Y})^2$$

+ That is, it is the deviance of the predicted scores from the mean of Y.

+ But it is easier to simply take:

$$SS_{model} = SS_{total} - SS_{residual}$$

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle("SS Model") +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, c(res$fitted.values), xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]


---
# Calculations

.pull-left[

$$SS_{model} = SS_{total} - SS_{residual}$$

```{r}
ss_tab %>%
  summarize(
    ss_tot = sum(y_dev2),
    ss_resid = sum(pred_dev2)
  ) %>%
  mutate( #<<
    ss_mod = ss_tot - ss_resid #<<
  )
```

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle("SS Model") +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, c(res$fitted.values), xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]


---
#  Coefficient of determination 
+ Now we can finally come back to $R^2$.

$$R^2 = 1 - \frac{SS_{residual}}{SS_{total}}$$

+ Or

$$R^2 = \frac{SS_{model}}{SS_{total}}$$

+ So in our example:

$$R^2 = \frac{SS_{model}}{SS_{total}} = \frac{22.9}{44.1} = 0.519$$

+ **$R^2$ = 0.519 means that 52% of the variation in test scores is accounted for by hours of revision.**

---
#  Our example 

```{r}
res <- lm(score ~ hours, data = test)
summary(res)
```

???
As at the end of last session, we can check this against the R-ouput:
Be sure to flag small amounts of rounding difference from working through "by hand" and so presenting to less decimal places.

---
#  Significance of the overall model 
+ The test of the individual predictors (IVs, or X?s) does not tell us if the overall model is significant or not.
	+ Neither does R-square
	+ But both are indicative

+ To test the significance of the model as a whole, we conduct an $F$-test.

---
#  F-ratio
+ $F$-ratio tests the null hypothesis that all the regression slopes in a model are all zero
  + We are currently talking about a model with only one $x$, thus one slope.
  + But the $F$-ratio test will generalise.

+ $F$-ratio is a ratio of the explained to unexplained variance:

$$F = \frac{MS_{model}}{MS_{residual}}$$

+ Where MS = mean squares

+ **What are mean squares?**
  + Mean squares are sums of squares calculations divided by the associated degrees of freedom.
  + The degrees of freedom are defined by the number of values associated with the different calculations.

---
# F-table

```{r, echo=FALSE}
ftest <- tibble(
  SS = c("Model", "Residual", "Total"),
  df = c("k", "n-k-1", " "),
  MS = c("SS_model/df_model", "SS_residual/df_residual", " "),
  Fratio = c("MS_model/MS_residual", " ", " "),
  pvalue = c("F(df_model,df_residual)", "", "")
)

kable(ftest)%>%
  kable_styling(., full_width = F)
```

---
# Our example: F-table

```{r, echo=FALSE}
tibble(
  Component = c("Model", "Residual", "Total"),
  SS = c(22.9,21.2,44.1),
  df = c(1, 8, " "),
  MS = c(22.9, 2.65, " "),
  Fratio = c(8.641509, " ", " "),
  pvalue = c("F(1,8)", "", "")
) %>%
  kable(.)%>%
  kable_styling(., full_width = F)
```


---
# F-ratio
+ Bigger $F$-ratios indicate better models.
  + It means the model variance is big compared to the residual variance.

+ The null hypothesis for the model says that the best guess of any individuals $Y$ value is the mean of $Y$ plus error.
	+ Or, that the $X$ variables carry no information collectively about $Y$.
	
+ $F$-ratio will be close to 1 when the null hypothesis is true
  + If there is equivalent residual to model variation, $F$=1
	+ If there is more model than residual $F$ > 1
	
+ $F$-ratio is then evaluated against an $F$-distribution with $df_{model}$ and $df_{residual}$ and a pre-defined $\alpha$

+ Testing the $F$-ratio evaluates statistical significance of the overall model

---
# Visualize the test

.pull-left[

```{r, echo=FALSE}
ggplot(NULL, aes(x = c(-1,10))) +
  stat_function(fun=df,
                geom = "line",
                args = list(df1=1, df2=8)) +
  stat_function(fun = df, 
                geom = "area",
                xlim = c(qf(0.95, 1,8), 10),
                alpha=.25,
                fill = "blue",
                args = list(df1=1, df2=8)) +
  geom_vline(xintercept = 8.641509, col="red") +
  xlab("\n F") +
  ylab("") +
  ggtitle("F-distribution (1,8); f-ratio (8.64; red line)")
```
]

.pull-right[

+ Critical value and $p$-value:
```{r}
tibble(
  Crit = round(qf(0.95, 1, 8),3),
  Exactp = 1-pf(8.64, 1, 8)
)
```

+ From this we would **reject the null**. 

]

---
#  Our example 

```{r}
res <- lm(score ~ hours, data = test)
summary(res)
```

???
As at the end of last session, we can check this against the R-ouput:
Comment on the minor differences for rounding.

---
# Summary of today

---
# Next tasks