---
title: "<b> Effects Interactions </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(emmeans)

hosp_tbl <- read_csv("./hospital.csv", col_types = "dff")

```

# Week's Learning Objectives
1. Interpretation of interactions with effects coded variables.
2. Use contrasts to code specific interaction hypotheses


---
# Hypotheses we test in Factorial Designs
+ Main effects
  + An overall, or average, effect of a condition.
  + Is there an effect of `Treatment` averaged over `Hospital`? 
  + Is there an effect of `Hospital` averaged over `Treatment`? 


+ **Simple contrasts/effects**
  + An effect of one condition at a specific level of another.
  + Is there an effect of `Hospital` for those receiving `Treatment A`? (...and so on for all combinations.)
  
+ **Interactions (categorical-categorical)**
  + A change in the effect of some condition as a function of another.
  + Does the effect of `Treatment` differ by `Hospital`? 
  + With effects coding, we can also think of this as a difference in simple effects.
  

---
# Our model and coefficients

+ Remember our effects coded model:

$$y_{ijk} = b_0 + \underbrace{(b_1E_1 + b_2E_2)}_{\text{Treatment}} + \underbrace{b_3E_3}_{\text{Hospital}} + \underbrace{b_4E_{13} + b_5E_{23}}_{\text{Interactions}} + \epsilon_{i}$$
+ Where the variables represent the following comparisons: 

```{r, echo=FALSE}
tibble(
  Treatment = c("A", "A", "B", "B", "C", "C"),
  Hospital = rep(c("Hosp1", "Hosp2"),3),
  E1 = c(1,1,0,0,-1,-1),
  E2 = c(0,0,1,1,-1,-1),
  E3 = c(1,-1,1,-1,1,-1),
  E13 = c(1,-1,0,0,-1,1),
  E23 = c(0,0,1,-1,-1,1)
)
```

---
# Model coefficients


---
# Our model and coefficients

+ And our model in R:

```{r}
m1 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl,
         contrasts = list(Treatment = contr.sum, 
                          Hospital = contr.sum)) # you can code contrasts within lm
```


---
# Table of means

+ Useful to keep in mind the group means:

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```


---
# Our results
```{r, echo=FALSE}
m1sum <- summary(m1)
summary(m1)
```

---
# Visualizing the interaction

.pull-left[
```{r, echo=FALSE, warning=FALSE}
emmip(m1, Hospital~Treatment) +
  geom_hline(yintercept = mean(hosp_tbl$SWB))
```
]

.pull-right[
```{r, eval=FALSE, warning=FALSE}
emmip(m1, Hospital~Treatment) +
  geom_hline(yintercept = mean(hosp_tbl$SWB))
```
]

---
# Interpretation with effects coding
```{r, echo=FALSE}
round(m1sum$coefficients,2)
```

+ $b_0$ (`Intercept`) = Grand mean.
+ $b_1$ (`Treatment1`) = Difference between row marginal for treatment A and the grand mean. 
+ $b_2$ (`Treatment2`) = Difference between row marginal for treatment B and the grand mean.
+ $b_3$ (`Hospital1`) = Difference between column marginal for Hospital 1 and the grand mean.
+ $b_4$ (`Treatment1:Hospital1`) = Difference between Treatment A and grand mean, in Hospital 1 and Hospital 2
+ $b_5$ (`Treatment2:Hospital1`) = Difference between Treatment B and grand mean, in Hospital 1 and Hospital 2


---
# Interpretation with effects coding
.pull-left[
```{r, echo=FALSE}
round(m1sum$coefficients,2)[,1:2]
```
]

.pull-right[
```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```
]

+ $b_0$ (`Intercept`) = Grand mean.
+ $b_1$ (`Treatment1`) = Difference between row marginal for treatment A and the grand mean. 
+ $b_2$ (`Treatment2`) = Difference between row marginal for treatment B and the grand mean.
+ $b_3$ (`Hospital1`) = Difference between column marginal for Hospital 1 and the grand mean.
+ $b_4$ (`Treatment1:Hospital1`) = Difference between Treatment A and grand mean, in Hospital 1 and Hospital 2
+ $b_5$ (`Treatment2:Hospital1`) = Difference between Treatment B and grand mean, in Hospital 1 and Hospital 2


---
class: center, middle
# Brief pause

---
# Interacting manual contrasts
+ Last week we spoke about creating manual contrasts. 

+ We can create an interaction based on these contrasts.

+ Suppose we wanted to test whether there is a difference between TreatmentA vs TreatmentB and TreatmentC across hospitals.

+ Steps:
  + Define the contrasts for each factor
  + Calculate the product (remember all interactions are products)
  + Create the contrast code
  + Run the contrast
  
---
# Interacting manual contrasts in R

+ Step 1: Individual contrasts

```{r}
treat_con <- c("TreatA" = 1, "TreatB" = -1/2, "TreatC" = -1/2)
hosp_con <- c("Hops1" = 1, "Hosp2" = -1)
```

+ Step 2: Product

```{r}
contr_prod <- outer(treat_con, hosp_con)
contr_prod
```


---
# Interacting manual contrasts in R

+ Step 3: Create the contrast

```{r}
m1_emm <- emmeans(m1, ~Treatment*Hospital)
m1_emm
```


```{r}
int_comp <- list("TreatA vs B and C across hospital" = 
                   c(1, -1/2, -1/2, -1, 1/2, 1/2))
```


---
# Interacting manual contrasts in R

+ Step 4: Run

```{r}
int_comp_test <- contrast(m1_emm, int_comp)
int_comp_test
```


---
class: center, middle
# Thanks for listening!
