---
title: "<b>Week 4: LM Assumptions </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(car)
```

# Weeks Learning Objectives
1. Understand the meaning of model coefficients in the case of a binary predictor.

2. Be able to state the assumptions underlying a linear model.

3. Understand how to assess if a fitted model satisfies the linear model assumptions.

4. Understand how to use transformations when the model violates assumptions.

---
# Topics for today
+ What are the assumptions of regression and how can we test them?
	+ Normally distributed errors
	+ Homoscedasticity
	+ Linearity
	+ Independence of errors

---
# Regression assumptions 
+ So far, we have discussed evaluating regression models with respect to:
	+ Overall model fit (F-ratio, R^2)
	+ Individual predictors
	+ Individual cases
+ However, the regression model is also built on a set of assumptions
+ If these assumptions are violated, the model will not be very accurate
+ Thus, we also need to test these assumptions

---
# Example: 


---
# The lm() results

+ We do all our assumption testing after fitting the lm( ) model, using the lm( ) object (?M1?) we save from the analysis


---
# Plots vs tests
+ about to talk about assumption checks
+ will present statistical test and plots
+ In general, graphical methods are often more useful
	+ Easier to see the nature and magnitude of the assumption violation

---
# Normally distributed errors 
+ It is assumed that the residuals (differences between predicted and observed scores) in a linear regression are normally distributed around each predicted DV score
+ Normality of residuals can be tested via a range of methods:
+ Graphical methods
	+ QQ-plots, Histograms
+ Statistical tests:
	+ Shapiro-Wilk test

	
---
# hist() 
+ Histograms plot the frequency distribution of the residuals
+ In R, we can plot a histogram of the residuals using the hist () function

---
# Q-Q Plots 
+ Quantile comparison plots (Q-Q plots)
	+ Plot the studentised residuals from the model against their theoretically expected values
	+ If the residuals are normally distributed, the points should fall neatly on the diagonal of the plot
	+ Non-normally distributed residuals cause deviations of points from the diagonal
		+ The specific shape of these deviations are characteristic of the distribution of the residuals

---
# qqPlot() 


---
# shapiro.test() 

+ The Shapiro-Wilk test provides a significance test on the departure from normality
+ A non-significant *p*-value suggests that the residuals don?t deviate from normality




---
#  Homoscedasticity 

+ Homoscedasticity is the assumption that error variance is constant and not related to the values of the predictors or the predicted Y values
	+ Heteroscedasticity refers to when this assumption is violated (non-constant variance)
+ Graphical methods
	+ Plot residual values against each predictor and the predicted Y values
+ Statistical tests
	+ Breusch-Pagan test

---
#  Residual-vs-predicted values plot 
+ In R, we can plot the Pearson residuals against the predictor values and predicted y-values using the residualPlots ( ) function from the car package

---
#  residualPlots() output 
+ Categorical predictors should show a similar spread of residual values across their levels
+ The plots for continuous predictors should look like a random array of dots
	+ The solid line should follow the dashed line closely

---
#  Breusch-Pagan test 
+ Also called the non-constant variance test
+ Tests whether residual variance depends on the predicted values
+ Implemented using the ncvTest () function in R
+ Non-significant *p* -value suggests homoscedasticity assumption holds

---
#  Linearity 
+ Linearity is the assumption that each IV has a straight line relation with the DV
+ Assuming a linear relation when the true relation is non-linear can result in under-estimating that relation

+ show plots of linear vs no linear

---
#  Non-linearity in simple regression 
+ In simple regression, plotting the relation between X and Y and looking for curvature can often identify non-linearity
	+ Adding a lowess line and comparing it with the line of best fit from an lm ( ) can highlight departures from linearity



---
# Violated Assumptions
+ What to do about non-normality of residuals, heteroscedasticity and non-linearity 



---
# Non-linear transformations 
+ Often non-normal residuals, heteroscedasticity and non-linearity can be ameliorated by a non-linear transformation of the DV and/or IVs
+ This involves applying a function to the values of a variable to change its values and overall shape of its distribution
	+ For non-normal residuals and heteroscedasticity, skewed DVs can be transformed to normality
	+ Non-linearity may be helped by a transformation of both IV and DV

---
#  Transforming variables to normality 
+ Positively skewed data can be made more normally distributed using a log-transformation
+ Negatively skewed data can be made more normally distributed using same procedure but first reflecting the variable (make biggest values the smallest and smallest the biggest) and then applying the log-transform

---
#  Log-transformations 
+ Log-transformations can be implemented in R using the log() function
+ If your variable contains zero or negative values, you need to first add a constant to make all your values positive
	+ A good strategy is to add a constant so that your minimum value is one
	+ E.g., if your minimum value is -1.5, add 2.5 to all your values

---
#  Other potential solutions 
+ Adding higher-order terms to the model 
+ Bootstrapping 

---
#  Independence of errors 
+ Independence of errors is the assumption that residual values are not correlated with one another
+ Difficult to test unless we know the potential source of correlation between cases
+ We can test a limited form of the assumption by testing for autocorrelation between errors
	+ We can test the correlation between each case an adjacent cases in the dataset
	+ Achieved using the Durbin-Watson test

---
#  Durbin-Watson test 
+ Durbin-Watson test implemented in R using the durbinWatsonTest () function:
+ The D-W statistic can take values between 0 and 4
	+ 2= no autocorrelation
+ Therefore, we ideally want D-W values close to 2 and a non-significant *p*-value
+ Values <1 or >3 may indicate problems

---
# Summary of today
+ The accuracy of regression results depends on a number of assumptions:
	+ Normality of residuals
		+ QQ-plots, histograms of studentised residuals, Shapiro Wilk
	+ Homoscedasticity
		+ Plots of residuals against predicted and IV values, non-constant variance test
	+ Linearity
		+ Component-residual plots
	+ Independence of errors
		+ Durbin-Watson test
+ Non-linear transformations of DV and/or IVs can often resolve issues with the first three assumptions

---
# Next tasks


