---
title: "Calculating the Slope and Intercept with binary predictor"
author: "dapR2 Team"
output: html_document
---

In the lecture we noted the nice properties of 0 and 1 as values for binary variables. This becomes really clear if we run through the slope and intercept calculations with a binary variable. So here we will follow the same steps as week 2, using a binary.

## Packages
For this example we will only need the `tidyverse` package.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Data
Below is code to create a smaller version of the data from the lecture, where we have the `method` variable for mode of study coded as 0 and 1, and the score on the test.

```{r}
test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  study = c(0,0,0,0,0,1,1,1,1,1),
  score = c(1,3,1,2,2,6,3,3,4,8)
)
```

It will be useful for us to have a mental note of the means of each group.

```{r}
test %>%
  group_by(factor(study)) %>%
  summarise(
    mean = mean(score)
  )
```

So we can see the mean for those in the `0` group (study alone) is 1.8, and the mean for those in the `1` group (study with others) is 4.8.

## Calculating slope

From lecture, we can calculate the slope using the following:

$$
\hat \beta_1 = \frac{SP_{xy}}{SS_x}
$$

Where $SP_{xy}$ = sum of cross-products:


$$
SP_{xy} = \sum_{i = 1}^{n}(x_i - \bar{x})(y_i - \bar{y})
$$

and $SS_x$ = sums of squared deviations of $x$:


$$
SS_x = \sum_{i = 1}^{n}(x_i - \bar{x})^2
$$

### Sum of cross-products

$$
SP_{xy} = \sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})
$$

We will follow the same calculations as in week 2, for extended discussion of the steps, read this file side by side with week 2.

```{r}
calc <- test %>%
  mutate(
    m_study = mean(study),
    m_score = mean(score)
  ) %>%
  select(1,2,4,3,5)

calc
```

Next, we can calculate the deviances:

```{r}
calc <- calc %>%
  mutate(
    dev_study = study - m_study,
    dev_score = score - m_score
  ) %>%
  select(1:3, 6, 4,5,7)
calc
```

The last step is to multiply the two deviances for each person.

```{r}
calc <- calc %>%
  mutate(
    SPXYi = dev_study * dev_score
  )
calc
```

And there we go. Our `SPXYi` variable contains the product of the deviations of each observation from the means of hours studied and test score. Our final step is just to add this up.

```{r}
SPXY <- sum(calc$SPXYi)
SPXY
```

### Sum of square deviations
So now to calculate $SS_x$. Let's look at the equation again:

$$
SS_x = \sum_{i=1}^{n}(x_i - \bar{x})^2
$$

We have already calculated the individual deviations $(x_1 - \bar{x}), \dots, (x_n - \bar{x})$. They are the column called `dev_hours` in the `calc` table. So all we need to do is square these for each observation and sum them. 

```{r}
calc <- calc %>%
  mutate(
    SSxi = dev_study^2
  )
calc
```

and

```{r}
SSx = sum(calc$SSxi)
SSx
```

### Return to the slope
We now have the two values we need to calculate the slope:

$$
\hat \beta_1 = \frac{SP_{xy}}{SS_x} = \frac{7.5}{2.5} = 3
$$

```{r}
beta1_hat <- SPXY/SSx
beta1_hat
```

Now to the intercept.

## Calculating intercept
Calculations for intercept:

$$
\hat \beta_0 = \bar{y} - \hat \beta_1 \bar{x}
$$

So this one is much easier as all we need are:

- $\hat \beta_1$ = slope estimate
- $\bar{y}$ = mean of $y$
- $\bar{x}$ = mean of $x$

So...


```{r}
beta0_hat <- mean(calc$score) - (beta1_hat * mean(calc$study))
beta0_hat
```

So we arrive at 1.8 as the intercept - the mean of the group coded 0, and a slope of 3, the difference between the means of the two groups.


If we were to repeat this with using values of 1 and 2, we would see different results.

