---
title: "<b>Course Introduction </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #  base_color = "#0F4C81", # DAPR1
  base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  # base_color = "#95A5A6", #intro
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)

library(tidyverse)
```


```{r welcome, echo=FALSE}
cohort <- 2021
team <- c("Aja", "Alex", "Emma", "Josiah", "Tom", "Umberto", "Zach", "all the tutors")
lecture_hours <- 10+10+50
lab_hours <- 20+20
```


# In this intro...
+ The dapR2 team

--

+ What each week will look like.

--

+ Expectations




---
# dapR2 Team
+ Tom Booth (CO, Lectures, Labs)
+ Emma Waterston (CO, Labs)
+ Monica Truelove-Hill (Lectures, Labs)
+ Zach Horne (Lectures)
+ Umberto Noe
+ Josiah King
+ Aja Murray (Stats Curriculum Coordinator)
+ Wonderful tutors (lots)



---
# What is R? 
+ A very flexible, very cool, programming language for all things data.
  + It does pretty much any statistical method you can think of. (Cool? I think so)
  + But it does a lot more.

+ We will also continue to teach you Rmarkdown 
  + this is a cool way to integrate text and analysis
  + and to write reproducible documents and analyses

---
# What is R? 
+ Open source (free) software for all things data.

+ Huge community worldwide which means lots of free resources in addition to your course materials

--

+ Some examples:
  + [interactive plots](https://shiny.rstudio.com/gallery/movie-explorer.html) 
  + [interactive dashboards](https://gallery.shinyapps.io/086-bus-dashboard/) 
  + Documents with automatically include results from analysis
  + [books](https://bookdown.org/csgillespie/efficientR/)
  + [websites](https://rmarkdown.rstudio.com/)
  + Presentations (like the one you are looking at)


---
# dapR 2 
+ In dapR2, we will teach you how to...

--

  + Deal with data in R, tidy it, describe it and visualize it

--

  + Run, interpret and use linear models for observational designs
  
--

  + Run, interpret and use linear models for experimental designs
  
--

  + Introduce the basic concepts of the generalized linear model

--

  + Talk about some common issues when applying the linear model to data.

---
# dapR2

+ Semester 1:
  + Weeks 1 to 4: Basics of the linear model
  + Weeks 5 to 7: Interactions
  + Weeks 8 to 10: Robustness checks for our models

--

+ Semester 2:
  + Weeks 1 to 5: Working with categorical data and analyzing experiments
  + Weeks 6 to 10: What if.....
  
---
# Basis for dapR 3 (Hons Stats) 
+ In dapR3, we will teach you how to...

--

  + Run, interpret and use linear mixed models to analyse repeated measures data

--

  + Run, interpret and use data reduction techniques

--

  + Run, interpret and use path models for testing more complex relations between variables

---
# dapR2 Course Delivery

+ **Lectures**
  + Two 50 minute lectures per week.
  + These will be a mix of new content are practical demonstration/Q&A (like the Live R's last year)

--

+ **Labs**
  + 1 hour per week in person labs (note these may take more than 1 hour to complete = self-study expected)
  + Structured reading and exercises with R.
  + VERY important sessions for practical skills.

--

+ **Supported by**
  + Office hours (see LEARN for times)
  + Piazza discussion boards (monitored daily)

--

+ **Reading**
  + All reading is from free to access sources.
  + You do not need to buy a book and should be able to access everything electronically.


???
+ Requires you to work steadily and consistently across the year. 
+ It builds, and it is hard to cram.
+ To help you build this study skills, course has a specific assessment structure.


---
# dapR2: Assessment 
+ **Weekly quizzes** (10%)
  + 20 quizzes in total 
  + Quizzes 1 and 2 are practices.
  + The rest comprise your grade.
  + Mark is the average of your best 14/18 scores.

--

+ **Coursework report** (30%)
  + Early semester 2
  + Provided a dataset and some questions
  + Task is to construct models to answer the questions, describe them, run them, report the results.

--

+ **Exam** (60%)
  + Summer exam block
  + Mix of MCQ, short answer R/calculation, and larger interpretation questions.
  + Closed book (equation sheets provided)
  + Main course assessment


---
# Materials
+ All of your materials for each week will be within the weekly folders on LEARN.

--

+ It will contain:
  + Slide decks and links to lecture recordings
  + Links to lab material
  + Links to the weekly quizzes
  + Any comments or specific instructions for the week
  + Links to discussion boards
  + Reading (all freely available on-line)

--

+ The weekly folders will become available incrementally.


---
# What you can expect from us
1. We will work hard to help you learn.

2. We will be open and communicate with you.

3. We will be polite, respectful and treat you like adults.


---
# What we expect of you
1. You work hard.

2. That you talk to us.

3. That you are polite, and respect the teaching team and your classmates.

4. Try and have fun.

---
# A brief word on engagement
+ dapR courses require consistent work across the course. 
  + You can not really cram this content.

--

+ We want to make sure everyone keeps on top of the course.
  + To do that, we need to know when people are struggling. 

--

+ Please help us by being proactive:
  + Come to all sessions
  + Use office hours to clarify any questions you have.
  + We can help construct peer study groups to learn from one another.
  + Monitor your quiz performance.

---
class: center, middle
# And that is it for this intro!
