---
title: 'dapR2 live R: Week 3'
author: "Tom Booth"
date: "04/10/2021"
output: html_document
---

# Searching objects and in-line code
Some data:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(10101)
service <- rnorm(100, 4, 1)
b0 = 15
b1 = 5
eps <- rnorm(100, 4, 5)
salary <- b0 + (b1*service) + eps

tib <- tibble(
  ID = paste("ID",101:200, sep=""),
  service = service,
  salary = salary
)

```

And a simple linear model

```{r}
lm(salary ~ service, data = tib)
```

With the explicit intercept

```{r}
lm(salary ~ 1 + service, data = tib)
```

```{r}
res <- lm(salary ~ service, data = tib)
```


## Exploring the object

Seeing what is within an object:

1. Using `names()`
2. `?lm`
3. In the global environment

```{r}
names(res)
```


## Extracting elements
To print:

```{r}
res$coefficients
```

Getting a specific value:

```{r}
res$coefficients[2]
```

To save for later:

```{r}
slope <- res$coefficients[2]
slope
```


## Using these in-line 
Writing a standard statement for a linear model interpretation:

Here is the slope `r round(res$coefficients[2], 2)` and again `r round(slope,1) `

# Extracting more information
Some are things missing from this model object from both week 2 and up and coming this week.

```{r}
res_sum <- summary(res)
res_sum
```

Look inside this object:

```{r}
names(res_sum)
```

See the coefficients:

```{r}
res_sum$coefficients
```

Select individual elements from this:

```{r}
res_sum$coefficients[2,4]
```

```{r}
res_sum$r.squared
```

