---
title: "<b>Why Do We Need a Reference Group?</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(simglm)

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.retina = 1.5)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(3119)

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation + study + method,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1),
               study = list(var_type = 'factor',
                            levels = c('alone', 'others'),
                            prob = c(0.53, 0.47)),
               method = list(var_type = 'factor',
                            levels = c('read', 'summarise', 'self-test'),
                            prob = c(0.3, 0.4, 0.3))),
  error = list(variance = 20),
  sample_size = 250,
  reg_weights = c(0.6, 1.4, 1.5, 6, 6, 2)
)

df3 <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study3 <- df3 %>%
  dplyr::select(y, hours, motivation, study, method) %>%
  mutate(
    ID = paste("ID", 101:350, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2),
    study = factor(study),
    method = factor(method)
  ) %>%
  dplyr::select(ID, score, hours, motivation, study, method)

mod1 <- lm(score ~ method, data = test_study3)
```


# Warning!
+ From a practical perspective, what we have covered in the lecture is what you need to know.

+ However, to understand how some of the other coding schemes work, there are a couple of technical bits you may find useful to consider.

---
# Why do we need a reference group?

.pull-left[

+ Consider our example and dummy coding.

+ Our `method` variable has three groups/levels (`read`, `self-test` & `summarise` )

+ We want a model that represents our data (observations), but all we "know" is what group an observation belongs to. So;

$$y_{ij} = \mu_i + \epsilon_{ij}$$

+ Where 
  + $y_{ij}$ are the individual observations
  + $\mu_i$ is the mean of group $i$ and
  + $\epsilon_{ij}$ is the individual deviation from that mean.

]

.pull-right[
```{r, echo = F}
 test_study3 %>%
   dplyr::select(ID,score,method) %>%
   slice(1:10)
```

]

???
+ And this hopefully makes sense.
  + Given we know someone's group, our best guess is the mean
  + But people wont all score the mean, so there is some deviation for every person.


---
# Why do we need a reference group?
+ An alternative way to present this idea looks much more like our linear model:

$$y_{ij} = \beta_0 + \underbrace{(\mu_{i} - \beta_0)}_{\beta_i} + \epsilon_{ij}$$
+ Where 
  + $y_{ij}$ are the individual observations
  + $\beta_0$ is an estimate of reference/overall average
  + $\mu_i$ is the mean of group $i$ 
  + $\beta_1$ is the difference between the reference and the mean of group $i$, and
  + $\epsilon_{ij}$ is the individual deviation from that mean.

---
# Why do we need a reference group?
+ We can write this equation more generally as:

$$\mu_i = \beta_0 + \beta_i $$

+ or for the specific groups (in our case 3):

$$\mu_{read} = \beta_0 + \beta_{1read}$$

$$\mu_{self-test} = \beta_0 + \beta_{2self-test}$$

$$\mu_{summarise} = \beta_0 + \beta_{3summarise}$$

+ **The problem**: we have four parameters ( $\beta_0$ , $\beta_{1read}$ , $\beta_{2self-test}$ , $\beta_{3summarise}$ ) to model three group means ( $\mu_{read}$ , $\mu_{self-test}$ , $\mu_{summarise}$ )

+ We are trying to estimate too much with too little.
    + We need to estimate at least 1 parameter less

---
# Constraints fix identification
+ Let's think again about dummy coding.

+ Suppose we make `read` the reference. Then, 

$$\mu_{read} = \beta_0$$

$$\mu_{self-test} = \beta_0 + \beta_{1self-test}$$

$$\mu_{summarise} = \beta_0 + \beta_{summarise}$$
+ **Fixed** ! 

+ We now only have three parameters ( $\beta_0$ , $\beta_{self-test}$ , $\beta_{summarise}$ ) for the three group means ( $\mu_{read}$ , $\mu_{self-test}$ , $\mu_{summarise}$ ). 


> So when we code categorical variables, we need a constraint so that we can estimate our models.

---
# One last look at our model
```{r}
summary(mod1)
```

