---
title: "<b>Planned Contrast - Testing specific hypotheses </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives
1. Interpret the output from a model using dummy coding and sum-to-zero coding

2. Create specific contrast matrices to test specific effects.

3. Recognise other forms of contrasts.

---
# Topics for today


---
# Manual contrast testing
+ We can structure a wide variety of contrasts so long as they can be written

1. A as a linear combination of population means
2. The associated coefficients sum to zero

+ So

$$H_0: c_1\mu_1 + c_1\mu_2 + c_3\mu_3 $$

+ With

$$c_1 + c_2 + c_3 = 0$$

---
# Manual contrast testing
+ For example, say we wanted to compare TreatmentA to TreatmentB and TreatmentC combined.

+ Our null becomes:

$$H_0: \mu_{treatmentA} = \frac{1}{2}(\mu_{treatmentB} + \mu_{treatmentC})$$

+ Or, is the average SWB of TreatmentA group equal to the average SWB of the combined TreatmentB and TreatmentC

+ We can set the contrasts here:
  + $c_1 = 1$
  + $c_2 = -1/2$
  + $c_3 = -1/2$


---
# Summary of today


