---
title: "<b> Effects Interactions </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(emmeans)

hosp_tbl <- read_csv("./hospital.csv", col_types = "dff")

```

# Week's Learning Objectives
1. Interpretation of interactions with effects coded variables.
2. Use contrasts to code specific interaction hypotheses


---
# Hypotheses we test in Factorial Designs
+ Main effects
  + An overall, or average, effect of a condition.
  + Is there an effect of `Treatment` averaged over `Hospital`? 
  + Is there an effect of `Hospital` averaged over `Treatment`? 


+ **Simple contrasts/effects**
  + An effect of one condition at a specific level of another.
  + Is there an effect of `Hospital` for those receiving `Treatment A`? (...and so on for all combinations.)
  
+ **Interactions (categorical-categorical)**
  + A change in the effect of some condition as a function of another.
  + Does the effect of `Treatment` differ by `Hospital`? 
  + With effects coding, we can also think of this as a difference in simple effects.


---
# Understanding our tests with effects codes
+ As discussed previously, each $\beta$ in the effects code model is the difference between a marginal mean and the grand mean

$$\mu_{maginal} - \mu_{..}$$

+ Lets look at this more closely.

+ Suppose we have the following data:

```{r, echo = FALSE}
dat2 <- tibble(
  cond1 = c("A", "A", "B", "B"),
  cond2 = c("C", "D", "C", "D"),
  score = c(12, 14, 20, 5)
)

eg2 <- matrix(c(12,14,13,20,5,12.5,16,9.5,12.75),ncol=3,byrow=TRUE)
colnames(eg2) <- c("C","D", "Mean")
rownames(eg2) <- c("A","B","Mean")
eg2 <- as.table(eg2)

kable(eg2) %>%
  kable_styling(., full_width = FALSE)
```

+ Lets visualize the tests of each condition

---
# Understanding our tests with effects codes
.pull-left[
```{r, echo=FALSE}
kable(eg2) %>%
  kable_styling(., full_width = FALSE)
```


]


.pull-right[

```{r, echo=FALSE}
gm <- mean(dat2$score)
dat2 %>%
  ggplot(., aes(x = cond1, y = score, group = cond2, colour = cond2, shape = cond2)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,30)) +
  annotate("point", x = 1, y = 13, size = 2) +
  annotate("point", x = 2, y = 12.5, size = 2) +
  geom_hline(yintercept = 13, linetype = "dashed") +
  geom_hline(yintercept = 12.5, linetype = "dashed") +
  geom_hline(yintercept = gm, linetype = "dashed", color = "blue" ) +
  ggtitle("No effect cond1")
```

]

---
# Understanding our tests with effects codes
.pull-left[
```{r, echo=FALSE}
kable(eg2) %>%
  kable_styling(., full_width = FALSE)
```


]


.pull-right[

```{r, echo=FALSE}
dat2 %>%
  ggplot(., aes(x = cond1, y = score, group = cond2, colour = cond2, shape = cond2)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,30)) +
  annotate("point", x = 1.5, y = 16, size = 2) +
  annotate("point", x = 1.5, y = 9.5, size = 2) +
  geom_hline(yintercept = 16, linetype = "dashed") +
  geom_hline(yintercept = 9.5, linetype = "dashed") +
  geom_hline(yintercept = gm, linetype = "dashed", color = "blue" ) +
  ggtitle("Effect cond2")
```

]


---
# Understanding our tests with effects codes
+ For the interaction it is a little more complicated.

+ However, fundamentally keep in mind a categorical interaction is saying is there a difference in a differences.
  + In the case of effects coding, we are asking if a marginal mean differs from the grand mean.
  + So we are thinking about whether the distance of the row or column marginals from the grand means differs depending on level of another condition.


---
# Understanding our tests with effects codes
.pull-left[
```{r, echo=FALSE}
kable(eg2) %>%
  kable_styling(., full_width = FALSE)
```


]


.pull-right[

```{r, echo=FALSE}
dat2 %>%
  ggplot(., aes(x = cond1, y = score, group = cond2, colour = cond2, shape = cond2)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,30)) +
  geom_hline(yintercept = gm, linetype = "dashed", color = "blue" ) +
  geom_segment(aes(x= 0.9 , xend = 0.9 , y=12, yend=14), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  geom_segment(aes(x= 2.1 , xend = 2.1 , y=20, yend=5), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  annotate("text", x = 0.85 , y = 14, label = "-2") +
  annotate("text", x = 2.15 , y = 11.5, label = "15") +
  ggtitle("Interaction")
```

]


---
# Understanding our tests with effects codes
.pull-left[
```{r, echo=FALSE}
dat5 <- tibble(
  cond1 = c("A", "A", "B", "B"),
  cond2 = c("C", "D", "C", "D"),
  score = c(20, 19, 10, 9)
)

eg5 <- matrix(c(20,19,19.5,10,9,9.5,15,14,14.5),ncol=3,byrow=TRUE)
colnames(eg5) <- c("C","D", "Mean")
rownames(eg5) <- c("A","B","Mean")
eg5 <- as.table(eg5)

kable(eg5) %>%
  kable_styling(., full_width = F)
```


]


.pull-right[

```{r, echo=FALSE}
dat5 %>%
  ggplot(., aes(x = cond1, y = score, group = cond2, colour = cond2, shape = cond2)) +
  geom_line() +
  geom_point(aes(size = 1.5), show.legend = FALSE) +
  ylim(c(0,30)) +
  geom_segment(aes(x= 0.9 , xend = 0.9 , y=19, yend=20), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  geom_segment(aes(x= 2.1 , xend = 2.1 , y=9, yend=10), 
               arrow=arrow(type = "closed", end = "both", length = unit(0.2, "cm")), 
               colour = "black", linetype = "dashed") +
  annotate("text", x = 0.8 , y = 19.5, label = "1") +
  annotate("text", x = 2.2 , y = 9.5, label = "1") +
  ggtitle("No Interaction")
```

]


---
# Our model and coefficients

+ Remember our effects coded model:

$$y_{ijk} = b_0 + \underbrace{(b_1E_1 + b_2E_2)}_{\text{Treatment}} + \underbrace{b_3E_3}_{\text{Hospital}} + \underbrace{b_4E_{13} + b_5E_{23}}_{\text{Interactions}} + \epsilon_{i}$$
+ Where the variables represent the following comparisons: 

```{r, echo=FALSE}
tibble(
  Treatment = c("A", "A", "B", "B", "C", "C"),
  Hospital = rep(c("Hosp1", "Hosp2"),3),
  E1 = c(1,1,0,0,-1,-1),
  E2 = c(0,0,1,1,-1,-1),
  E3 = c(1,-1,1,-1,1,-1),
  E13 = c(1,-1,0,0,-1,1),
  E23 = c(0,0,1,-1,-1,1)
)
```


---
# Our model and coefficients

+ And our model in R:

```{r}
m1 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl,
         contrasts = list(Treatment = contr.sum, 
                          Hospital = contr.sum)) # you can code contrasts within lm
```


---
# Table of means

+ Useful to keep in mind the group means:

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```


---
# Our results
```{r, echo=FALSE}
m1sum <- summary(m1)
summary(m1)
```

---
# Visualizing the interaction

.pull-left[
```{r, echo=FALSE, warning=FALSE}
emmip(m1, Hospital~Treatment) +
  geom_hline(yintercept = mean(hosp_tbl$SWB))
```
]

.pull-right[
```{r, eval=FALSE, warning=FALSE}
emmip(m1, Hospital~Treatment) +
  geom_hline(yintercept = mean(hosp_tbl$SWB))
```
]


---
# Interpretation with effects coding
.pull-left[
```{r, echo=FALSE}
round(m1sum$coefficients,2)[,1:2]
```
]

.pull-right[
```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```
]

+ $b_0$ (`Intercept`) = Grand mean.
+ $b_1$ (`Treatment1`) = Difference between row marginal for treatment A and the grand mean. 
+ $b_2$ (`Treatment2`) = Difference between row marginal for treatment B and the grand mean.
+ $b_3$ (`Hospital1`) = Difference between column marginal for Hospital 1 and the grand mean.


---
# Interpretation with effects coding
.pull-left[
```{r, echo=FALSE}
round(m1sum$coefficients,2)[,1:2]
```
]

.pull-right[
```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```
]

+ $b_4$ (`Treatment1:Hospital1`) = Does the effect of Treatment A differ across Hospital 1 and Hospital 2
+ $b_5$ (`Treatment2:Hospital1`) = Does the effect of Treatment B differ across Hospital 1 and Hospital 2

$$(\mu_{11} - \mu_{..}) - (\mu_{1.} - \mu_{..}) - (\mu_{.1} - \mu_{..})$$


---
# Overall effect of interaction
+ We looked at how to assess this when discussing model comparisons. 

+ We can compare nested models based on the $F$-test

```{r}
m0 <- lm(SWB ~ Treatment + Hospital, data = hosp_tbl,
         contrasts = list(Treatment = contr.sum, Hospital = contr.sum))
m1 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl,
         contrasts = list(Treatment = contr.sum, Hospital = contr.sum))
anova(m0, m1)
```

---
# Overall effect of interaction
+ or use `anova()` on the single model.

```{r}
anova(m1)
```




---
class: center, middle
# Brief pause

---
# Interacting manual contrasts
+ Last week we spoke about creating manual contrasts. 

+ We can create an interaction based on these contrasts.

+ Suppose we wanted to test whether there is a difference between TreatmentA vs TreatmentB and TreatmentC across hospitals.

+ Steps:
  + Define the contrasts for each factor
  + Calculate the product (remember all interactions are products)
  + Create the contrast code
  + Run the contrast
  
---
# Interacting manual contrasts in R

+ Step 1: Individual contrasts

```{r}
treat_con <- c("TreatA" = 1, "TreatB" = -1/2, "TreatC" = -1/2)
hosp_con <- c("Hops1" = 1, "Hosp2" = -1)
```

+ Step 2: Product

```{r}
contr_prod <- outer(treat_con, hosp_con)
contr_prod
```


---
# Interacting manual contrasts in R

+ Step 3: Create the contrast

```{r}
m1_emm <- emmeans(m1, ~Treatment*Hospital)
m1_emm
```


```{r}
int_comp <- list("TreatA vs B and C across hospital" = 
                   c(1, -1/2, -1/2, -1, 1/2, 1/2))
```


---
# Interacting manual contrasts in R

+ Step 4: Run

```{r}
int_comp_test <- contrast(m1_emm, int_comp)
int_comp_test
```


---
class: center, middle
# Thanks for listening!
