---
title: "Exam Prep"
author: "DAPR2 Team"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)
source('_theme/theme_quarto.R')
```


# Course Overview {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Introduction to Linear Models"
block1_lecs = c("Intro to Linear Regression",
                "Interpreting Linear Models",
                "Testing Individual Predictors",
                "Model Testing & Comparison",
                "Linear Model Analysis")
block2_name = "Analysing Experimental Studies"
block2_lecs = c("Categorical Predictors & Dummy Coding",
                "	Effects Coding & Coding Specific Contrasts",
                "Assumptions & Diagnostics",
                "Bootstrapping",
                "	Categorical Predictor Analysis")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Interactions"
block3_lecs = c("Interactions I",
                "Interactions II",
                "Interactions III",
                "Analysing Experiments",
                "Interaction Analysis")
block4_name = "Advanced Topics"
block4_lecs = c("Power Analysis",
                "Binary Logistic Regression I",
                "Binary Logistic Regression II",
                "Logistic Regression Analysis",
                "	Exam Prep and Course Q&A")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=10)
```

:::

::::


## Outline

+ Things to take into exam  

+ Process  

+ Structure of exam  

+ Exam *'strategy'*  

+ Example Questions  


## Things You Need for the Exam

:::: {.columns}

::: {.column width="50%"}

**What you need to take**

+ Student Card

+ Pens (blue or black ink)  

+ Pencils and rubber

+ Ruler
  + If you like, not really needed, but might be useful 
  
+ Calculator
  + See list on Learn of permitted calculators  
  + **Phone/watch cannot be your calculator**  

:::

::: {.column width="50%"}

**What you will be given**

+ Attendance slip    

+ The exam question paper (inc. equation & formulae section)    

+ A MCQ answer sheet for Section A questions    

+ An answer book for Section B questions    

:::

::::


## Process

+ You need to check your calendar [https://exams.is.ed.ac.uk/](https://exams.is.ed.ac.uk/)

+ Arrive in plenty of time (approx. 30 mins before exam starts)  

+ There will be direction to room    

+ Follow instructions for where to leave jackets, bags, phones, etc.    

+ Take all stationary, calculator, and **student ID card** to your desk  

+ Invigilators will give you exam instructions  


## Filling in Front Covers

::: {.grid layout-ncol=3}

```{=html}
<div style="aspect-ratio:4/3;">
<iframe height="315" src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_uwsdz3vt&config[provider]={"widgetId":"1_ppozvu21"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to complete an examination attendance form"></iframe>
</div>
```

```{=html}
<div style="aspect-ratio:4/3;">
<iframe height="315" src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_spaem93h&config[provider]={"widgetId":"1_rb96nq0i"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to prepare barcoded exam scripts and MCQ sheets"></iframe>
</div>
```

```{=html}
<div style="aspect-ratio:4/3;">
<iframe height="315" src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_ms01z6h1&config[provider]={"widgetId":"1_p1u7vrqf"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to fill in the cover of an exam script book?"></iframe>
</div>
```

:::

## Things You Can Take Away

<center>
<br><br>
<h2><b>NOTHING*!</b></h2>
*<i>Other than the things you bring in with you like pens, pencils, calculator</i>

<br><br>
<b>
When you finish, you must leave all papers (question paper (inc. equation & formulae section), answer sheets/books) in the exam hall  
</b></center>


## Exam Structure

:::: {.columns}

::: {.column width="50%"}

**Section A - 30 marks**

+ 15 MCQs

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from

:::

::: {.column width="50%"}

**Section B - 70 marks**  

+ 8 Questions  

+ Marks range from 3 to 22 marks per question:  
    + 1 x 22 marks
    + 1 x 14 marks
    + 2 x 8 marks
    + 1 x 6 marks
    + 1 x 5 marks
    + 1 x 4 marks
    + 1 x 3 marks

+ On range of topics  

:::

::::


## Exam Strategy: Section A

:::: {.columns}

::: {.column width="50%"}

**Section A - 30 marks**

+ 15 MCQs  

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from

:::

::: {.column width="50%"}

**Suggested Approach**    
  
+ Similar to quiz questions on Learn      

+ Spend 15-30 minutes on MCQs:    

  + Approx. 1-2 minutes per MCQ   
  
  + You can answer these quickly at the end   
  
  + Do the ones you know, then move on    
  
  + Keep track of how many you have not done, so you can go back to them at the end
  
:::

::::


## Exam Strategy: Section B {.smaller}


:::: {.columns}

::: {.column width="50%"}

**Section B - 70 marks**  

+ 8 questions, where marks range from 3 to 22 marks per question:  
    + 1 x 22 marks
    + 1 x 14 marks
    + 2 x 8 marks
    + 1 x 6 marks
    + 1 x 5 marks
    + 1 x 4 marks
    + 1 x 3 marks

+ On range of topics, where some questions will involve:
    + Short calculations
    + Asking you to explain statistical concepts  
    + Giving you model output and ask you to provide interpretation  

:::

::: {.column width="50%"}

**Suggested Approach** 

+ 70% of marks in this section
+ Spend more time on higher mark questions
    + Marks arranged from highest to lowest marks (i.e., Q1 = 22 marks, ..., Q8 = 3 marks)
+ Skim read all 8 questions
    + Prioritise highest mark questions
    + **BUT** also do the ones you know first

:::

::::


## General Tips for Approaching Exam

+ Split your time according to the marks for the sections and questions  
  
+ Make an attempt to answer all questions
    + Section A: Guess if you need to - don't leave an MCQ answer blank 
    + Section B: A brief answer is better than no answer  
     
+ Section B: Read the questions very carefully. For the longer questions, the question will include information on the things we expect your answers to comment on. Questions in Section B have the number of marks indicated next to each question and sub-question  
    + As a **(very rough)** rule of thumb, a mark will typically correspond to a sentence (so for a 5-mark question a 5-6 sentence answer would likely be appropriate)  
    
+ You will not necessarily need to directly use all the equations on the equation sheet, but they may help you answer questions


## Example Questions

+ Example questions provided as this weeks lab exercises 

+ Sample solutions shared on Friday


# Questions?





