---
title: "<b>Semester 2 Introduction</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```

# What we have covered so far...

---
# Aims for semester 2
1. Discuss analysis of experimental designs using linear model.
  + Models with multiple nominal categorical variables.
  + Overall effects as well as comparisons between levels.
  + Categorical interactions

--

2. Recap assumption and diagnostics

--

3. Discuss ways to conduct inferential tests with violated assumptions (bootstrapping)

--

4. Extend to linear models to include other types of dependent variale (binary, categorical)

--

5. Discuss a number of important modelling and inference topics
  + Missing data
  + Multiple comparisons
  + Power analysis

--

6. Draw all of this together and emphasize the skills we have been developing in the context of reproducible analysis and science.

---
# Key concepts from semester 1
1. Basic structure of the linear model:

$$
y_i = b_0 + b_1x_1 + b_2x_2 ...b_kx_k + \epsilon_i
$$

+ And in R:

```{r, eval=FALSE}
lm(y ~ x1 + x2 + x3, data = tibble)
```

+ From which we have:
  + Individual tests of $\beta$ coefficients
  + $F$-tests of overall model
  + $R^2$ estimating variance accounted for

---
# Key concepts from semester 1
2. Assumptions

+ **L**inearity: The relationship between $y$ and $x$ is linear.

+ **I**ndependence of errors: The error terms should be independent from one another.

+ **N**ormality: The errors $\epsilon$ are normally distributed

+ **E**qual variances ("Homoscedasticity"): The scale of the variability of the errors  $\epsilon$ is constant at all values of $x$. 

---
# Key concepts from semester 1
3. Interactions

+ We have looked at interactions between two numeric/continuous variables, and between a categorical and numeric variable.

+ When the effect of one variable changes as a function of another.
  + We can explore these visually
  + Calculate simple slopes (the effect of $x$ at a specific value of $z$)


---
# Key concepts from semester 1
4. Model comparisons

+ When we have two nested models, we can compute an incremental $f$-test that tells us the improvement in the model for the inclusion of a (set) of variables. 

---
# Semester plan
+ Weeks 1 to 5
  + Recap experimental designs and the types of data they produce
  + Discuss analysis for different types of design, linking linear models to ANOVA.
  + Introduce categorical interactions
  + Discuss assumptions and multiple comparisons

--

+ Weeks 7 to 11
  + Bootstrapping
  + Binary logistic regression
  + Multinomial regression & generalized linear model
  + Missing data (conceptual)
  + Power Analysis
  + Reproducibility
  
---
# Weekly material
1. Lectures

--

2. Lecture exercises

--

3. Practical exercises

--

4. Reading

---
# Weekly support
+ "Live R" and Q&A

--

+ Lab drop-in (new format)

--

+ Office hours (new time for Tom)

--

+ Study groups (still live, not used in S1)

--

+ Discussion boards (new format)


---
# Assessments
+ Weekly quizzes (10 more in weeks 1 5, 7 to 11)

+ Report 2 
  + Second half of the semester.
  + Format near identical to semester 1

---
class: center, middle
# Let's get to it!