---
title: "<b>Week 3: Categorical Variables </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives
1. Understand the meaning of regression coefficients in the case of a categorical predictor.
2. Be able to state the assumptions underlying a linear regression model.
3. Understand how to assess if a fitted model satisfies the regression assumptions
4. Understand how to use transformations when the model violates the regression assumptions.

---
# Topics for today

---
#  Categorical variables 

+ Up until now we?ve focused on continuous IVs
	+ Continuous variables can take any value within a given range
		+ E.g., height in cm
+ Regression can also accommodate categorical IVs
	+ Categorical variables can only take discrete values
		+ E.g., animal type: 1= duck, 2= cow, 3= wasp
	+ They are mutually exclusive
		+ No duck-wasps or cow-ducks!

---
#  Categorical variables in R 

+ In R:
	+ Continuous variables should be of class ?numeric?
	+ Categorical variables (including binary variables) should be of class ?factor?
+ It is possible to convert a numeric variable to a factor:
+ Here each unique number become a level of a categorical variable


---
#  Categorical variables in R 

+ Similarly, character variables can be converted to factors
+ Here each unique character string becomes a level of a categorical variable


---
#  Regression with binary IVs 

+ Binary variables are categorical variables with two levels
+ Returning to our revision hours and test score example, suppose we now have a sample of 100 students who come from 2 classes
	+ We have our continuous outcome (Score, Y)
	+ We have a binary predictor (Class, X)
		+ We could code our binary predictor:
			+ Class A = 0
			+ Class B = 1
	+ We would have dataset that looks something like


---
#  Regression with binary IVs 

+ We could include the ? study.class ? variable in a simple regression, just like any other IV:
+ We use ? ? rather than to represent the fact that the variable codes ** category membership

---
#  Including a binary IV in a simple regression 


In R:
+ The regression coefficient for study class is *the difference between the mean score for class A and class B*
	+ As we coded class A=0 and class B=1, class A serves as the baseline
	+ If we had coded them the other way round the coefficient would have the same magnitude but opposite sign
+ The intercept is the *mean test score for the group coded 0*


---
#  Equation for the two levels of binary predictors 

+ We can verify this by writing the regression equation in terms of separate equations for our two groups:
+ For class A, = 0, giving us:
+ That is, the expected score for a person in class A is equal to the intercept


---
#  Equations for the two levels of a binary predictor 

+ For class B, =1, giving us:
+ The expected score for a person in class B is the expected score for a person in class A plus the regression coefficient .
+ The difference between the expected scores in classes A and B is :


---
#  Overall model evaluation for categorical IVs 

+ R 2 and F-ratio interpretation are identical to their interpretation in models with only continuous IVs

---
#  Statistical significance of categorical IVs 

+ We assess the statistical significance of regression coefficients for categorical IVs in exactly the same way as we did for continuous IVs:
+ We use the standard error of the coefficient to construct:
	+ *t-* tests and associated *p-* values  for the coefficient
	+ Confidence intervals around the coefficient

---
#  Statistical significance of categorical IVs 

+ Here we can say that there was a mean difference in test scores between class A and class B of 1.67 points on the test and this difference was statistically significant ( t **=4.22,** ***p*** **<.001** )

---
# Summary of today

---
# Next tasks