---
title: 'dapR2 Live : Semester 2 week 5'
author: "dapR2 Team"
date: "14/02/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(emmeans)
library(kableExtra)
```

# The study
## Background
The data used for the current report are simulated, drawing on recent work on sleep deprivation and placekeeping. For example, see: 

Stepan, M. E., Altmann, E. M., & Fenn, K. M. (2020). Effects of total sleep deprivation on procedural placekeeping: More than just lapses of attention. Journal of Experimental Psychology: General, 149(4), 800.

**Note you are not expected to provide an introduction, you do not have to read this article.**

Sleep deprivation has been shown to impact many cognitive processes. One area of cognitive function is placekeeping. Placekeeping concerns one's ability to follow through a set of actions in order without skipping steps or actions. 

## Procedure
Prior to the study participants had a minimum of 7 hours sleep, and on the day before and day of the study, did not consume caffeine, alcohol or drugs. Participants arrived at 9pm. On arrival participants were randomly assigned into the sleep deprived, or not deprived conditions, before being shown to the room in which they were to complete all questionnaire and test measures. In this testing block, participants completed a set of demographic questionnaire items, were interviewed about their typical sleep patterns, and completed a short cognitive test battery to provide a measure of general cognitive ability. Information from the participants' interviews were coded into low, medium and high quality typical sleep groups.

The total time for all assessments was approximately 2 hours. 

After all measures were completed, participants in the non-sleep deprived condition were sent home and asked to return at 9am the next day for the second testing session. The participants in the sleep deprived condition stayed overnight and were kept awake by members of the research team. 

At 9am the next day, all participants sat a second batch of testing. At this session, the outcome of interest, placekeeping, was assessed using the UNRAVEL task. 

UNRAVEL is an acronym. Each letter defines a task step to be completed. When working through the task, participants need to keep track of where they are in the sequence and perform the next step. When they reach L, then loop back to U and begin again. The stimulus screen contains no prompts as to where in the sequence a participant is (see paper above for example stimuli). Periodically participants are distracted by the presentation of a text string participants need to type out. You can try out the UNRAVEL task yourself at [https://www.millisecond.com/download/library/v6/unravel/unravel.web](https://www.millisecond.com/download/library/v6/unravel/unravel.web). The key dependent variable for the current analysis is the total number of errors made.  

## Research Questions {#research-questions}
The researchers want to know whether there is an effect of sleep deprivation on UNRAVEL performance, and whether this effect is greater for those who have high quality typical sleep. They also want to make sure that any effects identified are not due to higher cognitive ability, age and sex. 

They are specifically interested in the hypothesis that the impact of sleep deprivation is different for people who report "high" quality sleep (as opposed to those reporting either "low" or "medium").

# Data
```{r, message=FALSE, warning=FALSE}
df<-read_csv("liveR_s2w5.csv")
```

```{r eval=TRUE, echo=FALSE}
dict <- tibble(
  Variable = c("ID", "unravel", "sleep_dep", "sleep_qual", "cog", "sex", "age"),
  Description = c("Unique participant identifier",
                  "Total errors on UNRAVEL task (min = 0, max = 66)", 
                  "Experimental Condition: 0 = non-deprivation; 1 = deprivation",
                  "Sleep quality grouping: 0 = low; 1 = medium; 2 = high",
                  "z-score combining cognitive tests scores assessing cognitive ability",
                  "sex; 0 = male; 1 = female",
                  "Age of participant (in years)")
)
knitr::kable(dict, "html") %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```

Do some checks:

```{r}
str(df)
```

Changes:

```{r}
df <- df %>% 
  mutate(
  sleep_dep = factor(sleep_dep, labels=c("non-dep","dep")),
  sleep_qual = factor(sleep_qual),
  sex = factor(sex, labels=c("male","female"))
)
```

Run model:

```{r}
m1 <- lm(unravel ~ sleep_dep*sleep_qual, data = df)
```

Review the model:

```{r}
anova(m1)
```
```{r}
summary(m1)
```


Plot interaction:

```{r}
m1_means <- emmeans(m1, ~sleep_dep*sleep_qual)
emmip(m1_means, sleep_qual~sleep_dep)
```

Test specific hypothesis: *"specifically interested in the hypothesis that the impact of sleep deprivation is different for people who report "high" quality sleep (as opposed to those reporting either "low" or "medium")."*

```{r}
dep_coef  <- c('dep' = 1, 'non-dep' = -1)
qual_coef  <- c('low' = -.5, 'med' = -.5, 'high' = 1)
qual_coef %o% dep_coef

comp_res <- contrast(m1_means, 
                     method = list('Research Hyp' = c(-1.0, 1, 0.5, -0.5, 0.5, -0.5)))
comp_res
```

```{r}
m1_means
```


```{r}
confint(comp_res)
```

Assumption checks:

```{r}
plot(m1, which = 1)
```

```{r}
plot(m1, which = 2)
```

```{r}
hist(m1$residuals)
```


```{r}
plot(m1, which = 3)
```

```{r}
plot(m1, which = 5)
```

# covariates
```{r}
m2 <- lm(unravel ~ age + cog + sex + sleep_dep*sleep_qual, data = df)
```

```{r}
summary(m2)
```

```{r}
m2_means <- emmeans(m2, ~sleep_dep*sleep_qual)
comp_res2 <- contrast(m2_means, 
                     method = list('Research Hyp' = c(-1.0, 1, 0.5, -0.5, 0.5, -0.5)))

comp_res2
```
```{r}
comp_res
```

