---
title: "<b>Week 2: Calculating coefficients</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives
1. Be able to specify a simple linear model. 
2. Understand what fitted values and residuals are. 
3. Be able to interpret the coefficients of a fitted model.
4. Be able to test hypotheses and construct confidence intervals for the regression coefficients.

---
# Topics for today
+ Now we know how to position our model line, we can consider:
  + Interpreting the values
  + Testing the significance
  
---
# Recap model
+ overall
+ slope
+ intercept

---
# Interpretation
+ Slope
+ Intercept

---
#  Significance of individual effects 

+ A general way to ask this question would be to state: *?Does our model capture an informative relationship between X and Y??*
+ In the context of our example from last lecture, we could ask, *?Is study time a useful predictor of test score??*
+ The above is a research question/hypothesis. As we have been doing throughout, we need to turn this into a testable statistical hypothesis.

---
#  Evaluating individual IVs 

+ As well as evaluating overall model fit and significance, we can evaluate individual IVs
+ Last lecture we discussed the interpretation of regression slopes
+ We also need to look at:
	+ The sampling variability of (its standard error)
	+ The statistical significance of 


---
#  Standard error of?<U+0001D44F>1 

#  ? 

+ The standard error (SE) provides a measure of how different values would be across repeated samples
+ Smaller SEs suggest is estimated more precisely
	+ Smaller SEs are better!


---
#  Formula for the standard error of?<U+0001D44F>1 

#  ? 

+ The formula for the standard error of is:
+ Where:
	+ SSE is the sum of squared error (i.e., )
	+ N is the sample size
	+ k is the number of predictors (=1 for simple regression)
+ From this formula, we can see that the SEs for regression slopes will be smaller when:
	+ Residual variance ( is smaller
	+ Sample size, N is larger


---
#  Significance of ?<U+0001D44F>1 

#  ? 

+ The SE is important in significance tests for
+ We can think of testing the significance of as follows:
	+ If an IV yields no information about the DV then
	+ Why?
		+ gives the predicted change in Y for a unit change in X
		+ If Y does not change as X changes, then = 0
+ We can state this formally as a null and alternative hypothesis:

---
#  Significance of ?<U+0001D44F>1 

#  ? 

+ For , the appropriate test statistic is a t-statistic:
+ In regression, the degrees of freedom for *t* is N-k-1
	+ In simple regression with only one predictor, this gives us N-2 df
+ We compare our t-value against a t-distribution with N-2 df to assess statistical significance


---
#  Significance of ?<U+0001D44F>1 

#  ? 

+ Applying this to our exam revision and test score example from last week where we had SSE=21.26 , 10 participants , = 20.625 and =1.055 gives us:

So;
+ We compare our t-value against a t-distribution with df=8 to assess statistical significance, giving us *p=* 0.019
+ We can conclude that *revision is a statistically significant predictor of test scores at alpha=0.05*


---
#  Confidence intervals for ?<U+0001D44F>1 

#  ? 

+ We can also compute confidence intervals for
+ The 100(1-alpha), e.g., 95%, confidence interval for the slope is:
+ E.g., the 95% confidence interval for in our revision and test score example would be:

The confidence interval of 0.229 to 1.881 does not include zero, therefore, we can conclude that *revision is a statistically significant predictor of test scores* ( *p<.* 05).


---
#  Significance of individual effects 

+ Conceptually:
	+ If x yields no information on y, then
	+ Why?
		+ Beta gives the predicted change in y for a unit change in x.
		+ If y does not change as x changes, then beta = 0.
+ We can state this formally as a null and alternative:

---
#  Significance of individual effects 

+ So we have a null, what we need to do now is construct a test statistic and the associated sampling distribution.
+ For beta?s, we use a t-statistic (again!):

where

And


---
#  Significance of individual effects 

+ This can be evaluated against a t-distribution with n-2 degree for a model with one predictor.
+ More generally this is written as N-K-1
	+ Where K is the number of predictors, and the additional -1 catches the intercept.


---
#  Significance of individual effects 

+ For our example:

So

And?

---
#  Significance of individual effects 

+ For our example:

---
#  A decision about the null 

+ So we have a t-value associated with our beta coefficient.
	+ t = 3.656
+ And we know we will evaluate it against a t-distribution with df = n-2 = 3.
+ As with all tests we need to set our alpha.
	+ Let?s take 0.05 two tailed.
+ This gives us a critical value of 3.11.
	+ What is our decision?

---
#  Decision about the null 

+ For our example:


---
# Important notation
+ Model

---
# Summary of today

---
# Next tasks