---
title: "<b>Week 1: Functions & Models</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```



# Weeks Learning Objectives
1. Reviw the main concepts from dapR1.
2. Understand the concept of a function.
3. Be able to discuss what models are.
4. Understand the link between models and functions.

---
# Topics for today
+ Functions and models
  + What is a function?
  + Linear and non-linear functions
  + What is a model?
  + How do functions and models relate
  
---
# What is a function?
+ A function maps one set of values to some other set of values.
+ Put differently, a function takes an input, does something, and provides an output.

$$
x = \begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$


$$
f(x) = x-2
$$


$$
f(x) = \sqrt{x}
$$

---
# Visualising Simple Functions
+ An important tool in understanding functions is to plot them.
  + This helps us both understand plots
  + And gain intuition about functions.
+ So let's look at the following:

$$
f(x) = 10 + 2x
$$

---
# Visualising Simple Functions
+ Where $x$ is:

$$
x = \begin{bmatrix}
1 \\ 
2 \\ 
3 \\
4 \\
5 \\ 
6 \\ 
7 \\
8 \\
\end{bmatrix}
$$

--- 
# Visualising Simple Functions

```{r, echo=FALSE}
func_x <- tibble(
  x = c(1,2,3,4,5,6,7,8), 
  fx = 10 + (2*x) 
)
  
kable(func_x) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)


```

---
# Visualising Simple Functions

```{r, echo=FALSE}
ggplot(func_x[1,], aes(x, fx)) + 
  xlim(0, 10) +
  ylab("f(x)") +
  ylim(0, 30) 
```

---
# Visualising Simple Functions

```{r, echo=FALSE}
ggplot(func_x[1,], aes(x, fx)) + 
  xlim(0, 10) +
  ylab("f(x)") +
  ylim(0, 30) +
  geom_point(colour = "red", size = 2) +
  geom_segment(aes(x = x, y = fx, xend = x, yend = 0), 
               arrow=arrow(type = "closed", length = unit(0.25, "cm"))) +
  geom_segment(aes(x = x, y = fx, xend = 0, yend = fx), 
               arrow=arrow(type = "closed", length = unit(0.25, "cm")))
```

---
# Visualising Simple Functions

```{r, echo=FALSE}
ggplot(func_x, aes(x, fx)) + 
  xlim(0, 10) +
  ylab("f(x)") +
  ylim(0, 30) +
  geom_point(colour = "red", size = 2)
```

```{r, echo=FALSE}
fun1 <- function(x) 10+(2*x)

ggplot(func_x, aes(x, fx)) + 
  geom_point(colour = "red", size = 2) +
  stat_function(fun = fun1) +
  xlim(0, 10) +
  ylab("f(x)") +
  ylim(0, 30)
```

---
# Multiple arguments
+ Functions can take multiple arguments. Consider:

$$
f(x,y) = 10 + (x*y)
$$
+ Where:

$$
x = \begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$

$$
y = \begin{bmatrix}
1 \\ 
2 \\ 
3 \\ 
\end{bmatrix}
$$

--- 
# Multiple arguments

```{r,echo=FALSE}
func_xy <- tibble(
  x = c(1,1,1,2,2,2,3,3,3),
  y = c(1,2,3,1,2,3,1,2,3)
)

kable(func_xy) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

--- 
# Multiple arguments
```{r,echo=FALSE}
func_xy2 <- tibble(
  x = c(1,1,1,2,2,2,3,3,3),
  y = c(1,2,3,1,2,3,1,2,3),
  "f(x,y)" = 10 + (x*y)
)

kable(func_xy2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

---
# Why are functions important?
+ There are going to be lots of examples of functions in action.
+ Two primary examples are:
    + **Data transformations**
    + **Describing formal models**

---
# What is a model?
+ Pretty much all statistics is about models.
+ A model is a formal representation of a system.
+ Put another way, a model is an idea about the way the world is.

---
# A model as a function
+ We tend to represent mathematical models as functions.
  + which can be very helpful.
+ It allows for the precise specification about what is important (arguments) and what those things do (operations)
  + This leads to predictions
  + And these predictions can be tested.

---
# An Example
+ To think through these relations, we can use a simpler example.
+ Suppose I have a model for growth of babies.

$$
Length = 55 + 4*month
$$

---
# Visualizing a model

```{r, warning=FALSE, message=FALSE}
fun1 <- function(x) 55 + (4*x)
m1 <- ggplot(data = data.frame(x=0), aes(x=x)) +
  stat_function(fun = fun1) +
  xlim(0,24) +
  ylim(0,150) +
  ylab("Length (cm)") +
  xlab("Age (months)")
```

---
# Visualizing a model
```{r, warning=FALSE, message=FALSE}
m1
```

---
# Models as "a state of the world"
+ Let's suppose my model is true.
  + That is, it is a perfect representation of how babies grow.
+ What are the implications of this?


---
# Models and predictions
+ My models creates predictions.
+ **IF** my model is a true representation of the world, **THEN** data from the world should closely match my predictions.
  + Sometimes this is referred to as my model matching the *data generation process*


---
# Predictions and data

```{r, eval=FALSE}
tibble(
  Age = seq(10, 12, 0.25)
) %>%
  mutate(
    Prediction = 55 + (Age*4)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

---
# Predictions and data

```{r, echo=FALSE}
tibble(
  Age = seq(10, 12, 0.25)
) %>%
  mutate(
    Prediction = 55 + (Age*4)
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

+ Our predictions are points which fall on our line (representing the model, as a function)

---
# Predictions and data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
m1
```

+ Consider the predictions when the children get a lot older...

---
# Predictions and data

```{r, eval=FALSE}
tibble(
  Age = seq(216,300, 12)
) %>%
  mutate(
    Year = Age/12,
    Prediction = 55 + (Age*4),
    Predction_M = Prediction/100
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


---
# Predictions and data

```{r, echo=FALSE}
tibble(
  Age = seq(216,300, 12)
) %>%
  mutate(
    Year = Age/12,
    Prediction = 55 + (Age*4),
    Predction_M = Prediction/100
  ) %>%
  kable(.) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

+ What do you think this would mean for our actual data?
+ Will the data fall on the line?

---
# How good is my model?
+ How might we judge how good our model is?

  1. Model is represented as a function
  2. We see that as a line (or surface if we have more things to consider)
  3. That yields predictions (or values we expect if our model is true)
  4. We can collect data
  5. If the predictions do not match the data (points deviate from our line), that says something about our model.


---
# Models and Statistics
+ In statistics we (roughly) follow this process.
+ We define a model that represents one state of the world (probabilistically)
+ We then collect data to compare to it.
+ These comparisons lead us to make inferences about how the world actually is, by comparison to a world that we specify by our model. 

---
# Length & Age is non-linear

```{r, eval=FALSE, warning=FALSE, message=FALSE}
df <- read_csv("length_age.csv", col_names = c("Month", "Mean", "SD", "Lower", "Upper"))
m1 + 
  geom_line(data = df, aes(x=Month, y = Mean), colour = "red") + 
  geom_line(data = df, aes(x=Month, y = Lower), colour = "red", linetype = "dashed") +
  geom_line(data = df, aes(x=Month, y = Upper), colour = "red", linetype = "dashed")
```


---
# Length & Age is non-linear

```{r, echo=FALSE, warning=FALSE, message=FALSE}
df <- read_csv("length_age.csv", col_names = c("Month", "Mean", "SD", "Lower", "Upper"))
m1 + 
  geom_line(data = df, aes(x=Month, y = Mean), colour = "red") + 
  geom_line(data = df, aes(x=Month, y = Lower), colour = "red", linetype = "dashed") +
  geom_line(data = df, aes(x=Month, y = Upper), colour = "red", linetype = "dashed")
```

+ Our red line is plotted based on the mean length for different ages [real data](https://www.cdc.gov/growthcharts/who/boys_length_weight.htm)


---
# Summary of today

---
# Next tasks