---
title: "Class Feedback: Assessed Group Report"
author: "DAPR2 Team"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)
source('_theme/theme_quarto.R')
```


# Course Overview {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Introduction to Linear Models"
block1_lecs = c("Intro to Linear Regression",
                "Interpreting Linear Models",
                "Testing Individual Predictors",
                "Model Testing & Comparison",
                "Linear Model Analysis")
block2_name = "Analysing Experimental Studies"
block2_lecs = c("Categorical Predictors & Dummy Coding",
                "	Effects Coding & Coding Specific Contrasts",
                "Assumptions & Diagnostics",
                "Bootstrapping",
                "	Categorical Predictor Analysis")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Interactions"
block3_lecs = c("Interactions I",
                "Interactions II",
                "Interactions III",
                "Analysing Experiments",
                "Interaction Analysis")
block4_name = "Advanced Topics"
block4_lecs = c("Power Analysis",
                "Binary Logistic Regression I",
                "Binary Logistic Regression II",
                "Logistic Regression Analysis",
                "	Exam Prep and Course Q&A")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=10)
```

:::

::::


## Outline

+ Brief overview of correct models for RQ1 & RQ2    
+ Outline some of the most common mistakes across reports   
  + See 'Common Error' slides        
  + Non-exhaustive list - look at your individual feedback comments
  
## RQ1: Interaction Model

 **After accounting for differences in personality traits, do any of the differences in accuracy ratings between neuroscience conditions depend on level of education?**

```{r, eval = FALSE}
m1 <-lm(acc_rate ~ O + C + E + A + N + educ * neuro, data = dat)
```

## RQ2: Contrast Analysis {.smaller}

**Based on the above model, we also wanted to know: Do differences in accuracy ratings between the visual and text-based neuroscience conditions differ between people who have a degree (either UG or PG) and people without a degree?**

```{r, eval = FALSE}
# compare those with no degree to degree
educ_contrast  <- c('ND' = -1, 'PG' = 0.5, 'UG' = 0.5)

#compare text and visual neuroscience conditions
neuro_contrast <- c('None' = 0, 'Desc' = -1, 'Pic' = 1)

#combine individual weights into the interaction
contr = educ_contrast %o% neuro_contrast
contr

# call emmeans to check ordering of coefficients 
emm <- emmeans(m1, ~educ * neuro)
emm

# insert the coefficients following the order specified by the rows of emm above
comp <- contrast(emm, method = list('Hyp' = c(0, 0, 0, 1, -0.5, -0.5, -1, 0.5, 0.5)))
summary(comp, infer=T)
```


## Critical Errors: Specification

+ No covariates included in model 

+ No interaction term included in model

+ No contrast analysis

+ Including analyses not appropriate to address RQs

## Critical Errors: Interpretation

+ Not interpreting coefficients in context of RQs

+ Drawing conclusions that did not follow from results presented


## General Common Errors Doc

+ For more details on errors seen across reports, see 'Assessed Report: Common Error Slides'


# Questions?





