---
title: "<b> Bootstrapping Linear Models </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(ggmosaic)
library(kableExtra)
library(performance)
```

# Weeks Learning Objectives
1. Identify and provide examples of binary psychological outcomes.
2. Understand why a standard LM is not appropriate for binary data.
3. Fit and interpret a logistic model
4. Run and intepret assumption checks for logistic lm.

---
# Topics for this week

1. Bootstrapping theory (recap)
2. Confidence intervals (recap)
3. Why this is useful for linear models?
4. Applying bootstrap inference to linear models


---
# Inference with assumption violations
- Note that for us to make inferences about our statistics, we need a known sampling distribution under the null.
  - If we have this, we can use our normal tools of inference.

--

- But these sampling distributions are only accurate when model assumptions are met.

--

- If they are not, we are in a tricky position.
  - We can not trust our estimates or inferences from these models.

--

- So what can we do?

---
# Bootstrapping as an alternative
- We saw that we can compute a confidence interval using bootstrap methods.
  - And we know we can use confidence intervals to make inferences
  - Does the CI include 0?

- The key difference with the bootstrapping procedure is it does not make distributional assumptions.
  - The bootstrap distriubtion is drawn from our sample.
  - And the sample can have any distribution it likes.

---
# Bootstrapping a linear model


---
# Summary of today


