---
title: "<b>Week 3: Standardized Coefficients</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Weeks Learning Objectives
1. Understand the calculation and interpretation of the coefficient of determination. 
2. Understand the calculation and interpretation of the F-test of model utility.
3. Understand how to standardized model coefficients and when this is appropriate to do.
4. Understand the relationship between the correlation coefficient and the regression slope.

---
# Topics for today
+ Standardization of coefficients

---
# Unstandardized Coefficient

+ make point b is unstandardized
+ interpretaion in units of variables

---
# Standardized units
+ why it might be useful

---
# Standardized coefficients

+ how to calculate

--- 
# Interpretation


---
# Relationship to r


---
# Coverting b to r

--- 
# Example in R

---
# Which should we use? 

---
# Summary of today

---
# Next tasks


---
#  Standardised regression coefficients 

+ Unstandardised regression coefficients are often more useful when the variables are on  meaningful scales
	+ E.g. X additional hours of exercise per week adds Y years of healthy life
+ Sometimes it?s useful to obtain standardised regression coefficients
	+ When the scales of variables are arbitrary
	+ When there is a desire to compare the effects of variables measured on different scales
+ For continuous IVs, transforming both the IV and DV to z-scores (mean=0, SD=1) prior to fitting the regression yields standardised betas.
+ Alternatively, unstandardised betas can be converted to standardised betas using:


---
#  Standardised regression coefficients 

+ For our revision and test score example, we can compute the SDs for IV and DV:
	+ SD for revision hours (X) = 1.513
	+ SD for test scores (Y) = 2.213
+ Using these to compute the standardised regression coefficient:
+ *For every SD unit increase in hours revised, test scores are expected to increase by* *0.72* *SD units*

---
#  Interpreting standardised regression coefficients  

+ R 2 , *F* and *t-test* remain the same for the standardised coefficients as for unstandardised coefficients
+ The intercept becomes zero when all slopes are standardised
+ The interpretation of the coefficients becomes the increase in Y in standard deviation units for every standard deviation increase in X
+ Standardised betas in simple regression are equal to the Pearson correlation between X and Y
+ Cautions
	+ Just because you can put regression coefficients on a common metric doesn?t mean they can be meaningfully compared.
	+ The SD is a poor measure of spread for skewed distributions, therefore, be cautious of their use with skewed variables

