---
title: "<b>Exam Prep</b>"
subtitle: "Data Analysis for Psychology in R 2"
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


# Outline

+ Process

+ Structure of exam

+ Things to take into exam

+ Exam "strategy"

+ Example Questions

---
# Process

+ You need to check your calendar [https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets](https://www.ed.ac.uk/timetabling-examinations/exams/exam-diets)

+ Arrive in plenty of time

+ There will be direction to room

+ When you go into the room, you will need your student ID on your desk

+ Invigilators will give you exam instructions


---
# Things YOU Need to Take to the Exam  

+ Student Card

+ Pens (blue or black ink)  

+ Pencils and rubber

+ Ruler
  + If you like, not really needed, but might be useful 
  
+ Calculator
  + See list on Learn of permitted calculators  
  + No advantage to scientific vs normal  
  + **Phone/Watch can not be your calculator**  

---
# Things WE Will Give You

**What you are given**

+ a specific sheet for your MCQ answers  

+ an answer book for all other questions  

+ the equation sheet

---

**FILLING IN FRONT COVERS**

.pull-left[

MCQ sheet:  

  - matriculation number (UUN but without the "s")  

Answer book:  

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_ms01z6h1&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_j0xl893i" width="400" height="285" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to fill in the cover of an exam script book?"></iframe>

]

.pull-right[

Attendance Form: 

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_uwsdz3vt&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_ft64ev29" width="400" height="285" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to complete an examination attendance form"></iframe>

]

---
# Things You Can Take Away

<center>
<br><br>
<h2>NOTHING!</h2>  

(Other than the things you bring in with you like pens, pencils, calculator)

<br><br>
<b>
When you finish, you leave all papers (questions, answers, equation sheet) in the exam hall.  
</b></center>


---
class: center, middle, animated, rotateInDownLeft

# Questions?


---
# Exam Structure

.pull-left[
**Section A - 30 MARKS**

+ 15 MCQs

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from

]
.pull-right[
**Section B - 70 MARKS**  

+ 10 Questions  

+ Marks range from 4 to 12 marks per question:  
    + 2 x 4 marks
    + 2 x 5 marks
    + 2 x 6 marks
    + 2 x 8 marks
    + 2 x 12 marks

+ On range of topics  
]

---
# Exam Strategy: Section A

.pull-left[

**Section A - 30 MARKS**

+ 15 MCQs  

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from
]

.pull-right[

**Suggested Approach**
+ Similar to quiz questions on Learn  
+ Spend 15-30 minutes on MCQs:
  + Approx. 1-2 minutes per MCQ
  + You can answer these quickly at the end
  + Do the ones you know, then move on
  + Keep track of how many you have not done, so you can go back to them at the end
  
]

---
# Exam Strategy: Section B


.pull-left[
**Section B - 70 MARKS**  

+ 10 Questions, where marks range from 4 to 12 marks per question:  
    + 2 x 4 marks
    + 2 x 5 marks
    + 2 x 6 marks
    + 2 x 8 marks
    + 2 x 12 marks

+ On range of topics, where some questions will involve:
    + Short calculations
    + Asking you to explain statistical concepts discussed in the course
    + Giving you a research design/question and task you with detailing an appropriate analysis plan
    + Giving you model output and ask you to provide interpretation  
]
.pull-right[

**Suggested Approach** 

+ 70% of marks in this section - spend approx 70% of time on it!
+ Spend more time on higher mark questions
    + Marks arranged from highest to lowest marks (i.e., Q1 = 12 marks, Q10 = 4 marks)
+ Skim read all 10 questions
    + Prioritise highest mark questions
    + **BUT** also do the ones you know first
]

---
# General Tips for Approaching Exam

+ Split your time according to the marks for the sections and questions
+ Make an attempt to answer all questions
    + Section A: Guess if you need to - don't leave an MCQ answer blank 
    + Section B: A brief answer is better than no answer
+ Section B: Read the questions very carefully. For the longer questions, the question will include information on the things we expect your answers to comment on. Questions in Section B have the number of marks indicated next to each question. As a **(very rough)** rule of thumb, a mark will typically correspond to a sentence (so for a 4-mark question a 4-5 sentence answer would likely be appropriate)
+ You will not necessarily need to directly use all the equations on the equation sheet, but they may help you answer questions

---
class: center, middle, animated, rotateInDownLeft

# Questions?

---
# Mock Questions

+ Mock questions to work on during lab

+ Live Q&A session via Collaborate to discuss approach to answering questions on DATE


---
class: center, inverse, middle, animated, rotateInDownLeft

# End
