---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

You have the following coefficients from a binary logistic model.

Intercept = 0.13
$\beta_1x_1$ = -0.54
$\beta_2x_2$ = 1.1

Using R, convert these values into odds ratios and interpret $\beta_1$ and $\beta_2$. What other information might you want to know before interpreting?

---
# Question 1: Answer


```{r}
exp(c(0.13, -0.54, 1.1))
```

- For a unit increase in $x_1$, there is a decrease in the odds of success of approximaely 42%.
- For a unit increase in $x_2$, the odds of success increase by a factor of 3.  

- What else might we want to know? Well it is useful to know the scale of the $x$ variables. We would also want to know the confidence interval to know if it includes 1.

---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 