---
title: "<b> Testing and Evaluating LM</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)

test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  hours = seq(0.5,5,.5),
  score = c(1,3,1,2,2,6,3,3,4,8)
)
```


# Week's Learning Objectives
1. Be able to test hypotheses and construct confidence intervals for the model coefficients.

2. Understand the calculation and interpretation of the coefficient of determination. 

3. Understand the calculation and interpretation of the F-test of model utility.


---
# Things to recap
+ We will look again at significance testing.

+ And also discuss sampling variability.

+ If at any point you feel like you are not following this section of the material, go back and review sampling and hypothesis testing material linked on LEARN.

---
# `lm` in R
+ We do not generally calculate by hand like this, so we also briefly introduced the `lm()` function.

```{r}
lm(score ~ hours, data = test)
```

---
# Evaluating our model
+ At this point, we have estimated values for the key parameters of our model (intercept and slope).

+ Now we have to think about how we evaluate the model.

+ There are three ways to think about evaluation:

  1. Evaluating the individual coefficients
  2. Evaluating the overall model quality
  3. Evaluating the model assumptions

+ Before accepting a set of results, it is important to consider all three of these aspects of evaluation.

???
Important to really emphasize this is a package of information and we want it all before we decide to accept our model.

---
#  Significance of individual effects 
+ A general way to ask this question would be to state: 

> **Is our model model informative about the relationship between X and Y?**

--

+ In the context of our example from last lecture, we could ask, 

> **Is study time a useful predictor of test score?**

--

+ The above is a research question/hypothesis. As we have done before, we need to turn this into a testable statistical hypothesis.

---
#  Evaluating individual predictors 
+ Steps in hypothesis testing:

--
  + Research questions
    
--
  
  + Statistical hypothesis
    
--
  
  + Define the null
    
--
  
  + Calculate an estimate of effect of interest.
  
--
  
  + Calculate an appropriate test statistic.
    
--
  
  + Evaluate the test statistic against the null.
    

---
# Research question and hypotheses

+ **Research questions** are statements of what we intend to study. 

+ A good question defines:

--

  + Constructs under study
  + the relationship being tested
  + A direction of relationship
  + target populations etc.

> **Does increased study time improve test scores in school age children?**

--

+ **Statistical hypotheses** are testable mathematical statements.

--

  + In typical testing in Psychology, we define have a **null ( $H_0$ )** and an **alternative ( $H_1$ )** hypothesis.
  + $H_0$ is precise, and states a specific value for the effect of interest.
  + $H_1$ is not specific, and simply says "something else other than the null is more likely"

???
Flag here that if these comments are completely alien to them, they should go back and recap the hypothesis testing material from dapR1-lectures 12 to 15 (20-21) or 12-14 (19-20).

---
# Defining null

+ Conceptually:
	+ If $x$ yields no information on $y$, then $\beta_1 = 0$
	
+ **Why would this be the case?**

--
	+ $\beta$ gives the predicted change in $y$ for a unit change in $x$.
	+ If $x$ and $y$ are unrelated, then a change in $x$ will not result in any change to the predicted value of $y$
	+ So for a unit change in $x$, there is no (=0) change in $y$.
	
+ We can state this formally as a null and alternative:

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$

???
+ For the null to be testable, we need to formally define it. 
+ Point out here the difference in the specificity of the hypotheses. $H_0$ is that the $b_1$ takes a specific value. $H_1$ is that $b_1$ has some value that is not this specific value. i..e one is directly testable, the other is not.


---
class: center, middle
# Time for a break

**Quiz time**

We are about to look at hypothesis tests for coefficients.

So our quiz is on the concept of the standard error


---
class: center, middle
# Welcome Back!

**Where we left off... **

We have defined a null, now let's look at constructing the test


---
# Point estimate and test statistic

+ We have already seen how we calculate $\hat \beta_1$.

+ The associated test statistic to for $\beta$ coefficients is a $t$-statistic

$$t = \frac{\hat \beta}{SE(\hat \beta)}$$

+ where

  + $\hat \beta$ = any beta coefficient we have calculated
  + $SE(\hat \beta)$ = standard error of $\beta$ 

--

+ **Recall** that the standard error describes the spread of the sampling distribution.
  + The standard error (SE) provides a measure of sampling variability
  + Smaller SE's suggest more precise estimate (=good)
  
???
+ brief reminders on test statistics
  + every quantity we wish to calculate a significance test for needs an test statistic.
  + the test statistic is a value that has a known sampling distribution
+ If sampling distribution is unfamiliar, again, recap the hypothesis testing material


---
#  SE( $\hat \beta_1$ )
+ The formula for the standard error of the slope is:

$$SE(\hat \beta_1) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_i - \bar{x})^2}}$$

+ Where:
	+ $SS_{Residual}$ is the residual sum of squares
	+ $n$ is the sample size
	+ $k$ is the number of predictors (= 1 for simple linear regression)

+ **Given the above, think about what things would make the SE smaller.**

--

+ From this formula, we can see that the SE's for $\beta$ will be smaller when:
	+ Residual variance ( $SS_{Residual}$ ) is smaller
	+ Sample size, $n$, is larger


---
# Back to the example

$$t = \frac{\hat \beta_1}{SE(\hat \beta_1)}$$

+ Let's calculate $t$ for our example.
+ We will use some values we have already calculated
  + $\hat \beta_1 = 1.055$ 

--

+ So we need $SE(\hat \beta_1)$

$$SE(\hat \beta_1) = \sqrt{\frac{SS_{Residual}/(n-k-1)}{\sum(x_i - \bar{x})^2}}$$

+ $n$ = 10
+ $k$ = 1
+ $\sum(x_i - \bar{x})^2$ = 20.625

---
# Back to the example
+ So all we have left is $SS_{Residual}$.

+ From earlier this week, we know:

$$SS_{Residual} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2$$

+ $SS_{Residual}$ = residual sum of squares = sum of the squared residuals

```{r}
res <- lm(score ~ hours, data = test)
SSRes = sum(res$residuals^2)
SSRes
```

---
# Back to our example

+ So pull all this together:

$$SE(\hat \beta_1) = \sqrt{\frac{SS_{Residual}/(n-k-1)}{\sum(x_i - \bar{x})^2}} 
= \sqrt{\frac{21.16364 / (10-1-1)}{20.625}} = \sqrt{\frac{2.65455}{20.625}} = 0.35814$$

--

+ and finally

$$t = \frac{\hat \beta_1}{SE(\hat \beta_1)} = \frac{1.055}{0.35814} = 2.945773 = 2.95$$


---
# Sampling distribution for the null

+ Now we have our $t$-statistic, we need to evaluate it.

+ For that, we need sampling distribution for the null.

+ For $\beta$, this is a $t$-distribution with $n-k-1$ degrees of freedom.
	+ Where $k$ is the number of predictors, and the additional -1 represents the intercept.

--

+ So for linear models with 1 predictor this is $n-2$
  + In our case = 8

---
#  A decision about the null 
+ So we have a $t$-value associated with our $\beta$ coefficient.
	+ t = 2.95

+ And we know we will evaluate it against a $t$-distribution with 8 df.

+ As with all tests we need to set our $\alpha$.
	+ Let's take 0.05 two tailed.

--

+ Now we need a critical value to compare our observed $t$-value to.

---
# Visualize the null

.pull-left[
```{r, echo=FALSE}
ggplot() + 
  xlim(-8, 8) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=8)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.025, 8), -8),
                alpha=.25,
                fill = "blue",
                args = list(df=8)) +
    stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.975, 8), 8),
                alpha=.25,
                fill = "blue",
                args = list(df=8)) +
  geom_vline(xintercept = 2.95, col="red") +
  xlab("\n t") +
  ylab("") +
  ggtitle("t-distribution (df=8); t-statistic (2.95; red line)")
```

]

.pull-right[

+ Critical value and $p$-value:
```{r}
tibble(
  LowerCrit = round(qt(0.025, 8), 3),
  UpperCrit = round(qt(0.975, 8), 3),
  Exactp = (1 - pt(2.95, 8)) * 2
)
```

]

???
+ discuss this plot.
+ remind them of 2-tailed
+ areas
+ % underneath each end
+ comment on how it would be different one tailed
+ remind about what X is, thus where the line is

---
# `lm` in R

.pull-left[
```{r}
summary(res)
```
]

.pull-right[
+ So in our example, we **reject the null**.
+ **Spend a little bit of time looking at this output, and comparing the various values to those things we calculated in our example.**
  + Some we haven't yet looked at. 
  + They are coming next.

]

---
#  Confidence intervals for $\beta_1$
+ We can also compute confidence intervals for $\hat \beta_1$
+ The $100 (1 - \alpha)$, e.g., 95%, confidence interval for the slope is:

$$\hat \beta_1 \pm t^* \times SE(\hat \beta_1)$$

+ So, 95% confidence interval for in our revision and test score example would be:

```{r}
tibble(
  LowerCI = round(1.055 - (2.306 * 0.358), 3),
  UpperCI = round(1.055 + (2.306 * 0.358), 3)
)
```



+ The confidence interval of 0.229 to 1.881 does not include zero, 
  + Therefore, we can conclude that **revision is a statistically significant predictor of test scores** ( $p < .05$).

---
# What changes when we have multiple predictors?
+ Broadly follows the same procedure as in simple regression:
	+ Standard errors (SEs) for each regression slope are computed
	
	+ SE gives a measure of the sampling variability of a regression coefficient
	
	+ $t$-tests and confidence intervals evaluate the statistical significance of regression slopes

---
#  Standard errors 

.pull-left[

$$SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_i - \bar{x})^2}}$$

]


.pull-right[

$$SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{ij}})^2(1-R_{xj}^2)}}$$

+ $1-R_{xj}^2$ is capturing the correlation between $x_j$ and all other $x$'s

]

---
#  Standard errors

$$SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{ij}})^2(1-R_{xj}^2)}}$$

+ Examining the above formula we can see that:
	+ $SE$ is smaller when residual variance ( $SS_{residual}$ ) is smaller
	+ $SE$ is smaller when sample size ( $N$ ) is larger
	+ $SE$ is larger when the number of predictors ( $k$ ) is larger
	+ $SE$ is larger when a predictor is strongly correlated with other predictors ( $R_{xj}^2$ )

???
+ Well return to this later when we discuss multi-collinearity issues

---
#  Significance of coefficients 

+ Once we have the standard error, all else is the same:

$$t = \frac{\hat \beta_1}{SE(\hat \beta_1)}$$

+ A $t$-test of the null hypothesis that $b_j = 0$

+ The $t$-value is compared to a $t$-distribution with N-k-1 degrees of freedom to assess statistical significance at a given $\alpha$.

---
#  Our academic performance example

```{r}
res$coefficients
```


**Self-control (t(647)=14.05, p<.001) was a significant predictor of academic performance ( $\alpha = 0.05$ ), and so we reject the null hypothesis of no effect. However, we failed to reject the null for class interaction (t(647)=-0.34, p>.05).**


---
#  Confidence intervals 

+ Like in simple regression, we can also compute confidence intervals for slopes in multiple regression.

+ The 100(1-alpha) confidence interval for the slope is:

$$\hat \beta_1 \pm t^* \times SE(\hat \beta_1)$$

```{r, echo=FALSE, message=FALSE}
data <- read_csv("./aggression.csv")
data <- data %>%
  transmute(
    z_perf = scale(aggression),
    z_SC = scale(self_control),
    z_interaction = scale(resting_HR)
    )
perf <- lm(z_perf ~ z_SC + z_interaction, #<<
          data = data)
res2 <- summary(perf)
```


```{r}
tibble(
  lower = res2$coefficients[2,1] - round(qt(0.975, 647), 3) *  res2$coefficients[2,2],
  upper = res2$coefficients[2,1] + round(qt(0.975, 647), 3) *  res2$coefficients[2,2]
)

```


---
#  Confidence intervals 

+ Or the much easier version...

```{r}
confint(perf)
```



---
class: center, middle
# Break

---
#  Quality of the overall model 
+ When we measure an outcome ( $y$ ) in some data, the scores will vary (we hope).
  + Variation in $y$ = total variation of interest.

--

+ The aim of our linear model is to build a model which describes $y$ as a function of $x$.
	+ That is we are trying to explain variation in $y$ using $x$.

--

+ But it won't explain it all.
  + What is left unexplained is called the residual variance.

--

+ So we can breakdown variation in our data based on sums of squares as;

$$SS_{Total} = SS_{Model} + SS_{Residual}$$

---
#  Coefficient of determination 
+ One way to consider how good our model is, would be to consider the proportion of total variance our model accounts for. 

$$R^2 = \frac{SS_{Model}}{SS_{Total}} = 1 - \frac{SS_{Residual}}{SS_{Total}}$$

+ $R^2$ = coefficient of determination

--

  + Quantifies the amount of variability in the outcome accounted for by the predictors.
  + More variance accounted for, the better.
  + Represents the extent to which the prediction of $y$ is improved when predictions are based on the linear relation between $x$ and $y$.

--

+ Let's see how it works.
  + To do so, we need to calculate the different sums of squares.


---
# Total Sum of Squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Total} = \sum_{i=1}^{n}(y_i - \bar{y})^2$$

+ Squared distance of each data point from the mean of $y$.

+ Mean is our baseline. 

+ Without any other information, our best guess at the value of $y$ for any person is the mean.

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Total}')) +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]

---
# Calculations

.pull-left[
```{r}
ss_tab <- test %>%
    mutate(
        y_dev = score - mean(score), 
        y_dev2 = y_dev^2
    )
```

```{r}
ss_tab %>%
    summarize(
        ss_tot = sum(y_dev2)
    )
```

]

.pull-right[

```{r, echo=FALSE}
kable(ss_tab) %>%
  kable_styling(., full_width = F)
```

]

---
# Residual sum of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Residual} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2$$

+ Which you may recognise.

+ Squared distance of each point from the predicted value.
]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Residual}')) +
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(res$fitted.values)), color = "red", lty =2)

```

]

---
# Calculations

.pull-left[
```{r}
ss_tab <- ss_tab %>%
  mutate(
    y_pred = round(res$fitted.values,2),
    pred_dev = round((score - y_pred),2),
    pred_dev2 = round(pred_dev^2,2)
  )
```

```{r}
ss_tab %>%
  summarize(
    ss_tot = sum(y_dev2),
    ss_resid = sum(pred_dev2) #<<
  )
```

]

.pull-right[

```{r, echo=FALSE}
kable(ss_tab[-c(4,5)]) %>%
  kable_styling(., full_width = F)
```

]


---
# Model sums of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Model} = \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2$$

+ That is, it is the deviance of the predicted scores from the mean of $y$.

+ But it is easier to simply take:

$$SS_{Model} = SS_{Total} - SS_{Residual}$$

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Model}')) +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, c(res$fitted.values), xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]


---
# Calculations

.pull-left[

$$SS_{Model} = SS_{Total} - SS_{Residual}$$

```{r}
ss_tab %>%
  summarize(
    ss_tot = sum(y_dev2),
    ss_resid = sum(pred_dev2)
  ) %>%
  mutate( #<<
    ss_mod = ss_tot - ss_resid #<<
  )
```

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle("SS Model") +
  geom_hline(aes(yintercept = mean(score)),color="blue", size=1) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", size=1) + 
  geom_segment(aes(x = hours, c(res$fitted.values), xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]


---
#  Coefficient of determination 
+ Now we can finally come back to $R^2$.

$$R^2 = 1 - \frac{SS_{Residual}}{SS_{Total}}$$

+ Or

$$R^2 = \frac{SS_{Model}}{SS_{Total}}$$

+ So in our example:

$$R^2 = \frac{SS_{Model}}{SS_{Total}} = \frac{22.9}{44.1} = 0.519$$

+ ** $R^2$ = 0.519 means that 52% of the variation in test scores is accounted for by hours of revision.**

---
#  Our example 

```{r}
res <- lm(score ~ hours, data = test)
summary(res)
```

???
As at the end of last session, we can check this against the R-output:
Be sure to flag small amounts of rounding difference from working through "by hand" and so presenting to less decimal places.

---
#  Adjusted $R^2$ 
+ We can also compute an adjusted $R^2$ when our lm has 2+ predictors.
  + $R^2$ is an inflated estimate of the corresponding population value

+ Due to random sampling fluctuation, even when $R^2 = 0$ in the population, it's value in the sample may $\neq 0$ 

+ In **smaller samples** , the fluctuations from zero will be larger on average

+ With **more IVs** , there are more opportunities to add to the positive fluctuation


$$\hat R^2 = 1 - (1 - R^2)\frac{N-1}{N-k-1}$$

+ Adjusted $R^2$ adjusts for both sample size ( $N$ ) and number of predictors ( $k$ )


---
#  In our academic performance example 

```{r}
res2
```

---
#  In our academic performance example 

+ **Based on adjusted R-squared, self-control and class interaction together explain `r round(res2$adj.r.squared,3)*100`% of the variance in academic performance**

+ As the sample size is large and the number of predictors small, unadjusted ( `r round(res2$r.squared,3)` ) and adjusted R-squared ( `r round(res2$adj.r.squared,3)` ) are similar.




---
class: center, middle
# Time for a break

**Quiz time!**


---
class: center, middle
# Welcome Back!

**Where we left off... **

We had just calculated $R^2$

Now let's look at calculating significance tests for our model


---
#  Significance of the overall model 
+ The test of the individual predictors (IVs, or $x$'s) does not tell us if the overall model is significant or not.
	+ Neither does R-square
	+ But both are indicative

+ To test the significance of the model as a whole, we conduct an $F$-test.

---
#  F-ratio
+ $F$-ratio tests the null hypothesis that all the regression slopes in a model are all zero
  + We are currently talking about a model with only one $x$, thus one slope.
  + But the $F$-ratio test will generalise.

--

+ $F$-ratio is a ratio of the explained to unexplained variance:

$$F = \frac{MS_{Model}}{MS_{Residual}}$$

+ Where MS = mean squares

--

+ **What are mean squares?**
  + Mean squares are sums of squares calculations divided by the associated degrees of freedom.
  + The degrees of freedom are defined by the number of "independent" values associated with the different calculations.

---
# F-table

```{r, echo=FALSE}
ftest <- tibble(
  SS = c("Model", "Residual", "Total"),
  df = c("k", "n-k-1", " "),
  MS = c("SS model/df model", "SS residual/df residual", " "),
  Fratio = c("MS model/ MS residual", " ", " "),
  pvalue = c("F(df model,df residual)", "", "")
)

kable(ftest)%>%
  kable_styling(., full_width = F)
```

---
# Our example: F-table

```{r, echo=FALSE}
tibble(
  Component = c("Model", "Residual", "Total"),
  SS = c(22.9,21.2,44.1),
  df = c(1, 8, " "),
  MS = c(22.9, 2.65, " "),
  Fratio = c(8.641509, " ", " "),
  pvalue = c("F(1,8)", "", "")
) %>%
  kable(.)%>%
  kable_styling(., full_width = F)
```


---
# F-ratio
+ Bigger $F$-ratios indicate better models.
  + It means the model variance is big compared to the residual variance.

--

+ The null hypothesis for the model says that the best guess of any individuals $y$ value is the mean of $y$ plus error.
	+ Or, that the $x$ variables carry no information collectively about $y$.

--

+ $F$-ratio will be close to 1 when the null hypothesis is true
  + If there is equivalent residual to model variation, $F$=1
	+ If there is more model than residual $F$ > 1

--

+ $F$-ratio is then evaluated against an $F$-distribution with $df_{Model}$ and $df_{Residual}$ and a pre-defined $\alpha$

--

+ Testing the $F$-ratio evaluates statistical significance of the overall model

---
# Visualize the test

.pull-left[

```{r, echo=FALSE}
ggplot() +
  xlim(-1,10) + 
  stat_function(fun=df,
                geom = "line",
                args = list(df1=1, df2=8)) +
  stat_function(fun = df, 
                geom = "area",
                xlim = c(qf(0.95, 1,8), 10),
                alpha=.25,
                fill = "blue",
                args = list(df1=1, df2=8)) +
  geom_vline(xintercept = 8.641509, col="red") +
  xlab("\n F") +
  ylab("") +
  ggtitle("F-distribution (1,8); f-ratio (8.64; red line)")
```
]

.pull-right[

+ Critical value and $p$-value:
```{r}
tibble(
  Crit = round(qf(0.95, 1, 8),3),
  Exactp = 1-pf(8.64, 1, 8)
)
```

+ From this we would **reject the null**. 

]

---
#  Our example 

```{r}
res <- lm(score ~ hours, data = test)
summary(res)
```

???
As at the end of last session, we can check this against the R-output:
Comment on the minor differences for rounding.

---
#  $F$-ratio 
+ Like in simple regression, the $F$-ratio is used to test the null hypothesis that **all** model slopes are zero.

+ It is calculated in exactly the same way as in simple linear model:

$$F = \frac{MS_{Model}}{MS_{Residual}} = \frac{\frac{SS_{model}}{df_{model}}}{\frac{SS_{residual}}{df_{residual}}}$$

+ Where
	+ df model = $k$
	+ df residual = $N$ - $k$ - 1
		  + $N$ = sample size
		  + $k$ = number of predictors

---
#  In our academic performance example? 

```{r}
res2
```

---
#  In our academic performance example? 

+ Our overall model was significant (*F(2,647)=99.32, p<.001*). 




---
# Summary of today



---
class: center, middle
# Thanks for listening!