---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)

dat <- read_csv("./salary2.csv")
mod <- lm(salary ~ perf, data = dat)
```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

.pull-left[
Which assumption is the plot of the right useful for assessing?

a. Linearity

b. Independence of errors

c. Normality of errors

d. Homoscedasticity

]


.pull-right[

```{r, echo=FALSE}

plot(mod, which = 2)

```


]

---
# Question 1: Answer

.pull-left[
Which assumption is the plot of the right useful for assessing?


c. Normality of errors


This is a qq-plot

]


.pull-right[

```{r, echo=FALSE}

plot(mod, which = 2)

```


]



---
# Question 2

.pull-left[
Which assumption is the plot of the right useful for assessing?

a. Linearity

b. Independence of errors

c. Normality of errors

d. Homoscedasticity

]


.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(dat, aes(x=perf, y=salary)) +
  geom_point()+
  geom_smooth(method = "lm", se = F) +
  geom_smooth(method = "loess", se=F, col = "green")

```


]

---
# Question 2: Answer

.pull-left[
Which assumption is the plot of the right useful for assessing?


a. linearity

This is a scatterplot with a linear and loess line included.
]


.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(dat, aes(x=perf, y=salary)) +
  geom_point()+
  geom_smooth(method = "lm", se = F) +
  geom_smooth(method = "loess", se=F, col = "green")

```


]



---
# Question 3

.pull-left[
Which assumption is the plot of the right useful for assessing?

a. Linearity

b. Independence of errors

c. Normality of errors

d. Homoscedasticity

]


.pull-right[

```{r, echo=FALSE}

hist(mod$residuals)

```


]

---
# Question 3: Answer

.pull-left[
Which assumption is the plot of the right useful for assessing?


c. Normality of errors

Axis and title were useful here! This is a histogram of model residuals (errors)


]


.pull-right[

```{r, echo=FALSE}

hist(mod$residuals)

```


]


---
# Question 4

.pull-left[
Which assumption is the plot of the right useful for assessing?

a. Linearity

b. Independence of errors

c. Normality of errors

d. Homoscedasticity

]


.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(car)
residualPlot(mod)

```


]

---
# Question 3: Answer

.pull-left[
Which assumption is the plot of the right useful for assessing?


d. Homoscedasticity

This is a residual-vs-predicted values plot


]


.pull-right[

```{r, echo=FALSE}

residualPlot(mod)

```


]


---
# Extra task
+ Go to the weekly discussion boards and discuss whether you think any of the plots above suggest an assumption is violated.


---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 