---
title: 'dapR2: Week 2 Quiz'
author: "Tom Booth"
date: "14/09/2020"
output: html_document
---

```{r, echo=FALSE, warning=FALSE}
library(MASS)
set.seed(7284)

mu <- c(0,0)
Sigma <- matrix(c(1.0, 0.6,
                  0.6, 1.0), byrow = T, ncol = 2)

rawvars <- mvrnorm(n=10000, mu=mu, Sigma=Sigma)

df <- rawvars[sample(nrow(rawvars), 150),]


Sigma2 <- matrix(c(1.0, -0.4,
                  -0.4, 1.0), byrow = T, ncol = 2)

rawvars2 <- mvrnorm(n=10000, mu=mu, Sigma=Sigma2)

df2 <- rawvars2[sample(nrow(rawvars2), 150),]


Sigma3 <- matrix(c(1.0, 0,
                  0, 1.0), byrow = T, ncol = 2)

rawvars3 <- mvrnorm(n=10000, mu=mu, Sigma=Sigma3)

df3 <- rawvars3[sample(nrow(rawvars3), 150),]


res <- data.frame(df, df2, df3)
colnames(res) <- c("x1", "y1","x2", "y2","x3", "y3")
```




# Lecture 2.1: Best line quiz

## Question 1
Which colour line is the best fit to the data (points)?

a. Black
b. Red
c. Blue
d. Green


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
ggplot(res, aes(x = x1, y = y1)) +
  geom_point() +
  geom_abline(slope = -0.64, intercept = -0.11) +
  geom_abline(slope = 0.4, intercept = 1.5, col = "red") +
  geom_abline(slope = 0.64, intercept = -0.11, col = "blue") + #correct
  geom_abline(slope = 0.3, intercept = 0.1, col = "green") -> p1

ggsave(filename = "../docs/quiz_images/bestfit1.png",plot = p1)

p1
```


## Question 2
Which colour line is the best fit to the data (points)?

a. Black
b. Red
c. Blue
d. Green


```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(res, aes(x = x2, y = y2)) +
  geom_point() +
  geom_abline(slope = -0.5, intercept = 1.5) +
  geom_abline(slope = -0.31, intercept = -0.07, col = "red") + #correct
  geom_abline(slope = -.1, intercept = 1, col = "blue") +
  geom_abline(slope = 0.3, intercept = 0.1, col = "green") -> p2

ggsave(filename = "../docs/quiz_images/bestfit2.png",plot = p2)

p2
```



## Question 3

Which colour line is the best fit to the data (points)?

a. Black
b. Red
c. Blue
d. Green


```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(res, aes(x = x3, y = y3)) +
  geom_point() +
  geom_abline(slope = 0.7, intercept = .1) +
  geom_abline(slope = 0.08, intercept = -2, col = "red") +
  geom_abline(slope = 0.08, intercept = 3, col = "blue") +
  geom_abline(slope = 0.08, intercept = -0.08, col = "green") -> p3 #correct 

ggsave(filename = "../docs/quiz_images/bestfit3.png",plot = p3)

p3
```


# Lecture 2.4: Interpretation of coefficients

## Question 1
I conduct a study where I measure the time it takes (in minutes) for a participant to complete a task, where the score for the task is the number of errors they make. My outcome measure is errors, and my predictor is time. I get the following results:

```{r, echo = FALSE}
colnames(res) <- c("x1", "y1","time", "error","x3", "y3")
q1 <- lm(error ~ time, res)
summary(q1)
```

Write an interpretation of both the intercept and the coefficient for time.


## Question 2
A second set of researchers runs the same study, and conducts the same analysis. There results are shown below:

```{r, echo = FALSE}
colnames(res) <- c("time", "error","x2", "y2","x3", "y3")
q2 <- lm(error ~ time, res)
summary(q2)
```

Write an interpretation of both the intercept and the coefficient for time.

How do the results from question 1 and question 2 differ? Which results make most sense given the variables measured?


# Lecture 2.5: Null distributions and SE

## Just one question....

The standard error is what?

a. variance of the normal distribution
b. standard deviation of the predictor
c. standard deviation of the sampling distribution
d. variance of the sampling distribution

If this questions seems like jibberish, before continuing, go back to week 1 materials and look again at the recap material on sampling and hypothesis testing. Note down any questions, and come to office hours to chat them over.
