---
title: "Class Feedback: Assessed Group Report Common Errors"
author: "DAPR2 Team"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)
source('_theme/theme_quarto.R')
```


# Course Overview {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block1_name = "Introduction to Linear Models"
block1_lecs = c("Intro to Linear Regression",
                "Interpreting Linear Models",
                "Testing Individual Predictors",
                "Model Testing & Comparison",
                "Linear Model Analysis")
block2_name = "Analysing Experimental Studies"
block2_lecs = c("Categorical Predictors & Dummy Coding",
                "	Effects Coding & Coding Specific Contrasts",
                "Assumptions & Diagnostics",
                "Bootstrapping",
                "	Categorical Predictor Analysis")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=0)
```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| warning: false
#| results: asis

block3_name = "Interactions"
block3_lecs = c("Interactions I",
                "Interactions II",
                "Interactions III",
                "Analysing Experiments",
                "Interaction Analysis")
block4_name = "Advanced Topics"
block4_lecs = c("Power Analysis",
                "Binary Logistic Regression I",
                "Binary Logistic Regression II",
                "Logistic Regression Analysis",
                "	Exam Prep and Course Q&A")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=10)
```

:::

::::

## Covariates: Inclusion Based on Significance {.smaller}

Selecting only a small number of the covariates (i.e., openness) based on their relation to the DV

+ Lacked rationale given RQs. Why?
  + If we have covariates we want to control for, we want them in a model whether their effect is significant or not. This is because although the effects may not be large enough to meet our $\alpha$ threshold, it is unlikely their effect is 0 
  + By wrongfully excluding some covariates, our resulting estimates of education & neuroscience condition may differ in part be due to differences in those covariates

+ When would this have been appropriate?
  + If a covariate had very high multicollinearity
  + If we believed that there were many model influential cases based on responses to the covariate
  + If we had a small sample size and lots of covariates to control for  
    + Very few groups provided any sort of rationale based on the above for only including a subset of covariates, most focused on significance


## Covariates: Model Comparison with and without Covariates

+ Fitting a model without the covariates, then fitting a model with covariates, and comparing these based on $F$-tests  
+ This was a redundant analysis - we need to control for covariates in all models, not interested in whether their inclusion improved model fit

## Contrast: Not Specified
    
    
**Combining levels**    
    
+ No contrast given for RQ2, instead a second interaction model used where a new binary variable was created where observation condition now contained only 2 levels - 'no degree' and 'degree'        
  + Not optimal or efficient        
  + Question specifically asked "**Based on the above model**, we also wanted to know" 
    
        
**Simple effects / pairwise tests**       
    
+ The particular combination of averaging across levels of education conditions conditions could not be achieved through either simple or pairwise effects
  + Unable to answer RQ2 using this approach

## Specification: Models {.smaller}

+ The model should have contained 14 $\beta$ coefficients (including the intercept):
  + 1 for intercept
  + 1 for openness
  + 1 for conscientiousness
  + 1 for extraversion
  + 1 for agreeableness
  + 1 for neuroticism
  + 2 for education condition $(k-1 = 3-1 = 2)$
  + 2 for neuroscience condition $(r-1 = 3-1 = 2)$
  + 4 for interaction $((k-1)(r-1) = (k-1)(r-1) = (2)(2) = 4)$

+ Common for reports to contain model in the form of `R` code (i.e., included 9 $\beta$ coefficients (including the intercept) where there was:
  + 5 (though in some cases 1 collapsed into 'personality') for each of the Big 5 traits
  + 1 for eduction condition
  + 1 for neuroscience condition
  + 1 for interaction


## Specification: Null & Alternative Hypotheses

+ Errors mostly followed from previous mis-specification of model, where many reports then suggested that RQ1 could be answered by evaluating a single $\beta$ coefficient (e.g., $H_0: \beta_{9} = 0$)

+ Some suggested that all four interaction $\beta$ coefficients had to $\neq$ 0 in alternative hypothesis

+ In some cases, no hypotheses formally stated or described in text - unclear what models were being used to test


## Assumptions: Details {.smaller}
    
**Analysis Strategy**       
    
+ In many cases there was insufficient information given concerning assumption checks. What needs to be included:        
  + the assumption being evaluated        
  + the plot used to evaluate this assumption       
  + how you would determine whether the assumption was met        
  + what you will do if there is a violation        
      
          
**Results**       
    
+ Linearity of model assumed, but the covariates were continuous - needed to check this          
+ Many just referred to appendix with no comment on how the assumptions were met. 
    
What needs to be included:        
    
  + reference to all assumptions mentioned in analysis strategy       
  + evaluation of plot - how we know the assumption was met       
  + any action taken if assumption violated (e.g., removed $n$ outliers since [insert criteria to justify])       
  
      
## Analysis Strategy: Unable to Replicate {.smaller}

+ Report instructions: "Your goal should be to describe everything clearly enough that another researcher could fully reproduce your analysis without ever seeing your code"
  + Data checks not mentioned (e.g., were scores in range? Were there `NA` values?)
  + No statement of alpha (e.g., were statements of significance made in relation to $\alpha = .05$, $\alpha = .01$, or something else?)
  + Not clear which coding scheme was used (dummy or effects? why?)
  + Not stated what reference levels were used or why certain levels were chosen (i.e., lacking rationale)
  + Models and/or hypotheses not formally presented or described in words (or there was a mis-match)
  + Weights for contrast not provided, and so could not replicate
  + Specifying a hypothesis for contrast analysis, but not stating what values represented 
  + No mention or limited discussion of assumptions (what criteria was used?)
  
   
## Interpretation: Errors or Issues with Detail {.smaller}

**Errors**        
    
+ Mis-interpretation of results (e.g., estimates and/or significance), especially of interaction coefficents       
+ Incorrect contrast estimate       
  + Failed to check that the ordering of the weights matched the order of the levels within `emmeans`. This resulted in the incorrect value for the contrast       
  + How/why did this happen? Typically when factors were re-ordered or re-labelled incorrectly       
      
**Missing / Lacking Detail**        
    
+ Level of interpretation in some reports was very low, largely focusing on statistical significance        
+ Tables / figures presented, but not interpreted (e.g., only stating that descriptive statistics are presented in Table 1)        
    
**Too Much Detail**
    
+ Reporting every single coefficient contained in the model output. Important in the results to focus on what is relevant to the RQs       
+ Repetition of findings - e.g., repeated across tables/figures/in-text       
  

## Interpretation: What we Wanted to See
    
**Results**         
    
+ All key model results presented in main body of report        
+ A focus on the coefficients of interest - a full interpretation of *key* results        
+ Interpretation of the direction and magnitude of effects that were relevant         
    
**Discussion**        
    
+ Summary of findings for *both* RQs framed in relation to the questions not specific results          
+ Results summarised into concise and accurate take-home messages that provided a clear answer to the study RQs           
  
      
## Instructions: Structure

+ Report instructions: "Your report should include the following **three sections - analysis strategy, results, and discussion**"
  + Many reports did not follow the above structure - some included no sections, others included many more sections

+ Report instructions: "Your analysis strategy should **not** contain any results."
  + Some reports included descriptive statistics in this section and/or results from assumption checks
  
## Instructions: Structure

+ Report instructions: "The discussion section should be **very brief** (one or two sentences). Give the reader **succinct take-home messages** about what you conclude for the research question(s).This section should not introduce any new statistical results, nor repeat detailed statistical results"
  + Discussion sections were often too long
  + Some included reporting of statistical results (either repeated from results section or new)
  + Lack of linking to RQs, especially RQ2 (e.g., no statement of direction of difference)

+ Report instructions: "Your report should be a **maximum of four sides of A4** when the default formatting and font settings within RStudio.... **Except for the appendix, no content beyond the fourth page will be read or marked**"
  + Some reports had content beyond page 4 - render and check files before submission

## Instructions: R Code/Output

+ Report instructions: "There should be **no visible R code or direct output** in your report. This means that code chunks should be hidden in the pdf produced by your rmd file"

+ Common issues:
  
  + Many reports included `R` code and/or direct `R` output -- make sure to check your PDF files before submission
  + Not creating own tables (e.g., printing contrast weights `R` object instead of creating table)
  + Not checking chunk options
  
## Instructions: Key Tables/Figures in Appendix

Report instructions:        
    
+ "The appendix is **only for figures and results from assumption and diagnostic checks**. Any results from your main models that appear in the appendix will not be marked"


+ Any interpretation, table of results, figures of interaction models, etc. should have been in the main text if they contain the results for models you set out in your analysis strategy


## Instructions: Formatting

+ Report instructions: "Reporting should be **clear and consistent**. If in doubt, follow **APA 7th Edition** guidelines"
  + Often numbers rounded to varying decimal points
  
+ Report instructions: "Avoid unnecessary repetition of information/duplication of tables/plots"
  + Most commonly observed in relation to descriptive tables and plots
