---
title: "<b>Week 4: LM Assumptions </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(car)
```

# Weeks Learning Objectives
1. Understand the meaning of model coefficients in the case of a binary predictor.

2. Be able to state the assumptions underlying a linear model.

3. Understand how to assess if a fitted model satisfies the linear model assumptions.

4. Understand how to use transformations when the model violates assumptions.

---
# Topics for today
+ What are the assumptions of regression and how can we test them?
	+ Normally distributed errors
	+ Homoscedasticity
	+ Linearity
	+ Independence of errors

---
# Regression assumptions 
+ So far, we have discussed evaluating regression models with respect to:
	+ Overall model fit ($F$-ratio, $R^2$)
	+ Individual predictors
	+ Individual cases

+ However, the regression model is also built on a set of assumptions

+ If these assumptions are violated, the model will not be very accurate

+ Thus, we also need to test these assumptions

---
# Example: 


---
# The lm() results
```{r}

```

+ We do all our assumption testing after fitting the lm( ) model.
  + We will need to use information from the object `M1`

???
+ This is a good time to make sure we are happy with the idea of objects
+ `M1` is a lm() model object
+ Contains information about the model we ran, estimates of the residuals, predicted scores etc.

---
# Visualizations vs tests
+ In talking about assumption checks, will present statistical test and visualizations

+ In general, graphical methods are often more useful
	+ Easier to see the nature and magnitude of the assumption violation

+ Statistical tests often suggest assumptions are violated when problem is small

???
Make the general point about power and tests.

---
# Normally distributed errors 
+ **Assumption 1**: The residuals ($\epsilon_i$) are normally distributed around each predicted values.

+ Investigated with:
  + QQ-plots
  +	Histograms
	+ Shapiro-Wilk test

	
---
# Visualizations 
+ **Histograms**: Plot the frequency distribution of the residuals.

--
+ **Q-Q Plots**: Quantile comparison plots.
	+ Plot the studentised residuals from the model against their theoretically expected values.
	+ If the residuals are normally distributed, the points should fall neatly on the diagonal of the plot.
	+ Non-normally distributed residuals cause deviations of points from the diagonal.
		+ The specific shape of these deviations are characteristic of the distribution of the residuals.

---
# Visualizations

.pull-left[

```{r}

```

]


.pull-right[

```{r}

```

]

---
# shapiro.test() 
+ The Shapiro-Wilk test provides a significance test on the departure from normality.
+ A significant $p$-value ($\alpha = .05$) suggests that the residuals deviate from normality.

```{r}

```



---
#  Homoscedasticity 

+ **Assumption**: Error variance is constant and not related to the values of the predictors or the predicted values
	+ Heteroscedasticity refers to when this assumption is violated (non-constant variance)

+ Investigated with:
  + Plot residual values against each predictor and the predicted Y values
	+ Breusch-Pagan test (Non-constant variance test)

---
#  Residual-vs-predicted values plot 
+ In R, we can plot the Pearson residuals against the predictor values and predicted y-values using the `residualPlots()` function from the `car` package

--
+  residualPlots() output 
  + Categorical predictors should show a similar spread of residual values across their levels
  + The plots for continuous predictors should look like a random array of dots
	  + The solid line should follow the dashed line closely


---
#  Residual-vs-predicted values plot 

.pull-left[
+cont
```{r}

```

]

.pull-right[
+cat
```{r}

```

]

---
#  Breusch-Pagan test 

.pull-left[
+ Also called the non-constant variance test

+ Tests whether residual variance depends on the predicted values

+ Implemented using the `ncvTest()` function in R
  + Non-significant $p$-value suggests homoscedasticity assumption holds
]

.pull-right[

```{r}

```

]


---
#  Linearity 
+ **Assumption**: Each predictor has a linear relationship (straight line) with the outcome
  + Assuming a linear relation when the true relation is non-linear can result in under-estimating that relation

+ Investigated with:
  + Scatterplots with lowess lines. 


---
# Linear vs non-linear

.pull-left[
+linear
```{r}

```

]

.pull-right[
+non-linear
```{r}

```

]

---
#  What is a lowess line?


---
# Visualization

```{r}

```


---
#  Independence of errors 
+ **Assumption**: Residual values are not correlated with one another

+ Difficult to test unless we know the potential source of correlation between cases.

+ We can test a limited form of the assumption by testing for autocorrelation between errors.
	+ We can test the correlation between each case an adjacent cases in the dataset
	+ Achieved using the Durbin-Watson test

---
#  Durbin-Watson test 
+ Durbin-Watson test implemented in R using the `durbinWatsonTest()` function:

```{r}

```


+ The D-W statistic can take values between 0 and 4
	+ 2= no autocorrelation
+ Therefore, we ideally want D-W values close to 2 and a non-significant $p$-value
  + Values <1 or >3 may indicate problems


---
class: center, middle

# Time for a break
Head to LEARN and try quiz X on assumption visualizations

---
class: center, middle
# Violated Assumptions
What do we do about non-normality of residuals, heteroscedasticity and non-linearity? 


---
# Non-linear transformations 
+ Often non-normal residuals, heteroscedasticity and non-linearity can be ameliorated by a non-linear transformation of the outcome and/or predictors.

+ This involves applying a function (see first week) to the values of a variable. 
  + This changes the values and overall shape of the distribution

+ For non-normal residuals and heteroscedasticity, skewed outcomes can be transformed to normality

+ Non-linearity may be helped by a transformation of both predictors and outcomes

---
#  Transforming variables to normality 
+ Positively skewed data can be made more normally distributed using a log-transformation.

+ Negatively skewed data can be made more normally distributed using same procedure but first reflecting the variable (make biggest values the smallest and smallest the biggest) and then applying the log-transform

+ What does skew look like?

--- 
# Visualizing Skew

.pull-left[

+ Positive skew

```{r}

```

]

.pull-right[

+ Negative skew

```{r}

```

]

---
#  Log-transformations 
+ Log-transformations can be implemented in R using the `log()` function.

+ If your variable contains zero or negative values, you need to first add a constant to make all your values positive
	+ A good strategy is to add a constant so that your minimum value is one
	+ E.g., if your minimum value is -1.5, add 2.5 to all your values

---
# Log-transformation in action

.pull-left[

+ Positive skew

```{r}

```

]

.pull-right[

+ Log-transformed

```{r}

```

]


---
# Log-transformation in action

.pull-left[

+ Negative skew

```{r}

```

]

.pull-right[

+ Log-transformed

```{r}

```

]

---
#  Other potential solutions 
+ Adding higher-order terms to the model
  + Explicitly model non-linear relationships
  + We will look at this when we consider interactions.

+ Bootstrapping 
  + Discussed in dapR1
  + We will come back to this later in the course

---
# Summary of today


---
# Next tasks


