---
title: "<b>Week 2: Calculating coefficients</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


# Weeks Learning Objectives
1. Be able to specify a simple linear model. 
2. Understand what fitted values and residuals are. 
3. Be able to interpret the coefficients of a fitted model.
4. Be able to test hypotheses and construct confidence intervals for the regression coefficients.

---
# Topics for today
+ Now we know how to position our model line, we can consider:
  + Interpreting the values
  + Testing the significance

---
# Things to recap
+ We will look again at significance testing.

+ So we will also discuss sampling variability.

+ If at any point you feel like you are not following this section of the material, go back and review sampling and hypothesis testing material linked on LEARN.

---
# Recap linear model
+ The linear model for a single predictor is written as:

$$y_i = b_0 + b_1x_{1i} + \epsilon_i$$

+ **Intercept**: the point where the line cross $y$, and $x$ = 0

+ **Slope**: the gradient of the line, or rate of change

+ In our worked example we saw how to calculate the intercept and slope from the raw data.


---
# `lm` in R
+ We do not generally calculate by hand like this, so we also briefly introduced the `lm()` function.

```{r, echo=FALSE}
test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  hours = seq(0.5,5,.5),
  score = c(1,3,1,2,2,6,3,3,4,8)
)
```

```{r}
lm(score ~ hours, data = test)
```

---
# Interpretation
+ So we know in a general sense what the intercept and slope are, but what do they mean with respect to our data and question?

+ **Slope is the number of units by which Y increases for a unit increase in X.**

--
    + Unit of Y = 1 point on the test
    + Unit of X = 1 hour of study
    
--

+ So, for every hour of study, test score increases by 1.055 points.

---
# Interpretation
+ **Slope is the number of units by which Y increases for a unit increase in X.**
    + Unit of Y = 1 point on the test
    + Unit of X = 1 hour of study
    + So, for every hour of study, test score increases by 1.055 points.

+ **Intercept is the expected value of Y when X is 0.**

--
    + X = 0 is a student who does not study.

--
+ So, a student who does no study would be expected to score 0.40 on the test.


---
# Note of caution on intercepts
+ In our example, 0 has a meaning.
    + It is a student who has studied for 0 hours.
    
+ But it is not always the case that 0 is meaningful.

+ Suppose our predictor variable was not hours of study, but age.

+ **Look back at the interpretation of the intercept, and instead of hours of study, insert age. Read this aloud a couple of times.**

--

+ This is the first instance of a very general lesson about interpreting statistical tests. 
    + The interpretation is always in the context of the constructs and how we have measured them.


---
# Evaluating our model
+ At this point, we have estimated values for the key parameters of our model (intercept and slope).

+ Now we have to think about how we evaluate the model.

+ There are three ways to think about evaluation:

1. Evaluating the individual coefficients
2. Evaluating the overall model quality
3. Evaluating the model assumptions

+ Before accepting a set of results, it is important to consider all three of these aspects of evaluation.

???
Important to really emphasize this is a package of information and we want it all before we decide to accept our model.

---
#  Significance of individual effects 
+ A general way to ask this question would be to state: 

> **Does our model capture an informative relationship between X and Y?**


+ In the context of our example from last lecture, we could ask, 

> **Is study time a useful predictor of test score?**

+ The above is a research question/hypothesis. As we have done before, we need to turn this into a testable statistical hypothesis.

---
#  Evaluating individual predictors 
+ Revise the steps in hypothesis testing:

--
    + Research questions
    
--
    + Statistical hypothesis
    
--
    + Define the null
    
--
    + Calculate an estimate of effect of interest.
  
--
    + Calculate an appropriate test statistic.
    
--
    + Evaluate the test statistic against the null.
    

---
# Research question and hypotheses


---
# Defining null

+ Conceptually:
	+ If x yields no information on y, then
	+ Why?
		+ Beta gives the predicted change in y for a unit change in x.
		+ If y does not change as x changes, then beta = 0.
+ We can state this formally as a null and alternative:

---
# Point estimate and test statistic
+ We have this, this is $b_1$
  
  + So, that was nice and easy.

+ 
+ So we have a null, what we need to do now is construct a test statistic and the associated sampling distribution.
+ For beta?s, we use a t-statistic (again!):

where

And


---
# Test statistic 

+ This can be evaluated against a t-distribution with n-2 degree for a model with one predictor.
+ More generally this is written as N-K-1
	+ Where K is the number of predictors, and the additional -1 catches the intercept.

+ For , the appropriate test statistic is a t-statistic:
+ In regression, the degrees of freedom for *t* is N-k-1
	+ In simple regression with only one predictor, this gives us N-2 df
+ We compare our t-value against a t-distribution with N-2 df to assess statistical significance

---
#  Standard error of slope (b1)
+ The standard error (SE) provides a measure of how different values would be across repeated samples
+ Smaller SEs suggest is estimated more precisely
	+ Smaller SEs are better!


---
#  Formula for the standard error b1

+ The formula for the standard error of is:
+ Where:
	+ SSE is the sum of squared error (i.e., )
	+ N is the sample size
	+ k is the number of predictors (=1 for simple regression)
+ From this formula, we can see that the SEs for regression slopes will be smaller when:
	+ Residual variance ( is smaller
	+ Sample size, N is larger


---
#  Significance of b1

+ The SE is important in significance tests for
+ We can think of testing the significance of as follows:
	+ If an IV yields no information about the DV then
	+ Why?
		+ gives the predicted change in Y for a unit change in X
		+ If Y does not change as X changes, then = 0
+ We can state this formally as a null and alternative hypothesis:


---
#  A decision about the null 

+ So we have a t-value associated with our beta coefficient.
	+ t = 3.656
+ And we know we will evaluate it against a t-distribution with df = n-2 = 3.
+ As with all tests we need to set our alpha.
	+ Let?s take 0.05 two tailed.
+ This gives us a critical value of 3.11.
	+ What is our decision?

---
#  Decision about the null 
+ Applying this to our exam revision and test score example from last week where we had SSE=21.26 , 10 participants , = 20.625 and =1.055 gives us:

So;
+ We compare our t-value against a t-distribution with df=8 to assess statistical significance, giving us *p=* 0.019
+ We can conclude that *revision is a statistically significant predictor of test scores at alpha=0.05*



---
#  Confidence intervals for b1
+ We can also compute confidence intervals for
+ The 100(1-alpha), e.g., 95%, confidence interval for the slope is:
+ E.g., the 95% confidence interval for in our revision and test score example would be:

The confidence interval of 0.229 to 1.881 does not include zero, therefore, we can conclude that *revision is a statistically significant predictor of test scores* ( *p<.* 05).

---
# Summary of today

---
# Next tasks