---
title: "<b>Week 9: Example & Write Up </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Weeks Learning Objectives
1. Understand the difference between outliers and influential points and how to detect them.
2. Be able to comment on the issues of correlation vs causation and generalisability of the sample results.
3. Be able to run a full simple regression analysis from start to finish and to correctly interpret the results.

---
# Topics for today
+ Run an example analysis
+ Evaluate the write up


---
# Study

+ find a study with open data
+ re-run their model


---
# Write up
+ have students evaluate the paper write up
  + what is good
  + what is bad
  + give them a set of questions (essentially could they construct the model from the write up)
  + What is missing in reporting
  + check APA for tables and references (give them a link, ask them to check and comment)

---
# Summary of today

---
# Next tasks

