---
title: "<b> Factorial Designs </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(emmeans)
library(kableExtra)
```

# Week's Learning Objectives
+ Understand how to interpret simple effects for experimental designs
+ Understadn how to interpret exploratory pairwise analyses for factorial designs
+ Know how to do both analyses using `emmeans()`

---
# Just a little bit more of the hospital example
```{r}
hosp_tbl <- read_csv("hospital.csv", col_types = "dff")
m1 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl)
anova(m1)
```

---
# Our results
```{r}
m1sum <- summary(m1)
round(m1sum$coefficients,2)
```

---
# But where do we go next?
+ It is typically a bad idea to focus on main effects in the presence of an interaction.
  + The interaction means the effect of the condition differs dependent on the interacting variable.

+ So we need to understand more about the interaction

+ We will use the `emmeans` package to explore this futher:
  + We will start by looking at the visualizations
  + And then consider the simple effects
  
  
---
# Visualizing the interaction

.pull-left[
```{r, out.width ="80%"}
emmip(m1, Hospital ~ Treatment)
```
]

.pull-right[

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```

]

---
# Visualizing the interaction

.pull-left[
```{r, out.width ="80%"}
emmip(m1, Treatment ~ Hospital)
```
]

.pull-right[

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```

]

---
# Simple Effects
+ We noted previously that simple contrasts/effects consider the effect of one condition at a specific level of the other.
  + Is there an effect of `Hospital` for those receiving `Treatment A`? (and so on for all combinations)
  + Or, put another way, is there a difference in SWB between Hospitals 1 and 2 for people receiving Treatment A

+ We also know an interaction is defined as the change in the effect of one variable given the value of another.
  + So here, value = a specific level.
  + So by considering the simple effects, we can identify at which levels of the interacting condition we see different effects.

---
# Simple Effects with `emmeans`

```{r}
m1_emm <- emmeans(m1, ~Treatment*Hospital)
m1_simple1 <- pairs(m1_emm, simple = "Hospital")
m1_simple1
```

---
# Simple Effects with `emmeans`

```{r}
m1_simple2 <- pairs(m1_emm, simple = "Treatment")
m1_simple2
```


---
# Simple effects with plots

.pull-left[
```{r, echo=FALSE, out.width="90%"}
emmip(m1, Treatment ~ Hospital)
```

]

.pull-right[
```{r}
m1_simple1
```

]

---
# Simple effects with plots

.pull-left[
```{r}
m1_simple2
```

]

.pull-right[
```{r, echo=FALSE, out.width="90%"}
emmip(m1, Hospital ~ Treatment)
```

]




---
class: center, middle
# Time for a break

---
class: center, middle
# Welcome Back!
**One more step in exploration**

---
# Pairwise comparisons
+ So far we have been discussing tests which move from the very general to the specific:
  + Overall model F
  + Incremental F/ F per condition
  + Contrasts and codes
  + Simple effects
  
+ But we have one more layer that more closely aligns to looking at the $\beta$ coefficients, namely a fully exploratory pairwise analysis.

---
# Pairwise comparisons
+ As the name suggests, pairwise comparisons compare all levels of a given predictor variable with all levels of the other.

```{r}
pairs_res <- pairs(m1_emm)
pairs_res
```

---
# Visualize the comparison
```{r}
plot(m1_emm, comparisons = TRUE)
```

---
# Summary

---
class: center, middle
# Thanks for listening!
