---
title: "<b> Power Analysis </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(pwr)
library(paramtest)
library(tidyverse)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# `pwr` for F-tests
+ For linear models, we use `pwr.f2.test()`

```{r, eval=FALSE}
pwr.f2.test(u = , #numerator degrees of freedom (model)
            v = , #denominator degrees of freedom (residual)
            f2 = , #stat to be calculated (below)
            sig.level = ,
            power = 
            )
```

+ `u` and `v` come from study design.
  + `u` = predictors in the model ( $k$ ) 
  + `v` = n-k-1

+ There are two versions of $f^2$
  + these are specified as formula
  + you can also use a pre-selected value; Cohen suggests f2 values of .02, .15, and .35 reflect small, moderate, and large effect sizes. 
  
  
---
# `pwr` for F-tests
+ The first is:

$$f^2 = \frac{R^2}{1-R^2}$$
+ This should be used when we want to see the overall power of a set of predictors
  + Think overall model $F$-test

+ For example, if we wanted sample size for an overall $R^2$ of 0.10, with 5 predictors, power of 0.8 and $\alpha$ = .05


```{r, eval=FALSE}
pwr.f2.test(u = 5, #numerator degrees of freedom (model)
            #v = , #denominator degrees of freedom (residual)
            f2 = 0.10/(1-0.10), #stat to be calculated (below)
            sig.level = .05,
            power = .80
            )
```



---
# `pwr` for F-tests

```{r}
pwr.f2.test(u = 5, #numerator degrees of freedom (model)
            #v = , #denominator degrees of freedom (residual)
            f2 = 0.10/(1-0.10), #stat to be calculated (below)
            sig.level = .05,
            power = .80
            )
```

+ We need a sample of ~121 (115 + 5 + 1)

---
# `pwr` for F-tests
+ The second is:

$$f^2 = \frac{R^2_{AB} - R^2_{A}}{1-R^2_{AB}}$$

+ This is the power for the incremental-F or the difference between a restricted ( $R^2_A$ ) and a full ( $R^2_{AB}$ ) model.

+ For example, if we wanted sample size for a difference between 0.10 (model with 2 predictors) and 0.15 (model with 5 predictors), power of 0.8 and $\alpha$ = .05


```{r, eval=FALSE}
pwr.f2.test(u = 3, #numerator degrees of freedom (model)
            #v = , #denominator degrees of freedom (residual)
            f2 = (0.15 - 0.10)/(1-0.15), #stat to be calculated (below)
            sig.level = .05,
            power = .80
            )
```


---
# `pwr` for F-tests

```{r}
pwr.f2.test(u = 3, #numerator degrees of freedom (model)
            #v = , #demoninator degrees of freedom (residual)
            f2 = (0.15 - 0.10)/(1-0.15), #stat to be calculated (below)
            sig.level = .05,
            power = .80
            )
```

+ We need a sample of ~180 (174.4 + 5 + 1)
