---
title: "<b>Week 1: Prelims</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives
1. Review the main concepts from introductory statistics.
2. Understand the concept of a function.
3. Be able to discuss what a statistical model is.
4. Understand the link between models and functions.

---
# Topics for today
+ What we need to review:
  + Types of data
  + Principle of visualization
  + Hypothesis testing
  
---
# Data types


---
# Data visualization
+ We will make extensive use of data visulization
+ Don't try to learn everything now.
+ Work through the `ggplot2` tutorial in your own time
+ Spend time thinking about all the plots we show

---
# Structure of plots

---
# What to look for in plots?
+ What does the x-axis represent?
+ What does the y-axis represent?
+ What are the scales of x and y?
+ What is the within the plot (e.g. line, dots, bars)?
+ What does that represent?

---
# Hypothesis testing

---
# Next tasks
+ If everything in this short lecture made sense, nothing.
+ If you feel you need a refresher on any of the topics;
  + Go to LEARN dapR2 week 1 folder
  + Click on recap material
  + Select the topic you wish to recap and work through
  + Check for the open office hours related to the topic at the end of the week for questions