---
title: "<b>Week 5: LM Diagnostics </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Weeks Learning Objectives
1. Understand the difference between outliers and influential points and how to detect them.
2. Be able to comment on the issues of correlation vs causation and generalisability of the sample results.
3. Be able to run a full simple regression analysis from start to finish and to correctly interpret the results.

---
# Topics for today
+ Regression Outliers
+ Leverage
+ Influence
+ Dealing with ?problematic? cases

---
#  Regression Diagnostics 
+ In previous lectures we have discussed evaluating regression models:
	+ Evaluating model fit (R^2 and F-ratio)
	+ Evaluating individual predictors
+ Another important aspect of evaluating regression models pertains to the cases (individuals):
	+ Does the regression model fit poorly for some individuals?
	+ Do some individuals have a lot of influence on the results?
+ Regression diagnostics allow us to explore individual cases in the context of a regression model

---
#  Three important features 
+ Regression outliers
	+ Cases for which there is a large discrepancy between their predicted y-value and their observed y-value
+ High leverage cases
	+ Cases with an unusual combination of predictor values
+ High influence cases
	+ Cases who are having a large influence on the regression results

---
#  Regression Outliers 
+ Regression outliers are cases that have unusual DV values given their IV values
	+ They will show a large difference between its predicted and its observed value
	+ i.e. they will show a large *residuals*

---
#  Why are we interested in regression outliers? 
+ They can (but do not necessarily) have a strong influence on the model
	+ We'll get to this?
+ It may be informative to follow-up and investigate outliers
	+ Is it a data entry error?
	+ Does the case somehow not belong with the rest of the data?
		+ E.g., a male in a sample of females
+ ADD PLOT

---
#  How do we determine if a case is an outlier? 
+ We judge regression outlying-ness of a case on the basis of the size of its residual
+ Unstandardised residuals are
	+ They are in the same units as the DV
	+ Fine for comparison across cases within the same regression model
	+ Difficult to compare across models where the DVs will have different units

---
#  Standardised residuals 
+ Standardised residuals
	+ Divide the unstandardised residuals by an estimate of their standard deviation
	+ Converts the residuals to z-score units
	+ However, their calculation includes the potential outlier
+ Studentised residuals
	+ Provide a version of the standardised residual excluding the case
	+ Values **>+2 or <-2** indicate potential outlyingness

---
#  Studentised residuals in R 
+ Let's imagine we have collected data on n=95 adolescents
+ We're interested in whether we can predict violence from gender, alcohol use, and low self-control
+ We have fit our model using lm() and now want to look for regression outliers

---
#  studres( ) function 
+ After fitting our model...
+ we can extract the studentised residuals using the studres () function from the MASS package

---
#  Identifying studentised residuals >2 
+ The `stdz` object we created stores the studentised residual values for every case in the dataset, e.g., for the first 20 cases:
+ We can check which cases have studentised residuals > , using:
+ Specifies absolute value to capture both >+2 and <-2
+ Used to identify cases for which a specified statement is true
+ Row numbers where the value of stdz is greater than

---
#  Leverage 

+ High leverage cases are those with an unusual IV or combination of IV values
	+ In simple regression: an X value far from the mean of X
	+ In multiple regression: a set of X values far from the centroid of X
		+ Centroid= generalisation of the mean to multivariate space

---
#  Why are we interested in leverage? 
+ High leverage cases have considerable potential to influence the regression model
	+ i.e., change the estimates of the intercept and regression slopes
+ High leverage cases, when they are also regression outliers, will have high **influence**
+ ADD PLOTS

---
#  How do we determine if a case has high leverage? 
+ hat values are used to assess leverage in regression
+ For simple regression the hat value for a case would be:
  + ADD EQUATION
+ n is the sample size
+ is the squared deviation of an X value from the mean X
+ is the sum of squared deviations of each X value from the mean


---
#  Hat values 
+ Hat values range between 1/n and n
+ The average hat value for a regression is:
  + ADD EQUATION
+ k is the number of predictors
+ n is the sample size
+ As a rough heuristic, values more than 2x are considered high leverage

---
#  hatvalues () function 
+ After fitting our model...
+ we can extract the hat values using the hatvalues () function

---
#  Identifying leverage values >2x the average 
+ The hats object stores all our hat values, e.g., for the first 20 cases:
+ We can check which cases have hat values >2x the average value:


---
#  Influence 
+ Cases with high influence, have a strong effect on the coefficients
	+ Occurs when a case has both high leverage and is a regression outlier
+ If we deleted such a case, the regression coefficients would change substantially
+ If a handful of influential cases are responsible for the regression results, the conclusions might not generalise very well

+ ADD PLOT

---
#  How do we identify influential cases? 
+ We can look at the effect of deleting a case:
	+ DFFit
		+ The difference between the predicted DV value for a case with versus without a case included
	+ DFbeta
		+ For each regression coefficient, the difference between that coefficient with and without a case included
	+ DFbetas
		+ A standardised version of DFbeta , obtained by dividing by an estimate of the standard error of the regression coefficient with the case removed
	+ Cooks distance
		+ A single summary index of the influence of a case
		+ Most commonly used measure of influence

---
#  Cooks Distance 
+ ADD EQUATIONS SHOWING COOKS D COMBINES OUTLIERS WITH HAT VALUES (LEVERAGE)

+ So;

+ PUT IT IN WORDS

---
#  Cook's Distance 
+ We look for Cook's distance values where:

+ ADD EQUATION

+ The cut-off for our violence example with n=95, k=3 would, therefore, be 4/(95-3-1)=0.044.

---
#  Cook?s distance in R 
+ After fitting our model...
+ we can extract the Cook?s distance values using the cooks.distance () function

---
#  What should be done about outliers, high leverage and high influence values? 

+ Easier to identify unusual cases than it is to decide what to do about them ?
+ In general, not a good idea to delete cases just because they exceed a threshold on a measure of outlyingness , leverage, or influence
	+ Especially if they don?t show large influence
+ Instead, try to investigate why a case is unusual
+ Think of regression diagnostics more as a way of learning about the limitations of the model
	+ Which cases can?t it explain very well?
	+ Do results depend quite strongly on certain cases ?
+ ? rather than as a justification for deleting cases


---
#  Investigating  and dealing with unusual cases 
+ Is there a data entry error?
	+ Is the value within the plausible range of values for a variable?
	+ Can it be corrected? If not, the incorrect value should be deleted
+ Is the data legit but extreme?
	+ Consider *winsorisation* ** over deletion to reduce its influence
		+ Replacing the extreme value with the next largest value for that variable
		+ Avoids missingness/ preserves information
	+ Note that deleting or winsorising values can change the model, therefore, different cases might then show up as outlying, high leverage, or influential
		+ Iterative approach to removing/ winsorising cases is needed

---
#  Investigating and dealing with unusual cases 
+ Is the data legit but extreme?
	+ Might there be unexplained variance that could be explained by predictors missing from your model?
		+ Or an interaction between predictors (see lecture 11 and 12)
	+ Is there something about a case that suggests it does not belong with the rest of the sample?
		+ Does a case come from a different population that would be expected to have substantively different IV-DV relations?
		+ E.g., do you have clinical cases in a non-clinical sample, children in a largely adult sample etc.

---
#  Investigating and dealing with unusual cases 
+ Are there ?unusual? values as a result of skewness?
	+ Unusual values can sometimes result from skewed distributions
	+ Look at a histogram and skewness values to identify problems
	+ Consider transforming skewed variables to normality  (see lecture tomorrow)

---
#  Sensitivity Analyses 
+ Sensitivity analysis refers to the idea of checking whether you get similar results irrespective of the methodological decisions you make
+ Sensitivity analysis can be used to check whether you get the same pattern of results:
	+ With versus without including certain unusual cases
	+ With versus without transforming a variable
	+ Etc.
+ If results are highly similar, you have more confidence that the results aren?t very dependent on those decisions
+ If they differ a lot, this should be reported as a limitation
  
---
# Summary of today
+ Regression case diagnostics allow us to learn about:
	+ Cases that are poorly described by the model (regression outliers)
	+ Cases that have unusual combinations of predictor values (leverage)
	+ Cases that strongly affect the regression results (influence)
+ Typically we are most concerned about influential cases
	+ They can undermine the generalisability of the conclusions
+ Cases flagged by regression diagnostics should not necessarily be deleted, but followed-up to see if the reasons can be identified

---
# Next tasks

