---
title: "<b>Week 3: Model evaluation</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Weeks Learning Objectives
1. Understand the coefficient of determination and the F-test of model utility.
2. Understand the relationship between the correlation coefficient and the regression slope

---
# Topics for today
+ Overall model evaluation
  + Coefficent of determination
  + F-test for the model

---
#  Quality of the overall model 

+ The aim of our linear model is to build a model which describes Y as a function of X.
	+ That is we are trying to explain Y using X.
+ This model gives us predicted values for Y, and the difference between these an the actual values is:
+ SSE captures the misfit of our model.
	+ If SSE is small, then our model is good.

---
#  Quality of the overall model 

+ But how do we quantify small?
+ Well if the total variation is given by:
+ Then we might quantify small as the proportion of residual variance as compared to total variance.

---
#  Coefficient of determination 

+ R 2 = coefficient of determination
+ Quantifies the amount of variability in the outcome accounted for by the predictors.
+ Can be written as:

Or:

---
#  Our example 



---
#  Significance of the overall model 

+ The test of the individual predictors (IVs, or X?s) does not tell us if the overall model is significant or not.
	+ Neither does R-square
	+ But both are indicative
+ To test the significance of the model as a whole, we conduct an F-test.
	+ You will talk much more about F-tests in the context of ANOVA so this is just a brief intro

---
#  F-statistic 

+ Model variation is regression, is a little more complex than in ANOVA. So we will wait to discuss this in detail.
+ In short:
	+ If there is equivalent residual to model variation, F=1
	+ If there is more model than residual F > 1

---
#  Null hypothesis for F 

+ In words, the null hypothesis for the model says that the best guess of any individuals Y value is the mean of Y plus error.
	+ Or, that the X variables carry no information collectively about Y.
+ The alternative is that the best prediction of an individual Yi value includes the predictors.

---
#  F-distribution 

+ The *F-* distribution is a continuous probability (sampling) distribution used when we want to compare two chi-square distributions.
	+ Why is that relevant here?
+ Chi-square distributions are used when we have ?sums of squared? quantities.
	+ Think about the calculation of the chi-square statistic.
+ And what is the *F-* statistic?
	+ Ratio of variances (which are sums of squares)
+ The parameters of the F-distribution are the degrees of freedom association which each of the variance estimates (model and residual).

---
#  Decision about the null (model) 

+ For our example:
  
---
# Summary of today

---
# Next tasks