---
title: "<b> Testing and Evaluating LM</b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(simglm)

test <- tibble(
  student = paste(rep("ID",10),1:10, sep=""),
  hours = seq(0.5,5,.5),
  score = c(1,3,1,2,2,6,3,3,4,8)
)

res <- lm(score ~ hours, data = test)
```


# Week's Learning Objectives
1. Understand how to interpret significance tests for $\beta$ coefficients.
2. Understand how to calculate the interpret $R^2$ and adjusted- $R^2$ as a measure of model quality.
3. Be able to locate each of these tests in R `lm` model output.


---
class: inverse, center, middle

# Part 1: Recap & Overview

---
# Recap
+ Last week we introduced the general linear model equation:

$$y_i = \beta_0 + \beta_1 x_{i} + \epsilon_i$$

+ And we ran an example concerning test scores:

$$score_i = \beta_0 + \beta_1 hours_{i} + \beta_2 motivation_{i} + \epsilon_i$$

+ And we looked at how to run this model in R:

```{r, eval=FALSE}
lm(score ~ hours + motivation, data = test_study2)
```

---
# Evaluating our model
+ At this point, we have estimated values for the key parameters of our model ( $\beta$s ).

+ Now we have to think about how we evaluate the model.

+ There are three ways to think about evaluation:

  1. Evaluating the individual coefficients
  2. Evaluating the overall model quality
  3. Evaluating the model assumptions (later in the course)

+ Before accepting a set of results, it is important to consider all three of these aspects of evaluation.

???
Important to really emphasize this is a package of information and we want it all before we decide to accept our model.

---
#  Significance of individual effects 
+ A general way to ask this question would be to state: 

> **Is our model model informative about the relationship between X and Y?**

--

+ In the context of our example from last lecture, we could ask, 

> **Is study time a useful predictor of test score?**

--

+ The above is a research question/hypothesis. As we have done before, we need to turn this into a testable statistical hypothesis.

---
#  Evaluating individual predictors 
+ Steps in hypothesis testing:

--
  + Research questions
    
--
  
  + Statistical hypothesis
    
--
  
  + Define the null
    
--
  
  + Calculate an estimate of effect of interest.
  
--
  
  + Calculate an appropriate test statistic.
    
--
  
  + Evaluate the test statistic against the null.
    

---
# Research question and hypotheses

+ **Research questions** are statements of what we intend to study. 

+ A good question defines:

--

  + Constructs under study
  + the relationship being tested
  + A direction of relationship
  + target populations etc.

> **Does increased study time improve test scores in school age children?**

--

+ **Statistical hypotheses** are testable mathematical statements.

--

  + In typical testing in Psychology, we define have a **null ( $H_0$ )** and an **alternative ( $H_1$ )** hypothesis.
  + $H_0$ is precise, and states a specific value for the effect of interest.
  + $H_1$ is not specific, and simply says "something else other than the null is more likely"


---
# Statistical significance: Overview

+ Remember, we only ever test the null.

+ We select a significance level, $\alpha$ (typically .05)

+ Then we calculate the $p$-value associated with our test statistic (here $\beta$ )

+ If the associated $p$ is smaller, then we **reject** the null.

+ If it is larger, then we **fail to reject** the null.


---
# Our results
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(7284) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1)),
  error = list(variance = 20),
  sample_size = 150,
  reg_weights = c(0.6, 1.4, 1.5)
)

df <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study2 <- df %>%
  dplyr::select(y, hours, motivation) %>%
  mutate(
    ID = paste("ID", 101:250, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2)
  ) %>%
  dplyr::select(ID, score, hours, motivation)

```

```{r}
performance <- lm(score ~ hours + motivation, data = test_study2); summary(performance)
```

---
class: inverse, center, middle

# Part 2: Steps in significance testing (the details)

---
# Defining null

+ Conceptually:
	+ If $x$ yields no information on $y$, then $\beta_1 = 0$
	
+ **Why would this be the case?**

--
	+ $\beta$ gives the predicted change in $y$ for a unit change in $x$.
	+ If $x$ and $y$ are unrelated, then a change in $x$ will not result in any change to the predicted value of $y$
	+ So for a unit change in $x$, there is no (=0) change in $y$.
	
+ We can state this formally as a null and alternative:

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$

???
+ For the null to be testable, we need to formally define it. 
+ Point out here the difference in the specificity of the hypotheses. $H_0$ is that the $b_1$ takes a specific value. $H_1$ is that $b_1$ has some value that is not this specific value. i..e one is directly testable, the other is not.


---
# Point estimate and test statistic

+ We have already seen how we calculate $\hat \beta_1$.

+ The associated test statistic to for $\beta$ coefficients is a $t$-statistic

$$t = \frac{\hat \beta}{SE(\hat \beta)}$$

+ where

  + $\hat \beta$ = any $\beta$ coefficient we have calculated
  + $SE(\hat \beta)$ = standard error of $\beta$ 

--

+ **Recall** that the standard error describes the spread of the sampling distribution.
  + The standard error (SE) provides a measure of sampling variability
  + Smaller SE's suggest more precise estimate (=good)
  
???
+ brief reminders on test statistics
  + every quantity we wish to calculate a significance test for needs an test statistic.
  + the test statistic is a value that has a known sampling distribution
+ If sampling distribution is unfamiliar, again, recap the hypothesis testing material


---
# Lets look at the output from `lm` again
```{r}
summary(performance)
```


---
# And work out the $t$-values

+ We can check the value for `motivation` first:

$$t = \frac{\hat \beta_1}{SE(\hat \beta_1)} = \frac{0.9163}{0.3838} = 2.388 (3dp)$$

+ Check `hours` in your own time.

+ So we know where the $\beta$ values come from, and we have just seen $t$, what about the $SE$ and $p$

---
#  SE( $\hat \beta_1$ )

+ The formula for the standard error of the slope is:

$$SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{j}})^2(1-R_{xj}^2)}}$$
+ Where:
	+ $SS_{Residual}$ is the residual sum of squares
	+ $n$ is the sample size
	+ $k$ is the number of predictors
	+ $x_{ij}$ is the observed value of a predictor ( $j$ ) for an individual ( $i$ )
	+ $\bar{x_{ij}$ is the mean of a predictor
	+ $(1 - R_{xj}^2)$ is the multiple correlation coefficient of the predictors

+ $(1 - R_{xj}^2)$ captures to degree to which all of our predictors are related.
  + For simple linear models, this = 0 as there is only 1 predictor
  
---
# SE( $\hat \beta_1$ )

$$SE(\hat \beta_j) = \sqrt{\frac{ SS_{Residual}/(n-k-1)}{\sum(x_{ij} - \bar{x_{j}})^2(1-R_{xj}^2)}}$$  
+ We want our $SE$ to be smaller - this means our estimate is precise

+ Examining the above formula we can see that:
	+ $SE$ is smaller when residual variance ( $SS_{residual}$ ) is smaller
	+ $SE$ is smaller when sample size ( $N$ ) is larger
	+ $SE$ is larger when the number of predictors ( $k$ ) is larger
	+ $SE$ is larger when a predictor is strongly correlated with other predictors ( $R_{xj}^2$ )

+ So all we have left is $p$

???
+ Well return to this later when we discuss multi-collinearity issues




---
# Sampling distribution for the null

+ Now we have our $t$-statistic, we need to evaluate it.

+ For that, we need sampling distribution for the null.

+ For $\beta$, this is a $t$-distribution with $n-k-1$ degrees of freedom.
	+ Where $k$ is the number of predictors, and the additional -1 represents the intercept.

--

+ So in our example:
  + n = 150
  + k = 2
  + 150 - 2 - 1 = 147

---
#  A decision about the null 
+ So we have a $t$-value associated with our $\beta$ coefficient.
	+ t = 2.388

+ And we know we will evaluate it against a $t$-distribution with 147 df.

+ As with all tests we need to set our $\alpha$.
	+ Let's take 0.05 two tailed.

--

+ Now we need a critical value to compare our observed $t$-value to.

---
# Visualize the null

.pull-left[
```{r, echo=FALSE}
ggplot() + 
  xlim(-8, 8) +
  stat_function(fun=dt,
                geom = "line",
                args = list(df=147)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.025, 8), -8),
                alpha=.25,
                fill = "blue",
                args = list(df=147)) +
    stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.975, 8), 8),
                alpha=.25,
                fill = "blue",
                args = list(df=147)) +
  geom_vline(xintercept = 2.388, col="red") +
  xlab("\n t") +
  ylab("") +
  ggtitle("t-distribution (df=147); t-statistic (2.388; red line)")
```

]

.pull-right[

+ Critical value and $p$-value:
```{r}
tibble(
  LowerCrit = round(qt(0.025, 147), 3),
  UpperCrit = round(qt(0.975, 147), 3),
  Exactp = (1 - pt(2.388, 147)) * 2
)
```

]

???
+ discuss this plot.
+ remind them of 2-tailed
+ areas
+ % underneath each end
+ comment on how it would be different one tailed
+ remind about what X is, thus where the line is

---
class: inverse, center, middle

# Part 3: An alternative using confidence intervals

---
# Refresher: What is a confidence interval?

[PINCH FROM DAPR1?]

---
#  Confidence intervals for $\beta$
+ We can also compute confidence intervals for $\hat \beta$
+ The $100 (1 - \alpha)$, e.g., 95%, confidence interval for the slope is:

$$\hat \beta_1 \pm t^* \times SE(\hat \beta_1)$$

+ So, 95% confidence interval for in our example for the effect of `motivation` would be:

```{r}
tibble(
  LowerCI = round(0.91634 - (qt(0.975, 147) * 0.38376), 3),
  UpperCI = round(0.91634 + (qt(0.975, 147)* 0.38376), 3)
)

```



+ The confidence interval of 0.158 to 1.675 does not include zero, 
  + Therefore, we can conclude that **motivation is a statistically significant predictor of test scores** ( $p < .05$).


---
# `confint` function

+ We can get confidence intervals for our models more easily than this:

```{r}
confint(performance)
```


---
# Where are we up to...

```{r}
performance <- lm(score ~ hours + motivation, data = test_study2)
summary(performance)
```


---
class: inverse, center, middle

# Part 4: Cofficient of determination ( $R^2$ )


---
#  Quality of the overall model 
+ When we measure an outcome ( $y$ ) in some data, the scores will vary (we hope).
  + Variation in $y$ = total variation of interest.

--

+ The aim of our linear model is to build a model which describes $y$ as a function of $x$.
	+ That is we are trying to explain variation in $y$ using $x$.

--

+ But it won't explain it all.
  + What is left unexplained is called the residual variance.

--

+ So we can breakdown variation in our data based on sums of squares as;

$$SS_{Total} = SS_{Model} + SS_{Residual}$$

---
#  Coefficient of determination 
+ One way to consider how good our model is, would be to consider the proportion of total variance our model accounts for. 

$$R^2 = \frac{SS_{Model}}{SS_{Total}} = 1 - \frac{SS_{Residual}}{SS_{Total}}$$

+ $R^2$ = coefficient of determination

--

  + Quantifies the amount of variability in the outcome accounted for by the predictors.
  + More variance accounted for, the better.
  + Represents the extent to which the prediction of $y$ is improved when predictions are based on the linear relation between $x$ and $y$.

--

+ Let's see how it works.
  + To do so, we need to calculate the different sums of squares.


---
# Total Sum of Squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Total} = \sum_{i=1}^{n}(y_i - \bar{y})^2$$

+ Squared distance of each data point from the mean of $y$.

+ Mean is our baseline. 

+ Without any other information, our best guess at the value of $y$ for any person is the mean.

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Total}')) +
  geom_hline(aes(yintercept = mean(score)),color="blue", linewidth=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]


---
# Residual sum of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Residual} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2$$

+ Which you may recognise.

+ Squared distance of each point from the predicted value.
]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Residual}')) +
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", linewidth=1) + 
  geom_segment(aes(x = hours, y = score, xend = hours, yend = c(res$fitted.values)), color = "red", lty =2)

```

]


---
# Model sums of squares

.pull-left[
+ Sums of squares quantify difference sources of variation.

$$SS_{Model} = \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2$$

+ That is, it is the deviance of the predicted scores from the mean of $y$.

+ But it is easier to simply take:

$$SS_{Model} = SS_{Total} - SS_{Residual}$$

]

.pull-right[

```{r, echo=FALSE}
test %>%
  ggplot(., aes(hours, score)) +
  geom_point(color="red", size = 2) +
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0))) +
  theme(axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0))) +
  xlab("Hours of Study") +
  ylab("Test Score") +
  ggtitle(latex2exp::TeX('SS_{Model}')) +
  geom_hline(aes(yintercept = mean(score)),color="blue", linewidth=1) + 
  geom_abline(aes(intercept=res$coefficients[1], slope=res$coefficients[2]), color="lightblue", linewidth=1) + 
  geom_segment(aes(x = hours, c(res$fitted.values), xend = hours, yend = c(rep(mean(score),10))), color = "red", lty =2)
```

]

---
# Values in our sample
+ In the current example, these values are:

  + $SS_{total}$ = 8556.06
  + $SS_{residual}$ = 2826.83
  + $SS_{model}$ = 5729.23

+ In the LEARN folder there is a document which shows the calculations from the raw data


---
#  Coefficient of determination 
+ Now we can finally come back to $R^2$.

$$R^2 = 1 - \frac{SS_{Residual}}{SS_{Total}}$$

+ Or

$$R^2 = \frac{SS_{Model}}{SS_{Total}}$$

+ So in our example:

$$R^2 = \frac{SS_{Model}}{SS_{Total}} = \frac{5729.23}{8556.06} = 0.6695$$

+ ** $R^2$ = 0.6695 means that 66.95% of the variation in test scores is accounted for by hours of revision and student motivation.**

---
#  Our example 

```{r}
summary(performance)
```

???
As at the end of last session, we can check this against the R-output:
Be sure to flag small amounts of rounding difference from working through "by hand" and so presenting to less decimal places.

---
#  Adjusted $R^2$ 
+ We can also compute an adjusted $R^2$ when our `lm` has 2+ predictors.
  + $R^2$ is an inflated estimate of the corresponding population value

+ Due to random sampling fluctuation, even when $R^2 = 0$ in the population, it's value in the sample may $\neq 0$ 

+ In **smaller samples** , the fluctuations from zero will be larger on average

+ With **more IVs** , there are more opportunities to add to the positive fluctuation


$$\hat R^2 = 1 - (1 - R^2)\frac{N-1}{N-k-1}$$

+ Adjusted $R^2$ adjusts for both sample size ( $N$ ) and number of predictors ( $k$ )


---
#  In our example 

```{r}
summary(performance)
```

---
#  In our example 
```{r, echo=FALSE}
sum_performance <- summary(performance)
```

+ **Based on adjusted R-squared, hours studying and student motivation explain `r round(sum_performance$adj.r.squared,3)*100`% of the variance in test scores**

+ As the sample size is large and the number of predictors small, unadjusted ( `r round(sum_performance$r.squared,3)` ) and adjusted R-squared ( `r round(sum_performance$adj.r.squared,3)` ) are similar.


---
# Summary

+ Key take homes:
  1. We have an inferential test, based on a $t$-distribution, for the significance of $\beta$.
  2. We are more likely to find a significant effect when we have picked good variables (smaller residual SS) and we have a large sample.
  3. We can assess the degree to which our model explains variance in the outcome based on $R^2$
  4. When we have multiple predictors, we can use the adjusted $R^2$ to account for the random fluctuations due to the model being more complex.
  
+ Next week we will look at model comparisons and standardization of coefficients.

---
class: inverse, center, middle

# Thanks for listening