---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)

set.seed(1512)
df <- tibble(
  RT = c(round(rnorm(100, 175, 10),0), round(rnorm(100, 215, 20),0)),
  Group = c(rep(0, 100), rep(1, 100))
)
```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

.pull-left[
I conduct an experiment looking at reaction times (in milliseconds) of participants with and without the presence of cute puppy distractor in the room. I have 100 participants in the "No puppy" = 0 group, and 100 participants in the "puppy" = 1 group.

The group means are shown below on the right handside.

I run a liner model:

```{r, eval=FALSE}
lm(RT ~ Group, data = df)
```

**What will the values for the intercept and slope be?**
]

.pull-right[

```{r}
average <- df %>%
  group_by(Group) %>%
  summarise(
    Mean = mean(RT)
  ) 

kable(average)
```

]
---
# Question 1: Answer

+ **Intercept**: `r average[[1,2]]`
  + Mean of group = 0, "no puppy"

+ **Slope (Group)**: `r average[[2,2]] - average[[1,2]]`
  + Group1 - Group0 (or the difference between groups)


---
# Question 2

**Write an interpretation of the effect of Group (or puppy).**


---
# Question 2: Answer

+ The expected RT of those in the "puppy" group is `r average[[2,2]] - average[[1,2]]` milliseconds slower than the "no puppy" group.
+ The puppy is a distraction!!



---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 