---
title: "dapR2 Lab Report 1 - Live R"
author: "Tom Booth"
date: "08/02/2021"
output: 
  bookdown::html_document2
---

## Analysis Strategy
**Points to include, not necessarily in this order:**

- Data checks: Missing data, out of range values, coding of categorical variables
- Personality data: At some point, either here or in model discussion, need to standardized.

- Assumptions:
  - What assumptions need to be tested?
  - What tests/plots are used?
  - How are they evaluated?
  - Could also comment here where the main results will be reported
- Diagnostics:
  - How will you check for influential cases, what are cut-off etc.

- Alpha level
  - Any other adjustments to alpha/ methods of inference

### Models
The key thing here is to describe the model, the variables within the model, and the information you will use to answer the question

For all questions:

- Models
- What information is used to answer the questions?
  - Significance of F, t, incremental model fit tests etc.
  - Magnitude of the beta coefficients
- This links to back to the question.
- Rationales for decision

**Q1 and Q2** answered with the same model, with reference level changed.

- Outcome, self-rated health
- Predictor location with countryside (or suburb) as reference level.
- Q1 answer by F-test significance and comment on variance explained.
- Q2 answered by the significance of the beta coefficient for the difference between countryside and suburb, with an interpretation of that effect.

**Q3** answered with single lm with all data.

- Outcome, self-rated health
- Predictors, all personality variables. z-scored so as to mean the beta coefficients are standardized and to allow comparison. This is also efficient for the final model which needs an interaction.
- Predictors location and age.
- Question answered by comparison of the beta-coefficients.
- Additional comments could be made on the overall model significance, significance of individual personality variables and variance explained. 

**Q4** add the interaction and probe.

- Outcome, self-rated health
- Predictors, everything!
- Focal test is the significance of the interaction term fro N and C
- Support for healthy-N is a buffering interaction assessed via the direction of the coefficients (-N, +C, + N*C), and a simple slopes plot.

## Results
Report the above.

### Descriptives
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(sjPlot)
tib <- read_csv("C:/Work/Teaching/psychstats/assignments/dapr2/dapr2_report.csv", col_types = "fdfdddddd")
```

### Q1 and Q2
```{r, boxplot, fig.cap = "<br> *Hello, is it a boxplot you are looking for*"}
tib %>%
  ggplot(., aes(x=location, y=srh)) + 
  geom_boxplot()
```

As you can see in **Figure\@ref(fig:boxplot)** 



```{r}
tib %>%
  group_by(location) %>%
  summarise(
    mean = mean(srh),
    sd = sd(srh)
  )
```

```{r}
levels(tib$location)
```

```{r}
contrasts(tib$location) <- contr.treatment(3, 2)
m1 <- lm(srh ~ location, data = tib)
summary(m1)
```

(*F*(2,747)=3.253, *p*<.05 )

```{r}
tab_model(m1, show.se = T, show.stat = T, string.se = "Bob",
          pred.labels = c("Intercept", "Location (City)", "Location (Countryside)"))
```


### Q3

```{r}
m2 <- lm(srh ~ age + location + scale(N) + scale(E) + scale(O) + scale(A) + scale(C), data = tib)
summary(m2)
```


### Q4

```{r}
m3 <- lm(srh ~ age + location + scale(N) + scale(E) + scale(O) + scale(A) + scale(C) + scale(N)*scale(C), data = tib)
summary(m3)
```

```{r}
anova(m2,m3)
```

```{r}
library(interactions)
interact_plot(m3, pred = N, modx = C, data = tib)
```


Include: 
  - Tables
  - Figures
  - Interaction plot
  - Simple slopes or regions of significance plot
  - Comment on whether the interaction is buffering or not.
  
## Discussion
Essentially 1 - 2 sentences per model/question.
Key points:
  - Don't repeat results verbatim
  - Comment on all questions
  - Be accurate with respect to the results shown
  
## General Writing

## Assumption/Diagnostic Appendix
Label!!!!