---
title: "<b>Categorical Predictors </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(effsize)
library(simglm)

theme_set(theme_gray(base_size = 15))

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.retina = 1.5)
```

# Weeks Learning Objectives
1. Understand the meaning of model coefficients in the case of a binary predictor.

2. Understand how to apply dummy coding to include categorical predictors with 2+ levels.

3. Be able to include categorical predictors into an `lm` model in R

---
#  Recap: Categorical variables 
+ Categorical variables can only take discrete values
	+ E.g., animal type: 1= duck, 2= cow, 3= wasp

+ They are mutually exclusive
  + No duck-wasps or cow-ducks!

+ In R, categorical variables should be of class `factor`
  + The discrete values are `levels`
  + Levels can have numeric values (1, 2 3) and labels (e.g. "duck", "cow", "wasp")
  + All that these numbers represent is category membership.
  

---
# Recap: Binary variable
+ Binary variable is a categorical variable with two levels.

+ Traditionally coded with a 0 and 1

+ In `lm`, binary variables are often referred to as dummy variables
  + when we use multiple dummies, we talk about the general procedure of dummy coding

+ Why 0 and 1?

--
  
+ Quick version: It has some nice properties when it comes to interpretation.

--

  + What is the interpretation of the intercept?

--

  + What about the slope?


---
# Extending our example

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(3119) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1),
               study = list(),
               study_method = list()),
  error = list(variance = 20),
  sample_size = 200,
  reg_weights = c(0.6, 1.4, 1.5)
)

df3 <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study3 <- df3 %>%
  dplyr::select(y, hours, motivation) %>%
  mutate(
    ID = paste("ID", 101:250, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2)
  ) %>%
  dplyr::select(ID, score, hours, motivation)

```



.pull-left[
+ Our in class example so far has used test scores and revision time and motivation.

+ Let's we also collected data on who they studied with (`study`);
  + 0 = alone; 1 = with others
  
+ And also which of three different revisions methods they used for the test (`study_method`)
	+ 1 = Notes re-reading; 2 = Notes summarising; 3 = Self-testing ([see here](https://www.psychologicalscience.org/publications/journals/pspi/learning-techniques.html))

+ We collect a new sample of 200 students.
]


.pull-right[



]

---
#  LM with binary predictors 
+ Now lets ask the question:

  + **Do students who study with others score better than students who study alone?**

+ Our equation does not change: 

$$score_i = \beta_0 + \beta_1 study_{i} + \epsilon_i$$

+ And in R:

```{r}
performance_study <- lm(score ~ study, data = test_study3)
```

---
# Model results
```{r}
summary(performance_study)
```

---
# Interpretation

.pull-left[
+ As before, the intercept $\hat \beta_0$ is the expected value of $y$ when $x=0$

+ What is $x=0$ here?
  + It is the students who study alone.

+ So what about $\hat \beta_1$?
  + $\beta_1$ = 

+ **Look at the output on the right hand side.** 
  + What do you notice about the difference in averages?

]

.pull-right[
```{r warning=FALSE, message=FALSE}
test_study3 %>%
  group_by(., study) %>% #<<
  summarise(
    Average = round(mean(score),4) #<<
  )
```


]


---
# Interpretation
+ $\hat \beta_0$ = predicted expected value of $y$ when $x = 0$
  + Or, the mean of group coded 0 (those who study alone)
  
+ $\hat \beta_1$ = predicted difference between the means of the two groups.
  + Group 1 - Group 0 (Mean `score` for those who study with others - mean `score` of those who study alone)
  
+ Notice how this maps to our question. 
  + Do students who study with others score better than students who study alone?
  

---
class: center, middle
# Quick pause for questions


---
#  Equations for each group 
+ What would our linear model look like if we added the values for $x$.


$$\widehat{score} = \hat \beta_0 + \hat \beta_1 study$$


+ For those who study alone ( $study = 0$ ):


$$\widehat{score}_{alone} = \hat \beta_0 + \hat \beta_1 \times 0$$


+ So;


$$\widehat{score}_{alone} = \hat \beta_0$$



---
#  Equations for each group 
+ For those who study with others ( $study = 1$ ):


$$\widehat{score}_{others} = \hat \beta_0 + \hat \beta_1 \times 1$$


+ So;


$$\widehat{score}_{others} = \hat \beta_0 + \hat \beta_1$$


+ And if we re-arrange;


$$\hat \beta_1 = \widehat{score}_{others} - \hat \beta_0$$


+ Remembering that $\widehat{score}_{alone} = \hat \beta_0$, we finally obtain:

$$\hat \beta_1 = \widehat{score}_{others} - \widehat{score}_{alone}$$


---
#  Visualize the model

```{r, echo=FALSE}
bin <- df %>%
  ggplot(., aes(x=factor(study), y=score, colour = study)) +
  geom_point(alpha=0.4) +
  labs(x = "\n Study", y = "Test Score \n") +
  ylim(0,10) +
  scale_x_discrete(labels = c("alone", "others")) +
  theme(legend.position = "none")

bin
```


---
#  Visualize the model
```{r, echo=FALSE}
bin +
  geom_jitter(width = .1, height = 0, alpha=0.4)

```


---
#  Visualize the model
```{r, echo=FALSE}
gpM <- df %>%
  group_by(study) %>%
  summarise(
    score = mean(score)
  )

bin +
  geom_jitter(width = .1, height = 0, alpha=0.4) +
  geom_errorbar(data = gpM, width=0.6,aes(ymax=..y..,ymin=..y..), size=1)

```



---
#  Visualize the model

```{r, echo=FALSE}
library(latex2exp)

bin +
  geom_jitter(width = .1, height = 0, alpha=0.4) +
  geom_errorbar(data = gpM, width=0.6, aes(ymax=..y..,ymin=..y..), size=1)+
  geom_segment(x=1.5, y=gpM[[1,2]], xend=1.5, yend=gpM[[2,2]], size =1, col="red") +
  geom_segment(x=1.48, y=gpM[[1,2]], xend=1.52, yend=gpM[[1,2]], size =1, col="red") +
  geom_segment(x=1.48, y=gpM[[2,2]], xend=1.52, yend=gpM[[2,2]], size =1, col="red") +
  geom_text(x=1.55, y = 5.5, label = TeX('$\\hat{\\beta}_1$'), size=5, col = "red") +
  geom_text(x=0.65, y = gpM[[1,2]] , label = TeX('$\\hat{\\beta}_0$'), size=5, col = "red")

```


---
#  Evaluation of model and significance of $\beta_1$

+ $R^2$ and $F$-ratio interpretation are identical to their interpretation in models with only continuous predictors.

+ And we assess the significance of predictors in the same way

+ We use the standard error of the coefficient to construct:
  + We calculate the $\hat \beta_1$ = difference between groups
	+ $t$-value and associated $p$-value for the coefficient
	+ Or a confidence interval around the coefficient


---
# Hold on... it's a t-test

```{r}
df %>%
  t.test(score ~ study, .)
```

???
Yup!


---
class: center, middle
# Time for a break!


---
#  Including categorical predictors with >2 levels in a regression 
+ The goal when analyzing categorical data with 2+ levels, we want each of our $\beta$ coefficients to represent a specific difference between means.
  + e.g. like with the use of a single binary variable, $\beta_1$ is the difference between the two groups.

+ When we have 2+ levels, to be able to do this, we need to apply a **coding scheme**. 

+ Two common coding schemes are:
  + Dummy coding (which we will discuss now)
  + Effects coding (or sum to zero, which we will discuss at the start of next semester)
  + There are LOTS of ways to do this, if curious, see !(here)[] UCLA

---
#  Dummy coding 
+ Dummy coding uses 0's and 1's to represent group membership
	+ One level is chosen as a baseline
	+ All other levels are compared against that baseline
	
+ Notice, this is identical to binary variables already discussed.

+ Dummy coding is simply the process of producing a set of binary coded variables

+ For any categorical variable, we will create $k$-1 dummy variables
  + $k$ = number of levels

---
# Steps in dummy coding
1. Choose a baseline level

2. Assign everyone in the baseline  group `0` for all $k$-1 dummy variables

3. Assign everyone in the next group a `1` for the first dummy variable and a `0` for all the other dummy variables

4. Assign everyone in the next again group a `1` for the second dummy variable and a `0` for all the other dummy variables

5. Repeat step 5 until all $k$-1 dummy variables have had 0's and 1's assigned

6. Enter the $k$-1 dummy variables into your regression


---
#  Choosing a baseline? 
+ Each level of your categorical predictor will be compared against the baseline.

+ Good baseline levels could be:
	+ The control group in an experiment
	+ The group expected to have the lowest score on the outcome
	+ The largest group

+ It is best the baseline is not:
	+ A poorly defined level, e.g. an `Other` group
	+ Much smaller than the other groups



---
#  Dummy coding 

.pull-left[
+ We start out with  a dataset that looks like:

```{r, echo=FALSE}
dum_dat %>%
  select(ID:method) %>%
  slice(1:10)
```


]


.pull-right[
+ And end up with one that looks like:

```{r, echo=FALSE}
dum_dat %>%
  slice(1:10)
```

]


---
#  Dummy coding with `lm` 

+ `lm` automatically applies dummy coding when you include a variable of class `factor` in a model.

+ It selects the first group as the baseline group

+ We write:

```{r, eval=FALSE}

mod1 <- lm(exam ~ method, data = dum_dat)

```


+ And `lm` does all the dummy coding work for us

---
#  Dummy coding with `lm`

.pull-left[
+ The intercept is the mean of the baseline group (notes re-reading)

+ The coefficient for `method2` is the mean difference between the notes summarising group and the baseline group

+ The coefficient for `method3` is the mean difference between the self-test group and the baseline group

]

.pull-right[

```{r}
mod1 <- lm(exam ~ method, data = dum_dat)
mod1
```

]


---
#  Dummy coding with `lm` (full results)

```{r}
summary(mod1)
```

---
#  Changing the baseline group  

+ The level that `lm` chooses as it's baseline may not always be the best choice
	+ You can change it using:

```{r, eval=FALSE}
contrasts(dum_dat$method) <- contr.treatment(3, base = 2)
```


	
+ `contrasts` updates the variable with the new coding scheme

+ `contr.treatment` Specifies that you want dummy coding

+ `3` is No. of levels of your predictor

+ `base=2` is the level number of your new baseline


---
#  Results using the new baseline 

.pull-left[
+ The intercept is the now the mean of the second group (Notes summarising)

+ `method1` is now the difference between Notes re-reading and Notes summarising

+ `method3` is now the difference between Self-testing and Notes summarising

]

.pull-right[

```{r}
contrasts(dum_dat$method) <- contr.treatment(3, base = 2)
mod2 <- lm(exam ~ method, data = dum_dat)
mod2
```

]

---
#  New baseline (full results)

```{r}
summary(mod2)
```

???
+ Note that the choice of baseline does not affect the R^2 or F-ratio



---
# Summary of today


---
class: center, middle
# Thanks for listening!