---
title: "<b>Week 8: Further diagnostics </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Weeks Learning Objectives
1. Specify the assumptions underlying a LM with multiple predictors
2. Assess if a fitted model satisfies the assumptions
3. Test and assess the effect of influential cases on LM coefficients and overall model evaluations

---
# Topics for today
+ Recap diagnostics for LM
+ Additional checks with multiple predictors
+ Additional metrics for assessing influential cases.

---
#  Three important features 

+ Regression outliers
	+ Cases for which there is a large discrepancy between their predicted y-value and their observed y-value
+ High leverage cases
	+ Cases with an unusual combination of predictor values
+ High influence cases
	+ Cases who are having a large influence on the regression results


---
#  Influence 

+ Cases with high influence, have a strong effect on the coefficients
	+ Occurs when a case has both high leverage and is a regression outlier
+ If we deleted such a case, the regression coefficients would change substantially
+ If a handful of influential cases are responsible for the regression results, the conclusions might not generalise very well


---
# Influence of coefficients
+ We can look at the effect of deleting a case:
	+ DFFit
		+ The difference between the predicted DV value for a case with versus without a case included
	+ DFbeta
		+ For each regression coefficient, the difference between that coefficient with and without a case included
	+ DFbetas
		+ A standardised version of DFbeta , obtained by dividing by an estimate of the standard error of the regression coefficient with the case removed

---
#  Influence on standard errors 

+ As well as having a strong influence on the regression coefficients, a case might be influential with respect to the standard errors of regression coefficients
+ Recall, the standard error for a regression slope in simple regression is:

+ ADD EQUATION

+ Where:
	+ SSE is the sum of squared error (i.e., )
	+ N is the sample size
	+ k is the number of predictors (=1 for simple regression)

---
#  Influence on standard errors 

+ ADD EQUATION

+ is the sum of squared deviations from each X value from the mean of X
+ This term implies that increasing the variance in X will decrease the standard error of
+ High leverage cases (which are far from the mean of X) can affect by increasing X variance
	+ If the case has high leverage but is not a regression outlier they can affect without necessarily affecting the regression coefficients

---
#  COVRATIO 
+ Influence on standard errors can be measured using the COVRATIO statistic
	+ COVRATIO value <1 show that precision is decreased (SE increased)  by a case
	+ COVRATIO value >1 show that precision is increased (SE decreased) by a case
+ Cases with COVRATIOS >1+[3( k +1)/ n ] or < 1-[3( k +1)/ n ] can be considered to have a strong influence on the standard errors
	+ For our violence example with n=95 , k=3 would, cases with a strong influence on the standard errors would have values :


---
#  COVRATIO in R 
+ COVRATIO values can be extracted using the covratio ( ) function


---
# Summary of today

---
# Next tasks

