---
title: "<b> Categorical Predictors: Effects (sum to zero) coding </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(car)
library(patchwork)
library(kableExtra)
```

# Weeks Learning Objectives
1. Understand the difference between dummy and sum-to-zero coding.

2. Understand the core principle of different coding schemes.

3. Interpret the output from a model using sum-to-zero coding.


---
# Overview of next 3 weeks
+ The next few weeks will focus on how we can work with linear models with categorical predictors to test different questions.

+ In particular, we will look at the type of questions typically asked as part of experiments.

+ We will see that the tools we saw in discussing model comparisons will be important.

+ As well the principles of categorical data coding discussed when talking about binary and dummy variables.
  + We are going to generalise some of these concepts, and then show how we can use the general principles to test questions.


---
# Recap on dummy coding (reference group)
+ Create $k$-1 dummy variables/contrasts
  + where $k$ is the number of levels of the categorical predictor. 

+ Assign reference group 0 on all dummies.

+ Assign 1 to the focal group for a particular dummy.

+ Enter the dummies into the linear model and they code the difference in means between the focal group/level and the reference.

---
# `Hospital` & `Treatment` data

+ **Condition 1**: `Treatment` (Levels: TreatA, TreatB, TreatC).
  
+ **Condition 2**: `Hospital` (Levels: Hosp1, Hosp2). 
  
+ Total sample n = 180 (30 patients in each of 6 groups).
  + Between person design. 

+ **Outcome**: Subjective well-being (`SWB`)
  + An average of multiple raters (the patient, a member of their family, and a friend). 
  + SWB score ranged from 0 to 20.

---
# The data
```{r}
hosp_tbl <- read_csv("hospital.csv", col_types = "dff")
hosp_tbl %>%
  slice(1:10)
```


---
class: center, middle
# General Principle 1
**We need to apply a constraint in order to test models with categorical data**


---
# Why do we need a reference group?
+ Consider our example and dummy coding.

+ We have three groups each given a specific Treatment A, B or C

+ We want a model that represents our data (observations), but all we "know" is what group an observation belongs to. So;

$$y_{ij} = \mu_i + \epsilon_{ij}$$

+ Where 
  + $y_{ij}$ are the individual observations
  + $\mu_i$ is the mean of group $i$ and
  + $\epsilon_{ij}$ is the individual deviation from that mean.


???
+ And this hopefully makes sense.
  + Given we know someone's group, our best guess is the mean
  + But people wont all score the mean, so there is some deviation for every person.


---
# Why do we need a reference group?
+ An alternative way to present this idea looks much more like our linear model:

$$y_{ij} = \beta_0 + \underbrace{(\mu_{i} - \beta_0)}_{\beta_i} + \epsilon_{ij}$$
+ Where 
  + $y_{ij}$ are the individual observations
  + $\beta_0$ is an estimate of reference/overall average
  + $\mu_i$ is the mean of group $i$ 
  + $\beta_1$ is the difference between the reference and the mean of group $i$, and
  + $\epsilon_{ij}$ is the individual deviation from that mean.

---
# Why do we need a reference group?
+ We can write this equation more generally as:

$$\mu_i = \beta_0 + \beta_i $$

+ or for the specific groups (in our case 3):

$$\mu_{treatmentA} = \beta_0 + \beta_{1A}$$

$$\mu_{treatmentB} = \beta_0 + \beta_{2B}$$

$$\mu_{treatmentC} = \beta_0 + \beta_{3C}$$

+ **The problem**: we have four parameters ( $\beta_0$ , $\beta_{1A}$ , $\beta_{2B}$ , $\beta_{3C}$ ) to model three group means ( $\mu_{TreatmentA}$ , $\mu_{TreatmentB}$ , $\mu_{TreatmentC}$ )

+ We are trying to estimate too much with too little.
    + This is referred to as under-identification.
    + We need to estimate at least 1 parameter less

---
# Constraints fix identification
+ Consider dummy coding.

+ Suppose we make Treatment A the reference. Then, 

$$\mu_{treatmentA} = \beta_0$$

$$\mu_{treatmentB} = \beta_0 + \beta_{2B}$$

$$\mu_{treatmentC} = \beta_0 + \beta_{3C}$$
+ **Fixed** ! 

+ We now only have three parameters ( $\beta_0$ , $\beta_{2B}$ , $\beta_{3C}$ ) for the three group means ( $\mu_{TreatmentA}$ , $\mu_{TreatmentB}$ , $\mu_{TreatmentC}$ ). 


> So when we code categorical variables, we need a constraint so that we can estimate our models.


---
class: center, middle
# Any questions?

---
# Why not always use dummy coding?

+ We might not always want to compare against a reference group.

+ We might want to compare to:
  + The overall or grand mean
  + Group 1 vs groups 2, 3, 4 combined
  + and on we go!

+ Let's consider the example of the grand mean... 

---
# Effects coding (sum to zero coding)

```{r, echo=FALSE, message=FALSE}
set.seed(1)
gpM <- hosp_tbl %>%
  group_by(Treatment) %>%
  summarise(
    SWB = round(mean(SWB))
  ) %>%
  mutate(
    GM = c(rep(mean(SWB),3))
  )

hosp_tbl %>%
  ggplot(., aes(x=Treatment, y=SWB, colour = Treatment, shape = Treatment)) +
  geom_jitter(width = 0.1, size = 2) + 
  geom_hline(yintercept = mean(hosp_tbl$SWB), lty = 2, colour = "darkgrey") +
  theme_classic() +
  geom_errorbar(data = gpM, width=0.8,aes(ymax=..y..,ymin=..y..))
```

---
# Sum to zero constraint

+ With dummy coding we had a reference group constraint, and the mean of that group was equal to the value of $\beta_0$, or 

$$\mu_{reference} = \beta_0$$ 

+ Alternately, we can apply what is referred to as the sum to zero constraint (again using example of three levels).

$$\beta_1 + \beta_2 + \beta_3 = 0$$

+ This constraints leads to the following interpretations:

+ $\beta_0$ is the grand mean (mean of all observations) or $\mu$

$$\beta_0 = \frac{\mu_1 + \mu_2 + \mu_3}{3} = \mu$$

+ $\beta_i$ are the differences between the coded group and the grand mean:

$$\beta_i = \mu_i - \mu$$

---
# Sum to zero constraint

+ Finally, we can get back to our group means from the coefficients as follows: 

$$\mu_1 = \beta_0 + \beta_1$$

$$\mu_2 = \beta_0 + \beta_2$$

$$\mu_3 = \beta_0 - (\beta_1 + \beta_2)$$

---
# OK, but how do we apply the constraint?

+ Answer, in the same way as we did with dummy coding.

+ We can create a set of sum to zero (sometimes called effect, or deviation) variables
  + Or the equivalent contrast matrix.
  
+ For effect code variables we:
  + Create $k-1$ variables
  + For observations in the focal group, assign 1
  + For observations in the last group, assign -1
  + For all other groups assign 0


---
# Comparing coding matrices

.pull-left[
```{r echo = FALSE}
tibble(
  Level = c("Treatment A", "Treatment B", "Treatment C"),
  D1 = c(0,1,0),
  D2 = c(0,0,1)
) %>%
  kable(.)%>%
  kable_styling(., full_width = F)
```

$$y_{ij} = \beta_0 + \beta_1D_1 + \beta_2D_2 + \epsilon_{ij}$$

]


.pull-right[

```{r echo = FALSE}
tibble(
  Level = c("Treatment A", "Treatment B", "Treatment C"),
  E1 = c(1,0,-1),
  E2 = c(0,1,-1)
) %>%
  kable(.)%>%
  kable_styling(., full_width = F)
```

$$y_{ij} = \beta_0 + \beta_1E_1 + \beta_2E_2 + \epsilon_{ij}$$

]

---
# Sum to zero/effects for group means

.pull-left[
```{r echo = FALSE}
tibble(
  Level = c("Treatment A", "Treatment B", "Treatment C"),
  E1 = c(1,0,-1),
  E2 = c(0,1,-1)
) %>%
  kable(.)%>%
  kable_styling(., full_width = F)
```

$$\mu_1 = \beta_0 + \beta_1$$

$$\mu_2 = \beta_0 + \beta_2$$

$$\mu_3 = \beta_0 - (\beta_1 + \beta_2)$$
]

.pull-right[
$$\mu_1 = \beta_0 + 1*\beta_1 + 0*\beta_2 = \beta_0 + \beta_1$$

$$\mu_2 = \beta_0 + 0*\beta_1 + 1*\beta_2 = \beta_0 + \beta_2$$

$$\mu_3 = \beta_0 -1*\beta_1 -1*\beta_2 = \beta_0 - \beta_1 -\beta_2$$

+ Now we will look practically at the implementation and differences

]


---
# Group Means

```{r}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3),
    sd = round(sd(SWB),1),
    N = n()
  )
```
+ ADD THE GRAND MEAN, PRESENT LIKE A SINGLE ROW CROSS-TAB

---
# Effects (sum to zero) model

+ We need to change the contrast scheme from default.

```{r}
contrasts(hosp_tbl$Treatment) <- contr.sum 
contrasts(hosp_tbl$Treatment)
```

---
# Effects (sum to zero) model
```{r}
summary(lm(SWB ~ Treatment, data = hosp_tbl))
```


---
# Effects (sum to zero) model

.pull-left[
```{r, echo=FALSE}
Eres <- summary(lm(SWB ~ Treatment, data = hosp_tbl))
round(Eres$coefficients[,1],3)
```

+ Coefficients from group means


$$\beta_0 = \frac{\mu_1 + \mu_2 + \mu_3}{3}$$ 

$$\beta_1 = \mu_1 - \mu$$

$$\beta_2 = \mu_2 - \mu$$

]


.pull-right[

```{r, echo=FALSE}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3)
  ) %>%
  mutate(
    Gmean = round(mean(mean),3)
  ) %>%
  kable(.) %>%
  kable_styling(., full_width = F)
```

+ CHANGE THIS TO SAME TABLE AS ABOVE

]


---
# Effects (sum to zero) model

.pull-left[
```{r, echo=FALSE}
Eres <- summary(lm(SWB ~ Treatment, data = hosp_tbl))
round(Eres$coefficients[,1],3)
```

+ Group means from coefficients:

$$\mu_1 = \beta_0 + \beta_1$$

$$\mu_2 = \beta_0 + \beta_2$$

$$\mu_3 = \beta_0 - (\beta_1 + \beta_2)$$
]


.pull-right[

```{r, echo=FALSE}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3)
  ) %>%
  mutate(
    Gmean = round(mean(mean),3)
  ) %>%
  kable(.) %>%
  kable_styling(., full_width = F)
```

+ CHANGE THIS TO SAME TABLE AS ABOVE

]


---
# The wide world of contrasts 
+ We have now seen two examples of coding schemes (dummy and effect).

+ There are **lots** of different coding schemes we can use for categorical variables to make different comparisons.
  + If you are interested, see the excellent resource on [UCLA website](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)

+ But always remember...

**The data is the same, the tested contrast differs**

---
class: center, middle
# Questions

---
class: center, middle
# General Principle 2
**The data is the same, the tested contrasts differ**


---
# The data is the same, the tested contrasts differ

+ Run both models:

```{r}
contrasts(hosp_tbl$Treatment) <- contr.treatment
m_dummy <- lm(SWB ~ Treatment, data = hosp_tbl)

# Change the contrasts and run again
contrasts(hosp_tbl$Treatment) <- contr.sum
m_zero <- lm(SWB ~ Treatment, data = hosp_tbl)

```

+ Create a small data set:

```{r}
treat <- tibble(Treatment = c("TreatA", "TreatB", "TreatC"))
```

---
# The data is the same, the tested contrasts differ

+ Add the predicted values from our models

```{r}
treat %>%
  mutate(
    pred_dummy = predict(m_dummy, newdata = .),
    pred_zero = predict(m_zero, newdata = .)
  )
```

+ No matter what coding or contrasts we use, we are still modelling the group means!



---
# Summary of today

+ We have considered different ways in which we can code categorical predictors.

+ Take home:
  + Use of coding matrices allows us to compare groups (or levels) in lots of ways.
  + Our $\beta$'s will represent differences in group means.
  + The scheme we use determines which group or combination of groups we are comparing.
  + **In all cases the underlying data is unchanged.**

+ This makes coding schemes a very flexible tool for testing hypotheses.


---
class: center, middle
# Thanks for listening! Any questions?
