---
title: "<b> Preregistration and Analytic Freedom </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```




# Learning Objectives

- What and why would we preregister studies. 

- Limitations to Preregistration and the relationship between Preregistration and exploratory analyses

- Concrete example of the benefits and difficults of preregistration



---
# Title

Content

---
# Title

- Point one
  - Subpoint one
  - Subpoint two
  - Subpoint three
  
--

- Another point


---
# New slide

- More content

--

- Even more content

- Hello!


---
# Summary

- The summary of this lecture.

- Bye bye!

