---
title: "Linear Model Analysis: Study Brief"
author: "Emma Waterston"
format: 
  html:
    toc: true
    number_sections: false
    toc-location: left
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)
source('_theme/theme_quarto.R')
```

# Data & Research Questions {#task}

> Conduct and report on an analysis that addresses the research aims.  
The data is available at: <https://uoepsy.github.io/data/FOMOdataset.csv>   


### Study Background and Aims 

The data used for this write-up exercise are simulated, drawing on recent work on Fear of Missing Out (FoMO), socio-demographic factors, and the Big Five personality traits. The simulated data are based on the findings of this work, and acted to expand upon the methods and results reported in the following paper:

Rozgonjuk, D., Sindermann, C., Elhai, J. D., & Montag, C. (2021). Individual differences in Fear of Missing Out (FoMO): Age, gender, and the Big Five personality trait domains, facets, and items. *Personality and Individual Differences, 171*, 110546.  [https://doi.org/10.1016/j.paid.2020.110546](https://doi.org/10.1016/j.paid.2020.110546)

*Note you are not expected to provide an introduction, you do not have to read this article.*

### Method & Procedure

In the current study, participants were invited to an online study investigating the associations among FoMO, socio-demographic factors, and personality. The final sample comprised 3370 people. Participants completed a FOMO scale and a personality inventory. The 10-item FOMO scale measured the extent of experiencing apprehension regarding missing out on interesting events of others on a 5-point scale (1 = “not at all true of me” to 5 = “extremely true of me”), producing a possible range of scores between 10 and 50. The Big Five Inventory (BFI) is a 45-item personality assessment questionnaire (note that only 43 items were used to match the study above) that uses a five-point response scale (1 = “very inapplicable” to 5 = “very applicable”). The BFI consists of five domains: Neuroticism (8 items; possible range of scores 8-40), Extraversion (8 items; possible range of scores 8-40), Openness to Experience (10 items; possible range of scores 10-50), Agreeableness (8 items; possible range of scores 8-40), and Conscientiousness (9 items; possible range of scores 9-45). We extended the aforementioned study to include an extra socio-demographic variable - a measure of popularity on social media based on the number of followers. Unlike the original study, we do not have measures of gender, education level, or specific country of residence.   


### Research Aim & Questions

> **Research Aim**
> 
> Explore the associations among Fear of Missing Out (FoMO), age, social media networks, and the Big Five personality traits. 
>
> **Research Questions**
> 
> + RQ1: Does age predict FoMO?
> + RQ2: Does the number of Instagram followers explain a significant amount of variance in FoMO over and above age?
> + RQ3: Does personality predict FoMO?


### Data Dictionary

The data in `FOMOdataset.csv` contain eight attributes collected from a simulated sample of $n=3370$ hypothetical individuals across the UK, and include: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
fomo <- read_csv("https://uoepsy.github.io/data/FOMOdataset.csv")
tibble(
Variable = names(fomo),
Description = c("FoMO Score (as measured by the 10-item FoMO scale)", "Age (in years)", "Score on personality items assessing Neuroticism from the Big Five Inventory (BFI)", "Score on personality items assessing Extraversion from the Big Five Inventory (BFI)", "Score on personality items assessing Openness from the Big Five Inventory (BFI)", "Score on personality items assessing Agreeableness from the Big Five Inventory (BFI)", "Score on personality items assessing Conscientiousness from the Big Five Inventory (BFI)", "Total Number of Instagram Followers")
) |> gt::gt()
```
 
