---
title: "<b>Week 4: Binary Predictors </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives
1. Understand the meaning of regression coefficients in the case of a categorical predictor.
2. Be able to state the assumptions underlying a linear regression model.
3. Understand how to assess if a fitted model satisfies the regression assumptions
4. Understand how to use transformations when the model violates the regression assumptions.

---
# Topics for today

---
#  Recap: Categorical variables 
+ Categorical variables can only take discrete values
	+ E.g., animal type: 1= duck, 2= cow, 3= wasp
+ They are mutually exclusive
		+ No duck-wasps or cow-ducks!
+ In R, categorical variables should be of class `factor`
  + The discrete values are `levels`
  + Levels can have numeric values and labels

---
# Recap: Binary variable
+ Binary variable is a categorical variable with two levels.
+ Traditionally coded with a 0 and 1
  + Referred to as dummy coding
+ Why 0 and 1?
  + We will come to this a little later

---
#  LM with binary predictors 
+ Returning to our example...
	+ We have our continuous outcome (Score, Y)
	+ We have a binary predictor (Class, ???)

+ use equation notion to show values

---
#  LM with binary predictors 
+ We could include XXX variable in a simple regression, just like any other IV:


---
# In R
+ Run it & show result


---
# Interpretation
+ Show means of groups
+ what do we notice?

---
# Interpretation
+ b0
+ b1
+ How does this work?


---
#  Equations for each group 


---
#  Visualize the model
+ show data
+ show only two values for X
+ everyone in group best represented as mean
+ 0 = link to interpreation of intercept + group mean
+ b1 as difference "slope"

---
#  Evaluation of model and significance of b 
+ R 2 and F-ratio interpretation are identical to their interpretation in models with only continuous IVs
+ We assess the statistical significance of regression coefficients for categorical IVs in exactly the same way as we did for continuous IVs:
+ We use the standard error of the coefficient to construct:
	+ *t-* tests and associated *p-* values  for the coefficient
	+ Confidence intervals around the coefficient

---
# Hold on...it's a t-test


--- 
# Standardizing b

---
# Summary of today

---
# Next tasks