---
title: "Quarto template for DAPR2 lecture slides"
author: "Firstname Lastname"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(patchwork)
source('_theme/theme_quarto.R')
```


# Course Overview {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r echo = FALSE, results='asis', warning = FALSE}
block1_name = "Introduction to Linear Models"
block1_lecs = c("Intro to Linear Regression",
                "Interpreting Linear Models",
                "Testing Individual Predictors",
                "Model Testing & Comparison",
                "Linear Model Analysis")
block2_name = "Analysing Experimental Studies"
block2_lecs = c("Categorical Predictors & Dummy Coding",
                "	Effects Coding & Coding Specific Contrasts",
                "Assumptions & Diagnostics",
                "Bootstrapping",
                "	Categorical Predictor Analysis")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block1_name,block2_name,block1_lecs,block2_lecs,week=3)
```

:::

::: {.column width="50%"}

```{r echo = FALSE, results='asis', warning = FALSE}
block3_name = "Interactions"
block3_lecs = c("Interactions I",
                "Interactions II",
                "Interactions III",
                "Analysing Experiments",
                "Interaction Analysis")
block4_name = "Advanced Topics"
block4_lecs = c("Power Analysis",
                "Binary Logistic Regression I",
                "Binary Logistic Regression II",
                "Logistic Regression Analysis",
                "	Exam Prep and Course Q&A")

source("https://raw.githubusercontent.com/uoepsy/junk/main/R/course_table.R")

course_table(block3_name,block4_name,block3_lecs,block4_lecs,week=0)
```

:::

::::

## Some code

::: {.fragment}

```{r}
pi
```

:::

<br>

::: {.fragment}

Multiple things can appear together as long as they're within a single `fragment` element.

<br>

```{r}
1+2
```

:::

<br>

::: {.fragment}

And some more `verbatim` text.

:::

# This week

## We're gonna ...

:::: {.columns}

::: {.column width="50%" .incremental}
- Eat spaghetti
- Drink wine
:::

::: {.column width="50%" .incremental}
- Drink coffee
- Eat a hamburger
:::

::::


# Broad ideas 

## Two questions

:::{.incremental}

- Q: To do anything with [construct Y], how do we get one number to represent an observation of Y?  
    - is one number enough - are $y1,y2,...,yk$ really unidimensional?  
    
<br>
    
- Q: How does [set of scores $y1,y2,...,yk$] get at [construct Y]?  
    - is it a reliable measure?
    - are the variables equally representative?
    - is there just one dimension to Y or are there multiple?

:::



## scoring multi-item measures


::::{.columns}
:::{.column width="50%"}
**scale scores**  
add 'em all up, you've got Y  

- clinically 'meaningful'?  
- but only 'meaningful' if underlying model holds (which it almost definitely doesn't!)  
    
:::

:::{.column width="50%"}
**dimension reduction**  
identify dimensions that capture how people co-vary across across the items

- PCA: reduce to set of orthogonal dimensions sequentially capturing most variability

- FA: develop theoretical model of underlying dimensions (possibly correlated) that explain variability in items  

:::
::::


## Understanding multi-item measures

::::{.columns}
:::{.column width="50%"}

:::

:::{.column width="50%"}
**dimension reduction**  

:::{style="opacity:.4"}
identify dimensions that capture how people co-vary across across the items

- PCA: reduce to set of orthogonal dimensions sequentially capturing most variability
::: 
- FA: develop theoretical model of underlying dimensions (possibly correlated) that explain variability in items  

:::
::::

## Understanding multi-item measures

$$
\begin{align}
\text{Outcome} &=& \text{Model} &\quad + \quad& \text{Error} \\
\quad \\
\text{observed correlation} &=& \text{factor loadings and} &\quad + \quad& \text{unique variance for} \\
\text{matrix of items}& &\text{factor correlations} &\quad + \quad& \text{each item} \\
\end{align}
$$

# Section 2

## Lorem ipsum?

Lorem ipsum.

# End!  









