---
title: "Interactions: Study brief"
author: "Emma Waterston"
format: 
  html:
    toc: true
    number_sections: false
    toc-location: left
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(simglm)

source('_theme/theme_quarto.R')
```


> Conduct and report on an analysis that addresses the research aims.  
The data is available at: <https://uoepsy.github.io/data/laptop_vs_longhand.csv>.



# Study background and aims 

The data used for this write-up exercise are simulated, drawing on a research paper that explored the association between student note taking and performance when answering different types of questions.
The simulated data are loosely based on the findings of this work, and act to expand upon the methods and results reported in the paper:

Mueller, P. A., & Oppenheimer, D. M. (2014). The pen is mightier than the keyboard: Advantages of longhand over laptop note taking. Psychological Science, 25(6), 1159–1168. <https://doi.org/10.1177/0956797614524581>

*NOTE: You are not expected to write an introduction, so you do not have to read this article.*


# Method and procedure

In the current study, participants were invited to take part in a study investigating the the medium of note taking and study time on test scores.
The sample comprised 160 students who took notes on a lecture via one of two mediums—either on a laptop or longhand (i.e., using pen and paper).
After watching the lecture and taking notes, they were then randomly allocated to one of four study time conditions, either engaging in no, minimal, moderate, or extensive study of the notes taken on their assigned medium.
After engaging in study for their allocated time, participants took a test on the lecture content.
The test involved a series of questions, where participants could score a maximum of 100 points.


# Research aim and questions

### Research Aim

Explore the associations among study time and note-taking medium on test scores.
 

### Research Questions

- RQ1: Do differences in test scores between study conditions differ by the note-taking medium used?
- RQ2: Are there differences in test scores between participants when comparing pairs of study and note-taking conditions? If so, what are these specific differences?


# Data dictionary

The data in `laptop_vs_longhand.csv` contain three attributes collected from a simulated sample of $n=160$ hypothetical individuals, and includes:

```{r echo=FALSE, message=FALSE, warning=FALSE}
notes <- read_csv("https://uoepsy.github.io/data/laptop_vs_longhand.csv")
tibble(
Variable = names(notes),
Description = c("Test Score (0-100)", "Medium of note-taking (levels = Longhand, Laptop)", "Study time (levels = No, Minimal, Moderate, Extensive)")
) %>% gt::gt()
```


The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/laptop_vs_longhand.csv') %>%  head %>% gt::gt()
```


