---
title: "<b>Exam Prep</b>"
subtitle: "Data Analysis for Psychology in R 2"
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


# Outline

+ Things to take into exam  

+ Process  

+ Structure of exam  

+ Exam *'strategy'*  

+ Example Questions  

---
# Things You Need for the Exam

.pull-left[

**What you need to take**

+ Student Card

+ Pens (blue or black ink)  

+ Pencils and rubber

+ Ruler
  + If you like, not really needed, but might be useful 
  
+ Calculator
  + See list on Learn of permitted calculators  
  + **Phone/watch cannot be your calculator**  

]

.pull-right[

**What you will be given**

+ Attendance slip    

+ The exam question paper (inc. equation & formulae section)    

+ A MCQ answer sheet for Section A questions    

+ An answer book for Section B questions    

]

---
# Process

+ You need to check your calendar [https://exams.is.ed.ac.uk/](https://exams.is.ed.ac.uk/)

+ Arrive in plenty of time (approx. 30 mins before exam starts)  

+ There will be direction to room    

+ Follow instructions for where to leave jackets, bags, phones, etc.    

+ Take all stationary, calculator, and **student ID card** to your desk  

+ Invigilators will give you exam instructions  

---

# Filling in Front Covers

<br>

.center[

<iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_uwsdz3vt&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_ft64ev29" width="320" height="300" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to complete an examination attendance form"></iframe>   <iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_spaem93h&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[mediaProxy.mediaPlayFrom]=96&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_pim3wi24" width="320" height="300" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to prepare barcoded exam scripts and MCQ sheets"></iframe>   <iframe id="kaltura_player" src="https://cdnapisec.kaltura.com/p/2010292/sp/201029200/embedIframeJs/uiconf_id/32599141/partner_id/2010292?iframeembed=true&playerId=kaltura_player&entry_id=1_ms01z6h1&flashvars[streamerType]=auto&amp;flashvars[localizationCode]=en&amp;flashvars[sideBarContainer.plugin]=true&amp;flashvars[sideBarContainer.position]=left&amp;flashvars[sideBarContainer.clickToClose]=true&amp;flashvars[chapters.plugin]=true&amp;flashvars[chapters.layout]=vertical&amp;flashvars[chapters.thumbnailRotator]=false&amp;flashvars[streamSelector.plugin]=true&amp;flashvars[EmbedPlayer.SpinnerTarget]=videoHolder&amp;flashvars[dualScreen.plugin]=true&amp;flashvars[Kaltura.addCrossoriginToIframe]=true&amp;&wid=1_j0xl893i" width="320" height="300" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" frameborder="0" title="How to fill in the cover of an exam script book?"></iframe>   

]
---
# Things You Can Take Away

<center>
<br><br>
<h2><b>NOTHING*!</b></h2>
*<i>Other than the things you bring in with you like pens, pencils, calculator</i>

<br><br>
<b>
When you finish, you must leave all papers (question paper (inc. equation & formulae section), answer sheets/books) in the exam hall.  
</b></center>

---
# Exam Structure

.pull-left[
**Section A - 30 marks**

+ 15 MCQs

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from

]
.pull-right[
**Section B - 70 marks**  

+ 8 Questions  

+ Marks range from 5 to 12 marks per question:  
    + 1 x 5 marks
    + 1 x 6 marks
    + 1 x 7 marks
    + 1 x 8 marks
    + 2 x 10 marks
    + 2 x 12 marks

+ On range of topics  
]

---
# Exam Strategy: Section A

.pull-left[

**Section A - 30 marks**

+ 15 MCQs  

+ 2 marks per MCQ  

+ On all topics   

+ Similar style to weekly quizzes - all questions have 4 proposed answers to choose from
]

.pull-right[

**Suggested Approach**
+ Similar to quiz questions on Learn  
+ Spend 15-30 minutes on MCQs:
  + Approx. 1-2 minutes per MCQ
  + You can answer these quickly at the end
  + Do the ones you know, then move on
  + Keep track of how many you have not done, so you can go back to them at the end
  
]

---
# Exam Strategy: Section B


.pull-left[
**Section B - 70 marks**  

+ 8 questions, where marks range from 5 to 12 marks per question:  
    + 1 x 5 marks
    + 1 x 6 marks
    + 1 x 7 marks
    + 1 x 8 marks
    + 2 x 10 marks
    + 2 x 12 marks

+ On range of topics, where some questions will involve:
    + Short calculations
    + Asking you to explain statistical concepts  
    + Giving you a research design/question and task you with detailing an appropriate analysis plan
    + Giving you model output and ask you to provide interpretation  
]
.pull-right[

**Suggested Approach** 

+ 70% of marks in this section
+ Spend more time on higher mark questions
    + Marks arranged from highest to lowest marks (i.e., Q1 = 12 marks, ..., Q8 = 5 marks)
+ Skim read all 8 questions
    + Prioritise highest mark questions
    + **BUT** also do the ones you know first
]

---
# General Tips for Approaching Exam

+ Split your time according to the marks for the sections and questions  
  
+ Make an attempt to answer all questions
    + Section A: Guess if you need to - don't leave an MCQ answer blank 
    + Section B: A brief answer is better than no answer  
     
+ Section B: Read the questions very carefully. For the longer questions, the question will include information on the things we expect your answers to comment on. Questions in Section B have the number of marks indicated next to each question.
    + As a **(very rough)** rule of thumb, a mark will typically correspond to a sentence (so for a 5-mark question a 5-6 sentence answer would likely be appropriate)  
    
+ You will not necessarily need to directly use all the equations on the equation sheet, but they may help you answer questions

---
# Example Questions

+ Example questions provided as this weeks lab exercises 

+ Video walkthrough of how to approach questions shared on Friday 

---
class: center, middle, animated, rotateInDownLeft

# **Questions?**

---
class: center, inverse, middle, animated, rotateInDownLeft

# End
