---
title: "Week 11: Practical Example"
output: bookdown::html_document2
---

```{r, echo = F, message=FALSE}
library(tidyverse)
library(gt)
library(kableExtra)
library(car)
library(emmeans)
library(sjPlot)

dat <- read.csv('https://uoepsy.github.io/data/DapR2_S1B2_PracticalPart1.csv')
dat2 <- read.csv('https://uoepsy.github.io/data/DapR2_S1B2_PracticalPart2.csv')
```

# Overview of the Week

This week, we'll be reviewing what we've learned in weeks 7-10 and applying it to a practical example. Specifically, we'll cover:

1) Multiple linear regression with categorical variables;
2) Dummy, effects, and manual contrast coding;
3) Checking model assumptions & diagnostics;
4) Bootstrapping


Our example is based on the paper _[Class Attendance in College: A Meta-Analytic Review of the Relationship of Class Attendance With Grades and Student Characteristics](https://www.ucsd.ac.uk/wp-content/uploads/Benefits-of-attendance-for-students.pdf)_ This paper looks at the association between class attendance and a range of other variables (such as student personality, academic performance, class marks, etc.) Imagine that we've read that paper and we want to perform further investigation on the variables that contribute to attendance. Specifically, we will investigate a range of categorical variables here. We'll also use our data to try and replicate the association between attendance and marks that is found in this paper.

During this week's example, we will conduct two separate, but related, studies. In the first, we are interested in exploring possible predictors of attendance in university courses. In the second, we will investigate the association between attendance and final marks in our sample. 


# Study 1

We collected data from 397 university students across years 1-4, as well as students in MSc and PhD programs. From these students, we gathered data on class time, the use of online materials, and student's conscientiousness. 

```{r, echo = F}
tibble(Variable = names(dat),
Description = c("Participant ID number", "Total attendance in days", "Level of conscientiousness (levels = Low; Moderate; High)", "Class time (levels = 9AM; 10AM; 11AM; 12PM; 1PM; 2PM; 3PM; 4PM)", "Frequency of access to online course materials (levels = Rarely; Sometimes; Often)", "Year in university (levels = Y1; Y2; Y3; Y4; MSc; PhD)")
) %>% gt()|> opt_horizontal_padding(scale = 3)
```

We want to use these data to investigate the following research questions:

**Research Question 1:** Can student's conscientiousness, frequency of access to online materials, and year in University predict course attendance?

**Research Question 2:** Is the time at which the class is scheduled associated with student attendance?


## Setup

First, let's load the data and all necessary packages. Note that by using the `stringsAsFactors` argument in the `read.csv` function, all of our character variables are automatically imported as factors, which is very handy when you're working with a dataset with lots of factors in it.

```{r, eval = F}
library(tidyverse)
library(kableExtra)
library(car)
library(sjPlot)
library(emmeans)

dat <- read.csv('https://uoepsy.github.io/data/DapR2_S1B2_PracticalPart1.csv', stringsAsFactors = T)
```


## Checking the Data

First, we'll have a look at our data using the `summary` function:

```{r}
  summary(dat)
```
You'll notice that we have a continuous outcome variable, `Attendance`, and discrete predictor variables, `Conscientiousness`, `Time`, `OnlineAccess` and `Year`. 

We'll also have a look at how our variables are distributed, using a histogram for our continuous outcome and bar plots for our factors:

```{r}
ggplot(dat, aes(Attendance)) + geom_histogram(colour = 'black', binwidth = 5)
```

```{r}
ggplot(dat, aes(Conscientiousness, fill = Conscientiousness)) + geom_bar() +
  theme(legend.position = 'none')
```

```{r}
ggplot(dat, aes(Year, fill = Year)) + geom_bar() +
  theme(legend.position = 'none')
```

```{r}
ggplot(dat, aes(Time, fill = Time)) + geom_bar() +
  theme(legend.position = 'none')
```


## Investigating RQ 1

Given the first research question we've specified, **Can student's conscientiousness, frequency of access to online materials, and year in University predict course attendance?**, we'll use the following model:

$$Attendance \sim Conscientiousness + OnlineAccess + Year$$

### Dummy Coding
Between the 3 predictors, we have 12 levels. Before we run our model, we have to make a few decisions in terms of the coding and baseline comparisons we'll be making across predictors. We decide to use dummy coding. We'll use be using `Moderate` as the baseline level for `Conscientiousness` and `Sometimes` as the baseline level for `OnlineAccess`. We'll keep `Y1` as the baseline level for `Year`. We can use the `factor` function to order our levels accordingly:

```{r}
dat$Conscientiousness <- factor(dat$Conscientiousness, levels = c('Moderate', 'Low', 'High'))

summary(dat$Conscientiousness)
```


```{r}
dat$OnlineAccess <- factor(dat$OnlineAccess, levels = c('Sometimes', 'Rarely', 'Often'))

summary(dat$OnlineAccess)
```

```{r}
dat$Year <- factor(dat$Year, levels = c('Y1', 'Y2', 'Y3', 'Y4', 'MSc', 'PhD'))

summary(dat$Year)
```

Now that we have this done, we should end up with a model that looks like the one below. Note that some variable names have been shortened for sizing purposes:

$$Attendance \sim \beta_0+\beta_1Consc_{low} + \beta_2Consc_{high} + \beta_3Online_{rarely} + \beta_4Online_{often} + \beta_5Year_{Y2} + \beta_6Year_{Y3} + \beta_7Year_{Y4} + \beta_8Year_{MSc} + \beta_9Year_{PhD} + \epsilon$$

Now we can run our model:

```{r}
m1 <- lm(Attendance~Conscientiousness+OnlineAccess+Year, dat)
summary(m1)
```

This looks interesting, but before we interpret, let's check assumptions:

**Linearity** 
We can assume linearity when all x-variables entered into the model are categorical (see [here](https://www.bookdown.org/rwnahhas/RMPH/mlr-linearity.html))

**Independence of Errors**
We are using between-subjects data, so we'll also assume independence of our error terms. 

**Normality of Residuals**
We can check this using histograms and QQ plots:

```{r}
hist(m1$residuals)
```

```{r}
plot(m1, which = 2)
```


Tails are a bit fatter than we would expect (particularly on the right), but overall, this looks pretty good. No major concerns.

**Equality of Variance (Homoscedasticity)**
We can check for heteroscedasticity using residuals vs predicted values plots. We can get these using the `residualPlot` from the `car` package:

```{r}
residualPlot(m1)
```


Looks good! Now we can interpret the model.

```{r}
summary(m1)
```

<br>


### Effects Coding

To get some practice with coding schemes, let's run the same model with effects coding. 

First, let's consider which level we want to drop in each case. Perhaps we're more interested in how the "extreme" groups differ from average. We'll drop the `Moderate` and `Sometimes` levels from the `Conscientiousness` and `OnlineAccess` variables, respectively. We'll drop `Y3` from our `Year` variable. To do this, we'll need to reorder our variables accordingly:

```{r}
dat$Conscientiousness <- factor(dat$Conscientiousness, levels = c('Low', 'High', 'Moderate'))

summary(dat$Conscientiousness)
```

```{r}
dat$OnlineAccess <- factor(dat$OnlineAccess, levels = c('Rarely', 'Often', 'Sometimes'))

summary(dat$OnlineAccess)
```

```{r}
dat$Year <- factor(dat$Year, levels = c('Y1', 'Y2', 'Y4', 'MSc', 'PhD', 'Y3'))

summary(dat$Year)
```

We also need to specify effects coding using the `contr.sum` function:

```{r}
contrasts(dat$Conscientiousness) <- contr.sum
contrasts(dat$OnlineAccess) <- contr.sum
contrasts(dat$Year) <- contr.sum
```

Now we can run our model. Keep in mind that we don't need to recheck assumptions. The data are the same, it's just the comparisons that differ.

```{r}
m2 <- lm(Attendance~Conscientiousness+OnlineAccess+Year, dat)

summary(m2)

```

Before I include the model summary in my write-up, I have to format it into a nicer looking table. 

```{r}
tab_model(m2, pred.labels = c('Intercept', 'Low Conscientiousness', 'High Conscientiousness', 'Online Access - Rarely', 'Online Access - Often', 
                              'Y1', 'Y2', 'Y4', 'MSc', 'PhD'),
          title = 'Regression Results - Predictors of Student Attendance')
```

<br>

## Investigating RQ 2
Our second research question is **Is the time at which class is scheduled associated with student attendance?**. To test this, we'll use the `emmeans` function to write our own contrasts. We're going to be testing the following model:

$$Attendance \sim Time$$

First, let's relevel our `Time variable` so that it's in chronological order:

```{r}
dat$Time <- factor(dat$Time, levels = c('9AM', '10AM', '11AM','12PM', '1PM', '2PM', '3PM', '4PM'))

summary(dat$Time)
```

We'll look at how `Time` is distributed using a bar plot: 

```{r}
ggplot(dat, aes(Time, fill = Time)) + geom_bar() + 
  theme(legend.position = 'none')
```


First let's run our model and check assumptions. Remember, we don't need to check linearity or independence.

**Normality of residuals:**
```{r}
m3 <- lm(Attendance~Time, dat)

hist(m3$residuals)
```

```{r}
plot(m3, which = 2)
```


The distribution is a bit flat, but if you'll remember from class, distributional problems are usually not a strong concern (and we can always bootstrap for extra certainty!). Heteroscedasticity is much more of a problem. Let's check that using our residuals by predicted values plot.

```{r}
residualPlot(m3)
```

This looks great! We're going to keep our model as is.


Now let's specify our contrasts with emmeans:

```{r}
timeMean <- emmeans(m3, ~Time)
```


```{r}
plot(timeMean)
```

I'd guess we're going to see some significant differences here. Let's say, however, we're less worried about specific times and more about times of day in general. We decide to look at midday compared to earlier (9AM and 10AM) and later (3PM and 4PM) classes. Before running our analysis, we have to manually group our levels:

```{r}
levels(dat$Time)
timeComp <- list('Middle of the Day vs Early or Late' = c(-1/4,-1/4, 1/4, 1/4, 1/4, 1/4, -1/4, -1/4))
```

In this case, our $H_0$ is 

$$\frac{1}{4}(\mu_3+\mu_4+\mu_5+\mu_6) - \frac{1}{4}(\mu_1+\mu_2+\mu_7+\mu_8) = 0$$

```{r}
timeTest <- contrast(timeMean, timeComp)

timeTest

confint(timeTest)
```

# Study 2

We collected attendance and final mark data from 200 university students. 

```{r, echo = F}
tibble(
Variable = names(dat2),
Description = c("Final Mark in points", "Total attendance in days")
) %>% gt()|> opt_horizontal_padding(scale = 3)
```

We want to use these data to investigate the following research question:

**Research Question:** Can student attendance be used to predict marks?

Our hypotheses in this case, can be stated in the following way:

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$

## Study 2 - Setup

First, let's load the data.

```{r, eval = F}
dat2 <- read.csv('https://uoepsy.github.io/data/DapR2_S1B2_PracticalPart2.csv', stringsAsFactors = T)
```


## Study 2 - Checking the Data

First, let's have a look at the data:

```{r}
summary(dat2)
```
You'll notice that here we have two continuous outcome variables, `Attendance` and `Marks`. 

Let's look at how our variables are distributed, using histograms:

```{r, message = F}
ggplot(dat2, aes(Attendance)) + geom_histogram(colour = 'black', binwidth = 2)
```

Ok, that looks a bit weird...but let's keep going.

```{r}
ggplot(dat2, aes(Marks)) + geom_histogram(colour = 'black', binwidth = 4)
```

## Study 2 - Investigating the RQ

We'll fit the following model to investigate our research question:

$$Marks \sim \beta_0 + \beta_1Attendance + \epsilon$$

Let's run our model, then we'll check assumptions:

```{r}
m4 <- lm(Marks~Attendance, dat2)
summary(m4)
```

This model looks highly significant, but we'll have to check assumptions before interpretation, especially given how the distribution of attendance looked.

**Linearity** 
It's a simple regression, so we can just look at the scatterplot:

```{r}
ggplot(dat2, aes(Attendance, Marks)) + geom_point() + 
  geom_smooth(method = 'lm') + 
  geom_smooth(method = 'loess', colour = 'red')
```

There's a very slight curve in the loess line, but based on the points, I think that's due to possible issues in the distribution of the observations rather than a nonlinear relationship. We'll say linearity assumption is not violated.

**Independence of Errors**
We are using between-subjects data, so we'll assume independence of our error terms. 

**Normality of Residuals**
Let's check with a histogram and QQ plot:

```{r}
hist(m4$residuals)
```

A possible outlier to the right, but not bad...

```{r}
plot(m4, which = 2)
```

Ok, this actually looks alright, I think. Just the slight tail to the right.

**Equality of Variance (Homoscedasticity)**
We'll use the `residualPlot` to check for heteroscedasticity:

```{r}
residualPlot(m4)
```

Oh no! It's violated! We can't trust the SEs in our model! We'll need to bootstrap. We can do this using the `Boot` function in the `car` package. Let's resample 1000 times.

## Bootstrapping - Study 2

```{r}
k <- 1000
boot_m4 <- Boot(m4, R = k)

summary(boot_m4)
```

Let's calculate confidence intervals to test our hypothesis:

```{r}
confint(boot_m4)
```

We can make this look nicer for our write-up:

```{r}
confint(boot_m4) %>%
  kable(digits = 2, caption = 'Bootstrapped 95% Confidence Intervals') %>%
  kable_styling(full_width = F)
```

Because our confidence interval for the beta associated with attendance does not include 0, we can be more certain that attendance is, in fact, a significant predictor of marks. 

# Next steps

The lab this week involves writing up the results from the three RQs above

* Sample analysis code similar to the above provided in the lab materials
* This analysis is not complete:
  * Lab analysis code includes more examples of descriptive statistics
  * We have looked at model assumptions here, but not checked for influential data points and multicollinearity
  
<br>

