---
title: "Exam prep"
author: "DAPR2 team"
---

# What we'll talk about today

## What we'll talk about today

- FAQs around final course marks, progression to Honours, etc.

- What to expect from the exam

  - What you should bring into (and out of) the exam room
  
  - The process on exam day
  
- The exam itself

  - How the exam is structured
  
  - Recommended strategy for approaching the exam
  
- Getting prepared
  
  - Tips and methods for learning and studying
  
  - This week's lab: Example questions in a mock exam


# Course mark FAQs

## How your final course mark is calculated

The final course mark draws on three components:


| Component                      | Percentage of final mark  |
| ------------------------------ | ------------------------- |
| Weekly quizzes (best 14 of 18) | 10%                       |
| Assessed group report          | 30%                       |
| Exam                           | 60%                       |


<br>

For example, imagine:

```{r}
quiz_mark   <- 75
report_mark <- 52
exam_mark   <- 60

(quiz_mark * .10) + (report_mark * .30) + (exam_mark * .60)
```

**this ^ would be your final mark, given those marks on each component**


## FAQs about marks, progression {.smaller}


:::{.dapr2callout}
**What overall course grade do I need to pass DAPR2?**

- 40%.
:::


:::{.dapr2callout}
**If I fail the exam, can I still pass the course?**

- Yes, if your quiz and report marks are high enough. For example:

```{r}
quiz_mark <- 65;  report_mark <- 65;  exam_mark <- 25

(quiz_mark * .10) + (report_mark * .30) + (exam_mark * .60)
```
:::


:::{.dapr2callout}
**If I fail the exam, can I still progress to Honours?**

- Yes, as long as your course mark is above 40% at first attempt (that is, the first time you sit the exam, unless you have exceptional circumstances).
:::


:::{.dapr2callout}
**What happens if I fail the course?**

- If a student fails the course and does not have exceptional circumstances, they are required to switch degree programme, since DAPR2 is a core course in the Psychology degree programme.
- If a student does have exceptional circumstances, then they will be offered a reassessment/null sit, which will count as their first attempt. **Check with your student adviser!**
:::


# What to expect from the exam

## Things you'll need for the exam

:::: {.columns}

::: {.column width="48%"}

**What you need to bring along:**

- your student card

- pens with blue or black ink

- pencils and eraser

- a calculator
  - calculators that are too clever aren't permitted; see list on Learn
  - your phone/watch cannot be your calculator

:::


::: {.column width="4%"}

:::

::: {.column width="48%"}

**What you will be given:**

- an attendance slip    

- the exam paper, which contains
  - the exam questions
  - the equation/formula sheet

- a multiple choice answer sheet for Section A (MCQs)

- a script book for Section B (short answer questions)

:::

::::

## What you can take away from the exam hall:

<br>

<center>
<h2><b>Only what you brought with you!</b></h2>
</center>

<br><br>
When you finish the exam, you will leave all papers in the exam hall.

That includes

- the exam paper with the exam questions + equation/formula sheet,
- your MCQ answer sheet, and
- your script book with written short answer questions.

Before anybody is allowed to leave the exam hall, we will make sure all of these papers have been collected.



## The process on exam day

- Check when and where your exam will take place on your calendar: [https://exams.is.ed.ac.uk/](https://exams.is.ed.ac.uk/){target="_blank"}

- Arrive early (approx. 30 minutes before the exam starts)

- Follow directions to your assigned exam room

- Follow instructions for where to leave jackets, bags, phones, etc.

- Take your stationery, calculator, and **student ID card** to your desk

- Pay attention to the invigilators for instructions about the exam


## Guides to filling in the forms

::: {.grid layout-ncol=3}

```{=html}
<div style="aspect-ratio:4/3;">
<iframe height="315" id="kaltura_player" type="text/javascript"  src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_uwsdz3vt&config[provider]={"widgetId":"1_ac8nxmb9"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to complete an examination attendance form"></iframe>
</div>
<!-- ^ link: https://media.ed.ac.uk/media/1_uwsdz3vt -->
```

```{=html}
<div style="aspect-ratio:4/3;">
<iframe height="315" id="kaltura_player" type="text/javascript"  src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_spaem93h&config[provider]={"widgetId":"1_bi6zhby5"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to prepare barcoded exam scripts and MCQ sheets"></iframe>
</div>
<!-- ^ link: https://media.ed.ac.uk/media/1_spaem93h -->
```

```{=html}
<div style="aspect-ratio:4/3;">
<iframe id="kaltura_player" type="text/javascript"  src='https://cdnapisec.kaltura.com/p/2010292/embedPlaykitJs/uiconf_id/55171522?iframeembed=true&entry_id=1_ms01z6h1&config[provider]={"widgetId":"1_n8vm6mb3"}'  style="width: 100%;height: 100%;border: 0;" allowfullscreen webkitallowfullscreen mozAllowFullScreen allow="autoplay *; fullscreen *; encrypted-media *" sandbox="allow-downloads allow-forms allow-same-origin allow-scripts allow-top-navigation allow-pointer-lock allow-popups allow-modals allow-orientation-lock allow-popups-to-escape-sandbox allow-presentation allow-top-navigation-by-user-activation" title="How to fill in the cover of an exam script book?"></iframe>
</div>
<!-- ^ link: https://media.ed.ac.uk/media/1_ms01z6h1 -->
```

:::


# The exam itself

## Exam structure {.smaller}

:::dapr2callout
**Section A (30 marks)**

- 15 multiple choice questions (MCQs)
- 2 marks per MCQ
- MCQs cover topics from across the course
- Similar style to weekly quizzes: each MCQ gives you four possible responses to choose from
:::


:::dapr2callout
**Section B (70 marks)**  

- 8 short-answer questions, which each include several related sub-questions

- Covering a range of topics, where some questions will involve:
  - short calculations
  - explaining statistical concepts  
  - interpreting model output provided in the question

- Marks range from TODO to TODO marks per question:  
  - TODO

- Marks range for each block's content:
  - TODO
:::



## Strategy: Suggested approach for Section A {.smaller}

:::dapr2callout
**Section A (30 marks)**

- 15 multiple choice questions (MCQs)
- 2 marks per MCQ
- MCQs cover topics from across the course
- Similar style to weekly quizzes: each MCQ gives you four possible responses to choose from
:::

<br>

**Suggested approach:**    

- Aim to spend 15–30 minutes on the MCQs, about 1–2 minutes per question.
- Answer all the questions that you know.
- Leave questions you don't know until later.
- Keep track of the questions you haven't done, so you can return to them at the end.
- If you're not sure about an answer, go with your gut instinct (implicit knowledge can be powerful!)
- If you don't feel any gut instincts, then just guess, because a 25% chance of guessing right is better than no response at all!



## Strategy: Suggested approach for Section B {.smaller}



:::dapr2callout
**Section B (70 marks)**  

- 8 short-answer questions, which each include several related sub-questions

- Covering a range of topics, where some questions will involve:
  - short calculations
  - explaining statistical concepts  
  - interpreting model output provided in the question
:::


<br>

**Suggested approach:**    

- Aim to spend about 90 minutes on the short answer questions.
- Questions are arranged from the highest number of marks to the lowest (i.e., Question 1 is worth the most marks, Question 8 is worth the fewest marks).
- Skim over all eight questions first.
- Prioritise questions that are worth more marks and questions that you know how to answer.
- Read the question text carefully, so that your answers include all the relevant information.
- Each question and sub-question will tell you how many marks it's worth.
  - As a rough rule of thumb, one mark → one sentence.
  - So for a five-mark question, expect to write five/six sentences.
    


## General tips for approaching the exam

- **No questions require you to write or memorise R code.** The closest thing will be interpreting some familiar output (e.g., model summaries).

- Split your time in the exam according to how many marks each section/question is worth.
  
- Try to give an answer for every question.
  - Section A: Guess if you need to!
  - Section B: A brief answer is better than no answer.
  
- You will not need every equation on the equation sheet, but some of them might help you out.

- **Remember: We are trying to give you marks!** Show us what you can do :)


# Getting prepared for the exam

## Study strategies: Active recall and self-testing

- Don't just read over labs/re-watch lectures! Instead, **practice retrieving information from memory, so that your brain is trained to give it to you when you need it.**

- **Develop and test yourself with your own flash cards.** (I designed the learning objectives in my lectures to be a good starting point for self-testing.)
  
- **Develop conceptual understanding using the Feynman technique:** Teach or explain an idea as simply as you can. Wherever you're not sure what to say or how to explain something, brush up on those areas and try again.

- **Practice for the MCQs** using the Learn quizzes marked "REVISION": you can take these as many times as you like.

- **Practice for the short answer questions** by working through lab exercises with pen and paper, where possible, especially questions that ask you to interpret model estimates.


## Mock exam in this week's lab

- The mock exam contains questions that are very similar in style and structure to the questions in the actual exam.

- We'll share sample solutions to the mock exam on Friday.



# Time for your questions!

