# Functions and models {#chap-functions-models}

```{r echo=FALSE}
# Solutions
SHOW_SOLS = TRUE

# Chunk options
knitr::opts_chunk$set(
    message = FALSE, 
    warning = FALSE, 
    fig.align = 'center',
    out.width = '60%'
)

# ggplot settings
library(ggplot2)
theme_set(
  theme_bw(base_size = 15) + 
    theme(plot.title = element_text(hjust = 0.5))
)

# reproducibility
set.seed(1)
```


:::lo
**LEARNING OBJECTIVES**

1. Recall the main concepts from introductory statistics.
2. Understand the concept of a function.
3. Be able to discuss what models are.
4. Understand the link between models and functions.
:::


## Refresher of basic terminology

`r qbegin(1)`
Provide a short definition for each of these terms:

- *(Observational) unit*
- *Variable*
- *Categorical variable*
- *Quantitative variable*
- *Response/dependent variable*
- *Explanatory/independent variable*
- *Treatment*
- *Observational study*
- *Experiment*
`r qend()`

`r solbegin(show = SHOW_SOLS)`
| Term | Definition |
|:-----|:-----------|
|*(Observational) unit* | The individual entities on which data are collected. |
|*Variable* | Any characteristic recorded on the observational units. |
|*Categorical variable* | A categorical variable, also known as a factor, places units into one of several groups. Examples are "country of birth", "dominant hand", and "eye colour". |
|*Quantitative variable* | A variable that records a numerical quantity for each case. For such variables standard arithmetic operations make sense. For example, average height or weight makes sense. |
|*Response/dependent variable* | A response variable (also called a dependent variable) measures the outcome of interest in a study. |
|*Explanatory/independent variable* | Explanatory variables (also called independent variables or predictors) are used to explain changes in the response variable. |
|*Treatment* | A treatment is any level of a factor or combination of levels from two or more factors. |
|*Observational study* | An observational study is a study in which the researcher does not manipulate any of the variables involved in the study, but merely records/observes the values as they naturally exist. |
|*Experiment* | An experiment is a study in which the researcher imposes the values of the explanatory variable on the units before measuring the response variable. |
`r solend()`


## Functions


`r qbegin(2)`
Consider the function $y = 2 + 5 \ x$.

- Identify the independent variable
- Identify the dependent variable
- Describe in words what the function means and give an example for $x = 2, 6$.

`r qend()`

`r solbegin()`
- The independent variable is $x$

- The dependent variable is $y$

- The function says that the $y$ value is obtained as a transformation of the $x$ value. That is, the $y$ value is obtained as five times $x$, plus two. 
In particular, if $x$ equals 2, the corresponding value of $y$ will be $2 + 5 * 2 = 12$. Similarly, if $x$ equals 6, the corresponding value of $y$ will be $2 + 5 * 6 = 32$.
`r solend()`



`r qbegin(3)`
Write down in words and in symbols the function summarising the relationship between the side of a square and its perimeter.

_**Hint:** We are interested in how the perimeter varies as a function of its side. Hence, the perimeter is the dependent variable, and the side is the independent variable._
`r qend()`


`r solbegin()`
_In words:_

The perimeter of a square is four times the length of its side.


<br />

_In symbols:_

The relationship between side and perimeter of squares is given by:
$$
Perimeter = 4 \ Side
$$

If you denote by $x$ the independent variable _Side_ and by $y$ the dependent variable _Perimeter_, this can be written as
$$
y = 4 \ x
$$
`r solend()`


<br>

In today's lab you will be using functions from the tidyverse package such as `tibble`, `mutate`, and `ggplot`.


`r qbegin(4)`
Load the tidyverse package.

Create a tibble named `squares` which contains the perimeters $y$ of squares having sides of length $x = 0, 2, 5, 9$.

_**Hint:** Remember that to collect multiple numbers together, you use the function `c()`._
`r qend()`

`r solbegin()`
```{r}
library(tidyverse)

squares <- tibble(
  side = c(0, 2, 5, 9),
  perimeter = 4 * side
)

squares
```
`r solend()`

`r qbegin(5)`
Plot the `squares` data as points.
`r qend()`

`r solbegin()`
```{r}
ggplot(data = squares, aes(x = side, y = perimeter)) +
  geom_point()
```
`r solend()`


`r qbegin(6)`
Plot the functional relationship between side and perimeter of squares using `geom_line()`.

_**Hint:** Create a sequence of 100 points for x going, for example, from 0 to 3._
`r qend()`

`r solbegin()`
```{r}
squares_grid <- tibble(
  side = seq(0, 3, length.out = 100),
  perimeter = 4 * side
)

ggplot(data = squares_grid, aes(x = side, y = perimeter)) +
  geom_line(colour = 'blue') +
  labs(title = 'Perimeter = 4 Side')
```
`r solend()`

<br>

The function $y = 4 \ x$ that you plotted above is an example of a function representing a mathematical model.

We typically validate a model using experimental data. 
However, we all know how squares work and that two squares with the same side will have the same perimeter. Hence this is a model of an *exact* relationship.

`r qbegin(7)`
The Scottish National Gallery kindly provided us with measurements of side and perimeter (in metres) for a sample of 10 square paintings.

The data are provided below:
```{r eval=FALSE}
sng <- tibble(
  side = c(1.3, 0.75, 2, 0.5, 0.3, 1.1, 2.3, 0.85, 1.1, 0.2),
  perimeter = c(5.2, 3.0, 8.0, 2.0, 1.2, 4.4, 9.2, 3.4, 4.4, 0.8)
)
```

Plot the mathematical model of the relationship between side and perimeter for squares, and superimpose on top the experimental data from the Scottish National Gallery.
`r qend()`

`r solbegin()`
```{r squares-scatterplot, fig.cap='The exact relationship between side and perimeter of squares.'}
sng <- tibble(
  side = c(1.3, 0.75, 2, 0.5, 0.3, 1.1, 2.3, 0.85, 1.1, 0.2),
  perimeter = c(5.2, 3.0, 8.0, 2.0, 1.2, 4.4, 9.2, 3.4, 4.4, 0.8)
)

ggplot() +
  geom_line(data = squares_grid, aes(x = side, y = perimeter), colour = 'blue') +
  geom_point(data = sng, aes(x = side, y = perimeter), colour = 'black', alpha = 0.5, size = 3) +
  labs(x = 'Side (m)', y = 'Perimeter (m)')
```
The above plot shows perfect agreement between observed data and the model.
`r solend()`


`r qbegin(8)`
Use the mathematical model to predict the perimeter of a painting with a side of 1.5 metres.
`r qend()`

`r solbegin()`
We do not have a painting with a side of 1.5 metres within the random sample of paintings from the Scottish National Gallery.
We predict the perimeter of an unobserved squared painting having a 1.5 metre side using the mathematical model.

You can obtained this prediction either using a visual approach or an algebraic one.

_Visual approach_

```{r}
ggplot() +
  geom_line(data = squares_grid, aes(x = side, y = perimeter), colour = 'blue') +
  geom_point(data = sng, aes(x = side, y = perimeter), colour = 'black', alpha = 0.5, size = 3) +
  geom_segment(aes(x = 1.5, xend = 1.5, y = 0, yend = 4 * 1.5), linetype = 2, colour = 'red',
               arrow = arrow(length = unit(0.5, "cm"))) +
  geom_segment(aes(x = 1.5, xend = 0 , y = 4 * 1.5, yend = 4 * 1.5), linetype = 2, colour = 'red',
               arrow = arrow(length = unit(0.5, "cm"))) +
  labs(x = 'Side (m)', y = 'Perimeter (m)')
```

Sometimes, we can read predicted values from the graph of the function representing the relationship.

Consider the plot created in the previous question. First, we need to check where x = 1.5. Then we draw a vertical dashed line until it meets the blue line. The y value corresponding to x = 1.5 can be read off the y-axis.

However, in this case it is not that easy to read from the drawing... Let's try the next approach.


_Algebraic approach_

Alternatively, you can substitute the x value in the formula and calculate the corresponding y value.
$$
y = 4 \ x = 4 \ (1.5) = 6
$$

<br>

**Don't forget to always include the measurement units when reporting/writing-up results!**

:::int
The predicted perimeter for squared paintings having a 1.5m side is 6m.
:::
`r solend()`



## Models

Consider now the relationship between height (in inches) and handspan (in cm). @Utts2015 provides data for a sample of 167 students which reported their height and handspan as part of a class survey. To download and preview the data, check the [data codebook](#dc-handheight).

`r qbegin(9)`
Investigate the relationship between $x$ = height and $y$ = handspan for the students in the sample.

Comment on any main differences you notice with the relationship between side and perimeter of squares.

_**Hint:** To read an '.xlsx' file you need to use the function `read_excel()` from the `readxl` package. Make sure to load that package before calling that function. Then, use a scatterplot to visualise the relationship between two quantitative variables._
`r qend()`

`r solbegin()`
```{r handheight-scatterplot, fig.cap='The statistical relationship between height and handspan.'}
library(readxl)
handheight <- read_excel('data/handheight.xlsx')

ggplot(handheight, aes(x = height, y = handspan)) +
  geom_point(size = 3, alpha = 0.5) +
  labs(x = 'Height (in.)', y = 'Handspan (cm)')
```

Recall Figure \@ref(fig:squares-scatterplot), displaying the relationship between side and perimeters of squares.
From the plot we notice two points on top of each other, reflecting the fact that two squares having the same side will always have the same perimeter.
In fact, the data from the Scottish National Gallery include two squared paintings with a side of 1.1m, both having a measured perimeter of 4.4m.

Figure \@ref(fig:handheight-scatterplot), instead, displays the relationship between height and handspan of a sample of students. The first thing that grabs our attention is the fact that students having the same height do not necessarily have the same handspan. Rather, we clearly see a variety of handspan values for students all having a height of, for example, 70in.
`r solend()`


`r qbegin(10)`
Using the following command, add on top of the scatterplot a best-fit line describing how handspan varies as a function of height.
```
geom_smooth(method = lm, se = FALSE)
```

Comment on any differences you notice with the line summarising the linear relationship between side and perimeter.
`r qend()`

`r solbegin()`
```{r}
ggplot(handheight, aes(x = height, y = handspan)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = 'Height (in.)', y = 'Handspan (cm)')
```

The line representing the relationship between side and perimeter of squares is able to predict the actual perimeter value from the measurement of the side of a square.
This is possible because the relationship between side and perimeter is an **exact** one. 
That is, any squares having the same side will have the same perimeter, and there will be no variation in those values.

The line that best fits the relationship between height and handspan, instead, is only able to predict the **average** handspan from a given value of height.
This is because there will be a distribution of handspans at each value of height.
The line will fit the trend/pattern in the values, but there will be individual-to-individual variability that we must accept around that average pattern.
`r solend()`


