# Simple linear regression {#chap-slr}

```{r echo=FALSE}
# Solutions
SHOW_SOLS = TRUE

# Chunk options
knitr::opts_chunk$set(
    message = FALSE, 
    warning = FALSE, 
    fig.align = 'center',
    out.width = '60%'
)

# ggplot settings
library(ggplot2)
theme_set(
  theme_bw(base_size = 15) + 
    theme(plot.title = element_text(hjust = 0.5))
)

# reproducibility
set.seed(1)
```

<center>
<h3 style="color: #BF1932;">
Response: Quantitative<br />Predictor: Quantitative
</h3>
</center>

<br />


:::red
**PRE-LAB ACTIVITIES**

Before attempting the lab please read [Chapter 1 of Lewis-Beck by clicking this link](https://discovered.ed.ac.uk/permalink/f/gfso8q/44UOE_ALMA51187817940002466).
:::

:::lo
**LEARNING OBJECTIVES**

1. Being able to specify a linear model for a quantitative response variable and a quantitative explanatory variable
2. Understand the concept of fitted values and residuals
3. Understand how the parameters of a linear regression model are estimated
4. Test hypotheses and construct confidence intervals for the slope of a linear model
5. Partition variability and determine the proportion of variability explained by the model
:::


## Introduction and research question

> Suppose a group of researchers was studying income differences among local government employees in Riverview, a hypothetical medium-size Midwestern city in the US. Exploratory interviews suggest a relationship between income and education. Specifically, those employees with more formal training appear to receive better pay. In an attempt to verify whether this is so, they gathered relevant data.
> The researchers did not have the time or money to interview the entire population: all 320 employees on the city payroll. Therefore, they decided to interview a simple random sample of 32, selected from the personnel list that the city clerk kindly provided.
>
> --- @Lewis-Beck2015


In this lab we will use the [riverview.csv](#dc-riverview) data to examine whether education level is related to income. The data contain five attributes collected from a random sample of $n=32$ employees working for the city of Riverview, a hypothetical Midwestern city in the US. Please look at the [data codebook](#dc-riverview) for a preview of the data and a link to the file.



`r qbegin(1)`
Load the required libraries and import the riverview data into R. Assign them to a variable called `riverview`.
`r qend()`

`r solbegin()`
```{r, warning=FALSE, message=FALSE}
library(tidyverse)

# Read in data
riverview = read_csv(file = "data/riverview.csv")
head(riverview)
```
`r solend()`



## Choose

### Data exploration {-}

The first step in any analysis should always be to explore the data. In this step, we typically want to explore:

1. the distribution of each variable used in the analysis separately, called *marginal distribution*;
2. the relationships between the variables.

Data exploration will help you specify the model, interpret results from your analysis, and will anticipate potential problems with your analysis.

Basic exploratory data analysis consists of the following:

1. *Plotting the data.* The type of plot depends on the type of data collected.
1. *Looking for overall patterns/trends.* Look at the shape, centre and spread of the data. Are the data symmetric or skewed? If histograms of quantitative variables are plotted, are the data unimodal or bimodal?
1. *Identifying any unusual observations.* An outlier is an individual observation that falls outside the overall pattern on the graph.
1. *Explore relationships.* Do you notice any association between some variables?


#### Marginal distribution of income {-}

`r qbegin(2)`
Investigate the marginal distribution of employee incomes.

_**Hint:** Create a density plot of the variable and provide summary statistics such as the centre and standard deviation of the distribution._
`r qend()`

`r solbegin()`
We can plot the marginal distribution of employee incomes as a density curve, and add a boxplot underneath to check for the presence of outliers.

```{r fig.cap="Density plot and boxplot of employee incomes."}
ggplot(data = riverview, aes(x = income)) +
  geom_density() +
  geom_boxplot(width = 1/300) +
  labs(x = "Income (in thousands of dollars)", y = "Probability density")
```

The plot suggests that the distribution of employee incomes is unimodal and most of the incomes are between roughly \$50,000 and \$70,000. The smallest income in the sample is about \$25,000 and the largest income is over \$80,000. (We could find the exact values using the `summary()` function.) This suggests there is a fair amount of variation in the data. 

To further summarize the distribution, it is typical to compute and report numerical summary statistics such as the mean and standard deviation. One way to compute these values is to use the `summary() `function from the `tidyverse` library:

```{r}
riverview %>% 
  summarize(
    M = mean(income), 
    SD = sd(income)
    )
```
`r solend()`

We can describe this variable as follows:

:::int
The marginal distribution of income is unimodal with a mean of approximately \$53,700. There is variation in employees' salaries (SD = \$14,553). 
:::


#### Marginal distribution of education level {-}

`r qbegin(3)`
Examine the marginal distribution of employee incomes.
`r qend()`

`r solbegin()`
We can visualise the marginal distribution of the education level variable using a density curve, and add a boxplot underneath to check for the presence of outliers.


```{r fig.cap='Density plot and boxplot of employee education levels.'}
ggplot(data = riverview, aes(x = education)) +
  geom_density() +
  geom_boxplot(width = 1/100) +
  labs(x = "Education level", y = "Probability density")
```

Below are the summary statistics for the employees' level of education:
```{r}
riverview %>%
  summarize(
    M = mean(education),
    SD = sd(education)
    )
```
`r solend()`

Again, we might write,

:::int
The marginal distribution of education is unimodal with a mean of 16 years. There is variation in employees' level of education (SD = 4.4 years).
:::



#### Relationship between variables {-}

After examining the marginal distributions of the variables of interest in the analysis, we typically move on to examining relationships between the variables.

`r qbegin(4)`
To better understand any relationship between income and education level we need to explore how the distribution of income differs as a function of education.

Create a scatterplot of incomes versus education.

_**Hint**: Use the `geom_point()` function_
`r qend()`

`r solbegin()`
```{r riverview-scatterplot, fig.cap='Scatterplot displaying the relationship between employee education levels and incomes.'}
ggplot(data = riverview, aes(x = education, y = income)) +
  geom_point(alpha = 0.5) +
  xlab("Education (in years)") +
  ylab("Income (in U.S. dollars)")
```
The plot suggests a relationship in the sample between level of education and income. 
`r solend()`


When describing the relationship between two variables we want to touch on four characteristics of the relationship:

1. Functional form of the relationship
2. Direction
3. Strength
4. Unusual observations that do not seem to fit with the trend (outliers)


`r qbegin(5)`
Describe the relationship between employees' level of education and income in the sample.
`r qend()`

`r solbegin()`
1. The functional form of the relationship appears to be linear.
2. The relationship seems to be increasing (i.e. the slope of the line is positive)
3. The strength of the relationship is obtained by computing a summary statistic, the correlation between the two variables. The correlation coefficient ranges from -1 to 1, with 0 meaning no correlation.
4. This is determined by visual examination of the scatterplot.


In order to investigate for the presence of outliers, we can add marginal boxplots to the scatterplot.

```{r fig.cap='Scatterplot displaying the relationship between employee education levels and incomes.'}
library(ggExtra)

p <- ggplot(data = riverview, aes(x = education, y = income)) +
  geom_point(alpha = 0.5) +
  xlab("Education (in years)") +
  ylab("Income (in U.S. dollars)")
ggMarginal(p, type = 'boxplot', margins = 'both', size = 10)
```

To investigate the strength of the relationships, we need to compute the correlation coefficient:
```{r}
riverview %>%
  select(education, income) %>%
  cor()
```
that is, 
$$
r_{\text{education, income}} = 0.79
$$

We might write,

:::int
There is a moderate, positive, linear relationship between education level and income for the employees in the sample.
Employee incomes tend to increase, on average, with the number of years of formal education.
There appears to be some variation around the linear trend, so we will investigate the strength of the relationship in more detail later on.
It appears that there are no extreme observations, i.e. points that do not fit with the rest of the data.
:::
`r solend()`




### Model specification {-}

The scatterplot in Figure \@ref(fig:riverview-scatterplot) shows a moderate, positive, and linear relationship between education level and income.

A possible model for the relationship between $x$ = education level and $y$ = income in the population is given by:
$$
y = \beta_0 + \beta_1 x + \epsilon \quad \text{where} \quad \epsilon \sim N(0, \sigma)
$$
where 

- $\mu_{y|x} = \beta_0 + \beta_1 x$ represents the systematic part of the model giving the mean of $y$ at each value of $x$;
- $\epsilon$ represents the error (deviation) from that mean, and the errors are independent from one another.

This model indicates that employee incomes should be scattered around a straight line with deviations from the line determined by the random error component, $\epsilon$.




## Fit


### Parameter estimates {-}

The parameters of the model are:

- the intercept $\beta_0$, 
- the slope $\beta_1$, 
- the standard deviation of the errors $\sigma$.

The standard deviation of the errors, denoted by $\sigma$ is an important quantity to estimate because it measures how much individual data points tend to deviate above and below the regression line. 
A small $\sigma$ indicates that the points hug the line closely and we should expect fairly accurate predictions, while a large $\sigma$ suggests that, even if we estimate the line perfectly, we can expect individual values to deviate from it by substantial amounts.

The estimates of those parameters are computed from the available sample data. These are denoted $\hat \beta_0$, $\hat \beta_1$, and $\hat \sigma$ respectively.

The *fitted model* predicts the average response for all values of the explanatory variable $x$
$$
\hat y = \hat \beta_0 + \hat \beta_1 x
$$
The residuals $\hat \epsilon = y - \hat y$ represent the discrepancy between the observations and the model-predicted responses.




`r qbegin(6)`
Fit the linear model in R using the function `lm()` and assign it to a variable called `mdl`.

The syntax is: `lm( <response variable> ~ 1 + <explanatory variable>, data = <dataframe> )`

_**Hint:** The fitted model, $\hat y = \hat \beta_0 + \hat \beta_1 x$, can be equivalently written as $\hat y = \hat \beta_0 \cdot 1+ \hat \beta_1 x$. When we specify the linear model in R, we include after the tilde `~` the variables that appear to the right of the $\hat \beta$s._
`r qend()`

`r solbegin()`
We fit the linear regression model with `income` as response variable and `education` as explanatory variable. These variables are included in the dataframe `riverview`. We assign the fitted model to an object called `mdl`.
```{r}
mdl <- lm(income ~ 1 + education, data = riverview)
mdl
```
`r solend()`

`r qbegin(7)`
What are the estimates of the model parameters?

_**Hint**: There are several equivalent options to extract them from the fitted model `mdl`:_

- `mdl`
- `mdl$coefficients`
- `coef(mdl)`
- `coefficients(mdl)`
- `summary(mdl)`

_For $\hat \sigma$, you can either use:_

- `sigma(mdl)`
- `summary(mdl)`
`r qend()`

`r solbegin()`
To obtain the estimated coefficients you can either:

- type `mdl`, i.e. simply invoking the name of the object;
- type `mdl$coefficients`;
- use the `coef(mdl)` function;
- use the `coefficients(mdl)` function;
- use the `summary(mdl)` function.


```{r}
mdl
mdl$coefficients
coef(mdl)
coefficients(mdl)
summary(mdl)
```

The estimated parameters returned by the above methods are all equivalent. However, `summary()` returns more information and you need to look under the column "Estimate".

The estimated intercept is $\hat \beta_0 = 11.32$, while the estimated slope is $\hat \beta_1 = 2.65$.

The estimated standard deviation of the errors can be obtained equivalently by:

- typing `sigma(mdl)`
- looking at the "Residual standard error" entry of the `summary(mdl)` output

```{r}
sigma(mdl)
summary(mdl)
```
The estimated standard deviation $\hat \sigma = 8.98$.
`r solend()`


`r qbegin(8)`
Plot the fitted regression line by:

1. extracting the intercept and slope of the fitted line with the function `coef(mdl)`;
2. providing the intercept and slope to the function `geom_abline()`.
`r qend()`


`r solbegin()`
We can plot the model as follows:
```{r}
intercept <- coef(mdl)[1]
slope <- coef(mdl)[2]

ggplot(data = riverview, aes(x = education, y = income)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = intercept, slope = slope, color = 'blue', size = 1) + 
  xlab("Education (in years)") +
  ylab("Income (in U.S. dollars)")
```
`r solend()`


The model-predicted values are obtained using the function `predict()`, which takes as first input the fitted regression model.

There are several call options:

- `predict(mdl)`: returns the fitted values for the units in the sample;
- `predict(mdl, newdata = <NEW DATA>)`: returns the predictions for the new explanatory variable you provide in `<NEW DATA>`.

```{r}
# Fitted values
predict(mdl)

# Predictions for those not in the sample
education_query <- tibble(education = c(0, 20.5))
predict(mdl, newdata = education_query)
```


`r qbegin(9)`
Use `predict(mdl)` to compute the fitted values and residuals, and add them as columns to the `riverview` dataframe.

Assign to the following symbols the corresponding numerical values:

- $y_{3}$ = reponse variable for unit $i = 3$ in the sample data
- $\hat y_{3}$ = fitted value for the third unit
- $\hat \epsilon_{5} = y_{5} - \hat y_{5}$ = the residual corresponding to the 5th unit.
`r qend()`

`r solbegin()`
```{r}
riverview_fitted <- riverview %>%
  mutate(
    income_hat = predict(mdl),
    resid = income - income_hat
  )

head(riverview_fitted)
```

- $y_{3}$ = 47.03
- $\hat y_{3}$ = 37.83
- $\hat \epsilon_{5} = y_{5} - \hat y_{5}$ = -12.4

`r solend()`


`r qbegin(10)`
Plot the original income vs education data points as black dots, along with the regression line in blue, the residuals as vertical red lines, and fitted values as blue dots.
`r qend()`

`r solbegin()`
```{r}
ggplot(riverview_fitted) +
  geom_point(aes(x = education, y = income), color = 'black', size = 2, alpha = 0.5) +
  geom_abline(intercept = intercept, slope = slope, color = 'blue') +
  geom_segment(aes(x = education, y = income_hat, xend = education, yend = income), 
               color = 'tomato1', alpha = 0.5) +
  geom_point(aes(x = education, y = income_hat), color = 'blue', size = 3)
```
`r solend()`


### Tests and confidence intervals for slope and intercept {-}

Consider again the output of the `summary()` function, which takes as input the fitted regression object:
```{r}
summary(mdl)
```

We already saw that the "Estimate" column returns the estimated regression coefficients.

To quantify the amount of uncertainty in each estimated coefficient that is due to sampling variability, we use the standard error of the coefficient. 
_Recall that a standard error gives a numerical answer to the question of how variable a statistic will be because of random sampling._
These are found in the column "Std. Error". 

In this example the slope, 2.651, has a standard error of 0.37. One way to envision this is as a distribution. Our best guess (mean) for the slope parameter is 2.651. The standard deviation of this distribution is 0.37, which indicates the precision (uncertainty) of our estimate.

```{r echo=FALSE, fig.cap='Sampling distribution of the slope coefficient. The distribution is approximately bell-shaped with a mean of 2.651 and a standard error of 0.36.'}
ggplot(tibble(x = c(-3 * 0.37 + 2.651, 3 * 0.37 + 2.651)), aes(x = x)) +
    stat_function(fun = dnorm, args = list(mean = 2.651, sd = 0.37)) +
  labs(x = "Estimate for employee incomes", y = '')
```

It shouldn't surprise you that the reference distribution in this case is a t-distribution with $n-2$ degrees of freedom, where $n$ is the sample size.
Recall the main formulas for obtaining a confidence interval and a test-statistic:

:::frame
**Confidence interval**: A confidence interval for the population slope is
$$
\hat \beta_1 \pm t^* \cdot SE
$$
where $t^*$ denotes the critical value chosen from t-distribution with $n-2$ degrees of freedom for a desired $\alpha$ level of confidence. 

**Test statistic**: A test statistic for $H_0: \beta_1 = 0$ is
$$
t = \frac{\hat \beta_1 - 0}{SE}
$$
which follows a t-distribution with $n-2$ degrees of freedom.
:::

In the riverview example, for 95\% confidence we have $t^* = 2.04$. The confidence interval is:
```{r}
n <- nrow(riverview)
tstar <- qt(0.975, df = n - 2)
tstar

beta1_ci <- tibble(
  lower = 2.6513 - tstar * 0.3696,
  upper = 2.6513 + tstar * 0.3696,
)
beta1_ci
```

In R it is easy to obtain the confidence intervals for the regression coefficients using the command `confint()`:
```{r}
confint(mdl, level = 0.95)
```

We typically report our uncertainty in a statistic by providing $\text{estimate} \pm t^* \cdot \text{SE}$. Here we would say that because of sampling variation, we are 95\% confident that the slope is between 1.896 and 3.406. Interpreting this, we might say,

:::int
For all Riverview city employees, each one-year difference in formal education is associated with a difference in income between \$1,896 and \$3,406, on average.
:::
    
Similarly, we could express the uncertainty in the intercept $\hat \beta_0$ as:

:::int
The average income for all Riverview city employees with zero years of education is between \$-1184 and \$23,827.
:::


`r qbegin(11)`
Test the hypothesis that the population slope is zero.
That is, that there is no linear association between income and education level in the population.
`r qend()`

`r solbegin()`
We calculate the test statistic
$$
t = \frac{\hat \beta_1 - 0}{SE} = \frac{ 2.6513 - 0 }{0.3696} = 7.173
$$
and compare it with the 5% critical value from a t-distribution with $n-2$ degrees of freedom, which is:
```{r}
n <- nrow(riverview)
tstar <- qt(0.975, df = n - 2)
tstar
```

As $|t| >> t^*$, we reject then null hypothesis as have strong evidence against it.

The p-value, shown below, also confirms the conclusion.
```{r}
2 * (1 - pt(7.173, n - 2))
```

Please note that the same information was already contained in the row corresponding to the variable "education" in the output of `summary(mdl)`
```{r}
summary(mdl)
```

:::int
We performed a t-test against the null hypothesis that education is not a significant predictor of income: $t(30) = 7.173,\ p < .001$, two-sided.
The large t-statistic leads to a very small p-value, meaning that we have strong evidence against the null hypothesis.
:::


`r solend()`


### ANOVA for regression {-}

The term ANOVA is an acronym for **AN**alysis **O**f **VA**riance. In a nutshell it means that we are examining/partitioning the total variability of a response variable.

To assess how well the model is explaining the repose, we can split the total variability in the response $y$ as the sum of the variability explained by the model (the fitted regression line) plus the variability that is left unexplained in the residuals/errors.

$$
\text{total variability in response = variability explained by model + unexplained variability in residuals}
$$

Each of these part is measured by a sum of squares:

$$
\begin{aligned}
SSTotal &= SSModel + SSE \\
\sum_i (y_i - \bar y)^2 &= \sum_i (\hat y_i - \bar y)^2 + \sum_i (y_i - \hat y_i)^2 
\end{aligned}
$$

`r qbegin(12)`
What is the proportion of the total variability in incomes explained by the linear relationship with education level?

_**Hint:** What is the value of $R^2 = SSModel / SSTotal$ for this model?_
`r qend()`

`r solbegin()`
The proportion of the total variability in incomes explained by the linear relationship with education level is given by R-squared.

**Option 1**

```{r}
summary(mdl)
```

The output of `summary()` returns the R-squared value in the following line:
```
Multiple R-squared:  0.6317
```

**Option 2**

This value can also be obtained as:
$$
R^2 = \frac{SSModel}{SSTotal} = 1 - \frac{SSE}{SSTotal}
$$

We can compute it using R as follows:
```{r}
head(riverview_fitted)

riverview_fitted %>%
  summarise(
    SSModel = sum( (income_hat - mean(income))^2 ),
    SSTotal = sum( (income- mean(income))^2 )
  ) %>%
  summarise(
    RSq = SSModel / SSTotal
  )
```


**Option 3**

The `anova` function returns the sum of squares of interest:
```{r}
mdl_anova <- anova(mdl)
mdl_anova
```

Consider the column `Sum Sq`. The entry corresponding to `education` gives SSModel = 4147.3, as education is the explanatory variable, and the entry corresponding to `Residuals` is SSE = 2418.2.

```{r}
SSModel <- mdl_anova$`Sum Sq`[1]
SSE <- mdl_anova$`Sum Sq`[2]
SSTotal <- SSModel + SSE

RSq <- SSModel / SSTotal
RSq
```

:::int
Approximately 63\% of the total variability in employee incomes is explained by the linear association with education level.
:::

`r solend()`



How do we tell if the model explains a significant amount of variability or if the explained variability is due to chance alone? 
To assess if the sample results could have happened by random chance alone we resort to a formal F-test. In simple linear regression the hypotheses being testes are:

$H_0: \text{the model is ineffective, } \beta_1 = 0$

$H_1: \text{the model is effective, } \beta_1 \neq 0$

The relevant test-statistic is a standardised version of $R^2$, which measures the amount of variation in the response explained by the model. 
We divide $R^2$ by a measure of the amount of unexplained variation, $1 - R^2$, and then adjust by the number of predictors used in the model.

$$
F = \frac{R^2 / k}{(1 - R^2) / (n - k - 1) } = \frac{R^2 / 1}{(1 - R^2) / (n - 2)}
$$
where $k$ is the number of predictors.

:::frame
**NOTE**

In simple linear regression, which is our case, the number of predictors is $k = 1$ as we only have one explanatory variable (education level).
:::

The sample F-statistic is compared to an F-distribution with $df_1 = k = 1$ and $df_2 = n - k - 1 = n - 2$ degrees of freedom.

`r optbegin()`
With some algebra we can also show that :
$$
F = \frac{R^2 / k}{(1 - R^2) / (n - k - 1) } = \frac{SSModel / k}{SSError / (n - k - 1)}
$$
Using the fact that $R^2 = SSModel / SSTotal = 1 - SSE / SSTotal$, we have

$$
\begin{aligned}
F &= \frac{R^2 / k}{(1 - R^2) / (n - k - 1) } \\
&= \frac{R^2}{1 - R^2} \cdot \frac{n - k -1}{k} \\
&= \frac{\frac{SSModel}{SSTotal}}{1 - \frac{SSModel}{SSTotal}} \cdot \frac{n - k -1}{k} \\
&= \frac{\frac{SSModel}{SSTotal}}{\frac{SSTotal - SSModel}{SSTotal}} \cdot \frac{n - k -1}{k} \\
&= \frac{\frac{SSModel}{SSTotal}}{\frac{SSE}{SSTotal}} \cdot \frac{n - k -1}{k} \\
&= \frac{SSModel}{SSE} \cdot \frac{n - k -1}{k} \\
&= \frac{SSModel / k}{SSE / (n - k - 1)} \\
&= \frac{MSModel}{MSE}
\end{aligned}
$$
`r optend()`

`r qbegin(13)`
Look at the output of `summary(mdl)` and `anova(mdl)`.

For each output, identify the relevant information to conduct an F-test against the null hypothesis that the model is ineffective at predicting income using education level.
`r qend()`

`r solbegin()`
```{r}
summary(mdl)
```

The relevant row is the following:
```
F-statistic: 51.45 on 1 and 30 DF,  p-value: 5.562e-08
```

<br />

Instead, the `anova()` output is:
```{r}
anova(mdl)
```

And the relevant entries are:

- the `Df` column, which contains the degrees of freedom;
- `F value = 51.452`, which is the F-statistic;
- `Pr(>F) = 5.562e-08`, which is the p-value.

:::int
We performed an F-test for the overall significance of the regression, $F(1, 30) = 51.45, p < .001$.
The large F-statistic leads to a very small p-value (<.001), meaning that we have very strong evidence against the null hypothesis that the model is ineffective.
The data provide strong evidence that education is an effective predictor of income.
:::
`r solend()`

