---
title: "Interactions"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('assets/setup.R')

set.seed(953)
```

:::green
Be sure to check the [**solutions to last week's exercises**](04_mlr.html).<br>You can still ask any questions about previous weeks' materials if things aren't clear!
:::



:::lo
**LEARNING OBJECTIVES**

1. Understand the concept of an interaction.
2. Interpret the meaning of a numeric $\times$ categorical interaction.  
3. Understand the principle of marginality and why this impacts modelling choices with interactions.
4. Visualize and probe interactions.

:::

`r qbegin(1)`  
Reseachers have become interested in how the number of social interactions might influence mental health and wellbeing differently for those living in rural communities compared to those in cities and suburbs. They want to assess whether the effect of social interactions on wellbeing _is moderated by_ (depends upon) whether or not a person lives in a rural area.  

Create a new RMarkdown, load the **tidyverse** package read in the wellbeing data into R.  
The data is available at https://uoepsy.github.io/data/wellbeing.csv".  

Count the number of respondents in each location (City/Location/Rural).   

**Open-ended:** Do you think there is enough data to answer this question?
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(tidyverse)
mwdata <- read_csv("https://uoepsy.github.io/data/wellbeing.csv")

mwdata %>% count(location)
```

+ We have only 7 respondents who are from a rural location, and 25 from the city & suburbs. Intuitively, this doesn't seem very many to rely on as representative of the population of those living in rural areas in Edinburgh & Lothians. Another thing to think about is that we probably don't expect _large_ differences between rural and city dwellers in the effect of social interaction on wellbeing (i.e., we might not expect differences in these sub-groups to be stronger than the overall relationship between social interaction and wellbeing).  
`r solend()`

> **Research Question:** 
> Does the relationship between number of social interactions and mental wellbeing differ between rural and non-rural residents?  

To investigate how the relationship between the number of social interactions and mental wellbeing might be different for those living in rural communities, the researchers conduct a new study, collecting data from 200 randomly selected residents of the Edinburgh & Lothian postcodes. 


`r optbegin("Wellbeing/Rurality data codebook.", olabel=FALSE, toggle=params$TOGGLE)`  

__Download link__

The data is available at https://uoepsy.github.io/data/wellbeing_rural.csv.  

__Description__

From the Edinburgh & Lothians, 100 city/suburb residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), was used to measure mental health and well-being. 
Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), whether a daily routine is followed (yes/no). For those respondents who had an activity tracker app or smart watch, they were asked to provide their average weekly number of steps.  
  
The data in `wellbeing_rural.csv` contain seven attributes collected from a random sample of $n=200$ hypothetical residents over Edinburgh & Lothians, and include:  

- `wellbeing`: Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and well-being. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70.  
- `outdoor_time`: Self report estimated number of hours per week spent outdoors  
- `social_int`: Self report estimated number of social interactions per week (both online and in-person)
- `routine`: Binary 1=Yes/0=No response to the question "Do you follow a daily routine throughout the week?"
- `location`: Location of primary residence (City, Suburb, Rural)
- `steps_k`: Average weekly number of steps in thousands (as given by activity tracker if available)
- `age`: Age in years of respondent

__Preview__

The first six rows of the data are:

```{r echo=FALSE}
read_csv('https://uoepsy.github.io/data/wellbeing_rural.csv') %>% head %>% gt::gt()
```
  
`r optend()`


  

`r qbegin(2)`
Specify a multiple regression model to answer the research question.  
Read in the data, and assign it the name "mwdata2". Then fully explore the variables and relationships which are going to be used in your analysis.   


> "Except in special circumstances, a model including a product term for interaction between two explanatory variables should also include terms with each of the explanatory variables individually, even though their coefficients may not be significantly different from zero. Following this rule avoids the logical inconsistency of saying that the effect of $X_1$ depends on the level of $X_2$ but that there is no effect of $X_1$."  
> @Ramsey2012

---
 
+ **Tip 1:** Install the `psych` package (remember to use the console, not your script to install packages), and then load it (load it in your script). The `pairs.panels()` function will plot all variables in a dataset against one another. This will save you the time you would have spent creating individual plots.  
+ **Tip 2:** Check the "location" variable. It currently has three levels (Rural/Suburb/City), but we only want two (Rural/Not Rural). You'll need to fix this. One way to do this would be to use `ifelse()` to define a variable which takes one value ("Rural") if the observation meets from some condition, or another value ("Not Rural") if it does not. Type `?ifelse` in the **console** if you want to see the help function. You can use it to add a new variable either inside `mutate()`, or using `data$new_variable_name <- ifelse(test, x, y)` syntax.  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
To address the research question, we are going to fit the following model, where $y$ = wellbeing; $x_1$ = weekly outdoor time; and $x_2$ = whether or not the respondent lives in a rural location or not. 
$$
y = \beta_0 + \beta_1 \cdot x_1 + \beta_2 \cdot x_2 + \beta_3 \cdot x_1 \cdot x_2 + \epsilon \\ 
\quad \\ \text{where} \quad \epsilon \sim N(0, \sigma) \text{ independently}
$$

First we read in the data, and take a quick look at our variables:
```{r}
mwdata2 <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
summary(mwdata2)
```

First let's create a new variable for Rural/Not Rural
```{r}
mwdata2 <- 
  mwdata2 %>% 
  mutate(
    isRural = ifelse(location == "rural", "rural","not rural")
  )
```

Now let's use `psych::pairs.panels()` function.  
We could use it on the whole dataset, but for now we'll just do it on the variables we're interested in:
```{r}
library(psych)
pairs.panels(mwdata2 %>% select(wellbeing, social_int, isRural))
```
`r solend()`

`r qbegin(3)`
Produce a visualisation of the relationship between weekly number of social interactions and well-being, with separate _facets_ for rural vs non-rural respondents. 
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
ggplot(data = mwdata2, aes(x = social_int, y = wellbeing)) + 
  geom_point() + 
  facet_wrap(~isRural)
```
`r solend()`

`r qbegin(4)`
Fit your model using `lm()`, and assign it as an object with the name "rural_mod".   

**Hint:** When fitting a regression model in R with two explanatory variables A and B, and their interaction, these two are equivalent:  

+ y ~ A + B + A:B
+ y ~ A*B  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
rural_mod <- lm(wellbeing ~ 1 + social_int * isRural, data = mwdata2)
```
`r solend()`

:::statbox
__Interpreting coefficients for A and B in the presence of an interaction A:B__   

When you include an interaction between $x_1$ and $x_2$ in a regression model, you are estimating the extent to which the effect of $x_1$ on $y$ is different across the values of $x_2$.  

What this means is that the effect of $x_1$ on $y$ *depends on/is conditional upon* the value of $x_2$.  
(and vice versa, the effect of $x_2$ on $y$ is different across the values of $x_1$).   
This means that we can no longer talk about the "effect of $x_1$ _holding $x_2$ constant_". Instead we can talk about a _marginal effect_ of $x_1$ on $y$ at a specific value of $x_2$. 

:::frame
When we fit the model $y = \beta_0 + \beta_1\cdot x_1 + \beta_2 \cdot x_2 + \beta_3 \cdot x_1 \cdot x_2 + \epsilon$ using `lm()`:  

- the parameter estimate $\hat \beta_1$ is the _marginal effect_ of $x_1$ on $y$ where $x_2 = 0$  
- the parameter estimate $\hat \beta_2$ is the _marginal effect_ of $x_2$ on $y$ where $x_1 = 0$  
:::

<div style="margin-left: 15px">
<small>
__side note:__ Regardless of whether or not there is an interaction term in our model, all parameter estimates in multiple regression are "conditional" in the sense that they are dependent upon the inclusion of other variables in the model. For instance, in $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$ the coefficient $\hat \beta_1$ is conditional upon holding $x_2$ constant. 
</small>
</div>

:::

:::statbox
__Interpreting the interaction term A:B__  

The coefficient for an interaction term can be thought of as providing an _adjustment to the slope._   
  
In our model: $\text{wellbeing} = \beta_0 + \beta_1\cdot\text{social-interactions} + \beta_2\cdot\text{isRural} + \beta_3\cdot\text{social-interactions}\cdot\text{isRural} + \epsilon$, we have a numeric*categorical interaction.  
The estimate $\hat \beta_3$ is the adjustment to the slope $\hat \beta_1$ to be made for the individuals in the $\text{isRural}=1$ group. 

:::


`r qbegin(5)`  
Look at the parameter estimates from your model, and write a description of what each one corresponds to on the plot shown in Figure \@ref(fig:plot-annotate-int) (it may help to sketch out the plot yourself and annotate it).  

> "The best method of communicating findings about the presence of significant interaction may be to present a table of graph of the estimated means at various combinations of the interacting variables."    
> @Ramsey2012  


```{r plot-annotate-int, echo=FALSE, fig.cap="Multiple regression model: Wellbeing ~ Social Interactions * is Rural<br><small>Note that the dashed lines represent predicted values below the minimum observed number of social interactions, to ensure that zero on the x-axis is visible</small>"}
nd = expand_grid(social_int=0:13,isRural=c("rural","not rural"))
nd = nd %>% mutate(wellbeing = predict(rural_mod, newdata = .))
sjPlot::plot_model(rural_mod, type="int")+
  scale_fill_manual(NULL, values=c(NA,NA))+xlim(0,28)+
  geom_line(inherit.aes=FALSE,data=nd,aes(x=social_int,col=isRural,y=wellbeing), lty="longdash")+
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 1))
```

`r optbegin("Hints.", olabel=FALSE, toggle=params$TOGGLE)`
Here are some options to choose from:

+ The point at which the blue line cuts the y-axis (where social_int = 0)
+ The point at which the red line cuts the y-axis (where social_int = 0)
+ The average vertical distance between the red and blue lines. 
+ The vertical distance from the blue to the red line _at the y-axis_ (where social_int = 0)
+ The vertical distance from the red to the blue line _at the y-axis_ (where social_int = 0)
+ The vertical distance from the blue to the red line _at the center of the plot_
+ The vertical distance from the red to the blue line _at the center of the plot_
+ The slope (vertical increase on the y-axis associated with a 1 unit increase on the x-axis) of the blue line
+ The slope of the red line
+ How the slope of the line changes when you move from the blue to the red line
+ How the slope of the line changes when you move from the red to the blue line
`r optend()`

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
We can obtain our parameter estimates using various functions such as `summary(rural_mod)`,`coef(rural_mod)`, `coefficients(rural_mod)` etc. 

```{r}
coefficients(rural_mod)
```


+ $\hat \beta_0$ = `(Intercept)` = `r round(coef(rural_mod)[1],2)`: The point at which the blue line cuts the y-axis (where social_int = 0).  
+ $\hat \beta_1$ = `social_int` = `r round(coef(rural_mod)[2],2)`: The slope (vertical increase on the y-axis associated with a 1 unit increase on the x-axis) of the blue line.
+ $\hat \beta_2$ = `isRuralrural` = `r round(coef(rural_mod)[3],2)`: The vertical distance from the blue to the red line _at the y-axis_ (where social_int = 0).  
+ $\hat \beta_3$ = `social_int:isRuralrural` = `r round(coef(rural_mod)[4],2)`: How the slope of the line changes when you move from the blue to the red line. 

`r solend()`

`r qbegin(6)`
Load the __sjPlot__ package and try using the function `plot_model()`.  
The default behaviour of `plot_model()` is to plot the parameter estimates and their confidence intervals. This is where `type = "est"`. 
Try to create a plot like Figure \@ref(fig:plot-annotate-int), which shows the two lines (**Hint:** what are this weeks' exercises all about? `type = ???`.)
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(sjPlot)
plot_model(rural_mod, type="int")
```
`r solend()`

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>