---
title: "Writing-up"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---


```{r setup, include=FALSE}
source('assets/setup.R')

# knitr::opts_chunk$set(cache = TRUE)
set.seed(1)
```


:::lo
**LEARNING OBJECTIVES**

1. Understand the difference between outliers and influential points. 
1. Understand and apply methods to detect outliers and influential points.
1. Comment on the issues of correlation vs causation and generalisability of the sample results.
1. Run a full simple regression analysis from start to finish and to correctly interpret the results.
:::


# Data and research question

A school is interested in investigating how children' performance is affected by anxiety.

Consider the following data on anxiety and academic performance for a sample of 100 children.


`r optbegin('Data: anx_perf.csv', FALSE, TRUE, toggle = params$TOGGLE)`
**Download link**

[Download the data here](https://uoepsy.github.io/data/anx_perf.csv)

**Description**

Level of anxiety and academic performance were measured for one hundred children randomly selected from Green Tree primary school, a hypothetical school.

The measured variables were:

- `anxiety`: pupils' anxiety score
- `performance`: pupils' academic performance score

**Preview**

The first six rows of the data are:

```{r echo=FALSE}
library(tidyverse)
library(kableExtra)

df <- read_csv(file = 'https://uoepsy.github.io/data/anx_perf.csv')
kable(head(df), align = 'c', digits = 2) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1:2, width = '10em')
```
`r optend()`


# Analysis

In today's lab the analysis will be provided, the task will be for you to correctly report what has been done.

Read the data:
```{r}
library(tidyverse)
library(car)

anxperf <- read_csv(file = 'https://uoepsy.github.io/data/anx_perf.csv')
head(anxperf)
```

Investigate the marginal distributions of each variable and the relationship between the two variables.

**Note:** The patchwork library is used to combine ggplots into a single figure made up of multiple panels. You can add them using `+` or `|`, and put plots underneath with the `/` operator.
```{r fig.height=4, fig.width=10, out.width='100%'}
library(patchwork)

plt_anx <- ggplot(anxperf, aes(x = anxiety)) +
  geom_density() +
  geom_boxplot(width = 0.01)

plt_perf <- ggplot(anxperf, aes(x = performance)) +
  geom_density() +
  geom_boxplot(width = 0.01)

plt_joint <- ggplot(anxperf, aes(x = anxiety, y = performance)) +
  geom_point() +
  labs(x = 'Anxiety score', y = 'Performance score')

plt_anx | plt_perf | plt_joint
```

```{r}
anxperf %>%
  summarise(M_anx = mean(anxiety), 
            SD_anx = sd(anxiety),
            MED_anx = median(anxiety), 
            IQR_anx = IQR(anxiety),
            M_perf = mean(performance), 
            SD_perf = sd(performance),
            MED_perf = median(performance), 
            IQR_perf = IQR(performance))
```


Correlation between the variables:
```{r}
cor(anxperf)
```

Linear model:
```{r}
mdl1 <- lm(performance ~ 1 + anxiety, data = anxperf)
summary(mdl1)
```

:::green
**TIP**

When you call `plot()` on a fitted model to show the diagnostic plots, you can display all four plots at once by telling R
```{r eval=FALSE}
par(mfrow = c(2,2))
```

This means that each figure should be made of 2 by 2 plots.

To go back to one figure made by a single plot, just type"
```{r eval=FALSE}
par(mfrow = c(1,1))
```

**Note that this will not work with ggplot.**
:::


Diagnostics:
```{r out.width='90%', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
plot(mdl1)
```

```{r}
im <- influence.measures(mdl1)
summary(im)
```

Focus on Cook's distance for now:
```{r}
anxperf <- anxperf %>%
    slice(-41)
```

Re-fit the model and check assumptions:
```{r}
mdl2 <- lm(performance ~ 1 + anxiety, data = anxperf)
summary(mdl2)
```

```{r out.width='90%', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
plot(mdl2)
```

```{r}
ncvTest(mdl2)
```


Try transforming the response:
```{r}
anxperf <- anxperf %>%
    mutate(log_performance = log(performance))

mdl3 <- lm(log_performance ~ 1 + anxiety, data = anxperf)
summary(mdl3)
```

```{r out.width='90%', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
plot(mdl3)

shapiro.test(mdl3$residuals)
```

Try removing outlier:
```{r}
anxperf <- anxperf %>%
    slice(-62)
```

```{r out.width='90%', fig.height=7, fig.width=7}
mdl4 <- lm(log_performance ~ 1 + anxiety, data = anxperf)
summary(mdl4)

par(mfrow = c(2,2))
plot(mdl4)

shapiro.test(mdl4$residuals)

ncvTest(mdl4)
```

Go back to figures with only one plot:
```{r}
par(mfrow = c(1,1))
```


Table for reporting:
```{r}
library(sjPlot)
tab_model(mdl4)
```


Plot of the final model:
```{r}
betas <- coef(mdl4)

ggplot(anxperf, aes(x = anxiety, y = log_performance)) +
  geom_point() +
  geom_abline(intercept = betas[1], slope = betas[2], color = 'blue') +
  labs(x = 'Anxiety score', y = 'Performance score (in log-scale)')
```



# Reporting

A good data analysis report should follow three steps: *Think, Show, and Tell.*

For a detailed explanation of each step, check this [document.](https://uoepsy.github.io/files/writeup_guide.pdf){target="_blank"}


## Think

:::yellow
What do you know? What do you hope to learn? What did you learn during the exploratory analysis?
:::

`r qbegin('1: Describe design', FALSE)`
Describe the study design, the data collection strategy, etc.
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
The data consist of measurements of anxiety and academic performance scores for a random sample of 100 children from Green Tree primary school.
`r solend()`



`r qbegin('2: Describe the data', FALSE)`
- How many observational units? (If not mentioned above.)
- Are there any observations that have been excluded based on pre-defined criteria? How/why, and how many?
- Describe and visualise the variables of interest. How are they scored? have they been transformed at all?
- Describe and visualise relationships between variables. Report covariances/correlations.
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
```{r marg-distrib, echo=FALSE, fig.height=4, fig.width=10, out.width='100%', fig.cap = 'Marginal and joint distributions.'}
library(patchwork)
library(ggExtra)

plt_anx <- ggplot(anxperf, aes(x = anxiety)) +
  geom_density() +
  geom_boxplot(width = 0.01) +
  labs(title = 'a')

plt_perf <- ggplot(anxperf, aes(x = performance)) +
  geom_density() +
  geom_boxplot(width = 0.01) +
  labs(title = 'b')

plt_joint <- ggplot(anxperf, aes(x = anxiety, y = performance)) +
  geom_point() +
  labs(x = 'Anxiety score', y = 'Performance score', title = 'c')

plt_anx | plt_perf | plt_joint
```

The distribution of anxiety scores appears to be roughly symmetric around the mean anxiety score of 20.4, with standard deviation 5.49.
The boxplot in Figure \@ref(fig:marg-distrib)(a) does not highlight any evident outliers.

The distribution of performance scores appears to be right skewed, with median performance score 28.5, and interquartile range 24.
A boxplot of the distribution, Figure \@ref(fig:marg-distrib)(b), flags two potential outliers which are worth keeping an eye on, but have not been removed from the dataset.

Finally, the scatterplot in Figure \@ref(fig:marg-distrib)(c) displays the relationship between anxiety and performance score. There appears to be a strong ($r_{Anxiety, Performance} = 0.81$), positive linear relationship between the two variables.
The scatterplot also highlights a couple of points which do not fit with the rest of the data. One has a high anxiety score but an unusually low performance score, and another point having an anxiety score of approximately 15 and a performance score of approximately 50, which is 25 scores above the rest of the points.
`r solend()`


`r qbegin('3: Describe the analytical approach', FALSE)`
- What type of statistical analysis do you use to answer the research question? (e.g., t-test, simple linear regression,
multiple linear regression)
- Describe the model/analysis structure
- What is your outcome variable? What is its type?
- What are your predictors? What are their types?
- Any other specifics?
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
We want to investigate whether children' performance is affected by anxiety and what is the intensity of that relationship.

To do so, we will use a simple linear regression model involving two numeric variables: anxiety (the predictor) and performance (the response).
`r solend()`



`r qbegin('4: Planned analysis vs actual analysis', FALSE)`
- Was there anything you had to do differently than planned during the analysis? Did the modelling highlight issues in your data?
- Did you have to do anything (e.g., transform any variables, exclude any observations) in order to meet assumptions?
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
Such model, however, highlighted one influential observation (namely, case 41) having a Cook's distance of 1.11.
Careful inspection of case 41 showed a high anxiety score (33), far from the average anxiety score (20.4) --- hence, the point has high leverage. Furthermore, the point was an outlier as the actual performance score was 8, which was roughly 57 points lower than the model-predicted value.

After removing such observation from the dataframe and re-fitting the linear model, the diagnostic plots raised concerns about the normality and equal variance assumptions not being satisfied.

The Normal quantile plot showed that the distribution of the residuals was skewed to the right, and the Scale-Location plot displayed a thickening pattern (that is, a fan-shape) rather than a roughly constant vertical spread.

To mitigate the violation of the equal variance assumption, we log transformed the response variable.

After refitting the model, the diagnostic plots only raised concern about a single point (case 62) not fitting the linear pattern in the Normal quantile plot, and also being particularly unusual in the Residuals vs Fitted values plot.

The Shapiro-Wilk test for normality of the errors showed a significant p-value, indicating violation of the normality assumption.

After careful inspection, and because the point was not unusual in x-value, we decided to remove case 62 from the dataset and re-fit the linear model.
`r solend()`




## Show

:::yellow
Show the mechanics and visualisations which will support your conclusions
:::


`r qbegin('5: Present and describe the final model', FALSE)`
Present and describe the model or test which you deemed best to answer your question.
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
The final model which was chosen to answer the research question of interest was:
$$
\widehat{\log (Performance)} = 1.41 + 0.0949 \ Anxiety
$$

and was fitted excluding cases 41 and 62. The former was found to be an influential point, while the latter an outlier affecting normality.
`r solend()`



`r qbegin('6: Are the assumptions and conditions of your final test or model satisfied?', FALSE)`
Are the assumptions and conditions of your final test or model satisfied?
_For the final model (the one you report results from), were all assumptions met? (Hopefully yes, or there is more work to do…). Include evidence (tests or plots)._
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
```{r echo=FALSE, out.width='90%', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
plot(mdl4)
```

The final model was found to not violate the regression assumptions. A plot of the residuals vs fitted values shows randomly scattered points with no pattern and a horizontal direction.

The Normal quantile plot (qq-plot) shows a roughly linear trend in the standardized residuals vs the normal quantiles.
We performed a Shapiro-Wilk test, at the 5\% significance level, against the null hypothesis that the residuals came from a normal distribution ($W = 0.98$, $p = 0.14$).
The large $W$-statistic leads to a p-value (0.14) larger than the 0.05 threshold.
Hence, the sample results do not provide sufficient evidence to reject the null hypothesis that the errors follow a normal distribution.

The Residuals vs Fitted plot and the Scale-Location plot show that the vertical spread of the residuals is roughly the same everywhere. 

Hence, we see no violation of the constant variance assumption.
At the 5\% significance level, we also performed a Breusch-Pagan test against the null hypothesis of homoscedasticity ($\chi^2(1) = 0.196, p = 0.658$). The large p-value indicates that the sample results do not provide sufficient evidence to reject the null hypothesis of equal variance.

Finally, a plot of residuals vs leverage does not highlight any influential observations.
`r solend()`



`r qbegin('7: Report your test or model results', FALSE)`
- Provide a table of results if applicable (for regression tables, try `tab_model()` from the **sjPlot** package).
- Provide plots if applicable.
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
The table below displays the estimated regression coefficients along with the corresponding test of significance:
<center>
```{r echo=FALSE}
sjPlot::tab_model(mdl4)
```
</center>

<br>
The fitted model is displayed below:
```{r echo=FALSE}
betas <- coef(mdl4)

ggplot(anxperf, aes(x = anxiety, y = log_performance)) +
  geom_point() +
  geom_abline(intercept = betas[1], slope = betas[2], color = 'blue') +
  labs(x = 'Anxiety score', y = 'Performance score (in log-scale)')
```

`r solend()`


## Tell

:::yellow
Communicate your findings
:::


`r qbegin('8: Interpret your results in the context of your research question.', FALSE)`
- What do your results suggest about your research question?
- Make direct links from hypotheses to models (which bit is testing hypothesis)
- Be specific - which statistic did you use/what did the statistical test say? Comment on effect sizes.
- Make sure to include measurement units where applicable.
`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`
The fitted model predicts log performance for a given anxiety score.
Both intercept ($t(96) = 19.26, p < .001$, two-sided) and slope ($t(96) = 27.14, p < .001$, two-sided) are significantly different from zero.

As an anxiety score of zero is meaningless and far from the allowed range of scores, the estimated intercept $\hat \beta_0 = 1.41$ has no meaning if not a practical one of where the line meets the y-axis.

The estimated slope, $\hat \beta_1 = 0.0949$, indicates that, each unit increase in anxiety score is associated, on average, to a 0.0949 units increase in log performance.

That is, there is a positive/increasing linear relationship between anxiety and log performance. However, the magnitude of the rate of increase, even if significant, is small and in the order of 0.1.

The F-test for model utility is also significant ($F(1,96) = 736.7, p < .001$). At the 5\% significance level, anxiety is a significant predictor of log performance.^[
This information, however, is equivalent to the t-test for the significance of the slope as the F-statistic is the square t-statistic.
]

Approximately 88.5\% of the variability in log performance scores is explained by the linear association with anxiety scores.
`r solend()`




# Put everything together

If you followed the steps above, it is just a matter of taking all answers to the above questions and combining them, in order to have a reasonable draft of a statistical report.

