# Multiple linear regression {#chap-mlr}

```{r echo=FALSE}
# Solutions
SHOW_SOLS = FALSE

# Chunk options
knitr::opts_chunk$set(
    message = FALSE, 
    warning = FALSE, 
    fig.align = 'center',
    out.width = '60%'
)

# ggplot settings
library(ggplot2)
theme_set(
  theme_bw(base_size = 15) + 
    theme(plot.title = element_text(hjust = 0.5))
)

# reproducibility
set.seed(953)
```

<center>
<h3 style="color: #BF1932;">
Response: Quantitative<br />Predictor<b>s</b>: ... + ... + ...
</h3>
</center>

<br />

:::red
**PRE-LAB ACTIVITIES**

Before attempting the lab please read ?.
:::

:::lo
**LEARNING OBJECTIVES**

1. Extend the ideas of SLR to consider regression models with two or more predictors.
2. Understand and interpret the coefficients in multiple linear regression models
3. Test hypotheses and construct confidence intervals for the coefficients of a multiple linear regression model
4. (Categorical pred.) Understand the meaning of a MLR model with categorical predictors (no interaction case). For example, two parallel regression lines, one for each of two groups.

:::

In this lab, we move from the simple linear regression model (one outcome variable, one explanatory variable) to the _multiple regression model_ (one outcome variable, multiple explanatory variables). Everything you learned about simple linear regression can be extended (with minor modification) to the multiple regresion model. The key conceptual difference is that for simple linear regression we think of the distribution of errors at some fixed value of the explanatory variable. For multiple linear regression, we think about the distribution of errors at fixed set of values for all our explanatory variables. 

## ~ Quantitative + Quantitative

As our first example, we will revisit the [riverview.csv](#dc-riverview) data used in Chapter \@ref(chap-slr). The data contain five attributes collected from a random sample of $n=32$ employees working for the city of Riverview, a hypothetical Midwestern city in the US. The [data codebook](#dc-riverview) contains a preview of the data and a link to download the file. 

> Suppose that the group of researchers studying income differences in Riverview (our hypothetical Midwestern city in the US) wanted to know if the relationship between income and education held even when taking into account the relationship between income and seniority.

We previously used simple linear regression to examine whether the level of income was related to the amount of education a person had received. We will now extend this and build a multiple regression model which we can use to predict income based on both education *and* seniority (years in the job).

`r qbegin(1)`
Create a new RMarkdown, and in the first code-chunk, load the required libraries and import the riverview data into R. Assign them to a object called `riverview`.
`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork) #used in the next section for plots

# Read in data
riverview = read_csv(file = "data/riverview.csv")
head(riverview)
```
`r solend()`

`r qbegin(2)`
Produce plots of the _marginal distributions_ (the distributions of each variable in the analysis without reference to the other variables) of the income, education, and seniority variables. 
`r qend()`
`r solbegin(show=SHOW_SOLS)`
 Two of these we saw in Chapter \@ref(chap-slr):
```{r fig.cap="Marginal distribution plots of employee incomes and education levels", out.width="100%"}
income_plot <- 
  ggplot(data = riverview, aes(x = income)) +
  geom_density() +
  geom_boxplot(width = 1/300) +
  labs(x = "Income (thousands of US $)", y = "Probability density")

education_plot <- 
  ggplot(data = riverview, aes(x = education)) +
  geom_density() +
  geom_boxplot(width = 1/100) +
  labs(x = "Education (in years)", y = "Probability density")

# the "patchwork" library allows us to arrange multiple plots
income_plot | education_plot
```

And we can do the same for the seniority variable, as it is also quantitative:
```{r fig.cap="Marginal distribution plot of employee seniority"}
ggplot(data = riverview, aes(x = seniority)) +
  geom_density() +
  geom_boxplot(width = 1/300) +
  labs(x = "Seniority (in years)", y = "Probability density")
```

:::int
The marginal distribution of seniority is unimodal with a mean of approximately `r round(mean(riverview$seniority),1)` years. There is variation in employees' salaries (SD = `r round(sd(riverview$seniority),1)` years). 
:::

`r solend()`

`r qbegin(3)`
Produce plots of the _marginal relationships_ between the outcome variable (income) and each of the explanatory variables.  
`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r riverview-mlr-rels, fig.cap='Scatterplots displaying the relationships between employee incomes and a) education levels, and b) seniority'}
income_edu <- 
  ggplot(data = riverview, aes(x = education, y = income)) +
  geom_point(alpha = 0.5) +
  labs(x = "Education (in years)", y = "Income (in thousands of U.S. dollars)")

income_sen <- 
  ggplot(data = riverview, aes(x = seniority, y = income)) +
  geom_point(alpha = 0.5) +
  labs(x = "Seniority (in years)", y = "Income (in thousands of U.S. dollars)")

income_edu | income_sen
```
`r solend()`
  
`r qbegin(4)`
Produce a correlation matrix of the variables which are to be used in the analysis, and write a short paragraph describing the relationships. 

:::yellow
__Correlation matrix__  

A table showing the correlation coefficients - $r_{(x,y)}=\frac{\mathrm{cov}(x,y)}{s_xs_y}$ - between variables. Each cell in the table shows the relationship between two variables. The diagonals show the correlation of a variable with itself (and are therefore always equal to 1).  

**Hint:** We can create a correlation matrix easily by giving the `cor()` function a dataframe. However, we only want to give it 3 columns here. Think about how we select specific columns, either using `select()`, or giving the column numbers inside `[]`. 
:::

`r qend()`
`r solbegin(show=SHOW_SOLS)`
We can either use:
```{r eval=FALSE}
# correlation matrix of the first 3 columns
cor(riverview[,1:3])
```
or:
```{r}
# select only the columns we want by name, and pass this to cor()
riverview %>% 
  select(income, education, seniority) %>%
  cor()
```


:::int
There is a moderate, positive, linear relationship between education level and income for the employees in the sample.
Employee incomes tend to increase, on average, with the number of years of formal education.
There is a moderate, positive, linear relationship between seniority and income for the employees in the sample.
Employee incomes tend to increase, on average, with seniority. 
There is also a weak positive correlation between number of years of education and seniority. 
::: 

<br>
Note that there is a weak correlation between our two explanatory variables (education and seniorty). We will return to how this might affect our model when we look at the assumptions of multiple regression in Chapter \@ref(chap-mlr-assumptions). 
`r solend()`

`r qbegin(5)`
The scatterplots in Figure \@ref(fig:riverview-mlr-rels) shows moderate, positive, and linear relationships both between education level and income and between seniority and income.  

Specify the form of your model, $x_1$ = education level, $x_2$ = seniority and $y$ = income.  
What are the _parameters_ of the model. How do we denote _parameter estimates_?  
`r qend()`
`r solbegin(show=SHOW_SOLS)`
A model for the relationship between $x_1$ = education level, $x_2$ = seniority and $y$ = income is given by:
$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon \\ \quad \\ \text{where} \quad \epsilon \sim N(0, \sigma) \text{ independently}
$$

In the model specified above,

- $\mu_{y|x_1, x_2} = \beta_0 + \beta_1 x + \beta_2 x_2$ represents the systematic part of the model giving the mean of $y$ at each combination of values of $x_1$ and $x_2$;
- $\epsilon$ represents the error (deviation) from that mean, and the errors are independent from one another.  
  
  
The parameters of our model are:

+ $\beta_0$ (The intercept);
+ $\beta_1$ (The slope across values of $x_1$);
+ $\beta_2$ (The slope across values of $x_2$);  
+ $\sigma$ (The standard deviation of the errors).

When we estimate these parameters from the available data, we have a _fitted model_ (recall that the h$\hat{\textrm{a}}$ts are used to distinguish our _estimates_ from the _true unknown parameters_): 
$$
\widehat{Income} = \hat \beta_0 + \hat \beta_1 \cdot Education + \hat \beta_2 \cdot Seniority
$$
And we have residuals $\hat \epsilon = y - \hat y$ which are the deviations from the observed values and our model-predicted responses.  
`r solend()`


:::frame   
__Visual__

Note that for simple linear regression we talked about our model as a _line_ in 2 dimensions: the systematic part $\beta_0 + \beta_1 x$ defined a line for $\mu_y$ across the possible values of $x$, with $\epsilon$ as the random deviations from that line. But in multiple regression we have more than two variables making up our model. 

In this particular case of three variables (one outcome + two explanatory), we can think of our model as a _regression surface_ (See Figure \@ref(fig:regsurf)). The systematic part of our model defines the surface across a range of possible values of both $x_1$ *and* $x_2$. Deviations from the surface are determined by the random error component, $\hat \epsilon$.  

```{r regsurf, echo=FALSE, fig.cap = "Regression surface for income ~ education + seniority, from two different angles", out.width="100%"}
fit<-lm(income~education+seniority, data=riverview)
steps=50
education <- with(riverview, seq(min(education),max(education),length=steps))
seniority <- with(riverview, seq(min(seniority),max(seniority),length=steps))
newdat <- expand.grid(education=education, seniority=seniority)
income <- matrix(predict(fit, newdat), steps, steps)


par(mfrow=c(1,2))
p <- persp(education,seniority,income, theta = 35,phi=10, col = NA)
obs <- with(riverview, trans3d(education,seniority, income, p))
pred <- with(riverview, trans3d(education, seniority, fitted(fit), p))
points(obs, col = "red", pch = 16)
#points(pred, col = "blue", pch = 16)
segments(obs$x, obs$y, pred$x, pred$y)

p <- persp(education,seniority,income, theta = -35,phi=10, col = NA)
obs <- with(riverview, trans3d(education,seniority, income, p))
pred <- with(riverview, trans3d(education, seniority, fitted(fit), p))
points(obs, col = "red", pch = 16)
#points(pred, col = "blue", pch = 16)
segments(obs$x, obs$y, pred$x, pred$y)

par(mfrow=c(1,1))
```

Don't worry about trying to figure out how to visualise it if we had any more explanatory variables! We can only concieve of 3 spatial dimensions. One could imagine this surface changing over time, which would bring in a 4th dimension, but beyond that, it's not worth trying!.

:::

`r qbegin(6)`
Fit the linear model in R, assigning the output to an object called `mdl1`.  
  
As we did for simple linear regression, we can fit our multiple regression model using the `lm()` function. We can add as many explanatory variables as we like, separating them with a `+`.  
<br>
<center>`lm( <response variable> ~ 1 + <explanatory variable 1> + <explanatory variable 2> + ... , data = <dataframe> )`</center>
<br>

`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r}
mdl1 <- lm(income ~ 1 + education + seniority, data=riverview)
```
`r solend()`


`r qbegin(7)`
Using any of:  

- `mdl1`
- `mdl1$coefficients`
- `coef(mdl1)`
- `coefficients(mdl1)`
- `summary(mdl1)`

Write out the estimated parameter values of: 

1. $\hat \beta_0$, the estimated average income associated with zero years of education and zero years of seniority.  
2. $\hat \beta_1$, the estimated increase in average income associated with one year increase in education, _holding seniority constant_ (i.e., when the remaining explanatory variables are held at the same value or are fixed).
3. $\hat \beta_2$, the estimated increase in average income associated with one year increase in seniority, _holding education constant_.  

`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r}
coef(mdl1)
```

1. $\hat \beta_0$ = `r round(coef(mdl1)[1],2)`  
2. $\hat \beta_1$ = `r round(coef(mdl1)[2],2)`  
3. $\hat \beta_2$ = `r round(coef(mdl1)[3],2)`  

`r solend()`

:::frame 
__Interpretation of Muliple Regression Coefficients__

You'll hear a lot of different ways that people explain multiple regression coefficients. For a model $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$, the estimate $\hat \beta_1$ will often be reported as the increase in $y$ for a one unit increase in $x_1$ when...

- holding the effect of $x_2$ constant.
- controlling for differences in $x_2$.
- partialling out the effects of $x_2$.
- holding $x_2$ equal. 
- accounting for effects of $x_2$. 

:::int
```{r, echo=FALSE}
summary(mdl1)$coefficients
```

The coefficient 2.25 of Education for predicting Income says that among employes of the same Seniority, those who have one additional year of education tend to, on average, have $2.25 thousand dollars more income. The multiple regression coefficient measures that average _conditional_ relationship.
:::

:::

`r qbegin(8)`
Within what distance from the model predicted values (the regression surface) would we expect 95% of employee incomes to be?

**Hint:** either `sigma(mdl1)` or part of the output from `summary(mdl1)` will help you here.  
`r qend()`
`r solbegin(show=SHOW_SOLS)`
The estimated standard deviation of the errors is $\hat \sigma$ = `r round(sigma(mdl1),2)`. We would expect 95% of employee incomes to be within about `r paste0("$",round(sigma(mdl1)*2,1)*1000)` ($2 \hat \sigma$) from the model fit. 

`r solend()`

`r qbegin(9)`
Obtain 95\% confidence intervals for the regression coefficients, and write a sentence for each.  
`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r}
confint(mdl1, level = 0.95)
```

:::int 

+ The average income for all Riverview city employees with zero years of education and zero years of seniority is between `r paste0("$",round(confint(mdl1, level=.95)[1,1],2)*1000)` and `r paste0("$",round(confint(mdl1, level=.95)[1,2],2)*1000)`.  
+ For all Riverview city employees, when _holding seniority level constant_, each one-year difference in formal education is associated with a difference in income between `r paste0("$",round(confint(mdl1, level=.95)[2,1],2)*1000)` and `r paste0("$",round(confint(mdl1, level=.95)[2,2],2)*1000)`, on average.  
+ For all Riverview city employees, when _holding education level constant_, each one-year difference in seniority is associated with a difference in income between `r paste0("$",round(confint(mdl1, level=.95)[3,1],2)*1000)` and `r paste0("$",round(confint(mdl1, level=.95)[3,2],2)*1000)`, on average.  

:::

`r solend()`

    
## ~ Quantitative + Qualitative

> Suppose that the group of researchers studying income differences in Riverview (our hypothetical Midwestern city in the US) were concerned about the gender pay gap - that is, they were concerned that men were being paid more than women. Moreover, they were worried that this discrimination was happening even *after* taking into account peoples' seniority.  


`r qbegin(9)`
We have already visualised the marginal distribution of employee years of seniority in an earlier question, as well as its relationship with employee incomes.  
Produce visualisations of:  

1. the distribution of employee genders
2. the relationship between gender and income. 

**Note:** We cannot visualise the distribution of gender as a density curve or boxplot, because it is a _qualitative_ variable (observations can only take one of a set of discrete response values).  
<small>This set of possible response values will depend on how 'Gender' is measured. The [riverview.csv](#dc-riverview) dataset contains only two ("male" and "female"), but this is not an endorsement to the view that gender is a binary variable.</small>  
`r qend()`
`r solbegin(show=SHOW_SOLS)`
`geom_bar()` will count the number of observations falling into each unique level of the gender variable:
```{r, fig.cap="Marginal distribution plots of employee genders"}
ggplot(data = riverview, aes(x = gender)) +
  geom_bar()+
  labs(x = "Gender", y = "Frequency")
```

We might plot the relationship between gender and income as two boxplots:
```{r, fig.cap="Marginal distribution plots of employee genders"}
ggplot(data = riverview, aes(x = gender, y = income)) +
  geom_boxplot()+
  labs(x = "Gender", y = "Income (thousands of US $)")
```
`r solend()`  

:::frame  
__Qualitative predictors__

When we have a _qualitative_ explanatory variable such as gender, what actually gets inputted into our model is a number of _dummy variables_ which are numeric variables that represents categorical data.  

For a qualitative variable with $k$ levels, we can express it in $k-1$ dummy variables:  
```{r echo=FALSE, out.width="80%"}
data.frame(
  species = c("cat","cat","dog","parrot","dog","cat","..."),
  species_dog = c(0,0,1,0,1,0,"..."),
  species_parrot = c(0,0,0,1,0,0,"...")
)
```

+ The "cat" level is expressed whenever both the "species_dog" and "species_parrot" variables are 0.
+ The "dog" level is expressed whenever the "species_dog" variable is 1 and the "species_parrot" variable is 0.
+ The "parrot" level is expressed whenever the "species_dog" variable is 0 and the "species_parrot" variable is 1.  

R will do all of this for us. If we include in our model a qualitative explanatory variable 4 different categories, the model will estimate 3 parameters - one for each dummy variable. We can interpret the parameter estimates (the coefficients we obtain using `coefficients()`,`coef()` or `summary()`) as the estimated increase in the outcome variable associated with an increase of one in each dummy variable (holding all other variables equal). Note that in the above example, an increase in 1 of "species_dog" is the difference between a "cat" and a "dog". An increase in one of "species_parrot" is the difference between a "cat" and a "parrot". We think of the "cat" category in this example as the _reference level_ - it is the category against which other categories are compared against. 

:::  


`r qbegin(10)`
Fit the multiple regression model below using `lm()`, and assign it to an object named `mdl2`. Examine the summary output of the model. 

$$
Income = \beta_0 + \beta_1 \cdot Seniority + \beta_2 \cdot Gender
$$
`r qend()`
`r solbegin(show=SHOW_SOLS)`
```{r}
mdl2 <- lm(income ~ seniority + gender, data = riverview)
summary(mdl2)
```
`r solend()`

`r qbegin(11)`
$\hat \beta_0$ (the intercept) is the estimated average income associated with zero years of seniority and zero in the gender variable.  

What does zero in the gender variable mean? What group is this the estimated income for when they have zero years of seniority? Why?  

`r qend()`
`r solbegin(show=SHOW_SOLS)`
As you can see in the output of the model, we have a coefficient called `gendermale`. This is the parameter estimate for a dummy variable which has been inputted into the model. The `lm()` function will automatically name the dummy variables (and therefore the coefficients) according to what level is identified by the 1. So we know that `gendermale` is 1 for "male" and 0 for "female".  

The intercept is therefore the estimated income for female employees with zero years of seniority.  
`r solend()`

`r qbegin(12)`
In the previous example, we had a visualisation of our model as a regression surface (Figure \@ref(fig:regsurf)).  
Here, one of our explanatory variable has only two possible responses. How might we visualise the model?    

a. one line
a. one surface
a. two lines
a. two surfaces
a. a curved (not flat) surface 

`r qend()`
`r solbegin(show=SHOW_SOLS)`
We can visualise the model $\widehat{Income} = \hat \beta_0 + \hat \beta_1 \cdot Seniority + \hat \beta_2 \cdot Gender$ as two lines.  
Each line represents the model predicted values for employee income across the range of years of seniority, with one line for "male" and one for "female".  
`r solend()`

`r qbegin(13)`
Get a pen and paper, and sketch out the plot shown in Figure \@ref(fig:plot-annotate).  

```{r plot-annotate, echo=FALSE, fig.cap="Multiple regression model: Income ~ Seniority + Gender"}
sjPlot::plot_model(mdl2, type="pred", terms=c("seniority","gender"), show.data=FALSE)+
  scale_fill_manual(NULL, values=c(NA,NA))
```

Annotate your plot with labels for each of parameter estimates from your model: 

| Parameter Estimate   |      Model Coefficient      |  Estimate |
|----------|:-------------:|------:|
| $\hat \beta_0$ | `(Intercept)` | `r round(coef(mdl2)[1],2)` |
| $\hat \beta_1$ | `seniority`   |  `r round(coef(mdl2)[2],2)` |
| $\hat \beta_2$ | `gendermale` | `r round(coef(mdl2)[3],2)` |


`r optbegin("Hint",olabel=FALSE)`
Below you can see where to add the labels, but we have not said which is which. 

```{r echo=FALSE}
sjPlot::plot_model(mdl2, type="pred", terms=c("seniority","gender"), show.data=FALSE)+
  scale_fill_manual(NULL, values=c(NA,NA))+
  geom_vline(xintercept = 0) +
  geom_segment(aes(x=1,xend=0,
                   y=predict(mdl2,newdata=data.frame(gender="female",seniority=0))-1,
                   yend=predict(mdl2,newdata=data.frame(gender="female",seniority=0))),
               col="black",lwd=.2)+
  geom_label(aes(x=2,y=predict(mdl2,newdata=data.frame(gender="female",seniority=0))-3), 
             label="B", col="black")+
  geom_segment(aes(x=20,xend=20,
                   y=predict(mdl2,newdata=data.frame(gender="female",seniority=20)),
                   yend=predict(mdl2,newdata=data.frame(gender="male",seniority=20))),
               col="black",lwd=.2,
               arrow = arrow(length = unit(3, "mm")))+
  geom_label(aes(x=21,y=predict(mdl2,newdata=data.frame(gender="male",seniority=20))-5),
               label="A", col="black")+
  geom_segment(aes(x=10,xend=11,
                 y=predict(mdl2,newdata=data.frame(gender="female",seniority=10)),
                 yend=predict(mdl2,newdata=data.frame(gender="female",seniority=10))),
             col="black",lwd=.2)+
  geom_segment(aes(x=11,xend=11,
                 y=predict(mdl2,newdata=data.frame(gender="female",seniority=10)),
                 yend=predict(mdl2,newdata=data.frame(gender="female",seniority=11))),
             col="black",lwd=.2)+
  geom_label(aes(x=12,y=predict(mdl2,newdata=data.frame(gender="female",seniority=10))),
                 label="C",col="black")+
  NULL -> plot_annotate

plot_annotate

```

+ A is the vertical distance between the red and blue lines (the lines are parallel, so this distance is the same wherever you cut it on the x-axis).  
+ B is the point at which the blue line cuts the y-axis.  
+ C is the vertical increase (increase on the y-axis) for the blue line associated with a 1 unit increase on the x-axis (the lines are parallel, so this is the same for the red line).  

`r optend()`
`r qend()`
`r solbegin(show=SHOW_SOLS)`

+ A = $\hat \beta_2$ = `gendermale` coefficient = `r round(coef(mdl2)[3],2)`
+ B = $\hat \beta_0$ = `(Intercept)` coefficient  = `r round(coef(mdl2)[1],2)`
+ C = $\hat \beta_1$ = `seniority` coefficient = `r round(coef(mdl2)[2],2)` 

```{r echo=FALSE}
plot_annotate
```

`r solend()`

