# Interactions {#chap-int}

```{r echo=FALSE}
# Solutions
SHOW_SOLS = TRUE

# Chunk options
knitr::opts_chunk$set(
    message = FALSE, 
    warning = FALSE, 
    fig.align = 'center',
    out.width = '60%'
)

# ggplot settings
library(ggplot2)
theme_set(
  theme_bw(base_size = 15) + 
    theme(plot.title = element_text(hjust = 0.5))
)

# reproducibility
set.seed(953)
```

:::red
**PRE-LAB ACTIVITIES**

Before attempting the lab please read ?.
:::

:::lo
**LEARNING OBJECTIVES**

1. Understand the concept of an interaction
2. Being able to interpret the meaning of an interaction in the quant-quant and mixed predictors cases
3. Test hypothesis on the coefficients of interactions terms and the principle of marginality
4. (Categorical pred.) Understand the meaning of a MLR model with categorical predictors (interaction case). For example, two non-parallel regression lines, one for each of two groups.

:::

<center>
<h3 style="color: #BF1932;">
Response: Quantitative<br />Predictor<b>s</b>: Quantitative*Qualitative
</h3>
</center>
<br />


## Introduction and research question  


## Choose 

### Data exploration {-}


### Model Specification {-}


## Fit  

## Understanding the fit



<center>
<h3 style="color: #BF1932;">
Response: Quantitative<br />Predictor<b>s</b>: Quantitative*Quantitative
</h3>
</center>
<br />





### Overclaiming {-}


As our first example, we will use some explore some data from a study investigating whether peoples' self-perception of how knowledgeable they are is related to 'overclaiming' (claiming to know about things which do not exist).  

@Atir2015 recruited 202 participants, and asked them to complete a measure of self-perceived knowledge, and an overclaiming task. 
The self-perceived knowledge questionnaire asked participants to indicate their level of knowledge in personal finance.  
The overclaiming task asked participants to indicate how much they knew about 15 terms related to personal finance. However, three of these terms did not exist ("pre-rated stocks", "fixed-rate deduction", "annualized credit"). 
Finally, participants also completed a test of financial literacy called the FINRA, which measured actual knowledge (rather than self-perceived knowledge) of personal finance.  

> Overclaiming was measured by calculating the false alarm rate, which is the proportion of nonexistent foils about which a participant claimed knowledge. We averaged the false alarm rates for each of the six potential knowledge cutoff points (i.e., we computed the proportion of foils rated as 2 or higher, the proportion of foils rated as 3 or higher, and so on for 4, 5, 6, and 7, and then averaged these proportions), which resulted in an overclaiming value ranging from 0 to 1.
>
> Accuracy was obtained by subtracting the averaged false alarm rate from the averaged hit rate (i.e., the proportion of real items about which each participant claimed knowledge, averaged across all six potential cutoff points). 
>
> --- @Atir2015

The [atir1b.csv](#dc-atir1b) contains data from this study. The [data codebook](#dc-atir1b) contains a preview of the data and a link to download the file. 



`r qbegin(1)`
Create a new RMarkdown, and in the first code-chunk, load the required libraries and import the data into R. Assign them to a variable called `atir1b`.
`r qend()`
`r solbegin()`
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork) #used in the next section for plots

# Read in data
atir1b = read_csv(file = "data/atir1b.csv")
head(atir1b)
```
`r solend()`



```{r fig.cap="Marginal distribution plots of overclaiming, self-perceived knowledge, and accuracy", out.width="80%"}
ggplot(data = atir1b, aes(x = overclaiming_proportion)) +
  geom_density() +
  geom_boxplot(width = 1/10)

ggplot(data = atir1b, aes(x = self_perceived_knowledge)) +
  geom_density() +
  geom_boxplot(width = 1/20)

ggplot(data = atir1b, aes(x = accuracy)) +
  geom_density() +
  geom_boxplot(width = 1/10)
```








## model building 

:::frame
__Adjusted $R^2$__

We saw in Chapter \@ref(chap-slr) that the R-squared can be obtained as:
$$
R^2 = \frac{SSModel}{SSTotal} = 1 - \frac{SSE}{SSTotal}
$$

However, when we add more and more predictors into a multiple regression model, $SSE$ cannot increase, and may decrease by pure chance alone, even if the predictors are unrelated to the outcome variable. Because $SSTotal$ is constant, the calculation $1-\frac{SSE}{SSTotal}$ will increase by chance alone. 

An alternative, the Adjusted-$R^2$, does not necessarily increase with the addition of more explanatory variables, by including a penalty according to the number of explanatory variables in the model. It is not by itself meaningful, but can be useful in determining what predictors to include in a model. 
$$
Adjusted{-}R^2=1-\frac{(1-R^2)(n-1)}{n-k-1} \\
\quad \\
\text{where} \quad n = \text{sample size} \\
k = \text{number of explanatory variables} 
$$
:::

