---
title: "Writing-up"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: FALSE
---

```{r setup, include=FALSE}
source('assets/setup.R')
```
  
:::lo
**LEARNING OBJECTIVES**

1. Construct a linear model from information in published paper. 
2. Evaluate a published analysis with respect to the thoroughness of information presented. 
3. Write-up and provide interpretation of a linear model analysis with multiple predictors

:::
  
:::red
**Pre-Lab activity**

Please read @Przybylski2017.   
You can find it at https://journals.sagepub.com/doi/full/10.1177/0956797616678438

:::
  



> **Research question**   
> Previous research has identified an association between an individual's perception of their social rank and symptoms of depression, anxiety and stress. We are interested in the individual differences in this relationship.  
Specifically: 
> 
> + Controlling for other personality traits, does neuroticism moderate effects of social comparison on symptoms of depression, anxiety and stress?  
>  

```{r}
library(tidyverse)
scs_study <- read_csv("https://uoepsy.github.io/data/scs_study.csv")

# scale scs score
scs_study <- 
  scs_study %>% 
    mutate(
      zscs = scale(scs)
    )

dass_mdl <- lm(dass ~ 1 + zscs*zn + zo + zc + ze + za, data = scs_study)
```

```{r echo=FALSE}
par(mfrow=c(2,2))
```

```{r}
plot(dass_mdl2)
```

```{r echo=FALSE}
par(mfrow=c(1,1))
```

```{r}
dass_mdl2 <- lm(dass ~ 1 + zscs*zn + zo + zc + ze + za, data = scs_study[-35, ])

library(car)

residualPlots(dass_mdl2)
ncvTest(dass_mdl2)

shapiro.test(residuals(dass_mdl2))

dwt(dass_mdl2)

vif(dass_mdl2)
```


0.
- describe study design, data collection, etc.  

1. Describe data

- how many observational units? 
- describe the variables of interest - how are they scored, have they been transformed at all?  
- covariances/correlations

2. describe analytical approach

- what type of statistical analysis do you use to answer the question? (e.g., t-test, simple linear regression, multiple linear regression)
- describe the model/analysis structure
  - what is your outcome variable?
  - what are your predictors? 
  - any other specifics?

3. planned analysis vs actual analysis

- Was there anything you had to do differently than planned during the analysis?  
  - Did you exclude any observations?
  - Did you have to do anything (e.g., transform any variables) in order to meet assumptions? 
  - 

4. present results

- provide table of results if applicable (for regression tables, try `tab_model()` from the **sjPlot** package). 
- provide plot if applicable

5. interpret results. 

- what do your results suggest about your research question?
  - be specific - which statistic/test says what?



