---
title: "More About Logistic Regression"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('assets/setup.R')
library(tidyverse)
library(kableExtra)
library(patchwork)
```



:::yellow
__PREREQUISITES__

Let $A$ and $B$ be two positive numbers:

$$
\begin{aligned}
e^{A + B} &= e^A \times e^B \\
e^{\log(A)} &= A \\
\log \left( \frac{A}{B} \right) &= \log(A) - \log(B) \\
\log (A \times B) &= \log(A) + \log(B)
\end{aligned}
$$

:::



# Senility and WAIS


A sample of elderly people was given a psychiatric examination to determine symptoms of senility were present. Other measurements taken at the same time included the score on a subset of the Wechsler Adult Intelligent Scale (WAIS). 

The data represent symptoms of senility (`senility` = 1 if symptoms are present and `senility`= 0 otherwise) and WAIS scores (`wais`) for $n = 54$ people. 
You can download the data at following link: https://uoepsy.github.io/data/SenilityWAIS.csv

```{r}
library(tidyverse)

sen <- read_csv(file = 'https://uoepsy.github.io/data/SenilityWAIS.csv')
head(sen)
```

First, we should plot the data to understand what's going on. We use `geom_jitter` rather than `geom_point` otherwise we will have multiple points on top of each other as there are multiple individuals with the same wais' values:
```{r}
ggplot(sen, aes(wais, senility)) +
  geom_jitter(width = 0, height = 0.05) +
  labs(x = "x, WAIS score", 
       y = "y, Has senility symptoms?")
```

Since we are interested in understanding how the probability of having senility symptoms changes as a function of the WAIS score, the response variable is $y$ = `senility` and the predictor is $x$ = `wais`. 

We will fit the following logistic regression model:
$$
\log \left( \frac{p}{1-p} \right) = \beta_0 + \beta_1 \ x
$$
where $p = P(y = 1)$ is the probability that $y = 1$ for an individual with WAIS score equal to $x$.

```{r}
mdl1 = glm(senility ~ wais, family = binomial, data = sen)
summary(mdl1)
```




# Interpretation of coefficients

To interpret the fitted coefficients, we first exponentiate the model:
$$
\begin{aligned}
\log \left( \frac{p}{1-p} \right) &= \beta_0 + \beta_1 x \\
e^{ \log \left( \frac{p}{1-p} \right) } &= e^{\beta_0 + \beta_1 x } \\
\frac{p}{1-p} &= e^{\beta_0} \ e^{\beta_1 x}
\end{aligned}
$$

and recall that the probability of success divided by the probability of failure is the odds of success:
$$
\frac{p}{1-p} = \text{odds}(y = 1)
$$

__Intercept__

If $x = 0$:
$$
\frac{p_0}{1-p_0} = e^{\beta_0} \ e^{\beta_1 x} = e^{\beta_0}
$$

That is, $e^{\beta_0}$ represents the odds of having symptoms of senility for individuals with a WAIS score of 0. 

In other words, for those with a WAIS score of 0, the probability of having senility symptoms is $e^{\beta_0}$ times that of non having them.


__Slope__

If $x = 1$:
$$
\frac{p_1}{1-p_1} = e^{\beta_0} \ e^{\beta_1 x} = e^{\beta_0} \ e^{\beta_1}
$$

Now consider taking the ratio of the odds of senility symptoms when $x=1$ to the odds of senility symptoms when $x = 0$:
$$
\frac{p_1 / (1 - p_1)}{p_0 / (1 - p_0)} 
= \frac{e^{\beta_0} \ e^{\beta_1}}{e^{\beta_0}}
= e^{\beta_1}
$$


So, $e^{\beta_1}$  represents the odds ratio for a 1 unit increase in WAIS score.

It is typically interpreted by saying that for a one-unit increase in WAIS score the odds of senility symptoms increase by a factor of $e^{\beta_1}$.


__In our model__

```{r}
exp(coef(mdl1))
```

The odds of senility symptoms for individuals scoring 0 on the WAIS test is 11:1. Or, the probability of having senility symptoms for those scoring 0 on the WAIS test is 11 times the probability of not having senility symptoms.

For every unit increase in WAIS score, the odds of senility symptoms reduces by a factor of 0.72



<br>

`r optbegin("Distinct covariate values (grouped data)")`

If you explore the full data, you will see that even though there are $n = `nrow(sen)`$ individuals, there are only 17 distinct values of the covariate $x$ (representing the WAIS scores), as some people have the same WAIS scores but different $y$ values.

Let's summarise the original data by computing for each covariate, the total number of people who showed senility symptoms ($y$) and the total number of people with that distinct covariate value $n$. This is the "grouped data":

```{r}
sen_grp <- sen %>%
  group_by(wais) %>%
  summarise(num_1s = sum(senility), total = n())

sen_grp
```

You can fit a logistic regression model to grouped data as follows. This is __completely equivalent__ to the one we fitted above.

1. Provide two columns, one giving the number of successes (s = 1) and the other giving the number of failures (s = 0) which is just $n - y$:

```{r}
mdl1a <- glm(cbind(num_1s, total - num_1s) ~ wais, data = sen_grp, family = binomial)
summary(mdl1a)
```

2. Provide the proportion of successes and the number of people with that distinct covariate in weights:
```{r}
sen_grp <- sen_grp %>%
  mutate(prop_1s = num_1s / total)

head(sen_grp)
```

```{r}
mdl1b <- glm(prop_1s ~ wais, weights = total, data = sen_grp, family = binomial)
summary(mdl1b)
```

`r optend()`




# Diagnosing model fit: Residuals

Unlike linear regression, logistic regression has 2 main types of residuals.

1. Pearson residuals
2. Deviance residuals

We will now discuss each in turn.

In addition, just like in linear regression, each of the above can be standardized or studentized if you wish to. We won't discuss how to standardize each of those as the formula is difficult, but the idea is to standardize them to make them have a mean of zero and standard deviation = 1.

Remember the difference?

- $i$th standardized residual = residual / SD(residual)
- $i$th studentized residual = residual / SD(residual from model fitted without observation $i$)



## Pearson residuals

Pearson residuals are similar to the Pearson residuals you might remember from the chi-squared test in DAPR1.
$$
Pres_i = \frac{Observed - Expected}{\sqrt{Expected}}
$$

In logistic regression, this is 
$$
Pres_i = \frac{y_i - \hat p_i}{\sqrt{\hat p_i (1 - \hat p_i)}}
$$

where 

- $y_i$ is the observed response for unit $i$, either 0 (failure) or 1 (success)
- $\hat p_i$ is the model-predicted probability of success

There is one residual per each case (row) in the dataset:
```{r}
Pres <- residuals(mdl1, type = 'pearson')
```


The standardized Pearson residuals (zero mean and unit standard deviation) are obtained as follows:
```{r}
SPres <- rstandard(mdl1, type = 'pearson')
```


The standardized Pearson residuals (also zero mean and unit standard deviation) are obtained as follows:
```{r}
StuPres <- rstudent(mdl1, type = 'pearson')
```


__In the following, I will use the Studentized residuals for the plots, but if you wish to, you can use the standardized ones.__


First, we plot the Studentized Pearson Residuals against their index, and we check whether there are any values larger than 2 in absolute value (that is larger than 2 or smaller than -2). Some people prefer to be strict and deem as outlier those greater than 3 in absolute value as most values should be within -3 and 3.

```{r}
plot(StuPres, ylab = "Studentized Pearson Residuals")
```


Yes, there appear to be 3 residuals with a value slightly larger than 2 in absolute value. We will keep these in mind and check later if they are also influential points.


<br>

Plotting the (Studentized) Pearson residuals against the fitted values or the predictor is not very informative in logistic regression as you can see below:
```{r}
plot(fitted(mdl1), StuPres, 
     xlab = 'Prob. of Success', ylab = 'Studentized Pearson Residuals')
```

```{r}
plot(sen$wais, StuPres, 
     xlab = 'Prob. of Success', ylab = 'Studentized Pearson Residuals')
```


Sometimes a __binned plot__ [^gelman-hill] can be more informative, but not always! It works by combining together all responses for people having the same covariate $x_i$ value, and taking the average Studentized Pearson residual for those.

Before using this function, make sure you have installed the **arm** package!

```{r}
arm::binnedplot(fitted(mdl1), StuPres, 
                xlab = 'Prob. of Success', ylab = 'Studentized Pearson Residuals')
```

There doesn't appear to be any extreme values.



[^gelman-hill]: Gelman, A., & Hill, J. (2006). Data Analysis Using Regression and Multilevel/Hierarchical Models (Analytical Methods for Social Research). Cambridge: Cambridge University Press. doi:10.1017/CBO9780511790942


## Deviance residuals

What does _deviance_ mean? In logistic regression, deviance is a measure of deviation, discrepancy, mismatch between the data and the model. You can think of it as a generalisation of the terms making up the residual sum of squares in simple linear regression.


```{r}
Dres <- residuals(mdl1, type = 'deviance')
```


The standardized deviance residuals (zero mean and unit standard deviation) are obtaiend with

```{r}
SDres <- rstandard(mdl1, type = 'deviance')
```

And the Studentized deviance residuals (also zero mean and unit standard deviation) are found via:
```{r}
StuDres <- rstudent(mdl1, type = 'deviance')
```


Again, we check whether any residuals are larger than 2 or 3 in absolute value:
```{r}
plot(StuDres, ylab = 'Studentized Deviance Residuals')
```

All fine!



Again, a plot against the fitted values or the predictors isn't that useful in logistic regression...
```{r}
plot(fitted(mdl1), StuDres, 
     xlab = 'Prob. of Success', ylab = 'Studentized Deviance Residuals')
```

```{r}
plot(sen$wais, StuDres, 
     xlab = 'Prob. of Success', ylab = 'Studentized Deviance Residuals')
```

So we will check against a binned plot:
```{r}
arm::binnedplot(fitted(mdl1), StuDres, 
                xlab = 'Prob. of Success', ylab = 'Studentized Deviance Residuals')
```

Again, it seems like everything is fine, no extremely high values.



## Which residuals should I use???

Both! Compute both, and for visual exploration most of the time it's fine to use the __deviance__ residuals.

<!-- You will get away with it -->

If you use standardized or studentized ones it's easier to explore extreme values as we expect most residuals to be within -2, 2 or -3, 3.

If you do not provide the `type = ...` argument to the function `residuals()`, then it will use the Deviance residuals by default, and so will the function `rstandard()` and `rstudent()`!



# Influential values

In logistic regression we typically check for influential observations by checking if there are any of the `r nrow(sen)` cases in the dataset that have a Cook's distance greater than 0.5 (moderately influential) or 1 (highly influential):

```{r}
plot(cooks.distance(mdl1), ylab = "Cook's distance")
```

None of the units in the dataset appears to have a Cook's distance value greater than 0.5, hence there does not seem to be issues with influential points.



# Diagnosing model fit: Goodness of fit statistics

In logistic regression there are two measures of goodness of fit. These are an extension of the ANOVA F-test in linear regression which test whether all slopes were 0 or at least one was not.

We test whether
$$
\begin{aligned}
H_0 &: \beta_1 = ... = 0 \\
H_1 &: \text{at least one }\beta \neq 0
\end{aligned}
$$

using either:

- Pearson chi-squared statistic
- Hosmer–Lemeshow statistic

We are super lucky, as we already saved both Pearson residuals and the Deviance residuals, __not the standardized__ and __not the studentized__ ones, we just need  to do respectively:

$$
X^2 = \sum Pres_i^2
$$

$$
D = \sum Dres_i^2
$$

## Pearson chi-squared statistic

The Pearson chi-squared statistic is computed as
$$
X^2 = \sum Pres_i^2
$$

where $Pres_i$ is the $i$th Pearson residual from `residuals(<fitted model>, type = 'pearson')`.

The test-statistic follows a chi-squared distribution with $m - p$ degrees of freedom where

- $m$ = number of distinct covariate combinations
- $p$ = number of estimated parameters



We are so lucky, because we already did compute the required residuals.
```{r}
m <- length(unique(sen$wais))
m

p <- length(coef(mdl1))
p

X2 <- sum(Pres^2)
X2

1 - pchisq(X2, df = m - p)
```


:::red
__WARNING__

This Pearson chi-squared goodness-of-fit test works reasonably well when the number of distinct covariate combinations is neither too small nor too large. 
If the number of patterns is too small, the test for goodness of fit is too coarse, whereas if the number of patterns is too large, the predicted number of successes for many of the patterns in the independent variables will be too small (i.e., the data are "thin" with less than five successes for many patterns), and so the distribution of the computed chi-squared is not reasonably described by the theoretical chi-squared distribution. 

This situation is particularly likely to occur when one or more of the independent variables are continuous or in clinical studies when the total sample size is relatively small (below 100 patients).

Do not despair though! There is the Hosmer-Lemeshow goodness-of-fit test...
:::



# Hosmer-Lemeshow goodness-of-fit test

The Hosmer-Lemeshow goodness-of-fit test can be implemented via the `performance` package. Don't forget to install it first with `install.packages("performance")`!

It divides the fitted probabilities into 10 bins by default, specified into `n_bins = 10`. Sometimes when we have small datasets, this might be too much, and we need to reduce it. It is good practice to try different values and see how much the p-value changes. If it is fairly consistent and above the significance level, you should be safe to accept goodness of fit!

```{r}
# performance::performance_hosmer(mdl1, 10)   # returns an error!
# performance::performance_hosmer(mdl1, 9)    # returns an error!
performance::performance_hosmer(mdl1, 8)
performance::performance_hosmer(mdl1, 7)
performance::performance_hosmer(mdl1, 6)
performance::performance_hosmer(mdl1, 5)
```

At the 5% significance level, the p-values are non-significant. Hence, the sample results do not provide sufficient evidence to reject the null hypothesis that the model fits the data well.


## Comparison of nested models

When moving from linear regression to more advanced and flexible models, testing of goodness of fit is more often done by comparing a wanted model to a simpler one.
The only caveat is that the two models need to be nested, i.e. all predictors of one model needs to be within the other.

We want to compare the model we previously fitted against a model where all slopes are 0, i.e. a baseline model:
$$
\begin{aligned}
M_1 : \qquad\log \left( \frac{p}{1 - p} \right) &= \beta_0 \\
M_2 : \qquad \log \left( \frac{p}{1 - p} \right) &= \beta_0 + \beta_1 x
\end{aligned}
$$

The null hypothesis will be that the simpler model is a good fit, while the alternative is that the more complex model is needed.

In R we do the comparison as follows:
```{r}
mdl_red <- glm(senility ~ 1, family = binomial, data = sen)

anova(mdl_red, mdl1, test = 'Chisq')
```

The above code shows the two fitted models
```
Model 1: senility ~ 1
Model 2: senility ~ wais
```

And then reports the _Residual Deviance_ of each model, 61.806 and 51.017 respectively. Remember the deviance is the equivalent of residual sum of squares in linear regression.

So, by adding the predictor $x$ = wais to the model, we reduce our deviance from 61.806 to 51.017, i.e. we reduce it by 10.789. 
Is this reduction sufficient to be attributed solely to the contribution of the predictor, or could it just be due to random sampling variation? This is what the chi-squared test tells us! 


:::int
The reduced model ($M_1$) has the slope set to zero $\beta_1 = 0$. Its deviance is obtained by fitting a logistic regression model without any explanatory variables (but including a constant term). This deviance is found to be 61.806, with 53 degrees of freedom.
The deviance statistic from a fit of the model with $x$ = wais included is 51.017, with 52  degrees of freedom. The deviance dropped by 10.789, with a drop of one degree of freedom. The associated p-value from a chi-squared distribution with 1df is $p = .001$.

There is strong evidence of an association between senility symptoms and scores on the Wechsler Adult Intelligent Scale.
:::



<!-- Above was Drop-in-deviance GOF test -->
<!-- # Deviance GOF test -->
<!-- mdl_sat = glm(senility ~ factor(1:nrow(sen)), family = binomial, data = sen) -->
<!-- anova(mdl1, mdl_sat, test = 'Chisq') -->






# Exercises


Kalkhoran et al. (2015) [^1] investigated predictors of dual use of cigarettes and smokeless tobacco products. Although their original sample size was large ($n$ = 1324), they were interested in running separate logistic regression analyses within subgroups. Most of the sample used only cigarettes. One of the smaller subgroups contained subjects who used both cigarettes and smokeless tobacco products ($n$=61). For this problem we focus on this smaller subgroup of dual cigarette and smokeless tobacco users. The dependent variable, Q, is whether the subject made an attempt to quit using tobacco products (0 = no attempt to quit; 1 = attempted to quit). There is one multi-category independent variable, intention to quit, I, with four levels: never intend to quit, may intend to quit but not in the next 6 months, intend to quit in the next 6 months, intend to quit in the next 30 days. 

The data can be found at the following link: https://uoepsy.github.io/data/QuitAttempts.csv

```{r echo=FALSE}
library(tidyverse)
library(kableExtra)

cb <- tibble(
  Variable = c('S', 'Q', 'I'),
  Description = c(
    'Subject',
    'Quitting (0 = No attempt to quit; 1 = Attempted to quit)',
    'Intentions to Quit (1 = Never intend to quit; 2 = May intend to quit but not in the next 6 months; 3 = Intend to quit in the next 6 months; 4 = Intend to quit in the next 30 days)'
  )
)

kable(cb, caption = 'Quit Attempts and Intentions Among Dual Users of Cigarettes and Smokeless Tobacco Products') %>%
  kable_styling(full_width = FALSE)
```


`r qbegin(1)`
Read the data into R.

Perform a preliminary exploratory analysis of the data by computing summary statistics and visualising the distribution of the variables.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
qa <- read_csv('https://uoepsy.github.io/data/QuitAttempts.csv')
qa
```

```{r}
qa %>%
  group_by(I) %>%
  count(Q)
```


```{r}
ggplot(qa, aes(I, fill = factor(Q))) +
  geom_bar() +
  labs(fill = 'Attempted quitting?')
```

`r solend()`


`r qbegin(2)`
Fit a logistic regression of Q on I using ordinary logistic regression. 

Be sure to treat the independent variable I as categorical rather than as continuous in these analyses (hint: you will need to make sure R will create for you dummy variables for each level of I).
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
First, we must make sure `I` is a factor otherwise R won't create the dummy variables for us:
```{r}
qa$I <- as.factor(qa$I)
```

```{r}
qa_fit1 = glm(Q ~ I, data = qa, family = binomial)
summary(qa_fit1)
```
`r solend()`


`r qbegin(3)`
What do you observe from the results?

Interpret the coefficients in the context of the study.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(4)`
Investigate whether or not the studentized deviance residuals raise any concerns about outliers.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
plot(residuals(qa_fit1), ylab = "Studentized Deviance Residuals")
plot(fitted(qa_fit1), residuals(qa_fit1),
     xlab = "Fitted values", ylab = "Studentized Deviance Residuals")
arm::binnedplot(fitted(qa_fit1), residuals(qa_fit1),
                ylab = "Average Studentized Deviance Residuals")
```

`r solend()`


`r qbegin(5)`
Investigate whether or not there are influential observations.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
plot(cooks.distance(qa_fit1), ylab = "Cook's Distance")
```

`r solend()`


`r qbegin(6)`
Compute the Pearson goodness-of-fit statistic and report its p-value.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
m <- length(unique(qa$I))
m

p <- length(coef(qa_fit1))
p

X2 <- sum(residuals(qa_fit1, type = 'pearson')^2)
X2

1 - pchisq(X2, m - p)
```

`r solend()`



`r qbegin(7)`
Compute the Hosmer-Lemeshow goodness-of-fit statistic and report its p-value.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
D <- sum(residuals(qa_fit1, type = 'deviance')^2)
D
deviance(qa_fit1)

1 - pchisq(D, m - p)
```
`r solend()`


`r qbegin(8)`
Perform a Deviance goodness-of-fit test by compare the following nested models:

$$
\begin{aligned}
M_1 &: \qquad \log \left( \frac{p}{1 - p}\right) = \beta_0 \\
M_2 &: \qquad \log \left( \frac{p}{1 - p}\right) = \beta_0 + \beta_1 I2 + \beta_2 I3 + \beta_3 I4
\end{aligned}
$$
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
qa_null <- glm(Q ~ 1, data = qa, family = binomial)
summary(qa_null)
```

```{r}
anova(qa_null, qa_fit1, test = 'Chisq')
```

`r solend()`






[^1]: Kalkhoran S, Grana RA, Neilands TB, and Ling PM. Dual use of smokeless tobacco or e-cigarettes with cigarettes and cessation. _Am J Health Behav._ 2015;39:277–284.


<!-- Formatting -->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>


