---
title: "Dummy Coding"
link-citations: TRUE
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')
set.seed(3)

library(DT)
library(tidyverse)
```

:::lo

### <i class="fa fa-graduation-cap"></i> Learning Objectives
At the end of this lab, you will:

1. Understand how to specify a new baseline/reference level for categorical variables
2. Understand how to specify dummy coding
3. Interpret the output from a model using dummy coding
4. Understand how to specify contracts to test specific effects

### <i class="fa fa-check-square-o fa-2"></i> What You Need

1. Be up to date with lectures

### <i class="fab fa-r-project"></i> Required R Packages
Remember to load all packages within a code chunk at the start of your RMarkdown file using `library()`. If you do not have a package and need to install, do so within the console using `install.packages(" ")`. For further guidance on installing/updating packages, see Section C [here](https://uoepsy.github.io/files/install-update-r#update-pkgs). 

For this lab, you will need to load the following package(s):

* **tidyverse** 
* **psych** 
* **kableExtra**
* **emmeans**

### <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Presenting Results
All results should be presented following [APA guidelines](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf). If you need a reminder on how to hide code, format tables/plots, etc., make sure to review the [rmd bootcamp](https://uoepsy.github.io//rmd-bootcamp/).

The example write-up sections included as part of the solutions are **not perfect** - they instead should give you a good example of what information you should include and how to structure this. Note that you must **not** copy any of the write-ups included below for future reports - if you do, you will be committing plagiarism, and this type of academic misconduct is taken very seriously by the University. You can find out more [here](https://www.ed.ac.uk/academic-services/students/conduct/academic-misconduct).

### <i class="fa fa-file"></i> Lab Data
You can download the data required for this lab [here](https://uoepsy.github.io/data/caffeinedrink.csv) or read it in via this link https://uoepsy.github.io/data/caffeinedrink.csv

:::

# Study Overview 

> **Research Question** 
>
> Does WPM differ by caffeine treatment condition?  

To investigate if the number of words typed per minute (WPM) differs among caffeine treatment conditions, the researchers conducted an experiment where participants were randomly allocated to one of four treatment conditions. Two of these conditions included non-caffeinated drinks - control (water) and mint tea, and the other two caffeinated drinks - coffee and red bull. 

| Drink           | Caffeine | Temp  | 
|:---------------:|:--------:|:-----:|
| Control (Water) | No       | Cold  | 
| Red Bull        | Yes      | Cold  | 
| Coffee          | Yes      | Hot   |
| Mint Tea        | No       | Hot   |

In addition to the above research question, the researchers were also interested in the following:

> **Comparisons** 
>
> + Whether having some kind of caffeine (i.e., red bull or coffee), rather than no caffeine (i.e., control - water or mint tea), resulted in a difference in average WPM
>
> + Whether there was a difference in average WPM between those with hot drinks (i.e., mint tea / coffee) in comparison to those with cold drinks (control - water / red bull)

`r optbegin("Caffeine data codebook", olabel=FALSE,toggle=params$TOGGLE)`  

__Data Dictionary__

The data in `caffeinedrink.csv` contained two attributes collected from $n=40$ participants:

```{r echo=FALSE, message=FALSE, warning=FALSE}
caffeine  <- read_csv("https://uoepsy.github.io/data/caffeinedrink.csv")
tibble(
variable = names(caffeine),
description = c("which one of four treatment conditions the participant was assigned - control (water), coffee, mint_tea, or red_bull", "average number of words typed per minute")
) %>% gt::gt()
```

__Preview__

The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/caffeinedrink.csv') %>%  head %>% gt::gt()
```

`r optend()`

<div class="divider div-transparent div-dot"></div>

# Setup

`r qbegin("Setup", qlabel = FALSE)`  

1. Create a new RMarkdown file
2. Load the required package(s)
3. Read the caffeinedrink dataset into R, assigning it to an object named `caffeine` 
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r message=FALSE}
#load packages
library(tidyverse)
library(psych)
library(kableExtra)
library(emmeans)

#read in data
caffeine <- read_csv("https://uoepsy.github.io/data/caffeinedrink.csv")
```

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Exercises 

## Study & Analysis Plan Overview 

`r qbegin(1)`

Examine the dataset, and perform any necessary and appropriate data management steps.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint
- Convert categorical variables to factors
- Label appropriately factors to aid with your model interpretations if required
- If needed, provide better variable names

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Let's have a look at the data to see what we're working with:

```{r}
#first look at dataset structure
str(caffeine)

#now lets look at top 6 rows (or the head) of the dataset
head(caffeine)

#check for NAs - there are none - all FALSE
table(is.na(caffeine))
```

Let's start with the 'treatment' condition variable. This should be coded as factor (`<fctr>`), but can see from above it is currently coded as a character (`<chr>`). Let's correct this.

```{r}
#Code treatment as a factor
caffeine <- caffeine %>%
  mutate(treatment = as_factor(treatment))
```

Next, let's look at the 'wpm' variable. Here we want to check for impossible values - i.e., cannot have a negative WPM.

```{r}
# all looks ok - min and max both positive values
describe(caffeine$wpm)
```

::: {.callout-important icon=false appearance="minimal"}

All participant data was complete (no missing values), with WPM scores within possible ranges. Treatment was coded as a factor with four levels (control (water), coffee, red bull, and mint tea).

:::

`r solend()`

<br>

`r qbegin(2)`

Choose an appropriate reference level for the Treatment condition.

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The Treatment factor has a group coded 'Control (Water)' which lends itself naturally to be the reference category. 

```{r}
#set 'Control' caffeine treatment condition as our reference group 
caffeine$treatment <- relevel(caffeine$treatment, "control")

#check levels - control should be first in the list
levels(caffeine$treatment)
```

`r solend()`

<br>

`r qbegin(3)`

Provide a brief overview of the study design and data, before detailing your analysis plan to address the research question.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint 

- Give the reader some background on the context of the study
- State what type of analysis you will conduct in order to address the research question
- Specify the model to be fitted to address the research question (note that you will need to specify the reference level of your categorical variable)
- Specify your chosen significance ($\alpha$) level
- State your hypotheses

Much of the information required can be found in the [Study Overview] codebook.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The `caffeine` dataset contained information on 40 hypothetical participants who took part in an experiment examining whether the number of words typed per minute (WPM) differed among caffeine treatment conditions. Using a between-subjects design, the researchers collected information on participants' WPM (average number of words typed per minute), and which one of four treatment conditions they were randomly assigned to (control (water), coffee, mint_tea, or red_bull).

Boxplots will be used to visualise the associations among WPM and caffeine treatment conditions. To address the research question of whether WPM differs by caffeine treatment condition, we first need to define the dummy variables for Treatment:

$$
Treatment_\text{Coffee} = \begin{cases}  
1 & \text{if Treatment is Coffee} \\  
0 & \text{otherwise}  
\end{cases}  
\quad    
$$

$$
Treatment_\text{Mint Tea} = \begin{cases}  
1 & \text{if Treatment is Mint Tea} \\  
0 & \text{otherwise}  
\end{cases}  
\quad  
$$

$$
Treatment_\text{Red Bull} = \begin{cases}  
1 & \text{if Treatment is Red Bull} \\  
0 & \text{otherwise}  
\\  
\end{cases}  
\quad  
$$

$$
(\text{Control (Water) is base level})  
$$

Based on the above dummy coding, we are going to fit the following regression model:

$$
\text{WPM} = \beta_0 + \beta_1 \cdot \text{Treatment(Coffee)} + \beta_2 \cdot \text{Treatment(Red Bull)} \\   
+ \beta_3 \cdot \text{Treatment(Mint Tea)} + \epsilon
$$

Effects will be considered statistically significant at $\alpha=.05$

Our hypotheses are:

$H_0:$ All $\beta_j = 0$ (for $j = 1, 2, 3$)

There are no differences in WPM based on caffeine treatment conditions. 

$H_1:$ At least one $\beta_j \neq  0$ (for $j = 1, 2, 3$)

There are differences in WPM based on caffeine treatment conditions. 


`r solend()`

<div class="divider div-transparent div-dot"></div>

## Descriptive Statistics & Visualisations

`r qbegin(4)`

Provide a table of descriptive statistics and visualise your data.

Remember to interpret your plot in the context of the study (i.e., comment on any observed differences among treatment groups). 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

1. For your table of descriptive statistics, both the `group_by()` and `summarise()` functions will come in handy here.
2. Recall that when visualising categorical variables, `geom_boxplot()` may be most appropriate to use.
3. Make sure to comment on any observed differences among the sample means of the four treatment conditions.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.panel-tabset}

## Numeric

Descriptive statistics presented in a well formatted table:

```{r message=FALSE, warning=FALSE}
#| label: tbl-caf-descript
#| tbl-cap: Descriptive Statistics
caf_desc <- caffeine %>%
  group_by(treatment) %>%
  summarise(n = n(), 
            Mean = mean(wpm), 
            SD = sd(wpm),
            Minimum = min(wpm),
            Maximum = max(wpm)) %>%
    kable(caption = "Descriptive Statistics", digits = 2) %>%
    kable_styling()

caf_desc
```

## Visual 

Since we have a continuous outcome and a categorical predictor - a boxplot would be most appropriate for visualisations:

```{r}
#| label: fig-caf-desc
#| fig-cap: "Association between Treatment Conditions and WPM"
caf_plt <- ggplot(data = caffeine, aes(x = treatment, y = wpm, fill = treatment)) +
  geom_boxplot() +
  labs(x = 'Treatment Condition', y = 'WPM')

caf_plt
```

:::

::: {.callout-important icon=false appearance="minimal"}

+ From the boxplots, it seems that those in the Red Bull condition, on average, typed the most WPM, whilst those in the Mint Tea condition the fewest. 
+ Overall, the average WPM appears to be lower for those in the non-caffeine conditions (i.e., control - water / mint tea) in comparison to those in the caffeine drinks condition (red bull / coffee).

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Model Fitting & Interpretation 

`r qbegin(5)`

Fit the specified model, and assign it the name "caf_mdl1". 

Interpret your coefficients in the context of the study.

:::{.callout-note}

# How do we specify dummy coding in `R`?

Fortunately, `R` computes the dummy variables for us! Thus, each row in the `summary()` output of the model will correspond to one of the estimated $\beta$'s in the equation above.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#fit model
caf_mdl1 <- lm(wpm ~ treatment, data=caffeine)
```

```{r}
#check model output
summary(caf_mdl1)
```

Let's first map our coefficients and estimates:

| Coefficient | Estimate | Corresponds to |
|:-:|:-:|:-:|
| (Intercept)   |  112.1460 | $\beta_0 = \hat \mu_1$
| treatmentcoffee       | 2.3350  | $\beta_0 + \beta_1 = \hat \mu_2$ |
| treatmentred_bull        | 4.5060  | $\beta_0 + \beta_2 = \hat \mu_3$ |
| treatmentmint_tea       | -1.0550  | $\beta_0 + \beta_3 = \hat \mu_4$ |

::: {.callout-important icon=false appearance="minimal"}

+ The estimate corresponding to (Intercept) contains $\hat \beta_0 = \hat \mu_1 = 112.15$. The estimated average WPM for those in the control condition (water) was approximately 112.15.

+ The next estimate corresponds to `treatmentcoffee` and was $\hat \beta_1 = 2.34$. The difference in mean WPM between `Control` and `Coffee` was estimated to be $2.34$. Thus, $\hat \mu_2 = 112.15 + 2.34 = 114.49$. In other words, people who have had coffee typed approximately 114.49 words, which was 2.34 words per minute more than those who have had water. This difference was statistically significant $(p = .01)$.

+ The estimate corresponding to `treatmentred_bull` was $\hat \beta_2 = 4.51$. This was the estimated difference in mean WPM between `Control` and `Red Bull`, estimated to be $4.51$. Thus, $\hat \mu_3 = 112.15 + 4.51 = 116.66$. In other words, people who had red bull typed approximately 116.66 words, 4.51 words per minute more than those who  had water. This difference was statistically significant $(p < .001)$.

+ The estimate corresponding to `treatmentmint_tea` was $\hat \beta_3 = -1.06$. This was the estimated difference in mean WPM between `Control` and `Mint Tea`, estimated to be $-1.06$. Thus, $\hat \mu_4 = 112.15 + (-1.06) = 111.09$. In other words, people who had mint tea typed approximately 111.09 words, 1.06 words per minute less than those who had water. This difference was not statistically significant $(p = .25)$.

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Planned Comparisons / Contrasts

`r qbegin(6)`

Formally state the two planned comparisons that the researchers were interested in as testable hypotheses. 

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Recall that the researchers were also interested in addressing these two questions:

1. Whether having some kind of caffeine (i.e., red bull or coffee), rather than no caffeine (i.e., control - water or mint tea), resulted in a difference in average WPM
2. Whether there was a difference in average WPM between those with hot drinks (i.e., mint tea / coffee) in comparison to those with cold drinks (control - water / red bull)

We can specify the two hypotheses as follows:

::: {.panel-tabset}

## Caffeine vs No Caffeine

$$
\begin{aligned}
1. \quad H_0 &: \mu_\text{No Caffeine} = \mu_\text{Caffeine} \\
    \quad H_0 &: \frac{1}{2} (\mu_\text{Control} + \mu_\text{Mint Tea}) = \frac{1}{2} (\mu_\text{Coffee} + \mu_\text{Red Bull}) \\
\\
    \quad H_1 &: \mu_\text{No Caffeine} \neq \mu_\text{Caffeine} \\
    \quad H_1 &: \frac{1}{2} (\mu_\text{Control} + \mu_\text{Mint Tea}) \neq \frac{1}{2} (\mu_\text{Coffee} + \mu_\text{Red Bull}) \\
\\ 
\end{aligned}
$$

## Hot vs Cold

$$
\begin{aligned}
2. \quad H_0 &: \mu_\text{Hot Drink} = \mu_\text{Cold Drink} \\
    \quad H_0 &: \frac{1}{2} (\mu_\text{Coffee} + \mu_\text{Mint Tea}) = \frac{1}{2} (\mu_\text{Control} + \mu_\text{Red Bull})
\\
\\
    \quad H_1 &: \mu_\text{Hot Drink} \neq \mu_\text{Cold Drink} \\
    \quad H_1 &: \frac{1}{2} (\mu_\text{Coffee} + \mu_\text{Mint Tea}) \neq \frac{1}{2} (\mu_\text{Control} + \mu_\text{Red Bull})
    \\ 
\end{aligned}
$$

:::

`r solend()`

<br> 

`r qbegin(7)`

After checking the levels of the factor `treatment`, use `emmeans()` to obtain the estimated treatment means and uncertainties for your factor. 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

Use `plot()` to visualise this.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
levels(caffeine$treatment)
```

Use the `emmeans()` to get the estimated means of our groups:
```{r}
treatment_mean <- emmeans(caf_mdl1, ~ treatment)
treatment_mean
plot(treatment_mean)
```

`r solend()`

<br>

`r qbegin(8)`

Specify the coefficients of the comparisons and run the contrast analysis, obtaining 95% confidence intervals. 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

Ordering matters here - look again at the output of `levels(caffeine$treatment)` as this will help you when assigning your weights. 

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

As shown above via `levels()`, the ordering of the treatment factor is:

1. Control (no caffeine / cold drink)
2. Coffee (caffeine / hot drink)
3. Red Bull (caffeine / cold drink)
4. Mint Tea (no caffeine / hot drink)

From this ordering, we can specify our weights - based on the hypothesis, lets assign positive values to the no caffeine and hot drink conditions:

```{r}
treatment_comp <- list("No Caffeine - Caffeine" = c(1/2, -1/2, -1/2, 1/2),
             "Hot Drink - Cold Drink" = c(-1/2, 1/2, -1/2, 1/2)
             )
```

Now lets run our contrast analysis and get confidence intervals - to do so we use the `contrast()` function from `emmeans()`:

```{r}
#run contrast analysis
treatment_comp_test <- contrast(treatment_mean, method = treatment_comp)

#examine output
treatment_comp_test

#obtain confidence intervals
confint(treatment_comp_test)
```

`r solend()`

<br>

`r qbegin(9)`

Interpret the results of the contrast analysis in the context of the researchers hypotheses.

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.panel-tabset}

## Hypothesis 1: Caffeine vs No Caffeine

We performed a test against $H_0: \frac{1}{2}(\mu_1 + \mu_4) - \frac{1}{2}(\mu_2 + \mu_3) = 0$. At the 5\% significance level, there was evidence that the mean WPM for those who were in the no caffeine condition was significantly different from those in a caffeine condition $(t(36) = -6.17, p < .001, \text{two-sided})$, and this difference was estimated to be -3.95. We are 95\% confident that those who consumed no caffeine typed, on average, between 2.7 and 5.3 words less per minute than those who consumed some form of caffeine $CI_{95}[-5.25, -2.65]$.

## Hypothesis 2: Hot vs Cold

We performed a test against $H_0: \frac{1}{2}(\mu_2 + \mu_4) - \frac{1}{2}(\mu_1 + \mu_3) = 0$. At the 5\% significance level, there was evidence that the average WPM for those in the hot drink condition significantly differed from those in the cold drink condition $(t(36) = -2.52, p = .02, \text{two-sided})$, and this difference was estimated to be -1.61. We are 95\% confident that those who consumed a hot drink typed, on average, between 0.3 and 2.9 words less per minute than those who consumed a cold drink $CI_{95}[-2.91, -0.32]$.

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Study Design

`r qbegin(10)`

For each of the below experiment descriptions, note (1) the design, (2) number of variables of interest, (3) levels of categorical variables, (4) what you think the reference group should be and why.

::: {.panel-tabset}

## Experiment 1

A group of researchers were interested in whether sleep deprivation influenced reaction time. They hypothesised that sleep deprived individuals would have slower reaction times than non-sleep deprived individuals. 

To test this, they recruited 60 participants who were matched on a number of demographic variables including age and sex. One member of each pair (e.g., female aged 18) was placed into a different sleep condition - 'Sleep Deprived' (4 hours per night) or 'Non-Sleep Deprived' (8 hours per night).

## Experiment 2

A group of researchers were interested in replicating an experiment testing the Stroop Effect. 

They recruited 50 participants who took part in Task A (word colour and meaning are congruent) and Task B (word colour and meaning are incongruent) where they were asked to name the color of the ink instead of reading the word. The order of presentation was counterbalanced across participants. The researchers hypothesised that participants would take significantly more time ('response time' measured in seconds) to complete Task B than Task A.

You can test yourself here for fun: [Stroop Task](https://www.psytoolkit.org/lessons/experiment_stroop.html)

## Experiment 3

A group of researchers wanted to test a hypothesised theory according to which patients with amnesia will have a deficit in explicit memory but not implicit memory. Huntingtons patients, on the other hand, will display the opposite: they will have no deficit in explicit memory, but will have a deficit in implicit memory.

To test this, researchers designed a study that included two variables: 'Diagnosis' (Amnesic, Huntingtons, Control) and 'Task' (Grammar, Classification, Recognition) where participants were randomly assigned to a Task condition. The first two tasks (Grammar and Classification) are known to reflect implicit memory processes, whereas the Recognition task is known to reflect explicit memory processes.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.panel-tabset}

## Experiment 1

1. Design = Between-person: Matched pairs
2. No of variables of interest = 2 - Sleep Condition and Reaction Time
3. Levels of variables = Sleep Condition has 2 levels - Sleep Deprived and Non-Sleep Deprived; Reaction Time is a continuous measure, so has no associated levels
4. Reference Group = Sleep Condition - Non-Sleep Deprived because the research question stated that the researchers were interested in how the sleep deprived group *differed from* the non-sleep deprived group. 

## Experiment 2

1. Design = Within-person: repeated measures
2. No of variables of interest = 2 - Task and Response Time 
3. Levels of variables = Task has 2 levels - A and B; Response Time is a continuous measure, so has no associated levels
4. Reference Group = Task - A because the research question stated that the researchers were interested in how response time in Task B *differed from* the response time in Task A. 

## Experiment 3

1. Design = Between-person: 3×3 factorial design 
2. No of variables of interest = 2 - Diagnosis and Task
3. Levels of variables = Diagnosis has 3 levels - Amnesic, Huntingtons, and Control; Task has 3 levels - Grammar, Classification, and Recognition
4. Reference Groups = Diagnosis - Control; Task - Recognition. We have chosen Control since the other two groups have some form of cognitive impairment; and the Recognition task since it measures explicit memory whilst the other two task types implicit. 

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Compile Report

`r qbegin("Compile Report", qlabel = FALSE)`  

Knit your report to PDF, and check over your work. To do so, you should make sure:

- Only the output you want your reader to see is visible (e.g., do you want to hide your code?)
- Check that the **tinytex** package is installed
- Ensure that the ‘yaml’ (bit at the very top of your document) looks something like this:

```{}
---
title: "this is my report title"
author: "B1234506"
date: "07/09/2024"
output: bookdown::pdf_document2
---
```

:::{.callout-tip appearance="simple" collapse="true"}
# What to do if you cannot knit to PDF
If you are having issues knitting directly to PDF, try the following:  

- Knit to HTML file  
- Open your HTML in a web-browser (e.g. Chrome, Firefox)  
- Print to PDF (Ctrl+P, then choose to save to PDF)  
- Open file to check formatting
:::

:::{.callout-tip appearance="simple" collapse="true"}
# Hiding Code and/or Output

:::{.panel-tabset}
## Hiding R Code

To not show the code of an R code chunk, and only show the output, write:

````
```{{r, echo=FALSE}}
# code goes here
```
````

## Hiding R Output

To show the code of an R code chunk, but hide the output, write:

````
```{{r, results='hide'}}
# code goes here
```
````

## Hiding R Code and Output

To hide both code and output of an R code chunk, write:

````
```{{r, include=FALSE}}
# code goes here
```
````
:::

:::

:::{.callout-tip appearance="simple" collapse="true"}

### Tinytex
You must make sure you have **tinytex** installed in R so that you can “Knit” your Rmd document to a PDF file:

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

:::

`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`

You should end up with a PDF file. If you have followed the above instructions and still have issues with knitting, speak with a Tutor. 

`r solend()`
