---
title: "Interactions: Numeric * Numeric"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('assets/setup.R')

set.seed(953)

library(tidyverse)
```


:::green
Be sure to check the [**solutions to last week's exercises**](1_07_int_nn.html).  
You can still ask any questions about previous weeks' materials if things aren't clear!
:::


:::lo
**LEARNING OBJECTIVES**

1. Understand the concept of an interaction.
2. Interpret the meaning of an numeric $\times$ numeric interaction.  
3. Understand the principle of marginality and why this impacts modelling choices with interactions.
4. Visualize and probe interactions.

:::


## Data Codebook

__Description__  

A group of researchers wants to test an hypothesised theory according to which patients with amnesia will have a deficit in explicit memory but not on implicit memory. Huntingtons patients, on the other hand, will be just the opposite: they will have no deficit in explicit memory, but will have a deficit in implicit memory.

The researchers designed a study yielding a 3 by 3 factorial design to test this theory. The first factor, "Diagnosis", classifies the three types of individuals:

- 1 denotes amnesic patients;
- 2 denotes Huntingtons patients; and
- 3 denotes a control group of individuals with no known neurological disorder.

The second factor, "Task", tells us to which of three tasks each study participant was randomly assigned to:

- 1 = artificial grammar task, which consists of classifying letter sequences as either following or not following grammatical rules;
- 2 = classification learning task, which consists of classifying hypothetical patients as either having or not having a certain disease based on symptoms probabilistically related to the disease; and
- 3 = recognition memory task, which consists of recognising particular stimuli as stimuli that have previously been presented during the task.

Keep in mind that each person has been randomly assigned to one of the three tasks, so there are five observations per cell of the design.^[Some researchers may point out that a design where each person was assessed on all three tasks might have been more efficient. However, the task factor in such design would then be within-subjects, meaning that the scores corresponding to the same person would be correlated. To analyse such design we will need a different method which is discussed next year!]

The tasks chosen by the researchers have been picked to map onto the theoretical differences between the three types of research participants. The first two tasks (grammar and classification) are known to reflect implicit memory processes, whereas the recognition task is known to reflect explicit memory processes. If the theory is correct, we would expect to see relatively higher scores on the first two tasks for the amnesiac group but relatively higher scores on the third task for the Huntington group.

__Download link__  

https://uoepsy.github.io/data/cognitive_experiment.csv


__Preview__  

The first ten rows of the data are:

```{r echo=FALSE}
library(tidyverse)

data <- read_csv("https://uoepsy.github.io/data/cognitive_experiment.csv")

DT::datatable(head(data, 10), options = list(lengthChange = FALSE))
```



## Exercises

`r qbegin(1)`
Read the data into R, and perform all the appropriate
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`



<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
