---
title: "Interactions: Categorical * Categorical"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('assets/setup.R')

set.seed(953)

library(tidyverse)
```


:::green
Be sure to check the [**solutions to last week's exercises**](1_07_int_nn.html).  
You can still ask any questions about previous weeks' materials if things aren't clear!
:::


:::lo
**LEARNING OBJECTIVES**

1. Understand the concept of an interaction.
2. Interpret the meaning of an numeric $\times$ numeric interaction.  
3. Understand the principle of marginality and why this impacts modelling choices with interactions.
4. Visualize and probe interactions.

:::


## Research question

A group of researchers wants to test an hypothesised theory according to which the difference in performance between explicit and implicit memory tasks will be greatest for Huntington patients in comparison to controls.  
On the other hand, the difference in performance between explicit and implicit memory tasks will not significantly differ between patients with amnesia in comparison to controls.

## Data

#### Description
The researchers designed a study yielding a 3 by 2 factorial design to test this theory. The first factor, "Diagnosis", classifies the three types of individuals:

- 1 denotes amnesic patients;
- 2 denotes Huntingtons patients; and
- 3 denotes a control group of individuals with no known neurological disorder.

The second factor, "Task", tells us to which of two tasks each study participant was randomly assigned to:

- 1 = grammar task, which consists of classifying letter sequences as either following or not following grammatical rules; and
- 2 = recognition task, which consists of recognising particular stimuli as stimuli that have previously been presented during the task.

Keep in mind that each person has been randomly assigned to one of the two tasks, so there are five observations per cell of the design.^[Some researchers may point out that a design where each person was assessed on both tasks might have been more efficient. However, the task factor in such design would then be within-subjects, meaning that the scores corresponding to the same person would be correlated. To analyse such design we will need a different method which is discussed next year!]

The tasks chosen by the researchers have been picked to map onto the theoretical differences between the three types of research participants. The first task (grammar) is known to reflect implicit memory processes, whereas the recognition task is known to reflect explicit memory processes. If the theory is correct, we would expect to see relatively higher scores on the first task for the amnesiac group but relatively higher scores on the second task for the Huntington group.


#### Download link

https://uoepsy.github.io/data/cognitive_experiment_3_by_2.csv


#### Preview

The first ten rows of the data are:

```{r echo=FALSE}
library(tidyverse)

data <- read_csv("https://uoepsy.github.io/data/cognitive_experiment_3_by_2.csv")

DT::datatable(head(data, 10), options = list(lengthChange = FALSE))
```



## Exercises

`r qbegin(1)`
Read the data into R, and perform all the appropriate data management steps:

- Convert categorical variables to factors
- Label appropriately factors to aid with your model interpretations
- If needed, provide better variable names
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(tidyverse)

cog <- read_csv("https://uoepsy.github.io/data/cognitive_experiment_3_by_2.csv")
head(cog)
```

The columns Diagnosis and Task should be coded into factors with better labels. The function `factor()` can be used by specifying the current levels and what labels each level should map to.

```{r}
cog <- cog %>%
    mutate(
        Diagnosis = factor(Diagnosis, 
                           levels = c(1, 2, 3),
                           labels = c('Amnesic', 'Huntingtons', 'Control')),
        Task = factor(Task, 
                      levels = c(1, 2),
                      labels = c('Grammar', 'Recognition'))
    ) %>%
    rename(
        Score = Y
    )

head(cog)
```

`r solend()`


`r qbegin(2)`
Choose appropriate reference levels for the factors.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
The Diagnosis factor has a group coded "Control" which lends itself naturally to be the reference category.

```{r}
cog$Diagnosis <- relevel(cog$Diagnosis, 'Control')

levels(cog$Diagnosis)
```

There is no natural reference category for the Task factor, so we will leave it unaltered. However, if you are of a different opinion, please note that there is no absolute correct answer. As long as you will interpret the model correctly, you will reach to the same conclusions as someone that has chosen a different baseline category.

`r solend()`


`r qbegin(3)`
Specify a multiple regression model to test the research hypothesis. 
Store the model in an object named `mdl_int`.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Define the dummy variables for Diagnosis:

$$
D_\text{Amnesic} = \begin{cases}
1 & \text{if Diagnosis = Amnesic} \\
0 & \text{otherwise}
\end{cases}
\quad
D_\text{Huntingtons} = \begin{cases}
1 & \text{if Diagnosis = Huntingtons} \\
0 & \text{otherwise}
\end{cases}
$$

Define the dummy variable for Task:

$$
T_\text{Recognition} = \begin{cases}
1 & \text{if Task = Recognition} \\
0 & \text{otherwise}
\end{cases}
$$


The model becomes

$$
\begin{aligned}
Score &= b_0 \\
      &+  b_1 D_\text{Amnesic} + b_2 D_\text{Huntingtons}  \\
      &+  b_3 T_\text{Recognition}  \\
      &+  b_4 (D_\text{Amnesic} * T_\text{Recognition}) + b_5 (D_\text{Huntingtons} * T_\text{Recognition})  \\
      &+ \epsilon
\end{aligned}
$$


Fortunately, R computes the dummy variables for us! Each row in the summary of the model will correspond to one of the $b$'s in the equation above.

```{r}
mdl_int <- lm(Score ~ Diagnosis * Task, data = cog)
summary(mdl_int)
```

`r solend()`



`r qbegin(4)`
Create a table of group means, and map each coefficient to the group means.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
means <- cog %>%
    group_by(Diagnosis, Task) %>% 
    summarise(M = mean(Score))
means
```

- $b_0$ = 80  
  = Mean(Control, Grammar)

- $b_1$ = -20  
  = Mean(Amnesic, Grammar) - Mean(Control, Grammar)  
  = 60 - 80

- $b_2$ = -40  
  = Mean(Huntingtons, Grammar) - Mean(Control, Grammar)  
  = 40 - 80
  
- $b_3$ = 15  
  = Mean(Control, Recognition) - Mean(Control, Grammar)  
  = 95 - 80

- $b_4$ = -10  
  = [Mean(Amnesic, Recognition) - Mean(Amnesic, Grammar)] -  
  [Mean(Control, Recognition) - Mean(Control, Grammar)]  
  = [65 - 60] - [95 - 80] = 5 - 15 = -10

- $b_5$ = 40  
  = [Mean(Huntingtons, Recognition) - Mean(Huntingtons, Grammar)] -  
  [Mean(Control, Recognition) - Mean(Control, Grammar)]  
  = [95 - 40] - [95 - 80] = 55 - 15 = 40

`r solend()`



`r qbegin(5)`
Load the sjPlot package and try using the function plot_model() to investigate the interactions.

**Hint**

Remember from last week that the default behaviour of plot_model() is to plot the parameter estimates and their confidence intervals. This is where type = "est" will come handy.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(sjPlot)

plot_model(mdl_int, type = "int")
```


The red dots ("Grammar") represent the baseline for Task, while "Control" is the baseline for Diagnosis.

If you imagine connecting the dots of the same color with a line, you can see that the two virtual lines are not parallel, suggesting the presence of an interaction. The difference in score between recognition and grammar tasks for Huntingtons patients (consider the vertical difference) is larger than the difference in score between recognition and grammar tasks for the Control patients. If those vertical differences were the same, there would be no interaction.


An alternative package is `emmeans`. Install it first by typing `install.packages("emmeans")`. You need to use the function `emmip` which stands for EMMeans Interaction Plot. You must provide as first argument the fitted model, then a special formula saying what to plot on the x axis and what to use to differentiate the colors: `colors ~ x`, and then `CIs = TRUE` tells the function to display uncertainty intervals.


```{r}
library(emmeans)

emmip(mdl_int, Task ~ Diagnosis, CIs = TRUE)
```


`r solend()`




`r qbegin(6)`
Interpretation the model output in the context of the research hypothesis.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
plot_model(mdl_int, type = "int")
```

Compared to controls, amnesiac patients will have a deficit in explicit memory (recognition task), but not on implicit memory (grammar and classification tasks).

Compared to controls, Huntingtons patients will have a deficit in implicit memory  (grammar and classification tasks) but not in explicit memory memory (recognition task).


Let's recall the researchers' hypothesis:

> A group of researchers wants to test an hypothesised theory according to which the difference in performance between explicit and implicit memory tasks will be greatest for Huntington patients in comparison to controls.  
> On the other hand, the difference in performance between explicit and implicit memory tasks will not significantly differ between patients with amnesia in comparison to controls.

```{r message = FALSE, warning = FALSE, info = FALSE}
tab_model(mdl_int)
```


:::int
The difference in scores between the recognition and grammar tasks, respectively measuring explicit and implicit memory, for Huntigton patients in comparison to controls was significant and indicated a difference of $b_5 = 40$ points in explicit vs implicit memory performance: $t(36) = 3.41, p = 0.002$.

The difference in scores between the recognition and grammar tasks, respectively measuring explicit and implicit memory, for amnesiac patients in comparison to controls was estimated to be $b_5 = -10$ points but it is not found to be significantly different from 0: $t(36) = -0.85, p = 0.40$.

This indicates that the researchers' hypothesis that the difference in performance between explicit and implicit memory tasks does not differ significantly between amnesic and control patients, while it does differ significantly between Huntington and control patients.
:::



`r solend()`


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
