---
title: "Writing-up"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')
set.seed(953)
```
  
:::lo
**LEARNING OBJECTIVES**

1. Evaluate a published analysis with respect to the thoroughness of information presented. 
2. Write-up and provide interpretation of a linear model analysis with multiple predictors.

:::
  

# Write-up task  

Here you will be presented with the output from a statistical analysis, and your job will be to write-up and present the results. We're going to use an example analysis using one of the datasets we have worked with on a number of exercises in previous labs concerning personality traits, social comparison, depression, and anxiety.  

:::frame
The aim in writing should be that a reader is able to more or less replicate your analyses **without** referring to your R code. This requires detailing all of the steps you took in conducting the analysis.  
The point of using RMarkdown is that you can pull your results **directly** from the code. If your analysis changes, so does your report!  
:::

`r optbegin("Research question and analysis.", olabel=FALSE)`
> **Research Question & Analysis Output**   
> Previous research has identified an association between an individual's perception of their social rank and symptoms of depression, anxiety and stress. We are interested in the individual differences in this relationship.  

> Specifically: 
> 
> + Controlling for other personality traits, does neuroticism moderate effects of social comparison on symptoms of depression, anxiety and stress?  
>  

```{r}
library(tidyverse) # for all things!
library(psych) # good for descriptive stats
library(kableExtra) #useful for creating nice tables
library(car) # for assumption tests
library(sjPlot) # for plotting models

scs_study <- read_csv("https://uoepsy.github.io/data/scs_study.csv")

# standardise scs score
scs_study <- 
  scs_study %>% 
    mutate(
      zscs = (scs-mean(scs))/sd(scs)
    )
#alternatively, you could do zscs = scale(scs, center = TRUE, scale = TRUE)

# the describe() function is from the psych package, and kable() from kableExtra which is used to make a nice table where the values are rounded to 2 decimal places using digits = 2. 
describe(scs_study) %>% 
    kable(digits = 2)

```


```{r, out.width="90%"}
dass_mdl <- lm(dass ~ 1 + zscs*zn + zo + zc + ze + za, data = scs_study)
par(mfrow=c(2,2))
plot(dass_mdl)
# 35 seems to be a very influential point, lets remove it and re-run the model
```

```{r, out.width="90%"}
dass_mdl2 <- lm(dass ~ 1 + zscs*zn + zo + zc + ze + za, data = scs_study[-35, ])
par(mfrow=c(2,2))
plot(dass_mdl2)
par(mfrow=c(1,1))
# looks better, now lets check our assumptions individually by visual inspection
```

```{r, out.width="90%"}
# SIMPLIFY PLOT STRUCTURE: WHICH = 1 SHOWS x, WHICH = 2 SHOWS Y.

#linearity - remember we cannot use crPlots for interactions
plot(dass_mdl2, which=1)

# normality - equivalent to plot(dass_mdl2, which = 2)
qqnorm(resid(dass_mdl2))

# equal variances(Homoscedasticity) - 
residualPlots(dass_mdl2)


# independence
par(mfrow = c(1,2)) 
plot(resid(dass_mdl2))
plot(fitted(dass_mdl2), resid(dass_mdl2))
par(mfrow = c(1,1)) 

# multicollinearity
vif(dass_mdl2)

```

```{r}
#model output
summary(dass_mdl2)
```

`r optend()`

## Analysis Strategy
  
Your analysis strategy will contain a number of different elements, informed by your description above. Remember, your analysis strategy should **not** contain any results. You may wish to include the following sections:  

-  Brief data and design description:
     - Give the reader some background on the context of your write-up. For example, you may wish to describe the data source, data collection strategy, study design, number of observational units, and specify/describe the variables of interest, etc. 
-  Specify variables of interest in relation to the research question:
     -  Describe any data cleaning and/or recoding.
     -  How are the variables scored? Are there any factors - if so, what is the order of the levels.
     -  \* Include descriptive statistics table. 
     -  \* Visualise the variables of interest, and relationships between variables.
     -  \* Report covariances/correlations.

-  Data management:  
     -  Are there any observations that have been excluded based on pre-defined criteria? How/why, and how many? 
     -  \* Describe any transformations performed to aid your interpretation (i.e., log transform, mean center, standardizing, etc.)

-  Model specification:  
     -  Clearly state your hypotheses and specify your chosen significance level.
     -  What type of statistical analysis do you use to answer the research question? (e.g., t-test, simple linear regression, multiple linear regression)
     -  Specify the model(s) to be fitted to answer your given research question and analysis structure. Clearly specify the response and explanatory variables included in your model(s) and remember to describe the coding of categorical variables (i.e., factors) so the reader is aware of any reference levels. 
     -  Detail the steps that you will undetake to ensure that your model(s) do not violate the required appropriate assumptions.
     -  If applicable, detail any required changes/modifications to the model specification to satisfy assumptions. Consider the following: Was there anything you had to do differently than planned during the analysis? Did the modelling highlight issues in your data? Did you have to do anything (e.g., transform any variables, exclude any observations) in order to meet assumptions?

:::red
Note that the * used on occasion in the above indicates that you may/should in some cases repeat these steps if you decide to make any modifications to your data (e.g., removing outliers, etc.)
:::

As noted at the start of the lab, one of the main benefits of using RMarkdown is the ability to include inline R code in your document. Try to incorporate this in your write up so you can automatically pull the specified values from your code. If you need a reminder on how to do this, see Lesson 4 of the Rmd Bootcamp: https://uoepsy.github.io//rmd-bootcamp/.

`r optbegin("Example Write-Up of Analysis Strategy Section", olabel=FALSE)`

Data was obtained from https://uoepsy.github.io/data/scs_study.csv: a dataset containing information on `r nrow(scs_study)` participants, including Z-scores on the 5 personality traits assessed by the Big-Five Aspects Scale (BFAS) (Openness, Conscientiousness, Extraversion, Agreeableness and Neuroticism). Participants were also assessed on the Social Comparison Scale (SCS), which is an 11-item scale measuring self-perception (relative to others) of social rank, attractiveness and belonging, and the Depression Anxiety and Stress Scale (DASS-21) - a 21 item measure with higher scores indicating higher severity of symptoms. For both of these measures, only total scores are available. Items in the SCS are measured on a 5-point scale, giving minimum and maximum possible scores of 11 and 55 respectively. Items in the DASS-21 are measured on a 4-point scale, meaning that scores can range from a possible 21 to 84. 
  
All participant data was complete (no missing values), with scores on the SCS and the DASS-21 all within possible ranges (see Table \@ref(tab:scsdasstab)). Bivariate correlations show a moderate negative relationship between DASS-21 and SCS scores; a moderate positive relationship between DASS-21 and Neuroticism, and a weak positive correlation between SCS and Neuroticism. Additionally, a strong positive relationship is evident between Extraversion and Agreeableness (see Figure \@ref(fig:splom)). 

```{r scsdasstab, echo = FALSE}
# the kable() function makes tables nice for html:
describe(scs_study %>% select(scs, dass))[,c(2:4,8:9)] %>% round(2) %>%
  knitr::kable(., caption = "SCS and DASS-21 descriptive statistics") %>%
  kableExtra::kable_styling()
```

```{r splom, echo=FALSE, fig.cap="Bivariate scatter plots (below diagonal), histograms (diagonal), and Pearson correlation coefficient (above diagonal), of personality trait measures and scores on the SCS and the DASS-21"}
# scatterplot matrix of dataset without the zscs variable
pairs.panels(scs_study %>% select(-zscs))
```

To investigate whether, when controlling for other personality traits, neuroticism moderates the effect of social comparison on symptoms of depression, anxiety and stress, total scores on the DASS-21 were modelled using multiple linear regression. The Z-scored measures on each of the big-five personality traits were included as predictors, along with scores on the SCS (Z-scored) and its interaction with the measure of Neuroticism. Effects will be considered statistically significant at $\alpha = 0.01$.  

One observation was excluded from the final analysis as it was judged to be too influential on the model (Cook's Distance = `r round(cooks.distance(dass_mdl)[35],2)`). 

The  model was fitted to the remaining 655 observations, and took the form: 
$$
\text{DASS-21} = \beta_0 + \beta_1 \text{O} + \beta_2 \text{C} + \beta_3 \text{E} + \beta_4 \text{A} + \beta_5 \text{N} + \beta_6 \text{SCS} + \beta_7 \text{SCS} \cdot \text{N} + \epsilon  \\
\begin{align}
\text{Where} \\
& \text{O = Openness} \\
& \text{C = Conscientiousness} \\
& \text{E = Extraversion} \\
& \text{A = Agreeableness} \\
& \text{N = Neuroticism} \\
\end{align}
$$

To address the research question of whether neuroticism moderates the effect of social comparison on depression and anxiety, we will consider the hypothesis test that the interaction coefficient is equal to zero, where:  

$H_0: \beta_7 = 0$. The interaction between SCS and Neuroticism is equal to zero.  
$H_1: \beta_7 \neq 0$. The interaction between SCS and Neuroticism is not equal to zero.  


For the final model we visually assessed the following assumptions using diagnostic plots: linearity and independence (via plot of residuals vs fitted values), equal variances (via a scale-location plot), and normality (via a qqplot of the residuals). We will also check if there is any evidence of multicollinearity by checking the VIF values, where values > 5 will be considered to indicate moderate multicollinearity, and values > 10 severe.

`r optend()`

## Results

The results section should follow from your analysis strategy. This is where you would present the results of all aspects of your approach outlined above. In this section, it is useful to include tables and plots to clearly present your findings to your reader. You should also interpret all of the results presented, and remember to make reference to and comment on your assumption and diagnostic checks.

`r optbegin("Example Write-Up of Results Section", olabel=FALSE)`

The model met assumptions of linearity and independence (see top left panel of Figure \@ref(fig:diagplots)), homoscedasticity (see bottom left panel of Figure \@ref(fig:diagplots)), and normality (see top right panel of Figure \@ref(fig:diagplots)) where QQplot shows very little deviation from the diagonal line.

```{r diagplots, echo = FALSE, out.width="90%"}
par(mfrow = c(2,2))
plot(dass_mdl2)
par(mfrow = c(1,1))
```

```{r echo=FALSE}
fres <- summary(dass_mdl2)$fstatistic
fres <- round(fres,2)
```

Full regression results including 95\% Confidence Intervals are shown in Table \@ref(tab:tabmodel). The interaction between social comparison and neuroticism in predicting depression and anxiety is visually presented in Figure \@ref(fig:intplot). The F-test for model utility was significant (F(`r fres[2]`,`r fres[3]`)=`r fres[1]`, p<.001), and the model explained approximately `r round(summary(dass_mdl2)$adj.r.squared * 100, 1)`% of the variability in DASS-21 Scores. 

<br>
<center>
```{r tabmodel, echo=FALSE}
tab_model(dass_mdl2,
          dv.labels = c("DASS-21"),
          pred.labels = c("zscs"="Social Comparison Scale", 
                          "zn"="Neuroticism", 
                          "zo"="Openness", 
                          "zc"="Conscientiousness",
                          "ze"="Extraversion",
                          "za"="Agreeableness",
                          "zscs:zn"="Social Comparison Scale : Neutoricism"),
          title = "(\\#tab:tabmodel) Regression table for DASS-21 model. Outcome variable is raw total score on DASS-21, all predictors are Z-scored")
```
</center>
<br> 
```{r echo=FALSE}
res <- summary(dass_mdl2)$coefficients %>% as.data.frame
res[,1:3]<-round(res[,1:3],2)
res[,5] <- ifelse(res[,4]<.01, "<.01",paste0("= ",round(res[,4],3)))
```

Results showed a significant conditional association between SCS scores (Z-scored) and DASS-21 Scores ($\beta$ = `r res[1,1]`,SE = `r res[1,2]`, p `r res[1,5]`), suggesting that for those at the mean level of neuroticism, scores on the DASS-21 decrease by `r res[1,1]` for every 1 standard deviation increase in SCS scores. A significant conditional association was also evident between Neuroticism (Z-scored) and DASS-21 Scores ($\beta$ = `r res[2,1]`,SE = `r res[2,2]`, p `r res[2,5]`), suggesting that for those who score the mean on the SCS, scores on the DASS-21 increase by `r res[2,1]` for every 1 standard deviation increase in neuroticism. 

Crucially, the association between social comparison and symptoms of depression and anxiety was found to be dependent upon the level of neuroticism, with a greater negative association between the two for those with high levels of neuroticism ($\beta$ = `r res[8,1]`,SE = `r res[8,2]`, p `r res[8,5]`). This interaction is visually presented in Figure \@ref(fig:intplot).

```{r intplot, echo=FALSE, out.width="90%", fig.cap="Predicted DASS-21 score across SCS scores, for +/-1 SD Neuroticism"}
plot_model(dass_mdl2, type="pred", terms = c("zscs","zn [-1,1]")) +
  labs(title="Neuroticism moderating the effect of\nsocial comparison on depression and anxiety", 
       x = "Social Comparison Scale (Z-scored)",
       y = "DASS-21")+
  scale_color_manual("Neuroticism (Z-scored)", labels = c("-1 SD", "+1 SD"),
                     values = c("dodgerblue","tomato1")) + 
  scale_fill_manual("Neuroticism (Z-scored)", labels = c("-1 SD", "+1 SD"),
                     values = c("dodgerblue","tomato1"))
```

<br> 


The results presented here indicate that the association between social comparison and depression and anxiety may depend upon individuals' levels of neuroticism, with perceived social rank perhaps leading to more symptoms of depression and anxiety for highly neurotic individuals. However, it is important to note that we can make no claims on the directions of these associations from these data - it may be that social comparison leads to more depression and anxiety in neurotic individuals, but also consistent is the view that - for these individuals - higher levels of depression leads to a greater reduction in perceived social rank.  

`r optend()`

## Discussion 

In the discussion section, you should summarise the formal results and clearly like these to each of the research questions. The discussion should be relatively brief, and should not include any statistical analysis - instead think of the discussion as a conlusion drawing the analysis together and relating it back to the original question. 

`r optbegin("Example Write-Up of Discussion Section", olabel=FALSE)`

Previous research has identified an association between an individual's perception of their social rank and symptoms of depression, anxiety and stress. We investigated if neuroticism moderated effects of social comparison on symptoms of depression, anxiety, and stress after controlling for other personality traits. 

We reject the null hypothesis that the interaction between SCS and Neuroticism is equal to zero, as our results indicated that the association between social comparison and depression and anxiety may depend upon individuals' levels of neuroticism, with perceived social rank perhaps leading to more symptoms of depression and anxiety for highly neurotic individuals. However, it is important to note that we can make no claims on the directions of these associations from these data - it may be that social comparison leads to more depression and anxiety in neurotic individuals, but also consistent is the view that - for these individuals - higher levels of depression leads to a greater reduction in perceived social rank.

`r optend()`

## Tying it all together  

All the component parts we have just written in the exercises above can be brought together to make a reasonable draft of a statistical report. There is a lot of variability in how to structure the reporting of statistical analyses, for instance you may be using the same model to test a selection of different hypotheses.  
  
The answers contained above are just an *example* of how you can structure and write each section. While we hope it will serve as a useful reference when you are writing your report, but it should **not** be taken as an exemplary template for a report which would score 100%.  

We have also included the RMarkdown file used to create this, which may be useful to see how things such as formatting and using inline R code can be used.  

You can find find the .Rmd file for the draft below at https://uoepsy.github.io/files/mlr_writeup_example.Rmd. 
Why not try downloading and compiling it to see how it works?  

**PLEASE NOTE:** A weird issue with hosting .Rmd files like this means that when you download it, it will remove the top bit of *meta-data* (i.e., the bit which includes author, title etc) meaning it will not compile without a small amount of editing. You can see the .Rmd file more clearly including these first 6 lines [here](https://raw.githubusercontent.com/uoepsy/uoepsy.github.io/master/files/mlr_writeup_example.Rmd).  


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>