---
title: "Interactions I: Num x Cat"
link-citations: TRUE
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')
library(tidyverse)
library(psych)
library(sjPlot)
library(patchwork)
library(kableExtra) 
library(plotly)
library(pander)
set.seed(953)
```

:::lo
### <i class="fa fa-graduation-cap"></i> Learning Objectives
At the end of this lab, you will:

1. Understand the concept of an interaction.
2. Be able to interpret the meaning of a numeric $\times$ categorical interaction.
3. Visualize and probe interactions.

### <i class="fa fa-check-square-o fa-2"></i> What You Need

1. Be up to date with lectures
2. Have completed all labs from Semester 1

### <i class="fab fa-r-project"></i> Required R Packages
Remember to load all packages within a code chunk at the start of your RMarkdown file using `library()`. If you do not have a package and need to install, do so within the console using `install.packages(" ")`. For further guidance on installing/updating packages, see Section C [here](https://uoepsy.github.io/files/install-update-r#update-pkgs). 

For this lab, you will need to load the following package(s):

* **tidyverse** 
* **psych** 
* **patchwork** 
* **sandwich**
* **interactions**

### <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Presenting Results
All results should be presented following [APA guidelines](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf). If you need a reminder on how to hide code, format tables/plots, etc., make sure to review the [rmd bootcamp](https://uoepsy.github.io//rmd-bootcamp/).

### <i class="fa fa-file"></i> Lab Data
You can download the data required for this lab [here](https://uoepsy.github.io/data/wellbeing_rural.csv) or read it in via this link https://uoepsy.github.io/data/wellbeing_rural.csv. 

:::

# Study Overview {#sec-studyview} 

> **Research Question** 
>
> Does the association between wellbeing and the number of social interactions differ between rural and non-rural residents?  

Researchers were specifically interested in how the number of social interactions might influence mental health and wellbeing differently for those living in rural communities compared to those in cities and suburbs. They want to assess whether the effect of social interactions on wellbeing _is moderated by_ (depends upon) whether or not a person lives in a rural area. 

To investigate how the association between the number of social interactions and mental wellbeing might be different for those living in rural communities, the researchers conducted a study, where they collected data from 200 randomly selected residents of the Edinburgh & Lothian postcodes. 

`r optbegin("Wellbeing/Rurality data codebook.", olabel=FALSE, toggle=params$TOGGLE)`  

__Description__

From the Edinburgh & Lothians, 100 city/suburb residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), was used to measure mental health and well-being. 

Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), whether a daily routine is followed (yes/no). For those respondents who had an activity tracker app or smart watch, they were asked to provide their average weekly number of steps.  
  
__Data Dictionary__

The data in `wellbeing_rural.csv` contain seven attributes collected from a random sample of $n=200$ hypothetical residents over Edinburgh & Lothians, and include: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
mwdata  <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
tibble(
variable = names(mwdata),
description = c("Age in years of respondent","Self report estimated number of hours per week spent outdoors ", "Self report estimated number of social interactions per week (both online and in-person)", "Binary 1=Yes/0=No response to the question 'Do you follow a daily routine throughout the week?'", "Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and well-being. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70", "Location of primary residence (City, Suburb, Rural)", "Average weekly number of steps in thousands (as given by activity tracker if available)")
) %>% gt::gt()
```
  
  
__Preview__

The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/wellbeing_rural.csv') %>% head %>% gt::gt()
```
   
`r optend()`

# Setup

`r qbegin("Setup", qlabel = FALSE)`  

1. Create a new RMarkdown file
2. Load the required package(s)
3. Read the wellbeing_rural dataset into R, assigning it to an object named `mwdata`
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r message=FALSE}
#Loading the required package(s)
library(tidyverse)
library(psych)
library(patchwork)
library(sandwich)
library(interactions)

#Reading in data and storing in object named 'mwdata'
mwdata <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
```

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Exercises 

## Study & Analysis Plan Overview 

`r qbegin(1)`

Provide a brief overview of the study design and data, before detailing your analysis plan to address the research question.


:::{.callout-tip appearance="simple" collapse="true"}

### Hint 

- State what type of analysis you will conduct in order to address the research question
- Specify the model to be fitted to address the research question (note that you will need to specify the reference level of your categorical variable)
- Specify your chosen significance ($\alpha$) level
- State your hypotheses

Much of the information required can be found in the [Study Overview] codebook.
:::

:::frame

:::quote
"Except in special circumstances, a model including a product term for interaction between two explanatory variables should also include terms with each of the explanatory variables individually, even though their coefficients may not be significantly different from zero. Following this rule avoids the logical inconsistency of saying that the effect of $X_1$ depends on the level of $X_2$ but that there is no effect of $X_1$."  
--- @Ramsey2012
:::

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The `mwdata` dataset contained information on 200 hypothetical participants who lived in Edinburgh & Lothians area. Using a between-subjects design, the researchers collected information on participants' wellbeing (measured via WEMWBS), outdoor time (hours per week), social interactions (number per week), routine (whether or not one was followed), location of residence (City, Suburb, or Rural), average weekly steps (in thousands), and age (in years).

To address the research question of whether the association between the number of social interactions and wellbeing differs between rural and non-rural residents, we are going to fit the following interaction model, where $y$ = wellbeing; $x_1$ = social interactions; and $x_2$ = whether or not the respondent lives in a rural location (where not rural will be specified as the reference group). 

$$
y = \beta_0 + \beta_1  x_1 + \beta_2  x_2 + \beta_3 (x_1 \cdot x_2) + \epsilon \\ 
\quad \\ \text{where} \quad \epsilon \sim N(0, \sigma) \quad \text{independently}
$$

or 

$$
\begin{split}
\text{Wellbeing} = \beta_0 + \beta_1 \cdot Social~Interactions + \beta_2 \cdot Location_{Rural} \\+ \beta_3 \cdot (Social~Interactions \cdot Location_{Rural}) + \epsilon \\
\end{split}
$$

Effects will be considered statistically significant at $\alpha=.05$

Our hypotheses are:

$H_0: \beta_3 = 0$

The association between wellbeing and social interactions is not moderated by whether or not a person lives in a rural area. 

$H_1: \beta_3 \neq 0$

The association between wellbeing and social interactions is moderated by whether or not a person lives in a rural area. 

`r solend()`

<br>

`r qbegin(2)`

Check coding of variables (e.g., that categorical variables are coded as factors). 

Note that the "location" variable currently has three levels (Rural/Suburb/City). In order to address the research question, we only want two (Rural/Not Rural) locations - you will need to fix this.

As specified in Q1, we want 'not rural' as the reference group, so make sure to specify this.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint
One way to do this would be to use `ifelse()` to define a variable which takes one value ("Rural") if the observation meets from some condition, or another value ("Not Rural") if it does not. Type `?ifelse` in the *console* if you want to see the help function. You can use it to add a new variable either inside `mutate()`, or using `data$new_variable_name <- ifelse(test, x, y)` syntax.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Create a new variable for Rural/Not Rural:
```{r}
#if location is rural, assign name 'rural'. If another value (i.e., city or suburb) assign name 'not rural'.
# In other words, if location = rural assign name rural; if location != rural then assign name not rural.
mwdata <- mwdata %>% 
  mutate(
    isRural = ifelse(location == "rural", "rural", "not rural")
  )
```

Check coding of variables within `mwdata` and ensure `isRural` is a factor with two levels, 'rural' and 'not rural':
```{r}
str(mwdata) #returns overall 'structure' of data. Or could run is.factor() for specific variable of interest
mwdata$isRural <- as_factor(mwdata$isRural)
is.factor(mwdata$isRural) #check that isRural is now a factor
```

```{r}
#specify 'not rural' as reference group
mwdata$isRural <- relevel(mwdata$isRural, 'not rural')
```

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Descriptive Statistics & Visualisations

`r qbegin(3)`

Visualise your data, and interpret your plots. 

In particular:

1. Explore the associations among the variables included in your analysis
2. Produce a visualisation of the association between weekly number of social interactions and well-being, with separate _facets_ for rural vs non-rural respondents **OR** with _different colours_ for each level of the `isRural` variable.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

1. The `pairs.panels()` function from the `psych` package can plot all variables in a dataset against one another. This will save you the time you would have spent creating individual plots, but is only useful for *continuous* variables.  
2. To include facets, within your `ggplot()` argument you will need to specify `+ facet_wrap()` in order to produce facets for each location. It would also be useful to specify `geom_smooth(method="lm")`

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Let's first plot the continuous variables included within our model (note that we could use this for the whole dataset, but we don't want to include irrelevant- / non-continuous variables):

```{r}
mwdata %>% 
  select(wellbeing, social_int) %>%
  pairs.panels()
```

::: {.callout-important icon=false appearance="minimal"}

Wellbeing and social interactions appear to follow unimodal distributions. There was a weak, positive association between wellbeing and social interactions ($r$ = .24). 

::: 

Now lets look at wellbeing scores by location:

```{r}
ggplot(data = mwdata, aes(x = isRural, y = wellbeing)) +
  geom_boxplot() + 
  labs(x = "Location", y = "Wellbeing (WEMWBS Scores)")
```

::: {.callout-important icon=false appearance="minimal"}

Those in rural locations appear to have lower wellbeing scores in comparison to those in non-rural locations.

::: 

Next, lets produce our plots with a facet for rural vs non-rural residents: 

```{r message=FALSE, warning = FALSE}
ggplot(data = mwdata, aes(x = social_int, y = wellbeing)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  facet_wrap(~isRural, labeller = "label_both") + 
  labs(x = "Social Interactions (number per week)", y = "Wellbeing (WEMWBS Scores)")
```

Or instead of facets, we could use different colours for each location (rural vs non-rural):

```{r message=FALSE, warning = FALSE}
ggplot(data = mwdata, aes(x = social_int, y = wellbeing, colour = isRural)) +
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
    scale_colour_discrete(
    name ="Location",
    labels=c("Not Rural", "Rural")) + 
    labs(x = "Social Interactions (number per week)", y = "Wellbeing (WEMWBS Scores)")
```

::: {.callout-important icon=false appearance="minimal"}

Those in non-rural locations appear to have higher wellbeing scores across almost all levels of social interactions. The slopes appear to be different for each location, where the greatest difference in wellbeing scores by location is most visible the highest number of social interactions. This suggests that there may be an interaction. 

::: 

::: {.callout-important icon=false collapse=true}
# How can we tell that there is an interaction?

The lines in the two plots above are not running in *parallel* - this suggests the presence of an interaction. Specifically in our example, the non-parallel lines suggest an interaction effect based on location, as the number of social interactions does not appear to have the same influence on rural and non-rural residents' wellbeing scores. 

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Model Fitting & Interpretation 

`r qbegin(4)`

Fit your model using `lm()`, and assign it as an object with the name "rural_mod".   

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

When fitting a regression model in `R` with two explanatory variables A and B, and their interaction, these three are equivalent:  

+ y ~ A + B + A:B
+ y ~ A + B + A*B
+ y ~ A*B  

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#fit model including interaction between social_int and isRural
rural_mod <- lm(wellbeing ~  social_int * isRural, data = mwdata)

#check model output
summary(rural_mod)
```

`r solend()`

<br>

`r qbegin(5)`  

Look at the parameter estimates from your model, and write a description of what each one corresponds to on the plot shown in @fig-annotate-int (it may help to sketch out the plot yourself and annotate it).  

```{r fig-annotate-int, echo=FALSE, message = FALSE, fig.cap="Multiple regression model: Wellbeing ~ Social Interactions * is Rural<br><small>Note that the dashed lines represent predicted values below the minimum observed number of social interactions, to ensure that zero on the x-axis is visible</small>"}
#| label: fig-annotate-int
#| fig-cap: "Multiple regression model: Wellbeing ~ Social Interactions * is Rural"
nd = expand_grid(social_int=0:13,isRural=c("rural","not rural"))
nd = nd %>% mutate(wellbeing = predict(rural_mod, newdata = .))
sjPlot::plot_model(rural_mod, type="int")+
  scale_fill_manual(NULL, values=c(NA,NA))+xlim(0,28)+
  geom_line(inherit.aes=FALSE,data=nd,aes(x=social_int,col=isRural,y=wellbeing), lty="longdash")+
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 1))
```

`r optbegin("Options", olabel=FALSE, toggle=params$TOGGLE)`
Here are some options to choose from:

+ The point at which the red line cuts the y-axis (where social_int = 0)
+ The point at which the blue line cuts the y-axis (where social_int = 0)
+ The vertical distance from the red to the blue line _at the y-axis_ (where social_int = 0)
+ The vertical distance from the blue to the red line _at the y-axis_ (where social_int = 0)
+ The vertical distance from the blue to the red line _at the center of the plot_
+ The vertical distance from the red to the blue line _at the center of the plot_
+ The slope (vertical increase on the y-axis associated with a 1 unit increase on the x-axis) of the red line
+ The slope (vertical increase on the y-axis associated with a 1 unit increase on the x-axis) of the blue line
+ How the slope of the line changes when you move from the red to the blue line

`r optend()`

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Recall that we can obtain our parameter estimates using various functions such as `summary()`,`coef()`, `coefficients()`, etc. 

```{r}
coefficients(rural_mod)
```

+ $\beta_0$ = `(Intercept)` = `r round(coef(rural_mod)[1],2)`
    - On plot: The point at which the red line cuts the y-axis
    - Interpretation: The intercept, or predicted wellbeing score when the number of social interactions per week is 0, and when location is not rural.

+ $\beta_1$ = `social_int` = `r round(coef(rural_mod)[2],2)`
    - On plot: The slope (vertical increase on the y-axis associated with a 1 unit increase on the x-axis) of the red line.
    - Interpretation: The simple slope of social interactions (number per week) for location reference group (not rural).

+ $\beta_2$ = `isRuralrural` = `r round(coef(rural_mod)[3],2)`
    - On plot: The vertical distance from the red to the blue line _at the y-axis_ (where social_int = 0).  
    - Interpretation: The simple effect of location (or the difference in wellbeing scores between rural and non rural residents) when number of social interactions is 0. 

+ $\beta_3$ `social_int:isRuralrural` = `r round(coef(rural_mod)[4],2)`
    - On plot: How the slope of the line changes when you move from the red to the blue line. 
    - Interpretation: The interaction between social interactions (number per week) and location (rural/not rural) - the difference in the simple slopes of social interactions for rural vs non-rural residents. 

`r solend()`

<br>

`r qbegin(6)`

Mean center the continuous IV(s), and re-run your model with mean centered variable(s).

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Create mean centered variable for 'social_int':

```{r}
mwdata <-
 mwdata %>%
  mutate(
   mc_social_int = social_int - mean(social_int)
    )
```

Re-run model:

```{r}
#fit model including interaction between social_int and isRural
rural_mod1 <- lm(wellbeing ~  mc_social_int * isRural, data = mwdata)

#check model output
summary(rural_mod1)
```

`r solend()`

<br>

`r qbegin(7)`

Note any differences between the `summary()` output between the "rural_mod" and "rural_mod1" models. Pay particular attention to your coefficients and their significance values. Why do you think these differences have been observed?

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

These plots illustrate the difference between the "rural_mod" and "rural_mod1" models. 

```{r echo = FALSE, message=FALSE, warning=FALSE}
#| label: fig-mc_difference
#| fig-cap: "Difference when social interactions is not vs is mean centered.<br><small>Note that the lines without SE intervals on the left plot represent \npredicted values below the minimum observed number of social \ninteractions, to ensure that zero on the x-axis is visible</small>"
nd = expand_grid(social_int=-1:13,isRural=c("rural","not rural"))
nd = nd %>% mutate(wellbeing = predict(rural_mod, newdata = .))

p1 <- ggplot(data = mwdata, aes(x = social_int, y = wellbeing, colour = isRural)) +
    geom_point() +
    geom_smooth(method="lm", se=TRUE) +
    geom_smooth(method="lm", se=FALSE, fullrange=TRUE) +
    geom_point(aes(x=0, y=30.9986), colour="#F8766D", size = 7) +
    geom_point(aes(x=0, y=32.3852), colour="#00BFC4", size = 7) +
    geom_vline(xintercept = 0, colour="black", linetype = "longdash", size = 0.5) +
    scale_fill_manual(NULL, values=c(NA,NA))+xlim(-5,30)+
    geom_line(inherit.aes=FALSE,data=nd,aes(x=social_int,col=isRural,y=wellbeing), lty="solid", lwd=1) + 
    scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 1)) +
    labs(x = "Social Interactions \n(n per week)", y = "Wellbeing (WEMWBS Scores)", caption = "") +
    theme(plot.caption = element_text(size = 8, hjust = 0, face = "italic"), legend.position="none") +
    scale_colour_discrete(
    name ="Location",
    labels=c("Not Rural", "Rural"))

p2 <- ggplot(data = mwdata, aes(x = mc_social_int, y = wellbeing, colour = isRural)) +
    geom_point() +
    geom_smooth(method="lm", se=TRUE) +
    geom_smooth(method="lm", se=FALSE, fullrange=TRUE) +
    geom_point(aes(x=0, y=38.8263), colour="#F8766D", size = 7) +
    geom_point(aes(x=0, y=34), colour="#00BFC4", size = 7) +
    geom_vline(xintercept = 0, colour="black", linetype = "longdash", size = 0.5) +
    scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 1)) +
    labs(x = "Mean Centred Social \nInteractions (n per week)", y = "Wellbeing (WEMWBS Scores)") +
    scale_colour_discrete(
    name ="Location",
    labels=c("Not Rural", "Rural"))+
    guides(fill="none")


(p1 | p2) + plot_layout(guides = "collect")
```

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


Recall that when there is an interaction A$\times$B, the coefficients A and B are no longer *main effects*. Instead, they are *conditional effects* upon the other being zero. 

In our "rural_mod", the `isRural` coefficient is the difference in rural vs non-rural when social interactions is 0. In our "rural_mod1", this difference is when social interactions is the mean (12.06). 

Whilst the difference in rural vs non-rural may not be significantly different when social interactions is zero, there is a significant difference at the average number of social interactions (as you can see from the plot below - note that this is the same plot as in the hint).

```{r echo = FALSE, message=FALSE, warning=FALSE}
#| label: fig-mc_difference-2
#| fig-cap: "Difference when social interactions is not vs is mean centered" 
(p1 | p2) + plot_layout(guides = "collect")
```

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Visualise Interaction Model

`r qbegin(8)`

Using the `probe_interaction()` function from the interactions package, visualise the interaction effects from your model.

Try to summarise the interaction effects in a short and concise sentence. 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

Make sure to give your plot informative titles/labels. You, for example, likely want to give your plot:

- a clear and concise title (specify `main.title = `)
- axis labels with units or scale included (specify `x.label = ` and `y.label = `)
- a legend title (specify `legend.main = `)

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r warning=FALSE, message=FALSE}
plt_rural_mod <- probe_interaction(model = rural_mod1, 
                  pred = mc_social_int, 
                  modx = isRural, 
                  interval = T,
                  main.title = "Predicted Wellbeing Scores across \n Social Interactions by Location",
                  x.label = "Number of Social Interactions per Week (Mean Centred)",
                  y.label = "Wellbeing (WEMWBS Scores)",
                  legend.main = "Location")
```

Let's look at our plot:

```{r}
#| label: fig-int
#| fig-cap: "Predicted Wellbeing Scores across Social Interactions by Location"

plt_rural_mod$interactplot
```

::: {.callout-important icon=false appearance="minimal"}

This suggested that for individuals living in non-rural locations, wellbeing scores increased at a steeper rate across the number of social interactions in comparison to those in rural locations.

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Writing Up & Presenting Results

`r qbegin(9)`

Provide key model results in a formatted table.


`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| label: tbl-rural-mc-modresults
#| tbl-cap: Regression Table for Wellbeing Model
#create table for results
tab_model(rural_mod1,
          dv.labels = "Wellbeing (WEMWBS Scores)",
          pred.labels = c("mc_social_int" = "Social Interactions (number per week; mean centred)",
                          "isRuralrural" = "Location - Rural",
                          "mc_social_int:isRuralrural" = "Social Interactions (number per week; mean centred) * Location - Rural"),
          title = "Regression Table for Wellbeing Model")
```

`r solend()`

<br>

`r qbegin(10)`

Interpret your results in the context of the research question and report your model in full.

Make reference to the interaction plot and regression table.

:::frame

:::quote
"The best method of communicating findings about the presence of a significant interaction may be to present a table or graph of the estimated means at various combinations of the interacting variables."    
--- @Ramsey2012  
:::

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.callout-important icon=false appearance="minimal"}

Full regression results including 95% Confidence Intervals are shown in @tbl-rural-mc-modresults. The $F$-test for model utility was significant $(F(3,196) = 27.49, p<.001)$, and the model explained approximately 28.54% of the variability in wellbeing scores.

There was a significant conditional association between wellbeing (WEMWBS Scores) and social interactions $(\beta = 0.65, SE = 0.12, p < .001)$, which suggested that for those living in non-rural locations, wellbeing scores increased by 0.65 for every additional social interaction per week. A significant conditional association was also evident between wellbeing and location $(\beta = -4.86, SE = 0.65, p < .001)$, which suggested that for those with the average number of social interactions per week $(M = 12.06)$, wellbeing scores were 4.86 points lower for those in rural areas in comparison to those in non-rural.

The association between wellbeing (WEMWBS Scores) and social interactions was found to be dependent upon location (rural/non-rural), where the slope was less steep for those in rural locations $(\beta = -0.52, SE = 0.16, p = .002)$. This interaction is visually presented in @fig-int. The effect of every additional social interaction per week on wellbeing (WEMWBS Scores) was 0.52 less for those living in rural locations in comparison to those in non-rural locations. Therefore, we have evidence to reject the null hypothesis (that the association between wellbeing and social interactions is not moderated by whether or not a person lives in a rural area).

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Compile Report

`r qbegin("Compile Report", qlabel = FALSE)`  

Knit your report to PDF, and check over your work. To do so, you should:

- Make sure the **tinytex** package is installed
- Makes sure the ‘yaml’ (bit at the very top of your document) looks something like this:

```{}
---
title: "this is my report title"
author: "B1234506"
date: "07/09/2024"
output: bookdown::pdf_document2
---
```

::: {.callout-important icon=false collapse=true}
# What to do if you cannot knit to PDF
If you are having issues knitting directly to PDF, try the following:  

- Knit to HTML file  
- Open your HTML in a web-browser (e.g. Chrome, Firefox)  
- Print to PDF (Ctrl+P, then choose to save to PDF)  
- Open file to check formatting
:::

:::{.callout-tip appearance="simple" collapse="true"}

### Hint
You must make sure you have **tinytex** installed in R so that you can “Knit” your Rmd document to a PDF file:

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

:::

`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`

You should end up with a PDF file. If you have followed the above instructions and still have issues with knitting, speak with a tutor. 

`r solend()`

