---
title: "Interactions II: Num x Num"
link-citations: TRUE
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')
library(tidyverse)
library(plotly)
library(patchwork)
library(pander)
set.seed(953)
```

:::lo
### <i class="fa fa-graduation-cap"></i> Learning Objectives
At the end of this lab, you will:

1. Understand the concept of an interaction
2. Be able to interpret the meaning of a numeric $\times$ numeric interaction
3. Understand the principle of marginality and why this impacts modelling choices with interactions
4. Be able to visualize and probe interactions

### <i class="fa fa-check-square-o fa-2"></i> What You Need

1. Be up to date with lectures
2. Have completed previous lab exercises from [Week 1](https://uoepsy.github.io/dapr2/2425/labs/2_01_int1_nc.html)  

### <i class="fab fa-r-project"></i> Required R Packages
Remember to load all packages within a code chunk at the start of your RMarkdown file using `library()`. If you do not have a package and need to install, do so within the console using `install.packages(" ")`. For further guidance on installing/updating packages, see Section C [here](https://uoepsy.github.io/files/install-update-r#update-pkgs).  

For this lab, you will need to load the following package(s):

* **tidyverse** 
* **psych** 
* **sjPlot**
* **kableExtra**
* **sandwich**
* **interactions**

### <i class="fa fa-pencil-square-o" aria-hidden="true"></i> Presenting Results
All results should be presented following [APA guidelines](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf).If you need a reminder on how to hide code, format tables/plots, etc., make sure to review the [rmd bootcamp](https://uoepsy.github.io/scs/rmd-bootcamp/).

The example write-up sections included as part of the solutions are **not perfect** - they instead should give you a good example of what information you should include and how to structure this. Note that you must **not** copy any of the write-ups included below for future reports - if you do, you will be committing plagiarism, and this type of academic misconduct is taken very seriously by the University. You can find out more [here](https://www.ed.ac.uk/academic-services/students/conduct/academic-misconduct).

### <i class="fa fa-file"></i> Lab Data
You can download the data required for this lab [here](https://uoepsy.github.io/data/scs_study.csv) or read it in via this link https://uoepsy.github.io/data/scs_study.csv

:::

# Study Overview {#sec-studyview} 

> **Research Question** 
>
> Does the effect of social comparison on symptoms of depression, anxiety and stress vary depending on level of Neuroticism?

`r optbegin("Social Comparison Study data codebook", olabel=FALSE,toggle=params$TOGGLE)`  

__Description__

Previous research has identified an association between an individual's perception of their social rank and symptoms of depression, anxiety and stress. We are interested in the individual differences in this association. 

Before we think about fitting our model, it is important that we understand the data available - how many participants? What type (or class) of data will we be working with? What was the measurement scale? How were scales scored? Look at the social comparison study data codebook below. 
  
__Data Dictionary__

The data in `scs_study.csv` contain seven attributes collected from a random sample of $n=656$ participants, and include: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
scs_study <- read_csv("https://uoepsy.github.io/data/scs_study.csv")
tibble(
variable = names(scs_study),
description = c("Openness (Z-scored), measured on the Big-5 Aspects Scale (BFAS)","Conscientiousness (Z-scored), measured on the Big-5 Aspects Scale (BFAS)", "Extraversion (Z-scored), measured on the Big-5 Aspects Scale (BFAS)", "Agreeableness (Z-scored), measured on the Big-5 Aspects Scale (BFAS)", "Neuroticism (Z-scored), measured on the Big-5 Aspects Scale (BFAS)", "Social Comparison Scale (SCS): An 11-item scale that measures an individual’s perception of their social rank, attractiveness and belonging relative to others. The scale is scored as a sum of the 11 items (each measured on a 5-point scale), with higher scores indicating more favourable perceptions of social rank", "Depression Anxiety and Stress Scale (DASS-21): The DASS-21 includes 21 items, each measured on a 4-point scale. The score is derived from the sum of all 21 items, with higher scores indicating higher a severity of symptoms")
) %>% gt::gt()
```
  
__Preview__

The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/scs_study.csv') %>%  head %>% round(2) %>% gt::gt()
```

`r optend()`

# Setup

`r qbegin("Setup", qlabel = FALSE)`  

1. Create a new RMarkdown file
2. Load the required package(s)
3. Read the scs_study dataset into R, assigning it to an object named `scs_study` 
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r message=FALSE}
#Loading the required package(s)
library(tidyverse)
library(psych)
library(sjPlot)
library(kableExtra) 
library(sandwich)
library(interactions)

#Reading in data and storing in object named 'scs_study'
scs_study <- read_csv("https://uoepsy.github.io/data/scs_study.csv")
```

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Exercises 

## Study & Analysis Plan Overview 

`r qbegin(1)`

Provide a brief overview of the study design and data, before detailing your analysis plan to address the research question.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint 

- Give the reader some background on the context of the study
- State what type of analysis you will conduct in order to address the research question
- Specify the model to be fitted to address the research question 
- Specify your chosen significance ($\alpha$) level
- State your hypotheses

Much of the information required can be found in the [Study Overview] codebook.   

The [statistical models](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#statistical-models) flashcards may also be useful to refer to. Specifically the [interaction models flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#interaction-models) and [numeric x numeric example flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example) might be of most use.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The `scs_study` dataset contained information on `r nrow(scs_study)` participants, including $z$-scores on 5 personality traits assessed by the Big-Five Aspects Scale (BFAS; Openness, Conscientiousness, Extraversion, Agreeableness and Neuroticism). Participants were also assessed on the Social Comparison Scale (SCS), which is an 11-item scale measuring self-perception (relative to others) of social rank, attractiveness and belonging, and the Depression Anxiety and Stress Scale (DASS-21) - a 21 item measure with higher scores indicating higher severity of symptoms. For both of these measures, only total scores were available. Items in the SCS were measured on a 5-point scale, giving minimum and maximum possible scores of 11 and 55 respectively. Items in the DASS-21 were measured on a 4-point scale, meaning that scores could range from 21 to 84. 

Density plots and histograms will be used to visualise the marginal distributions of DASS-21 Scores, SCS Scores, and Neuroticism. To understand the strength of association among the variables, we will estimate the correlation coefficients; and to visualise these associations scatterplots will be used. To address the research question of whether Neuroticism moderated the effect of social comparison on depression and anxiety, we are going to fit the following interaction model: 

$$
\begin{align}
\text{DASS-21 Score} ~=~ & \beta_0 + \beta_1 \cdot \text{SCS Score} + \beta_2 \cdot \text{Neuroticism} \\  
& + \beta_3 \cdot (\text{Neuroticism} \cdot \text{SCS Score}) + \epsilon
\end{align}
$$ 
Effects will be considered statistically significant at $\alpha=.05$

Our hypotheses are:

$H_0: \beta_3 = 0$

The effect of social comparison on symptoms of depression, anxiety and stress does not vary depending on level of Neuroticism.

$H_1: \beta_3 \neq 0$

The effect of social comparison on symptoms of depression, anxiety and stress does vary depending on level of Neuroticism.

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Descriptive Statistics & Visualisations

`r qbegin(2)`

Provide a table of descriptive statistics and visualise your data. You may also want to consider estimating the associations among the variables of interest. 
 
Remember to interpret these in the context of the study.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint
Review the many ways to numerically and visually explore your data by reading over the [data exploration flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#data-exploration).

For examples, see flashcards on [descriptives statistics tables - numeric values only examples](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#data-exploration) and [numeric x numeric example - visualise data](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example).  

The `pairs.panels()` function may come in handy here to complete multiple tasks at once!  

:::
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.panel-tabset}

## Numeric

Descriptive statistics presented in a well formatted table:

```{r}
#| label: tbl-descript
#| tbl-cap: Descriptive Statistics - DASS-21, SCS, and Neuroticism (Z-Scored)
# note that we are selecting only our three variables of interest (dass, scs, zn)

scs_study %>%
    select(dass, scs, zn) %>% 
    describe() %>%
    kable(caption = "Descriptive Statistics - DASS-21, SCS, and Neuroticism (Z-Scored)", digits = 2) %>%
    kable_styling()
```


## Visual

Visualise associations among variables of interest:

```{r}
scs_study %>% 
  select(dass, scs, zn) %>%
  pairs.panels()
```

:::

Description of individual variables:

::: {.callout-important icon=false appearance="minimal"}
+ The marginal distribution of scores on the Depression, Anxiety and Stress Scale (DASS-21) was unimodal with a mean of `r round(mean(scs_study$dass), 2)` and a standard deviation of `r round(sd(scs_study$dass), 2)`. 
+ The marginal distribution of scores on the Social Comparison Scale (SCS) was unimodal with a mean of `r round(mean(scs_study$scs), 2)` and a standard deviation of `r round(sd(scs_study$scs), 2)`.
+ The marginal distribution of Neuroticism (Z-scored) was positively skewed.
:::

Description of correlations:

::: {.callout-important icon=false appearance="minimal"}
+ There was a weak, negative association between scores on the Depression Anxiety and Stress Scale and scores on the Social Comparison Scale for the participants in the sample $(r = -.23)$  
    + Severity of symptoms measured on the DASS-21 were lower, on average, for those who more favorably perceived their social rank    
+ There was a weak, positive association between DASS-21 Scores and levels of Neuroticism $(r = .20)$  
    + Participants who were more neurotic tended to, on average, display a higher severity of symptoms of depression, anxiety and stress    
:::

`r solend()`

<br> 

`r qbegin(3)`

For demonstration purposes to help us visualise and understand the associations among our variables a little better, copy and run the two code chunks below. It takes the dataset, and uses the `cut()` function to add a new variable called "zn_group", which is the "zn" variable split into 4 groups.  

```{r}
scs_study <-
  scs_study %>%
  mutate(
    zn_group = cut(zn, 4)
  )
```

We can see how it has split the "zn" variable by plotting the two against one another (note that the levels of the new variable are named according to the cut-points):  

```{r}
ggplot(data = scs_study, aes(x = zn_group, y = zn)) + 
  geom_point()
```

Plot the association between scores on the SCS and scores on the DASS-21, for each group of the variable we just created.  

How does the pattern differ across groups? Does it suggest an interaction?  

:::{.callout-tip appearance="simple" collapse="true"}

### Hint
Rather than creating four separate plots, you might want to map some feature of the plot to the variable we created in the data, or make use of `facet_wrap()` / `facet_grid()`. 

Remember that you can specify `geom_smooth()` to add a trend line. For a recap, review the facet examples contained within the [visual exploration flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#visual-exploration). Alternatively, review the [numeric x numeric example - visualise data flashcard](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example). 

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r message = FALSE, warning=FALSE, fig.width = 8, fig.height = 4, out.width = '90%'}
ggplot(data = scs_study, aes(x = scs, y = dass, col = zn_group)) + 
  geom_point() + 
  geom_smooth(method='lm', se = FALSE) +
  facet_grid(~zn_group) +
  labs(x = "SCS Scores ", y = "DASS-21 Scores") +
  theme(legend.position = "none") # removes the legend
```

The association between DASS-21 scores and SCS scores appears to be different across these groups. For those with a relatively high Neuroticism score, the association seems stronger, while for those with a low Neuroticism score there is almost no discernible association. 

This does suggest an interaction - the association of DASS-21 ~ SCS differed across the values of Neuroticism.

`r solend()`

<br>

:::frame

**Visualising Interaction Terms**

Cutting one of the explanatory variables up into groups essentially turns a numeric variable into a categorical one. We did this just to make it easier to visualise how an association differs across the values of another variable, because we can imagine a separate line for the association between SCS and DASS-21 scores for each of the groups of Neuroticism. However, in grouping a numeric variable like this we lose information. Neuroticism is measured on a continuous scale, and we want to capture how the association between SCS and DASS-21 differs across that continuum (rather than cutting it into chunks). 

We could imagine cutting it into more and more chunks (see @fig-reglinescut), until what we end up with is an infinite number of lines - i.e., a three-dimensional plane/surface (recall that in for a multiple regression model with 2 explanatory variables, we can think of the model as having three-dimensions). The inclusion of the interaction term simply results in this surface no longer being necessarily flat. You can see this in @fig-3dint. 
 
```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
#| label: fig-reglinescut
#| fig-cap: "Separate regression lines DASS ~ SCS for Neuroticism when cut into 4 (left) or 6 (center) or 12 (right) groups."

p1 <- ggplot(data = scs_study, aes(x = scs, y = dass, col = zn_group)) + 
  geom_point(alpha=.1) + 
  stat_smooth(method="lm",se=FALSE)+theme(legend.position = "none")

scs_study %>%
  mutate(
    zn_group = cut(zn, 6)
  ) %>% ggplot(data = ., aes(x = scs, y = dass, col = zn_group)) + 
  geom_point(alpha=.1) + 
  stat_smooth(method="lm",se=FALSE)+theme(legend.position = "none") -> p2

scs_study %>%
  mutate(
    zn_group = cut(zn, 12)
  ) %>% ggplot(data = ., aes(x = scs, y = dass, col = zn_group)) + 
  geom_point(alpha=.1) + 
  stat_smooth(method="lm",se=FALSE)+theme(legend.position = "none") -> p3

p1 | p2 | p3
```


```{r include=FALSE, echo=FALSE}
fit<-lm(dass ~ scs*zn, data = scs_study)
steps=50
scs <- with(scs_study, seq(min(scs),max(scs),length=steps))
zn <- with(scs_study, seq(min(zn),max(zn),length=steps))
newdat <- expand.grid(scs=scs, zn=zn)
dass <- matrix(predict(fit, newdat), steps, steps)
p <- persp(scs,zn,dass, theta = -25,phi=5, col = NA)
```


```{r echo=FALSE, fig.align = 'center', out.width = '90%'}
#| label: fig-3dint
#| fig-cap: "3D plot of regression surface with interaction. You can explore the plot in the figure below from different angles by moving it around with your mouse."

plot_ly(x=scs,y=zn,z=dass, type="surface") %>% layout(
    scene = list(
      xaxis = list(title = "SCS"),
      yaxis = list(title = "Neuroticism"),
      zaxis = list(title = "DASS-21")
    ))
```

:::

<div class="divider div-transparent div-dot"></div>

## Model Fitting & Interpretation 

`r qbegin(4)`

Consider that Neuroticism has already been $z$-scored, but scs has not. To ensure that we can compare the effects of our estimates (and so they are both on meaningful scales), standardize the scs variable.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

Review the [standardisation flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#data-transformations) for a recap if needed. Note, it would be best to create a new z-scored variable to then use within the model in this instance. 
:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
# standardize scs score
scs_study <- 
  scs_study %>% 
    mutate(
      zscs = (scs-mean(scs))/sd(scs)
    )
```

`r solend()`

<br>

`r qbegin(5)`

Fit your model (including the standardized predictor variables) using `lm()`, and assign it the name "dass_mdl".   

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

We can fit interaction models using the `lm()` function.  
  
For an overview, see the [interaction models flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#interaction-models).  
  
For an example, review the [interaction models > numeric x numeric example > model building flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example).

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#fit interaction model
dass_mdl <- lm(dass ~  zn*zscs, data = scs_study)

#check model output
summary(dass_mdl)
```

`r solend()`

<br>

`r qbegin(6)`

Interpret your coefficients in the context of the study.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

Recall that we can obtain our parameter estimates using various functions such as `summary()`,`coef()`, `coefficients()`, etc. 

For an overview of how to interpret coefficients, review the [interaction models > interpreting coefficients flashcard](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#interaction-models). It is important to remember that you have standardised the variables in your model, and this will affect your interpretation. For a recap on what standardisation is and how it affects the scales of your variables, review the [standardisation flashcard](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#data-transformations). 
   
For a specific example of coefficient interpretation, review the [interaction models > numeric x numeric example > results interpretation flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example).  

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Obtain parameter estimates:

```{r}
coefficients(dass_mdl)
```

::: {.panel-tabset}
## `(Intercept)`   

$\beta_0$ = `(Intercept)` = `r round(coef(dass_mdl)[1],2)`  

+ The intercept, or predicted DASS-21 score for an individual with an average SCS score and average Neuroticism score. In this case, the mean scores of SCS and Neuroticism are both 0 (since both have been standardised).  
    - An individual with a Neuroticism score of 0 ($Z$-scored) and an SCS score of 0 ($Z$-scored) was expected to have a DASS-21 score of `r round(coef(dass_mdl)[1],2)`.  

## `zn`  
$\beta_1$ = `zn` = `r round(coef(dass_mdl)[2],2)`  

+ The simple slope of Neuroticism on DASS-21 scores with average SCS scores (0; $Z$-scored).  
    - For an individual with an average SCS score (0; $Z$-scored), every 1 standard deviation increase in Neuroticism score was associated with a `r round(abs(coef(dass_mdl)[2]),2)` point increase in DASS-21 scores.
      
## `zscs`  
$\beta_2$ = `zscs` = `r round(coef(dass_mdl)[3],2)`  

+ The simple slope of SCS scores on DASS-21 scores with average Neuroticism scores (0; $Z$-scored).  
    - For an individual with an average Neuroticism score (0; $Z$-scored), every 1 standard deviation increase in SCS score was associated with a `r round(abs(coef(dass_mdl)[3]),2)` decrease in DASS-21 scores.  

## `zscs:zn`
$\beta_3$ = `zscs:zn` = `r round(coef(dass_mdl)[4],2)`  

+ The interaction between SCS score and Neuroticism on DASS-21 Scores - the change in the slope of SCS Scores as a function of Neuroticism.  
    - For every 1 standard deviation increase in SCS scores, when Neuroticism scores increased by 1 standard deviation, the slope with DASS-21 scores was adjusted by `r round(coef(dass_mdl)[4],2)`.   
    - The higher Neuroticism scores, the stronger the negative association between SCS Scores and DASS-21 scores - the association between SCS and DASS-21 scores became more negative by `r round(coef(dass_mdl)[4],2)` points.   
    
:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Visualise Interaction Model

`r qbegin(7)`

Using the `probe_interaction()` function from the interactions package, visualise the interaction effects from your model.

Try to summarise the interaction effects in a short and concise sentence. 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

For an overview and example, review the [interaction models > numeric x numeric example > model visualisation flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example). 

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r warning=FALSE, message=FALSE}
plt_dass_mdl <- probe_interaction(model = dass_mdl, 
                  pred = zscs, 
                  modx = zn, 
                  cond.int = T,
                  interval = T, 
                  jnplot = T,
                  main.title = "Neuroticism moderating the effect of\nsocial comparison on depression and anxiety",
                  x.label = "Social Comparison Scale (Z-scored)",
                  y.label = "DASS-21 Scores",
                  legend.main = "Neuroticism (Z-scored)")
```

Let's look at the plot - to do so you need to call `interactplot` from your object `plt_dass_mdl`:

```{r}
#| label: fig-int-dass
#| fig-cap: "Simple Slopes for +/- 1 SD and Mean Neuroticism Scores"

plt_dass_mdl$interactplot
```

::: {.callout-important icon=false appearance="minimal"}


Recall that higher DASS-21 scores indicate higher severity of symptoms. Based on this, we can state:

+ For individuals 1 SD below the sample mean on Neuroticism, as their SCS Score increases, it appears that their DASS-21 scores remain flat / there is a *very* slight increase (i.e., their wellbeing do not change much)
+ For individuals with average levels of Neuroticism, as their SCS Score increases, it appears that their DASS-21 scores decreases (i.e., their wellbeing increases)
+ For individuals 1 SD above the sample mean on Neuroticism, as their SCS Score increases, it appears that their DASS-21 scores more steeply decreases (i.e., their wellbeing increases)

:::

`r solend()`

<br>

`r qbegin(8)`

Conduct a simple slopes and regions of significance analysis.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint 

For an overview and example, review the [interaction models > numeric x numeric example > model visualisation flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example). Pay particular attention to how you can extract specific parts of output. 

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| label: fig-sim-slopes
#| fig-cap: "Johnson-Neyman Plot"
plt_dass_mdl$simslopes
```

::: {.callout-important icon=false appearance="minimal"}

The Johnson-Neyman technique (see @fig-sim-slopes) indicated that the association between DASS-21 scores and SCS was significant when Neuroticism scores were less than 1.28 standard deviations below the mean or greater than -0.55 standard deviations above the mean.

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

## Writing Up & Presenting Results

`r qbegin(9)`

Provide key model results in a formatted table.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint 

Use `tab_model()` from the __sjPlot__ package. For a quick guide, review the [tables flashcard](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#tables).

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| label: tbl-dass-modresults
#| tbl-cap: Regression table for DASS-21 model
#create table for results
tab_model(dass_mdl,
          dv.labels = "DASS-21 Scores",
          pred.labels = c("zscs" = "Social Comparison Scale (Z-scored)",
                          "zn" = "Neuroticism (Z-scored)",
                          "zscs:zn" = "Social Comparison Scale (Z-scored): Neutoricism (Z-scored)"),
          title = "Regression Table for DASS-21 Model")
```

`r solend()`

<br>

`r qbegin(10)`

Interpret your results in the context of the research question and report your model in full.

Make reference to the interaction plot and regression table.

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

For an example of coefficient interpretation, review the [interaction models > numeric x numeric example > results interpretation flashcards](https://uoepsy.github.io/dapr2/2425/labs/1_b3_reading.html#numeric-x-numeric-example). 

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::: {.callout-important icon=false appearance="minimal"}

Full regression results including 95% confidence intervals are shown in @tbl-dass-modresults. The $F$-test for model utility was significant $(F(3,652) = 48.50, p<.001)$, and the model explained approximately 17.87% of the variability in DASS-21 scores.

There was a significant conditional association between DASS-21 Scores and SCS scores ($Z$-scored) $(\beta = -1.57, SE = 0.24, p < .001)$, suggesting that for those with Neuroticism scores of 0, DASS-21 scores decreased by 1.57 for every 1 standard deviation increase in SCS scores. 

A significant conditional association was also evident between DASS-21 Scores and Neuroticism ($Z$-scored) $(\beta = 1.58, SE = 0.24, p <.001)$, suggesting that for those with SCS scores of 0, DASS-21 scores increased by 1.58 for every 1 standard deviation increase in Neuroticism. 

The association between symptoms of depression and anxiety (DASS-21 scores) and social comparison was found to be dependent upon the level of Neuroticism, with a greater negative association between the two for those with higher levels of Neuroticism $(\beta = -1.83, SE = 0.23, p <.001)$. For every standard deviation increase in SCS Scores, the change in DASS-21 scores associated with an increase of 1 SD in Neuroticism was adjusted by -1.83. Thus, Neuroticism buffered the association between DASS-21 scores and SCS - this is visually displayed in @fig-int-dass. We further used the Johnson-Neyman technique to probe the interaction, and to identify regions of significance. We identified that Neuroticism values (z-scored) outside the range of -1.28 to -0.55 were significant (see @fig-sim-slopes).

Therefore, we have evidence to reject the null hypothesis (that the effect of social comparison on symptoms of depression, anxiety and stress does not vary depending on level of Neuroticism).

:::

`r solend()`

<div class="divider div-transparent div-dot"></div>

# Compile Report

`r qbegin("Compile Report", qlabel = FALSE)`  

Knit your report to PDF, and check over your work. To do so, you should make sure:

- Only the output you want your reader to see is visible (e.g., do you want to hide your code?)
- Check that the **tinytex** package is installed
- Ensure that the ‘yaml’ (bit at the very top of your document) looks something like this:

```{}
---
title: "this is my report title"
author: "B1234506"
date: "07/09/2024"
output: bookdown::pdf_document2
---
```

:::{.callout-tip appearance="simple" collapse="true"}
# What to do if you cannot knit to PDF
If you are having issues knitting directly to PDF, try the following:  

- Knit to HTML file  
- Open your HTML in a web-browser (e.g. Chrome, Firefox)  
- Print to PDF (Ctrl+P, then choose to save to PDF)  
- Open file to check formatting
:::

:::{.callout-tip appearance="simple" collapse="true"}
# Hiding Code and/or Output

:::{.panel-tabset}
## Hiding R Code

To not show the code of an R code chunk, and only show the output, write:

````
```{{r, echo=FALSE}}
# code goes here
```
````

## Hiding R Output

To show the code of an R code chunk, but hide the output, write:

````
```{{r, results='hide'}}
# code goes here
```
````

## Hiding R Code and Output

To hide both code and output of an R code chunk, write:

````
```{{r, include=FALSE}}
# code goes here
```
````
:::

:::

:::{.callout-tip appearance="simple" collapse="true"}

### Tinytex
You must make sure you have **tinytex** installed in R so that you can “Knit” your Rmd document to a PDF file:

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

:::

`r qend()`

`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`

You should end up with a PDF file. If you have followed the above instructions and still have issues with knitting, speak with a Tutor. 

`r solend()`
