---
title: "Coding Factors"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
---


```{r setup, include=FALSE}
source('assets/setup.R')
```

```{r echo=FALSE}
set.seed(3)
```

:::lo
**LEARNING OBJECTIVES**

1. Understand how to specify dummy and effects coding
2. Understand how to specify contracts to test specific effects

:::

# Research question

```{r eval=FALSE, echo=FALSE}
library(tidyverse)

set.seed(3)

rest_spend <- tibble(
  type = c('Classical Music', 'Pop Music', 'No Music'),
  n = c(120, 120, 120),
  mu = c(24.13, 21.91, 21.70),
  sigma = c(2.243, 2.627, 3.332)
)

rest_spend <- rest_spend %>% 
  rowwise() %>% 
  mutate(data = list(rnorm(n, mu, sigma))) %>%
  unnest(data) %>%
  select(-n, -mu, -sigma)

rest_spend <- rest_spend %>%
  sample_n(nrow(.), replace = FALSE) %>%
  mutate(id = 1:nrow(.)) %>%
  select(id, everything()) %>%
  rename(amount = data)
rest_spend

write_csv(rest_spend, '../../data/RestaurantSpending.csv')
```



> _Do you think the type of background music playing in a restaurant influences the amount of money that diners spend on their meal?_


`r optbegin('Data: RestaurantSpending.csv', FALSE, show = TRUE, toggle = TRUE)`
**Download link**

[Download the data
here](https://uoepsy.github.io/data/RestaurantSpending.csv)

**Description**

A group of researchers wanted to test the claims reported in [@North2003] on whether the type of background music playing in a restaurant influences the average amount of money spent by diners on their meal.

The group researchers got in touch with a restaurant and asked to alternate silence, popular music, and classical music on successive nights over 18 days. On those nights they recorded the mean spend per head for each table.

The following variables were collected:

- `id`: Identifier for each diner
- `type`: type of music played ("Classical Music", "Pop Music", "No Music")
- `amount`: restaurant spending per person (in pounds - £)

**Preview**

```{r echo=FALSE}
library(tidyverse)

rest_spend <- read_csv('https://uoepsy.github.io/data/RestaurantSpending.csv')
```

The top six rows of the data are as follows:

<center>
```{r echo=FALSE}
pander::pander(head(rest_spend))
```
</center>
`r optend()`

## Data Exploration: Descriptives

`r qbegin(1)`
(1) Load the `tidyverse` package.
(2) Read the restaurant spending data into R using the function `read_csv()` and name the data `rest_spend`.
(3) Check for the correct coding of all variables (i.e., categorical variables should be factors and numeric variables should be numeric).

`r qend()`


`r solbegin(show = params$SHOW_SOLS, toggle = params$TOGGLE)`

Load the tidyverse library:

```{r}
library(tidyverse)
```

Read the data into R:
```{r}
rest_spend <- read_csv('https://uoepsy.github.io/data/RestaurantSpending.csv')
```

Check if the data were read into R correctly by examining the top six rows:
```{r}
head(rest_spend)
```

Music type is a categorical variable but it is encoded as a character (`<chr>`) variable rather than a factor (`<fctr>`). Let's fix this:

```{r eval=FALSE}
# We could do this the standard way
rest_spend$type <- as.factor(rest_spend$type)
```

```{r}
# Or the tidyverse way
rest_spend <- rest_spend %>%
  mutate(type = as.factor(type))
```

And check again:
```{r}
head(rest_spend)
```

`r solend()`


`r qbegin(2)`
Produce a boxplot displaying the relationship between restaurant spending and music type, and comment on any visually observed differences among the sample means of the three background music types.
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


```{r}
ggplot(data = rest_spend, aes(x = type, y = amount)) +
  geom_boxplot() +
  labs(x = 'Background music type', y = 'Restaurant spending (in GBP)')
```

:::int 

From the boxplots, it seems that customers without background music or pop music had a similar average restaurant spending.

Furthermore, the average restaurant spending seems to be higher for those who had a classical music background compared to the customers either without music or with a pop music background.

:::

`r solend()`


`r qbegin(3)`

For the restaurant spending data, what is the number of groups ($g$) and the numbers of observations per group ($n$)? What was the average spending and standard deviation of each group?
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
rest_spend %>%
  group_by(type) %>%
  summarise(n = n(), 
            M = mean(amount), 
            SD = sd(amount))
```


There are three groups, one for each music type: "Classical Music", "No Music", "Pop Music". Hence, $g = 3$. There are 120 observations for each group, i.e. $n = 120$.

`r solend()`


## Levels of Variable & Side Constraints

### Levels of variable

R by default orders the levels of a categorical variable in alphabetical order:
```{r}
levels(rest_spend$type)
```

If we follow the same approach and order the groups by alphabetical ordering, we can represent the data as follows:

| Population Name | Population ID | Sample observations  | Population mean |
|:---------------:|:-------------:|:--------------------:|:---------------:|
| Classical Music | 1 | $y_{1,1}, y_{1,2}, ..., y_{1,n}$ | $\mu_{1}$ |
| No Music        | 2 | $y_{2,1}, y_{2,2}, ..., y_{2,n}$ | $\mu_{2}$ |
| Pop Music       | 3 | $y_{3,1}, y_{3,2}, ..., y_{3,n}$ | $\mu_{3}$ |


:::frame 
#### Sum to zero constraint - effects coding

Under the constraint $\beta_1 + \beta_2 + \beta_3 = 0$, the model coefficients are interpreted as follows:

- $\beta_0 = \mu$ is interpreted as the overall or global mean --- that is, the population mean when all the groups are combined;
- $\beta_i = \mu_i - \mu$ is interpreted as the effect of group $i$ --- that is, the difference between the population mean for group $i$, $\mu_i$, and the global population mean, $\mu$.


We can fit the linear regression model:

$$
y 
= b_0 
+ b_1 \ \text{EffectLevel1} 
+ b_2 \ \text{EffectLevel2} 
+ \epsilon
$$

where

$$
\text{EffectLevel1} = \begin{cases}
1  & \text{if observation is from category 1} \\
0  & \text{if observation is from category 2} \\
-1 & \text{otherwise}
\end{cases}
\\
\text{EffectLevel2} = \begin{cases}
0  & \text{if observation is from category 1} \\
1  & \text{if observation is from category 2} \\
-1 & \text{otherwise}
\end{cases}
$$

Schematically,

$$
\begin{matrix}
\textbf{Level}  & \textbf{EffectLevel1} & \textbf{EffectLevel2} \\
\hline
\text{Classical Music} & 1   & 0    \\
\text{No Music}        & 0   & 1    \\
\text{Pop Music}       & -1  & -1
\end{matrix}
$$


:::blue
In **R** this is simply done by saying:

```{r}
contrasts(rest_spend$type) <- "contr.sum"
mdl_sum <- lm(amount ~ type, data = rest_spend)
coef(mdl_sum)
```

where `type` is a factor. In such case R will automatically create the two variables `EffectLevel1` and `EffectLevel2` for you!

The `summary(mdl)` will return 3 estimated coefficients:

- $b_0 = 22.74$ is estimated overall mean
- $b_1 = 1.44$ is the effect of Classical Music
- $b_2 = -0.6$ is the effect of No Music
- **NOTE**: The effect of Pop Music is not shown by `summary` as it is found via the side-constraint as $b_3 = -(b_1 + b_2) = -0.84$!
:::
:::


:::frame
#### Reference group constraint - dummy coding

Under the constraint $\beta_1 = 0$, meaning that the first factor level is the reference group,

- $\beta_0$ is interpreted as $\mu_1$, the mean response for the reference group (group 1);
- $\beta_i$ is interpreted as the difference between the mean response for group $i$ and the reference group.

We can fit the linear regression model:

$$
y 
= b_0 
+ b_1 \ \text{IsTypeNoMusic} 
+ b_2 \ \text{IsTypePopMusic} 
+\epsilon
$$

where

$$
\text{IsTypeNoMusic} = \begin{cases}
1 & \text{if observation is from the No Music category} \\
0 & \text{otherwise}
\end{cases}
\\
\text{IsTypePopMusic} = \begin{cases}
1 & \text{if observation is from the Pop Music category} \\
0 & \text{otherwise}
\end{cases}
$$

Schematically,

$$
\begin{matrix}
\textbf{Level}  & \textbf{IsTypeNoMusic} & \textbf{IsTypePopMusic} \\
\hline
\text{Classical Music} & 0   & 0   \\
\text{No Music}        & 1   & 0   \\
\text{Pop Music}       & 0   & 1
\end{matrix}
$$

:::blue
In **R** this is simply done by saying:

```{r}
contrasts(rest_spend$type) <- "contr.treatment"
mdl_trt <- lm(amount ~ type, data = rest_spend)
coef(mdl_trt)
```

where `type` is a factor. In such case R will automatically create the two dummy variables `IsTypeNoMusic` and `IsTypePopMusic` for you!

The `summary(mdl)` will return 3 estimated coefficients:

- $b_0 = 24.17$ is the mean of the base level (level 1)
- $b_1 = -2.03$ is the effect of No Music
- $b_2 = -2.28$ is the effect of Pop Music
- **NOTE**: The effect of Classical Music is not shown as it is equal to zero by the constraint!
:::
:::

### Side Contraints

Possible side-constraints on the parameters are:

|       Name      |             Constraint            |             Meaning of $\beta_0$            |         R         |
|:---------------:|:---------------------------------:|:-------------------------------------------:|:-----------------:|
| Sum to zero (Effects Coding) | $\beta_1 + \beta_2 + \beta_3 = 0$ | $\beta_0 = \mu$   |    `contr.sum`    |
| Reference group (Dummy Coding) |           $\beta_1 = 0$           | $\beta_0 = \mu_1$ | `contr.treatment` |


:::yellow
**IMPORTANT**

- By default R uses the reference group constraint

- If your factor has $g$ levels, your regression model will have $g-1$ dummy variables (R creates them for you)
:::



# ANOVA F-test

Since R uses the reference group constraint, in this section and the next one, we will not specify any side-constraints, meaning that $\beta_1 = 0$.

`r qbegin(4)`
Fit a linear model to the data, having `amount` as response variable and the factor `type` as predictor. Call the fitted model `mdl_rg` (for reference group constraint)

Identify the relevant pieces of information from the commands `anova(mdl_rg)` and `summary(mdl_rg)` that can be used to conduct an ANOVA F-test against the null hypothesis that all population means are equal.

Interpret the F-test results in the context of the ANOVA null hypothesis.
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
mdl_rg <- lm(amount ~ type, data = rest_spend)
mdl_rg

summary(mdl_rg)

anova(mdl_rg)
```

The model summary returns the F-test of model utility which, in this case, corresponds to the ANOVA F-test against the null hypothesis of equal population means.

The relevant line from `summary()` is:
```
F-statistic: 23.21 on 2 and 357 DF,  p-value: 3.335e-10
```

The relevant parts from `anova()` are:

- `F value` of `23.211`
- The `Df` column giving `2` and `357` degrees of freedom
- The p-value of the test, reported under `Pr(>F)` as `3.335e-10 ***`.

We can write this up as follows:

:::int
We performed an analysis of variance against the null hypothesis of equal population mean spending across three types of background music, $F(2, 357) = 23.21$, $p < .001$.

The large observed F statistic leads to a very small *p*-value, meaning that such a large observed variability among the mean restaurant spending across the different music types, compared to the variability in the residuals, is very unlikely to happen by chance alone if the population means where all the same.

For this reason, at the 5\% significance level we reject the null hypothesis as there is strong evidence that at least two population means differ.
:::
`r solend()`


# Reference group (dummy) constraint

`r qbegin(5)`
Examine and investigate the meaning of the coefficients in the output of `summary(mdl_rg)`. Next, obtain the estimated (or predicted) group means for the "Classical Music", "No Music", and "Pop Music" groups by using the `predict(mdl_rg, newdata = ...)` function.
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
Let's recall the model summary:
```{r}
summary(mdl_rg)
```

The interpretation is as follows

| Coefficient | Estimate | Corresponds to |
|:-:|:-:|:-:|
| (Intercept)   |  24.1741 | $\hat \beta_0 = \hat \mu_1$ |
| typeNo Music  | -2.0328  | $\hat \beta_2 = \hat \mu_2 - \hat \beta_0 = \hat \mu_2 - \hat \mu_1$ |
| typePop Music | -2.2752  | $\hat \beta_3 = \hat \mu_3 - \hat \beta_0 = \hat \mu_3 - \hat \mu_1$ |

The estimate corresponding to (Intercept) contains $\hat \beta_0 = \hat \mu_1 = 24.17$. The estimated average spending for those having a classical music background is approximately £24.2.

The next estimate corresponds to `typeNo Music` and is $\hat \beta_2 = -2.033$. The difference in mean spending between `No Music` and `Classical Music` is estimated to be $-2.033$. In other words, people with a silent background seem to spend approximately £2 less than those having a classical music background.

The estimate corresponding to `typePop Music` is $\hat \beta_3 = -2.275$. This is the estimated difference in mean spending between `Pop Music` and `Classical Music`. People with a pop music background seem to spend approximately £2.3 less than those with a classical music background.

Hence, for all levels except the reference group we see differences to the reference group while the estimate of the reference level can be found next to `(Intercept)`.

It is also important to notice how the coefficients names are written. They are a combination of `factor name` and `level name`, such as `typeNo Music`.
The only coefficient that is missing is `typeClassical Music`, the one corresponding to the reference category `Classical Music`.

To use `preditct`, you first need to define a data frame with a column having the same name as the factor in the fitted model. Then, specify all the groups (= levels) for which you would like the predicted mean.
```{r}
query_groups <- tibble(type = c("Classical Music", "No Music", "Pop Music"))
query_groups
```

Pass the data frame to the predict function using the `newdata =` argument. The predict function will match the column named `type` with the predictor called `type` in the fitted model `mdl_rg`.

```{r}
predict(mdl_rg, newdata = query_groups)
```

Or:
```{r}
query_groups %>%
  mutate(pred = predict(mdl_rg, newdata = .))
```


Hence,

- $\hat \mu_\text{Classical Music} = \hat \mu_1 = 24.17$
- $\hat \mu_\text{No Music} = \hat \mu_2 = 22.14$
- $\hat \mu_\text{Pop Music} = \hat \mu_3 = 21.90$

`r solend()`

`r qbegin(6)`
It actually makes more sense to have "No Music" as reference group.

Open the help page of the function `fct_relevel()`, and look at the examples.

Within the `rest_sped` data, reorder the levels of the factor `type` so that your reference group is "No Music".

:::yellow
**NOTE**

Because you reordered the levels, now 

- $\mu_1$ = mean of no music group
- $\mu_2$ = mean of classical music group
- $\mu_3$ = mean of pop music group
:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
Check the help page:
```{r eval=FALSE}
?fct_relevel
```

Select "No Music" to be the first level:
```{r}
rest_spend$type <- fct_relevel(rest_spend$type, "No Music")
```

Check that the factor levels are now in the correct order:
```{r}
levels(rest_spend$type)
```

`r solend()`

`r qbegin(7)`

Refit the linear model, and inspect the model summary once more.

Do you notice any change in the estimated coefficients? Do you notice any change in the F-test for model utility?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
mdl_rg <- lm(amount ~ type, data = rest_spend)
summary(mdl_rg)
```
The displayed coefficients have changed as now the (Intercept) represents the new reference group, which is "No Music".

However, by checking the calculations, this model is equivalent to the previous one as the predicted group means are the same.

- Predicted mean of "No Music" $\hat \mu_\text{No Music}$ = 22.1414
- Predicted mean of "Classical Music" $\hat \mu_\text{Classical Music}$ = 22.1414 + 2.0328 = 24.1742
- Predicted mean of "Pop Music" = $\hat \mu_\text{Pop Music}$ = 22.1414 - 0.2424 = 21.899

The `predict` function returns the same output:
```{r}
query_groups <- tibble(type = levels(rest_spend$type))
query_groups

query_groups %>% 
  mutate(pred = predict(mdl_rg, newdata = .))
```

`r solend()`


# Sum-to-zero (effects) constraint

Let's now change the side-constraint from the R default (the reference group or `contr.treatment` constraint) to the sum-to-zero constraint using `contr.sum`.

Recall that under this constraint the interpretation of the coefficients becomes:

- $\beta_0$ represents the global mean
- $\beta_i$ the effect due to group $i$ --- that is, the mean response in group $i$ minus the global mean


`r qbegin(8)`
Set the sum to zero constraint for the factor `type` of background music.

Fit again the linear model using `amount` as response variable and `type` as predictor. Call the fitted model using the sum to zero constraint `mdl_stz` (for sum to zero).

Examine the output of the `summary(mdl_stz)` function:

- Do the displayed coefficients change? What do they represent now?
- Do the predicted group means change?
- Is the model utility F-test still the same? Why do you think it's the case?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Set the sum to zero contrast to the `type` factor:
```{r}
contrasts(rest_spend$type) <- "contr.sum"
```

Re-fit the model and inspect the summary output:
```{r}
mdl_stz <- lm(amount ~ type, data = rest_spend)
summary(mdl_stz)
```

With the new side-constraint, we get different values for the model coefficients because the meaning of the parameters has changed. If we closely inspect the output, we can also notice that a slightly different naming has been used.

Instead of `factor name` and `level name` (such as `typeNo Music`), we get `factor name` and `level number` (e.g. `type1`). 
Here, `type1` simply means the first level of the factor `type`, while `type2` means the second level of the factor `type`.

```{r}
levels(rest_spend$type)
```

The first level is "No Music", while the second level is "Classical Music".

- The estimate for `(Intercept)` is now the *global mean* of the data.
- The estimate for `type1` is now the difference of the first group ("No Music") to the global mean
- The estimate for `type2` is now the difference of the second group ("Classical Music") to the global mean

The estimate for `type3`, representing the difference of "Pop Music" to the global mean is not shown. Because of the side-constraint, we know it must be $\beta_3 = - (\beta_1 + \beta_2) = - (-0.5968 + 1.4360) =  -0.8392$.

Underneath we see the group effects that need to be added to the global mean to obtain the predicted group means.

Compute the predicted group means:
```{r}
query_groups %>%
  mutate(pred = predict(mdl_stz, newdata = .))
```

We notice that we get the same predicted group means independently of the side-constraint.

The predicted means do _not_ depend on the side-constraint that we apply. However, the side-constraint influences the meaning of the parameters in the model.

`r solend()`


:::blue
__In **R**, we can switch back to the default reference group constraint by either of these__

```{r}
# Option 1
contrasts(rest_spend$type) <- NULL
# Option 2
contrasts(rest_spend$type) <- "contr.treatment"
```
:::


# Contrasts

When the ANOVA F-test against the null hypothesis $H_0: \mu_1 = \mu_2 = \cdots = \mu_g$ is rejected, it tells us that there are significant differences between at least two of the group means. However, it does not tells us *which* particular population means are different. Further investigation should be conducted to find out **exactly** which population means differ. In some cases the investigator will have pre-planned comparisons they would like to make; in other situations they may have no idea what differences to look for.

:::yellow
__Contrasts__ represent a wide range of hypotheses about the population means that can be tested using the fitted model.

However, they need to match the following general form. A contrast only allows us to test hypotheses that can be written as a linear combination of the population means with coefficients summing to zero:
$$
H_0 : c_1 \mu_1 + c_2 \mu_2 + c_3 \mu_3 = 0 \qquad \text{with} \qquad c_1 + c_2 + c_3 = 0
$$
:::

We have already performed some simple contrasts by looking at the model coefficients under the reference group constraint.
Those represented pairwise comparisons between each group and the reference group. To test other hypotheses, such as combinations of two groups against a third group, we can use a package called `emmeans`.

```{r}
#install.packages("emmeans") remember that you only install packages once, and you should do so in the console, not your .Rmd script
library(emmeans)
```

`r qbegin(9)`
We were interested in the following comparisons:

- Whether having some kind of background music, rather than no music, makes a difference in mean restaurant spending
- Whether there is any difference in mean restaurant spending between those with no background music and those with pop music

These are planned comparisons and can be translated to the following research hypotheses:
$$
\begin{aligned}
1. \quad H_0 &: \mu_\text{No Music} = \frac{1}{2} (\mu_\text{Classical Music} + \mu_\text{Pop Music}) \\
2. \quad H_0 &: \mu_\text{No Music} = \mu_\text{Pop Music}
\end{aligned}
$$

First of all check the levels of the factor `type`, as the contrast coefficients need to match the order of the levels:
```{r}
levels(rest_spend$type)
```

We specify the hypotheses by first giving each a name, and then specify the coefficients of the comparisons:
```{r}
comp <- list("No Music - Some Music" = c(1, -1/2, -1/2),
             "No Music - Pop Music" = c(1, 0, -1))
```

Use the `emmeans()` function to obtain the estimated treatment means and uncertainties for your factor:
```{r}
emm <- emmeans(mdl_stz, ~ type)
emm
```

Next, we run the contrast analysis. To do so, we use the `contrast()` function:
```{r}
comp_res <- contrast(emm, method = comp)
comp_res
```

Finally, we can ask for 95\% confidence intervals:
```{r}
confint(comp_res)
```

Interpret the result of the contrast analysis.

- What do the p-values tell us?
- Interpret the confidence intervals in the context of the study.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

The hypothesis test for the first contrast could be reported as follows:

:::int
We performed a test against $H_0: \mu_1 - \frac{1}{2}(\mu_2 + \mu_3) = 0$, resulting in $t(357) = -2.818, p = .005$, two-sided.
At the 5\% significance level, there is strong evidence that the mean spending for those with no music is different than the mean spending for those with some of background music.
:::


The 95\% confidence interval for $\mu_1 - \frac{1}{2}(\mu_2 + \mu_3)$ is given by $[-1.520, -0.270]$.

The corresponding confidence interval could be written up as:

:::int
We are 95\% confident that diners with no background music will spend, on average, between £1.52 and £0.27 pounds less than those with some background music.
:::

<br>

The hypothesis test for the second contrast could be reported as follows:

:::int
We performed a test against $H_0: \mu_1 - \mu_3 = 0$, resulting in $t(357) = 0.661, p = .509$, two-sided.
At the 5\% significance level, there is not sufficient evidence to reject the null hypothesis that the mean spending for those with no music is the same as the mean spending for those with pop music.
:::


The 95\% confidence interval for $\mu_1 - \mu_3$ is given by $[-0.479, 0.964]$.

Because we have established that there isn't a significant difference between the mean spending for those with no music and those with pop music, we will not interpret the confidence interval.

`r solend()`

# References


<!-- Formatting -->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
