---
title: "Binary Logistic Regression"
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')

set.seed(953)

library(DT)
library(pander)
library(sjPlot)
library(tidyverse)

```

:::lo

### <i class="fa fa-graduation-cap"></i> Learning Objectives
At the end of this lab, you will:

1. 1. Understand when to use a logistic model
2. Understand how to fit and interpret a logistic model

### <i class="fa fa-check-square-o fa-2"></i> What You Need

1. Be up to date with lectures
2. Have completed Labs 1-4

### <i class="fab fa-r-project"></i> Required R Packages
Remember to load all packages within a code chunk at the start of your RMarkdown file using `library()`. If you do not have a package and need to install, do so within the console using `install.packages(" ")`. For further guidance on installing/updating packages, see Section C [here](https://uoepsy.github.io/files/install-update-r#update-pkgs). 

For this lab, you will need to load the following package(s):

* **tidyverse** 
* **patchwork**
* **kableExtra**
* **emmeans**
* **pander**

### <i class="fa fa-file"></i> Lab Data
You can download the data required for this lab [here](https://uoepsy.github.io/data/drunkdoor.csv) or read it in via this link https://uoepsy.github.io/data/drunkdoor.csv. 

:::


# Study Overview

> **Research Question** 
>
> Is susceptibility to change blindness influenced by level of alcohol intoxication and perceptual load?

Watch the following video:

<p align="center">
<iframe width="560" height="315" src="https://www.youtube.com/embed/FWSxSQsspiQ" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe><br>
[Simons, D. J., & Levin, D. T. (1997). Change blindness. Trends in cognitive sciences, 1(7), 261-267.](https://www.cse.iitk.ac.in/users/se367/11/papers/simons-ambinder-05_change-blindness.pdf)
</p>

You may well have already heard of these series of experiments, or have seen similar things on [Netflix](https://www.netflix.com/gb/title/80190510).

`r optbegin("Cognitive Exp 3x3 Codebook", olabel=FALSE, toggle=params$TOGGLE)` 

__Description__

**Method**  
Researchers conducted a study in which they approached 120 people, recruited from within the vicinity of a number of establishments with licenses to sell alcohol to be consumed on-premises. Initially, experimenter A approached participants and asked if they were interested in participating in a short study, and obtained their written consent. While experimenter A subsequently talked each participant through a set of questions on multiple pieces of paper (with the pretense of explaining what the participant was required to do), experimenters B and C carrying a door passed between the participant and experimenter A, with experimenter C replacing A (as can be viewed in the video). 

The perceptual load of the experiment was manipulated via a) the presentation of the door and b) the papers held by the experimenters. For 60 of these participants, the door was painted with some detailed graffiti and had a variety of pieces of paper and notices attached to the side facing the participants. Additionally, for these participants, the experimenters handled a disorganised pile of 30 papers, with the top pages covered in drawings around the printed text. For the remaining 60, the door was a standard MDF construction painted a neutral grey, and the experimenters handled only 2 sheets of paper which had minimal printed text on them and nothing else. 

**Measures**   
After experimenters A and C had successfully swapped positions, the participant was asked (now by C) to complete small number of questions taking approximately 1 minute. Either after this set of questions, or if the participant made an indication that they had noticed the swap, the experimenters regrouped and the participant was explicitly asked whether they had noticed the swap.   
Immediately after this, participants were breathalysed, and their blood alcohol content was recorded.  

__Preview__

The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/drunkdoor.csv') %>%  head %>% gt::gt()
```

`r optend()`

# Setup

`r qbegin("Setup", qlabel = FALSE)`  

1. Create a new RMarkdown file
2. Load the required package(s)
3. Read the caffeinedrink dataset into R, assigning it to an object named `drunkdoor` 
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r message=FALSE}
#load packages
library(tidyverse)
library(psych)
library(kableExtra)

#read in data
drunkdoor <- read_csv("https://uoepsy.github.io/data/drunkdoor.csv")
```

`r solend()`

<br>

`r qbegin(1)`

Examine the dataset, and perform any necessary and appropriate data management steps.

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#look at structure of data
str(drunkdoor)

#check for NAs - there are none - all FALSE
table(is.na(drunkdoor))

```

`r solend()`














