---
title: "Sample Size and Power Analysis"
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')

set.seed(953)

library(DT)
library(pander)
library(sjPlot)
library(tidyverse)
library(car)

```

:::lo

### <i class="fa fa-graduation-cap"></i> Learning Objectives
At the end of this lab, you will:

1. Understand differences between Type I and Type II errors
1. Understand how varying factors can influence power
1. Be able to conduct power analyses using the `pwr` package

### <i class="fa fa-check-square-o fa-2"></i> What You Need

1. Be up to date with lectures
2. Have completed previous lab exercises from [Week 6](https://uoepsy.github.io/dapr2/2223/labs/2_06_glm1.html)

### <i class="fab fa-r-project"></i> Required R Packages
Remember to load all packages within a code chunk at the start of your RMarkdown file using `library()`. If you do not have a package and need to install, do so within the console using `install.packages(" ")`. For further guidance on installing/updating packages, see Section C [here](https://uoepsy.github.io/files/install-update-r#update-pkgs). 

For this lab, you will need to load the following package(s):

* **tidyverse** 
* **patchwork**
* **kableExtra**
* **psych**
* **sjPlot**
* **pwr**

### <i class="fa fa-file"></i> Lab Data
You can download the data required for this lab [here](https://uoepsy.github.io/data/QuitAttempts.csv) or read it in via this link https://uoepsy.github.io/data/QuitAttempts.csv. 

:::


# Study Overview

> **Research Question**
>
> Does intervention type influence recall?

`r optbegin("Smoking Codebook", olabel=FALSE, toggle=params$TOGGLE)` 

__Description__

A sample of 200 individuals aged 18-75 years old was randomly allocated to either take a free-recall test after an intervention. There were two types of intervention - exciting (1-hour long roller-coaster session; group = 0) or relaxing (1-hour long meditation session; group = 1). 

__Data Dictionary__
```{r echo=FALSE, message=FALSE, warning=FALSE}
recdata <- read_csv("https://uoepsy.github.io/data/QuitAttempts.csv")
tibble(
variable = names(recdata),
description = c("Percentage Recall - the percentage of correctly recalled items","Group - whether the individual was assigned to the roller-coaster (group = 0) or the meditation (group = 1) intervention condition", "Age - individual's age (in years)")
) %>% gt::gt()
```

__Preview__

The first six rows of the data are:

```{r echo=FALSE, message=FALSE}
read_csv('https://uoepsy.github.io/data/recall_med_coast.csv') %>% head %>% gt::gt()
```

`r optend()`

# Setup

`r qbegin("Setup", qlabel = FALSE)`  

1. Create a new RMarkdown file
2. Load the required package(s)
3. Read in the recall_med_coast dataset into R, assigning it to an object named `recdata` 
 
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r message=FALSE}
#load packages
library(tidyverse)
library(psych)
library(kableExtra)
library(patchwork)
library(sjPlot)
library(pwr)

#read in data
recdata <- read_csv("https://uoepsy.github.io/data/recall_med_coast.csv")
```

`r solend()`

<br>

`r qbegin(1)`

Examine the dataset, and perform any necessary and appropriate data management steps.

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#look at structure of data:
str(recdata)

#check for NAs - there are none - all FALSE:
table(is.na(recdata))

#Group should be a factor:
recdata$group <- factor(recdata$group, 
                        levels = c(0, 1), 
                        labels = c('rollercoaster', 'meditation'))

```

`r solend()`

<br>

`r qbegin(2)`

Provide a table of descriptive statistics and visualise your data.

Remember to interpret your plot in the context of the study. 

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

1. For your table of descriptive statistics, both the `group_by()` and `summarise()` functions will come in handy here.
2. Recall that when visualising a continuous outcome across groups, `geom_boxplot()` may be most appropriate to use.
3. Make sure to comment on any observed differences among the sample means of the four treatment conditions.

:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Let's first produce a descriptive statistics table:

```{r message=FALSE, warning=FALSE}
#| label: tbl-cog-descript
#| tbl-cap: Descriptive Statistics
recall_stats <- recdata %>%
    group_by(group) %>%
    summarise(
       n = n(),
       Avg_recall = mean(perc_recall)) %>%
    kable(caption = "Descriptive Statistics") %>%
    kable_styling()

recall_stats
```

We can explore the association between our variables as follows:

```{r}
#| label: fig-recall-desc1
#| fig-cap: "Association between Age, Intervention, and Recall"
recall_plt1 <- ggplot(data = recdata, aes(x = age, y = perc_recall, fill = group)) +
    geom_boxplot() + 
    labs(x = "Age (in years", y = "Recall (%)", title = "Association between Age, Intervention, and Recall")
recall_plt1
```

`r solend()`

<br>

`r qbegin(3)`
Using a significance level of .05, what sample size ($n$) would you require to check whether any of the predictors (and interaction) influences recall scores with a 90% chance?

Because you do not know the effect size, assume Cohen's guideline for linear regression and, to be on the safe side, consider the "Small" value.
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
k <- 3
f2 <- 0.02
pwr.f2.test(u = k, f2 = f2, sig.level = 0.05, power = 0.9)
```

The required sample size is $n = \text v + k + 1 = 709 + 3 + 1 = 713$.

`r solend()`

<br>

`r qbegin(4)`

Using the same $\alpha$ (.05) and power, what would be the sample size if you assumed effect size to be medium?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
k <- 3
f2 <- 0.15
pwr.f2.test(u = k, f2 = f2, sig.level = 0.05, power = 0.9)
```

The required sample size is $n = \text v + k + 1 = 95 + 3 + 1 = 99$ or $100$ (to be on the safe side).
`r solend()`

<br>

`r qbegin(5)`

Using the same $\alpha$ and power, what would be the sample size if you assumed effect size to be large?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
k <- 3
f2 <- 0.35
pwr.f2.test(u = k, f2 = f2, sig.level = 0.05, power = 0.9)
```

The required sample size is $n = \text v + k + 1 = 41 + 3 + 1 = 45$.

`r solend()`

<br>

`r qbegin(6)`
Fit the following model using `lm()`, and assign it as an object with the name "recall_mdl1".

$$
\text{Recall} = \beta_0 + \beta_1 \cdot Age  + \epsilon \\
$$
What is the R-squared?
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
recall_mdl1 <- summary(lm(perc_recall ~ age, data = recdata))
recall_mdl1

recall_mdl1$r.squared
recall_mdl1$adj.r.squared
```

We can use either, for multiple linear regression, you might want to use the adjusted R-squared: $R^2 = 0.16$.
`r solend()`

<br>

`r qbegin(7)`
Use a scatterplot to visualise the assocation between recall and age by group. Is there any evidence of an interaction between age and group?

:::{.callout-tip appearance="simple" collapse="true"}

### Hint

consider using `geom_smooth()` to superimpose the best-fitting line describing the association of interest for each intervention group. 
:::

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r, message = FALSE}
#| label: fig-recall-desc2
#| fig-cap: "Scatterplot displaying the association between age, intervention group, and recall"
recall_plt2 <- ggplot(data = recdata, aes(x = age, y = perc_recall, color = group)) +
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) + 
    labs(x = "Age (in years)", y = "Recall (%)", title = "Scatterplot displaying the association between age, intervention group, and recall")
recall_plt2
```

The slope appears to be stepper in the roller coaster intervention group than the meditation group, so we should investigate whether the interaction is significant.

`r solend()`

<br>

`r qbegin(8)`
Imagine you found the $R$-squared that you computed above (Q6) in a paper, and you are using that to base your next study.

A researcher believes that the inclusion of intervention group and its interaction with age should explain an extra 50% of the variation in recall scores.

Using a significance level of 5%, what sample size should you use for your next data collection in order to discover that effect with a power of 0.9?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
# smaller model m
k <- 1
R2m <- 0.16

# larger model M
K <- 3
R2M <- 0.16 + 0.5

# effect size
f2 <- (R2M - R2m) / (1 - R2M)
pwr.f2.test(u = K - k, f2 = f2, sig.level = 0.05, power = 0.9)
```

The sample size should be $n = \text v + K + 1 = 10 + 3 + 1 = 14$.

With such a big effect size, don't be surprised it's so small. When the effect size is much smaller, that will be harder to detect and you will require a bigger sample size.

`r solend()`

<br>

`r qbegin(9)`

Suppose that the addition of intervention group and its interaction with age, to the $R$-squared from the literature, was thought to explain only 5% more of the variability in recall scores.

Using a significance level of 5%, what sample size should you use for your next data collection in order to discover that effect with a power of 0.9?

`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
# smaller model m
k <- 1
R2m <- 0.16

# larger model M
K <- 3
R2M <- 0.16 + 0.05

# effect size
f2 <- (R2M - R2m) / (1 - R2M)
pwr.f2.test(u = K - k, f2 = f2, sig.level = 0.05, power = 0.9)
```

The sample size should be $n = \text v + K + 1 = 200 + 3 + 1 = 204$.

With such a small effect size, we need a bigger sample size for us to detect it with high confidence.

`r solend()`
