--- 
title: "Data Analysis for Psychology in R 2"
author: "Department of Psychology, The University of Edinburgh"
date: "Academic year 2020-2021"
site: bookdown::bookdown_site
documentclass: book
bibliography: [references.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "DAPR2 is a 2nd year course in data analysis for psychological research."
---


```{r book-setup, include=FALSE}
source("dropdowns.R")


# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')


knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  fig.align = 'center',
  out.width = '60%'
)
```



# Welcome {-}

Data Analysis for Psychology in R 2 (DAPR2) is a 2nd year course in data analysis for psychological research, using the R software. 

This course will introduce you to statistical modelling beyond what is typically taught in an introductory statistics course, empowering you with tools to analyse richer data and answer a broader set of research questions of interest to you.


<!-- ## Course team {-} -->

<!-- - [Dr Tom Booth](https://www.ed.ac.uk/profile/tom-booth) / [email me!](mailto:Tom.Booth@ed.ac.uk) -->
<!-- - [Dr Alex Doumas](https://www.ed.ac.uk/profile/leonidas-doumas) / [email me!](mailto:Alex.Doumas@ed.ac.uk) -->
<!-- - [Dr Josiah King](https://www.ed.ac.uk/profile/josiah-pj-king) / [email me!](mailto:ug.ppls.stats@ed.ac.uk?subject=DAPR2 - ) -->
<!-- - [Dr Umberto Noe](https://www.ed.ac.uk/profile/umberto-noe) / [email me!](mailto:ug.ppls.stats@ed.ac.uk?subject=DAPR2 - ) -->





## Getting started {-}

Do you already have the R software and RStudio installed on your pc? If yes, proceed to the section [Installing packages in R](#installing-packages-in-r). If not, please continue reading.


### Installing R {-}

R is one of the most commonly used programming language in statistics and data science.
It is open-source, free to use, and it comes with a large collection of user-contributed additional tools.

To install R, please follow the link corresponding to your operating system:

**Windows**

Download and install R:
https://cran.r-project.org/bin/windows/base/old/3.6.3/R-3.6.3-win.exe

**macOS**

1. Download and install XQuartz:
https://dl.bintray.com/xquartz/downloads/XQuartz-2.7.11.dmg

2. Download and install R:
https://cran.r-project.org/bin/macosx/R-3.6.3.nn.pkg

**Chromebook**

You can not install R on a chromebook.
Proceed to the next section.



### Installing RStudio {-}

RStudio is a fancy wrapper around R that makes it easy to interact with the R programming language.
It integrates a text editor, a file explorer, an help panel, and an R console.

**Windows and macOS**

Follow this link, download and install the recommended version for your system:
https://rstudio.com/products/rstudio/download/#download

**Chromebook**

You can not install RStudio on a chromebook.

Instead, you should use a cloud-based version of RStudio available here: https://rstudio.cloud/

You will need to sign up and create an account for a **Cloud Free** plan.



## Installing packages in R {-}

As real research involves big data, you would not be able to do the required calculations by hand.
Instead, you will be using the R statistical software to perform those calculations.

When you install R, the software comes with a set of functions that you can use out of the box, such as `c()`, `sample()`, `mean()`, `median()`, `sd()`, and many more.

However, R is so popular also because of the many user-contributed packages that extend the basic R functionality.
These packages need to be installed once for all, in order to extend the functionality of your local R installation, and then can be used in a specific session by loading that package.

For example, a group of users created a package called `tidyverse`, which includes functions such as `tibble()`, `summarise()`, `ggplot()`.
To install the package once for all, you need to type the following command in the R console:
```{r eval=FALSE}
install.packages("tidyverse")
```

Then, if you want to use a specific function from that package in your work, for example the `tibble()` function, you need to type the command below in order to make the functions from that package available in your current session:
```{r eval=FALSE}
library(tidyverse)
```

Remember the main difference between `install.package()` and `library()`.
You only need to install a package once for all, while you need to call `library()` in every new R session in which you intend to use that package.

```{r echo=FALSE, out.width='90%', fig.cap='Source: https://twitter.com/visnut/status/1248087845589274624?s=20'}
knitr::include_graphics('https://pbs.twimg.com/media/EVIZRvPU0AAwgdH?format=jpg&name=900x900')
```

When you will be running code from this book, if you will obtain the following error message:
```
library(<PACKAGE NAME>)
Error in library(<PACKAGE NAME>) : there is no package called ‘<PACKAGE NAME>’
```
this simply means that <PACKAGE NAME> is not installed on your machine.
Simply type `install.packages("<PACKAGE NAME>")`, wait for the installation to be complete, and try again running `library(<PACKAGE NAME>)`.





## Good practices {-}

The following instructions will help you keep a tidy workspace.

1. Create an R project named "DAPR2" for this course. 
Open RStudio, click File, click New Project, click New Directory, click New project, and under Directory name write "DAPR2".

1. In the DAPR2 project folder, create a subfolder called "data", in which you will download the data files which will be asked to download.

1. Each week, create a new Rmarkdown file in which you will answer the lab's questions. Remember to give the file a meaningful name such as "lab01.Rmd".



