---
title: "Sensitivity analysis"
params: 
    TOGGLE: TRUE
---

```{r setup, include=FALSE}
source('../assets/setup.R')
library(tidyverse)
library(sjPlot)
```

```{r}
mwdata <- read_csv('https://uoepsy.github.io/data/wellbeing_rural.csv')
```


A sensitivity analysis allows us to assess the **sensitivity** of our results (i.e., parameter estimates, p-values, confidence intervals) to changes in our modelling approach (here, the removal of observations).

We can re-fit our model after excluding potentially influential observations, and compare these results to the original model.

::: {.callout-important icon=false collapse=false}

# Remove all influential points at once, or one by one?

The current example involves removing all potentially influential observations at the same time.
**Ideally, and to ensure a more thorough sensitivity analysis, you would remove each of these observations one at a time,** assess the effects on the model by comparing to your original, reassessing the remaining pre-identified observations, and repeating the process if necessary. 

:::

::: {.panel-tabset}

### Original Model

```{r}
## wellbeing model
wb_mdl1 <- lm(wellbeing ~ outdoor_time + social_int, data = mwdata) 
summary(wb_mdl1)
```


### Model with Observations Removed

```{r}
## wellbeing model
wb_mdl2 <- lm(wellbeing ~ outdoor_time + social_int, data = mwdata[-c(16, 25, 50, 53, 56, 58, 59, 60, 62, 72, 73, 75, 76, 78, 79, 85, 101, 109, 125, 126, 127, 131, 149, 151, 159, 163, 165, 169, 173, 176, 179, 197), ])
summary(wb_mdl2)
```


### Compare `summary()` output

```{r}
tab_model(wb_mdl1, wb_mdl2,
          dv.labels = c("Wellbeing (WEMWBS Scores)", "Wellbeing (WEMWBS Scores)"),
          pred.labels = c("outdoor_time" = "Outdoor Time (hours per week)",
                          "social_int" = "Social Interactions (number per week)"),
          title = "Regression Table for Wellbeing Models wb1 and wb2")
```

:::

::: {.callout-important icon=false appearance="minimal"}

We conducted a sensitivity analysis to assess how robust our conclusions were regarding outdoor time and the weekly number of social interactions in the presence of previously identified potentially influential observations. We re-fit the model, excluding these 28 observations (14% of our original sample), and compared these model results (`wb_mdl2`) to those of our original model (`wb_mdl1`). 

There was little difference in the estimates from `wb_mdl1` and `wb_mdl2`, and so we can conclude that after conducting a sensitivity analysis, there were no meaningful differences in our results, and hence our conclusions from our original model hold. Specifically:

+ The direction of all model estimates are the same in `wb_mdl1` and `wb_mdl2` (i.e., all positive)
+ There is no difference in statistical significance, and the p-values were of a similar magnitude (i.e., all < .001)
+ The estimate and confidence intervals for `outdoor_time` are very similar
+ There are some quantitative differences in the estimate and confidence intervals for `social_int`. The estimate differs slightly in magnitude by 0.07), but given that this remains positive and significant, we do not need to be too concerned about this. 

:::

