---
title: "11 Playground – Interactions num cat, centering"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(interactions)  # for probe_interactions() plot
```


```{r}
salary <- read_csv("../../data/salary_lec.csv", show_col_types = FALSE) |>
  mutate(
    dept = ifelse(dept == 1, 'Accounts', 'Managers')
  )

# make the acct slope a bit steeper to illustrate the point better
salary <- salary |>
  mutate(
    salary_aug = ifelse(
      dept == 'Accounts',
      salary + (service^2)/1.5 - 10,
      salary
    ),
    salary = salary_aug
  )
```


# Plot data

```{r}
palette_probe <- c('#4ab8fc', '#ff7b01')

salary |>
  ggplot(aes(x = service, y = salary, colour = dept)) +
  geom_point(size = 5) +
  geom_smooth(method = 'lm', se = F) +
  # facet_wrap(~ dept) +
  # theme(legend.position = 'none') +
  scale_color_manual(values = palette_probe) +
  NULL
```




`dept`:

- 1 = store managers
- 0 = accounts


`salary`:

- salary in thosuands of pounds


`service`:

- years of service



# Model: no interaction

```{r}
m1 <- lm(salary ~ service + dept, data = salary)
summary(m1)
```

```{r warning=F}
salary_probedat <- salary |>
  mutate(
    pred = service,
    modx_group = dept
  )
  
plot_m1_probe <- probe_interaction(
  model = m1,
  pred = service,
  modx = dept,
  interval = T
)$interactplot +
  geom_point(data = salary_probedat, size = 5)

plot_data_probe <- plot_m1_probe

plot_data_probe$layers[[1]] <- NULL
plot_data_probe$layers[[1]] <- NULL
```

```{r}
plot_m1_probe
```

- blue line: too high at beginning, too low at end
- orange line: too low at beginning, too high at end

meeting in the middle means that neither line fits the data very well



## plot raw data

```{r}
plot_data_probe
```

## Compute simple slopes

```{r}
coef(m1)
```


Compute simple slopes.

$$
\begin{align}
\widehat{salary} &= \beta_0 + (\beta_1 \cdot service) + (\beta_2 \cdot dept) \\
\widehat{salary} &= 19    + (7.8 \cdot service)     + (-26.2 \cdot dept) \\ 
\end{align}
$$

When $dept = 0$ (Accounts):

$$
\begin{align}
\widehat{salary}_{dept=0} &= 19    + (7.8 \cdot service)     + (-26.2 \cdot 0) \\
\widehat{salary}_{dept=0} &= 19    + (7.8 \cdot service)      \\
\end{align}
$$
When $dept = 1$ (Managers):

$$
\begin{align}
\widehat{salary}_{dept=1} &= 19    + (7.8 \cdot service)     + (-26.2 \cdot 1) \\
\widehat{salary}_{dept=1} &= 19 - 26.2   + (7.8 \cdot service)      \\
\widehat{salary}_{dept=1} &= 9   + (7.8 \cdot service)      \\
\end{align}
$$



# Model: with interaction

```{r}
m2 <- lm(salary ~ service + dept + service:dept, data = salary)
summary(m2)
```


## Compute simple slopes

```{r}
coef(m2)
```


### By hand

$$
\begin{align}
\widehat{salary} &= \beta_0 + (\beta_1 \cdot serv) + (\beta_2 \cdot dept) + (\beta_3 \cdot serv \cdot dept)\\
\widehat{salary} &= -3.4    + (12.3 \cdot serv)     + (20.3 \cdot dept)    + (-9.6 \cdot serv \cdot dept)\\
\end{align}
$$


When $dept = 0$ (Accounts):

$$
\begin{align}
\widehat{salary}_{acct} &= -3.4    + (12.3 \cdot serv)     + (20.3 \cdot dept)    + (-9.6 \cdot serv \cdot dept)\\
\widehat{salary}_{acct} &= -3.4    + (12.3 \cdot serv)     + (20.3 \cdot 0)    + (-9.6 \cdot serv \cdot 0)\\
\widehat{salary}_{acct} &= -3.4    + (12.3 \cdot serv)\\
\end{align}
$$


When $dept = 1$ (Managers):

$$
\begin{align}
\widehat{salary}_{mng} &= -3.4    + (12.3 \cdot serv)     + (20.3 \cdot dept)    + (-9.6 \cdot serv \cdot dept)\\
\widehat{salary}_{mng} &= -3.4    + (12.3 \cdot serv)     + (20.3 \cdot 1)    + (-9.6 \cdot serv \cdot 1)\\
\widehat{salary}_{mng} &= -3.4 + 20.3   + (12.3 \cdot serv) + (-9.6 \cdot serv)\\
\widehat{salary}_{mng} &= 16.9   + (12.3 \cdot serv) + (-9.6 \cdot serv)\\
\widehat{salary}_{mng} &= 16.9   + ((12.3 - 9.6) \cdot serv) \\
\widehat{salary}_{mng} &= 16.9   + (2.7 \cdot serv)\\
\end{align}
$$

### in R

When dept = 0 (acct):

```{r}
coef(m2)[['(Intercept)']]  # intercept
coef(m2)[['service']]      # slope
```


When dept = 1 (mng)

```{r}
coef(m2)[['(Intercept)']] + coef(m2)[['deptManagers']]      # intercept
coef(m2)[['service']] + coef(m2)[['service:deptManagers']]  # slope
```




## Plot asis model

```{r warning = F}
plot_m2_probe <- probe_interaction(
  model = m2,
  pred = service,
  modx = dept,
  interval = T
)$interactplot +
  geom_point(data = salary_probedat, size = 3) +

  NULL

plot_m2_probe
```



### Asis y-intercept plot

```{r}
m2_simple_effs <- tibble(
  dept = c('Accounts', 'Managers'),
  modx_group = c('Accounts', 'Managers'),
  int = c(
    coef(m2)[['(Intercept)']], 
    coef(m2)[['(Intercept)']] + coef(m2)[['deptManagers']]
    ),
  slp = c(
    coef(m2)[['service']],
    coef(m2)[['service']] + coef(m2)[['service:deptManagers']]
    )
)

plot_m2_probe_intercept <- plot_m2_probe
plot_m2_probe_intercept$layers[[1]] <- NULL

plot_m2_probe_intercept +
  xlim(-7, 7) +
  ylim(-10, 100) +
  geom_abline(
    data = m2_simple_effs, 
    aes(intercept = int, slope = slp, colour = dept, linetype = dept),
    linewidth = 1.5
  ) +
  geom_point(
    data = m2_simple_effs,
    x = 0,
    aes(y = int),
    size = 5,
    shape = 15
  ) +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  labs(
    subtitle = 'service as-is'
  ) +
  theme(
    legend.position = 'bottom'
  ) +
  NULL
```



# Min-shift predictor

Plot predictor as-is.

```{r}
salary |>
  ggplot(aes(x = service, y = salary)) +
  geom_vline(xintercept = 0, colour = 'red', linewidth = 2) +
  geom_point() +
  NULL
```


Plot min-shifted predictor.

```{r}
salary <- salary |>
  mutate(
    service_min = service - min(service)
  )

salary |>
  ggplot(aes(x = service_min, y = salary)) +
  geom_vline(xintercept = 0, colour = 'red', linewidth = 2) +
  geom_point() +
  NULL
```

0 is meaningful now = the minimum length of service.


## Model

```{r}
m3 <- lm(salary ~ service_min + dept + service_min:dept, data = salary)
summary(m3)
```

## Compute simple slopes

### In R

When dept = 0 (acct):

```{r}
coef(m3)[['(Intercept)']]  # intercept
coef(m3)[['service_min']]      # slope
```


When dept = 1 (mng)

```{r}
coef(m3)[['(Intercept)']] + coef(m3)[['deptManagers']]      # intercept
coef(m3)[['service_min']] + coef(m3)[['service_min:deptManagers']]  # slope
```




## Plot shifted model

```{r warning=F}
salary_probedat <- salary_probedat |>
  mutate(
    service_min = service - min(service)
  )

plot_m3_probe <- probe_interaction(
  model = m3,
  pred = service_min,
  modx = dept,
  interval = T
)$interactplot +
  geom_point(data = salary_probedat, size = 3)

plot_m3_probe
```

### Shifted y-intercept plot

```{r}
m3_simple_effs <- tibble(
  dept = c('Accounts', 'Managers'),
  modx_group = c('Accounts', 'Managers'),
  int = c(
    coef(m3)[['(Intercept)']], 
    coef(m3)[['(Intercept)']] + coef(m3)[['deptManagers']]
    ),
  slp = c(
    coef(m3)[['service_min']],
    coef(m3)[['service_min']] + coef(m3)[['service_min:deptManagers']]
    )
)

plot_m3_probe_intercept <- plot_m3_probe
plot_m3_probe_intercept$layers[[1]] <- NULL

plot_m3_probe_intercept +
  xlim(-7, 7) +
    ylim(-10, 100) +
  geom_abline(
    data = m3_simple_effs, 
    aes(intercept = int, slope = slp, colour = dept, linetype = dept),
    linewidth = 1.5
  ) +
  geom_point(
    data = m3_simple_effs,
    x = 0,
    aes(y = int),
    size = 5,
    shape = 15
  ) +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  labs(
    subtitle = 'service min-shifted'
  ) +
  theme(
    legend.position = 'bottom'
  ) +
  NULL
```




# Centre predictor

Plot centered predictor.

```{r}
salary <- salary |>
  mutate(
    service_c = scale(service, center = TRUE, scale = FALSE)
  )

salary |>
  ggplot(aes(x = service_c, y = salary)) +
  geom_vline(xintercept = 0, colour = 'red', linewidth = 2) +
  geom_point() +
  NULL
```

```{r}
round(mean(salary$service_c))
```


## Model

```{r}
m4 <- lm(salary ~ service_c + dept + service_c:dept, data = salary)
summary(m4)
```


## Plot centered model

```{r warning = F}
salary_probedat <- salary_probedat |>
  mutate(
    service_c = scale(service, center = TRUE, scale = FALSE)
  )

plot_m4_probe <- probe_interaction(
  model = m4,
  pred = service_c,
  modx = dept,
  interval = T
)$interactplot +
  geom_point(data = salary_probedat, size = 3)
plot_m4_probe
```


### Centered y-intercept plot

```{r}
m4_simple_effs <- tibble(
  dept = c('Accounts', 'Managers'),
  modx_group = c('Accounts', 'Managers'),
  int = c(
    coef(m4)[['(Intercept)']], 
    coef(m4)[['(Intercept)']] + coef(m4)[['deptManagers']]
    ),
  slp = c(
    coef(m4)[['service_c']],
    coef(m4)[['service_c']] + coef(m4)[['service_c:deptManagers']]
    )
)

plot_m4_probe_intercept <- plot_m4_probe
plot_m4_probe_intercept$layers[[1]] <- NULL

plot_m4_probe_intercept +
  xlim(-7, 7) +
    ylim(-10, 100) +
  geom_abline(
    data = m4_simple_effs, 
    aes(intercept = int, slope = slp, colour = dept, linetype = dept),
    linewidth = 1.5
  ) +
  geom_point(
    data = m4_simple_effs,
    x = 0,
    aes(y = int),
    size = 5,
    shape = 15
  ) +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  labs(
    subtitle = 'service mean-centered'
  ) +
  theme(
    legend.position = 'bottom'
  ) +
  NULL
```


# Compare coef estims

## As-is

With `service` as-is:

```{r}
summary(m2)$coefficients |> round(3)
```


## Min-shifted

With `service` min-shifted:

```{r}
summary(m3)$coefficients |> round(3)
```


## Centered

With `service` mean-centered:

```{r}
summary(m4)$coefficients |> round(3)
```


# Appendix

- all simple slope calculations for all models




