---
title: "14 Playground – Interactions cat cat effect + manual coded AND ALSO pairwise comparisons"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(emmeans)
library(interactions)  # for probe_interactions() plot

hosp <- read_csv("../../data/hospital.csv") |>
  mutate(
    Treatment = factor(Treatment),
    Hospital = factor(Hospital),
  )

# get group means
mean_A1 <- mean(filter(hosp, Treatment == 'TreatA', Hospital == 'Hosp1')$SWB)
mean_B1 <- mean(filter(hosp, Treatment == 'TreatB', Hospital == 'Hosp1')$SWB)
mean_C1 <- mean(filter(hosp, Treatment == 'TreatC', Hospital == 'Hosp1')$SWB)
mean_A2 <- mean(filter(hosp, Treatment == 'TreatA', Hospital == 'Hosp2')$SWB)
mean_B2 <- mean(filter(hosp, Treatment == 'TreatB', Hospital == 'Hosp2')$SWB)
mean_C2 <- mean(filter(hosp, Treatment == 'TreatC', Hospital == 'Hosp2')$SWB)
```

# Plot the data

```{r}
hosp |>
  ggplot(aes(x = Treatment, y = SWB, colour = Treatment, fill = Treatment)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  facet_wrap(~ Hospital)
```


# Sum coding

```{r}
contrasts(hosp$Treatment) <- contr.sum(3)
contrasts(hosp$Hospital) <- contr.sum(2)
```

```{r}
contrasts(hosp$Treatment)
```


```{r}
contrasts(hosp$Hospital)
```



```{r}
m1 <- lm(SWB ~ Treatment*Hospital, data = hosp)
summary(m1)
```


# Post-hoc interactions using emmeans

- Treatment A vs. (Treatment B, Treatment C)
- is there a diff between those two chunks, across hospitals?

- basically kind of compresses data into a 2x2 design


## Plot the contrast we want to test

```{r}
hosp |>
  mutate(
    TreatmentPosthoc = ifelse(
      Treatment == 'TreatA',
      'TreatA',
      'TreatBC'
    )
  ) |>
  ggplot(aes(x = TreatmentPosthoc, y = SWB, colour = TreatmentPosthoc, fill = TreatmentPosthoc)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  facet_wrap(~ Hospital) +
  stat_summary(geom = 'pointrange', colour = 'black') +
  NULL
```


## Get estimated marginal means

Create manual contrasts for each of the things we want to compare.

Step by step from before, end up with

```{r}
treat_contrast <- c(
  'TreatA' = 1,
  'TreatB' = -1/2,
  'TreatC' = -1/2
)

hosp_contrast <- c(
  'Hosp1' = 1,
  'Hosp2' = -1
)
```

Get outer product (actually show table and fill in cell by cell)

```{r}
ixn_contrast <- outer(treat_contrast, hosp_contrast)
ixn_contrast
```


```{r}
m1_emm <- emmeans(m1, ~Treatment * Hospital)
m1_emm
```

Before we used `levels()` to get the order, because we were only looking at one variable.
Now we've got interplay between two variables, so we'll just base things on the order they appear in the emmeans table here.

```{r}
ixn_comp <- list(
  "A vs. BC in diff Hospitals" = 
    c(
      1,     # TreatA Hosp1 
      -1/2,  # TreatB Hosp1
      -1/2,  # TreatC Hosp1
      -1,    # TreatA Hosp2
      1/2,   # TreatB Hosp2
      1/2    # TreatC Hosp2
    )
)
```


Test the contrast using `contrast()` from emmeans.

```{r}
ixn_contrast_test <- contrast(m1_emm, ixn_comp)
ixn_contrast_test
```

We can reject the H0 that there's no difference between hospitals in the effectiveness of A vs. BC.
Symmetrically, we can reject the H0 that the relationship between A vs. BC is the same in both hospitals.


Where does this number 3.73 come from?

It's the sum of all of the group means, each multiplied by the coding from our matrix.

```{r}
ixn_contrast

1 * mean_A1 +
  -1   * mean_A2 +
  -0.5 * mean_B1 +
   0.5 * mean_B2 +
  -0.5 * mean_C1 +
   0.5 * mean_C2
```

It's positive, so that means that the means that received positive weights are bigger than the means that received negative weights.


