---
title: "<b>Missing Data </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

# Weeks Learning Objectives

+ Develop awareness of the glm for modelling non-continuous outcomes

+ Understand the types of missing data and common causes.

+ Develop awareness of approaches to dealing with missing data.

---
# Topics for today
- Conceptual introduction to missing data.

- Mechanisms of missingness

- Main methods you will read about in the literature.

---
# Missing Data
- Missing data is very common

- We worry about two main things with missing data:
  - Loss of efficiency (due to smaller sample size)
  - Bias (i.e., incorrect estimates)

- The effects of missing data depend on a combination of the missing data mechanism and how we deal with the missing data


---
# Missing Data Mechanisms
- Missing data can be of several types:
  - Missing completely at random (MCAR)
  - Missing at random (MAR)
  - Missing not at random (MNAR)

---
# MAR
- Missing at random (MAR) means:

**When the probability of missing data on a variable $Y$ is related to other variables in the model but not to the values of $Y$ itself**

- Example:
  - $X$ = self-control and $Y$ = aggression.
  - People with lower self-control are more likely to have missing data on aggression.
  - After taking into account self-control, people who are high in aggression are no more likely to have missing data on aggression.

- Challenge is that there is no way to confirm that there is no relation between aggression scores and missing data on aggression because that would knowledge of the missing scores.


---
# MCAR
- Missing completely at random (MCAR) means:

**Genuinely random missingness. No relation between $Y$ or any other variable in the model and missingness on $Y$**

- The data you have are a simple random sample of the complete data.
  - The ideal missing data scenario!

- Example:
  - $X$ = self-control and $Y$= aggression
  - People of all levels of self-control and aggression are equally likely to have missing data on aggression

---
# MNAR
- Missing not random (MNAR) means:

**When the probability of missingness on $Y$ is related to the values of $Y$ itself**

- Example:
  - $X$ = self-control, $Y$ = aggression
  - Those high in aggression are more likely to have missing data on the aggression variable, even after taking into account self-control.
  
- As with MAR, there is no way to verify that data are MNAR without knowledge of the missing values

---
# Methods for Missing Data
- Deletion methods:
  - Listwise deletion
  - Pairwise deletion

- Imputation methods:
  - Mean imputation
  - Regression imputation
  - Multiple imputation

- Maximum likelihood estimation

- Methods for MNAR data:
  - Pattern mixture models
  - Random coefficient models

---
# Deletion Methods
- Listwise deletion or complete case analysis
  - Delete everyone from the analysis who has missing data on either self-control or aggression

- Will give biased results unless data are MCAR

- Even if data are MCAR, power will be reduced by reducing the sample size

- Bottom line: **not recommended**

---
# Deletion Methods
- Pairwise deletion or available-case analysis
  - Uses available data for each analysis
  - Different cases contribute to different correlations in a correlation matrix

- Example:
  - Cases 2,3,7,18, 56, 100 not used in the self-control- aggression correlation
  - Cases 2,7,18,77, 103 not used in the aggression-substance use correlation

- Doesn’t reduce power as much as listwise deletion

- But still gives biased results whenever data are not MCAR

- Bottom line: **not recommended**

---
# Mean Imputation
- Replace missing values with the mean on that variable

- Two major issues:
  - artificially reduces the variability of the data
  - can give very biased estimates even when data are MCAR

- Bottom line: **not recommended**

> ‘possibly the worst missing data handling method available… you should absolutely avoid this approach’ Enders (2011)

---
# Regression Imputation
- Replaces the missing values with values predicted from a regression
  - Estimate a set of regression equations where the incomplete variables are predicted from the complete variables
  - Use the regression equations to calculate the predicted values on the incomplete variables

- Based on the principle of using information from the complete data to estimate the missing data

- Two forms:
 - ‘normal’ regression imputation 
  - Stochastic regression imputation (adds a residual term to overcome loss of variance)

- Stochastic regression is preferred and gives unbiased results if data are MAR

---
# Multiple Imputation
- Procedure:
  - Imputes missing data several times to create multiple complete datasets
  - Analyses are conducted for each dataset
  - Analysis results are pooled across datasets to get parameter estimates and standard errors
  - 3-5 datasets are often enough but more is better and 20+ is ideal

- Unlike in most single imputation approaches, the standard errors take account of the additional uncertainty due to missingness

- Gives unbiased parameter estimates under MAR

- Can be tedious if pooling has to be done by hand

-Bottom line: **recommended method if data are likely to be MAR**

- If using MI, include as many variables and higher-order effects as possible in the imputation phase

---
# Maximum Likelihood Estimation
- Estimation method 
  - Makes use of all the information in the model to arrive at the parameter estimates (e.g. regression coefficient) ‘as if’ the data were complete

- Does not ‘impute’ individual values

- Gives unbiased estimates under MAR

- Assumes multivariate normality

- Even under MCAR it is superior to listwise and pairwise deletion because it uses more information from the observed data.

- Practical advantage= easy to implement, usually much more so than multiple imputation

- Bottom line: **recommended method if data are likely to be MAR**

---
# Methods for MNAR
- Selection models:
  - Combines model for predicting missingness as well as the analysis model of interest, e.g.:
  - Selection model= predict missingness on aggression from covariates such as gender, age, ADHD scores
  - Substantive model= predict aggression from self-control

- The parameter estimates of the latter are adjusted by the inclusion of the former

- Makes strong, untestable assumptions

- Under many realistic conditions, gives worse results than MI or MLE, even when the MAR assumption is violated

- Bottom line: **Good to include as part of a sensitivity analysis but often between to use MI or MLE** 

---
# Methods for MNAR
- Pattern mixture models
  - Stratifies the sample according to different missing data patterns
  - E.g. pattern 1= complete data , pattern 2= ‘missing on aggression’ , pattern 3= ‘missing on self-control’
  - Estimates the substantive model in each subgroup
  - Pools the results from each subgroup to get a weighted average parameter estimate

- Similar to selection model, requires strong untestable assumptions

- Violations of the assumptions lead to bias

- Bottom line: **Good to include as part of a sensitivity analysis but often between to use MI or MLE**

---
# What to do?

.pull-left[
- When MLE may be better:
  - When the substantive model includes interactions
  - For structural equation models (more on this in dapR3)
  - For the inexperienced (easier to learn and implement)

]

.pull-right[
- When MI may be better:
  - When a structural equation model has categorical indicators
  - When there is missing data on the predictors
  - When including auxiliary variables

]

---
# Summary of today
- This was intended as a short overview of missing data and the methods you may read about.

- Goal was to make you aware of the issue.

- We will return to FIML approaches in dapR3.
  - Most likely to work effectively for your own practical research.
  
