---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)

set.seed(123)
n.sample <- 200
x1 <- rbinom(n.sample, size = 1, prob = 0.5)
x2 <- runif(n.sample, -5, 5)
a <- 5
b1 <- 3
b2 <- 4
b3 <- -3
e <- rnorm(n.sample, 0, 5)
y <- a + b1 * x1 + b2 * x2 + b3 * x1 * x2 + e
dat1 <- data.frame(y, x1, x2)


set.seed(711)
n.sample <- 200
x1 <- rbinom(n.sample, size = 1, prob = 0.6)
x2 <- runif(n.sample, -3, 3)
a <- 3
b1 <- 6
b2 <- 2
b3 <- 3
e <- rnorm(n.sample, 0, 5)
y <- a + b1 * x1 + b2 * x2 + b3 * x1 * x2 + e
dat2 <- data.frame(y, x1, x2)

```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

.pull-left[
Consider the interaction plot on the right hand side. It shows the interaction between a binary predictor ( $x_1$ ) and a continuous predictor ( $x_2$ ). The full model is:

$$y_i = b_0 + b_1x_1 + b_2x_2 + b_3x_1x_2 + \epsilon_1$$

Answer the following questions:

1. Will $b_1$ be positive or negative?

2. Will $b_3$ be positive or negative?
]

.pull-right[

```{r, warning=FALSE, echo=FALSE, message=FALSE}
dat1 %>%
  ggplot(., aes(x=x2, y=y, colour = factor(x1))) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)+
  scale_colour_discrete(name  ="Binary X1")
```

]

---
# Question 1: Answer

.pull-left[

```{r echo=FALSE}
res1 <- summary(lm(y~x1+x2+x1*x2, data=dat1))
res1$coefficients
```

+ $b_1$ = difference between groups 0/1 when `x2` = 0. 
  + If we read along the x-axis, we see the blue line (`x1` = 1) is higher.
  + So the mean of this group is bigger, and the coefficient positive

+ $b_3$ = the interaction, is the difference in slopes
  + Blue line is flatter
  + so the slope is less when `x1` = 1.
  + So $b_3$ is negative

]

.pull-right[
```{r, warning=FALSE, echo=FALSE, message=FALSE}
dat1 %>%
  ggplot(., aes(x=x2, y=y, colour = factor(x1))) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)+
  scale_colour_discrete(name  ="Binary X1")
```
]

---
# Question 2

.pull-left[
Consider the interaction plot on the right hand side. It shows the interaction between a binary predictor ( $x_1$ ) and a continuous predictor ( $x_2$ ). The full model is:

$$y_i = b_0 + b_1x_1 + b_2x_2 + b_3x_1x_2 + \epsilon_1$$

Answer the following questions:

1. Will $b_1$ be positive or negative?

2. Will $b_3$ be positive or negative?
]

.pull-right[

```{r, warning=FALSE, echo=FALSE, message=FALSE}
dat2 %>%
  ggplot(., aes(x=x2, y=y, colour = factor(x1))) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)+
  scale_colour_discrete(name  ="Binary X1")
```

]

---
# Question 2: Answer

.pull-left[

```{r echo=FALSE}
res1 <- summary(lm(y~x1+x2+x1*x2, data=dat2))
res1$coefficients
```

+ $b_1$ = difference between groups 0/1 when `x2` = 0. 
  + If we read along the x-axis, we see the blue line (`x1` = 1) is higher.
  + So the mean of this group is bigger, and the coefficient positive

+ $b_3$ = the interaction, is the difference in slopes
  + Blue line is steeper
  + so the slope is greater when `x1` = 1.
  + So $b_3$ is positive

]

.pull-right[

```{r, warning=FALSE, echo=FALSE, message=FALSE}
dat2 %>%
  ggplot(., aes(x=x2, y=y, colour = factor(x1))) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)+
  scale_colour_discrete(name  ="Binary X1")
```

]

---
# Question 3
+ How many lines will be on the plot if the categorical predictor had 4 levels?

---
# Question 3: Answer

+ 4
+ We can have a line per group.

---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 