---
title: "13 Playground – Interactions cat cat treatment coded"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(interactions)   
```

# data set-up

```{r eval=F}
set.seed(1)
anx <- read_csv("https://uoepsy.github.io/data/lmm_ef_sdmt.csv") |>
  slice_sample(prop = .6) |>
  rename(
   anx = SDMT,
  ) |>
  mutate(
    transport = ifelse(headphones == 'H', 'Cyclist', 'Motorist'),
    road_type  = case_when(
      audio == 'music' ~ 'RuralRoad',
      audio == 'no_audio' ~ 'CityStreet',
      audio == 'white_noise' ~ 'DualCarriageway'
    )
  ) |>
  select(anx, transport, road_type) |>
  rownames_to_column(var = 'ppt_id')

# write_csv(anx, '../../data/post_travel_anx.csv')
# write_csv(anx, '../data/post_travel_anx.csv')
```

```{r}
anx <- read_csv('../../data/post_travel_anx.csv')
```

```{r}
anx |> 
  ggplot(aes(x = road_type, y = anx)) +
  geom_violin() +
  stat_summary(geom = 'pointrange') +
  facet_wrap(~ transport)
```


# 2x2 data treatment coded

```{r}
anx1 <- filter(
  anx,
  road_type %in% c('CityStreet', 'RuralRoad')
)
```


## violin plots (symmetrical)

```{r}
anx1 |>
  ggplot(aes(x = road_type, y = anx, colour = road_type, fill = road_type)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, size = 5) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 8) +
  facet_wrap(~ transport) +
  theme(legend.position = 'none') +
  NULL
```

```{r}
anx1 |>
  ggplot(aes(x = transport, y = anx, colour = transport, fill = transport)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(alpha = 0.5, size = 5) +
  stat_summary(fun = mean, geom = 'point', colour = 'black', size = 8) +
  facet_wrap(~ road_type) +
  theme(legend.position = 'none') +
  NULL
```


TODO: add diff of diffs into plot as vertical line between means, at x = 1.5 in each facet.



## pointrange plots

```{r}
anx1 |>
  ggplot(aes(x = road_type, y = anx, colour = transport)) +
  stat_summary(geom = 'pointrange', position = position_dodge(width = .5)) +
  NULL
```

```{r}
anx1 |>
  ggplot(aes(x = transport, y = anx, colour = road_type)) +
  stat_summary(geom = 'pointrange', position = position_dodge(width = .5)) +
  NULL
```


## set up data

Set ref levels to the group that has the lower scores, so we can measure how anx increases in diff scenarios.

```{r}
anx1 <- anx1 |>
  mutate(
    transport = factor(transport, levels = c('Motorist', 'Cyclist')),
    road_type = factor(road_type, levels = c('RuralRoad', 'CityStreet'))
  )
```

```{r}
contrasts(anx1$transport)
```

```{r}
contrasts(anx1$road_type)
```


### contrast coding the interaction


| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Cyclist  | Rural |
| Cyclist  | City  |


| transport | road_type |
|---|---|
| 0 | 0 |
| 0 | 1 |
| 1 | 0 |
| 1 | 1 |


| transport | road_type | transport:road_type |
|---|---|---|
| 0 | 0 | 0 |
| 0 | 1 | 0 |
| 1 | 0 | 0 |
| 1 | 1 | 1 |



## fit the model

$$
\text{anx} = \beta_0 + 
(\beta_1 \cdot \text{transport}) + 
(\beta_2 \cdot \text{road_type}) + 
(\beta_3 \cdot \text{transport} \cdot \text{road_type}) + 
\epsilon
$$

```{r}
m1 <- lm(anx ~ transport * road_type, data = anx1)
summary(m1)
```

## cat_plots

```{r}
cat_plot(
  m1,
  pred = transport,
  modx = road_type,
  # geom = 'line'
)
```

```{r}
cat_plot(
  m1,
  pred = road_type,
  modx = transport,
  # geom = 'line'
)
```





## simple effects by hand

(simple slopes = continuous, simple effects = categorical)

```{r}
coef(m1) |> round(2)
```

$$
\text{anx} = \beta_0 + 
(\beta_1 \cdot \text{transport}) + 
(\beta_2 \cdot \text{road_type}) + 
(\beta_3 \cdot \text{transport} \cdot \text{road_type}) + 
\epsilon \\

\text{anx} = 27.03 +  (0.44 \cdot \text{transport}) +  (4.89 \cdot \text{road_type}) +  (7.76 \cdot \text{transport} \cdot \text{road_type}) +  \epsilon \\
$$


### RuralRoad

$$
\text{anx}_{\text{Rural}} = 27.03 +  (0.44 \cdot \text{transport}) +  (4.89 \cdot 0) +  (7.76 \cdot \text{transport} \cdot 0) +  \epsilon \\
\text{anx}_{\text{Rural}} = 27.03 +  (0.44 \cdot \text{transport}) + \epsilon \\
$$

### CityStreet

$$
\text{anx}_{\text{City}} = 27.03 +  (0.44 \cdot \text{transport}) +  (4.89 \cdot 1) +  (7.76 \cdot \text{transport} \cdot 1) +  \epsilon \\
\text{anx}_{\text{City}} = 27.03 + 4.89 +  (0.44 \cdot \text{transport}) + (7.76 \cdot \text{transport}) +  \epsilon \\
\text{anx}_{\text{City}} = 31.92 +  ((0.44 + 7.76) \cdot \text{transport}) +  \epsilon \\
\text{anx}_{\text{City}} = 31.92 +  (8.2 \cdot \text{transport}) +  \epsilon \\
$$


### Motorist

$$
\text{anx}_{\text{Motorist}} = 27.03 +  (0.44 \cdot 0) +  (4.89 \cdot \text{road_type}) +  (7.76 \cdot 0 \cdot \text{road_type}) +  \epsilon \\
\text{anx}_{\text{Motorist}} = 27.03 + (4.89 \cdot \text{road_type}) + \epsilon \\
$$

### Cyclist

$$
\text{anx}_{\text{Cyclist}} = 27.03 +  (0.44 \cdot 1) +  (4.89 \cdot \text{road_type}) +  (7.76 \cdot 1 \cdot \text{road_type}) +  \epsilon \\
\text{anx}_{\text{Cyclist}} = 27.03 + 0.44 +  (4.89 \cdot \text{road_type}) +  (7.76 \cdot \text{road_type}) +  \epsilon \\
\text{anx}_{\text{Cyclist}} = 27.47 +  ((4.89 + 7.76) \cdot \text{road_type})  +  \epsilon \\
\text{anx}_{\text{Cyclist}} = 27.47 +  (12.65 \cdot \text{road_type})  +  \epsilon \\
$$









# 2x3 data treatment coded

```{r}
anx2 <- anx
```


## set up the data

```{r}
anx2 <- anx2 |>
  mutate(
    transport = factor(transport, levels = c('Motorist', 'Cyclist')),
    road_type = factor(road_type, levels = c('RuralRoad', 'CityStreet', 'DualCarriageway')),
  )
anx2
```

```{r}
contrasts(anx2$transport)
```

```{r}
contrasts(anx2$road_type)
```


### contrast coding CityStreet ixn

| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Motorist | Dual  |
| Cyclist  | Rural |
| Cyclist  | City  |
| Cyclist  | Dual  |


| transport | road_typeCityStreet |
|---|---|
| 0 | 0 |
| 0 | 1 |
| 0 | 0 |
| 1 | 0 |
| 1 | 1 |
| 1 | 0 |


| transport | road_typeCityStreet | transport:road_typeCityStreet |
|---|---| --- |
| 0 | 0 | 0 |
| 0 | 1 | 0 |
| 0 | 0 | 0 |
| 1 | 0 | 0 |
| 1 | 1 | 1 |
| 1 | 0 | 0 |



### contrast coding DualCarriageway ixn

| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Motorist | Dual  |
| Cyclist  | Rural |
| Cyclist  | City  |
| Cyclist  | Dual  |


| transport | road_typeDualCarriageway |
|---|---|
| 0 | 0 |
| 0 | 0 |
| 0 | 1 |
| 1 | 0 |
| 1 | 0 |
| 1 | 1 |


| transport | road_typeDualCarriageway | transport:road_typeDualCarriageway |
|---|---| --- |
| 0 | 0 | 0 |
| 0 | 0 | 0 |
| 0 | 1 | 0 |
| 1 | 0 | 0 |
| 1 | 0 | 0 |
| 1 | 1 | 1 |



## fit the model


$$
\text{anx} = \beta_0 + 
(\beta_1 \cdot \text{transport}) + 
(\beta_2 \cdot \text{road_type}_{\text{CS}}) + 
(\beta_2 \cdot \text{road_type}_{\text{DC}}) +  \\
(\beta_3 \cdot \text{transport} \cdot \text{road_type}_{\text{CS}}) + 
(\beta_3 \cdot \text{transport} \cdot \text{road_type}_{\text{DC}}) + 
\epsilon
$$

```{r}
m2 <- lm(anx ~ transport * road_type, data = anx2)
summary(m2)
```


## cat_plots

```{r}
cat_plot(
  m2,
  pred = transport,
  modx = road_type,
  geom = 'line'
)
```

```{r}
cat_plot(
  m2,
  pred = road_type,
  modx = transport,
  # geom = 'line'
)
```

## simple effects by hand

### RuralRoad

To get RuralRoad, we have $\text{road_type}_{\text{CS}} = 0$ and $\text{road_type}_{\text{DC}} = 0$.
(We know this bc of the contrast coding)

$$
\begin{align}
\text{anx}_{\text{RR}} ~=&~ \beta_0 +  (\beta_1 \cdot \text{transport}) +  (\beta_2 \cdot \text{road_type}_{\text{CS}}) +  (\beta_3 \cdot \text{road_type}_{\text{DC}}) +  
\\ & (\beta_4 \cdot \text{transport} \cdot \text{road_type}_{\text{CS}}) +  (\beta_5 \cdot \text{transport} \cdot \text{road_type}_{\text{DC}}) +  \epsilon \\


\text{anx}_{\text{RR}} ~=&~ \beta_0 +  (\beta_1 \cdot \text{transport}) +  (\beta_2 \cdot 0) +  (\beta_3 \cdot 0) +  
\\ & (\beta_4 \cdot \text{transport} \cdot 0) +  (\beta_5 \cdot \text{transport} \cdot 0) +  \epsilon \\

\text{anx}_{\text{RR}} ~=&~ \beta_0 + (\beta_1 \cdot \text{transport}) + \epsilon \\
\end{align}
$$


### DualCarriageway

To get DualCarriageway, we have $\text{road_type}_{\text{CS}} = 0$ and $\text{road_type}_{\text{DC}} = 1$.
(We know this bc of the contrast coding)

$$
\begin{align}
\text{anx}_{\text{DC}} ~=&~ \beta_0 +  (\beta_1 \cdot \text{transport}) +  (\beta_2 \cdot \text{road_type}_{\text{CS}}) +  (\beta_3 \cdot \text{road_type}_{\text{DC}}) +  
\\ & (\beta_4 \cdot \text{transport} \cdot \text{road_type}_{\text{CS}}) +  (\beta_5 \cdot \text{transport} \cdot \text{road_type}_{\text{DC}}) +  \epsilon \\

\text{anx}_{\text{DC}} ~=&~ \beta_0 +  (\beta_1 \cdot \text{transport}) +  (\beta_2 \cdot 0) +  (\beta_3 \cdot 1) +  
\\ & (\beta_4 \cdot \text{transport} \cdot 0) +  (\beta_5 \cdot \text{transport} \cdot 1) +  \epsilon \\

\text{anx}_{\text{DC}} ~=&~ \beta_0 + \beta_3 + ((\beta_1 + \beta_5) \cdot \text{transport}) +  \epsilon \\

\end{align}
$$


### Motorist

To get Motorist, we set transport = 0.

$$
\begin{align}
\text{anx}_{\text{Motorist}} ~=&~ \beta_0 +  (\beta_1 \cdot \text{transport}) +  (\beta_2 \cdot \text{road_type}_{\text{CS}}) +  (\beta_3 \cdot \text{road_type}_{\text{DC}}) +  
\\ & (\beta_4 \cdot \text{transport} \cdot \text{road_type}_{\text{CS}}) +  (\beta_5 \cdot \text{transport} \cdot \text{road_type}_{\text{DC}}) +  \epsilon \\

\text{anx}_{\text{Motorist}} ~=&~ \beta_0 +  (\beta_1 \cdot 0) +  (\beta_2 \cdot \text{road_type}_{\text{CS}}) +  (\beta_3 \cdot \text{road_type}_{\text{DC}}) +  
\\ & (\beta_4 \cdot 0 \cdot \text{road_type}_{\text{CS}}) +  (\beta_5 \cdot 0 \cdot \text{road_type}_{\text{DC}}) +  \epsilon \\

\text{anx}_{\text{Motorist}} ~=&~ \beta_0 + (\beta_2 \cdot \text{road_type}_{\text{CS}}) +  (\beta_3 \cdot \text{road_type}_{\text{DC}}) + \epsilon \\

\end{align}
$$



# Imagine 2x3x2

What if this pattern of results looks different in different countries?

Might compare Scotland (not that much bike infrastructure) to the Netherlands (lots of bike infrastructure).

Reasonable question but big pain to analyse/interpret.

Only way to really interpret is to visualise.



# 2x2 data sum coded

## contrast coding the interaction

```{r}
contrasts(anx1$transport) <- contr.sum(2)
contrasts(anx1$road_type) <- contr.sum(2)
```

```{r}
contrasts(anx1$transport)
```


```{r}
contrasts(anx1$road_type)
```




| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Cyclist  | Rural |
| Cyclist  | City  |


| transport | road_type |
|---|---|
|  1 |  1 |
|  1 | -1 |
| -1 |  1 |
| -1 | -1 |

| transport | road_type | transport:road_type |
|---|---|
|  1 |  1 |  1 |
|  1 | -1 | -1 |
| -1 |  1 | -1 |
| -1 | -1 |  1 |


## fit the model

```{r}
m3 <- lm(anx ~ transport * road_type, data = anx1)
summary(m3)
```

```{r}
anx1 |>
  group_by(transport, road_type) |>
  summarise(m = mean(anx))
```

Intercept is mean of means.

But not mean of anx straight up, bc groups have diff sizes!

```{r}
mean(anx1$anx)
```


nice ref for interpreting sum contrasts in the flash cards:
https://uoepsy.github.io/dapr2/2425/labs/1_b4_reading.html#categorical-x-categorical-example 


# 2x3 data sum coded

maybe leave this for the appendix

## contrast coding the ixns

```{r}
contrasts(anx2$road_type) <- contr.sum(3)
contrasts(anx2$road_type)
```


### contrast coding road_type1 (Rural = 1)

| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Motorist | Dual  |
| Cyclist  | Rural |
| Cyclist  | City  |
| Cyclist  | Dual  |


| transport | road_type1 |
|---|---|
|  1 |  1 |
|  1 |  0 |
|  1 | -1 |
| -1 |  1 |
| -1 |  0 |
| -1 | -1 |


| transport | road_type1 | transport:road_type1 |
|---|---| --- |
|  1 |  1 |  1 |
|  1 |  0 |  0 |
|  1 | -1 | -1 |
| -1 |  1 | -1 |
| -1 |  0 |  0 |
| -1 | -1 |  1 |



### contrast coding road_type2 (City = 1)

| transport | road_type |
|---|---|
| Motorist | Rural |
| Motorist | City  |
| Motorist | Dual  |
| Cyclist  | Rural |
| Cyclist  | City  |
| Cyclist  | Dual  |


| transport | road_type2 |
|---|---|
|  1 |  0 |
|  1 |  1 |
|  1 | -1 |
| -1 |  0 |
| -1 |  1 |
| -1 | -1 |


| transport | road_type2 | transport:road_type2 |
|---|---| --- |
|  1 |  0 |  0 |
|  1 |  1 |  1 |
|  1 | -1 | -1 |
| -1 |  0 |  0 |
| -1 |  1 | -1 |
| -1 | -1 |  1 |



## fit the model

```{r}
m4 <- lm(anx ~ transport * road_type, data = anx2)
summary(m4)
```



