---
title: "Cat*Con Interactions"
author: "Tom Booth"
date: '2022-11-01'
output: html_document
---

This document is to briefly elaborate the difference between two linear models with a continous outcome ( $y$ ), a continuous predictor ( $x$ ) and a binary predictor ( $z$ ) First with no interaction. :

$$y = b_0 + b_1x + b_2z + \epsilon$$

And second with an interaction: 

$$y = b_0 + b_1x + b_2z + b_3xz + \epsilon$$

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(simglm)
library(sjPlot)
library(interactions)

set.seed(3119) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation + study + method,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1),
               study = list(var_type = 'factor', 
                            levels = c('alone', 'others'),
                            prob = c(0.53, 0.47)),
               method = list(var_type = 'factor', 
                            levels = c('read', 'summarise', 'self-test'),
                            prob = c(0.3, 0.4, 0.3))),
  error = list(variance = 20),
  sample_size = 250,
  reg_weights = c(0.6, 1.4, 1.5, 6, 6, 2)
)

df3 <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

data1 <- df3 %>%
  dplyr::select(y, hours, motivation, study, method) %>%
  mutate(
    ID = paste("ID", 101:350, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2),
    study = factor(study),
    method = factor(method)
  ) %>%
  dplyr::select(ID, score, hours, motivation, study, method)
```


```{r}
m1 <- lm(score ~ hours + study, data = data1)
summary(m1)
```

```{r}
tab_model(m1)
```


```{r}
sjPlot::plot_model(m1, type="pred", terms=c("hours","study"))
```


- $b_0$ = `r round(m1$coefficients[1],3)` the point where the red line crosses 0. For someone who studies no hours in the `alone` group, expected test score is 12.18
- $b_1$ = `r round(m1$coefficients[2],3)` the slope of both lines. For each additional hour of study, score increases by 1.28 points
- $b_2$ = `r round(m1$coefficients[3],3)` the point where the blue line crosses 0. The difference in the intercepts between study `alone` and with `others`. Those who study with others and have 0 hours of revision have an expected test score of `r round(m1$coefficients[1],3) + round(m1$coefficients[3],3)` (`r round(m1$coefficients[3],3)` higher than $b_0$ )


```{r, echo=FALSE, warning=FALSE}
set.seed(3119) 

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation + study + method + hours:study,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1),
               study = list(var_type = 'factor', 
                            levels = c('alone', 'others'),
                            prob = c(0.53, 0.47)),
               method = list(var_type = 'factor', 
                            levels = c('read', 'summarise', 'self-test'),
                            prob = c(0.3, 0.4, 0.3))),
  error = list(variance = 20),
  sample_size = 250,
  reg_weights = c(0, 1.4, 1.5, 6, 6, 2, 2)
)

df3 <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

data2 <- df3 %>%
  dplyr::select(y, hours, motivation, study, method) %>%
  mutate(
    ID = paste("ID", 101:350, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2),
    study = factor(study),
    method = factor(method)
  ) %>%
  dplyr::select(ID, score, hours, motivation, study, method)
```

```{r}
m2 <- lm(score ~ hours + study + hours*study, data = data2)
summary(m2)
```

```{r}
tab_model(m2)
```


```{r}
interact_plot(m2, 
              pred = hours,
              modx = study)
```


- $b_0$ = `r round(m2$coefficients[1],3)` the point where the red line crosses 0. For someone who studies no hours in the `alone` group, expected test score is 12.51
- $b_1$ = `r round(m2$coefficients[2],3)` the slope for those who study `alone`. For each additional hour of study, score increases by 1.24 points
- $b_2$ = `r round(m2$coefficients[3],3)` the point where the blue line crosses 0. The difference in the intercepts between study `alone` and with `others`. Those who study with others and have 0 hours of revision have an expected test score of `r round(m2$coefficients[1],3) + round(m2$coefficients[3],3)` (`r round(m2$coefficients[3],3)` higher than $b_0$ )
- $b_3$ = `r round(m2$coefficients[4],3)` the difference in slopes between red and blue lines. The difference in effect of hours between study `alone` and with `others`. Those who study with others have a more positive effect of hours studied (more effective) than those who study alone. The slope for others is `r round(m2$coefficients[2],3) + round(m2$coefficients[4],3)` ( $b_1 + b_3$ )
