---
title: "<b> Simple Effects </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR2 Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(emmeans)

hosp_tbl <- read_csv("./hospital.csv", col_types = "dff")

```

# Lecture Objective
1. Distinguish between main effects, simple effects and interactions.
2. Interpret and test simple effects


---
# Hypotheses we test in Factorial Designs
+ Main effects
  + An overall, or average, effect of a condition.
  + Is there an effect of `Treatment` averaged over `Hospital`? 
  + Is there an effect of `Hospital` averaged over `Treatment`? 


+ **Simple contrasts/effects**
  + An effect of one condition at a specific level of another.
  + Is there an effect of `Hospital` for those receiving `Treatment A`? (...and so on for all combinations.)
  
+ **Interactions (categorical-categorical)**
  + A change in the effect of some condition as a function of another.
  + Does the effect of `Treatment` differ by `Hospital`? 
  + With effects coding, we can also think of this as a difference in simple effects.
  

---
# Our model and coefficients

+ Remember our effects coded model:

$$y_{ijk} = b_0 + \underbrace{(b_1E_1 + b_2E_2)}_{\text{Treatment}} + \underbrace{b_3E_3}_{\text{Hospital}} + \underbrace{b_4E_{13} + b_5E_{23}}_{\text{Interactions}} + \epsilon_{i}$$
+ Where the variables represent the following comparisons: 

```{r, echo=FALSE}
tibble(
  Treatment = c("A", "A", "B", "B", "C", "C"),
  Hospital = rep(c("Hosp1", "Hosp2"),3),
  E1 = c(1,1,0,0,-1,-1),
  E2 = c(0,0,1,1,-1,-1),
  E3 = c(1,-1,1,-1,1,-1),
  E13 = c(1,-1,0,0,-1,1),
  E23 = c(0,0,1,-1,-1,1)
)
```

---
# Our model and coefficients

+ And our model in R:

```{r}
m1 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl,
         contrasts = list(Treatment = contr.sum, 
                          Hospital = contr.sum)) # you can code contrasts within lm
```


---
# Table of means

+ Useful to keep in mind the group means:

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```


---
# Simple Effects
+ We noted previously that simple effects consider the effect of one condition at a specific level of the other.
  + Is there an effect of `Hospital` for those receiving `Treatment A`? (and so on for all combinations)
  + Or, put another way, is there a difference in `SWB` between Hospitals 1 and 2 for people receiving Treatment A

---
# Simple Effects with `emmeans`

.pull-left[
```{r}
m1_emm <- emmeans(m1, ~Treatment*Hospital)
m1_simple1 <- pairs(m1_emm, 
                    simple = "Hospital")
m1_simple1
```
]

.pull-right[

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```

]
---
# Simple Effects with `emmeans`

.pull-left[
```{r}
m1_simple2 <- pairs(m1_emm, 
                    simple = "Treatment")
m1_simple2
```
]

.pull-right[

```{r, echo=FALSE}
knitr::kable(tibble::tribble(
   ~V1, ~V2, ~V3, ~V4,
  "TreatA", "10.80",  "7.85", "9.33",
  "TreatB", "9.43",  "13.11", "11.27",
  "TreatC", "10.10",  "7.98", "9.04", 
  "Marginal","10.11", "9.65", "9.88"
  ), col.names = c("", "Hosp1",  "Hosp2", "Marginal"))
```

]

---
# Simple effects with plots

.pull-left[
```{r, echo=FALSE, out.width="90%"}
emmip(m1, Treatment ~ Hospital)
```

]

.pull-right[
```{r}
m1_simple1
```

]

---
# Simple effects with plots

.pull-left[
```{r}
m1_simple2
```

]

.pull-right[
```{r, echo=FALSE, out.width="90%"}
emmip(m1, Hospital ~ Treatment)
```

]


---
class: center, middle
# Thanks for listening

