---
title: "dapR2 Week 3 Live R"
output: html_document
date: "2022-10-03"
---


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(simglm)
```

# Some things from class questions:
Create a random data set and run an `lm`

```{r}

```

Plot

```{r}

```

# Closer look at the t-distribution and inference


```{r}
tibble(
  t_value = rep(2, 10),
  t_df = c(seq(1,100, 10)),
  Crit_value = abs(round(qt(0.025, c(seq(1,100, 10))), 3)),
  Exactp = (1 - pt(2, c(seq(1,100, 10)))) * 2
)
```

```{r, echo=FALSE}
ggplot() + 
  xlim(-3, 3) +
  stat_function(fun=dt,
                geom = "line",
                colour = "black",
                args = list(df=1)) +
  stat_function(fun=dt,
                geom = "line",
                colour = "red",
                args = list(df=11)) +
  stat_function(fun=dt,
                geom = "line",
                colour = "blue",
                args = list(df=91)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.975, 11), 3),
                alpha=.25,
                fill = "red",
                args = list(df=11)) +
  stat_function(fun = dt, 
                geom = "area",
                xlim = c(qt(0.975, 91), 3),
                alpha=.25,
                fill = "blue",
                args = list(df=91)) +
  geom_vline(xintercept = 2, col="grey", cex=2) +
  xlab("\n t") +
  ylab("") +
  ggtitle("t-distribution: black = 1df; red = 21 df; blue = 91 df")
```

# Example

```{r}
set.seed(7284) # change this number to generate different data with approximately the same beta

sim_arguments <- list(
  formula = y ~ 1 + hours + motivation,
  fixed = list(hours = list(var_type = 'ordinal', levels = 0:15),
               motivation = list(var_type = 'continuous', mean = 0, sd = 1)),
  error = list(variance = 20),
  sample_size = 150, # change this number to change the sample size
  reg_weights = c(0.6, 1.4, 1.5) # change these numbers to change the beta. the 1st number is intercept, 2nd beta for hours, 3rd beta for motivation
)

df <- simulate_fixed(data = NULL, sim_arguments) %>%
  simulate_error(sim_arguments) %>%
  generate_response(sim_arguments)

test_study2 <- df %>%
  dplyr::select(y, hours, motivation) %>%
  mutate(
    ID = paste("ID", 101:250, sep = ""),
    score = round(y+abs(min(y))),
    motivation = round(motivation, 2)
  ) %>%
  dplyr::select(ID, score, hours, motivation)

```

Tasks for this data:

1. Plot `x1` and `x2` against `y`.
2. Run a linear mode with all 3 variables
3. Write down interpretations of the $\beta$ coefficients
4. Use the predict function for the values x1 = .., x2 = ...

# Principles of `ggplot()`

- Works in layers
- `geoms` are the things we see (points, bars etc.)
- `geoms` need asthetics, these are the variables/information used to make the things we see
- We can edit pretty much everytihng

Let's do a scatterplot in layers

```{r}

```

# Run `lm`

# Interpretations

# `predict`


