---
title: "dapR2 live R: Week 4"
author: "Tom Booth"
date: "11/10/2021"
output: html_document
---

# Data
Some data:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(10101)
service <- rnorm(100, 6, 1.2)
b0 = 16.5
b1 = 4.8
eps <- rnorm(100, 4, 5)
salary <- b0 + (b1*service) + eps

tib <- tibble(
  ID = paste("ID",101:200, sep=""),
  service = service,
  salary = salary
)

```

And a simple linear model:

```{r}
dave <- lm(salary ~ 1 + service, data = tib)
summary(dave)
```


# Predict function (and design)

How can we find the values that our model predicts?

$$y_i = b_0 + b_1x_1$$

```{r}
employee_prediction <- tibble(
  service = c(3.5, 5, 10, 20, 30, 40)
  )

bob <- tibble(
  service = 15
)

predict(m1, 
        bob
        )

```

```{r}
summary(m1)
```


```{r}
20.862 + 4.863*15

m1$coefficients[1] + m1$coefficients[2]*15
```


## Interpolation, extrapolation

- Interpolation = predicting values within the measured range of $x$
- Extrapolation = predicting outside the measured range of $x$

```{r}
tib %>%
  ggplot(., aes(x=service, y=salary)) +
  geom_abline(intercept = m1$coefficients[1], slope = m1$coefficients[2], col = "red") +
  geom_point(alpha = .3)
```

The point of intersection of the two lines will fall on our model line. The horizontal line comes from the predicted value for Bob's salary, the vertical line is his known years of service ($x_1$)

```{r}
tib %>%
  ggplot(., aes(x=service, y=salary)) +
  geom_abline(intercept = m1$coefficients[1], slope = m1$coefficients[2], col = "blue") +
  geom_point(alpha = 0.3) +
  ylim(c(0,250)) +
  xlim(c(0, 40)) +
  geom_vline(xintercept = 15, lty = "dashed") +
  geom_hline(yintercept = 93.80319, lty = "dashed")
```


# Lab Q7

The plot below shows the intersecting lines at the mean of $x$ (service) corresponding to the mean of $y$ (salary).

```{r}
mean_y = mean(tib$salary)
mean_x = mean(tib$service)

tib %>%
  ggplot(., aes(x=service, y=salary)) +
  geom_abline(intercept = m1$coefficients[1], slope = m1$coefficients[2], col = "blue") +
  geom_point(alpha = 0.3) +
  #ylim(c(0,250)) +
  #xlim(c(0, 40)) +
  geom_vline(xintercept = mean_x, lty = "dashed") +
  geom_hline(yintercept = mean_y, lty = "dashed")
```


# Quiz question approach
A quick note that when you are asked to do calculations in quizzes, the fastest thing for you to do is to create the data in R, and run a model/do the calculations. The code below is a very quick way...

```{r}
Dv = c(10, 13, 14, 15, 16)
Iv = c(1,2,3,4,5)

lm(Dv ~ Iv)
```

But with this be careful what else you have in your environment. You may want to take a minute or so longer to create a data set. For example:

```{r}
quiz1 <- tibble(
  y = c(10, 20, 30, 41, 48),
  x = c(1,2,3,4,5)
)

summary(lm(y~x, data = quiz1))
```

