---
title: "07: Sum coding"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: FALSE
    # TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')
# source('assets/setup.R')

library(tidyverse)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll TODO

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab07.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
   
:::

---

:::frame

TODO FIXME

To investigate if the number of words typed per minute (WPM) differs between people who've consumed different kinds of beverages, the researchers conducted an experiment where 40 participants were each randomly allocated to one of four beverage conditions.

The data in `caffeinedrink.csv` contained two attributes collected from $n=40$ participants:

```{r echo=FALSE, message=FALSE, warning=FALSE}
tibble(
    variable = names(read_csv("https://uoepsy.github.io/data/caffeinedrink.csv")),
    description = c("which one of four treatment conditions the participant was assigned - control (water), coffee, mint tea, or red bull", "average number of words typed per minute")
) |> gt::gt()
```

:::

# q1

