---
title: "08: Assumptions and diagnostics"
params: 
    # SHOW_SOLS: TRUE
    SHOW_SOLS: FALSE
    # TOGGLE: FALSE
    TOGGLE: TRUE
editor_options:
  chunk_output_type: console
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')
# source('assets/setup.R')

library(tidyverse)
library(car)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll revisit a multiple regression model you developed in Block 1.
Using that model, you'll practice checking assumptions and running diagnostics of multicollinearity and influence.

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab08.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
    - `car`
   
:::

---

:::frame

Research question (RQ): **Is there an association between wellbeing and time spent outdoors, when controlling for the effect that social interaction has on wellbeing?**

Data dictionary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
mwdata  <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
tibble(
variable = names(mwdata),
description = c("Age in years of respondent","Self report estimated number of hours per week spent outdoors ", "Self report estimated number of social interactions per week (both online and in-person)", "Binary 1=Yes/0=No response to the question 'Do you follow a daily routine throughout the week?'", "Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and wellbeing. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70", "Location of primary residence (City, Suburb, Rural)", "Average weekly number of steps in thousands (as given by activity tracker if available)")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

From the Edinburgh & Lothians, 100 city/suburb residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS) was used to measure mental health and wellbeing. 

Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), whether a daily routine is followed (yes/no). For those respondents who had an activity tracker app or smart watch, they were asked to provide their average weekly number of steps.

:::

:::

# Set up data and model

<!-- # q1 -->

`r qbegin(qcounter())`

Read in the data from  <https://uoepsy.github.io/data/wellbeing_rural.csv> and store it in a variable named `mwdata` ("mw" stands for "mental wellbeing").

Use the function `lm()` to fit the linear model represented by the mathematical model formulation below, and name the result `m1`.

$$
\text{wellbeing} = \beta_0 + 
(\beta_1 \cdot \text{outdoor\_time}) +
(\beta_2 \cdot \text{social\_int}) +
\epsilon
$$

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mwdata <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
m1     <- lm(wellbeing ~ outdoor_time + social_int, data = mwdata)
```

`r solend()`


# Check assumptions

<!-- # q2 -->

`r qbegin(qcounter())`

Do the associations between each predictor and the outcome appear sufficiently linear to justify using a linear model for this data?

:::red
🗂️
See [LM assumptions > Linearity of association](https://uoepsy.github.io/dapr2/2627/flashcards/block2/assump.html#l-linearity-of-association){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Assess visually using component-residual plots.

```{r}
crPlots(m1)
```

We want the wobbly pink line to match the dashed blue line.
I think the match is pretty good!
Associations look sufficiently linear.

`r solend()`


<!-- # q3 -->

`r qbegin(qcounter())`

Are the errors of `m1` independent?

:::red
🗂️
See [LM assumptions > Independence of errors](https://uoepsy.github.io/dapr2/2627/flashcards/block2/assump.html#i-independence-of-errors){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

As long as each observation comes from a different source—for example, as long as we don't have multiple observations from the same person—we can safely assume that errors are independent.

Let's double-check this in the data:

```{r}
mwdata |> head()
```

A clear give-away would be a column like `participant_id` with IDs that appear only once each.
But we can still tell that we have one observation per person because every row has a different value of `age`.

So we can assume here that errors are independent because each observation comes from a different source.

`r solend()`

<!-- # q4 -->

`r qbegin(qcounter())`

Are the errors of `m1` sufficiently normally distributed?

:::red
🗂️
See [LM assumptions > Normality of errors](https://uoepsy.github.io/dapr2/2627/flashcards/block2/assump.html#n-normality-of-errors){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Assess visually using a Q-Q plot.

```{r}
plot(m1, which = 2)
```

The match to the diagonal is pretty good, except for that one observation in row 109 (top right) that's pretty far off the diagonal.
But it's just one data point, which could be random chance, so I'd tend to say (for now at least) that the errors are sufficiently normally distributed.

`r solend()`

<!-- # q5 -->

`r qbegin(qcounter())`

Do the errors of `m1` have sufficiently equal variance across the whole range of fitted values?

:::red
🗂️
See [LM assumptions > Equal variance of errors](https://uoepsy.github.io/dapr2/2627/flashcards/block2/assump.html#e-equal-variance-of-errors){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Assess visually using a residuals-vs-fitted plot.

```{r}
plot(m1, which = 1)
```

We're looking for an even vertical spread of residuals across the whole x axis.
That's *not* what we see here!
The residuals are not a uniform cloud, but have a fan/funnel kind of shape.
Specifically, the vertical spread of the residuals is a lot smaller for fitted values around 32 and 34, and a lot bigger around 38 and 40.
So the the errors of this model do *not* have equal variance.

A pattern like this should suggest to you that we should bootstrap this model before interpreting its estimates; bootstrapping will be covered in lab/lecture 09.

Also, notice 109 sticking out again at top right ... I'm starting to get suspicious ...

`r solend()`

# Check multicollinearity

<!-- # q6 -->

`r qbegin(qcounter())`

Do you think multicollinearity between predictors is a problem for `m1`?

:::red
🗂️
See [Checking multicollinearity](https://uoepsy.github.io/dapr2/2627/flashcards/block2/multicollinearity.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

Measure multicollinearity using VIF (variance inflation factor).

```{r}
vif(m1)
```

We want values below 5, and the closer to 1, the better.
Because these values are very close to 1, I would conclude that multicollinearity between predictors is not inflating the model's variance estimates.

`r solend()`

# Check influence

<!-- # q7 -->

`r qbegin(qcounter())`

One particular observations influences the model's estimates more than others.
What row is that observation in?

:::red
🗂️
See [Influential observations](https://uoepsy.github.io/dapr2/2627/flashcards/block2/influence.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

You might have a guess already, based on the data point that's been behaving funny in a couple of the assumption plots...

To check: Measure influence using Cook's Distance.

```{r}
plot(m1, which = 4)
```

Observation 109 rears its ugly head again!

`r solend()`

# Run sensitivity analysis

<!-- # q8 -->

`r qbegin(qcounter())`

Fit the same model from Q1 to a version of the dataset that does not include the high-influence data point.
Call that model `m2`.

:::red
🗂️
See [Sensitivity analysis](https://uoepsy.github.io/dapr2/2627/flashcards/block2/sensitivity-analysis.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
m2 <- lm(
  wellbeing ~ outdoor_time + social_int,
  data = mwdata[-109,]  # "subtract" row 109
)
```

`r solend()`


<!-- # q9 -->

`r qbegin(qcounter())`

Compare the model summaries for `m1` and `m2`.

- How much do the coefficient estimates change after the high-influence data point has been removed?
- How much does the pattern of statistical significance change?

:::red
🗂️
See [Sensitivity analysis](https://uoepsy.github.io/dapr2/2627/flashcards/block2/sensitivity-analysis.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

::::{.columns}
:::{.column width="48%"}
```{r}
summary(m1)
```

:::
:::{.column width="4%"}

:::
:::{.column width="48%"}

```{r}
summary(m2)
```

:::
::::

The coefficient estimates change a little bit, but the sign stays the same (i.e., if an estimate was positive in `m1`, it's still positive in `m2`).
And the pattern of significance doesn't change at all: all coefs are still significant at $p$ < .001.

`r solend()`

<!-- # q10 -->

`r qbegin(qcounter())`

Does the high-influence observation impact the conclusions you would draw from the model?

:::red
🗂️
See [Sensitivity analysis](https://uoepsy.github.io/dapr2/2627/flashcards/block2/sensitivity-analysis.html){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

No.
Even though observation 109 has a Cook's distance that's much greater than any of its fellows, it does not have much practical impact on our conclusions.
When it's excluded, the signs of the coefficients and the pattern of significance do not change.
We can therefore conclude that observation 109 does not impact our conclusions, because it's not driving any of the effects we see in the model.

`r solend()`


# Optional: Knit your report to PDF

:::red

Rmarkdown is a useful format because it can be used as a foundation for generating different kinds of documents.
For the group report, you'll be asked to render your Rmd file to PDF.
Why not practice now?

First, make sure you have **tinytex** installed in R.
This is what lets you convert things into PDF format.

Run the following code to install tinytex (you'll only need to run it once).

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```


Then ensure that the "yaml" (rhymes with "mammal") header—the bit at the top of your Rmd document—looks something like this:

```{}
---
title: "this is my report title"
author: "B1234506"
date: "07/09/2025"
output: bookdown::pdf_document2
---
```

**Then follow [the instructions from the Rmd bootcamp](https://uoepsy.github.io/scs/rmd-bootcamp/02-compiling.html){target="_blank"} to knit your Rmd document to PDF.**

If you encounter errors, refer to [the DAPR1 formatting resources / knitting checklist](https://uoepsy.github.io/dapr1/2526/labs/zf_formatting_resources.html){target="_blank"}.


:::{.callout-tip appearance="simple" collapse="true"}
#### What to do if you cannot knit to PDF

If you are having issues knitting directly to PDF, try the following:  

- Knit to HTML file  
- Open your HTML in a web-browser (e.g. Chrome, Firefox)  
- Print to PDF (Ctrl+P, then choose to save to PDF)  
- Open file to check formatting
:::

:::{.callout-tip appearance="simple" collapse="true"}
#### Hiding code and/or output from code chunks

Review [Lesson 5 of the rmd bootcamp](https://uoepsy.github.io/scs/rmd-bootcamp/05-echoeval.html) for a detailed description/worked examples.

**Hiding R Code**

To not show the code of an R code chunk, and only show the output, write:

````
```{{r, echo=FALSE}}
# code goes here
```
````

**Hiding R Output:**

To show the code of an R code chunk, but hide the output, write:

````
```{{r, results='hide'}}
# code goes here
```
````

**Hiding R code AND output**

To hide both code and output of an R code chunk, write:

````
```{{r, include=FALSE}}
# code goes here
```
````
:::

:::

