---
title: "09: Bootstrapping"
params: 
    # SHOW_SOLS: TRUE
    SHOW_SOLS: FALSE
    # TOGGLE: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

<!-- when rendering in RStudio / not production profile (tutor-facing):      SHOW_SOLS: TRUE  -->
<!-- when rendering in terminal / with production profile (student-facing):  SHOW_SOLS: FALSE -->

```{r}
#| label: setup
#| include: false

source('../assets/setup.R')
# source('assets/setup.R')

library(tidyverse)
library(car)

qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
```

This week, you'll bootstrap some coefficient estimates for last week's model.

There's relatively little new coding this week, so if you have extra time after running through the exercises, why not start putting together some study materials?
Your future self will thank you :)
We give some recommendations in Q6 below.

:::{.callout-note}
### Get set up

1. Open RStudio.
1. Create a new .Rmd file for this week's exercises.
1. Save it somewhere you can find it again.
1. Give it a clear name (for example, `dapr2_lab09.Rmd`).
1. In the first code chunk, load the packages you'll need this week (and install them if you don't have them already):
    - `tidyverse`
    - `car`
   
:::

---

:::frame

Research question (RQ): **Is there an association between wellbeing and time spent outdoors, when controlling for the effect that social interaction has on wellbeing?**

Data dictionary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
mwdata  <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
tibble(
variable = names(mwdata),
description = c("Age in years of respondent","Self report estimated number of hours per week spent outdoors ", "Self report estimated number of social interactions per week (both online and in-person)", "Binary 1=Yes/0=No response to the question 'Do you follow a daily routine throughout the week?'", "Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS), a self-report measure of mental health and wellbeing. The scale is scored by summing responses to each item, with items answered on a 1 to 5 Likert scale. The minimum scale score is 14 and the maximum is 70", "Location of primary residence (City, Suburb, Rural)", "Average weekly number of steps in thousands (as given by activity tracker if available)")
) |> gt::gt()
```

:::{.callout-note collapse='true'}
### More detail about this dataset

From the Edinburgh & Lothians, 100 city/suburb residences and 100 rural residences were chosen at random and contacted to participate in the study. The Warwick-Edinburgh Mental Wellbeing Scale (WEMWBS) was used to measure mental health and wellbeing. 

Participants filled out a questionnaire including items concerning: estimated average number of hours spent outdoors each week, estimated average number of social interactions each week (whether on-line or in-person), whether a daily routine is followed (yes/no). For those respondents who had an activity tracker app or smart watch, they were asked to provide their average weekly number of steps.

:::

:::

# Bootstrapping

<!-- # q1 -->

`r qbegin(qcounter())`

Read in the data from  <https://uoepsy.github.io/data/wellbeing_rural.csv> and store it in a variable named `mwdata` ("mw" stands for "mental wellbeing").

Use the function `lm()` to fit the linear model represented by the mathematical model formulation below, and name the result `m1`.

$$
\text{wellbeing} = \beta_0 + 
(\beta_1 \cdot \text{outdoor\_time}) +
(\beta_2 \cdot \text{social\_int}) +
\epsilon
$$

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
mwdata <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
m1     <- lm(wellbeing ~ outdoor_time + social_int, data = mwdata)
```

`r solend()`

<!-- # q2 -->

`r qbegin(qcounter())`

Here's a plot from last week's lab that illustrates a problem with this model's residuals:

```{r}
plot(m1, which = 1)
```

Make an attempt to write down from memory (a) what assumption this plot is checking, (b) what each point represents, and (c) the problem with the model's results that this plot illustrates.

After you've made a good-faith effort to respond to each point from memory, you can revisit last week's lab/notes to complete your responses.

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

(a) This is a residuals-vs-fitted plot, and it's used to visually assess the assumption of equal variance of errors.

(b) Each point represents one observation from our dataset: specifically, its residual (y axis) and its model-fitted/model-predicted outcome value (x axis).

(c) The problem is that these data points appear to build a fan/funnel/cone shape, where the vertical spread is much smaller at one end of the x axis compared to the other. 
This pattern indicates unequal variance, and it means we can't totally trust the model's estimates of variability, because they won't apply equally to the whole dataset.

`r solend()`

<!-- # q3 -->

`r qbegin(qcounter())`

Because the assumption shown in the plot above is not met, we must bootstrap our linear model's estimates.

Use the function `Boot()` to draw 1000 bootstrapped samples.
Name the output `m1_boot`.

:::{.callout-tip collapse='true'}
### Code hint

Replace the `...` in the code below with the appropriate values.

```{r eval=F}
... <- Boot(..., R = ...)
```

:::

:::red
🗂️
See [Bootstrapping > In R](https://uoepsy.github.io/dapr2/2627/flashcards/block2/bootstrapping.html#in-r){target="_blank"} flash card.
:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
m1_boot <- Boot(m1, R = 1000)
```

`r solend()`

<!-- # q4 -->

`r qbegin(qcounter())`

Copy and run the following line of code.

```{r eval=F}
summary(m1_boot)
```

Identify (a) the column of the output that represents the coefficent estimate, and (b) the column of the output that represents the standard error (i.e., the standard deviation of the sampling distribution).

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
summary(m1_boot)
```

(a) `original` is the column containing the coefficient estimate.

(b) `bootSE` is the column containing the bootstrapped standard error.

`r solend()`

<!-- # q5 -->

`r qbegin(qcounter())`

Use the function `confint()` to get the 95% CIs for each coefficient estimate.

For each coefficient, can we reject the H0 that the coefficient is equal to zero?

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
confint(m1_boot)
```

None of these 95% CIs contain zero, so for all three coefficients, we can reject the H0 at $\alpha$ = .05.

`r solend()`


# Develop study materials/strategies

<!-- # q6 -->

`r qbegin(qcounter())`

If you reach this point and still have a bit of time to dedicate to DAPR this week, we'd recommend that you start thinking about study/revision materials and strategies.

It's true that report and exam season is still far away, but it's good to start these things early.
Your learning will develop best through little bits of continuous effort.
So the more prep you do now, the more your understanding will grow over time.

Here are some of Elizabeth's favourite learning strategies (note that it's the *process* here that will help you learn, much more than just getting an end product):

1. [Mind mapping](https://subjectguides.york.ac.uk/note-taking/mind-map){target="_blank"}: Create visual representations of how concepts are connected. You could start a mind map now and keep updating it as the course goes on.
1. [Flash cards (and other active recall methods)](https://subjectguides.york.ac.uk/study-revision/active-recall#s-lg-box-16274981){target="_blank"}: Testing yourself using active recall now will train your brain in retrieving key information, so that it's there when you need it later.
    - If you like using physical flash cards, try the [Leitner system](https://subjectguides.york.ac.uk/study-revision/leitner-system){target="_blank"}, a system that adds in spaced repetition to help build your long-term memory.
1. [The Feynman technique](https://subjectguides.york.ac.uk/study-revision/feynman-technique){target="_blank"}: Teach or explain an idea as simply as you can. Wherever you're not sure what to say or how to explain something, brush up on those areas and try again.

`r qend()`


# Optional: Knit your report to PDF

:::red

Rmarkdown is a useful format because it can be used as a foundation for generating different kinds of documents.
For the group report, you'll be asked to render your Rmd file to PDF.
Why not practice now?

First, make sure you have **tinytex** installed in R.
This is what lets you convert things into PDF format.

Run the following code to install tinytex (you'll only need to run it once).

```{r eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```


Then ensure that the "yaml" (rhymes with "mammal") header—the bit at the top of your Rmd document—looks something like this:

```{}
---
title: "this is my report title"
author: "B1234506"
date: "07/09/2025"
output: bookdown::pdf_document2
---
```

**Then follow [the instructions from the Rmd bootcamp](https://uoepsy.github.io/scs/rmd-bootcamp/02-compiling.html){target="_blank"} to knit your Rmd document to PDF.**

If you encounter errors, refer to [the DAPR1 formatting resources / knitting checklist](https://uoepsy.github.io/dapr1/2526/labs/zf_formatting_resources.html){target="_blank"}.


:::{.callout-tip appearance="simple" collapse="true"}
#### What to do if you cannot knit to PDF

If you are having issues knitting directly to PDF, try the following:  

- Knit to HTML file  
- Open your HTML in a web-browser (e.g. Chrome, Firefox)  
- Print to PDF (Ctrl+P, then choose to save to PDF)  
- Open file to check formatting
:::

:::{.callout-tip appearance="simple" collapse="true"}
#### Hiding code and/or output from code chunks

Review [Lesson 5 of the rmd bootcamp](https://uoepsy.github.io/scs/rmd-bootcamp/05-echoeval.html) for a detailed description/worked examples.

**Hiding R Code**

To not show the code of an R code chunk, and only show the output, write:

````
```{{r, echo=FALSE}}
# code goes here
```
````

**Hiding R Output:**

To show the code of an R code chunk, but hide the output, write:

````
```{{r, results='hide'}}
# code goes here
```
````

**Hiding R code AND output**

To hide both code and output of an R code chunk, write:

````
```{{r, include=FALSE}}
# code goes here
```
````
:::

:::


