---
title: "dapR2 Week 7 Live R"
output: html_document
---

Open up the data from lab:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(sjPlot)


w7 <- read_csv("https://uoepsy.github.io/data/wellbeing_rural.csv")
slice(w7, 1:10)
```

Covert location to be a factor:

```{r}
w7 <- w7 %>%
  mutate(
    location = factor(location)
  )
slice(w7, 1:10)
str(w7)
```

Remember this has 3 levels, and in lab you dropped this down to 2-levels:

```{r}
w7 %>% 
    count(location)
```

Average wellbeing across groups:

```{r}
w7 %>%
  group_by(location) %>%
  summarise(
    mean = mean(wellbeing)
  )
```

Look at the `wellbeing` predicted by `outdoor_time` relationship:

```{r}
w7 %>%
  ggplot(., aes(x=outdoor_time, y=wellbeing)) +
  geom_point() +
  geom_smooth(method="lm")+
  facet_wrap(~location)
```

Look at the `wellbeing` predicted by `outdoor_time` relationship:

```{r}
w7 %>%
  ggplot(., aes(x=outdoor_time, y=wellbeing, colour = location)) +
  geom_point(alpha = 0.2)+
  geom_smooth(method="lm")
```

Have a look at the dummy codes:

```{r}
contrasts(w7$location)
```

Fit the model:

```{r}
m1 <- lm(wellbeing ~ outdoor_time*location, w7)
m1a <- lm (wellbeing ~ outdoor_time + location + outdoor_time:location, data = w7)
summary(m1)
```

Look at the interaction plots:

```{r}
plot_model(m1,type="int")
```

Changing the reference level:

```{r}
w7 <- w7 %>%
  mutate(
    location = relevel(location, ref = "rural")
  )
```

```{r}
contrasts(w7$location)
```


```{r}
m1b <- lm(wellbeing ~ outdoor_time*location, w7)
summary(m1b)
```

```{r}
plot_model(m1,type="int")
```

Mean centering:

```{r}
w7 <- w7 %>%
  mutate(
    m_outdoor = scale(outdoor_time, center = T, scale = F)[,1]
  )
```

New model

```{r}
m2 <- lm(wellbeing ~ m_outdoor*location, w7)
summary(m2)
```

```{r}
plot_model(m2, type = "int")
```





