---
title: 'dapR2 Live : Semester 2 week 2'
author: "dapR2 Team"
date: "24/01/2022"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

# Data

```{r, warning=FALSE, message=FALSE}
n <- round(500*(c(.55, .2, .1, .05, .1)),0)
set.seed(7284)
wb_tib <- tibble(
  swb = c(rnorm(n[1], 11, 3.6),
          rnorm(n[2], 12, 4.2),
          rnorm(n[3], 8, 2.2),
          rnorm(n[4], 6, 1.1),
          rnorm(n[5], 9.5, 2.5)),
  status = factor(c(rep("Married/CP", n[1]),
             rep("Cohab", n[2]),
             rep("Single", n[3]),
             rep("Widowed", n[4]),
             rep("Divorced", n[5])))
)
```

Take a look at the structure of the data set and the variables present:

```{r}
str(wb_tib)
```

And the descriptive stats by group.

```{r}
wb_tib %>%
  group_by(status) %>%
  summarise(
    n = n(),
    mean = round(mean(swb),2),
    sd = round(sd(swb),2)
  ) 
```

# Coding Schemes: Dummy

```{r}
contrasts(wb_tib$status)
```

What does this mean? First to create a little table of group means for reference

```{r}
gpM <- wb_tib %>%
  group_by(status) %>%
  summarise(
    swb = round(mean(swb),2)
  ) %>%
  mutate(
    GM = c(rep(mean(swb),5))
  )
gpM
```

Visualise the comparisons:

```{r}
wb_tib %>%
  ggplot(., aes(x=status, y=swb, colour = status, shape = status)) +
  geom_jitter(width = 0.1, size = 2) + 
  geom_hline(yintercept = as.numeric(gpM[1,2]), lty = 2, colour = "darkgrey") +
  theme_classic() +
  geom_errorbar(data = gpM, width=0.8,aes(ymax=..y..,ymin=..y..))
```

Run our `lm`

```{r}
res_dummy <- lm(swb ~ status, data = wb_tib)
summary(res_dummy)
```

```{r}
contrasts(wb_tib$status) <- contr.treatment(n = 5, base = 3)
contrasts(wb_tib$status)
```

```{r}
res_dummy <- lm(swb ~ status, data = wb_tib)
summary(res_dummy)
```

Have a look at the group means to compare:

```{r}
gpM
```


# Coding Schemes: Effects

Switch our contrasts

```{r}
contrasts(wb_tib$status) <- contr.sum
contrasts(wb_tib$status)
```

Visualize

```{r}
wb_tib %>%
  ggplot(., aes(x=status, y=swb, colour = status, shape = status)) +
  geom_jitter(width = 0.1, size = 2) + 
  geom_hline(yintercept = mean(wb_tib$swb), lty = 2, colour = "darkgrey") +
  theme_classic() +
  geom_errorbar(data = gpM, width=0.8,aes(ymax=..y..,ymin=..y..))
```

Run the model

```{r}
wb_tib$status <- fct_relevel(wb_tib$status, "Widowed")
contrasts(wb_tib$status) <- contr.sum
contrasts(wb_tib$status)
```

```{r}

res_zero <- lm(swb ~ status, data = wb_tib)
summary(res_zero)
```

```{r}
summary(res_dummy)
```

Refer to group means:

```{r}
gpM
```


# Predicted values
```{r}
pred_swb <- tibble(status = c("Cohab", "Divorced", "Married/CP", "Single", "Widowed"))
pred_swb %>%
  mutate(
    pred_dummy = round(predict(res_dummy, newdata = .),2),
    pred_zero = round(predict(res_zero, newdata = .),2)
  )

```

```{r}
gpM
```

