---
title: 'dapR2 Live : Semester 2 wek 1'
author: "dapR2 Team"
date: "15/01/2022"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(psych)
library(sjPlot)
```

# Data

```{r, warning=FALSE, message=FALSE}
rep <- read_csv("https://uoepsy.github.io/data/dapr2_2122_report1.csv")
```

Take a look at the structure of the data set and the variables present:

```{r}
str(rep)
```

# Models

**Question 1**: Is the perceived persuasiveness of a speaker influenced by the rate at which participants speak, and is dependent of listeners having counter- or pro-attitudinal opinions towards the statements being presente?

```{r, eval=FALSE}
m1 <- lm(persuasive ~ sp_rate*endorse)
```

- Question is answered by the interaction.
- It is going to require us to run the model, interpret the effects, and plot the interaction.


**Question 2**: Does the age of the speaker, and whether or not the speaker was presented as an expert, or else was member of the general public, account for any identified effects of rate of speech and counter- or pro-attitudinal opinions?

```{r, eval=FALSE}
m2 <- lm(persuasive ~ age + expert + sp_rate*endorse)
```

- Is asking about accounting for effects (i.e. change)
- We need `m1` to know if there is change when we fit `m2`
- Key to answering second question is a descriptive comparison on the interaction effect in `m1` and `m2`

# Data cleaning steps

1. Check ranges of variables
2. Code expert and endorse as a factors
3. Decide on 0 point for speach rate

**Check data**

Check the range: 

```{r}
describe(rep)
```

Code expert as factor:

```{r}
rep <- rep %>%
  mutate(
    expert = factor(expert, labels = c("public", "expert")),
    sp_rate = scale(sp_rate),
    endorse = factor(endorse)
  )
```


Have a look at associations (excluding factors):
```{r}
rep %>%
  select(persuasive, sp_rate, age) %>%
  cor(.)
```

All small effects. 

# Run the models

```{r}
m1 <- lm(persuasive ~ sp_rate*endorse, data = rep)
m2 <- lm(persuasive ~ age + expert + sp_rate*endorse, data = rep)
```

# Main areas from report 1

## Interpreting coefficients

```{r}
summary(m1)
```

- *intercept* = predicted persuasiveness in the Counter group with mean speach rate.
- *sp_rate* = conditional effect of speach rate. For each SD increase in speach rate in the Counter group, persuasiveness increases by 3.718. 
- *endorsepro* = conditional effect of endorsement. The difference in persuasiveness between the Counter and Pro groups at mean speach rate was 5.055.
- *sp_rate:endorsepro* = The difference in slopes across the groups. For each standard deviation increase in speach rate, those in the Pro group decrease in persuasiveness by an additional 8.887.


```{r}
plot_model(m1, type = "int")
```


```{r}
summary(m2)
```

Main point here, interpretation of both `expert` and `age` are as they would be in a standard linear model. What does holding constant mean?

## Coding of categorical variables

## Impact of covariates
Descriptive here (we will talk about tests of multiple variables this semester)

- Conditional effect of speach rate = same direction, approximately same magnitude
- Conditional effect of endorsement = same direction, approximately same magnitude
- Interaction term = same direction, approximately same magnitude

Conclusion, whilst `age` does significantly predict persuasiveness, it does not change the the effects of the key variables from Q1.

