---
title: "<b> Exam Q&A 1 </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "dapR Team"
institute: "Department of Psychology<br>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #  base_color = "#0F4C81", # DAPR1
  base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  # base_color = "#95A5A6", #intro
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)

library(tidyverse)
```

# Please read the exam hall regulations [here](https://www.ed.ac.uk/sites/default/files/atoms/files/the_university_of_edinburgh_exam_hall_regulations_-_september_2018.pdf)


---
# dapR2 Exam

+ **Exam** (60%)
  + Summer exam block
  + Mix of MCQ, short answer R/calculation, and larger interpretation questions.
  + Closed book
  + You will be given an equation sheet. You will be allowed to make notes on this sheet and take it into the exam. 
    + There will be strict rules on this, if they are not followed, your notes sheet will be taken away and replaced with a blank equation sheet.

+ Primary assessment of individual learning.
  
---
# Section A: MCQ

+ 20 MCQ questions

+ Each worth 1 mark

+ Exactly the same style as your weekly quizzes
  + All questions have 4 possible options
  + Varied content across the whole course

+ Presented slightly differently as it is a written exam
  + You will see examples in final week lab

---
# Section B: Open Questions

+ Contains 10 questions

+ These questions include:
  + Defining concepts
  + Basic calculations, writing out models
  + Interpreting models

---
# What do you need to take in?

+ Black or blue ink pen (and spare)
+ Pencil
+ Rubber
+ Calculator


---
# Equation Sheet
+ This is available to you now in the Assessments tab on LEARN.

+ You can download and print it.

+ There is a box on the right hand side of each page.
  + You may add **hand written** notes to this section only

+ You will have your equation sheet taken off you and replace with a blank version if:
  + You have typed notes
  + You have gone outside of the box
  + You have written anything on any of the blank spaces on the front or back of the page

---
# What will be on the desk?

+ Pencil
+ Rubber
+ MCQ answer book
+ Exam answer booklet
+ Question booklet
+ Attendance slips

---
# What will the process be on exam day?
+ When timetable is released, you will know the time and room.
  + Arrive ~30mins before exam.
  + You may not be able to enter the room this early, but it will mean you are there when you can.
  + You will be instructed where to leave bags, phones etc.

+ Take all stationary, calculator, noted equation sheet, **student ID**, to your desk.

+ We will come and check your equation sheets and put a green sticker on it to say it has been checked.

---
# What will the process be on exam day?
+ Invigilator will then take over and explain timings, etc.
  + You will have 2hrs
  + there will be an MCQ answer sheet for section A
  + an exam book for section B
  + I will show how to fill these in in the last week of class.
  + do the exam

+ You must leave the question book, all answer sheets, your equation sheet, in the exam hall. 

---
# How should I prepare?

1. Look over the equation sheet and make sure you understand all equations and terms.
2. Add any notes to this sheet about use, or other things you want to have noted.
3. Review materials and make sure you have not missed anything on the sheet.
4. Look over MCQ questions and consider if your notes would help.
5. Look over interpretations in labs and consider if your notes would help.
6. Revise your notes.


---
class: center, middle
# And that is it for this intro!
