---
title: "dapr2 week 8 live R"
subtitle: "Department of Psychology, The University of Edinburgh"
output:
  html_document:
    theme: flatly
    number_sections: true
---

# The study
## Background
In the field of individual differences and health, the concept of healthy Neuroticism is being increasingly studied. In short, research has shown that Neuroticism (N) typically has negative associations with physical health and psychological well-being, whereas Conscientiousnes (C) has positive associations. This has led to the idea that for those high in N, having high C may buffer the negative impact of the trait on health.

For more background, see:

- [Associations with longevity](https://uoepsy.github.io/files/dapr2/HealthyN_Long.pdf)
- [Associations with health behaviours](https://uoepsy.github.io/files/dapr2/HealthyN_HB.pdf)
- [Associations with chronic conditions](https://uoepsy.github.io/files/dapr2/HealthyN_CC.pdf)

**Please note, you are not expected to read or comment on these papers, they are here for broader context if you wish to look them over**

## Procedure
Researchers were interested in assessing the degree to which urban environments and personality traits contribute towards perceptions of health. They conducted a survey study (n = 750) assessing individual's personality - as measured by a self report Big Five measure from the International Personality Item Pool (IPIP; 6point Likert scale ranging from 1 to 6), as well as individual's age, location of residence (city, suburb, countryside) and multiple self-ratings of different aspects of health (sleep, diet, physical health, mental health - high scores = better health). These variables were combined into a single z-score composite variable.

## Research Questions {#research-questions}
The researchers' final research questions were as follows:

1. Is there a difference in self-rated health by location of residence?
2. Is there a difference in self-rated health between those who live in the countryside vs suburb?
3. Controlling for age and location, which of the Big Five personality traits has the strongest association with self-rated health?
4. Controlling for age and location, is there evidence to support healthy neuroticism? 

```{r eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)
dict <- tibble(
  Variable = c("ID", "age", "location", "N", "E", "O", "A", "C", "srh"),
  Description = c("Unique participant identifier",
                  "Age of participant (in years)",
                  "Which of the following best describes where you live? 1: City; 2: Suburb; 3: Countryside",
                  "Sum score of 10 personality items assessing N from the IPIP Big Five measure",
                  "Sum score of 10 personality items assessing E from the IPIP Big Five measure",
                  "Sum score of 10 personality items assessing O from the IPIP Big Five measure",
                  "Sum score of 10 personality items assessing A from the IPIP Big Five measure",
                  "Sum score of 10 personality items assessing C from the IPIP Big Five measure",
                  "z-score composite variable assessing self-rated health")
)
knitr::kable(dict, "html") %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```

# Analysis

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(interactions)

dat <- read_csv("./")
```


## Visualise data

```{r}

```

## Clean and rescale data

```{r}

```

## Run model

```{r}

```

## Tabulate results

```{r}

```

## Visualise results

```{r}

```

