---
title: "DAPR2 report – EXAMPLE"
format:
  html:
    toc: true
    number_sections: false
    toc-location: left
editor_options: 
  chunk_output_type: console
---

```{css echo=FALSE}
div.blue, div.red, div.green, div.yellow, div.frame{ 
    border-radius: 5px; 
    padding: 20px 20px 10px 20px; 
    margin-top: 20px; 
    margin-bottom: 20px; 
}
.blue { background-color:#d9edf7 !important; }
.green { background-color:#dff0d8 !important; }
.yellow { background-color:#fcf8e3 !important; }
.red { background-color:#F3E3E5 !important; 
.frame {border: 1px solid #333333 !important; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval=FALSE, warning=FALSE, message=FALSE)
```

:::red
**THIS DOCUMENT IS ONLY AN EXAMPLE.**

**It's based on last year's report, and it's only meant to give you a preview of what the analysis task for the group report might look like.**

**You will not do this specific task.**

**The actual report task that you will do will be published in Semester 2.**
:::


The data is available at: [LINK REDACTED]

The data is simulated, designed to expand on the methods and results reported in the following paper:

> Weisberg, D. S., Keil, F. C., Goodstein, J., Rawson, E., & Gray, J. R. (2008). The seductive allure of neuroscience explanations. *Journal of Cognitive Neuroscience, 20*(3), 470-477. https://dx.doi.org/10.1162%2Fjocn.2008.20040


## Study background and procedure

In this study, we're interested in how people perceive inaccurate explanations of psychological phenomena.
Specifically, we want to know whether people judge inaccurate explanations to be more accurate if those explanations are accompanied by other kinds of information, and how that effect might be impacted by people's level of education.

In total, 270 participants were recruited: 90 with no degree, 90 with a UG degree, and 90 with a PG degree. 
These are the three education groups. 
The UG and PG degrees were all in cognitive neuroscience, cognitive psychology, or closely related fields.

Participants from each education group were equally assigned into one of three neuroscience information conditions: 90 participants to the "no neuroscience" condition, 90 to "text neuroscience", and 90 to "visual neuroscience". 
In the "no neuroscience" condition, no neuroscience information (i.e., no brain-related information in any format) was provided. 
In the "text neuroscience" condition, a description of the brain areas involved in the phenomenon was added to each explanation. 
In the "visual neuroscience" condition, an image showing the brain area(s) involved in the phenomenon was added to each explanation.  

Participants took part in a testing block that matched their neuroscience information condition.
In this testing block, participants judged the accuracy of 10 inaccurate explanations of psychological phenomena on a scale of 0 (inaccurate) to 10 (accurate). 
Each participants’ accuracy judgements across the 10 explanations were then summed up, which gives us an accuracy score from 0 to 100 for each person.


## Research question and prediction

RQ: Do differences in accuracy judgements between neuroscience conditions depend on people's level of education?

Specific prediction:

- People with UG and PG degrees will judge explanations with neuroscientific visuals to be less accurate than people without a degree.


## Data dictionary

The data contains:

```{r echo=FALSE, eval=TRUE}
library(tidyverse)

tibble(
  variable = names(read_csv("https://uoepsy.github.io/data/neuro_explanations.csv")),
  description = c(
    "Participant identifier",
    "Education level (ND = no degree; UG = undergraduate degree; PG = postgraduate degree)",
    "Neuroscience condition (None = no neuroscience; Desc = text description; Pic = visual presentation)",
    "Score on 10 personality items assessing Openness from the IPIP Big Five Measure (z-scored)",
    "Score on 10 personality items assessing Conscientiousness from the IPIP Big Five Measure (z-scored)",
    "Score on 10 personality items assessing Extraversion from the IPIP Big Five Measure (z-scored)",
    "Score on 10 personality items assessing Agreeableness from the IPIP Big Five Measure (z-scored)",
    "Score on 10 personality items assessing Neuroticism from the IPIP Big Five Measure (z-scored)",
    "Accuracy judgement summed across 10 explanations (0 to 100)")
) %>% knitr::kable() %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```

:::red
**THIS DOCUMENT IS ONLY AN EXAMPLE.**

**It's based on last year's report, and its purpose is to show you what the analysis task for the group report might look like.**

**You will not do this specific task.**

**The actual report task that you will do will be published in Semester 2.**
:::
